### 2012-09-17 doc/papers/renpar2013/papier.tex
Add first version of the paper to submit to renpar2013.

%===========================================================
%                              Choix: RenPar / SympA / CFSE
%===========================================================
% Une des trois options 'renpar', 'sympa' et 'cfse' doit
% être utilisée avec le style compas2013 en fonction 
% de l' l'appel à soumission choisi.
\documentclass[renpar]{compas2013}
\usepackage{todonotes}
\usepackage{listings}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{trees}
\usetikzlibrary{shadows}

%===========================================================
%                               Title
%===========================================================

\toappear{1} % Conserver cette ligne pour la version finale

\begin{document}

\title{Pythran~: Génération automatique de modules natifs parallèles à
partir de modules Python avec annotations OpenMP}
\shorttitle{Génération automatique de modules natifs parallèles pour
Python}

\author{Serge \textsc{Guelton}}%\thanks{Le texte a été relu par Thierry Gautier.}%

\address{École Normale Supérieure,\\
Département d'informatique\\
45 rue d'Ulm, F-75230 Paris Cedex 05 -- France\\
serge.guelton@telecom-bretagne.eu}

\date{6 septembre 2012}

\maketitle

%===========================================================         %
%Résumé
%===========================================================  
\begin{abstract}
  Cet article étudie l'utilisation des directives OpenMP dans le cadre de
  programmes écrits en Python. La méthode proposée est de traduire le
  code Python annoté en un code C++ conservant les annotations.

  L'approche est mise en pratique dans le compilateur pythran, qui est
  capable de compiler statiquement un module écrit dans un sous-ensemble
  statiquement typé du langage Python et annoté de directives OpenMP en un
  module natif dont certaines sections de code s'exécutent en parallèle.

  On peut ainsi générer des modules dont les temps d'exécutions sont 100
  fois inférieurs à ceux du module d'origine, grâce à la compilation
  statique et à l'utilisation du parallélisme.

  \MotsCles{Python, OpenMP, C++}
\end{abstract}

%=========================================================
\section{Introduction}
%=========================================================

Le langage Python~\cite{rossum97} est né en décembre 1989. Depuis, son
caractère extrêmement dynamique, sa bibliothèque standard riche, son large
écosystème et sa syntaxe concise lui ont permis de toucher de nombreux
domaines d'applications, dont par exemple l'administration système ou les
services webs.  Python a également percé dans le monde du calcul
scientifique~\cite{Oliphant2007}, notamment grâce au paquet
SciPy~\cite{scipy}. Le surcout d'interprétation du langage Python étant
rédhibitoire pour une utilisation dans un contexte de calcul intensif,
SciPy repose sur deux concepts pour atteindre un bon niveau de
performance~: les noyaux de calculs intensifs sont écrits dans des
langages de bas niveau, C et Fortran, dont les appels sont encapsulés par
un module dit \emph{natif}~; et la structure de données de base, le
tableau multidimensionnel~\cite{numpyarray2011} a été conçue pour être
accessible à la fois par les modules natifs et l'interpréteur Python sans
surcout de conversion.

Pour faciliter l'écriture par l'utilisateur de modules natifs, c'est à
dire sans avoir à écrire soi-même le code permettant de traduire des
objets Python en structures natives et d'appeler une fonction C depuis
l'interpréteur, le paquet SciPy fournit le module \texttt{weave} qui
permet d'intégrer de petits fragments de code C à un module python, ce
code étant par la suite compilé et chargé dynamiquement à l'exécution,
pour des temps d'exécution moindres.

Cette approche hybride, consistant à mélanger code interprété et code
compilé dans un même flot d'exécution, est présente sous de nombreuses
formes dans le paysage Python. La Section~\ref{sec:python-hybrid} étude
les différentes approches qui ont été explorées et montre le manque de
support du parallélisme dans ces approches. La
Section~\ref{sec:python-parallelism} fournit des éléments d'explications à
ce phénomène et propose l'ajout du support des directives OpenMP au langage
dans le cadre d'un \emph{domain specific language} embarqué dans Python et
compilable statiquement, Pythran. Ce langage et les contraintes qu'il doit
respecter pour être compilable statiquement et compatible avec la norme
OpenMP sont exposés à la Section~\ref{sec:pythran}. Un compilateur pour ce
langage a été écrit et a permis de mesurer des temps d'exécutions
significativement inférieur à ceux obtenus avec l'implémentation de
référence de l'interpréteur Python, comme présenté en
Section~\ref{sec:validation}.

%=========================================================
\section{Systèmes hybrides en Python}\label{sec:python-hybrid}
%=========================================================

Dans le cadre qui nous intéresse, un système hybride où une partie du code
spécifié par l'utilisateur est exécutée par l'interpréteur Python et une
partie est exécutée nativement, comme introduit dans~\cite{dongara2007}.

Il est possible d'écrire des modules natifs directement en langage C/C++.
Il faut dans ce cas respecter l'interface fournie dans~\cite{pythoncapi},
ce qui conduit à l'écriture d'une quantité non négligeable de code
d'encapsulation. Des outils ont été développé pour réduire ce coût de
développement, notamment Swig~\cite{swig2003} qui utilise un fichier
d'interface spécifique pour guider la traduction, et
\texttt{boost::python}~\cite{boostpython2007} qui se base lui sur une
le langage C++ pour guider la traduction.

Une approche concurrente est de se baser sur le langage hôte --- ici
Python --- pour décrire les deux parties du système, et de laisser un
outil tierce convertir une partie de l'application en code natif,
généralement la partie la plus demandeuse en ressources de calcul.
L'avantage est de permettre à des développeurs qui ne seraient pas
familiers avec les langages C/C++ ou qui n'auraient pas de temps à
consacrer à écrire une partie de leur application dans ces langages,
d'obtenir des performances satisfaisantes avec un investissement moindre.
Cette approche a fait l'objet de nombreuses études et solutions
techniques, que l'on peut classifier suivant leur caractère intrusif ou
non et sur le support de toute ou partie des fonctionnalités du langage.

\subsection{Approches intrusives}

Un des points contraignants du langage Python est son typage structurel
implicite. Il implique par exemple que chaque appel de méthode est résolu
dynamiquement. Il n'est alors pas surprenant que de nombreuses approches
s'attachent à l'annotation de code Python pour y apporter une couche de
typage statique, mais aussi pour typer les types flottants en entier plus
finement que les trois options fournies par Python, entiers 64 bits,
entiers multi-précision et nombres flottants double précision. Ainsi
cython~\cite{cython2010} propose un dialecte de Python permettant de
convertir un module Python fortement annoté à l'aide de types C ou d'appel
de fonctions natives. plw~\cite{dongara2007} propose une approche
similaire, en limitant la conversion à des portions de code délimitées par
des annotations, et en générant du C++. Le projet
numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
fonctionne sur le même principe, mais en générant du code LLVM.

\subsection{Approches non-intrusives}

Le problème des approches intrusives et qu'elles impliquent de modifier du
code existant de façon non bénigne, et demandent l'investissement dans un
nouveau langage, investissement dont la pérennité n'est pas toujours
garantie. \emph{A contrario} les approches non-intrusives, si elles
contraignent éventuellement le langage d'entrée, n'ajoutent pas de
nouvelles constructions et restent compatibles avec l'interpréteur

L'interpréteur PyPy~\cite{pypy2009}, lui-même écrit en Python, vise à
supporter la totalité du langage Python. Il utilise un JIT traceur qui
génère du code natif ---sans passer par un compilateur intermédiaire ---
pour les points chauds.

Le compilateur shedskin~\cite{shedskin2006} traduit du Python en C++. Il
pose néanmoins plusieurs contraintes sur le code d'entrée~: absence de
n-uplets, pas de variables polymorphes et pas d'introspection. De nombreux
modules de la bibliothèque standard sont supportés. La traduction se fait
pour un programme complet ou par module.

Le compilateur copperhead~\cite{copperhead2011} traduit quant à lui un DSL
embarqué dans Python, se limitant ainsi à un sous-ensemble en assignation
unique autorisant tableaux et n-uplets et restreignant énormément le flot
de contrôle (ni boucles, ni exceptions\dots). En contrepartie, ce DSL est
automatiquement parallélisé et traduit en CUDA ou C++ avec appels à la
bibliothèque Thrust~\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}}. La traduction se fait dynamiquement
pour les fonctions identifiées par un décorateur spécifique.

Seul copperhead permet de générer des modules natifs parallèles, mais
l'expressivité de ce langage ne lui permet que de toucher une petite
classe d'applications. La section suivante étudie les limitations de
Python vis-à-vis du parallélisme et conduit à une proposition originale
visant à pallier ses limitations.

%=========================================================
\section{Python et Parallélisme}\label{sec:python-parallelism}
%=========================================================

La parallélisation de langage de scripts est une approche naturelle pour
compenser leur manque relatif de performance. L'étude menée
dans~\cite{choy05} montre les nombreuses approches explorées autour du
langage MATLAB. Une approche purement automatique pour le langage R est
décrite dans~\cite{mals07}. Peux de travaux traitent du langage Python et
de ses spécificités.

\subsection{Python et processus lourds}

La parallélisation de calculs est supportée en Python à travers le module
\texttt{multiprocessing} fourni dans la bibliothèque standard.  Il permet
de piloter différent processus fils en mode client-serveur. La
communication entre processus se fait par queue ou pile en utilisant la
sérialisation native des objets. Il est également possible de partager des
segments mémoire. Cette approche est adaptée pour du parallélisme gros
grains, mais est moins importante qu'une approche basée sur des processus
légers communiquant par mémoire partagée pour du parallélisme à grain fin.

\subsection{Python et processus légers}

Le langage Python offre la capacité de démarrer plusieurs processus légers
dans le même interpréteur. Cependant, ce type de parallélisation n'est pas
adapté au calcul intensif en raison d'un mécanisme propre à CPython,
l'implémentation de référence de l'interpréteur: le \emph{Global
Interpreter Lock}~\cite{gil2012}~\footnote{D'autres implémentations, comme
IronPython et Jython, ne souffrent pas de cette limitation.}. Ce verrou
est utilisé, entre autres, par le ramasse-miette et il est pris à chaque
appel de méthode d'un objet.  Certains modules reposent également sur ce
verrou pour garantir certains invariants. Ce verrou rend impossible
l'utilisation de processus légers pour du calcul scientifique.

\subsection{Proposition}

Les développeurs Python désirant utiliser des machines multi-cœurs
utilisent souvent des modules natifs au sein desquels ils peuvent utiliser
des processus légers sans être contraints par le GIL. On obtient ainsi le
double intérêt d'un module natif et d'un module parallèle.

Cet article propose d'automatiser cette approche en~:
\begin{enumerate}
  \item donnant la possibilité d'annoter du code Python avec des
	directives OpenMP~\cite{openmp3.1};
  \item convertissant ce code Python en code C++11~\cite{isocxx11} en conservant la
	sémantique des annotations.
\end{enumerate}

%=========================================================
\section{Pythran}\label{sec:pythran}
%=========================================================

\subsection{Support d'OpenMP en Python}
\label{sec:python-openmp}

\subsubsection{Ajout des annotations}

Le langage Python permet de faire porter des annotations par des
fonctions, sous la forme de décorateurs. Aucun mécanisme n'est prévu pour
décorer des instructions. Une approche possible est d'utiliser des
commentaires, à la manière de l'intégration d'OpenMP dans le langage
FORTRAN. Cependant, le module \texttt{ast} de la bibliothèque standard,
qui fournit une API pour l'analyse lexicale et syntaxique du langage, ne
conserve pas les commentaires. Une alternative satisfaisante consiste
alors à interpréter les instructions ne comprenant qu'une expression de type
chaîne de caractères comme une annotation portant sur l'instruction
suivant directement. Par exemple, on pourra écrire~:

\begin{lstlisting}[language=python]
"omp parallel for reduction(r), private(auto)"
for auto in mon_conteneur:
  r+=x
\end{lstlisting}

Il est dangereux de représenter les annotations sous forme de chaîne de
caractères, car elles font éventuellement référence à des variables du
programme, par exemple \texttt{auto} dans l'exemple précédent, qui peuvent
être amenés à être renommée dans le flot de compilation.

L'arbre de syntaxe abstrait fourni par le module \texttt{ast} est alors
étendu en ajoutant un attribut \texttt{annotation} à toutes les
instructions~:

\begin{lstlisting}
annotation(str description, Name* identifiers)
\end{lstlisting}

\noindent où \texttt{description} est une chaîne de caractère décrivant
l'annotation et paramétrée par la liste d'indentifiants
\texttt{identifiers}. Le type \texttt{Name} est la façon canonique de
représenter un identifiant dans le module \texttt{ast}.  En opérant de la
sorte, toutes les opérations de renommage pourront utiliser un visiteur
générique spécialisé pour le type \texttt{Name}.

\subsubsection{Spécificités liées à Python}

À la manière de FORTRAN, le langage Python ne permet pas de créer des
blocs structurés sur lesquels l'on pourrait faire porter une annotation.
Le problème est résolu en FORTRAN en utilisant des commentaires pour
délimiter les débuts et fin de bloc. Cette solution ne se conforme pas aux
canons du langage, on lui préfèrera l'extraction du bloc concerné dans une
nouvelle fonction.

Il n'existe en Python que des variables locales à une fonction, non pas à
un bloc. Cela à une influence sur le mode de partage des variables à
l'intérieur, par exemple, d'une boucle parallèle. Dans le code suivant~:

\begin{lstlisting}[language=python]
for x in mon_conteneur:
  f=foo(x)
  b=bar(x)
  r+=f*b
\end{lstlisting}

\noindent l'instruction \texttt{for} référence les variables \texttt{x},
\texttt{mon\_conteneur}, \texttt{f}, \texttt{b} et \texttt{r} en plus des
appels de fonction. Le mode de partage par défaut de ses variables est
\texttt{shared}, il faut donc lister \emph{toutes} les variables privées
dans une clause \texttt{private} pour obtenir un bon degré de parallélisme.

À titre d'exemple, le Listing~\ref{lst:pythran-mat-mul} illustre une
multiplication de matrices annotées par des directives OpenMP.

\begin{lstlisting}[language=python, label={lst:pythran-mat-mul},caption={multiplication de matrice avec
  annotations OpenMP}]
def zero(n,m): return [[0 for row in xrange(n)] for col in xrange(m)]
def matrix_multiply(m0, m1):
  new_matrix = zero(len(m0),len(m1[0]))
  "omp parallel for private(i,j,k)"
  for i in xrange(len(m0)):
    for j in xrange(len(m1[0])):
      for k in xrange(len(m1)):
        new_matrix[i][j] += m0[i][k]*m1[k][j]
  return new_matrix
\end{lstlisting}

\subsection{Organisation de Pythran}

Le fonctionnement du compilateur Pythran est détaillé à la
Figure~\ref{fig:pythran}. L'outil convertit un module python complet,
éventuellement annoté de directives OpenMP et avec des indications de
types pour les fonctions exportées par le module. Il transforme le tout en
code C++11 annoté, qui, lié avec un environnement d'exécution nommé
\texttt{pythonic++} et l'encapsulation fournie par \texttt{boost::python},
conduit à un module natif.

\begin{figure}
  \centering
  \begin{tikzpicture}[
	file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center},
	tool/.style={draw=black!50,fill=black!10,circle, align=center}]
	\node[file] (python) {Module Python [\texttt{.py}]};
	\node[file] (annotation)	   [right=of python] {signature typée};
	\node[file] (omp)	   [left=of python] {annotations //};
	\node[tool] (pythran) [below=of python] {Pythran};
	\node[file] (cxx) [below=of pythran] {C++};
	\node[file] (boost) [left=of cxx] {\texttt{boost::python}};
	\node[file] (pythonic) [right=of cxx] {\texttt{pythonic++}};
	\node[tool] (gxx) [below=of cxx] {g++};
	\node[file] (so) [below=of gxx] {Module natif [\texttt{.so}]};

	\draw[very thick, ->] (python) -- (pythran);
	\draw[very thick,dashed, ->] (annotation) -- (pythran);
	\draw[very thick,dashed, ->] (omp) -- (pythran);
	\draw[very thick, ->] (pythran) -- (cxx);
	\draw[very thick, ->] (cxx) -- (gxx);
	\draw[very thick, ->] (boost) -- (gxx);
	\draw[very thick, ->] (pythonic) -- (gxx);
	\draw[very thick, ->] (gxx) -- (so);
  \end{tikzpicture}

  \caption{Fonctionnement schématique du compilateur Pythran}
  \label{fig:pythran}
\end{figure}

Le flot de compilation utilisé en interne par Pythran peut être modélisé de la façon
suivante~:

\begin{enumerate}
  \item\label{enum:lex-yacc} analyse lexicale \& syntaxique~;
  \item\label{enum:openmp} intégration des directives OpenMP~;
  \item\label{enum:simplification-ast} simplification de l'arbre de syntaxe abstrait~;
  \item\label{enum:typage-ast} typage paramétrique de l'arbre de syntaxe abstrait~;
  \item\label{enum:instanciation-types} instanciation des types paramétriques~;
  \item\label{enum:traduction-cxx} traduction de la représentation Python en une représentation C++~;
  \item\label{enum:compilation-extrene} compilation par un compilateur externe en une bibliothèque
	partagée.
\end{enumerate}

Ces différentes étapes sont décrites par la suite, à l'exception de
l'item~\ref{enum:lex-yacc}, et de l'item~\ref{enum:openmp} déjà discuté en
Section~\ref{sec:python-openmp}.

\subsubsection{Simplification de la grammaire}

Le langage Pythran supporte un sous ensemble de la grammaire du langage
Python. Les mots clés \texttt{class} et \texttt{exec} ne sont pas
supportés~\footnote{D'autres constructions, telles les variables globales
ou les gestionnaires de contexte à l'aide du mot-clé \texttt{with} ne sont
pas supportés, mais cela est plus lié à des priorités de développement
qu'à des problèmes fondamentaux.}. L'absence de classe utilisateur
simplifie l'inférence de type et la gestion de la mémoire. Plusieurs
classes natives, notamment tous les conteneurs, sont supportées.
L'évaluation dynamique de chaîne de caractère comme code source ne rentre
pas dans les possibilités du langage cible. De même, certaines fonction
intrinsèques telles \texttt{eval} ne sont pas supportées. Le langage
intermédiaire utilisé par PyPy, RPython~\cite{rpython2007}, est soumis à
des contraintes de staticité similaires.

Une fois que la validité de la syntaxe est assurée, l'arbre de syntaxe
abstrait est transformé en un nouvel arbre.
\begin{enumerate}

  \item Les fonctions anonymes sont transformées en fonctions imbriquées~:
	\texttt{lambda x:a*x} devient \texttt{def foo(x): return a*x}.

  \item Les fonctions imbriquées, utilisant des fermetures, sont
	transformées en fonctions globales prenant leur contexte en
	paramètre~: \texttt{def foo(x): return a*x} devient \texttt{def
	foo(a,x): return a*x}.

  \item Les listes/ensembles/dictionnaires en compréhension sont
	transformés en appels de fonction faisant apparaître explicitement la
	structure de boucle~:\texttt{[x*x for x in l]} devient
\begin{lstlisting}[language=python]
def foo(y):
	tmp=list()
	for x in y: tmp.append(x*x)
	return tmp
foo(l)
\end{lstlisting}

  \item Les attributs de classe sont transformés en indices constants~:
	\texttt{c.real+c.imag} devient \texttt{c[0]+c[1]}.

  \item Les appels de méthodes sont transformés en appels de fonctions~:
	\texttt{l.append(v)} devient \texttt{any.append(l,v)}.

  \item Les valeurs de retour implicites sont rendus
	explicites~\footnote{Toute fonction Python a une valeur de retour, qui
	se trouve être \texttt{None} par défaut.}:\texttt{def foo(): print
	"bar"} devient texttt{def foo(): print "bar" ; return None}.

  \item Tous les symboles importés de manière relative sont remplacés par
	leur chemin absolus~: \texttt{from math import cos as COS ; COS(3)}
	devient \texttt{import math ; math.cos(3)}.

\end{enumerate}

Ces étapes de raffinement conduisent à une représentation interne à
Pythran plus concise et facile à manipuler que l'arbre de syntaxe abstrait
de Python.

Un point crucial à respecter lors de ces transformations d'arbre est de
garantir la validité de ces transformations vis-à-vis des annotations
OpenMP portées par les instructions. Comme le compilateur Pythran n'a pas
de compréhension directe du sens de l'annotation, il garantit qu'une
annotation est toujours attachée à la même instruction, et qu'au cas où
une instruction devait être transformée en plusieurs, l'annotation serait
portée par le bloc d'instructions crée. Par exemple~:
%
\begin{lstlisting}[language=python]
"omp critical"
a,b=make_tuple()
\end{lstlisting}
%
\noindent peut être transformé en
\begin{lstlisting}[language=python]
"omp critical"
if 1:
	t=make_tuple()
	a=t[0] ; b=t[1]
\end{lstlisting}
%
\noindent afin de conserver l'annotation.

\subsubsection{Typage}

Pythran utilise une variante de l'algorithme d'inférence de type de
Hindley--Milner~\cite{milner78}. À l'issu de son application, tous les
nœuds de la représentation interne se voient dotés d'un type paramétrique,
dépendant du type symbolique de ses paramètres.

Cet article ne vise pas à expliquer en détail le fonctionnement de cette
étape. On admettra donc que cette analyse permet à Pythran d'accepter des
fonctions polymorphes, mais uniquement des variables monomorphes.

Pour faciliter la gestion de la mémoire, on ajoute une nouvelle
contrainte~: il ne peut y avoir de dépendance circulaire dans la
définition des types des objets. Ce genre de dépendance ne pause aucun
problème en Python, et demande l'utilisation de pointeurs en C/C++.

\subsubsection{Gestion de la mémoire}

En Python, une instance de classe est accessible via sa référence. Une
affectation est une copie de référence, et non pas de l'objet lui-même.
L'implémentation de référence, CPython, gère la mémoire en utilisant deux
stratégies~:compteurs de référence et ramasse-miette.

La première stratégie peut être employée --- avec un certain coût --- en
C++ par l'utilisation de pointeurs partagés, les \texttt{std::shared\_ptr}
disponible dans la bibliothèque standards.  Grâce à la contrainte sur
l'absence de dépendances de type circulaires, le comptage de référence est
un mécanisme suffisant pour garantir l'absence de fuites de mémoires. De
plus, il garantit une libération «~au plus tôt~» des ressources.

Cependant, l'utilisation de compteurs partagés dans un contexte parallèle,
implique de prendre un certain nombre de précautions. Pythran utilise pour
le moment un simple verrou, mais des techniques plus performantes ont été
développées par ailleurs~\cite{Levanoni2006}.

%=========================================================
\section{Validation et expériences}\label{sec:validation}
%=========================================================

\subsection{Validation}

La vérification de la validité de la traduction effectuée par Pythran
repose sur des test unitaires, constitués d'une fonction et de ses
arguments.  Cette fonction est généralement courte, elle retourne une
valeur servant de trace et est déterministe.

Pour chaque test unitaire, la fonction correspondante est exécutée par
l'interpréteur avec les arguments fournis et sa valeur de retour est
mémorisée. La même fonction est alors compilée par Pythran en un module
natif qui est ensuite chargé dynamiquement dans l'interpréteur pour être
exécuté avec les mêmes arguments pour récupérer une nouvelle trace. Le
test est considéré comme valide si les deux traces sont identiques. On
garantit ainsi le même comportement entre les deux versions.

En plus des test unitaires, des applications complètes sont utilisées pour
comparer les performances du code compilé à celles de l'interpréteur.

\subsection{Mesures de performances}

Les mesures faites ici comparent les temps d'exécution de plusieurs
applications exécutées par les interpréteurs CPython ou par PyPy, ou après
transformation en un module natif par les traducteurs Shedskin et Pythran.

Les codes utilisés sont tirés des bases de tests du défunt projet
\texttt{unladen swallow}, ou issus de la littérature. La seule
modification faite est l'ajout de la signature typée de la fonction
appelée. Quand cela était possible, des annotations OpenMP ont été
ajoutée. \texttt{extrema} recherche les valeurs extrêmes d'une liste,
\texttt{matmul} effectue une produit matriciel entre deux matrices
carrées, \texttt{mandel} fait un calcul de fractal, \texttt{nqueens}
résout une instance du problème des n-reines, \texttt{pi\_buffon} utilise
une méthode de monte-carlo pour calculer les décimales de $\pi$,
\texttt{quicksort} est une implémentation récursive du tri rapide et
\texttt{stone} est l'adaptation en Python du vénérable test de
performances \texttt{wetstone}.

Toutes les mesures de temps sont faites à l'aide du module
\texttt{timeit}. Elles représentent le temps d'exécution médian sur 100
itérations. Les mesures \emph{incluent} le coût de conversion entre les
structures de données Python et leur représentation native.

Les évaluations de performances sont effectuées sur une machine comportant
$X$ processeurs de type $Y$, avec $Z$G de RAM et tournant un noyau $X.Y$.
Les modules Pythran sont compilés avec g++-X-Y et les drapeaux
\texttt{-Ofast -DNDEBUG}.
Les temps reproduits ici sont obtenus à l'aides des commandes
\texttt{python setup.py bench} et \texttt{python setup.py bench
--parallel} fournie avec l'installation de Pythran.

Le tableau~\ref{tbl:pythran-vs-cpython} illustre les performances de
Pythran par rapport à CPython, en prenant en compte ou non les directives
OpenMP. Il permet d'observer des gains importants dus au changement de
langage. On observe aussi que les directives OpenMP sont bien prises en
compte et permettent d'obtenir des facteurs d'accélération du module natif
entre X.Y et Z.W.

%=========================================================
\section{Conclusion}
%=========================================================

Cet article présente l'ajout du support OpenMP au langage Python dans le
cadre du traducteur Python vers C++ Pythran. Partant d'un large
sous-ensemble du langage, il propose une syntaxe simple pour annoter des
instructions Python, et un processus de traduction conservatif pour
traduire ses annotations en directives pour le langage C++. Le code généré
repose sur une bibliothèque paramétrique compatible avec une utilisation
en mode parallèle.

Le code généré est évalué pour ses versions séquentielles et parallèles. La
comparaison avec des approches séquentielles existantes montre des gains
de $\times XY$ à $\times YZ$ très encourageants.

%=========================================================
\section{Remerciements}
%=========================================================

L'auteur tient à remercier la société Silkan pour avoir financé une partie des
développements du compilateur Pythran. Il remercie également Albert
\textsc{Cohen} pour lui avoir libéré du temps pour travailler sur ce projet.

Plusieurs discussions avec Pierre \textsc{Fiorini}, Bryan
\textsc{Catanzero}, Joël \textsc{Falcou} et Mathias \textsc{Gaunard} ne
sont pas étrangères aux idées exposées dans cet article.

Les nombreux conseils de Fabien \textsc{Dagnat} ont fortement contribué à
la réflexion sur le typage statique de code Python. Deux élèves de Télécom
Bretagne, Adrien \textsc{Merlini} et Pierrick \textsc{Brunet}, ont apporté
une aide inestimables pour le support des listes, des exceptions et des
ensembles.

\bibliography{biblio}

\end{document}

### 2012-09-18 doc/papers/renpar2013/papier.tex
Update the paper with Pierrick's comments.

@@
 d'interface spécifique pour guider la traduction, et
-\texttt{boost::python}~\cite{boostpython2007} qui se base lui sur une
-le langage C++ pour guider la traduction.
+\texttt{boost::python}~\cite{boostpython2007} qui se base lui sur le
+langage C++ pour guider la traduction.
 
@@
 langage MATLAB. Une approche purement automatique pour le langage R est
-décrite dans~\cite{mals07}. Peux de travaux traitent du langage Python et
+décrite dans~\cite{mals07}. Peu de travaux traitent du langage Python et
 de ses spécificités.
@@
 	se trouve être \texttt{None} par défaut.}:\texttt{def foo(): print
-	"bar"} devient texttt{def foo(): print "bar" ; return None}.
+	"bar"} devient \texttt{def foo(): print "bar" ; return None}.
 
@@
 dépendant du type symbolique de ses paramètres.
+
+L'instanciation des types paramétriques est déclenchée par une annotation
+fournie par l'utilisateur, qui spécifie la signature des fonctions
+exportées par le module sous la forme~:
+%
+\begin{lstlisting}[language=python]
+#pythran export matrix_multiply(int list list, int list list)
+#pythran export matrix_multiply(float list list, float list list)
+\end{lstlisting}
+%
+Plusieurs annotations pour la même fonction peuvent coexister~: plusieurs
+fonctions sont alors générées.
 

### 2012-09-18 doc/papers/renpar2013/papier.tex
Update renpar paper with real benchmark values.

@@
 Les évaluations de performances sont effectuées sur une machine comportant
-$X$ processeurs de type $Y$, avec $Z$G de RAM et tournant un noyau $X.Y$.
-Les modules Pythran sont compilés avec g++-X-Y et les drapeaux
+4 processeurs de type Intel Core i7 avec l'\emph{hypethreading}, avec 8G
+de RAM et tournant un noyau 3.0.0.
+Les modules Pythran sont compilés avec g++-4.7.1 et les drapeaux
 \texttt{-Ofast -DNDEBUG}.
@@
 compte et permettent d'obtenir des facteurs d'accélération du module natif
-entre X.Y et Z.W.
+parallèle allant jusqu'à $\times300$ par rapport à CPython et $\time8$ par
+rapport à PyPy.
+
+\begin{table}
+  \begin{tabular}{|l|c|c|c|c|}
+				&CPython	&	PyPy	&	Pythran	&	Pythran+OpenMP\\
+	nqueens		&1.754		&	0.603	&	0.228	&	0.231		  \\
+	pi\_buffon	&47.083		&	12.08	&	2.676	&	0.561		\\
+	extrema		&0.386		&	0.564	&	0.033	&	0.033\\
+	stone		&6.956		&	0.204	&	0.023	&	0.023\\
+	mandel		&976.81		&	6.304	&	5.563	&	2.565\\
+	caxpy		&0.588		&	0.204	&	0.340	&	0.340\\
+	matmul		&153.91		&	9.061	&	7.314	&	2.263\\
+	quicksort	&3.331		&	0.087	&	0.040	&	0.041\\
+  \end{tabular}
+
+  \caption{Comparaison de CPython, PyPy et Pythran, avec ou sans OpenMP.}
+  \label{tbl:pythran-vs-cpython}
+\end{table}
 

### 2012-09-20 doc/papers/renpar2013/papier.tex
Take into account Fabien Dagnat's comments.

--- 2012-09-18 doc/papers/renpar2013/papier.tex
@@
 
-\title{Pythran~: Génération automatique de modules natifs parallèles à
-partir de modules Python avec annotations OpenMP}
-\shorttitle{Génération automatique de modules natifs parallèles pour
-Python}
+\title{Pythran~: Génération automatique de modules natifs parallèles pour
+Python avec annotations OpenMP}
+
+\shorttitle{Génération automatique de
+modules natifs parallèles pour Python}
 
@@
 
-  On peut ainsi générer des modules dont les temps d'exécutions sont 100
-  fois inférieurs à ceux du module d'origine, grâce à la compilation
+  On peut ainsi générer des modules dont les temps d'exécutions sont
+  jusqu'à 300 fois inférieurs à ceux du module d'origine, grâce à la compilation
   statique et à l'utilisation du parallélisme.
@@
 domaines d'applications, dont par exemple l'administration système ou les
-services webs.  Python a également percé dans le monde du calcul
+services web.  Python a également percé dans le monde du calcul
 scientifique~\cite{Oliphant2007}, notamment grâce au paquet
@@
 SciPy repose sur deux concepts pour atteindre un bon niveau de
-performance~: les noyaux de calculs intensifs sont écrits dans des
-langages de bas niveau, C et Fortran, dont les appels sont encapsulés par
-un module dit \emph{natif}~; et la structure de données de base, le
-tableau multidimensionnel~\cite{numpyarray2011} a été conçue pour être
-accessible à la fois par les modules natifs et l'interpréteur Python sans
-surcout de conversion.
+performance. D'une part, les noyaux de calculs intensifs sont écrits dans
+des langages de bas niveau, C et Fortran, dont les appels sont encapsulés
+par un module dit \emph{natif}. D'autre part, la structure de données de
+base, le tableau multidimensionnel~\cite{numpyarray2011} a été conçue pour
+être accessible à la fois par les modules natifs et l'interpréteur Python
+sans surcout de conversion.
 
-Pour faciliter l'écriture par l'utilisateur de modules natifs, c'est à
-dire sans avoir à écrire soi-même le code permettant de traduire des
-objets Python en structures natives et d'appeler une fonction C depuis
-l'interpréteur, le paquet SciPy fournit le module \texttt{weave} qui
+Pour faciliter l'écriture par l'utilisateur de modules natifs, \emph{i.e.\
+} sans avoir à écrire soi-même le code permettant de traduire des objets
+Python en structures natives et d'appeler une fonction C depuis
+l'interpréteur, le paquet SciPy fournit le module \texttt{weave}. Il
 permet d'intégrer de petits fragments de code C à un module python, ce
@@
 compilé dans un même flot d'exécution, est présente sous de nombreuses
-formes dans le paysage Python. La Section~\ref{sec:python-hybrid} étude
+formes dans le paysage Python. La Section~\ref{sec:python-hybrid} étudie
 les différentes approches qui ont été explorées et montre le manque de
 support du parallélisme dans ces approches. La
-Section~\ref{sec:python-parallelism} fournit des éléments d'explications à
-ce phénomène et propose l'ajout du support des directives OpenMP au langage
-dans le cadre d'un \emph{domain specific language} embarqué dans Python et
-compilable statiquement, Pythran. Ce langage et les contraintes qu'il doit
-respecter pour être compilable statiquement et compatible avec la norme
-OpenMP sont exposés à la Section~\ref{sec:pythran}. Un compilateur pour ce
-langage a été écrit et a permis de mesurer des temps d'exécutions
-significativement inférieur à ceux obtenus avec l'implémentation de
-référence de l'interpréteur Python, comme présenté en
+Section~\ref{sec:python-parallelism} fournit des éléments d'explication à
+ce phénomène et propose l'ajout du support des directives OpenMP au
+langage dans le cadre d'un \emph{domain specific language} embarqué dans
+Python et compilé statiquement, Pythran. Ce langage et les contraintes
+qu'il doit respecter pour être compatible avec
+la norme OpenMP sont exposés à la Section~\ref{sec:pythran}. Un
+compilateur pour ce langage a été écrit et a permis de mesurer des temps
+d'exécutions significativement inférieurs à ceux obtenus avec
+l'implémentation de référence de l'interpréteur Python, comme présenté en
 Section~\ref{sec:validation}.
@@
 outil tierce convertir une partie de l'application en code natif,
-généralement la partie la plus demandeuse en ressources de calcul.
+généralement la partie la plus gourmande en ressources de calcul.
 L'avantage est de permettre à des développeurs qui ne seraient pas
@@
 numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
-fonctionne sur le même principe, mais en générant du code LLVM.
+fonctionne sur le même principe, mais en générant du code LLVM.  Le
+problème des approches intrusives et qu'elles impliquent de modifier du
+code existant de façon non triviale, et demandent l'investissement dans un
+nouveau langage, investissement dont la pérennité n'est pas toujours
+garantie.
 
@@
 
-Le problème des approches intrusives et qu'elles impliquent de modifier du
-code existant de façon non bénigne, et demandent l'investissement dans un
-nouveau langage, investissement dont la pérennité n'est pas toujours
-garantie. \emph{A contrario} les approches non-intrusives, si elles
-contraignent éventuellement le langage d'entrée, n'ajoutent pas de
-nouvelles constructions et restent compatibles avec l'interpréteur
+Les approches non-intrusives, si elles contraignent éventuellement le
+langage d'entrée, n'ajoutent pas de nouvelles constructions et restent
+compatibles avec l'interpréteur standard.
 
@@
 \texttt{multiprocessing} fourni dans la bibliothèque standard.  Il permet
-de piloter différent processus fils en mode client-serveur. La
+de piloter différents processus fils en mode client-serveur. La
 communication entre processus se fait par queue ou pile en utilisant la
 sérialisation native des objets. Il est également possible de partager des
-segments mémoire. Cette approche est adaptée pour du parallélisme gros
-grains, mais est moins importante qu'une approche basée sur des processus
+segments mémoires. Cette approche est adaptée pour du parallélisme gros
+grains, mais est moins utile qu'une approche basée sur des processus
 légers communiquant par mémoire partagée pour du parallélisme à grain fin.
@@
 IronPython et Jython, ne souffrent pas de cette limitation.}. Ce verrou
-est utilisé, entre autres, par le ramasse-miette et il est pris à chaque
+est utilisé, entre autre, par le ramasse-miette et il est pris à chaque
 appel de méthode d'un objet.  Certains modules reposent également sur ce
@@
 
+Pythran est un traducteur d'un sous-ensemble du langage Python vers le
+langage C++. Cette section étudie la chaine de compilation utilisée.
+
 \subsection{Support d'OpenMP en Python}
@@
 
-Le langage Python permet de faire porter des annotations par des
-fonctions, sous la forme de décorateurs. Aucun mécanisme n'est prévu pour
-décorer des instructions. Une approche possible est d'utiliser des
-commentaires, à la manière de l'intégration d'OpenMP dans le langage
-FORTRAN. Cependant, le module \texttt{ast} de la bibliothèque standard,
-qui fournit une API pour l'analyse lexicale et syntaxique du langage, ne
-conserve pas les commentaires. Une alternative satisfaisante consiste
-alors à interpréter les instructions ne comprenant qu'une expression de type
-chaîne de caractères comme une annotation portant sur l'instruction
-suivant directement. Par exemple, on pourra écrire~:
+Le langage Python permet d'ajouter des annotations par des fonctions, sous
+la forme de décorateurs. Aucun mécanisme n'est prévu pour décorer des
+instructions. Une approche possible est d'utiliser des commentaires, à la
+manière de l'intégration d'OpenMP dans le langage FORTRAN. Cependant, le
+module \texttt{ast} de la bibliothèque standard, qui fournit une API pour
+l'analyse lexicale et syntaxique du langage, ne conserve pas les
+commentaires. Une alternative satisfaisante consiste alors à interpréter
+les instructions ne comprenant qu'une expression de type chaîne de
+caractères comme une annotation portant sur l'instruction suivant
+directement. Par exemple, on pourra écrire~:
 
@@
 caractères, car elles font éventuellement référence à des variables du
-programme, par exemple \texttt{auto} dans l'exemple précédent, qui peuvent
+programme, par exemple \texttt{auto} et \texttt{r} dans l'exemple précédent, qui peuvent
 être amenés à être renommée dans le flot de compilation.
@@
 \texttt{mon\_conteneur}, \texttt{f}, \texttt{b} et \texttt{r} en plus des
-appels de fonction. Le mode de partage par défaut de ses variables est
-\texttt{shared}, il faut donc lister \emph{toutes} les variables privées
-dans une clause \texttt{private} pour obtenir un bon degré de parallélisme.
-
-À titre d'exemple, le Listing~\ref{lst:pythran-mat-mul} illustre une
-multiplication de matrices annotées par des directives OpenMP.
+appels de fonction. La visibilité par défaut de ses variables est
+partagée, ou \texttt{shared}. Pour permettre une exécution concurrente
+valide, Il faut donc lister \emph{toutes} les variables locales
+dans une clause \texttt{private} pour obtenir un bon degré de
+parallélisme.  À titre d'exemple, le Listing~\ref{lst:pythran-mat-mul}
+illustre une multiplication de matrices annotées par des directives
+OpenMP.
 
@@
   annotations OpenMP}]
-def zero(n,m): return [[0 for row in xrange(n)] for col in xrange(m)]
 def matrix_multiply(m0, m1):
-  new_matrix = zero(len(m0),len(m1[0]))
+  new_matrix = [[0 for row in xrange(n)] for col in xrange(m)]
   "omp parallel for private(i,j,k)"
@@
   \centering
+  \scalebox{.8}{
   \begin{tikzpicture}[
@@
   \end{tikzpicture}
+  }
 
@@
 
-Le langage Pythran supporte un sous ensemble de la grammaire du langage
+Le langage Pythran supporte un sous-ensemble de la grammaire du langage
 Python. Les mots clés \texttt{class} et \texttt{exec} ne sont pas
@@
 classes natives, notamment tous les conteneurs, sont supportées.
-L'évaluation dynamique de chaîne de caractère comme code source ne rentre
-pas dans les possibilités du langage cible. De même, certaines fonction
+L'évaluation dynamique de chaînes de caractère comme code source ne rentre
+pas dans les possibilités du langage cible. De même, certaines fonctions
 intrinsèques telles \texttt{eval} ne sont pas supportées. Le langage
 intermédiaire utilisé par PyPy, RPython~\cite{rpython2007}, est soumis à
-des contraintes de staticité similaires.
+des contraintes similaires.
 
@@
 
-  \item Les attributs de classe sont transformés en indices constants~:
+  \item Les classe sont transformés en $n$-uplets~:
 	\texttt{c.real+c.imag} devient \texttt{c[0]+c[1]}.
@@
 	explicites~\footnote{Toute fonction Python a une valeur de retour, qui
-	se trouve être \texttt{None} par défaut.}:\texttt{def foo(): print
+	se trouve être \texttt{None} par défaut.}: \texttt{def foo(): print
 	"bar"} devient \texttt{def foo(): print "bar" ; return None}.
@@
   \item Tous les symboles importés de manière relative sont remplacés par
-	leur chemin absolus~: \texttt{from math import cos as COS ; COS(3)}
+	leur chemin absolu~: \texttt{from math import cos as COS ; COS(3)}
 	devient \texttt{import math ; math.cos(3)}.
@@
 une instruction devait être transformée en plusieurs, l'annotation serait
-portée par le bloc d'instructions crée. Par exemple~:
+portée par le bloc d'instructions créé. Par exemple~:
 %
@@
 %
-\noindent peut être transformé en
+\noindent peut être transformé, afin de conserver l'annotation,  en~:
 \begin{lstlisting}[language=python]
@@
 %
-\noindent afin de conserver l'annotation.
 
@@
 
-Pythran utilise une variante de l'algorithme d'inférence de type de
+Pythran utilise une variante de l'algorithme d'inférence de
 Hindley--Milner~\cite{milner78}. À l'issu de son application, tous les
@@
 contrainte~: il ne peut y avoir de dépendance circulaire dans la
-définition des types des objets. Ce genre de dépendance ne pause aucun
+définition des types des objets. Ce genre de dépendance ne pose aucun
 problème en Python, et demande l'utilisation de pointeurs en C/C++.
@@
 L'implémentation de référence, CPython, gère la mémoire en utilisant deux
-stratégies~:compteurs de référence et ramasse-miette.
+stratégies~: compteurs de références et ramasse-miette.
 
@@
 C++ par l'utilisation de pointeurs partagés, les \texttt{std::shared\_ptr}
-disponible dans la bibliothèque standards.  Grâce à la contrainte sur
-l'absence de dépendances de type circulaires, le comptage de référence est
+disponible dans la bibliothèque standard.  Grâce à la contrainte sur
+l'absence de dépendances de type circulaires, le comptage de références est
 un mécanisme suffisant pour garantir l'absence de fuites de mémoires. De
@@
 La vérification de la validité de la traduction effectuée par Pythran
-repose sur des test unitaires, constitués d'une fonction et de ses
+repose sur plus de 400 tests unitaires, constitués d'une fonction et de ses
 arguments.  Cette fonction est généralement courte, elle retourne une
-valeur servant de trace et est déterministe.
+valeur servant de trace et doit être déterministe.
 
@@
 ajoutée. \texttt{extrema} recherche les valeurs extrêmes d'une liste,
-\texttt{matmul} effectue une produit matriciel entre deux matrices
+\texttt{matmul} effectue le produit de deux matrices
 carrées, \texttt{mandel} fait un calcul de fractal, \texttt{nqueens}
@@
 Les évaluations de performances sont effectuées sur une machine comportant
-$X$ processeurs de type $Y$, avec $Z$G de RAM et tournant un noyau $X.Y$.
-Les modules Pythran sont compilés avec g++-X-Y et les drapeaux
+4 processeurs de type Intel Core i7 avec l'\emph{hypethreading}, avec 8G
+de RAM et un noyau Linux 3.0.0.
+Les modules Pythran sont compilés avec g++-4.7.1 et les drapeaux
 \texttt{-Ofast -DNDEBUG}.
-Les temps reproduits ici sont obtenus à l'aides des commandes
-\texttt{python setup.py bench} et \texttt{python setup.py bench
---parallel} fournie avec l'installation de Pythran.
+Les temps reproduits ici sont obtenus à l'aide de la commande
+\texttt{python setup.py bench} fournie avec l'installation de Pythran.
 
@@
 compte et permettent d'obtenir des facteurs d'accélération du module natif
-entre X.Y et Z.W.
+généralement de l'ordre de $\times20$ par rapport à CPython et de l'ordre
+de $\times3$ par rapport à PyPy. Les gains liés à OpenMP permettent
+d'apporter un facteur $\times2$ à $\times4$ suivant l'application. Les
+gains importants pour \texttt{matmul} et \texttt{mandel} s'expliquent par
+l'utilisation de types natifs au lieu d'objets Python pour les opérations
+de base. Le cas de \texttt{quicksort} est à relier au coût prohibitif d'un
+appel de fonction en Python. Le cas d'\texttt{extrema}, implémenté à
+l'aide des intrinsèques \texttt{zip} et \texttt{map}, montrent des
+possibilités d'améliorations non exploitées.
+
+\begin{table}
+  \begin{tabular}{|l|c|c|c|c|}
+				&CPython	&	PyPy	&	Pythran	&	Pythran+OpenMP\\
+	nqueens		&1.754		&	0.603	&	0.228	&	-		  \\
+	pi\_buffon	&47.083		&	12.08	&	2.676	&	0.561		\\
+	extrema		&0.386		&	0.564	&	0.033	&	-\\
+	stone		&6.956		&	0.204	&	0.023	&	-\\
+	mandel		&976.81		&	6.304	&	5.563	&	2.565\\
+	matmul		&153.91		&	9.061	&	7.314	&	2.263\\
+	quicksort	&3.331		&	0.087	&	0.040	&	-\\
+  \end{tabular}
+
+  \caption{Comparaison de CPython, PyPy et Pythran, avec ou sans OpenMP.}
+  \label{tbl:pythran-vs-cpython}
+\end{table}
 
@@
 Bretagne, Adrien \textsc{Merlini} et Pierrick \textsc{Brunet}, ont apporté
-une aide inestimables pour le support des listes, des exceptions et des
+une aide inestimable pour le support des listes, des exceptions et des
 ensembles.

### 2012-09-20 pythran/typing.py
Improve error reporting for infinite function recursion.

@@ -43,7 +43,8 @@ class Reorder(ast.NodeVisitor):
             if isinstance(stmt, ast.FunctionDef):
                 olddef.append(stmt)
             else: newbody.append(stmt)
-        newdef = [ f for f in nx.topological_sort(self.typedeps, ordered_global_declarations(node)) if isinstance(f, ast.FunctionDef) ]
+            try: newdef = [ f for f in nx.topological_sort(self.typedeps, ordered_global_declarations(node)) if isinstance(f, ast.FunctionDef) ]
+            except nx.exception.NetworkXUnfeasible: raise PythranSyntaxError("Infinite function recursion",stmt)
         assert set(newdef) == set(olddef)
         node.body=newbody + newdef
 

### 2012-09-25 doc/papers/renpar2013/papier.tex
Update renpar paper with Pierrick's review.

@@
 scientifique~\cite{Oliphant2007}, notamment grâce au paquet
-SciPy~\cite{scipy}. Le surcout d'interprétation du langage Python étant
+SciPy~\cite{scipy}. Le surcoût d'interprétation du langage Python étant
 rédhibitoire pour une utilisation dans un contexte de calcul intensif,
@@
 par un module dit \emph{natif}. D'autre part, la structure de données de
-base, le tableau multidimensionnel~\cite{numpyarray2011} a été conçue pour
+base, le tableau multidimensionnel~\cite{numpyarray2011}, a été conçue pour
 être accessible à la fois par les modules natifs et l'interpréteur Python
-sans surcout de conversion.
+sans surcoût de conversion.
 
@@
 ce qui conduit à l'écriture d'une quantité non négligeable de code
-d'encapsulation. Des outils ont été développé pour réduire ce coût de
+d'encapsulation. Des outils ont été développés pour réduire ce coût de
 développement, notamment Swig~\cite{swig2003} qui utilise un fichier
@@
 fonctionne sur le même principe, mais en générant du code LLVM.  Le
-problème des approches intrusives et qu'elles impliquent de modifier du
+problème des approches intrusives est qu'elles impliquent de modifier du
 code existant de façon non triviale, et demandent l'investissement dans un
@@
 \begin{table}
-  \begin{tabular}{|l|c|c|c|c|}
+	\centering
+  \begin{tabular}{l|c|c|c|c}
 				&CPython	&	PyPy	&	Pythran	&	Pythran+OpenMP\\
+	  \hline
+	  \hline
 	nqueens		&1.754		&	0.603	&	0.228	&	-		  \\
@@
 comparaison avec des approches séquentielles existantes montre des gains
-de $\times XY$ à $\times YZ$ très encourageants.
+de $\times 6$ à $\times 390$ très encourageants.
 

### 2012-09-29 doc/papers/renpar2013/papier.tex
Take into account Sven's review for renpar paper.
* Some French typo
* Annotation part was not clear

--- 2012-09-20 doc/papers/renpar2013/papier.tex
@@
 
-Dans le cadre qui nous intéresse, un système hybride où une partie du code
-spécifié par l'utilisateur est exécutée par l'interpréteur Python et une
+Dans le cadre qui nous intéresse, un système est hybride quand une partie du
+code spécifié par l'utilisateur est exécutée par l'interpréteur Python et une
 partie est exécutée nativement, comme introduit dans~\cite{dongara2007}.
@@
 
-Il est dangereux de représenter les annotations sous forme de chaîne de
-caractères, car elles font éventuellement référence à des variables du
-programme, par exemple \texttt{auto} et \texttt{r} dans l'exemple précédent, qui peuvent
-être amenés à être renommée dans le flot de compilation.
+Comme l'annotation fait référence à des variables du programme, par exemple
+\texttt{auto} et \texttt{r} dans l'exemple précédent, qui peuvent être amenés à
+être renommée dans le flot de compilation, il faut en extraire les informations contextuelles.
 
@@
 \noindent où \texttt{description} est une chaîne de caractère décrivant
-l'annotation et paramétrée par la liste d'indentifiants
-\texttt{identifiers}. Le type \texttt{Name} est la façon canonique de
-représenter un identifiant dans le module \texttt{ast}.  En opérant de la
-sorte, toutes les opérations de renommage pourront utiliser un visiteur
-générique spécialisé pour le type \texttt{Name}.
+l'annotation et paramétrée par la liste d'indentifiants \texttt{identifiers}.
+Le type \texttt{Name} est la façon canonique de représenter un identifiant dans
+le module \texttt{ast}.  En opérant de la sorte, toutes les opérations de
+renommage pourront utiliser un visiteur générique spécialisé pour le type
+\texttt{Name}. L'exemple précédent sera alors représenté par un 2-uplet
+\texttt{("omp parallel for reduction(\{0\}), private(\{1\}", [r, auto])}.
 

### 2012-11-22 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
Added the .tex of the "RapportCourt" paper.

\documentclass[a4paper]{article}

\usepackage[utf8]{inputenc}   
\usepackage[T1]{fontenc}      
\usepackage[francais]{babel}

\usepackage{url}

\usepackage{graphicx}

\usepackage[a4paper]{geometry}

\title{Pythran : Rapport court}           
\author{Alan Raynaud}
\date{}

\sloppy

\begin{document}

%Page de garde

\maketitle

\tableofcontents

\clearpage

\section{Analyse bibliographique}

\subsection{Introduction : du Python et des performances}

TODO : Langage interprété, conçu pour optimiser dev et pas execution

TODO : Question de la nécessité de la performance + temps d'execution vs temps de dev

\subsection{Optimisation dans l'utilisation de python}

\label{OptCode}

\subsubsection{Améliorer les performances en améliorant le code}

Python permet génalement de résoudre un même problème de plusieurs façons. Si le choix d'un bon algorithme est crucial, la façon dont est implémenté cet algorithme peut faire varier considérablement les performance du programme final. C'est pourquoi il existe un ensemble de bonne pratiques permettant d'améliorer son code\cite{PythonSpeed}.

Exemples :
\begin{itemize}
\item Concaténer des strings avec .join() plutôt qu'avec l'opérateur +.
\item Utiliser plusieurs affectations plutôt qu'une affectation multiple : a=x;b=y; plutôt que a,b=x,y.
\item Utiliser les opérateurs propres au langage plutôt que les fonctions lamda : operator.add plutôt que lambda x,y : x+y.
\end{itemize}

\subsubsection{Iterateurs et generateur}

Dans la boucle de l'exemple suivant, python génère la liste  [1..100] puis créé un objet iterateur qui permet de parcourir un à un les élements de cette liste.

\begin{verbatim}
for i in range(100):
  print i
\end{verbatim}

On se rend compte dans cet exemple qu'il n'était pas nécessaire d'avoir toute la liste en mémoire. En effet, un seul élement est utilisé à la fois, et chaque élément n'est utilisé qu'une seule fois.

C'est pourquoi il existe en python des générateurs. Les générateurs permettent de réaliser des iterateurs qui ne parcourent pas un ensemble, mais génèrent les éléments de l'itération au fur et à mesure. Quand on utilise un iterateur produit par un generateur, il n'y a toujours qu'une seule valeur en mémoire : celle actuellement utilisée.

Python fournit également des iterateurs se comportant comme ceux produits par des générateurs :

\begin{itemize}
  \item xrange qui se comporte comme range, mais ne créé pas de liste en mémoire
  \item les iterateurs du module itertools comme imap et product
\end{itemize}

Afin de considérer l'intérêt en terme de performance, on peut considérer la figure~\ref{GenExpMapImapPy} , qui montre les différences de temps d'exécution entre trois implémentations du calcul de la somme des carrés des N premiers entiers.

Dans chaque cas, le temps de l'execution de 100 itérations du même programme est mesuré grâce au module timeit.

\begin{verbatim}
time100 = timeit.timeit('Expression', number=100)
\end{verbatim}

\begin{figure}[h]
  \includegraphics[scale=0.5]{./Pictures/GenExpMapImapPy}
  \caption{Compaison de performances entre trois implémentations en python}
  \label{GenExpMapImapPy}
\end{figure}

La première implémentation utilise un générateur :

\begin{verbatim}
sum(x*x for x in xrange(N))
\end{verbatim}

Le générateur créé permet de ne produire les carrés qu'au fur et à mesure de la sommation. A un instant donné, il n'y a qu'un seul carré en mémoire.

La seconde implémentation utilise l'operateur imap qui applique une fonction à un iterateur pour produire un nouvel iterateur :

\begin{verbatim}
sum(imap(lambda x:x*x, xrange(N)))
\end{verbatim}

Cette implémentation est moins performante que la première car elle utilise des appels de fonctions, qui sont couteux en python.

La troisième implémentation utilise l'operateur map qui applique une fonction à un iterateur pour produire une liste :

\begin{verbatim}
sum(map(lambda x:x*x, xrange(N)))
\end{verbatim}

Dans cette implémentation, la liste des carrés est créée avant la sommation. C'est l'opération de création de l'espace mémoire nécessaire pour la stocker qui est couteuse et diminue les performance par rapport à la seconde implémentation.

\subsection{Optimisation à partir du code python}

Les méthodes d'amélioration des performances présentées en \ref{OptCode} nécessitent d'effectuer des modifications au niveau du code source. Les gains en temps d'exécution se feront donc au prix d'augmentation du temps de développement. Il existe cependant des outils qui permettent d'augmenter les performances d'un programme sans avoir à modifier le code source.

\subsubsection{Optimisation de l'interpréteur}

CPython est l'implémentation de référence du langage python. Il en existe cependant d'autre comme Jython écrit en Java et IronPython , l'implémentation de python pour .NET. Le projet Pypy, quant à lui, a pour but d'obtenir une implémentation de python plus performante que CPython.

Un programme écrit en python pour CPython peut être exécuté avec Pypy sans modification\footnote{Il existe quelques incompatibilités dans des cas particuliers, souvent liées à des différence de stratégie des ramasse-miettes\cite{PypyDiff}.} et sera en moyenne 5,5 fois plus rapide\cite{PypySpeed}.

\subsubsection{La solution pythran : compiler en code natif}

L'outils pythran permet de traduire un module python en C++ pour pouvoir ensuite être compilé en code natif. Afin d'être utilisable par pythran, le code source doit seulement être annoté de descriptions d'interfaces. Le résultat est un module python en code natif qui peut être utilisé depuis un programme python usuel.

Avant même de traduire le code python en C++, pythran se charge de l'optimiser. Par exemple, il va analyser quels sont les expressions constantes et les remplacer par leurs valeurs.

Pythran ne traduit qu'un sous-ensemble de python. En particulier, les classes ne sont pas gérées. Cela destine pythran à être utilisé pour l'optimisation d'algorithmes dont les temps d'exécution sont critiques pour le programme les utilisant.

La figure~\ref{RangeXrangePyCpp} permet de comparer les performances entre un programme en python et son équivalent en C++.

\begin{figure}[h]
  \includegraphics[scale=0.5]{./Pictures/RangeXrangePyCpp}
  \caption{Comparaison de performances entre python et C++ compilé avec et sans le flag -O3.}
  \label{RangeXrangePyCpp}
\end{figure}

Les programme en python sont les suivants :

\begin{verbatim}
  sum(range(N))
\end{verbatim}

\begin{verbatim}
  sum(xrange(N))
\end{verbatim}

et leurs équivalents en C++ sont :

\begin{verbatim}
  pythonic::sum(pythonic::range(N))
\end{verbatim}

\begin{verbatim}
  pythonic::sum(pythonic::xrange(N))
\end{verbatim}

où pythonic est le namespace de pythonic++ (cf. \ref{pythonicpp} p. \pageref{pythonicpp}).

Les temps en python sont de nouveau mesurés avec timeit.timeit() et les temps en C++ sont mesurés avec clock().

On remarque que, contrairement à ce à quoi on pouvait s'attendre, les programme en C++ ne sont pas toujours plus rapides que ceux en python. Dans le cas du xrange, le programme en python est légèrement plus performant que celui en C++. Cependant, l'optimisation du compilateur C++ permet de réduire de façon drastiques le temps d'exécution. Enfin, le gain réalisé en remplaçant range par xrange se retrouve également en C++.

De cela on peut en conclure que pythran peut augmenter les performances d'un algorithme sur trois axes :

\begin{itemize}
  \item Optimisation du code d'origine, par exemple en utilisant des iterateurs comme xrange ou imap plutôt que range ou map.
  \item Traduction en code C++ pour pouvoir être compilé en code natif, en général plus rapide que du code interprété.
  \item Possibilité de bénéficier des optimisations supplémentaires du compilateur C++.
\end{itemize}

Pythran permet également d'annoter le programme de directives OpenMP, qui permettent de paralléliser les opérations qui peuvent l'être. Une bonne utilisation des directives OpenMP permet ainsi d'augmenter encore la vitesse d'execution d'un facteur x2 à x4\cite{PythranRenpar}.

\section{Analyse du problème}

Le code python lu par pythran est réprésenté sous forme d'un arbre syntaxique abstrait (AST) dont les noeuds représentent des opérateurs et les feuilles des opérandes. C'est sur cet arbre que pythran travaille pour optimiser le programme et le traduire en C++.

\subsection{Transformations de l'AST}

Avant de produire la traduction C++, pythran peut effectuer des opérations sur l'AST. Ces opérations visent à apporter des modifications ne changeant pas les résultats produits par le programme, mais utilisant des combinaisons d'opérateurs plus performantes celles initialement utilisées. On appelle ce type de modification une transformation de l'AST.

\subsubsection{Recherche de transformations valide}

La recherche de transformations pertinentes nécessitent d'avoir des connaissance pointues sur le coût des différentes opérations effectuées par un même opérateur. Par exemple, c'est parce que l'allocation d'une grande quantité de mémoire a un coût prohibitif que les transformation du type range -> xrange permet un gain de performances.

Un autre problème est celui de la recherche de la validité d'une transformation. On a vu qu'une transformation consiste à remplacer une combinaison d'opérateurs par une autre combinaison équivalente. Or, cette équivalence peut n'être vérifiée que sous certaines conditions. La transformation n'est alors valide que dans les cas où ces conditions sous toutes vérifiées.

Dans le cas de xrange, une condition suffisante pour que la transformation range -> xrange soit valide est que le résultat produit par range ne soit jamais modifié. Ce qui serait par exemple le cas dans l'exemple suivant\footnote{En supposant que range(100) soit bien une opération produisant la liste [1..100].} :

\begin{verbatim}
  for i in range(100):
    f(i)
\end{verbatim}

Ceci peut donc être remplacé par

\begin{verbatim}
  for i in xrange(100):
    f(i)
\end{verbatim}

\subsubsection{Le problème du contexte}

Un obstacle majeur est qu'il n'est pas trivial de savoir si les conditions de validité de la transformation sont remplies.

Par exemple, il est possible d'appliquer la transformation range -> xrange au programme suivant. Mais pour pouvoir savoir cela, il faut vérifier dans le reste du code pour s'assurer que la liste assignée à la variable l n'est jamais modifiée.

\begin{verbatim}
  l = range(100)
  .
  .
  .
  for i in l:
    f(i)
\end{verbatim}

Un autre problème vient du fait que dans l'AST, un opérateur est représenté par son nom, ce qui permet la situation de l'exemple suivant, où la transformation range() -> xrange() n'est pas valide.

\begin{verbatim}
  def g(n):
    return [9001,n,27,42]
  .
  .
  .
  range = g
  .
  .
  .
  for i in range(100):
    f(i)
\end{verbatim}

Avant d'effectuer la transformation, il faut donc analyser le programme pour déterminer dans quels cas la transformation est valide. Dans pythran, c'est ce que réalisent les analyses, qui parcourent et annotent l'AST. Les transformations appliquées ensuites utilisent les annotations laissées par les analyses.

Un exemple simple est celui de la réduction des expressions constantes. Une analyse parcourt l'AST pour déterminer quels noeuds correspondent à des expressions constantes. Une transformation peut ensuite être appliquée à ces noeuds pour les réduire en feuille.

\subsection{Compilation du code optimisé}

\label{pythonicpp}

C'est à partir de l'AST optimisé que pythran génère le code C++. Pour cela, il utilise pythonic++, un ensemble de headers C++ qui contiennent les équivalents en C++ des opérateurs et des collections natives au python.

Pour qu'un module de la bibliothèque standart de python puisse être utilisé dans du code traduit par pythran, il faut donc qu'un équivalent de ce module ai été développé pour pythonic++.

En particulier, le module itertools n'est pas encore implementé dans pythonic++, ce qui empêche d'utiliser des opérateurs comme imap ou product.

\section{Plan de travail}

L'ordre envisagé pour le plan de travail prend en compte plusieurs paramètres :

\begin{itemize}
\item Certains travaux dépendent des résultats des autres. En particulier, la plupart des transformations de l'AST visent à utiliser des fonctions du module itertools toute les fois où cela est possible. Ces transformations ne peuvent donc être implémentées qu'une fois le module itertools disponible dans pythonic++.
\item Certains travaux sont plus simples et permettent d'assimiler les concepts et outils nécessaires à la réalisation des travaux plus complexes du même type. Par exemple, la transformation de range en xrange est complexe et devrait donc être abordée après que la transformation de generator expression en imap ait été complétée.
\end{itemize}

\subsection{Implémentation du module Itertools}

\subsubsection{Ecriture du module pour Pythonic++}

Le module itertools doit être implémenté en C++ pour pythonic++. Le code produit doit être homogène avec le reste de pythonic++. A savoir :

\begin{itemize}
\item Etre conforme à la norme C++11.
\item BLABLABLA...
\end{itemize}

TODO Completer ou fusionner les subsubsection

\subsubsection{Méthodologie}

La méthodologie utilisée pour l'implémentation du module itertools est une itération des étapes suivantes :

\begin{enumerate}
\item Déterminer une fonction à implémenter
\item Ecrire des tests 
\item Implémenter la fonction
\item Tester la fonction pour vérifier qu'elle produit le résultat attendu pour tout les cas de test.
\item Tester les performances de la fonction et vérifier qu'il n'y a pas d'opération anormalement couteuse avec l'outils Callgrind. 
\end{enumerate}

\subsubsection{Priorités}

TODO imap, ifiler et product? cf S\P

\subsection{Transformation de l'AST}

Une seconde partie du travail consistera à TODO(concevoir?) des transformations de l'AST pertinentes. En particulier, il faudra déterminer dans quel cas elles sont valides et s'assurer de ne les appliquer que dans ces cas.

\subsubsection{Les generator expressions}

TODO Justifier pertinence

Il est toujours possible de remplacer

\begin{verbatim}
  (g(a,b,c...) for a in A if fa(a) for b in B if fb(a,b) for c in C if fc(a,b,c) ... )
\end{verbatim}

par

\begin{verbatim}
  imap(lambda a,b,c... : g(a,b,c...), ifilter(lambda a,b,c... : fa(a) and fb(a,b) and fc(a,b,c)..., product(A,B,C...)))
\end{verbatim}

Il faudra donc implémenter la transformation correspondante.

\subsubsection{Le cas xrange}

Les exemples précédents ont montré que la transformation range -> xrange n'est pas toujours valide. Il faudra donc :

\begin{enumerate}
  \item Identifier la plupart des cas dans lesquels la transformation est valide.
  \item Ecrire une ou des analyses permettant d'identifier ces cas.
  \item Implémenter la transformation.
\end{enumerate}

On note en particulier qu'il est à priori très difficile de trouver une analyse permettant de trouver tous les cas dans lesquels la transformation serait valide.

Par exemple, il serait valide de remplacer range par xrange dans la boucle for de l'exemple suivant, mais très difficile de le déterminer.

\begin{verbatim}
def g(n):
  return map(lambda x : - 1 * x, xrange(0,-n,-1))

range = g

for i in range(100):
    f(i)
\end{verbatim}

\clearpage

\appendix

\listoffigures

\bibliographystyle{plain}
\bibliography{RapportCourtBib}

\end{document}

### 2012-11-22 pythran/passes.py
Fix bug in expand import.
Function argument may shadows module name.

@@ -573,7 +573,7 @@ class ExpandImports(Transformation):
     def visit_FunctionDef(self, node):
         self.symbols.pop(node.name, None)
         gsymbols = self.symbols.copy()
-        self.visit(node.args)
+        [self.symbols.pop(arg.id, None) for arg in node.args.args]
         node.body = [k for k in (self.visit(n) for n in node.body) if k]
         self.symbols = gsymbols
         return node

### 2012-11-22 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
Complted TODO section and added a reference

@@
 
-TODO : Langage interprété, conçu pour optimiser dev et pas execution
+Python est un langage interprété conçu pour faciliter le développement mais pas pour obtenir des hautes performances.
+L'exécution d'algorithmes est généralement plus lentes qu'avec d'autres langages\cite{PythonPerf}.
 
@@
 Python permet génalement de résoudre un même problème de plusieurs façons.
-Si le choix d'un bon algorithme est crucial, la façon dont est implémenté cet algorithme peut faire varier considérablement les performance du programme final.
-C'est pourquoi il existe un ensemble de bonne pratiques permettant d'améliorer son code\cite{PythonSpeed}.
+Si le choix d'un bon algorithme est crucial, la façon dont est implémenté cet algorithme peut faire varier considérablement les performances du programme final.
+C'est pourquoi il existe un ensemble de bonne pratiques permettant d'améliorer les performances de son code\cite{PythonSpeed}.
 
@@
 
-Afin de considérer l'intérêt en terme de performance, on peut considérer la figure~\ref{GenExpMapImapPy} , qui montre les différences de temps d'exécution entre trois implémentations du calcul de la somme des carrés des N premiers entiers.
+Afin de considérer l'intérêt en terme de performances, on peut considérer la figure~\ref{GenExpMapImapPy} , qui montre les différences de temps d'exécution entre trois implémentations du calcul de la somme des carrés des N premiers entiers.
+
+\begin{figure}[h]
+  \includegraphics[scale=0.5]{./Pictures/GenExpMapImapPy}
+  \caption{Compaison de performances entre trois implémentations en python}
+  \label{GenExpMapImapPy}
+\end{figure}
 
@@
 \end{verbatim}
-
-\begin{figure}[h]
-  \includegraphics[scale=0.5]{./Pictures/GenExpMapImapPy}
-  \caption{Compaison de performances entre trois implémentations en python}
-  \label{GenExpMapImapPy}
-\end{figure}
 
@@
 
-Cette implémentation est moins performante que la première car elle utilise des appels de fonctions, qui sont couteux en python.
+Cette implémentation est moins performante que la première car elle utilise des appels de fonction, qui sont couteux en python.
 
@@
 CPython est l'implémentation de référence du langage python.
-Il en existe cependant d'autre comme Jython écrit en Java et IronPython , l'implémentation de python pour .NET.
+Il en existe cependant d'autre comme Jython écrit en Java et IronPython, l'implémentation de python pour .NET.
 Le projet Pypy, quant à lui, a pour but d'obtenir une implémentation de python plus performante que CPython.
@@
 L'outils pythran permet de traduire un module python en C++ pour pouvoir ensuite être compilé en code natif.
-Afin d'être utilisable par pythran, le code source doit seulement être annoté de descriptions d'interfaces.
+Afin d'être utilisable par pythran, le code source doit seulement être annoté de descriptions d'interface.
 Le résultat est un module python en code natif qui peut être utilisé depuis un programme python usuel.
-
-Avant même de traduire le code python en C++, pythran se charge de l'optimiser. Par exemple, il va analyser quels sont les expressions constantes et les remplacer par leurs valeurs.
 
@@
 
-On remarque que, contrairement à ce à quoi on pouvait s'attendre, les programme en C++ ne sont pas toujours plus rapides que ceux en python.
+\begin{figure}[h]
+  \includegraphics[scale=0.5]{./Pictures/RangeXrangeCppO3}
+  \caption{Comparaison entre range et xrange en C++ avec le flag O3.}
+  \label{RangeXrangeCppO3}
+\end{figure}
+
+On remarque que, contrairement à ce à quoi on pouvait s'attendre, les implémentations en C++ ne sont pas toujours plus rapides que celles en python.
 Dans le cas du xrange, le programme en python est légèrement plus performant que celui en C++.
-Cependant, l'optimisation du compilateur C++ permet de réduire de façon drastiques le temps d'exécution.
-Enfin, le gain réalisé en remplaçant range par xrange se retrouve également en C++.
+Cependant, l'optimisation du compilateur C++ permet de réduire de façon drastique le temps d'exécution.
+Enfin, le gain réalisé en remplaçant range par xrange se retrouve également en C++, comme le montre la figure~\ref{RangeXrangeCppO3}.
 
@@
 La recherche de transformations pertinentes nécessitent d'avoir des connaissance pointues sur le coût des différentes opérations effectuées par un même opérateur.
-Par exemple, c'est parce que l'allocation d'une grande quantité de mémoire a un coût prohibitif que les transformation du type range -> xrange permet un gain de performances.
+Par exemple, c'est parce que l'allocation d'une grande quantité de mémoire a un coût prohibitif que les transformation du type range -> xrange permettent un gain de performances.
 
@@
 Or, cette équivalence peut n'être vérifiée que sous certaines conditions.
-La transformation n'est alors valide que dans les cas où ces conditions sous toutes vérifiées.
+La transformation n'est alors valide que dans les cas où ces conditions sont toutes vérifiées.
 
@@
 
-Un autre problème vient du fait que dans l'AST, un opérateur est représenté par son nom, ce qui permet la situation de l'exemple suivant, où la transformation range() -> xrange() n'est pas valide.
+Un autre problème vient du fait que dans l'AST, un opérateur est représenté par son nom, ce qui permet la situation de l'exemple suivant, où la transformation range() -> xrange() n'est pas valide dans la boucle for.
 
@@
 
-Pour qu'un module de la bibliothèque standart de python puisse être utilisé dans du code traduit par pythran, il faut donc qu'un équivalent de ce module ai été développé pour pythonic++.
+Pour qu'un module de la bibliothèque standard de python puisse être utilisé dans du code traduit par pythran, il faut donc qu'un équivalent de ce module ai été développé pour pythonic++.
 
-En particulier, le module itertools n'est pas encore implementé dans pythonic++, ce qui empêche d'utiliser des opérateurs comme imap ou product.
+En particulier, le module itertools n'est pas encore implementé dans pythonic++, ce qui empêche d'utiliser des fonctions comme imap ou product.
 
@@
 \item Certains travaux dépendent des résultats des autres.
-En particulier, la plupart des transformations de l'AST visent à utiliser des fonctions du module itertools toute les fois où cela est possible.
+En particulier, la plupart des transformations de l'AST visent à utiliser des fonctions du module itertools dès que cela est possible.
 Ces transformations ne peuvent donc être implémentées qu'une fois le module itertools disponible dans pythonic++.
@@
 
-\subsubsection{Ecriture du module pour Pythonic++}
-
-Le module itertools doit être implémenté en C++ pour pythonic++. Le code produit doit être homogène avec le reste de pythonic++. A savoir :
-
-\begin{itemize}
-\item Etre conforme à la norme C++11.
-\item BLABLABLA...
-\end{itemize}
-
-TODO Completer ou fusionner les subsubsection
-
 \subsubsection{Méthodologie}
@@
 \begin{enumerate}
-\item Déterminer une fonction à implémenter
-\item Ecrire des tests 
-\item Implémenter la fonction
+\item Déterminer une fonction à implémenter.
+\item Écrire des tests.
+\item Implémenter la fonction en C++ dans pythonic++.
 \item Tester la fonction pour vérifier qu'elle produit le résultat attendu pour tout les cas de test.
-\item Tester les performances de la fonction et vérifier qu'il n'y a pas d'opération anormalement couteuse avec l'outils Callgrind. 
+\item Tester les performances de la fonction et vérifier qu'il n'y a pas d'opération anormalement coûteuse avec l'outils Callgrind. 
 \end{enumerate}
@@
 
-TODO imap, ifiler et product? cf S\P
+Les opérateurs à implémenter en priorités sont ceux nécessaires à l'implémentations des nouvelles transformations :
+
+\begin{itemize}
+  \item imap
+  \item product
+  \item ifilter
+\end{itemize}
 
@@
 
-Une seconde partie du travail consistera à TODO(concevoir?) des transformations de l'AST pertinentes.
+Une seconde partie du travail consistera à mettre au point des transformations de l'AST pertinentes.
 En particulier, il faudra déterminer dans quel cas elles sont valides et s'assurer de ne les appliquer que dans ces cas.
@@
 
-TODO Justifier pertinence
+En python, les generator expression sont plus performantes que l'utilisation de map ou imap.
+Cependant, cela est lié au cout prohibitif d'un appel de fonction en python.
+Une fois traduit en C++, ce surcoût devrait être moindre et devrait rendre l'implémentation de imap plus performante que l'équivalent des generator expressions.
 
-Il est toujours possible de remplacer
+Il sera alors pertinent de remplacer
 
@@
 
-Il faudra donc implémenter la transformation correspondante.
+Cette transformation est toujours valide et ne nécessitera pas d'écrire une analyse.
+
+Il faudra donc implémenter la transformation correspondante, et effecter des tests pour vérifier qu'elle apporte bien un gain de performances.
 
@@
   \item Identifier la plupart des cas dans lesquels la transformation est valide.
-  \item Ecrire une ou des analyses permettant d'identifier ces cas.
+  \item Écrire une ou des analyses permettant d'identifier ces cas.
   \item Implémenter la transformation.
@@
 
+\subsubsection{Contraintes}
+
+Le analyse et transformations de pythran sont écrites en python.
+Tout code produit doit être conforme à pep8, un ensemble de bonnes pratiques python.
+Pour cela, avant d'être commité, tout code ajouté doit être validé par l'outils pep8.
+
 \clearpage

### 2012-11-22 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
"RapportCourt" completed with a new graph (optimized c++ range/xrange)

@@
 
+\clearpage
+
 \tableofcontents
@@
 
-Python est un langage interprété conçu pour faciliter le développement mais pas pour obtenir des hautes performances.
-L'exécution d'algorithmes est généralement plus lentes qu'avec d'autres langages\cite{PythonPerf}.
+Le python, langage inteprété et dynamiquement typé, est souvent considéré comme moins performant que d'autre langages compilés comme C ou C++.
+Dans ce contexte, la performance est souvent réduite au temps d'éxecution des opérations du programme, et c'est cette définition restreinte qui sera utilisée sous le nom de performance dans la suite\footnote{Une meilleurs définition serait celle de Arnold Allen : "The word performance in computer performance means the same thing that performance means in other contexts, that is, it means 'How well is the computer doing the work it is supposed to do?'"\cite{Allen}}.
+Toutes les applications n'ont pas un même besoin en performances, et celles du python sont généralement considérées comme suffisantes dans la plupart des cas.
+Il existe cependant des applications qui effectuent des calculs intensifs ou doivent traiter de très grande quantités de données, et qui ont donc besoin que certains de leurs algorithmes soient effectués avec des performances supérieures à celles ordinairement obtenues avec python.
 
-TODO : Question de la nécessité de la performance + temps d'execution vs temps de dev
-
-\subsection{Optimisation dans l'utilisation de python}
+\subsection{Utilisation d'un python plus performant}
 
@@
 
-\subsubsection{Améliorer les performances en améliorant le code}
+\subsubsection{Bonne pratiques de codage}
 
@@
 Si le choix d'un bon algorithme est crucial, la façon dont est implémenté cet algorithme peut faire varier considérablement les performances du programme final.
-C'est pourquoi il existe un ensemble de bonne pratiques permettant d'améliorer les performances de son code\cite{PythonSpeed}.
+C'est pourquoi il existe un ensemble de bonnes pratiques permettant d'améliorer les performances de son code\cite{PythonSpeed}.
 
@@
 \begin{itemize}
-  \item xrange qui se comporte comme range, mais ne créé pas de liste en mémoire
-  \item les iterateurs du module itertools comme imap et product
+  \item xrange qui se comporte comme range, mais ne créé pas de liste en mémoire.
+  \item les iterateurs du module itertools comme imap et product.
 \end{itemize}
 
-Afin de considérer l'intérêt en terme de performances, on peut considérer la figure~\ref{GenExpMapImapPy} , qui montre les différences de temps d'exécution entre trois implémentations du calcul de la somme des carrés des N premiers entiers.
+Afin de considérer l'intérêt en terme de performances, on peut considérer la figure~\ref{GenExpMapImapPy}, qui montre les différences de temps d'exécution entre trois implémentations du calcul de la somme des carrés des N premiers entiers.
 
@@
 
-Le générateur créé permet de ne produire les carrés qu'au fur et à mesure de la sommation. A un instant donné, il n'y a qu'un seul carré en mémoire.
+Le générateur créé permet de ne produire les carrés qu'au fur et à mesure de la sommation. À un instant donné, il n'y a qu'un seul carré en mémoire.
 
-La seconde implémentation utilise l'operateur imap qui applique une fonction à un iterateur pour produire un nouvel iterateur :
+La seconde implémentation utilise la fonction imap qui applique une fonction à un iterateur pour produire un nouvel iterateur :
 
@@
 
-La troisième implémentation utilise l'operateur map qui applique une fonction à un iterateur pour produire une liste :
+La troisième implémentation utilise la fonction map qui applique une fonction à un iterateur pour produire une liste :
 
@@
 
+On remarque bien que l'implémentation choisie peut faire considérablement varier le temps d'exécution.
+On constate également que le simple fait de remplacer map par imap permet un gain de performance non néglieable dès que l'on manipule un large ensemble de données.
+La même constatation peut être faite quand à l'utilisation de xrange plutôt que range ou encore izip plutôt que zip.
+
+\subsubsection{Module spécialisés}
+
+Il existe des modules pour python destinés à permettre de réaliser des applications de calcul scientifiques en python.
+C'est par exemple le cas du package NumPy, qui fournit à python des fonctionnalités proches de celles de matlab.
+Les modules de numpy sont écrit en C, de façon à permettre à un programme les utilisant d'avoir des performances proches d'un programme écrit en C\cite{NumPyPerf}.
+
 \subsection{Optimisation à partir du code python}
@@
 Les méthodes d'amélioration des performances présentées en \ref{OptCode} nécessitent d'effectuer des modifications au niveau du code source.
-Les gains en temps d'exécution se feront donc au prix d'augmentation du temps de développement.
+Les gains en temps d'exécution se feront donc au prix d'une augmentation du temps de développement.
 Il existe cependant des outils qui permettent d'augmenter les performances d'un programme sans avoir à modifier le code source.
@@
 L'outils pythran permet de traduire un module python en C++ pour pouvoir ensuite être compilé en code natif.
-Afin d'être utilisable par pythran, le code source doit seulement être annoté de descriptions d'interface.
+Afin d'être utilisable par pythran, le code source doit seulement être annoté de descriptions d'interfaces.
 Le résultat est un module python en code natif qui peut être utilisé depuis un programme python usuel.
-
 Pythran ne traduit qu'un sous-ensemble de python.
@@
 \begin{itemize}
-  \item Optimisation du code d'origine, par exemple en utilisant des iterateurs comme xrange ou imap plutôt que range ou map.
+  \item Optimisation du code d'origine, par exemple en remplaçant automatiquement des fonctions comme range ou map par xrange et imap.
   \item Traduction en code C++ pour pouvoir être compilé en code natif, en général plus rapide que du code interprété.
@@
 Pythran permet également d'annoter le programme de directives OpenMP, qui permettent de paralléliser les opérations qui peuvent l'être.
-Une bonne utilisation des directives OpenMP permet ainsi d'augmenter encore la vitesse d'execution d'un facteur x2 à x4\cite{PythranRenpar}.
+Cela nécessite une analyse du programme pour déterminer les opérations parallélisables, mais cela permet d'augmenter encore la vitesse d'execution d'un facteur x2 à x4\cite{PythranRenpar}.
 
@@
 Par exemple, il est possible d'appliquer la transformation range -> xrange au programme suivant.
-Mais pour pouvoir savoir cela, il faut vérifier dans le reste du code pour s'assurer que la liste assignée à la variable l n'est jamais modifiée.
+Mais pour pouvoir savoir cela, il faut vérifier le reste du code pour s'assurer que la liste assignée à la variable l n'est jamais modifiée.
 
@@
 Pour qu'un module de la bibliothèque standard de python puisse être utilisé dans du code traduit par pythran, il faut donc qu'un équivalent de ce module ai été développé pour pythonic++.
-
-En particulier, le module itertools n'est pas encore implementé dans pythonic++, ce qui empêche d'utiliser des fonctions comme imap ou product.
+Ce n'est pas encore le cas du module itertools, ce qui empêche d'utiliser des fonctions comme imap ou product.
 
@@
 \begin{verbatim}
-  imap(lambda a,b,c... : g(a,b,c...), ifilter(lambda a,b,c... : fa(a) and fb(a,b) and fc(a,b,c)..., product(A,B,C...)))
+imap(lambda a,b,c... : g(a,b,c...), 
+  ifilter(lambda a,b,c... : fa(a) and fb(a,b) and fc(a,b,c)..., product(A,B,C...)))
 \end{verbatim}
@@
 Cette transformation est toujours valide et ne nécessitera pas d'écrire une analyse.
-
-Il faudra donc implémenter la transformation correspondante, et effecter des tests pour vérifier qu'elle apporte bien un gain de performances.
+Il faudra donc simplement implémenter la transformation, et effecter des tests pour vérifier qu'elle apporte bien un gain de performances.
 
@@
 
-On note en particulier qu'il est à priori très difficile de trouver une analyse permettant de trouver tous les cas dans lesquels la transformation serait valide.
-
+On note en particulier qu'il est à priori hors de portée de trouver une analyse permettant de trouver tous les cas dans lesquels la transformation serait valide.
 Par exemple, il serait valide de remplacer range par xrange dans la boucle for de l'exemple suivant, mais très difficile de le déterminer.

### 2012-11-23 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
Spellchecking and line length correction

@@
 
-Le python, langage inteprété et dynamiquement typé, est souvent considéré comme moins performant que d'autre langages compilés comme C ou C++.
-Dans ce contexte, la performance est souvent réduite au temps d'éxecution des opérations du programme, et c'est cette définition restreinte qui sera utilisée sous le nom de performance dans la suite\footnote{Une meilleurs définition serait celle de Arnold Allen : "The word performance in computer performance means the same thing that performance means in other contexts, that is, it means 'How well is the computer doing the work it is supposed to do?'"\cite{Allen}}.
-Toutes les applications n'ont pas un même besoin en performances, et celles du python sont généralement considérées comme suffisantes dans la plupart des cas.
-Il existe cependant des applications qui effectuent des calculs intensifs ou doivent traiter de très grande quantités de données, et qui ont donc besoin que certains de leurs algorithmes soient effectués avec des performances supérieures à celles ordinairement obtenues avec python.
+Le python, langage interprété et dynamiquement typé, est souvent
+considéré comme moins performant que d'autre langages compilés comme C
+ou C++. Dans ce contexte, la performance est souvent réduite au temps
+d'exécution des opérations du programme, et c'est cette définition
+restreinte qui sera utilisée sous le nom de performance dans la
+suite\footnote{Une meilleurs définition serait celle de Arnold Allen :
+  "The word performance in computer performance means the same thing
+  that performance means in other contexts, that is, it means 'How
+  well is the computer doing the work it is supposed to
+  do?'"\cite{Allen}}. Toutes les applications n'ont pas un même besoin
+en performances, et celles du python sont généralement considérées
+comme suffisantes dans la plupart des cas. Il existe cependant des
+applications qui effectuent des calculs intensifs ou doivent traiter
+de très grande quantités de données, et qui ont donc besoin que
+certains de leurs algorithmes soient effectués avec des performances
+supérieures à celles ordinairement obtenues avec python.
 
@@
 
-Python permet génalement de résoudre un même problème de plusieurs façons.
-Si le choix d'un bon algorithme est crucial, la façon dont est implémenté cet algorithme peut faire varier considérablement les performances du programme final.
-C'est pourquoi il existe un ensemble de bonnes pratiques permettant d'améliorer les performances de son code\cite{PythonSpeed}.
+Python permet généralement de résoudre un même problème de plusieurs
+façons. Si le choix d'un bon algorithme est crucial, la façon dont est
+implémenté cet algorithme peut faire varier considérablement les
+performances du programme final. C'est pourquoi il existe un ensemble
+de bonnes pratiques permettant d'améliorer les performances de son
+code\cite{PythonSpeed}.
 
@@
 \item Concaténer des strings avec .join() plutôt qu'avec l'opérateur +.
-\item Utiliser plusieurs affectations plutôt qu'une affectation multiple : a=x;b=y; plutôt que a,b=x,y.
-\item Utiliser les opérateurs propres au langage plutôt que les fonctions lamda : operator.add plutôt que lambda x,y : x+y.
+\item Utiliser plusieurs affectations plutôt qu'une affectation
+  multiple : a=x;b=y; plutôt que a,b=x,y.
+\item Utiliser les opérateurs propres au langage plutôt que les
+  fonctions lambda : operator.add plutôt que lambda x,y : x+y.
 \end{itemize}
 
-\subsubsection{Iterateurs et generateur}
+\subsubsection{Itérateurs et générateur}
 
-Dans la boucle de l'exemple suivant, python génère la liste  [1..100] puis créé un objet iterateur qui permet de parcourir un à un les élements de cette liste.
+Dans la boucle de l'exemple suivant, python génère la liste [1..100]
+puis créé un objet itérateur qui permet de parcourir un à un les
+éléments de cette liste.
 
@@
 
-On se rend compte dans cet exemple qu'il n'était pas nécessaire d'avoir toute la liste en mémoire. En effet, un seul élement est utilisé à la fois, et chaque élément n'est utilisé qu'une seule fois.
+On se rend compte dans cet exemple qu'il n'était pas nécessaire
+d'avoir toute la liste en mémoire. En effet, un seul élément est
+utilisé à la fois, et chaque élément n'est utilisé qu'une seule fois.
 
-C'est pourquoi il existe en python des générateurs.
-Les générateurs permettent de réaliser des iterateurs qui ne parcourent pas un ensemble, mais génèrent les éléments de l'itération au fur et à mesure.
-Quand on utilise un iterateur produit par un generateur, il n'y a toujours qu'une seule valeur en mémoire : celle actuellement utilisée.
+C'est pourquoi il existe en python des générateurs. Les générateurs
+permettent de réaliser des itérateurs qui ne parcourent pas un
+ensemble, mais génèrent les éléments de l'itération au fur et à
+mesure. Quand on utilise un itérateur produit par un générateur, il
+n'y a toujours qu'une seule valeur en mémoire : celle actuellement
+utilisée.
 
-Python fournit également des iterateurs se comportant comme ceux produits par des générateurs :
+Python fournit également des itérateurs se comportant comme ceux
+produits par des générateurs :
 
 \begin{itemize}
-  \item xrange qui se comporte comme range, mais ne créé pas de liste en mémoire.
-  \item les iterateurs du module itertools comme imap et product.
+\item xrange qui se comporte comme range, mais ne créé pas de liste en
+  mémoire.
+\item les itérateurs du module itertools comme imap et product.
 \end{itemize}
 
-Afin de considérer l'intérêt en terme de performances, on peut considérer la figure~\ref{GenExpMapImapPy}, qui montre les différences de temps d'exécution entre trois implémentations du calcul de la somme des carrés des N premiers entiers.
+Afin de considérer l'intérêt en terme de performances, on peut
+considérer la figure~\ref{GenExpMapImapPy}, qui montre les différences
+de temps d'exécution entre trois implémentations du calcul de la somme
+des carrés des N premiers entiers.
 
 \begin{figure}[h]
-  \includegraphics[scale=0.5]{./Pictures/GenExpMapImapPy}
-  \caption{Compaison de performances entre trois implémentations en python}
+  \includegraphics[scale=1]{./Pictures/GenExpMapImapPy}
+  \caption{Comparaison de performances entre trois implémentations en python}
   \label{GenExpMapImapPy}
@@
 
-Dans chaque cas, le temps de l'execution de 100 itérations du même programme est mesuré grâce au module timeit.
+Dans chaque cas, le temps de l'exécution de 100 itérations du même
+programme est mesuré grâce au module timeit.
 
@@
 
-Le générateur créé permet de ne produire les carrés qu'au fur et à mesure de la sommation. À un instant donné, il n'y a qu'un seul carré en mémoire.
+Le générateur créé permet de ne produire les carrés qu'au fur et à
+mesure de la sommation. À un instant donné, il n'y a qu'un seul carré
+en mémoire.
 
-La seconde implémentation utilise la fonction imap qui applique une fonction à un iterateur pour produire un nouvel iterateur :
+La seconde implémentation utilise la fonction imap qui applique une
+fonction à un itérateur pour produire un nouvel itérateur :
 
@@
 
-Cette implémentation est moins performante que la première car elle utilise des appels de fonction, qui sont couteux en python.
+Cette implémentation est moins performante que la première car elle
+utilise des appels de fonction, qui sont coûteux en python.
 
-La troisième implémentation utilise la fonction map qui applique une fonction à un iterateur pour produire une liste :
+La troisième implémentation utilise la fonction map qui applique une
+fonction à un itérateur pour produire une liste :
 
@@
 
-Dans cette implémentation, la liste des carrés est créée avant la sommation.
-C'est l'opération de création de l'espace mémoire nécessaire pour la stocker qui est couteuse et diminue les performance par rapport à la seconde implémentation.
+Dans cette implémentation, la liste des carrés est créée avant la
+sommation. C'est l'opération de création de l'espace mémoire
+nécessaire pour la stocker qui est coûteuse et diminue les performance
+par rapport à la seconde implémentation.
 
-On remarque bien que l'implémentation choisie peut faire considérablement varier le temps d'exécution.
-On constate également que le simple fait de remplacer map par imap permet un gain de performance non néglieable dès que l'on manipule un large ensemble de données.
-La même constatation peut être faite quand à l'utilisation de xrange plutôt que range ou encore izip plutôt que zip.
+On remarque bien que l'implémentation choisie peut faire
+considérablement varier le temps d'exécution. On constate également
+que le simple fait de remplacer map par imap permet un gain de
+performance non négligeable dès que l'on manipule un large ensemble de
+données. La même constatation peut être faite quand à l'utilisation de
+xrange plutôt que range ou encore izip plutôt que zip.
 
@@
 
-Il existe des modules pour python destinés à permettre de réaliser des applications de calcul scientifiques en python.
-C'est par exemple le cas du package NumPy, qui fournit à python des fonctionnalités proches de celles de matlab.
-Les modules de numpy sont écrit en C, de façon à permettre à un programme les utilisant d'avoir des performances proches d'un programme écrit en C\cite{NumPyPerf}.
+Il existe des modules pour python destinés à permettre de réaliser des
+applications de calcul scientifiques en python.  C'est par exemple le
+cas du package NumPy, qui fournit à python des fonctionnalités proches
+de celles de matlab. Les modules de numpy sont écrit en C, de façon à
+permettre à un programme les utilisant d'avoir des performances
+proches d'un programme écrit en C\cite{NumPyPerf}.
 
@@
 
-Les méthodes d'amélioration des performances présentées en \ref{OptCode} nécessitent d'effectuer des modifications au niveau du code source.
-Les gains en temps d'exécution se feront donc au prix d'une augmentation du temps de développement.
-Il existe cependant des outils qui permettent d'augmenter les performances d'un programme sans avoir à modifier le code source.
+Les méthodes d'amélioration des performances présentées en
+\ref{OptCode} nécessitent d'effectuer des modifications au niveau du
+code source. Les gains en temps d'exécution se feront donc au prix
+d'une augmentation du temps de développement. Il existe cependant des
+outils qui permettent d'augmenter les performances d'un programme sans
+avoir à modifier le code source.
 
@@
 
-CPython est l'implémentation de référence du langage python.
-Il en existe cependant d'autre comme Jython écrit en Java et IronPython, l'implémentation de python pour .NET.
-Le projet Pypy, quant à lui, a pour but d'obtenir une implémentation de python plus performante que CPython.
+CPython est l'implémentation de référence du langage python. Il en
+existe cependant d'autre comme Jython écrit en Java et IronPython,
+l'implémentation de python pour .NET. Le projet Pypy, quant à lui, a
+pour but d'obtenir une implémentation de python plus performante que
+CPython.
 
-Un programme écrit en python pour CPython peut être exécuté avec Pypy sans modification\footnote{Il existe quelques incompatibilités dans des cas particuliers, souvent liées à des différence de stratégie des ramasse-miettes\cite{PypyDiff}.} et sera en moyenne 5,5 fois plus rapide\cite{PypySpeed}.
+Un programme écrit en python pour CPython peut être exécuté avec Pypy
+sans modification\footnote{Il existe quelques incompatibilités dans
+  des cas particuliers, souvent liées à des différence de stratégie
+  des ramasse-miettes\cite{PypyDiff}.} et sera en moyenne 5,5 fois
+plus rapide\cite{PypySpeed}.
 
@@
 
-L'outils pythran permet de traduire un module python en C++ pour pouvoir ensuite être compilé en code natif.
-Afin d'être utilisable par pythran, le code source doit seulement être annoté de descriptions d'interfaces.
-Le résultat est un module python en code natif qui peut être utilisé depuis un programme python usuel.
-Pythran ne traduit qu'un sous-ensemble de python.
-En particulier, les classes ne sont pas gérées.
-Cela destine pythran à être utilisé pour l'optimisation d'algorithmes dont les temps d'exécution sont critiques pour le programme les utilisant.
+L'outils pythran permet de traduire un module python en C++ pour
+pouvoir ensuite être compilé en code natif. Afin d'être utilisable par
+pythran, le code source doit seulement être annoté de descriptions
+d'interfaces. Le résultat est un module python en code natif qui peut
+être utilisé depuis un programme python usuel. Pythran ne traduit
+qu'un sous-ensemble de python. En particulier, les classes ne sont pas
+gérées. Cela destine pythran à être utilisé pour l'optimisation
+d'algorithmes dont les temps d'exécution sont critiques pour le
+programme les utilisant.
 
-La figure~\ref{RangeXrangePyCpp} permet de comparer les performances entre un programme en python et son équivalent en C++.
+La figure~\ref{RangeXrangePyCpp} permet de comparer les performances
+entre un programme en python et son équivalent en C++.
 
 \begin{figure}[h]
-  \includegraphics[scale=0.5]{./Pictures/RangeXrangePyCpp}
+  \includegraphics[scale=1]{./Pictures/RangeXrangePyCpp}
   \caption{Comparaison de performances entre python et C++ compilé avec et sans le flag -O3.}
@@
 
-où pythonic est le namespace de pythonic++ (cf. \ref{pythonicpp} p. \pageref{pythonicpp}).
+où pythonic est le namespace de pythonic++ (cf. \ref{pythonicpp}
+p. \pageref{pythonicpp}).
 
-Les temps en python sont de nouveau mesurés avec timeit.timeit() et les temps en C++ sont mesurés avec clock().
+Les temps en python sont de nouveau mesurés avec timeit.timeit() et
+les temps en C++ sont mesurés avec clock().
 
 \begin{figure}[h]
-  \includegraphics[scale=0.5]{./Pictures/RangeXrangeCppO3}
+  \includegraphics[scale=1]{./Pictures/RangeXrangeCppO3}
   \caption{Comparaison entre range et xrange en C++ avec le flag O3.}
@@
 
-On remarque que, contrairement à ce à quoi on pouvait s'attendre, les implémentations en C++ ne sont pas toujours plus rapides que celles en python.
-Dans le cas du xrange, le programme en python est légèrement plus performant que celui en C++.
-Cependant, l'optimisation du compilateur C++ permet de réduire de façon drastique le temps d'exécution.
-Enfin, le gain réalisé en remplaçant range par xrange se retrouve également en C++, comme le montre la figure~\ref{RangeXrangeCppO3}.
+On remarque que, contrairement à ce à quoi on pouvait s'attendre, les
+implémentations en C++ ne sont pas toujours plus rapides que celles en
+python. Dans le cas du xrange, le programme en python est légèrement
+plus performant que celui en C++. Cependant, l'optimisation du
+compilateur C++ permet de réduire de façon drastique le temps
+d'exécution. Enfin, le gain réalisé en remplaçant range par xrange se
+retrouve également en C++, comme le montre la
+figure~\ref{RangeXrangeCppO3}.
 
-De cela on peut en conclure que pythran peut augmenter les performances d'un algorithme sur trois axes :
+De cela on peut en conclure que pythran peut augmenter les
+performances d'un algorithme sur trois axes :
 
 \begin{itemize}
-  \item Optimisation du code d'origine, par exemple en remplaçant automatiquement des fonctions comme range ou map par xrange et imap.
-  \item Traduction en code C++ pour pouvoir être compilé en code natif, en général plus rapide que du code interprété.
-  \item Possibilité de bénéficier des optimisations supplémentaires du compilateur C++.
+\item Optimisation du code d'origine, par exemple en remplaçant
+  automatiquement des fonctions comme range ou map par xrange et imap.
+\item Traduction en code C++ pour pouvoir être compilé en code natif,
+  en général plus rapide que du code interprété.
+\item Possibilité de bénéficier des optimisations supplémentaires du
+  compilateur C++.
 \end{itemize}
 
-Pythran permet également d'annoter le programme de directives OpenMP, qui permettent de paralléliser les opérations qui peuvent l'être.
-Cela nécessite une analyse du programme pour déterminer les opérations parallélisables, mais cela permet d'augmenter encore la vitesse d'execution d'un facteur x2 à x4\cite{PythranRenpar}.
+Pythran permet également d'annoter le programme de directives OpenMP,
+qui permettent de paralléliser les opérations qui peuvent l'être. Cela
+nécessite une analyse du programme pour déterminer les opérations
+parallélisables, mais cela permet d'augmenter encore la vitesse
+d'exécution d'un facteur x2 à x4\cite{PythranRenpar}.
 
@@
 
-Le code python lu par pythran est réprésenté sous forme d'un arbre syntaxique abstrait (AST) dont les noeuds représentent des opérateurs et les feuilles des opérandes.
-C'est sur cet arbre que pythran travaille pour optimiser le programme et le traduire en C++.
+Le code python lu par pythran est représenté sous forme d'un arbre
+syntaxique abstrait (AST) dont les noeuds représentent des opérateurs
+et les feuilles des opérandes. C'est sur cet arbre que pythran
+travaille pour optimiser le programme et le traduire en C++.
 
@@
 
-Avant de produire la traduction C++, pythran peut effectuer des opérations sur l'AST.
-Ces opérations visent à apporter des modifications ne changeant pas les résultats produits par le programme, mais utilisant des combinaisons d'opérateurs plus performantes celles initialement utilisées.
-On appelle ce type de modification une transformation de l'AST.
+Avant de produire la traduction C++, pythran peut effectuer des
+opérations sur l'AST. Ces opérations visent à apporter des
+modifications ne changeant pas les résultats produits par le
+programme, mais utilisant des combinaisons d'opérateurs plus
+performantes celles initialement utilisées. On appelle ce type de
+modification une transformation de l'AST.
 
@@
 
-La recherche de transformations pertinentes nécessitent d'avoir des connaissance pointues sur le coût des différentes opérations effectuées par un même opérateur.
-Par exemple, c'est parce que l'allocation d'une grande quantité de mémoire a un coût prohibitif que les transformation du type range -> xrange permettent un gain de performances.
+La recherche de transformations pertinentes nécessitent d'avoir des
+connaissance pointues sur le coût des différentes opérations
+effectuées par un même opérateur. Par exemple, c'est parce que
+l'allocation d'une grande quantité de mémoire a un coût prohibitif que
+les transformation du type range -> xrange permettent un gain de
+performances.
 
-Un autre problème est celui de la recherche de la validité d'une transformation.
-On a vu qu'une transformation consiste à remplacer une combinaison d'opérateurs par une autre combinaison équivalente.
-Or, cette équivalence peut n'être vérifiée que sous certaines conditions.
-La transformation n'est alors valide que dans les cas où ces conditions sont toutes vérifiées.
+Un autre problème est celui de la recherche de la validité d'une
+transformation. On a vu qu'une transformation consiste à remplacer une
+combinaison d'opérateurs par une autre combinaison équivalente. Or,
+cette équivalence peut n'être vérifiée que sous certaines
+conditions. La transformation n'est alors valide que dans les cas où
+ces conditions sont toutes vérifiées.
 
-Dans le cas de xrange, une condition suffisante pour que la transformation range -> xrange soit valide est que le résultat produit par range ne soit jamais modifié.
-Ce qui serait par exemple le cas dans l'exemple suivant\footnote{En supposant que range(100) soit bien une opération produisant la liste [1..100].} :
+Dans le cas de xrange, une condition suffisante pour que la
+transformation range -> xrange soit valide est que le résultat produit
+par range ne soit jamais modifié. Ce qui serait par exemple le cas
+dans l'exemple suivant\footnote{En supposant que range(100) soit bien
+  une opération produisant la liste [1..100].} :
 
@@
 
-Un obstacle majeur est qu'il n'est pas trivial de savoir si les conditions de validité de la transformation sont remplies.
+Un obstacle majeur est qu'il n'est pas trivial de savoir si les
+conditions de validité de la transformation sont remplies.
 
-Par exemple, il est possible d'appliquer la transformation range -> xrange au programme suivant.
-Mais pour pouvoir savoir cela, il faut vérifier le reste du code pour s'assurer que la liste assignée à la variable l n'est jamais modifiée.
+Par exemple, il est possible d'appliquer la transformation range ->
+xrange au programme suivant. Mais pour pouvoir savoir cela, il faut
+vérifier le reste du code pour s'assurer que la liste assignée à la
+variable l n'est jamais modifiée.
 
@@
 
-Un autre problème vient du fait que dans l'AST, un opérateur est représenté par son nom, ce qui permet la situation de l'exemple suivant, où la transformation range() -> xrange() n'est pas valide dans la boucle for.
+Un autre problème vient du fait que dans l'AST, un opérateur est
+représenté par son nom, ce qui permet la situation de l'exemple
+suivant, où la transformation range() -> xrange() n'est pas valide
+dans la boucle for.
 
@@
 
-Avant d'effectuer la transformation, il faut donc analyser le programme pour déterminer dans quels cas la transformation est valide.
-Dans pythran, c'est ce que réalisent les analyses, qui parcourent et annotent l'AST.
-Les transformations appliquées ensuites utilisent les annotations laissées par les analyses.
+Avant d'effectuer la transformation, il faut donc analyser le
+programme pour déterminer dans quels cas la transformation est
+valide. Dans pythran, c'est ce que réalisent les analyses, qui
+parcourent et annotent l'AST. Les transformations appliquées ensuite
+utilisent les annotations laissées par les analyses.
 
-Un exemple simple est celui de la réduction des expressions constantes.
-Une analyse parcourt l'AST pour déterminer quels noeuds correspondent à des expressions constantes.
-Une transformation peut ensuite être appliquée à ces noeuds pour les réduire en feuille.
+Un exemple simple est celui de la réduction des expressions
+constantes. Une analyse parcourt l'AST pour déterminer quels noeuds
+correspondent à des expressions constantes. Une transformation peut
+ensuite être appliquée à ces noeuds pour les réduire en feuille.
 
@@
 
-C'est à partir de l'AST optimisé que pythran génère le code C++.
-Pour cela, il utilise pythonic++, un ensemble de headers C++ qui contiennent les équivalents en C++ des opérateurs et des collections natives au python.
+C'est à partir de l'AST optimisé que pythran génère le code C++. Pour
+cela, il utilise pythonic++, un ensemble de headers C++ qui
+contiennent les équivalents en C++ des opérateurs et des collections
+natives au python.
 
-Pour qu'un module de la bibliothèque standard de python puisse être utilisé dans du code traduit par pythran, il faut donc qu'un équivalent de ce module ai été développé pour pythonic++.
-Ce n'est pas encore le cas du module itertools, ce qui empêche d'utiliser des fonctions comme imap ou product.
+Pour qu'un module de la bibliothèque standard de python puisse être
+utilisé dans du code traduit par pythran, il faut donc qu'un
+équivalent de ce module ai été développé pour pythonic++. Ce n'est pas
+encore le cas du module itertools, ce qui empêche d'utiliser des
+fonctions comme imap ou product.
 
@@
 
-L'ordre envisagé pour le plan de travail prend en compte plusieurs paramètres :
+L'ordre envisagé pour le plan de travail prend en compte plusieurs
+paramètres :
 
 \begin{itemize}
-\item Certains travaux dépendent des résultats des autres.
-En particulier, la plupart des transformations de l'AST visent à utiliser des fonctions du module itertools dès que cela est possible.
-Ces transformations ne peuvent donc être implémentées qu'une fois le module itertools disponible dans pythonic++.
-\item Certains travaux sont plus simples et permettent d'assimiler les concepts et outils nécessaires à la réalisation des travaux plus complexes du même type.
-Par exemple, la transformation de range en xrange est complexe et devrait donc être abordée après que la transformation de generator expression en imap ait été complétée.
+\item Certains travaux dépendent des résultats des autres. En
+  particulier, la plupart des transformations de l'AST visent à
+  utiliser des fonctions du module itertools dès que cela est
+  possible. Ces transformations ne peuvent donc être implémentées
+  qu'une fois le module itertools disponible dans pythonic++.
+\item Certains travaux sont plus simples et permettent d'assimiler les
+  concepts et outils nécessaires à la réalisation des travaux plus
+  complexes du même type. Par exemple, la transformation de range en
+  xrange est complexe et devrait donc être abordée après que la
+  transformation de generator expression en imap ait été complétée.
 \end{itemize}
@@
 
-La méthodologie utilisée pour l'implémentation du module itertools est une itération des étapes suivantes :
+La méthodologie utilisée pour l'implémentation du module itertools est
+une itération des étapes suivantes :
 
@@
 \item Implémenter la fonction en C++ dans pythonic++.
-\item Tester la fonction pour vérifier qu'elle produit le résultat attendu pour tout les cas de test.
-\item Tester les performances de la fonction et vérifier qu'il n'y a pas d'opération anormalement coûteuse avec l'outils Callgrind. 
+\item Tester la fonction pour vérifier qu'elle produit le résultat
+  attendu pour tout les cas de test.
+\item Tester les performances de la fonction et vérifier qu'il n'y a
+  pas d'opération anormalement coûteuse avec l'outils Callgrind.
 \end{enumerate}
@@
 
-Les opérateurs à implémenter en priorités sont ceux nécessaires à l'implémentations des nouvelles transformations :
+Les opérateurs à implémenter en priorités sont ceux nécessaires à
+l'implémentation des nouvelles transformations :
 
@@
 
-Une seconde partie du travail consistera à mettre au point des transformations de l'AST pertinentes.
-En particulier, il faudra déterminer dans quel cas elles sont valides et s'assurer de ne les appliquer que dans ces cas.
+Une seconde partie du travail consistera à mettre au point des
+transformations de l'AST pertinentes. En particulier, il faudra
+déterminer dans quel cas elles sont valides et s'assurer de ne les
+appliquer que dans ces cas.
 
@@
 
-En python, les generator expression sont plus performantes que l'utilisation de map ou imap.
-Cependant, cela est lié au cout prohibitif d'un appel de fonction en python.
-Une fois traduit en C++, ce surcoût devrait être moindre et devrait rendre l'implémentation de imap plus performante que l'équivalent des generator expressions.
+En python, les generator expression sont plus performantes que
+l'utilisation de map ou imap. Cependant, cela est lié au coût
+prohibitif d'un appel de fonction en python. Une fois traduit en C++,
+ce surcoût devrait être moindre et devrait rendre l'implémentation de
+imap plus performante que l'équivalent des generator expressions.
 
@@
 
-Cette transformation est toujours valide et ne nécessitera pas d'écrire une analyse.
-Il faudra donc simplement implémenter la transformation, et effecter des tests pour vérifier qu'elle apporte bien un gain de performances.
+Cette transformation est toujours valide et ne nécessitera pas
+d'écrire une analyse. Il faudra donc simplement implémenter la
+transformation, et effectuer des tests pour vérifier qu'elle apporte
+bien un gain de performances.
 
@@
 
-Les exemples précédents ont montré que la transformation range -> xrange n'est pas toujours valide. Il faudra donc :
+Les exemples précédents ont montré que la transformation range ->
+xrange n'est pas toujours valide. Il faudra donc :
 
 \begin{enumerate}
-  \item Identifier la plupart des cas dans lesquels la transformation est valide.
+\item Identifier la plupart des cas dans lesquels la transformation
+  est valide.
 \item Écrire une ou des analyses permettant d'identifier ces cas.
@@
 
-On note en particulier qu'il est à priori hors de portée de trouver une analyse permettant de trouver tous les cas dans lesquels la transformation serait valide.
-Par exemple, il serait valide de remplacer range par xrange dans la boucle for de l'exemple suivant, mais très difficile de le déterminer.
+On note en particulier qu'il est à priori hors de portée de trouver
+une analyse permettant de trouver tous les cas dans lesquels la
+transformation serait valide. Par exemple, il serait valide de
+remplacer range par xrange dans la boucle for de l'exemple suivant,
+mais très difficile de le déterminer.
 
@@
 
-Le analyse et transformations de pythran sont écrites en python.
-Tout code produit doit être conforme à pep8, un ensemble de bonnes pratiques python.
-Pour cela, avant d'être commité, tout code ajouté doit être validé par l'outils pep8.
+Le analyse et transformations de pythran sont écrites en python. Tout
+code produit doit être conforme à pep8, un ensemble de bonnes
+pratiques python. Pour cela, avant d'être commité, tout code ajouté
+doit être validé par l'outils pep8.
 

### 2012-11-23 pythran/pythonic++/core/string.h
A few more str methods.
And add an (almost empty) test bed for string methods.

@@ -137,8 +137,8 @@ namespace pythonic {
         };
     }
 }
-template<typename T>
-pythonic::core::string operator*(pythonic::core::string const & s, T const & t) {
+
+pythonic::core::string operator*(pythonic::core::string const & s, long t) {
     size_t n=t;
     pythonic::core::string other;
     other.resize(s.size()*n,'a');
@@ -147,6 +147,11 @@ pythonic::core::string operator*(pythonic::core::string const & s, T const & t) 
         std::copy(s.begin(), s.end(), where);
     return other;
 }
+
+pythonic::core::string operator*(long t, pythonic::core::string const & s) {
+    return s*t;
+}
+
 /* string_view implementation */
 pythonic::core::string_view::string_view(string & other, slice const &s) :
     data(other), slicing(s)

### 2012-11-25 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
A lot of rework and suggestions.
Note that I did not fix all the verbatim -> lstlisting conversion,
neither did I use texttt each time a command / module / intrinsic is
used.

The global layout is quite difficult to follow, use the abstract
to help you there !

@@
 \usepackage{url}
+\usepackage{listings}
+\lstset{language=python} % more options are welcome, RTFM!
 
 \usepackage{graphicx}
+\usepackage{todo}
 
@@
 \maketitle   
+\begin{abstract}
+    trop de la balouze ce sujet
+\end{abstract}
 
@@
 
-Le python, langage interprété et dynamiquement typé, est souvent
-considéré comme moins performant que d'autre langages compilés comme C
+Python, langage interprété et dynamiquement typé, est souvent
+considéré\todo{c'est plus qu'une considération, c'est un fait, cf. par example \url{http://shootout.alioth.debian.org/}} comme moins performant que d'autre langages compilés comme C
 ou C++. Dans ce contexte, la performance est souvent réduite au temps
-d'exécution des opérations du programme, et c'est cette définition
+d'exécution des opérations du programme,\todo{ça ne veut rien dire :-)} et c'est cette définition
 restreinte qui sera utilisée sous le nom de performance dans la
@@
   do?'"\cite{Allen}}. Toutes les applications n'ont pas un même besoin
-en performances, et celles du python sont généralement considérées
+en performances, et celles du Python sont généralement considérées
 comme suffisantes dans la plupart des cas. Il existe cependant des
@@
 certains de leurs algorithmes soient effectués avec des performances
-supérieures à celles ordinairement obtenues avec python.
+supérieures à celles ordinairement obtenues avec Python.\todo{tu peux citer numpy ici, dire que python est de plus en plus utilisé en calcul scientifique}
 
-\subsection{Utilisation d'un python plus performant}
+\subsection{Utilisation d'un Python plus performant}
 
@@
 de bonnes pratiques permettant d'améliorer les performances de son
-code\cite{PythonSpeed}.
+code\cite{PythonSpeed}.\todo{Attention, généralement un dev python ne se pose pas ce genre de questions. On dit qu'en python il existe une unique façon de résoudre un problème...}
 
@@
 \item Utiliser plusieurs affectations plutôt qu'une affectation
-  multiple : a=x;b=y; plutôt que a,b=x,y.
+  multiple : \lstinline|a=x;b=y;| plutôt que \lstinline|a,b=x,y|.
 \item Utiliser les opérateurs propres au langage plutôt que les
@@
 
-Dans la boucle de l'exemple suivant, python génère la liste [1..100]
+Dans la boucle de l'exemple suivant, Python génère la liste [1..100]
 puis créé un objet itérateur qui permet de parcourir un à un les
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 for i in range(100):
   print i
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-C'est pourquoi il existe en python des générateurs. Les générateurs
+C'est pourquoi il existe en Python des générateurs. Les générateurs
 permettent de réaliser des itérateurs qui ne parcourent pas un
@@
 \begin{itemize}
-\item xrange qui se comporte comme range, mais ne créé pas de liste en
+    \item \texttt{xrange} qui se comporte comme \texttt{range}, mais ne créé pas de liste en
   mémoire.
-\item les itérateurs du module itertools comme imap et product.
+\item les itérateurs du module \texttt{itertools} comme \texttt{imap} et \texttt{product}.\todo{toujours utiliser lstlisting / lstinline ou texttt pour le code}
 \end{itemize}
@@
 de temps d'exécution entre trois implémentations du calcul de la somme
-des carrés des N premiers entiers.
+des carrés des N premiers entiers.\todo{donner la version de python utilisée, la distrib, la version du noyau, le proc pour que ce soit reproductible}
 
 \begin{figure}[h]
-  \includegraphics[scale=1]{./Pictures/GenExpMapImapPy}
-  \caption{Comparaison de performances entre trois implémentations en python}
+  \includegraphics[width=\textwidth]{./Pictures/GenExpMapImapPy}
+  \caption{Comparaison de performances entre trois implémentations en Python.}
   \label{GenExpMapImapPy}
@@
 Dans chaque cas, le temps de l'exécution de 100 itérations du même
-programme est mesuré grâce au module timeit.
+programme est mesuré grâce au module \texttt{timeit}.\todo{expliquer un peu ce module}
 
-\begin{verbatim}
+\begin{lstlisting}
 time100 = timeit.timeit('Expression', number=100)
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 sum(x*x for x in xrange(N))
-\end{verbatim}
+\end{lstlisting}
 
@@
 Cette implémentation est moins performante que la première car elle
-utilise des appels de fonction, qui sont coûteux en python.
+utilise des appels de fonction, qui sont coûteux en Python.\todo{argumenter}
 
@@
 données. La même constatation peut être faite quand à l'utilisation de
-xrange plutôt que range ou encore izip plutôt que zip.
+xrange plutôt que range ou encore izip plutôt que zip.\todo{pour plus atrd: comment mesurer l'utilisation émmoire et la comparer?}
 
@@
 
-Il existe des modules pour python destinés à permettre de réaliser des
-applications de calcul scientifiques en python.  C'est par exemple le
-cas du package NumPy, qui fournit à python des fonctionnalités proches
+Il existe des modules pour Python destinés à permettre de réaliser des
+applications de calcul scientifiques en Python.  C'est par exemple le
+cas du package NumPy, qui fournit à Python des fonctionnalités proches
 de celles de matlab. Les modules de numpy sont écrit en C, de façon à
@@
 
-\subsection{Optimisation à partir du code python}
+\subsection{Optimisation à partir du code Python}
 
@@
 
-CPython est l'implémentation de référence du langage python. Il en
+CPython est l'implémentation de référence du langage Python. Il en
 existe cependant d'autre comme Jython écrit en Java et IronPython,
-l'implémentation de python pour .NET. Le projet Pypy, quant à lui, a
-pour but d'obtenir une implémentation de python plus performante que
+l'implémentation de Python pour .NET. Le projet Pypy, quant à lui, a
+pour but d'obtenir une implémentation de Python plus performante que
 CPython.
 
-Un programme écrit en python pour CPython peut être exécuté avec Pypy
+Un programme écrit en Python pour CPython peut être exécuté avec Pypy
 sans modification\footnote{Il existe quelques incompatibilités dans
@@
 
-L'outils pythran permet de traduire un module python en C++ pour
+L'outils pythran permet de traduire un module Python en C++ pour
 pouvoir ensuite être compilé en code natif. Afin d'être utilisable par
 pythran, le code source doit seulement être annoté de descriptions
-d'interfaces. Le résultat est un module python en code natif qui peut
-être utilisé depuis un programme python usuel. Pythran ne traduit
-qu'un sous-ensemble de python. En particulier, les classes ne sont pas
+d'interfaces. Le résultat est un module Python en code natif qui peut
+être utilisé depuis un programme Python usuel. Pythran ne traduit
+qu'un sous-ensemble de Python. En particulier, les classes ne sont pas
 gérées. Cela destine pythran à être utilisé pour l'optimisation
@@
 La figure~\ref{RangeXrangePyCpp} permet de comparer les performances
-entre un programme en python et son équivalent en C++.
+entre un programme en Python et son équivalent en C++.
 
@@
   \includegraphics[scale=1]{./Pictures/RangeXrangePyCpp}
-  \caption{Comparaison de performances entre python et C++ compilé avec et sans le flag -O3.}
+  \caption{Comparaison de performances entre Python et C++ compilé avec et sans le flag -O3.}
   \label{RangeXrangePyCpp}
@@
 
-Les programme en python sont les suivants :
+Les programme en Python sont les suivants :
 
@@
 
-où pythonic est le namespace de pythonic++ (cf. \ref{pythonicpp}
-p. \pageref{pythonicpp}).
+où pythonic est le namespace de Pythonic++ (cf. \ref{Pythonicpp}
+p. \pageref{Pythonicpp}).
 
-Les temps en python sont de nouveau mesurés avec timeit.timeit() et
+Les temps en Python sont de nouveau mesurés avec timeit.timeit() et
 les temps en C++ sont mesurés avec clock().
@@
   \includegraphics[scale=1]{./Pictures/RangeXrangeCppO3}
-  \caption{Comparaison entre range et xrange en C++ avec le flag O3.}
+  \caption{Comparaison entre range et xrange en C++ avec le drapeau -O3.}
   \label{RangeXrangeCppO3}
@@
 implémentations en C++ ne sont pas toujours plus rapides que celles en
-python. Dans le cas du xrange, le programme en python est légèrement
+Python. Dans le cas du xrange, le programme en Python est légèrement
 plus performant que celui en C++. Cependant, l'optimisation du
@@
 
-Le code python lu par pythran est représenté sous forme d'un arbre
+Le code Python lu par pythran est représenté sous forme d'un arbre
 syntaxique abstrait (AST) dont les noeuds représentent des opérateurs
@@
 suivant, où la transformation range() -> xrange() n'est pas valide
-dans la boucle for.
+dans la boucle for.\todo{pas en gros problème, juste une gestion des portées... on a déjà l'analyse qui faut.Mais qui de foo(range(100))? a=range(18) ?}
 
@@
 
-\label{pythonicpp}
+\label{Pythonicpp}
 
 C'est à partir de l'AST optimisé que pythran génère le code C++. Pour
-cela, il utilise pythonic++, un ensemble de headers C++ qui
+cela, il utilise Pythonic++, un ensemble de headers C++ qui
 contiennent les équivalents en C++ des opérateurs et des collections
-natives au python.
+natives au Python.
 
-Pour qu'un module de la bibliothèque standard de python puisse être
+Pour qu'un module de la bibliothèque standard de Python puisse être
 utilisé dans du code traduit par pythran, il faut donc qu'un
-équivalent de ce module ai été développé pour pythonic++. Ce n'est pas
+équivalent de ce module ai été développé pour Pythonic++. Ce n'est pas
 encore le cas du module itertools, ce qui empêche d'utiliser des
@@
   possible. Ces transformations ne peuvent donc être implémentées
-  qu'une fois le module itertools disponible dans pythonic++.
+  qu'une fois le module itertools disponible dans Pythonic++.
 \item Certains travaux sont plus simples et permettent d'assimiler les
@@
 \item Écrire des tests.
-\item Implémenter la fonction en C++ dans pythonic++.
+\item Implémenter la fonction en C++ dans Pythonic++.
 \item Tester la fonction pour vérifier qu'elle produit le résultat
-  attendu pour tout les cas de test.
+    attendu pour tout les cas de test.\todo{préciser que c'est fait automatiquement par la suite de test existante ?}
 \item Tester les performances de la fonction et vérifier qu'il n'y a
-  pas d'opération anormalement coûteuse avec l'outils Callgrind.
+    pas d'opération anormalement coûteuse avec l'outils Callgrind.\todo{module callgrind devalgrind ?}
 \end{enumerate}
@@
 
-En python, les generator expression sont plus performantes que
+En Python, les generator expression sont plus performantes que
 l'utilisation de map ou imap. Cependant, cela est lié au coût
-prohibitif d'un appel de fonction en python. Une fois traduit en C++,
+prohibitif d'un appel de fonction en Python. Une fois traduit en C++,
 ce surcoût devrait être moindre et devrait rendre l'implémentation de
@@
 transformation, et effectuer des tests pour vérifier qu'elle apporte
-bien un gain de performances.
+bien un gain de performances.\todo{à priori perte de perfs en python, et gain en C++...}
 
 \subsubsection{Le cas xrange}
+
+\todo[inline]{pas uniquement xrange: map ->imap, zip ->izip,[x for x..] ->(x for x...)}
 
@@
 
-Le analyse et transformations de pythran sont écrites en python. Tout
+Le analyse et transformations de pythran sont écrites en Python. Tout
 code produit doit être conforme à pep8, un ensemble de bonnes
-pratiques python. Pour cela, avant d'être commité, tout code ajouté
+pratiques Python. Pour cela, avant d'être commité, tout code ajouté
 doit être validé par l'outils pep8.
+
+\todo[inline]{et ne pas perturbé les 450+ cas test existants}
 

### 2012-11-26 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
Correction of formating issues (verbatim->lstlsting, texttt, pictures, ...)

@@
 \usepackage{graphicx}
-\usepackage{todo}
+\usepackage[marginpar]{todo}
 
@@
 \item Utiliser les opérateurs propres au langage plutôt que les
-  fonctions lambda : operator.add plutôt que lambda x,y : x+y.
+  fonctions lambda : \lstinline|operator.add| plutôt que \lstinline|lambda x,y : x+y|.
 \end{itemize}
@@
   mémoire.
-\item les itérateurs du module \texttt{itertools} comme \texttt{imap} et \texttt{product}.\todo{toujours utiliser lstlisting / lstinline ou texttt pour le code}
+\item les itérateurs du module \texttt{itertools} comme \texttt{imap} et \texttt{product}.
 \end{itemize}
@@
 Dans chaque cas, le temps de l'exécution de 100 itérations du même
-programme est mesuré grâce au module \texttt{timeit}.\todo{expliquer un peu ce module}
+programme est mesuré grâce au module \texttt{timeit}.
 
@@
 
-La seconde implémentation utilise la fonction imap qui applique une
+La seconde implémentation utilise la fonction \texttt{imap} qui applique une
 fonction à un itérateur pour produire un nouvel itérateur :
 
-\begin{verbatim}
+\begin{lstlisting}
 sum(imap(lambda x:x*x, xrange(N)))
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-La troisième implémentation utilise la fonction map qui applique une
+La troisième implémentation utilise la fonction \texttt{map} qui applique une
 fonction à un itérateur pour produire une liste :
 
-\begin{verbatim}
+\begin{lstlisting}
 sum(map(lambda x:x*x, xrange(N)))
-\end{verbatim}
+\end{lstlisting}
 
@@
 considérablement varier le temps d'exécution. On constate également
-que le simple fait de remplacer map par imap permet un gain de
+que le simple fait de remplacer \texttt{map} par \texttt{imap} permet un gain de
 performance non négligeable dès que l'on manipule un large ensemble de
 données. La même constatation peut être faite quand à l'utilisation de
-xrange plutôt que range ou encore izip plutôt que zip.\todo{pour plus atrd: comment mesurer l'utilisation émmoire et la comparer?}
+\texttt{xrange} plutôt que \texttt{range} ou encore \texttt{izip} plutôt que \texttt{zip}.\todo{pour plus atrd: comment mesurer l'utilisation émmoire et la comparer?}
 
@@
 \begin{figure}[h]
-  \includegraphics[scale=1]{./Pictures/RangeXrangePyCpp}
+  \includegraphics[width=\textwidth]{./Pictures/RangeXrangePyCpp}
   \caption{Comparaison de performances entre Python et C++ compilé avec et sans le flag -O3.}
@@
 
-\begin{verbatim}
+\begin{lstlisting}
   sum(range(N))
-\end{verbatim}
+\end{lstlisting}
 
-\begin{verbatim}
+\begin{lstlisting}
   sum(xrange(N))
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
   pythonic::sum(pythonic::range(N))
-\end{verbatim}
+\end{lstlisting}
 
-\begin{verbatim}
+\begin{lstlisting}
   pythonic::sum(pythonic::xrange(N))
-\end{verbatim}
+\end{lstlisting}
 
-où pythonic est le namespace de Pythonic++ (cf. \ref{Pythonicpp}
+où \texttt{pythonic} est le namespace de Pythonic++ (cf. \ref{Pythonicpp}
 p. \pageref{Pythonicpp}).
 
-Les temps en Python sont de nouveau mesurés avec timeit.timeit() et
-les temps en C++ sont mesurés avec clock().
+Les temps en Python sont de nouveau mesurés avec \texttt{timeit.timeit()} et
+les temps en C++ sont mesurés avec \texttt{clock()}.
 
 \begin{figure}[h]
-  \includegraphics[scale=1]{./Pictures/RangeXrangeCppO3}
-  \caption{Comparaison entre range et xrange en C++ avec le drapeau -O3.}
+  \includegraphics[width=\textwidth]{./Pictures/RangeXrangeCppO3}
+  \caption{Comparaison entre \texttt{range} et \texttt{xrange} en C++ avec le drapeau -O3.}
   \label{RangeXrangeCppO3}
@@
 implémentations en C++ ne sont pas toujours plus rapides que celles en
-Python. Dans le cas du xrange, le programme en Python est légèrement
+Python. Dans le cas du \texttt{xrange}, le programme en Python est légèrement
 plus performant que celui en C++. Cependant, l'optimisation du
 compilateur C++ permet de réduire de façon drastique le temps
-d'exécution. Enfin, le gain réalisé en remplaçant range par xrange se
+d'exécution. Enfin, le gain réalisé en remplaçant \texttt{range} par \texttt{xrange} se
 retrouve également en C++, comme le montre la
@@
 \item Optimisation du code d'origine, par exemple en remplaçant
-  automatiquement des fonctions comme range ou map par xrange et imap.
+  automatiquement des fonctions comme \texttt{range} ou \texttt{map} par \texttt{xrange} et \texttt{imap}.
 \item Traduction en code C++ pour pouvoir être compilé en code natif,
@@
 l'allocation d'une grande quantité de mémoire a un coût prohibitif que
-les transformation du type range -> xrange permettent un gain de
+les transformation du type \texttt{range} -> \texttt{xrange} permettent un gain de
 performances.
@@
 
-Dans le cas de xrange, une condition suffisante pour que la
-transformation range -> xrange soit valide est que le résultat produit
-par range ne soit jamais modifié. Ce qui serait par exemple le cas
+Dans le cas de \texttt{xrange}, une condition suffisante pour que la
+transformation \texttt{range} -> \texttt{xrange} soit valide est que le résultat produit
+par\texttt{range}ne soit jamais modifié. Ce qui serait par exemple le cas
 dans l'exemple suivant\footnote{En supposant que range(100) soit bien
@@
 
-\begin{verbatim}
+\begin{lstlisting}
   for i in range(100):
     f(i)
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
   for i in xrange(100):
     f(i)
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-Par exemple, il est possible d'appliquer la transformation range ->
-xrange au programme suivant. Mais pour pouvoir savoir cela, il faut
+Par exemple, il est possible d'appliquer la transformation \texttt{range} ->
+\texttt{xrange} au programme suivant. Mais pour pouvoir savoir cela, il faut
 vérifier le reste du code pour s'assurer que la liste assignée à la
@@
 
-\begin{verbatim}
+\begin{lstlisting}
   l = range(100)
@@
     f(i)
-\end{verbatim}
+\end{lstlisting}
 
@@
 représenté par son nom, ce qui permet la situation de l'exemple
-suivant, où la transformation range() -> xrange() n'est pas valide
+suivant, où la transformation \texttt{range()} -> \texttt{xrange()} n'est pas valide
 dans la boucle for.\todo{pas en gros problème, juste une gestion des portées... on a déjà l'analyse qui faut.Mais qui de foo(range(100))? a=range(18) ?}
 
-\begin{verbatim}
+\begin{lstlisting}
   def g(n):
@@
 équivalent de ce module ai été développé pour Pythonic++. Ce n'est pas
-encore le cas du module itertools, ce qui empêche d'utiliser des
-fonctions comme imap ou product.
+encore le cas du module \texttt{itertools}, ce qui empêche d'utiliser des
+fonctions comme \texttt{imap} ou \texttt{product}.
 
@@
   particulier, la plupart des transformations de l'AST visent à
-  utiliser des fonctions du module itertools dès que cela est
+  utiliser des fonctions du module \texttt{itertools} dès que cela est
   possible. Ces transformations ne peuvent donc être implémentées
-  qu'une fois le module itertools disponible dans Pythonic++.
+  qu'une fois le module \texttt{itertools} disponible dans Pythonic++.
 \item Certains travaux sont plus simples et permettent d'assimiler les
   concepts et outils nécessaires à la réalisation des travaux plus
-  complexes du même type. Par exemple, la transformation de range en
-  xrange est complexe et devrait donc être abordée après que la
-  transformation de generator expression en imap ait été complétée.
+  complexes du même type. Par exemple, la transformation de \texttt{range} en
+  \texttt{xrange} est complexe et devrait donc être abordée après que la
+  transformation de generator expression en \texttt{imap} ait été complétée.
 \end{itemize}
 
-\subsection{Implémentation du module Itertools}
+\subsection{Implémentation du module \texttt{Itertools}}
 
@@
 
-La méthodologie utilisée pour l'implémentation du module itertools est
+La méthodologie utilisée pour l'implémentation du module \texttt{itertools} est
 une itération des étapes suivantes :
@@
 \begin{itemize}
-  \item imap
-  \item product
-  \item ifilter
+  \item \texttt{imap}
+  \item \texttt{product}
+  \item \texttt{ifilter}
 \end{itemize}
@@
 En Python, les generator expression sont plus performantes que
-l'utilisation de map ou imap. Cependant, cela est lié au coût
+l'utilisation de \texttt{map} ou \texttt{imap}. Cependant, cela est lié au coût
 prohibitif d'un appel de fonction en Python. Une fois traduit en C++,
 ce surcoût devrait être moindre et devrait rendre l'implémentation de
-imap plus performante que l'équivalent des generator expressions.
+\texttt{imap} plus performante que l'équivalent des generator expressions.
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 (g(a,b,c...) for a in A if fa(a) for b in B if fb(a,b) for c in C if fc(a,b,c) ... )
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\begin{verbatim}
+\begin{lstlisting}
 imap(lambda a,b,c... : g(a,b,c...), 
   ifilter(lambda a,b,c... : fa(a) and fb(a,b) and fc(a,b,c)..., product(A,B,C...)))
-\end{verbatim}
+\end{lstlisting}
 
@@
 
-\subsubsection{Le cas xrange}
+\subsubsection{Le cas \texttt{xrange}}
 
@@
 
-Les exemples précédents ont montré que la transformation range ->
-xrange n'est pas toujours valide. Il faudra donc :
+Les exemples précédents ont montré que la transformation \texttt{range} ->
+ \texttt{xrange} n'est pas toujours valide. Il faudra donc :
 
@@
 transformation serait valide. Par exemple, il serait valide de
-remplacer range par xrange dans la boucle for de l'exemple suivant,
+remplacer \texttt{range} par \texttt{xrange} dans la boucle for de l'exemple suivant,
 mais très difficile de le déterminer.
 
-\begin{verbatim}
+\begin{lstlisting}
 def g(n):
@@
 
+\todos
+
 \end{document}

### 2012-11-27 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
More details about timeit. PC configuration(CPU, Kernen --version). Some other \todo

@@
 \begin{abstract}
-    trop de la balouze ce sujet
+    trop de la balouze ce sujet\todo{Ecrire un résumé plus explicite.}
 \end{abstract}
@@
 
-Python, langage interprété et dynamiquement typé, est souvent
-considéré\todo{c'est plus qu'une considération, c'est un fait, cf. par example \url{http://shootout.alioth.debian.org/}} comme moins performant que d'autre langages compilés comme C
-ou C++. Dans ce contexte, la performance est souvent réduite au temps
-d'exécution des opérations du programme,\todo{ça ne veut rien dire :-)} et c'est cette définition
-restreinte qui sera utilisée sous le nom de performance dans la
-suite\footnote{Une meilleurs définition serait celle de Arnold Allen :
-  "The word performance in computer performance means the same thing
-  that performance means in other contexts, that is, it means 'How
-  well is the computer doing the work it is supposed to
+Python, langage interprété et dynamiquement typé, est la plupart du
+temps moins performant que d'autres langages compilés comme C ou
+C++\cite{PythonVsCpp}. On se contente souvent de définir la
+performance d'un programme comme étant sa vitesse d'exécution par
+rapport à un programme semblabme écrit dans un langage différent ou
+utilisant un autre algorithme. C'est cette définition restreinte qui
+sera utilisée sous le nom de performance dans la suite\footnote{Il
+  existe de nombreux travaux sur la manière de mesurer scentifiquement
+  la performance d'un programme, qui s'appuient sur de nombreux
+  indicateurs en plus du temps d'exécution. L'un des spécialiste du
+  domaine résume cela de manière informelle en ces mots : "The word
+  performance in computer performance means the same thing that
+  performance means in other contexts, that is, it means 'How well is
+  the computer doing the work it is supposed to
   do?'"\cite{Allen}}. Toutes les applications n'ont pas un même besoin
@@
 certains de leurs algorithmes soient effectués avec des performances
-supérieures à celles ordinairement obtenues avec Python.\todo{tu peux citer numpy ici, dire que python est de plus en plus utilisé en calcul scientifique}
+supérieures à celles ordinairement obtenues avec Python.
 
@@
 
-Python permet généralement de résoudre un même problème de plusieurs
-façons. Si le choix d'un bon algorithme est crucial, la façon dont est
-implémenté cet algorithme peut faire varier considérablement les
-performances du programme final. C'est pourquoi il existe un ensemble
-de bonnes pratiques permettant d'améliorer les performances de son
-code\cite{PythonSpeed}.\todo{Attention, généralement un dev python ne se pose pas ce genre de questions. On dit qu'en python il existe une unique façon de résoudre un problème...}
+Un des principes de la philosophie python est ``There should be one--
+and preferably only one --obvious way to do it.''\cite{ZenPython} Le
+souci d'optimiser les performances d'un programme python induit
+cependant plus de reflexions quant à la manière de résoudre un
+problème. C'est pour cela qu'il existe des guides donnant bonnes
+pratique pour écrire un programme python le plus performant
+possible\cite{PythonSpeed}.
 
@@
 \item Utiliser les opérateurs propres au langage plutôt que les
-  fonctions lambda : \lstinline|operator.add| plutôt que \lstinline|lambda x,y : x+y|.
+  fonctions lambda : \lstinline|operator.add| plutôt que
+  \lstinline|lambda x,y : x+y|.
 \end{itemize}
@@
 \begin{itemize}
-    \item \texttt{xrange} qui se comporte comme \texttt{range}, mais ne créé pas de liste en
-  mémoire.
-\item les itérateurs du module \texttt{itertools} comme \texttt{imap} et \texttt{product}.
+\item \texttt{xrange} qui se comporte comme \texttt{range}, mais ne
+  créé pas de liste en mémoire.
+\item les itérateurs du module \texttt{itertools} comme \texttt{imap}
+  et \texttt{product}.
 \end{itemize}
@@
 de temps d'exécution entre trois implémentations du calcul de la somme
-des carrés des N premiers entiers.\todo{donner la version de python utilisée, la distrib, la version du noyau, le proc pour que ce soit reproductible}
+des carrés des N premiers entiers.
 
@@
 
-Dans chaque cas, le temps de l'exécution de 100 itérations du même
-programme est mesuré grâce au module \texttt{timeit}.
+Les évaluations de performances sont effectuées sur une machine
+comportant un processeur de type Intel Core i3 4G de RAM, un noyau
+Linux 3.2.0 et python 2.7.3.  Dans chaque cas, le temps de l'exécution
+de 100 itérations du même programme est mesuré grâce au module
+\texttt{timeit}. Le module \texttt{timeit} est un module de python
+fournissant des outils pour mesures les temps d'exécution de morceaux
+de code python. En particulier, la fonction
+\lstinline|timeit.timeit(expr, number = n)| permet de mesurer le temps
+de n executions du code contenu dans \texttt{expr}.
 
@@
 
-La seconde implémentation utilise la fonction \texttt{imap} qui applique une
-fonction à un itérateur pour produire un nouvel itérateur :
+La seconde implémentation utilise la fonction \texttt{imap} qui
+applique une fonction à un itérateur pour produire un nouvel
+itérateur~:
 
@@
 Cette implémentation est moins performante que la première car elle
-utilise des appels de fonction, qui sont coûteux en Python.\todo{argumenter}
+utilise des appels de fonction, qui sont coûteux en
+Python.\todo{argumenter}
 
-La troisième implémentation utilise la fonction \texttt{map} qui applique une
-fonction à un itérateur pour produire une liste :
+La troisième implémentation utilise la fonction \texttt{map} qui
+applique une fonction à un itérateur pour produire une liste :
 
@@
 considérablement varier le temps d'exécution. On constate également
-que le simple fait de remplacer \texttt{map} par \texttt{imap} permet un gain de
-performance non négligeable dès que l'on manipule un large ensemble de
-données. La même constatation peut être faite quand à l'utilisation de
-\texttt{xrange} plutôt que \texttt{range} ou encore \texttt{izip} plutôt que \texttt{zip}.\todo{pour plus atrd: comment mesurer l'utilisation émmoire et la comparer?}
+que le simple fait de remplacer \texttt{map} par \texttt{imap} permet
+un gain de performance non négligeable dès que l'on manipule un large
+ensemble de données. La même constatation peut être faite quand à
+l'utilisation de \texttt{xrange} plutôt que \texttt{range} ou encore
+\texttt{izip} plutôt que \texttt{zip}.\todo{pour plus atrd: comment
+  mesurer l'utilisation émmoire et la comparer?}
 
@@
   \includegraphics[width=\textwidth]{./Pictures/RangeXrangePyCpp}
-  \caption{Comparaison de performances entre Python et C++ compilé avec et sans le flag -O3.}
+  \caption{Comparaison de performances entre Python et C++ compilé
+    avec et sans le flag -O3.}
   \label{RangeXrangePyCpp}
@@
   \includegraphics[width=\textwidth]{./Pictures/RangeXrangeCppO3}
-  \caption{Comparaison entre \texttt{range} et \texttt{xrange} en C++ avec le drapeau -O3.}
+  \caption{Comparaison entre \texttt{range} et \texttt{xrange} en C++
+    avec le drapeau -O3.}
   \label{RangeXrangeCppO3}
@@
 implémentations en C++ ne sont pas toujours plus rapides que celles en
-Python. Dans le cas du \texttt{xrange}, le programme en Python est légèrement
-plus performant que celui en C++. Cependant, l'optimisation du
-compilateur C++ permet de réduire de façon drastique le temps
-d'exécution. Enfin, le gain réalisé en remplaçant \texttt{range} par \texttt{xrange} se
-retrouve également en C++, comme le montre la
+Python. Dans le cas du \texttt{xrange}, le programme en Python est
+légèrement plus performant que celui en C++. Cependant, l'optimisation
+du compilateur C++ permet de réduire de façon drastique le temps
+d'exécution. Enfin, le gain réalisé en remplaçant \texttt{range} par
+\texttt{xrange} se retrouve également en C++, comme le montre la
 figure~\ref{RangeXrangeCppO3}.
@@
 \item Optimisation du code d'origine, par exemple en remplaçant
-  automatiquement des fonctions comme \texttt{range} ou \texttt{map} par \texttt{xrange} et \texttt{imap}.
+  automatiquement des fonctions comme \texttt{range} ou \texttt{map}
+  par \texttt{xrange} et \texttt{imap}.
 \item Traduction en code C++ pour pouvoir être compilé en code natif,
@@
 l'allocation d'une grande quantité de mémoire a un coût prohibitif que
-les transformation du type \texttt{range} -> \texttt{xrange} permettent un gain de
-performances.
+les transformation du type \texttt{range} -> \texttt{xrange}
+permettent un gain de performances.
 
@@
   concepts et outils nécessaires à la réalisation des travaux plus
-  complexes du même type. Par exemple, la transformation de \texttt{range} en
-  \texttt{xrange} est complexe et devrait donc être abordée après que la
-  transformation de generator expression en \texttt{imap} ait été complétée.
+  complexes du même type. Par exemple, la transformation de
+  \texttt{range} en \texttt{xrange} est complexe et devrait donc être
+  abordée après que la transformation de generator expression en
+  \texttt{imap} ait été complétée.
 \end{itemize}
@@
 \item Tester la fonction pour vérifier qu'elle produit le résultat
-    attendu pour tout les cas de test.\todo{préciser que c'est fait automatiquement par la suite de test existante ?}
+  attendu pour tout les cas de test.\todo{préciser que c'est fait
+    automatiquement par la suite de test existante ? -> ah?}
   \item Tester les performances de la fonction et vérifier qu'il n'y a
-    pas d'opération anormalement coûteuse avec l'outils Callgrind.\todo{module callgrind devalgrind ?}
+    pas d'opération anormalement coûteuse avec le module CallGrind de
+    l'outils Vallgrind.
 \end{enumerate}
@@
 transformation, et effectuer des tests pour vérifier qu'elle apporte
-bien un gain de performances.\todo{à priori perte de perfs en python, et gain en C++...}
+bien un gain de performances après traduction en C++.
 
@@
 
-\todo[inline]{pas uniquement xrange: map ->imap, zip ->izip,[x for x..] ->(x for x...)}
+\todo{pas uniquement xrange: map ->imap, zip ->izip,[x for x..] ->(x for x...)}
 
-Les exemples précédents ont montré que la transformation \texttt{range} ->
- \texttt{xrange} n'est pas toujours valide. Il faudra donc :
+Les exemples précédents ont montré que la transformation
+\texttt{range} -> \texttt{xrange} n'est pas toujours valide. Il faudra
+donc :
 
@@
 transformation serait valide. Par exemple, il serait valide de
-remplacer \texttt{range} par \texttt{xrange} dans la boucle for de l'exemple suivant,
-mais très difficile de le déterminer.
+remplacer \texttt{range} par \texttt{xrange} dans la boucle for de
+l'exemple suivant, mais très difficile de le déterminer.
 
@@
 
-\todo[inline]{et ne pas perturbé les 450+ cas test existants}
+\todo{et ne pas perturber les 450+ cas test existants}
 

### 2012-11-29 doc/papers/F2P_Raynaud/RapportCourt/RapportCourt.tex
new reference for python call overhead. Abstract. Rest of the todos.

@@
 \begin{abstract}
-    trop de la balouze ce sujet\todo{Ecrire un résumé plus explicite.}
+  Ce rapport court présente tout d'abord les différents moyens
+  d'améliorer les performances d'un programme en python. En
+  particulier, il montre que l'utilisation d'itérateurs à la place de
+  créations de liste peut diminuer le temps d'exécution d'un programme
+  donné. Il présente également pythran, un outil permettant de
+  traduire du code python en C++ pour créer des modules hautement
+  performants. En plus d'une simple traduction, pythran peut optimiser
+  le programme initial, comme remplacer l'utilisation de listes par
+  l'utilisation d'itérateurs. Les problématiques liées à ce type
+  d'optimisation seront détaillées puis suivies d'un plan de travail
+  visant à les résoudre.
 \end{abstract}
@@
 utilise des appels de fonction, qui sont coûteux en
-Python.\todo{argumenter}
+Python\cite{CallOverhead}.
 
@@
 l'utilisation de \texttt{xrange} plutôt que \texttt{range} ou encore
-\texttt{izip} plutôt que \texttt{zip}.\todo{pour plus atrd: comment
-  mesurer l'utilisation émmoire et la comparer?}
+\texttt{izip} plutôt que \texttt{zip}.
 
@@
 
-Un autre problème vient du fait que dans l'AST, un opérateur est
-représenté par son nom, ce qui permet la situation de l'exemple
-suivant, où la transformation \texttt{range()} -> \texttt{xrange()} n'est pas valide
-dans la boucle for.\todo{pas en gros problème, juste une gestion des portées... on a déjà l'analyse qui faut.Mais qui de foo(range(100))? a=range(18) ?}
+ou encore
 
 \begin{lstlisting}
-  def g(n):
-    return [9001,n,27,42]
+  def f(l):
+    ...
+
   .
@@
   .
-  range = g
-  .
-  .
-  .
-  for i in range(100):
-    f(i)
+  f(range(100))
 \end{lstlisting}
@@
 
-La méthodologie utilisée pour l'implémentation du module \texttt{itertools} est
-une itération des étapes suivantes :
+La méthodologie utilisée pour l'implémentation du module
+\texttt{itertools} est une itération des étapes suivantes :
 
@@
 \item Tester la fonction pour vérifier qu'elle produit le résultat
-  attendu pour tout les cas de test.\todo{préciser que c'est fait
-    automatiquement par la suite de test existante ? -> ah?}
+  attendu pour tout les cas de test. Il suffit pour cela d'écrire les
+  tests unitaires et de les ajouter à la suite de test de pythran.
   \item Tester les performances de la fonction et vérifier qu'il n'y a
@@
 
-\subsubsection{Le cas \texttt{xrange}}
-
-\todo{pas uniquement xrange: map ->imap, zip ->izip,[x for x..] ->(x for x...)}
+\subsubsection{Les transformations en itérateur}
 
 Les exemples précédents ont montré que la transformation
-\texttt{range} -> \texttt{xrange} n'est pas toujours valide. Il faudra
-donc :
+\texttt{range} -> \texttt{xrange} n'est pas toujours valide. Les
+autres transformation du même type comme \texttt{zip} -> \texttt{izip}
+ou \texttt{map} -> \texttt{imap} posent le même problème et devront
+donc être résolu de façon semblable.
 
@@
 
-\todo{et ne pas perturber les 450+ cas test existants}
+Bien évidement, les modifications ne doivent pas apporter de
+regressions. La suite de test de pythran devra donc être utilisée pour
+vérifier que tous les tests unitaires sont toujours réussis après
+avoir implémenté les transformations.
 
@@
 
-\todos
-
 \end{document}

### 2012-12-03 doc/papers/renpar2013/papier.tex
A few updates to the renpar paper.
Take into account (some of) the reviewers comments.

@@
 
-\address{École Normale Supérieure,\\
-Département d'informatique\\
-45 rue d'Ulm, F-75230 Paris Cedex 05 -- France\\
+\address{Télécom Bretagne / École Normale Supérieure,\\
 serge.guelton@telecom-bretagne.eu}
 
-\date{6 septembre 2012}
+\date{15 janvier 2013}
 
@@
 intrinsèques telles \texttt{eval} ne sont pas supportées. Le langage
-intermédiaire utilisé par PyPy, RPython~\cite{rpython2007}, est soumis à
+intermédiaire utilisé par PyPy~\cite{rpython2007} est soumis à
 des contraintes similaires.
@@
 
-  \item Les appels de méthodes sont transformés en appels de fonctions~:
-	\texttt{l.append(v)} devient \texttt{any.append(l,v)}.
+  \item Les appels de méthodes sont convertis en appels de fonctions~:
+	\texttt{l.add(v)} devient \texttt{any.add(l,v)}.
 
@@
 
-Cet article ne vise pas à expliquer en détail le fonctionnement de cette
-étape. On admettra donc que cette analyse permet à Pythran d'accepter des
-fonctions polymorphes, mais uniquement des variables monomorphes.
+Cet article ne vise pas à expliquer en détail le fonctionnement de cette étape.
+On admettra donc que cette analyse permet à Pythran d'accepter des fonctions
+polymorphes. Cependant toutes les variables doivent être monomorphes.
 
@@
 Les codes utilisés sont tirés des bases de tests du défunt projet
-\texttt{unladen swallow}, ou issus de la littérature. La seule
-modification faite est l'ajout de la signature typée de la fonction
-appelée. Quand cela était possible, des annotations OpenMP ont été
-ajoutée. \texttt{extrema} recherche les valeurs extrêmes d'une liste,
-\texttt{matmul} effectue le produit de deux matrices
-carrées, \texttt{mandel} fait un calcul de fractal, \texttt{nqueens}
-résout une instance du problème des n-reines, \texttt{pi\_buffon} utilise
-une méthode de monte-carlo pour calculer les décimales de $\pi$,
-\texttt{quicksort} est une implémentation récursive du tri rapide et
-\texttt{stone} est l'adaptation en Python du vénérable test de
-performances \texttt{wetstone}.
+\texttt{unladen swallow}, ou issus de la littérature. La seule modification
+faite est l'ajout de la signature typée de la fonction appelée. Quand cela
+était possible, des annotations OpenMP ont été ajoutée. \texttt{matmul}
+effectue le produit de deux matrices carrées, \texttt{mandel} fait un calcul de
+fractal, \texttt{nqueens} résout une instance du problème des n-reines,
+\texttt{pi\_buffon} utilise une méthode de monte-carlo pour calculer les
+décimales de $\pi$ et \texttt{stone} est l'adaptation en Python du vénérable
+test de performances \texttt{wetstone}.
 
@@
 Les temps reproduits ici sont obtenus à l'aide de la commande
-\texttt{python setup.py bench} fournie avec l'installation de Pythran.
+\texttt{python setup.py bench} fournie avec l'installation de Pythran à partir du dépot \texttt{https://github.com/serge-sans-paille/pythran}.
 
@@
 l'utilisation de types natifs au lieu d'objets Python pour les opérations
-de base. Le cas de \texttt{quicksort} est à relier au coût prohibitif d'un
-appel de fonction en Python. Le cas d'\texttt{extrema}, implémenté à
-l'aide des intrinsèques \texttt{zip} et \texttt{map}, montrent des
-possibilités d'améliorations non exploitées.
+de base.
 
@@
 	  \hline
-	nqueens		&1.754		&	0.603	&	0.228	&	-		  \\
-	pi\_buffon	&47.083		&	12.08	&	2.676	&	0.561		\\
-	extrema		&0.386		&	0.564	&	0.033	&	-\\
-	stone		&6.956		&	0.204	&	0.023	&	-\\
-	mandel		&976.81		&	6.304	&	5.563	&	2.565\\
-	matmul		&153.91		&	9.061	&	7.314	&	2.263\\
-	quicksort	&3.331		&	0.087	&	0.040	&	-\\
+	nqueens		&1.754s		&	0.603s	&	0.228s	&	-		  \\
+	stone		&6.956s		&	0.204s	&	0.023s	&	-\\
+	pi\_buffon	&47.083s	&	12.08s	&	2.676s	&	0.561s		\\
+	mandel		&976.81s	&	6.304s	&	5.563s	&	2.565s\\
+	matmul		&153.91s	&	9.061s	&	7.314s	&	2.263s\\
   \end{tabular}
 
-  \caption{Comparaison de CPython, PyPy et Pythran, avec ou sans OpenMP.}
+  \caption{Comparaison des temps d'exécution par CPython, PyPy et Pythran, avec ou sans OpenMP.}
   \label{tbl:pythran-vs-cpython}

### 2013-01-23 pythran/intrinsic.py
Fix bug in constant propagation.
Functions with global effects were not considered const.
Intrinsics had a global side effect by default.

@@ -8,7 +8,7 @@ class ReadEffect(object):
 
 class Intrinsic:
     def __init__(self, argument_effects=(UpdateEffect(),) * 11,
-            global_effects=True,
+            global_effects=False,
             return_alias=lambda x: {None}):
         self.argument_effects = argument_effects
         self.global_effects = global_effects
@@ -30,7 +30,9 @@ class Intrinsic:
         return False
 
     def isconst(self):
-        return any(isinstance(x, UpdateEffect) for x in self.argument_effects)
+        return not any(
+                isinstance(x, UpdateEffect) for x in self.argument_effects
+                ) and not self.global_effects
 
 
 class FunctionIntr(Intrinsic):

### 2013-01-23 doc/papers/europar2013/paper.tex
Setup the europar paper.

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{llncs}
%
\begin{document}
%
\title{Pythran: Bringing OpenMP to a Python Subset}

\author{Serge Guelton\inst{1,2} \and Pierrick Brunet\inst{2}}

\institute{
    {\'E}cole Normale Sup{\'e}rieure, Paris, France
    \and
    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France
}

\maketitle

%
\begin{abstract}

    The Python language offers an interesting alternative as a scientific
    computing language thanks to the SciPy package. However, the lack of
    fine-grained parallelism support remains an important shortcoming. This
    paper proposes to add OpenMP directives to the Python language and shows
    that it is possible to turn a module written in a large Python subset
    enhanced with these directive into a native parallel module that runs as
    fast as its naive C equivalent. The approach is validated through the
    Pythran compiler.

\keywords{Python, OpenMP, C++, static compilation, scripting language}
\end{abstract}

%
\section{Introduction}

Since its birth in December 1989, the Python language~\cite{rossum97} has proved
to be useful in various domains, ranging from system administration to web
services, thanks to its dynamicity, expressiveness, rich ecosystem and
battery-included standard library. It is also getting widely used in scientific
computing~\cite{Oliphant2007}, mainly thanks to the SciPy~\cite{scipy} project
that adds Matlab-like functionalities to the language.

The prohibitive performance overhead of the language implied by its interpreted
and highly dynamic nature should prevent its usage as a language for high
performance computing. SciPy overcomes this issue through the usage of low-level
routines written in C or Fortran and encapsulated in Python through
\emph{native} modules. Moreover, its core data structure, the multi-dimensional
array~\cite{numpyarray2011}, was designed so that the underlying data are
available to both native modules and the Python interpreter without conversion
costs.

To make it easier for the user to write native functions, \emph{i.e.}\ without
having to write the glue code that turns Python object into native structures
back and forth, the SciPy package provides the \texttt{weave} module that makes
it possible to bundle C code snippet into Python code and compile then load them
at runtime.

This hybrid approach, based on a mix of interpreted and native code, is getting
wide spread in the Python landscape. Section~\ref{sec:python-hybrid} studies the
existing approaches and shows a critical lack of parallelism support.
Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives to the
Python language in order to add fine-grain parallelism support to Python in the
context of a Python subset called Pythran. Static compilation of this language
into C++ and the underlying runtime is described in
Section~\ref{sec:python-static-translation}.
Section~\ref{sec:python-optim} presents several optimizations applied by the compiler to
optimized generated code with respect to parallelism. The Pythran compiler is
benchmarked on several scientific applications and compared to both interpreted
and native code in Section~\ref{sec:validation}.

%=========================================================
\section{Hybrid Computations in Python}\label{sec:python-hybrid}
%=========================================================

In the context of interpreted language, an hybrid computation is defined
in~\cite{dongara2007} as a computation where part of the code is interpreted,
and part of the code is executed natively.

It is now common for scripting language to have C bindings. To take advantage of
compiled code, Python developers must write C/C++ functions and write the glue
code using the Python C API~\cite{pythoncapi}. Tools have been developed to
relieve the user from writing the encapsulating code, notably
Swig~\cite{swig2003} that relies on an enhanced interface specification, or
\texttt{boost::python}~\cite{boostpython2007} that relies on C++ facilities to
guide translation.

An opposite approach consists in using the host language --- herein Python ---
to describe both parts of the system, and to let an automated tool perform the
translation to native code of a part of the application, generally the
computation-intensive one. That way, developers not familiar with lower level
language or not eager to invest the additional development time can still
benefit from a fair performance boost. This approach has been subject to many
studies that can be classified based on their backward-compatibity with the host
language.

\subsection{Backward-Incompatible Languages}

A constraining (from the performance point of view) aspect of the Python
language is its type system. It has several consequences, among which the fact
that each method call is resolved dynamically, including a simple add
operation! It comes at no surprise that many approaches extend the Python
language to add a static typing overlay. Likewise, there is only two types of
integers (64bits integers and multi-precision integers) and one type of floating
point type (double-precision floats) in Python, while using a type with the
appropriate size may lead to significant performance boost.

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, among which the possibility to define
parallel loops, but no task parallelism. Plw~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code.
The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
compiler uses additional type information to generate sequential LLVM bytecode.
PyCUDA and PyOpenCL~\cite{klockner2012} also mixes python with kernels
embeded as raw strings to target accelerators.

The core issue of these approaches is that they imply to modify, in a more or
less trivial way, the original code. They require to invest in a new language,
and the long-term preservation of this investment is not ensured.

%=========================================================
\section{OpenMP Semantic Adaptations}\label{sec:python-openmp}
%=========================================================

\subsection{Variable Scope}

\subsection{Iterators}

%=========================================================
\section{Static Translation of Python Programs}\label{sec:python-static}
%=========================================================

\subsection{Directive Oblivious Translation}

\subsection{Runtime Support}

\subsubsection{Shared References}

\subsubsection{Iterator Category}

%=========================================================
\section{Optimizations for Parallelism}\label{sec:python-optim}
%=========================================================

\subsection{Limiting Copies}

\subsection{Transfer Costs}

%=========================================================
\section{Validation}\label{sec:validation}
%=========================================================

\subsection{Comparison with CPython}

\subsection{Parallelism Gain}

\subsection{Comparison with Cython}

%=========================================================
\section{Conclusion and Future Work}
%=========================================================

\bibliographystyle{splncs03}
\bibliography{biblio}

\end{document}
% vim:spell spelllang=en
% vim:tw=80

### 2013-01-25 doc/papers/europar2013/paper.tex
Finish related works.

@@
 This hybrid approach, based on a mix of interpreted and native code, is getting
-wide spread in the Python landscape. Section~\ref{sec:python-hybrid} studies the
+wide spread in the Python landscape. Section~\ref{sec:python-parallel} studies the
 existing approaches and shows a critical lack of parallelism support.
@@
 %=========================================================
-\section{Hybrid Computations in Python}\label{sec:python-hybrid}
+\section{Parallel Computations in Python}\label{sec:python-parallel}
 %=========================================================
+
+In the manycore area, relying on parallelism to balance the performance
+limitations of scripting languages, as described in~\cite{choy05} in the context
+of the MATLAB language, or in~\cite{mals07} in the context of the R language.
+In the context of Python, most approaches have focused on processes.
+
+\subsection{Python and Parallelism}
+
+Parallel computations are supported by the Python standard library through the
+\texttt{multiprocessing} module. It spawns several interpreters that can
+communicate through IPC, using Python built-in object serialization. This
+approach is only viable for computation intensive application, but the
+communication and synchronization overheads are much greater than a light-weight
+processes approach.
+
+Although the standard \texttt{threading} module makes it possible to start
+several light-weight threads within the same interpreter, this approach is not
+relevant for HPC, because of a specificity of CPython, the \emph{Global
+Interpreter Lock}~\cite{gil2012}~\footnote{Other Python interpreters, such as
+IronPython or Jython, do not have a GIL.} This lock ensures only one thread is
+active at a time in the interpreter. While it make it possible to have
+cooperative thread, say for a GUI, it does not take advantage of multiple cores.
+However, there are two notable exceptions: the GIL is released on I/O, and the
+GIl does not prevent the use of threads inside native modules, where the user
+has full control.
+
+As a consequence, Python developers need to write multi-threaded native
+modules in order to fully benefit from multiple cores. This lead to a kind of
+computations referred as hybrid computations.
+
+%There have been several approaches to replace the GIL by Transactional
+%Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
+%mainstream interpreters.
+
+\subsection{Hybrid Computations}
 
@@
 It is now common for scripting language to have C bindings. To take advantage of
-compiled code, Python developers must write C/C++ functions and write the glue
-code using the Python C API~\cite{pythoncapi}. Tools have been developed to
-relieve the user from writing the encapsulating code, notably
-Swig~\cite{swig2003} that relies on an enhanced interface specification, or
+compiled code, and to overcome the GIL limitations, Python developers must write
+parallel C/C++ functions and the associated boilerplate based on the Python C
+API~\cite{pythoncapi}. Tools have been developed to relieve the user from
+writing the encapsulating code, notably Swig~\cite{swig2003} that relies on an
+enhanced interface specification, or
 \texttt{boost::python}~\cite{boostpython2007} that relies on C++ facilities to
@@
 translation to native code of a part of the application, generally the
-computation-intensive one. That way, developers not familiar with lower level
-language or not eager to invest the additional development time can still
-benefit from a fair performance boost. This approach has been subject to many
-studies that can be classified based on their backward-compatibity with the host
-language.
+computation-intensive one where parallelism has been expressed in some ways.
+Thus, developers not familiar with lower level language or not eager to invest
+the additional development time can still benefit from a fair performance boost.
+This approach has been subject to many studies that can be classified based on
+their backward-compatibity with the host language.
 
-\subsection{Backward-Incompatible Languages}
+\subsubsection{Backward-Incompatible Approaches}
 
@@
 parallel loops, but no task parallelism. Plw~\cite{dongara2007} proposes a
-similar approach that mixes Python with C, using raw strings to hold the C code.
-The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
-compiler uses additional type information to generate sequential LLVM bytecode.
-PyCUDA and PyOpenCL~\cite{klockner2012} also mixes python with kernels
-embeded as raw strings to target accelerators.
+similar approach that mixes Python with C, using raw strings to hold the C code
+and the parallel directives, that are limited to parallel for loops.  The
+numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
+uses additional type information to generate sequential LLVM bytecode.  PyCUDA
+and PyOpenCL~\cite{klockner2012} also mixes python with kernels embeded as raw
+strings to target accelerators.
 
@@
 and the long-term preservation of this investment is not ensured.
+
+\subsubsection{Backward-Compatible Approaches}
+
+Most backward-compatible approaches also require to modify the input program,
+but they do not extend the Python language, rather restrict it. As a
+consequence, they remain compatible with the original language and do not suffer
+from the drawbacks of the previous approach.
+
+Copperhead~\cite{shedskin2006} is a functional, data parallel language embedded in
+Python. It uses n-uplets, Numpy arrays and lists as its core data structure and
+prohibits usage of many control-flow operators such as loops, enforcing the use
+of the \texttt{map}, \texttt{filter} or \texttt{reduce} intrinsics to exhibit
+parallelism. In returns, it can be efficiently compiled to either CUDA or C++
+with calls to the Thrust~\footnote{\emph{cf}.\
+\texttt{http://thrust.github.com/}} library. Python decorators are used to
+identify hot-spots that are dynamically compiled to native code.
+
+Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
+Shedskin~\cite{shedskin2006}, a Python to C++ compiler also are viable way to
+turn regular Python codes into optimized ones, but they do not offer support for
+fine-grained parallelism beyond what the standard library
+proposes.~\footnote{Support for STM in PyPy is a work in
+progress.}~\footnote{Shedskin is compatible with several Python modules that
+provide coarse-grained, fork-based parallelism.}
+
+This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
+with a Python subset called Pythran to bring fine-grain parallelism to Python
+while being backward compatible with both the Python language, and the
+sequential algorithm. It means that:
+\begin{enumerate}
+    \item Any Pythran code can be run (sequentially), with no module dependency or code change,
+        by any Python interpreter.
+    \item Parallelism is explicit and incrementally added to the original code
+        through directives.
+\end{enumerate}
 
@@
 
-\subsection{Parallelism Gain}
-
 \subsection{Comparison with Cython}

### 2013-01-27 doc/papers/europar2013/paper.tex
Take into account Pierrick's review.

@@
 
-In the manycore area, relying on parallelism to balance the performance
+In the many cores area, turning to parallelism to balance the performance
 limitations of scripting languages, as described in~\cite{choy05} in the context
-of the MATLAB language, or in~\cite{mals07} in the context of the R language.
-In the context of Python, most approaches have focused on processes.
+of the MATLAB language, or in~\cite{mals07} in the context of the R language,
+seems legitimate. In the context of Python, most approaches have
+focused on fork-based parallelism.
 
@@
 
-An opposite approach consists in using the host language --- herein Python ---
+An opposite approach consists in using the host language ---herein Python---
 to describe both parts of the system, and to let an automated tool perform the

### 2013-01-29 doc/papers/europar2013/paper.tex
WIP

@@
 into C++ and the underlying runtime is described in
-Section~\ref{sec:python-static-translation}.
+Section~\ref{sec:python-static}.
 Section~\ref{sec:python-optim} presents several optimizations applied by the compiler to
@@
 
-\subsection{Variable Scope}
+OpenMP is a standard API for parallel programming for Fortran, C and C++.
 
-\subsection{Iterators}
+\subsection{Directives}
+
+\subsubsection{Variable Scope}
+
+\subsubsection{Iterators}
+
+\subsection{Runtime Library}
+
+\subsection{Validation}
+
+\cite{wang2012}
 

### 2013-01-29 doc/papers/europar2013/paper.tex
Draft version.

@@
 %
+\usepackage{listings}
+\usepackage{url}
+\usepackage{todo}
+\newcommand\see{\emph{cf.\ }}
+
+\usepackage{tikz}
+\usetikzlibrary{positioning}
+\usetikzlibrary{shadows}
+\usetikzlibrary{arrows}
+
+\lstset{language=python}
+
 \begin{document}
@@
 This hybrid approach, based on a mix of interpreted and native code, is getting
-wide spread in the Python landscape. Section~\ref{sec:python-parallel} studies the
-existing approaches and shows a critical lack of parallelism support.
-Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives to the
-Python language in order to add fine-grain parallelism support to Python in the
-context of a Python subset called Pythran. Static compilation of this language
-into C++ and the underlying runtime is described in
-Section~\ref{sec:python-static}.
-Section~\ref{sec:python-optim} presents several optimizations applied by the compiler to
-optimized generated code with respect to parallelism. The Pythran compiler is
-benchmarked on several scientific applications and compared to both interpreted
-and native code in Section~\ref{sec:validation}.
+wide spread in the Python landscape. Section~\ref{sec:python-parallel} studies
+the existing approaches and shows a critical lack of parallelism support.
+Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives
+to the Python language in order to add fine-grain parallelism support to Python
+in the context of a Python subset called Pythran. Static compilation of this
+language into C++ and the underlying runtime is described in
+Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
+scientific applications and compared to both interpreted and native code in
+Section~\ref{sec:validation}.
 
@@
 fine-grained parallelism beyond what the standard library
-proposes.~\footnote{Support for STM in PyPy is a work in
-progress.}~\footnote{Shedskin is compatible with several Python modules that
-provide coarse-grained, fork-based parallelism.}
+proposes.
+%~\footnote{Shedskin is compatible with several Python modules that
+%provide coarse-grained, fork-based parallelism.}
 
@@
 
-OpenMP is a standard API for parallel programming for Fortran, C and C++.
+OpenMP is a standard API for parallel programming for Fortran, C and C++. It
+consists in a set of parallelizing directives and a few runtime library calls.
+If OpenMP is not activated, the directives are ignored, thus enabling
+incremental parallelization of the original source code while keeping the
+original code structure.
+
+The languages targeted by OpenMP are statically compiled. This section studies
+the semantic adaptation required to use the same API for a scripting language
+such as Python.
 
@@
 
-\subsubsection{Variable Scope}
+OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them apply
+on structured blocks, represented by an instruction inC/C++ and delimited by
+comment in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
+attached to a specific instruction.
 
-\subsubsection{Iterators}
+While Pythran does have a decorator mechanism~\footnote{\see PEP 319,
+\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation fo
+Python decorators.}, it only applies to functions, methods and classes and does
+not make it possible to attach decorations to random instructions.
+PLW~\cite{dongara2007} suggests to use string instructions to hold such
+decorations. As Python does not have anonymous block~\footnote{\see PEP340,
+    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
+anonymous block support in Python.}, one has to create a dummy \texttt{if 1:}
+instruction to apply an annotation to a whole block. Alternatively, the syntax
+\texttt{if 'my annotation':} is also supported.
+Listing~\ref{lst:instruction-annotation} shows how to annotate a parallel loop
+with reduction this way.
+
+\begin{figure}
+    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
+    loop with reduction annotated with OpenMP directive.}]
+'omp parallel for reduction(+:r)'
+for auto in my_container:
+  r+=x
+  \end{lstlisting}
+\end{figure}
+
+Many OpenMP annotations are parametrized by clauses that list variables,
+specifying their behavior with respect to parallel regions, e.g.\
+\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
+variables that have already been declared. However, there is no variable
+declaration in Python, and all variables assigned in a function have the
+function scope. Consequently, all variables referred to in a function are to be
+considered when building such variable lists: there is no concept of variable
+local to a block. Using the \texttt{default(none)} clause is possible to ensure
+that no variable gets forgotten when building such lists, as a compile-time
+error is raised if a variable is referred in a block but not in the associated
+parallel region.
+
+The \texttt{reduction(\emph{operator}: \emph{list}} is used to handle data
+dependencies when performing a parallel reduction. The supported operator
+depends on the input and backend language: Python has \texttt{min}/\texttt{max} operators but
+C/C++ have not. Python does not have \texttt{\&\&} or \texttt{||}
+operators~\footnote{Although similar, the \texttt{and} and \texttt{or} keywords
+have a special meaning in Python.} while C/C++ have.
+
+\subsubsection{Parallel For and Iterators}
+
+The core directive to handle data parallelism is the \texttt{for} directive that
+distribute the iteration space of the associated loop among the existing
+threads. To be compatible with OpenMP, the loop iteration space must be
+described by a random access iterator with a total order, that is it is possible
+to increase the iterator by any integer and to decide which iterator is the
+greatest among two. The integer used as loop indices in Fortran and C satisfy
+this conditions, so do C++ iterators with the \texttt{random\_access\_tag}
+trait. But a Python iterator only advance by a step of one until it is
+exhausted, in which case it raises an exception. They only have forward iterator
+behavior.
+
+The extension of Python iterator to random access iterators is direct for the
+standard containers: \texttt{list}~\footnote{Python \texttt{list}s behave as C++
+\texttt{vector}s.}, \texttt{set} or \texttt{dict}.
+
+Python also generalizes the concept of iterators through generators, which also
+behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
+\emph{stop}, \emph{step})}, successively yields value starting from
+\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}, is
+easily extended to support random access, but it is generally does not make
+sense to use a generator as a loop iterator, as the relation between two
+random states of the iterator may be of an arbitrary complexity.
+
+A particular class of generator behaves like adaptor, in the sense that they
+apply a particular expression on each value of the iterator. It is the case of
+the \texttt{enumerate} builtin that yields each element of the enumerated
+iterator associated with its index. Generator expressions such as \texttt{(x*x
+for x in l)} have the same behavior. These generators are random iterators only
+if their input iterator is a random iterator itself.
+
+Finally, if the iterator yields a tuple, it is possible to unpack it inside the
+\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
+all the unpacked variables are considered as iterators, especially with respect
+to default privatization rules. In the given example, \texttt{i} and \texttt{v}
+are private, and the parallel iteration is valid because the input of
+\texttt{enumerate} is a list, which allows random access iteration.
+
+\begin{figure}
+    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
+    loop with tuple unpacking.}]
+'omp parallel for'
+for i,v in enumerate([2, 3, 5, 7, 11]):
+  print i, ':', v
+  \end{lstlisting}
+\end{figure}
 
@@
 
+The OpenMP runtime library makes it possible to retrieve the active thread id,
+whether dynamic scheduling is enabled and so on. All the functions are declared
+in the \texttt{<omp.h>} header, and have a default behavior when OpenMP is not
+activated.
+
+Providing a biding to these library in the Python does not raise
+particular problems, as the signature of these functions only involves passing
+integers as parameters or return values, except for the mutex manipulation. In
+that case an opaque type is used to represent the native type.
+
+The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
+activated, and can be used for portability with system where OpenMP is not
+available. Python does not have a preprocessor, but is possible to catch the
+import exception if the \texttt{omp} module is not found.
+
 \subsection{Validation}
 
-\cite{wang2012}
+A validation suite for OpenMP has been proposed in~\cite{wang2012} for C and
+Fortran. We have ported it to Python, and we also extended it to validate the
+corner cases specific to Python described in this section. An example of such
+test case is given in Listing~\ref{lst:openmp-validation}.
+
+\begin{figure}
+    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
+    OpenMP validation test case.}]
+def omp_parallel_for_if():
+  import omp
+  using = num_threads = 0
+  sum = sum2 = 0
+  LOOPCOUNT=1000
+  'omp parallel for private(i) if(using == 1)'
+  for i in range(LOOPCOUNT+1):
+    num_threads = omp.get_num_threads()
+    sum+=i
+  known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
+  return known_sum == sum and num_threads == 1
+    \end{lstlisting}
+\end{figure}
+
+We have used the Pythran tool described in the following section to turn each
+Python test function into a C++ function with the same directives and runtime
+calls. Apart from the \texttt{threadprivate} directive and the
+\texttt{collapse(\emph{n}} clause, all tests were successful.
+\texttt{threadprivate} directives were held by global variables not supported in
+Pythran~; and the C++ code generated by Pythran does not preserve the perfect
+loop nesting required by the \texttt{collapse} clause.
 
@@
 
-\subsection{Directive Oblivious Translation}
+Pythran is subset of Python focused on scientific computing. It is implicitly
+statically typed and support most Python constructs except those that involve
+introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
+\texttt{eval}). A few standard module are supported in addition to the core
+language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
+partially supported. The associated compiler turns Pythran code, eventually
+annotated with OpenMP and exported function type annotation, into C++ code. Its
+processing is summarized in Figure~\ref{fig:pythran-compiler}. The purpose of
+this section is not to dig into the internal of the compiler, but rather to
+focus on the impact of the parallelism layer, especially on the runtime library,
+\texttt{pythonic}.
+
+\begin{figure}
+
+\centering
+\begin{tikzpicture}[
+file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center},
+    tool/.style={draw=black!50,fill=black!10,circle, align=center}]
+    \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
+\node[file] (annotation)	   [right=of python] {\textbf{Type Info}};
+\node[file] (omp)	   [left=of python] {\textbf{OpenMP}};
+\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
+\node[file] (cxx) [below=of pythran] {\textbf{C++}};
+\node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
+\node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
+\node[tool] (gxx) [below=of cxx] {\textbf{g++}};
+\node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
+
+\draw[very thick, ->] (python) -- (pythran);
+\draw[very thick,dashed, ->] (annotation) -- (python);
+\draw[very thick,dashed, ->] (omp) -- (python);
+\draw[very thick, ->] (pythran) -- (cxx);
+\draw[very thick, ->] (cxx) -- (gxx);
+\draw[very thick, ->] (boost) -- (gxx);
+\draw[very thick, ->] (pythonic) -- (gxx);
+\draw[very thick, ->] (gxx) -- (so);
+\end{tikzpicture}
+\caption{Pythran compiler workflow.}
+\label{fig:pythran-compiler}
+\end{figure}
 
@@
 
-\subsubsection{Shared References}
+The standard library of C++11, unlike its C++03 version, is thread-safe. Pythran
+runtime is based on the STL and provides the same guaranty, so that the data
+dependences in the original code remains explicit.
 
-\subsubsection{Iterator Category}
+A critical point when designing the pythonic runtime was memory management, the
+very same aspect that lead to the use of a GIL in CPython. Pythran handles the
+problem by refusing recursive types, which makes it possible to solely rely on
+reference counting for memory management. It can be implemented through the
+thread-safe shared pointer mechanism provided in C++11 standard library.
 
-%=========================================================
-\section{Optimizations for Parallelism}\label{sec:python-optim}
-%=========================================================
+Using shared references simplifies memory management, but the counterpart is an
+extra atomic operation for each copy. It limits parallelism, so reducing the
+number of copies remains an important goal. The move semantic introduced in
+C++11 avoids a few copies when working on temporary objects, but argument
+passing may be a trouble. To avoid a reference increment, one can use pass
+parameters by reference. Using an interprocedural memory effect analysis, Pythran
+determines for each argument whether it must be passed using reference or const
+reference to prevent this extra cost.
 
-\subsection{Limiting Copies}
+\subsection{Directive Oblivious Translation}
+
+During Python to C++ translation, Pythran adopts a blind strategy: it does not
+understand the semantic of the annotations. Instead, it just split each
+annotation into a context-sensitive part ---the variable names--- and a
+context-insensitive part ---the clauses---, and attach them to the proper
+instruction in the AST. It then ensures a bijective translation between Python
+instruction and generated C++ instruction so that the OpenMP directive is
+regenerated on the proper instruction. The same approach is used at the
+expression level, to be compatible with the \texttt{if} clause.
 
 \subsection{Transfer Costs}
+
+When passing containers from Python to C back and forth, a copy of the whole
+data is made to turn the type agnostic, non-contiguous Python data into dense
+typed ones. This extra copy implies an extra cost that is not negligible: the
+time taken to pass two lists of float from Python to C++ is only twice less than
+the time taken to compute the dot product of the same lists directly in Python.
+Following Amdahl's law, this transfer cost greatly hinders parallelization, and
+is a well known issue.
+
+The traditional solution is to use a native type that exposes a Python interface
+and has a constant translation cost. Numpy's \texttt{ndarray} is a typical
+implementation of this concept and is the keystone of the Scipy and Numpy
+packages. Basically, a numpy \texttt{ndarray} is a raw C pointer that is exposed
+at the Python level. As a tool for scientific computing, Pythran supports such a
+structure.
 
@@
 
-\subsection{Comparison with CPython}
+To validate the approach proposed in this paper, we first performed a simple
+experiment. Starting from the C code of a small geomatics application, we
+successively parallelized it with OpenMP, turned it into Python with a Pythran
+compatible kernel and parallelized the Python version using the approach
+described in this paper. We also measurer the SLOC for the two versions.
+Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
+laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It
+non only emphasises the use of a high level language prototyping, but also shows
+that it is possible to turn this prototype into reasonably efficient code
+through Pythran. It is also possible to prototype the parallel version while
+remaining at the Python level.
 
-\subsection{Comparison with Cython}
+\begin{table}
+
+    \centering
+    \begin{tabular}{|l|l|l|l|}
+        \hline
+        Language & SLOC & wall time (s) [gcc] & wall time (s) [clang]\\
+        \hline
+        C       & 102   & 25 & 22 \\
+        Python  & 30    & 676 & --\\
+        Pythran & 30    & 38 &  22 \\
+        Pythran+OMP    & 30    & 11 & no OpenMP support\\
+        \hline
+    \end{tabular}
+
+    \caption{Comparison of several versions of the Hyantes prototype.}
+    \label{tbl:hyantes}
+
+\end{table}
+
+We then compare Pythran with Cython. The approach share some similarities: they
+both generate code written in a rather low level language, with OpenMP
+directives. However, the inputs differ as Cython requires more typing
+information that Pythran to generate efficient code, and Cython is not
+backward-compatible with Python. They both translate explicit fine-grained
+parallelism through code annotations or specific python constructs,
+respectively.
+
+Pythran exposes the full OpenMP interface to the user, thus enabling both data
+and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
+not the case in Cython, where:
+
+\begin{itemize}
+    \item Only loops can be made parallel, through the use of a dedicated
+        intrinsic \texttt{prange} as loop iterator. To use more advanced
+        parallelism construction, one must rely on the possibility to write raw C
+        code embedded in cython.~\todo{SG->PB: tu m'explique ce truc de C dans
+        cython ?}
+
+    \item Cython automatically infers reductions and variable privacy, giving
+        less control to the end-user.
+
+    \item As Cython allows the mix of Python and Cython code, the user is
+        responsible from releasing the GIL.
+
+    \item The code included in a parallel region interacts with the Python code
+        in a restricted manner. For instance, it is impossible to use a function
+        imported from a Python module, while it is still possible to use native
+        C functions.
+\end{itemize}
+
+Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrates the
+difference between the two approaches. It shows that while being more integrated
+in Python, the Cython approach is more verbose.
+
+\begin{figure}[ht]
+
+    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython Impelmentation
+    of a Parallel Reduction.}]
+import math
+cimport cython
+from cython.parallel import parallel, prange
+from libc.math cimport sqrt
+def sum_sqrt():
+    cdef float r=0
+    cdef int i
+    with nogil, parallel():
+        for i in prange(10000000):
+            r+= sqrt(i)
+    return r
+\end{lstlisting}
+\end{figure}
+%
+\begin{figure}[ht]
+    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython Impelmentation
+    of a Parallel Reduction.}]
+#pythran export sum_sqrt()
+import math
+def sum_sqrt():
+    r=0
+    "omp parallel for private(i) reduction(+:r)"
+    for i in xrange(10000000):
+        r+= math.sqrt(i)
+    return r
+    \end{lstlisting}
+\end{figure}
+
+The performance of the two approaches is illustrated in
+Figure~\ref{fig:cython-pythran}.
+
+\begin{figure}[ht]
+    \includegraphics[width=\textwidth]{cython}
+    \caption{Comparison of Cython en Pythran generated code performance.}
+    \label{fig:cython-pythran}
+\end{figure}
+
+The benchmarked codes are typical mathematical, image-processing or linear
+algebra kernels. All these kernels have been written in Cython and Python
+---compatible with Pythran--- and annotated through the mechanism of each
+language to exhibit parallelism, then compiled into native code using GCC 4.1
+and the \texttt{-O2 -fopenmp} flag. Their execution time when called from the
+Python interpreter is measured through the \texttt{timeit} module. All results
+are normalized to Cython sequential execution time.
+
+This figure shows our inability to select proper benchmarking code, and to run
+them on a real multiprocessor machine, but this is gonna change!~\todo{SG->PB:
+c'est pour toi !}
 
@@
 %=========================================================
+
+This paper studies the relevancy of adding OpenMP annotations to Python code. It
+shows that providing minor semantic adaptations, these annotations are relevant
+and makes it possible for Python code to benefit from multi-core performance
+boost without worrying about the Global Interpreter Lock.
+
+To achieve this goal, a translator from a subset of the Python language to C++
+has been designed. This translator can turn regular Python modules
+annotated with OpenMP directives and a few type annotations into native parallel
+module. Both the input module and the generated one remain compatible with the
+standard interpreter.
+
+The approach is compared with Cython, an extension of the Python language used
+to generate native module with an hybrid Python-C syntax that also provide means
+to exhibits fine grained parallelism. Modules generated by Pythran outperform
+modules generated by Cython for various kernels and both benefit from the
+parallel boost.
+
+Future works include the extension of the approach to OpenMP 4 and the
+\texttt{target} clause that should make it possible to target MIC from Python.
+There is also many vectorization possibility in Python, especially in the list
+comprehension construction, that needs to be explored.
+
+%=========================================================
+\section*{Acknowledgments}
+%=========================================================
+
+This work has received fundings from the \texttt{Silkan} company and the French
+ANR through the CARP project. The authors would like to thanks Adrien
+\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime.
+Fabien \textsc{Dagnat} and Mehdi \textsc{Amini} also contributed to this paper
+through their many valuable advices.
 

### 2013-01-29 doc/papers/europar2013/cython.py
Draft version.

@@ -1,0 +1,38 @@
+#!/usr/bin/env python
+import numpy as np
+import matplotlib.mlab as mlab
+import matplotlib.pyplot as plt
+import math
+from matplotlib.backends.backend_pdf import PdfPages
+
+x = np.array([[44.2, 1410, 1310, 260, 9.44, 6.21],[29.2, 744, 704, 143, 5.13, 5.12], [994, 0, 1320, 888, 14.1, 5.59],[118, 0, 711, 481, 8.21, 5.83]])
+
+y = np.zeros(x.shape)
+
+for i,j in enumerate(x):
+    y[i] = j/x[2]
+
+for i in xrange(y.shape[0]):
+    for j in xrange(x.shape[1]):
+        if math.isinf(y[i][j]):
+            y[i][j] = x[i][j]/x[0][j]
+        if math.isnan(y[i][j]):
+            y[i][j] = 0
+    
+
+color_ = ["red", "blue", "orange", "purple"]
+p = range(4)
+for i,j in enumerate(y):
+    p[i] = plt.bar(range(i,y.size+len(y)+1, len(y) +1),j, color=color_[i])
+
+#plt.xlabel('Comparison between cython and pythran')
+plt.xticks(range(2,y.size+len(y)+1, len(y) +1), ('fbcorr', 'pi_buffon', 'matmul', 'mandel', 'pairwise', 'convolution') )
+plt.ylabel('execution time')
+#plt.title(r'$comparaison between cython and pythran with/without parallelism$')
+plt.grid(True)
+plt.legend( (p[0][0], p[1][0], p[2][0], p[3][0]), ('Pythran', 'Pythran + OMP', 'Cython', 'Cython + OMP') , bbox_to_anchor=(0., 1.02, 1., .102), loc=2, ncol=2, mode="expand", borderaxespad=0.)
+pp = PdfPages('cython.pdf')
+plt.savefig(pp,format='pdf')
+pp.close()
+
+

### 2013-02-02 doc/papers/europar2013/paper.tex
Few cosmetic modifications.

@@
 OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them apply
-on structured blocks, represented by an instruction inC/C++ and delimited by
+on structured blocks, represented by an instruction in C/C++ and delimited by
 comment in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
@@
 for auto in my_container:
-  r+=x
+  r += x
   \end{lstlisting}
@@
 
-The \texttt{reduction(\emph{operator}: \emph{list}} is used to handle data
+The \texttt{reduction(\emph{operator}: \emph{list})} is used to handle data
 dependencies when performing a parallel reduction. The supported operator
@@
 threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order, that is it is possible
+described by a random access iterator with a total order, i.e.\ it is possible
 to increase the iterator by any integer and to decide which iterator is the
@@
 \emph{stop}, \emph{step})}, successively yields value starting from
-\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}, is
-easily extended to support random access, but it is generally does not make
+\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}. It is
+easily extended to support random access, but it generally does not make
 sense to use a generator as a loop iterator, as the relation between two
@@
 'omp parallel for'
-for i,v in enumerate([2, 3, 5, 7, 11]):
+for i, v in enumerate([2, 3, 5, 7, 11]):
   print i, ':', v
@@
 activated, and can be used for portability with system where OpenMP is not
-available. Python does not have a preprocessor, but is possible to catch the
+available. Python does not have a preprocessor, but it is possible to catch the
 import exception if the \texttt{omp} module is not found.
@@
   sum = sum2 = 0
-  LOOPCOUNT=1000
+  LOOPCOUNT = 1000
   'omp parallel for private(i) if(using == 1)'
-  for i in range(LOOPCOUNT+1):
+  for i in range(LOOPCOUNT + 1):
     num_threads = omp.get_num_threads()
-    sum+=i
+    sum += i
   known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
@@
 calls. Apart from the \texttt{threadprivate} directive and the
-\texttt{collapse(\emph{n}} clause, all tests were successful.
+\texttt{collapse(\emph{n})} clause, all tests were successful.
 \texttt{threadprivate} directives were held by global variables not supported in
-Pythran~; and the C++ code generated by Pythran does not preserve the perfect
+Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
 loop nesting required by the \texttt{collapse} clause.
@@
 def sum_sqrt():
-    cdef float r=0
+    cdef float r = 0
     cdef int i
@@
         for i in prange(10000000):
-            r+= sqrt(i)
+            r += sqrt(i)
     return r
@@
 def sum_sqrt():
-    r=0
+    r = 0
     "omp parallel for private(i) reduction(+:r)"
     for i in xrange(10000000):
-        r+= math.sqrt(i)
+        r += math.sqrt(i)
     return r
@@
 c'est pour toi !}
+
+All the source codes used for these benchmarks are availabe on the Pythran
+reporsitory.~\footnote{\emph{cf}.\
+\texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
+branch.}
 

### 2013-02-02 doc/papers/europar2013/paper.tex
Few cosmetic modifications

@@
 threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order, i.e.\ it is possible
+described by a random access iterator with a total order, that is possible
 to increase the iterator by any integer and to decide which iterator is the
@@
 activated, and can be used for portability with system where OpenMP is not
-available. Python does not have a preprocessor, but it is possible to catch the
+available. As Python does not have a preprocessor, it is possible to catch the
 import exception if the \texttt{omp} module is not found.
@@
 \texttt{threadprivate} directives were held by global variables not supported in
-Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
+Pythran~ yet; and the C++ code generated by Pythran does not preserve the perfect
 loop nesting required by the \texttt{collapse} clause.
@@
 
-All the source codes used for these benchmarks are availabe on the Pythran
-reporsitory.~\footnote{\emph{cf}.\
-\texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
-branch.}
+Time mesure can be reproduce as procedure and test codes are available on 
+developer repository~\footnote{\emph{cf}.\
+\texttt{https://github.com/pbrunet/pythran} on \texttt{europar2013} branch}
 

### 2013-02-03 doc/papers/europar2013/paper.tex
Few cosmetic modifications
Apply Adrien Merlini's review.

--- 2013-02-02 doc/papers/europar2013/paper.tex
@@
 This hybrid approach, based on a mix of interpreted and native code, is getting
-wide spread in the Python landscape. Section~\ref{sec:python-parallel} studies
+widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
 the existing approaches and shows a critical lack of parallelism support.
@@
 relevant for HPC, because of a specificity of CPython, the \emph{Global
-Interpreter Lock}~\cite{gil2012}~\footnote{Other Python interpreters, such as
-IronPython or Jython, do not have a GIL.} This lock ensures only one thread is
-active at a time in the interpreter. While it make it possible to have
+Interpreter Lock}~\cite{gil2012}.~\footnote{Other Python interpreters, such as
+IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
+is active at a time in the interpreter. While it makes it possible to have
 cooperative thread, say for a GUI, it does not take advantage of multiple cores.
 However, there are two notable exceptions: the GIL is released on I/O, and the
-GIl does not prevent the use of threads inside native modules, where the user
+GIL does not prevent the use of threads inside native modules, where the user
 has full control.
@@
 computation-intensive one where parallelism has been expressed in some ways.
-Thus, developers not familiar with lower level language or not eager to invest
+Thus, developers not familiar with lower level languages or not eager to invest
 the additional development time can still benefit from a fair performance boost.

### 2013-02-03 doc/papers/europar2013/paper.tex
A few reformatting later.

@@
 
-    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython Impelmentation
-    of a Parallel Reduction.}]
+    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
+    of a parallel reduction.}]
 import math
@@
 \begin{figure}[ht]
-    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython Impelmentation
-    of a Parallel Reduction.}]
+    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython implementation
+    of a parallel reduction.}]
 #pythran export sum_sqrt()

### 2013-02-06 doc/papers/europar2013/paper.tex
Major Paper update.
Take into account A. Cohen's review.

@@
 \usetikzlibrary{arrows}
+\usetikzlibrary{shapes}
 
-\lstset{language=python}
+\lstset{language=python,morekeywords={cimport, cdef, with}}
 
@@
 
-    The Python language offers an interesting alternative as a scientific
+    The Python language offers an interesting solution as a scientific
     computing language thanks to the SciPy package. However, the lack of
@@
     that it is possible to turn a module written in a large Python subset
-    enhanced with these directive into a native parallel module that runs as
-    fast as its naive C equivalent. The approach is validated through the
-    Pythran compiler.
+    enhanced with these directives into a native parallel module that runs as
+    fast as its naive C equivalent, while keeping the input code totally
+    compatible with the python interpreter. The approach is validated through
+    the Pythran compiler.
 
@@
 The prohibitive performance overhead of the language implied by its interpreted
-and highly dynamic nature should prevent its usage as a language for high
+and highly dynamic nature does not prevent its usage as a language for high
 performance computing. SciPy overcomes this issue through the usage of low-level
@@
 available to both native modules and the Python interpreter without conversion
-costs.
+cost.
 
@@
 widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
-the existing approaches and shows a critical lack of parallelism support.
+the existing approaches and shows a critical lack of parallelism support, and
+emphasises on the need for a backward-compatible approach.
 Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives
@@
 Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
-scientific applications and compared to both interpreted and native code in
-Section~\ref{sec:validation}.
+scientific kernels and compared to Cython in Section~\ref{sec:validation}.
 
@@
 In the many cores area, turning to parallelism to balance the performance
-limitations of scripting languages, as described in~\cite{choy05} in the context
-of the MATLAB language, or in~\cite{mals07} in the context of the R language,
-seems legitimate. In the context of Python, most approaches have
-focused on fork-based parallelism.
+limitations of scripting languages, as described 
+% in~\cite{choy05} in the context of the MATLAB language, or
+in~\cite{mals07} in the context of the R language, seems legitimate. In the
+context of Python, most approaches have focused on fork-based parallelism.
 
@@
 
+To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
+estimate the value of $\pi$. It's execution time using a multi-threaded
+implementation on 4 cores is $\times 1.46$ the execution time of the sequential
+version. GIL contention actually makes the code run slower.
+
 As a consequence, Python developers need to write multi-threaded native
-modules in order to fully benefit from multiple cores. This lead to a kind of
+modules in order to fully benefit from multiple cores. This leads to a kind of
 computations referred as hybrid computations.
@@
 
-In the context of interpreted language, an hybrid computation is defined
-in~\cite{dongara2007} as a computation where part of the code is interpreted,
-and part of the code is executed natively.
+In the context of interpreted language, \cite{dongara2007} defines a computation
+as \emph{hybrid} when part of the code is interpreted, and part of it is
+executed natively.
 
@@
 API~\cite{pythoncapi}. Tools have been developed to relieve the user from
-writing the encapsulating code, notably Swig~\cite{swig2003} that relies on an
+writing the encapsulating code, notably SWIG~\cite{swig2003} that relies on an
 enhanced interface specification, or
@@
 limited set of parallelism constructs, among which the possibility to define
-parallel loops, but no task parallelism. Plw~\cite{dongara2007} proposes a
+parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
 similar approach that mixes Python with C, using raw strings to hold the C code
-and the parallel directives, that are limited to parallel for loops.  The
-numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
-uses additional type information to generate sequential LLVM bytecode.  PyCUDA
-and PyOpenCL~\cite{klockner2012} also mixes python with kernels embeded as raw
-strings to target accelerators.
+and the parallel directives, that are limited to parallel for loops.
+%The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
+%compiler uses additional type information to generate sequential LLVM bytecode.
+PyCUDA and PyOpenCL~\cite{klockner2012} also mixes python with kernels embedded
+as raw strings to target accelerators.
 
@@
 less trivial way, the original code. They require to invest in a new language,
-and the long-term preservation of this investment is not ensured.
+and the long-term preservation of this investment is not ensured. Moreover,
+there is no falling back if the code where to be deployed in an environment
+where the parallelizing tool/module is not available. For instance, once a code
+has been ported to PyCUDA and transformed into CUDA code embedded in Python
+strings, there is no way back and the developer has to maintain two versions of
+the algorithm.
 
@@
 
-Copperhead~\cite{shedskin2006} is a functional, data parallel language embedded in
-Python. It uses n-uplets, Numpy arrays and lists as its core data structure and
+Copperhead~\cite{copperhead2011} is a functional, data parallel language embedded in
+Python. It uses n-uplets, numpy arrays and lists as its core data structure and
 prohibits usage of many control-flow operators such as loops, enforcing the use
@@
 \texttt{http://thrust.github.com/}} library. Python decorators are used to
-identify hot-spots that are dynamically compiled to native code.
+identify hot-spots that are JIT-compiled to native code.
 
@@
 
+To be completely backward compatible, it must be possible to run the input code
+in an environment that is not aware of the existence of the parallelizing
+solution. This principle is not respected by parallel libraries, but code
+annotations satisfy it. This is where  OpenMP~\cite{openmp3.1} parallel
+annotations shines: the original code remains compatible with a compiler that is
+not aware of OpenMP.
+
 This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
-with a Python subset called Pythran to bring fine-grain parallelism to Python
+with a Python subset called Pythran to bring fine-grain parallelism to Python.
 while being backward compatible with both the Python language, and the
-sequential algorithm. It means that:
+sequential algorithm.It means that:
+%
 \begin{enumerate}
@@
 incremental parallelization of the original source code while keeping the
-original code structure.
-
-The languages targeted by OpenMP are statically compiled. This section studies
-the semantic adaptation required to use the same API for a scripting language
-such as Python.
+original code structure. The languages targeted by OpenMP are statically
+compiled. This section studies the semantic adaptation required to use the same
+API for a scripting language such as Python.
 
@@
 
-While Pythran does have a decorator mechanism~\footnote{\see PEP 319,
-\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation fo
+While Pythran does have a decorator mechanism~\footnote{\see PEP 318,
+\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
 Python decorators.}, it only applies to functions, methods and classes and does
@@
 PLW~\cite{dongara2007} suggests to use string instructions to hold such
-decorations. As Python does not have anonymous block~\footnote{\see PEP340,
-    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
-anonymous block support in Python.}, one has to create a dummy \texttt{if 1:}
-instruction to apply an annotation to a whole block. Alternatively, the syntax
-\texttt{if 'my annotation':} is also supported.
-Listing~\ref{lst:instruction-annotation} shows how to annotate a parallel loop
-with reduction this way.
+decorations. As Python does not have anonymous block,
+%~\footnote{\see PEP340, \url{http://www.python.org/dev/peps/pep-0340/} for a
+%discussion concerning anonymous block support in Python.}
+one has to create a dummy \texttt{if 1:} instruction to apply an annotation to a
+whole block. Alternatively, the syntax \texttt{if 'my annotation':} is also
+supported.
 
-\begin{figure}
-    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
-    loop with reduction annotated with OpenMP directive.}]
-'omp parallel for reduction(+:r)'
-for auto in my_container:
-  r += x
-  \end{lstlisting}
-\end{figure}
+%Listing~\ref{lst:instruction-annotation} shows how to annotate a
+%parallel loop with reduction this way.
+%
+%\begin{figure}
+%    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
+%    loop with reduction annotated with OpenMP directive.}]
+%'omp parallel for reduction(+:r)'
+%for auto in my_container: r += x
+%  \end{lstlisting}
+%\end{figure}
 
@@
 local to a block. Using the \texttt{default(none)} clause is possible to ensure
-that no variable gets forgotten when building such lists, as a compile-time
-error is raised if a variable is referred in a block but not in the associated
-parallel region.
+no variable gets forgotten when building such lists.
 
@@
 
-Python also generalizes the concept of iterators through generators, which also
+Python also generalizes the concept of iterators through generators, which 
 behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
@@
 apply a particular expression on each value of the iterator. It is the case of
-the \texttt{enumerate} builtin that yields each element of the enumerated
-iterator associated with its index. Generator expressions such as \texttt{(x*x
-for x in l)} have the same behavior. These generators are random iterators only
-if their input iterator is a random iterator itself.
+the \texttt{enumerate} builtin, that yields each element of the enumerated
+iterator associated with its index, and of generator expressions such as
+\texttt{(x*x for x in l)}. These generators are random iterators only if their
+input iterator is a random iterator itself.
 
@@
 all the unpacked variables are considered as iterators, especially with respect
-to default privatization rules. In the given example, \texttt{i} and \texttt{v}
+to default privatization rules: in the given example, \texttt{i} and \texttt{v}
 are private, and the parallel iteration is valid because the input of
@@
     \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
-    loop with tuple unpacking.}]
+    loop in Pythran with tuple unpacking.}]
 'omp parallel for'
-for i, v in enumerate([2, 3, 5, 7, 11]):
-  print i, ':', v
+for i, v in enumerate([2, 3, 5, 7, 11]): print i, ':', v
   \end{lstlisting}
@@
 The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
-activated, and can be used for portability with system where OpenMP is not
-available. Python does not have a preprocessor, but it is possible to catch the
-import exception if the \texttt{omp} module is not found.
+activated, and can be used to detect when OpenMP is not available. Python does
+not have a preprocessor, but it is possible to catch the import exception if the
+\texttt{omp} module is not found.
 
@@
     OpenMP validation test case.}]
-def omp_parallel_for_if():
+def omp_parallel_for_if(loop_count):
   import omp
-  using = num_threads = 0
-  sum = sum2 = 0
-  LOOPCOUNT = 1000
-  'omp parallel for private(i) if(using == 1)'
-  for i in range(LOOPCOUNT + 1):
+  using = sum = sum2 = 0
+  'omp parallel for if(using == 1)'
+  for i in range(loop_count + 1):
     num_threads = omp.get_num_threads()
     sum += i
-  known_sum = (LOOPCOUNT * (LOOPCOUNT + 1)) / 2
+  known_sum = (loop_count * (loop_count + 1)) / 2
   return known_sum == sum and num_threads == 1
@@
 annotated with OpenMP and exported function type annotation, into C++ code. Its
-processing is summarized in Figure~\ref{fig:pythran-compiler}. The purpose of
-this section is not to dig into the internal of the compiler, but rather to
-focus on the impact of the parallelism layer, especially on the runtime library,
-\texttt{pythonic}.
+processing is summarized in Figure~\ref{fig:pythran-compiler} and relies on type
+inference to remove most dynamic behavior, combined with a high-level runtime
+library, \texttt{pythonic}, to allow a one-to-one mapping between Python and C++
+constructs. The purpose of this section is not to dig into the internal of the
+compiler, but rather to focus on the impact of the parallelism layer, especially
+on the runtime library.
 
@@
 \begin{tikzpicture}[
-file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center},
-    tool/.style={draw=black!50,fill=black!10,circle, align=center}]
+file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
+node distance=0.6cm},
+    tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
+distance=0.6cm}]
 \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
@@
 
-A critical point when designing the pythonic runtime was memory management, the
-very same aspect that lead to the use of a GIL in CPython. Pythran handles the
-problem by refusing recursive types, which makes it possible to solely rely on
-reference counting for memory management. It can be implemented through the
-thread-safe shared pointer mechanism provided in C++11 standard library.
+A critical point of the design of the \texttt{pythonic} runtime is memory
+management, the very same aspect that lead to the use of a GIL in CPython.
+Memory management is implemented in Shedskin through the general-purpose Boehm
+garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
+object inside parallel regions, thus making memory management explicit for these
+parts.
+
+Pythran handles the problem by refusing recursive types, which makes it possible
+to solely rely on reference counting for memory management. It can be
+implemented through the thread-safe shared pointer mechanism provided in C++11
+standard library.
 
 Using shared references simplifies memory management, but the counterpart is an
-extra atomic operation for each copy. It limits parallelism, so reducing the
-number of copies remains an important goal. The move semantic introduced in
+extra atomic operation for each copy. As it limits parallelism, reducing the
+number of copies becomes an important goal. The move semantic introduced in
 C++11 avoids a few copies when working on temporary objects, but argument
-passing may be a trouble. To avoid a reference increment, one can use pass
-parameters by reference. Using an interprocedural memory effect analysis, Pythran
-determines for each argument whether it must be passed using reference or const
-reference to prevent this extra cost.
+passing still imply copies. To avoid a reference increment, one can pass
+parameters by reference. Using an out-of-scope interprocedural memory effect
+analysis, Pythran determines for each argument whether it must be passed using
+reference or const reference to prevent this extra cost.
 
@@
 context-insensitive part ---the clauses---, and attach them to the proper
-instruction in the AST. It then ensures a bijective translation between Python
-instruction and generated C++ instruction so that the OpenMP directive is
-regenerated on the proper instruction. The same approach is used at the
-expression level, to be compatible with the \texttt{if} clause.
+instruction in the AST.
+
+The Pythran compiler ensures a bijective translation between Python instruction
+and generated C++ instruction so that the OpenMP directive is regenerated on the
+proper instruction. The same approach is used at the expression level, to be
+compatible with the \texttt{if} clause.
+
+At the AST level, it means that the Python AST is first reduced to a Python AST
+where all nodes not available in C++ are transformed into a compatible
+representation. For instance, list comprehension expressions are turned into
+function calls or tuple unpacking is turned in multiple assignments. During this
+transformation process, an expression is always transformed into a single expression
+and a statement is always transformed into a single statement. Then this
+AST is converted into a C++ AST meant to be pretty-printed.
 
@@
 To validate the approach proposed in this paper, we first performed a simple
-experiment. Starting from the C code of a small geomatics application, we
-successively parallelized it with OpenMP, turned it into Python with a Pythran
-compatible kernel and parallelized the Python version using the approach
+experiment. Starting from the C code of a small geomatics application, Hyantes,
+we successively parallelized it with OpenMP, turned it into Python with a
+Pythran compatible kernel and parallelized the Python version using the approach
 described in this paper. We also measurer the SLOC for the two versions.
 Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
-laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It
-non only emphasises the use of a high level language prototyping, but also shows
+laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It non
+only emphasises the use of a high level language prototyping, but also shows
 that it is possible to turn this prototype into reasonably efficient code
@@
 \begin{table}
+
+    \caption{Comparison of several versions of the Hyantes prototype.}
+    \label{tbl:hyantes}
 
@@
 
-    \caption{Comparison of several versions of the Hyantes prototype.}
-    \label{tbl:hyantes}
+\end{table}
 
-\end{table}
+The Hyantes programs scales relatively well. Figure~\ref{fig:hyantes-speedup} shows
+the relative speedup of the application when increasing the number of OpenMP
+threads.
+
+\begin{figure}[ht]
+    \caption{Relative speedups of Pythran+OpenMP version of the Hyantes prototype.}
+    \label{fig:hyantes-speedup}
+    \centering
+    \includegraphics[width=\textwidth]{hyantes}
+\end{figure}
 
@@
 not the case in Cython, where:
-
+%
 \begin{itemize}
     \item Only loops can be made parallel, through the use of a dedicated
-        intrinsic \texttt{prange} as loop iterator. To use more advanced
-        parallelism construction, one must rely on the possibility to write raw C
-        code embedded in cython.~\todo{SG->PB: tu m'explique ce truc de C dans
-        cython ?}
+        intrinsic \texttt{prange} as loop iterator.
 
-    \item Cython automatically infers reductions and variable privacy, giving
-        less control to the end-user.
+    \item Reductions and variable privacy are inferred, but it chokes on
+        reduction on private variables.
 
@@
 
-    \item The code included in a parallel region interacts with the Python code
-        in a restricted manner. For instance, it is impossible to use a function
-        imported from a Python module, while it is still possible to use native
-        C functions.
+    \item It is impossible to use a function imported from a Python module in a
+        parallel region, while it is still possible to use native C functions.
+
 \end{itemize}
 
-Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrates the
-difference between the two approaches. It shows that while being more integrated
-in Python, the Cython approach is more verbose.
+Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrate the
+difference between the two and illustrates the intrusive behavior of Cython.
 
-\begin{figure}[ht]
+\begin{figure}
 
@@
     of a parallel reduction.}]
-import math
-cimport cython
+from libc.math cimport sqrt
 from cython.parallel import parallel, prange
-from libc.math cimport sqrt
-def sum_sqrt():
-    cdef float r = 0
+def sum_sqrt(double r):
     cdef int i
     with nogil, parallel():
-        for i in prange(10000000):
-            r += sqrt(i)
+        for i in prange(10000000): r += sqrt(i)
     return r
     \end{lstlisting}
-\end{figure}
 %
-\begin{figure}[ht]
     \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython implementation
     of a parallel reduction.}]
-#pythran export sum_sqrt()
+#pythran export sum_sqrt(float)
 import math
-def sum_sqrt():
-    r = 0
+def sum_sqrt(r):
     "omp parallel for private(i) reduction(+:r)"
-    for i in xrange(10000000):
-        r += math.sqrt(i)
+    for i in xrange(10000000): r += math.sqrt(i)
     return r
@@
 
-The performance of the two approaches is illustrated in
-Figure~\ref{fig:cython-pythran}.
+The performance of the two approaches is shown in
+Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
+mathematical, image-processing or linear algebra kernels. All these kernels have
+been written in Cython and Python ---compatible with Pythran--- and annotated
+through the mechanism of each language to exhibit parallelism, then compiled
+into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag. Their
+execution time when called from the Python interpreter is measured through the
+\texttt{timeit} module on a laptop with 4 Intel i7 cores at 2.10 GHz with
+hyperthreading deactivated. All results are normalized against Cython sequential
+execution time. They show that while being of a higher level than Cython,
+Pythran achieves comparable results.
+
+The source codes used for these benchmarks are available on the Pythran
+repository.~\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
+branch.}
 
@@
 
-The benchmarked codes are typical mathematical, image-processing or linear
-algebra kernels. All these kernels have been written in Cython and Python
----compatible with Pythran--- and annotated through the mechanism of each
-language to exhibit parallelism, then compiled into native code using GCC 4.1
-and the \texttt{-O2 -fopenmp} flag. Their execution time when called from the
-Python interpreter is measured through the \texttt{timeit} module. All results
-are normalized to Cython sequential execution time.
-
-This figure shows our inability to select proper benchmarking code, and to run
-them on a real multiprocessor machine, but this is gonna change!~\todo{SG->PB:
-c'est pour toi !}
-
-All the source codes used for these benchmarks are availabe on the Pythran
-reporsitory.~\footnote{\emph{cf}.\
-\texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
-branch.}
-
 %=========================================================
@@
 This paper studies the relevancy of adding OpenMP annotations to Python code. It
-shows that providing minor semantic adaptations, these annotations are relevant
-and makes it possible for Python code to benefit from multi-core performance
-boost without worrying about the Global Interpreter Lock.
+shows that providing minor semantic adaptations, these annotations make it
+possible for Python code to benefit from multi-core performance boost while
+retaining backward compatibility and without worrying about the Global
+Interpreter Lock.
 
 To achieve this goal, a translator from a subset of the Python language to C++
-has been designed. This translator can turn regular Python modules
-annotated with OpenMP directives and a few type annotations into native parallel
-module. Both the input module and the generated one remain compatible with the
-standard interpreter.
+has been designed. This translator turns regular Python modules annotated with
+OpenMP directives and a few type annotations into native parallel module. The
+input module remains compatible with the standard interpreter.
 
@@
 to generate native module with an hybrid Python-C syntax that also provide means
-to exhibits fine grained parallelism. Modules generated by Pythran outperform
-modules generated by Cython for various kernels and both benefit from the
-parallel boost.
+to exhibit fine grained parallelism. It shows that retaining Python
+compatibility does not prevent the achievement of comparable performances.
 
-Future works include the extension of the approach to OpenMP 4 and the
+Future work includes the extension of the approach to OpenMP 4 and the
 \texttt{target} clause that should make it possible to target MIC from Python.
@@
 ANR through the CARP project. The authors would like to thanks Adrien
-\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime.
-Fabien \textsc{Dagnat} and Mehdi \textsc{Amini} also contributed to this paper
-through their many valuable advices.
+\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime,
+Albert \textsc{Cohen}, B{\'e}atrice \textsc{Creusillet}, Fabien
+\textsc{Dagnat} and Mehdi \textsc{Amini} for their valuable advices.
 

### 2013-02-06 doc/papers/europar2013/bench/cython/fbcorr_cython.pyx

@@ -10,18 +10,16 @@ def fbcorr(np.ndarray[DTYPE_t, ndim=4, negative_indices=False] imgs,np.ndarray[D
     cdef int n_imgs = imgs.shape[0], n_rows = imgs.shape[1], n_cols = imgs.shape[2], n_channels = imgs.shape[3]
     cdef int n_filters = filters.shape[0], height = filters.shape[1], width = filters.shape[2], n_ch2 = filters.shape[3]
     cdef int ii, rr, cc, hh, ww, jj, ff
-    cdef double impval, filterval
+    cdef double imgval, filterval, tmp
     cdef np.ndarray[DTYPE_t, ndim=4, negative_indices=False] output = np.zeros((n_imgs, n_filters, n_rows - height + 1, n_cols - width + 1))
-    "omp parallel for private(ii,rr,cc,hh,ww,jj,ff,imgval,filterval)"
-    for ii in xrange(n_imgs):
-        for rr in xrange(n_rows - height + 1):
-            for cc in xrange(n_cols - width + 1):
-                for ff in xrange(n_filters):
-                    tmp = 0
-                    for hh in xrange(height):
-                        for ww in xrange(width):
-                            for jj in xrange(n_channels):
-                                imgval = imgs[ii,rr + hh,cc + ww,jj]
-                                filterval = filters[ff,hh,ww,jj]
-                                tmp += imgval * filterval
-                    output[ii,ff,rr,cc] = tmp
+    with nogil:
+        for ii in xrange(n_imgs):
+            for rr in xrange(n_rows - height + 1):
+                for cc in xrange(n_cols - width + 1):
+                    for ff in xrange(n_filters):
+                        for hh in xrange(height):
+                            for ww in xrange(width):
+                                for jj in xrange(n_channels):
+                                    imgval = imgs[ii,rr + hh,cc + ww,jj]
+                                    filterval = filters[ff,hh,ww,jj]
+                                    output[ii,ff,rr,cc] = output[ii,ff,rr,cc] + imgval * filterval

### 2013-02-06 doc/papers/europar2013/bench/cython/hyantes.pyx

@@ -12,12 +12,14 @@ ctypedef np.double_t DTYPE_t
 def run(double xmin, double ymin, double xmax, double ymax, double step, double range_, int range_x, int range_y, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] t):
     cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] pt = np.zeros((range_x, range_y))
     cdef int i,j,k
-    cdef double tmp
+    cdef double tmp, xi, yj
     "omp parallel for private(j,k,tmp)"
     for i in xrange(range_x):
         for j in xrange(range_y):
+            xi = xmin+step*i
+            yj = ymin+step*j
             for k in xrange(t.shape[0]):
-                tmp = 6368.* math.acos( math.cos(xmin+step*i)*math.cos( t[k,0] ) * math.cos((ymin+step*j)-t[k,1])+  math.sin(xmin+step*i)*math.sin(t[k,0]))
+                tmp = 6368.* math.acos( math.cos(xi)*math.cos( t[k,0] ) * math.cos(yj-t[k,1])+  math.sin(xi)*math.sin(t[k,0]))
                 if tmp < range_:
                     pt[i,j]+=t[k,2] / (1+tmp)
     return pt

### 2013-02-06 doc/papers/europar2013/bench/cython/hyantes_para.pyx

@@ -14,13 +14,15 @@ ctypedef np.double_t DTYPE_t
 def run(double xmin, double ymin, double xmax, double ymax, double step, double range_, int range_x, int range_y, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] t):
     cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] pt = np.zeros((range_x, range_y))
     cdef int i,j,k
-    cdef double tmp
+    cdef double tmp, xi, yj
     "omp parallel for private(j,k,tmp)"
     with nogil, parallel():
         for i in prange(range_x):
             for j in xrange(range_y):
+                xi = xmin+step*i
+                yj = ymin+step*j
                 for k in xrange(t.shape[0]):
-                    tmp = 6368.* math.acos( math.cos(xmin+step*i)*math.cos( t[k,0] ) * math.cos((ymin+step*j)-t[k,1])+  math.sin(xmin+step*i)*math.sin(t[k,0]))
+                    tmp = 6368.* math.acos( math.cos(xi)*math.cos( t[k,0] ) * math.cos(yj-t[k,1])+  math.sin(xi)*math.sin(t[k,0]))
                     if tmp < range_:
                         pt[i,j]+=t[k,2] / (1+tmp)
     return pt

### 2013-02-06 doc/papers/europar2013/bench/pythran/hyantes.py

@@ -3,11 +3,12 @@
 import math, numpy
 def run(xmin, ymin, xmax, ymax, step, range_, range_x, range_y, t):
     pt = numpy.zeros((range_x, range_y))
-    "omp parallel for private(j,k,tmp)"
+    "omp parallel for private(i,j,k,tmp)"
     for i in xrange(range_x):
         for j in xrange(range_y):
+            xi, yj = xmin+step*i, ymin+step*j
             for k in xrange(t.shape[0]):
-                tmp = 6368.* math.acos( math.cos(xmin+step*i)*math.cos( t[k,0] ) * math.cos((ymin+step*j)-t[k,1])+  math.sin(xmin+step*i)*math.sin(t[k,0]))
+                tmp = 6368.* math.acos( math.cos(xi)*math.cos( t[k,0] ) * math.cos((yj)-t[k,1])+  math.sin(xi)*math.sin(t[k,0]))
                 if tmp < range_:
                     pt[i,j]+=t[k,2] / (1+tmp)
     return pt

### 2013-02-06 doc/papers/europar2013/bench/pythran/setup.py

@@ -56,5 +56,6 @@ setup(
             Extension("pairwise_omp", ["pairwise.py"], extra_compile_args=['-fopenmp']),
             Extension("hyantes", ["hyantes.py"]),
             Extension("hyantes_omp", ["hyantes.py"], extra_compile_args=['-fopenmp']),
+            Extension("pi_buffon_omp", ["pi_buffon.py"], extra_compile_args=['-fopenmp']),
             ]
      )

### 2013-02-06 doc/papers/europar2013/bench/run_all.py

@@ -14,7 +14,7 @@ import timeit
 candidates = glob.glob("pythran/*.py")
 sys.path.append("pythran")
 sys.path.append("cython")
-nb_iter = 1
+nb_iter = 21
 median = lambda x: sorted(x)[len(x) / 2]
 for candidate in candidates:
     with file(candidate) as content:

### 2013-02-06 doc/papers/europar2013/cython.py

@@ -3,10 +3,8 @@ import numpy as np
 import matplotlib.mlab as mlab
 import matplotlib.pyplot as plt
 import math
-from matplotlib.backends.backend_pdf import PdfPages
 
 with file("cython.dat") as fd:
-
     r = list()
     for line in fd:
         r.append(filter(None, line.split(' ')))
@@ -31,20 +29,18 @@ with file("cython.dat") as fd:
             if math.isnan(y[i][j]):
                 y[i][j] = 0
         
+    fig = plt.figure(1, figsize=(8,4))
 
-    hatches = [ '/' , 'o', '\\' ,  'O' , '+' , 'x' , 'o' , 'O' , '.' , '*' ]
+    hatches = [ ' ' , 'x', ' ' , 'x' , '+' , 'x' , 'o' , 'O' , '.' , '*' ]
+    colors  = [ 'grey' , 'grey', 'w', 'w' ]
     p = [0]*len(x)
     for i,j in enumerate(y):
-        p[i] = plt.bar(range(i,y.size+len(y)+len(y)-1, len(y) +1), j, hatch=hatches[i], color='white')
+        p[i] = plt.bar(range(i,y.size+len(y)+len(y)-1, len(y) +1), j, hatch=hatches[i], color=colors[i])
 
     #plt.xlabel('Comparison between cython and pythran')
     plt.xticks(range(2,y.size+len(y)+1, len(y) +1), [_[0] for _ in r ] )
-    plt.ylabel('execution time')
+    plt.ylabel('normalized execution time')
     #plt.title(r'$comparaison between cython and pythran with/without parallelism$')
-    plt.grid(True)
+    plt.grid(True, axis='y')
     plt.legend( (p[0][0], p[1][0], p[2][0], p[3][0]), ('Pythran', 'Pythran + OMP', 'Cython', 'Cython + OMP') , bbox_to_anchor=(0., 1.02, 1., .102), loc=2, ncol=2, mode="expand", borderaxespad=0.)
-    pp = PdfPages('cython.pdf')
-    plt.savefig(pp,format='pdf')
-    pp.close()
-
-
+    plt.savefig('cython.pdf',format='pdf')

### 2013-02-07 doc/papers/europar2013/paper.tex
Take into account F. Dagnat's advices.

@@
 \usepackage{todo}
+\usepackage{xcolor}
 \newcommand\see{\emph{cf.\ }}
@@
 
-\lstset{language=python,morekeywords={cimport, cdef, with}}
+\lstset{language=python,morekeywords={cimport, cdef, with},
+basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray}}
 
 \begin{document}
+% update listing numbering
+\renewcommand{\thelstlisting}{\arabic{lstlisting}}
 %
@@
 
-In the context of interpreted language, \cite{dongara2007} defines a computation
+In the context of interpreted languages, \cite{dongara2007} defines a computation
 as \emph{hybrid} when part of the code is interpreted, and part of it is
@@
 the additional development time can still benefit from a fair performance boost.
-This approach has been subject to many studies that can be classified based on
-their backward-compatibity with the host language.
+This approach is the subject of many studies that can be classified based on
+their compatibility with the host language.
 
@@
 The core issue of these approaches is that they imply to modify, in a more or
-less trivial way, the original code. They require to invest in a new language,
+less complex way, the original code. They require to invest in a new language,
 and the long-term preservation of this investment is not ensured. Moreover,
@@
 consequence, they remain compatible with the original language and do not suffer
-from the drawbacks of the previous approach.
+from the drawbacks of the previous approach. They also keep the benefit of
+existing tools associated to the language.
 
@@
 The core directive to handle data parallelism is the \texttt{for} directive that
-distribute the iteration space of the associated loop among the existing
+distributes the iteration space of the associated loop among the existing
 threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order, i.e.\ it is possible
-to increase the iterator by any integer and to decide which iterator is the
-greatest among two. The integer used as loop indices in Fortran and C satisfy
-this conditions, so do C++ iterators with the \texttt{random\_access\_tag}
-trait. But a Python iterator only advance by a step of one until it is
-exhausted, in which case it raises an exception. They only have forward iterator
-behavior.
+described by a random access iterator with a total order. Integers used as
+loop indices in Fortran and C satisfy this conditions, as well as C++ iterators with
+the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
+step of one until it is exhausted, in which case it raises an exception: it
+behaves as a forward iterator. To be compatible with OpenMP, these iterators
+must be turned into random access iterators.
 
@@
 standard containers: \texttt{list}~\footnote{Python \texttt{list}s behave as C++
-\texttt{vector}s.}, \texttt{set} or \texttt{dict}.
+\texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
+care.
 
-Python also generalizes the concept of iterators through generators, which 
-behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
-\emph{stop}, \emph{step})}, successively yields value starting from
-\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}. It is
-easily extended to support random access, but it generally does not make
-sense to use a generator as a loop iterator, as the relation between two
-random states of the iterator may be of an arbitrary complexity.
+Generators, Python objects that behave like iterators, are commonly used in
+Python. The simplest one, \texttt{xrange(\emph{start}, \emph{stop},
+\emph{step})}, successively yields value starting from \texttt\emph{start} to
+\texttt\emph{stop} by a step of \texttt\emph{step}. It is easily extended to
+support random access, but it generally does not make sense to use a generator
+as a loop iterator, as the relation between two random states of the iterator
+may be of an arbitrary complexity.
 
-A particular class of generator behaves like adaptor, in the sense that they
-apply a particular expression on each value of the iterator. It is the case of
-the \texttt{enumerate} builtin, that yields each element of the enumerated
-iterator associated with its index, and of generator expressions such as
-\texttt{(x*x for x in l)}. These generators are random iterators only if their
+Generator expressions are generators whose content is built from an other
+iterator. For instance \texttt{(x*x for x in l)} successively yields the square
+of each element in \texttt{l}. They behave like adaptors: they apply a
+particular expression on each value of the iterator. It is also the case of the
+\texttt{enumerate} builtin, that yields each element of the enumerated iterator
+associated with its index. These generators are random iterators only if their
 input iterator is a random iterator itself.
@@
 
+Pythran runtime is aware of these three kind of iterators: random access
+iterators for standard containers, adaptor iterators for generator expressions
+and forward iterators for generators.
+
 \subsection{Runtime Library}
 
-The OpenMP runtime library makes it possible to retrieve the active thread id,
-whether dynamic scheduling is enabled and so on. All the functions are declared
-in the \texttt{<omp.h>} header, and have a default behavior when OpenMP is not
+OpenMP provides a small runtime library that, for instance, makes it possible to
+retrieve the active thread id. All the functions are declared in the
+\texttt{<omp.h>} header, and have a default behavior when OpenMP is not
 activated.
@@
 
-A validation suite for OpenMP has been proposed in~\cite{wang2012} for C and
-Fortran. We have ported it to Python, and we also extended it to validate the
-corner cases specific to Python described in this section. An example of such
-test case is given in Listing~\ref{lst:openmp-validation}.
+A validation suite for OpenMP is proposed in~\cite{wang2012} for C and Fortran.
+We ported it to Python, and we also extended it to validate the corner cases
+specific to Python described in this section. A typical test case is given in
+Listing~\ref{lst:openmp-validation}.
 
@@
 
-Pythran is subset of Python focused on scientific computing. It is implicitly
+Pythran is a subset of Python focused on scientific computing. It is implicitly
 statically typed and support most Python constructs except those that involve
@@
 language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
-partially supported. The associated compiler turns Pythran code, eventually
-annotated with OpenMP and exported function type annotation, into C++ code. Its
-processing is summarized in Figure~\ref{fig:pythran-compiler} and relies on type
-inference to remove most dynamic behavior, combined with a high-level runtime
-library, \texttt{pythonic}, to allow a one-to-one mapping between Python and C++
-constructs. The purpose of this section is not to dig into the internal of the
-compiler, but rather to focus on the impact of the parallelism layer, especially
-on the runtime library.
+currently only partially supported. The associated compiler turns Pythran code,
+possibly annotated with OpenMP and a few function type annotations, into C++
+code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
+relies on type inference to remove most dynamic behavior, combined with a
+high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
+between Python and C++ constructs. The purpose of this section is not to dig
+into the internal of the compiler, but rather to focus on the impact of the
+parallelism layer, especially on the runtime library.
 
@@
 runtime is based on the STL and provides the same guaranty, so that the data
-dependences in the original code remains explicit.
+dependencies in the original code remains explicit.
 
-A critical point of the design of the \texttt{pythonic} runtime is memory
-management, the very same aspect that lead to the use of a GIL in CPython.
+A critical point of the design of the \texttt{pythonic++} runtime is memory
+management, the very same aspect that led to the use of a GIL in CPython.
 Memory management is implemented in Shedskin through the general-purpose Boehm
@@
 C++11 avoids a few copies when working on temporary objects, but argument
-passing still imply copies. To avoid a reference increment, one can pass
+passing still implies copies. To avoid a reference increment, one can pass
 parameters by reference. Using an out-of-scope interprocedural memory effect
@@
 Pythran compatible kernel and parallelized the Python version using the approach
-described in this paper. We also measurer the SLOC for the two versions.
+described in this paper. We also measure the SLOC of the two versions.
 Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
-laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It non
+laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It not
 only emphasises the use of a high level language prototyping, but also shows
@@
         Pythran & 30    & 38 &  22 \\
-        Pythran+OMP    & 30    & 11 & no OpenMP support\\
+        Pythran+OMP    & 31    & 11 & no OpenMP support\\
         \hline
@@
 We then compare Pythran with Cython. The approach share some similarities: they
-both generate code written in a rather low level language, with OpenMP
+both generate code written in a lower level language, with OpenMP
 directives. However, the inputs differ as Cython requires more typing
@@
 \begin{itemize}
-    \item Only loops can be made parallel, through the use of a dedicated
-        intrinsic \texttt{prange} as loop iterator.
+
+    \item Only loops can be made parallel, using a new \texttt{prange} intrinsic
+        as iterator.
 
@@
 
-    \item As Cython allows the mix of Python and Cython code, the user is
-        responsible from releasing the GIL.
+    \item The user is responsible from releasing the GIL inside parallel
+        regions.
 
     \item It is impossible to use a function imported from a Python module in a
-        parallel region, while it is still possible to use native C functions.
+        parallel region, but it is still possible to use native C functions.
 
 \end{itemize}
-
+%
 Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrate the
@@
 
-This paper studies the relevancy of adding OpenMP annotations to Python code. It
-shows that providing minor semantic adaptations, these annotations make it
-possible for Python code to benefit from multi-core performance boost while
-retaining backward compatibility and without worrying about the Global
-Interpreter Lock.
+This paper studies the addition  of OpenMP annotations to Python. It shows that
+providing minor semantic adaptations, these annotations make it possible for
+Python code to benefit from multi-cores while retaining backward compatibility
+and without worrying about the Global Interpreter Lock.
 

### 2013-02-07 doc/papers/europar2013/paper.tex
Update nebchmark result thanks to Big's machine.

@@
 the relative speedup of the application when increasing the number of OpenMP
-threads.
+threads using up to 8 AMD Opteron 6176 SE.
 
@@
 Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
-mathematical, image-processing or linear algebra kernels. All these kernels have
-been written in Cython and Python ---compatible with Pythran--- and annotated
-through the mechanism of each language to exhibit parallelism, then compiled
-into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag. Their
-execution time when called from the Python interpreter is measured through the
-\texttt{timeit} module on a laptop with 4 Intel i7 cores at 2.10 GHz with
-hyperthreading deactivated. All results are normalized against Cython sequential
-execution time. They show that while being of a higher level than Cython,
-Pythran achieves comparable results.
+mathematical, image-processing or linear algebra kernels. All these kernels
+have been written in Cython and Python ---compatible with Pythran--- and
+annotated through the mechanism of each language to exhibit parallelism, then
+compiled into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.
+Their execution time when called from the Python interpreter is measured
+through the \texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.
+All results are normalized against Cython sequential execution time. They show
+that while being of a higher level than Cython, Pythran achieves comparable
+results.
 

### 2013-02-07 doc/papers/europar2013/bench/pythran/parsetab.py

@@ -1,49 +1,0 @@
-
-# parsetab.py
-# This file is automatically generated. Do not edit.
-_tabversion = '3.2'
-
-_lr_method = 'LALR'
-
-_lr_signature = 'Li5-\x8b+j.\xdf\x84\x1c\x1e&_\x05c'
-    
-_lr_action_items = {'FLOAT':([9,10,11,25,27,],[20,20,20,20,20,]),'RPAREN':([9,10,12,13,14,15,16,17,18,19,20,21,22,23,24,26,31,33,34,35,],[-5,-5,-17,-19,-7,-15,-16,-6,-9,29,-20,-18,30,31,-11,-10,-14,-8,-12,-13,]),'STR':([9,10,11,25,27,],[15,15,15,15,15,]),'INT':([9,10,11,25,27,],[21,21,21,21,21,]),'PYTHRAN':([3,],[5,]),'LIST':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,26,-15,-16,-9,-20,-18,-11,-10,-14,26,-12,-13,]),'RARRAY':([28,],[34,]),'LONG':([9,10,11,25,27,],[13,13,13,13,13,]),'COLUMN':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,25,-15,-16,-9,-20,-18,-11,-10,-14,25,-12,-13,]),'LARRAY':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,28,-15,-16,-9,-20,-18,-11,-10,-14,28,-12,-13,]),'COMPLEX':([9,10,11,25,27,],[12,12,12,12,12,]),'EXPORT':([5,6,],[6,7,]),'BOOL':([9,10,11,25,27,],[16,16,16,16,16,]),'LPAREN':([7,8,9,10,11,25,27,],[9,10,11,11,11,11,11,]),'SET':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,24,-15,-16,-9,-20,-18,-11,-10,-14,24,-12,-13,]),'SHARP':([0,2,29,30,],[3,3,-4,-3,]),'IDENTIFIER':([6,],[8,]),'DICT':([12,13,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,-15,-16,-9,-20,-18,-11,-10,-14,35,-12,-13,]),'COMMA':([12,13,14,15,16,
[…]

### 2013-02-08 doc/papers/europar2013/paper.tex
Take into account BC's review.

@@
 and highly dynamic nature does not prevent its usage as a language for high
-performance computing. SciPy overcomes this issue through the usage of low-level
-routines written in C or Fortran and encapsulated in Python through
-\emph{native} modules. Moreover, its core data structure, the multi-dimensional
-array~\cite{numpyarray2011}, was designed so that the underlying data are
+performance computing. SciPy overcomes this issue through low-level routines
+written in C or Fortran and encapsulated in Python \emph{native} modules.
+Moreover, its core data structure, the multi-dimensional
+array~\cite{numpyarray2011}, has been designed so that the underlying data are
 available to both native modules and the Python interpreter without conversion
@@
 
-To make it easier for the user to write native functions, \emph{i.e.}\ without
-having to write the glue code that turns Python object into native structures
-back and forth, the SciPy package provides the \texttt{weave} module that makes
-it possible to bundle C code snippet into Python code and compile then load them
-at runtime.
+To enable the user to write native functions, without having to write the glue
+code that turns Python object into native structures back and forth, the SciPy
+package provides the \texttt{weave} module that makes it possible to bundle C
+code snippet into Python code and compile then load them at runtime.
 
@@
 emphasises on the need for a backward-compatible approach.
-Section~\ref{sec:python-openmp} studies the validity of adding OpenMP directives
-to the Python language in order to add fine-grain parallelism support to Python
-in the context of a Python subset called Pythran. Static compilation of this
-language into C++ and the underlying runtime is described in
+Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
+within the Python language in order to add fine-grain parallelism support to
+Python in the context of a Python subset called Pythran. Static compilation of
+this language into C++ and the underlying runtime is described in
 Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
@@
 relevant for HPC, because of a specificity of CPython, the \emph{Global
-Interpreter Lock}~\cite{gil2012}.~\footnote{Other Python interpreters, such as
+Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
 IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
 is active at a time in the interpreter. While it makes it possible to have
-cooperative thread, say for a GUI, it does not take advantage of multiple cores.
-However, there are two notable exceptions: the GIL is released on I/O, and the
-GIL does not prevent the use of threads inside native modules, where the user
-has full control.
+cooperative threads, say for a GUI, it does not take advantage of multiple
+cores.  However, there are two notable exceptions: the GIL is released on I/O,
+and the GIL does not prevent the use of threads inside native modules, where the
+user has full control.
 
@@
 operation! It comes at no surprise that many approaches extend the Python
-language to add a static typing overlay. Likewise, there is only two types of
+language to add a static typing overlay. Likewise, there are only two types of
 integers (64bits integers and multi-precision integers) and one type of floating
-point type (double-precision floats) in Python, while using a type with the
+point type (double-precision floats) in Python, so using a type with the
 appropriate size may lead to significant performance boost.
@@
 %compiler uses additional type information to generate sequential LLVM bytecode.
-PyCUDA and PyOpenCL~\cite{klockner2012} also mixes python with kernels embedded
-as raw strings to target accelerators.
+PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
+Python with kernels embedded as raw strings to target accelerators.
 
-The core issue of these approaches is that they imply to modify, in a more or
-less complex way, the original code. They require to invest in a new language,
+The main issue of these approaches is that they imply to modify, in a more or
+less complex way, the original code. They require to invest in a new dialect,
 and the long-term preservation of this investment is not ensured. Moreover,
@@
 consequence, they remain compatible with the original language and do not suffer
-from the drawbacks of the previous approach. They also keep the benefit of
+from the drawbacks of the previous approaches. They also keep the benefit of
 existing tools associated to the language.
@@
 parallelism. In returns, it can be efficiently compiled to either CUDA or C++
-with calls to the Thrust~\footnote{\emph{cf}.\
+with calls to the Thrust\footnote{\emph{cf}.\
 \texttt{http://thrust.github.com/}} library. Python decorators are used to
@@
 Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
-Shedskin~\cite{shedskin2006}, a Python to C++ compiler also are viable way to
+Shedskin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
 turn regular Python codes into optimized ones, but they do not offer support for
-fine-grained parallelism beyond what the standard library
-proposes.
-%~\footnote{Shedskin is compatible with several Python modules that
-%provide coarse-grained, fork-based parallelism.}
+fine-grained parallelism beyond what the standard library proposes.
+%~\footnote{Shedskin is compatible with several Python modules that provide
+%coarse-grained, fork-based parallelism.}
 
@@
 annotations satisfy it. This is where  OpenMP~\cite{openmp3.1} parallel
-annotations shines: the original code remains compatible with a compiler that is
+annotations shine: the original code remains compatible with a compiler that is
 not aware of OpenMP.
@@
 %=========================================================
-\section{OpenMP Semantic Adaptations}\label{sec:python-openmp}
+\section{OpenMP Semantical Adaptations}\label{sec:python-openmp}
 %=========================================================
@@
 original code structure. The languages targeted by OpenMP are statically
-compiled. This section studies the semantic adaptation required to use the same
-API for a scripting language such as Python.
+compiled. This section studies the semantical adaptation required to use the
+same API for a scripting language such as Python.
 
@@
 
-OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them apply
-on structured blocks, represented by an instruction in C/C++ and delimited by
-comment in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
+OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them
+apply to structured blocks, represented by an instruction in C/C++ and delimited
+by comments in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
 attached to a specific instruction.
 
-While Pythran does have a decorator mechanism~\footnote{\see PEP 318,
+While Pythran has a decorator mechanism\footnote{\see PEP 318,
 \url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
 Python decorators.}, it only applies to functions, methods and classes and does
-not make it possible to attach decorations to random instructions.
-PLW~\cite{dongara2007} suggests to use string instructions to hold such
-decorations. As Python does not have anonymous block,
+not make it possible to attach decorations to other statements.  Pythran uses
+the path taken by PLW~\cite{dongara2007}, that uses string instructions to hold
+such decorations. As Python does not have anonymous block,
 %~\footnote{\see PEP340, \url{http://www.python.org/dev/peps/pep-0340/} for a
@@
 declaration in Python, and all variables assigned in a function have the
-function scope. Consequently, all variables referred to in a function are to be
-considered when building such variable lists: there is no concept of variable
-local to a block. Using the \texttt{default(none)} clause is possible to ensure
-no variable gets forgotten when building such lists.
+function scope. Consequently, all variables that are referenced in a function
+are considered when building such variable lists: there is no concept of
+variable local to a block. Using the \texttt{default(none)} clause is possible
+to ensure no variable gets forgotten when building such lists.
 
-The \texttt{reduction(\emph{operator}: \emph{list})} is used to handle data
-dependencies when performing a parallel reduction. The supported operator
-depends on the input and backend language: Python has \texttt{min}/\texttt{max} operators but
-C/C++ have not. Python does not have \texttt{\&\&} or \texttt{||}
-operators~\footnote{Although similar, the \texttt{and} and \texttt{or} keywords
-have a special meaning in Python.} while C/C++ have.
+The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
+characterize some data dependencies when performing a parallel reduction. The
+list of supported operators depend on the input and backend languages: Python
+has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
+\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
+the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
 
@@
 threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order. Integers used as
-loop indices in Fortran and C satisfy this conditions, as well as C++ iterators with
+described by a random access iterator with a total order. Integers used as loop
+indices in Fortran and C satisfy this conditions, as well as C++ iterators with
 the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
@@
 The extension of Python iterator to random access iterators is direct for the
-standard containers: \texttt{list}~\footnote{Python \texttt{list}s behave as C++
+standard containers: \texttt{list}\footnote{Python \texttt{list}s behave as C++
 \texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
@@
 Python. The simplest one, \texttt{xrange(\emph{start}, \emph{stop},
-\emph{step})}, successively yields value starting from \texttt\emph{start} to
-\texttt\emph{stop} by a step of \texttt\emph{step}. It is easily extended to
+\emph{step})}, successively yields value starting from \texttt{\emph{start}} to
+\texttt{\emph{stop}} by a step of \texttt{\emph{step}}. It is easily extended to
 support random access, but it generally does not make sense to use a generator
@@
 
-Providing a biding to these library in the Python does not raise
-particular problems, as the signature of these functions only involves passing
-integers as parameters or return values, except for the mutex manipulation. In
-that case an opaque type is used to represent the native type.
+Providing a binding to these library in Python does not raise particular
+problems, as the signature of these functions only involves integers passed as
+parameters or return values, except for the mutex manipulation. In that case an
+opaque type is used to represent the native type.
 
@@
 A validation suite for OpenMP is proposed in~\cite{wang2012} for C and Fortran.
-We ported it to Python, and we also extended it to validate the corner cases
+We ported it to Python, and also extended it to validate the corner cases
 specific to Python described in this section. A typical test case is given in
@@
 Pythran is a subset of Python focused on scientific computing. It is implicitly
-statically typed and support most Python constructs except those that involve
+statically typed and supports most Python constructs except those that involve
 introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
-\texttt{eval}). A few standard module are supported in addition to the core
+\texttt{eval}). A few standard modules are supported in addition to the core
 language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
@@
 between Python and C++ constructs. The purpose of this section is not to dig
-into the internal of the compiler, but rather to focus on the impact of the
+into the internals of the compiler, but rather to focus on the impact of the
 parallelism layer, especially on the runtime library.
@@
 
-The standard library of C++11, unlike its C++03 version, is thread-safe. Pythran
-runtime is based on the STL and provides the same guaranty, so that the data
-dependencies in the original code remains explicit.
+Pythran runtime is based on the STL. The standard library of C++11, unlike its
+C++03 version, is thread-safe, so the data dependencies in the original code
+remain explicit.
 
@@
 garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
-object inside parallel regions, thus making memory management explicit for these
-parts.
+objects inside parallel regions, thus making memory management explicit for
+these parts.
 
@@
 During Python to C++ translation, Pythran adopts a blind strategy: it does not
-understand the semantic of the annotations. Instead, it just split each
+understand the semantics of the annotations. Instead, it just splits each
 annotation into a context-sensitive part ---the variable names--- and a
-context-insensitive part ---the clauses---, and attach them to the proper
+context-insensitive part ---the clauses---, and attaches them to the proper
 instruction in the AST.
@@
 
-At the AST level, it means that the Python AST is first reduced to a Python AST
-where all nodes not available in C++ are transformed into a compatible
-representation. For instance, list comprehension expressions are turned into
-function calls or tuple unpacking is turned in multiple assignments. During this
-transformation process, an expression is always transformed into a single expression
-and a statement is always transformed into a single statement. Then this
-AST is converted into a C++ AST meant to be pretty-printed.
+At the AST level, it means that the Python AST is first reduced to a tree where
+all nodes not available in C++ are transformed into a compatible representation.
+For instance, list comprehension expressions are turned into function calls or
+tuple unpacking is turned in multiple assignments. During this transformation
+process, an expression is always transformed into a single expression and a
+statement is always transformed into a single statement. Then this AST is
+converted into a C++ AST meant to be pretty-printed.
 
@@
 data is made to turn the type agnostic, non-contiguous Python data into dense
-typed ones. This extra copy implies an extra cost that is not negligible: the
-time taken to pass two lists of float from Python to C++ is only twice less than
-the time taken to compute the dot product of the same lists directly in Python.
-Following Amdahl's law, this transfer cost greatly hinders parallelization, and
-is a well known issue.
+typed ones. This extra copy implies an extra cost that is not negligible:
+Passing two lists of float from Python to C++ requires as much as half the time
+to compute the dot product of the same lists directly in Python.  Following
+Amdahl's law, this transfer cost greatly hinders parallelization, and is a well
+known issue.
 
@@
 
-We then compare Pythran with Cython. The approach share some similarities: they
-both generate code written in a lower level language, with OpenMP
+We then compare Pythran with Cython. The approaches share some similarities:
+they both generate code written in a lower level language, with OpenMP
 directives. However, the inputs differ as Cython requires more typing
-information that Pythran to generate efficient code, and Cython is not
-backward-compatible with Python. They both translate explicit fine-grained
+information than Pythran to generate efficient code, and Cython is not
+backward-compatible with Python. They translate explicit fine-grained
 parallelism through code annotations or specific python constructs,
@@
 Pythran exposes the full OpenMP interface to the user, thus enabling both data
-and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
-not the case in Cython, where:
+and task parallelism, as described in Section~\ref{sec:python-openmp}. It is
+not the case in Cython:
 %
@@
 %
-    \begin{lstlisting}[label={lst:pythran-sample}, caption={Cython implementation
+    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran implementation
     of a parallel reduction.}]
@@
 The source codes used for these benchmarks are available on the Pythran
-repository.~\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
+repository.\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
 branch.}
@@
 The approach is compared with Cython, an extension of the Python language used
-to generate native module with an hybrid Python-C syntax that also provide means
-to exhibit fine grained parallelism. It shows that retaining Python
+to generate native module with an hybrid Python-C syntax that also provides
+means to exhibit fine grained parallelism. It shows that retaining Python
 compatibility does not prevent the achievement of comparable performances.
@@
 
-This work has received fundings from the \texttt{Silkan} company and the French
-ANR through the CARP project. The authors would like to thanks Adrien
+This work has received fundings from the \textsc{silkan} company and the French
+ANR through the CARP project. The authors would like to thank Adrien
 \textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime,

### 2013-02-09 doc/papers/europar2013/paper.tex

@@
 and the long-term preservation of this investment is not ensured. Moreover,
-there is no falling back if the code where to be deployed in an environment
+there is no falling back if the code were to be deployed in an environment
 where the parallelizing tool/module is not available. For instance, once a code
@@
 with a Python subset called Pythran to bring fine-grain parallelism to Python.
-while being backward compatible with both the Python language, and the
-sequential algorithm.It means that:
+While being backward compatible with both the Python language, and the
+sequential algorithm. It means that:
 %
 \begin{enumerate}
-    \item Any Pythran code can be run (sequentially), with no module dependency or code change,
-        by any Python interpreter.
+    \item Any Pythran code can be run (sequentially), with no module dependency
+        or code change, by any Python interpreter.
     \item Parallelism is explicit and incrementally added to the original code
@@
   import omp
-  using = sum = sum2 = 0
+  using = sum = 0
   'omp parallel for if(using == 1)'
@@
 
-    \item Only loops can be made parallel, using a new \texttt{prange} intrinsic
-        as iterator.
+    \item Only loops can be made parallel, using a new \texttt{prange} generator.
 

### 2013-02-10 doc/papers/europar2013/paper.tex
Take into account FI's review.

@@
 implementation on 4 cores is $\times 1.46$ the execution time of the sequential
-version. GIL contention actually makes the code run slower.
+version. GIL contention actually increase the execution time.
 
@@
 
-An opposite approach consists in using the host language ---herein Python---
-to describe both parts of the system, and to let an automated tool perform the
+An opposite approach consists in using the host language ---herein Python--- to
+describe both parts of the system, and to let an automated tool perform the
 translation to native code of a part of the application, generally the
@@
 the additional development time can still benefit from a fair performance boost.
-This approach is the subject of many studies that can be classified based on
+This approach is the subject of many studies that can be classified according to
 their compatibility with the host language.
@@
 A constraining (from the performance point of view) aspect of the Python
-language is its type system. It has several consequences, among which the fact
-that each method call is resolved dynamically, including a simple add
-operation! It comes at no surprise that many approaches extend the Python
-language to add a static typing overlay. Likewise, there are only two types of
-integers (64bits integers and multi-precision integers) and one type of floating
-point type (double-precision floats) in Python, so using a type with the
-appropriate size may lead to significant performance boost.
+language is its type system. It implies that each method call is resolved
+dynamically, even a simple add operation! It comes at no surprise that many
+approaches restrain the Python language to add a static typing overlay.  Also,
+only two types of integers (64bits integers and multi-precision integers) and
+one type of floating point type (double-precision floats) are available in
+Python, so using a type with the appropriate size may lead to significant
+performance boost.
 
@@
 typing informations, calls to native functions from third party libraries, and a
-limited set of parallelism constructs, among which the possibility to define
+limited set of parallelism constructs, such as the possibility to define
 parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
@@
 PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
-Python with kernels embedded as raw strings to target accelerators.
+Python with kernels embedded as raw strings containing accelerator code.
 
 The main issue of these approaches is that they imply to modify, in a more or
-less complex way, the original code. They require to invest in a new dialect,
+less complex way, the original code. They require to learn a new dialect,
 and the long-term preservation of this investment is not ensured. Moreover,
@@
 
-Most backward-compatible approaches also require to modify the input program,
-but they do not extend the Python language, rather restrict it. As a
-consequence, they remain compatible with the original language and do not suffer
-from the drawbacks of the previous approaches. They also keep the benefit of
-existing tools associated to the language.
+Most backward-compatible approaches also require to modify the input program.
+They do not extend the Python language, but restrict it. As a consequence, they
+remain compatible with the original language and do not suffer from the
+drawbacks of the previous approaches. They also benefit from existing tools
+associated to the language.
 
-Copperhead~\cite{copperhead2011} is a functional, data parallel language embedded in
-Python. It uses n-uplets, numpy arrays and lists as its core data structure and
-prohibits usage of many control-flow operators such as loops, enforcing the use
-of the \texttt{map}, \texttt{filter} or \texttt{reduce} intrinsics to exhibit
-parallelism. In returns, it can be efficiently compiled to either CUDA or C++
-with calls to the Thrust\footnote{\emph{cf}.\
+Copperhead~\cite{copperhead2011} is a functional, data parallel language
+embedded in Python. It uses n-uplets, numpy arrays and lists as its core data
+structure and prohibits usage of many control-flow operators such as loops,
+enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
+intrinsics to exhibit parallelism. But it can be efficiently compiled to either
+CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
 \texttt{http://thrust.github.com/}} library. Python decorators are used to
@@
 
-Pythran is a subset of Python focused on scientific computing. It is implicitly
-statically typed and supports most Python constructs except those that involve
-introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
-\texttt{eval}). A few standard modules are supported in addition to the core
-language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
-currently only partially supported. The associated compiler turns Pythran code,
-possibly annotated with OpenMP and a few function type annotations, into C++
-code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
+Pythran is a subset of Python designed for scientific computing. It is
+implicitly statically typed and supports most Python constructs except those
+that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
+(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
+core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
+is currently only partially supported. The associated compiler turns Pythran
+code, possibly annotated with OpenMP and a few function type annotations, into
+C++ code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
 relies on type inference to remove most dynamic behavior, combined with a
@@
 analysis, Pythran determines for each argument whether it must be passed using
-reference or const reference to prevent this extra cost.
+reference or const reference to prevent this overhead.
 
@@
 When passing containers from Python to C back and forth, a copy of the whole
-data is made to turn the type agnostic, non-contiguous Python data into dense
-typed ones. This extra copy implies an extra cost that is not negligible:
+container is made to turn the type agnostic, non-contiguous Python data into
+dense typed ones. This extra copy implies an extra cost that is not negligible:
 Passing two lists of float from Python to C++ requires as much as half the time
 to compute the dot product of the same lists directly in Python. Following
-Amdahl's law, this transfer cost greatly hinders parallelization, and is a well
-known issue.
+Amdahl's law, this copy overhead greatly hinders the benefits of
+parallelization, and is a well known issue.
 
@@
 
-This work has received fundings from the \textsc{silkan} company and the French
-ANR through the CARP project. The authors would like to thank Adrien
-\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime,
-Albert \textsc{Cohen}, B{\'e}atrice \textsc{Creusillet}, Fabien
-\textsc{Dagnat} and Mehdi \textsc{Amini} for their valuable advices.
+This work has received fundings from \textsc{silkan} and the French ANR through the
+CARP project. The authors thank Adrien \textsc{Merlini} and Alan
+\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
+B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Mehdi \textsc{Amini},
+Fran{\,c}ois \textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable
+reviews.
 

### 2013-02-11 doc/papers/europar2013/paper.tex
Begin of Europar2013 paper proofreading

@@
 
+\marginpar{The title is to be changed}
 %
@@
 
-\keywords{Python, OpenMP, C++, static compilation, scripting language}
+\keywords{Python, OpenMP, C++, static compilation, scripting language.}
 \end{abstract}
@@
 computing~\cite{Oliphant2007}, mainly thanks to the SciPy~\cite{scipy} project
-that adds Matlab-like functionalities to the language.
+that adds MATLAB-like functionalities to the language.
 
@@
 communicate through IPC, using Python built-in object serialization. This
-approach is only viable for computation intensive application, but the
+approach is only viable for computation intensive application. Furthermore, the
 communication and synchronization overheads are much greater than a light-weight
@@
 To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
-estimate the value of $\pi$. It's execution time using a multi-threaded
+estimate the value of $\pi$. Its execution time using a multi-threaded
 implementation on 4 cores is $\times 1.46$ the execution time of the sequential
-version. GIL contention actually increase the execution time.
+version. So the GIL contention actually \emph{increases} the execution time.
 
@@
 approaches restrain the Python language to add a static typing overlay.  Also,
-only two types of integers (64bits integers and multi-precision integers) and
+only two types of integers (64-bit integers and multi-precision integers) and
 one type of floating point type (double-precision floats) are available in
@@
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% ispell-local-dictionary: "american"
+%%% TeX-auto-untabify: t
+%%% TeX-PDF-mode: t
+%%% End:
+
 % vim:spell spelllang=en

### 2013-02-11 doc/papers/europar2013/paper.tex
Correct EuroPar 2013 article up to section 2 inclusive.
Be strict with the names of other projects!

@@
 Copperhead~\cite{copperhead2011} is a functional, data parallel language
-embedded in Python. It uses n-uplets, numpy arrays and lists as its core data
+embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
 structure and prohibits usage of many control-flow operators such as loops,
@@
 Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
-Shedskin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
+Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
 turn regular Python codes into optimized ones, but they do not offer support for
@@
 This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
-with a Python subset called Pythran to bring fine-grain parallelism to Python.
-While being backward compatible with both the Python language, and the
+with a Python subset called Pythran to bring fine-grain parallelism to Python,
+while being backward compatible with both the Python language and the
 sequential algorithm. It means that:
@@
 \begin{enumerate}
-    \item Any Pythran code can be run (sequentially), with no module dependency
-        or code change, by any Python interpreter.
-    \item Parallelism is explicit and incrementally added to the original code
+    \item any Pythran code can be run (sequentially), with no module dependency
+        or code change, by any Python interpreter;
+    \item parallelism is explicit and incrementally added to the original code
         through directives.
@@
 management, the very same aspect that led to the use of a GIL in CPython.
-Memory management is implemented in Shedskin through the general-purpose Boehm
+Memory management is implemented in Shed Skin through the general-purpose Boehm
 garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
@@
 The traditional solution is to use a native type that exposes a Python interface
-and has a constant translation cost. Numpy's \texttt{ndarray} is a typical
-implementation of this concept and is the keystone of the Scipy and Numpy
-packages. Basically, a numpy \texttt{ndarray} is a raw C pointer that is exposed
+and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
+implementation of this concept and is the keystone of the Scipy and NumPy
+packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
 at the Python level. As a tool for scientific computing, Pythran supports such a

### 2013-02-11 doc/papers/europar2013/paper.tex
[Europar 2013] Few thoughts

@@
 code snippet into Python code and compile then load them at runtime.
+\marginpar{Strange to have a "partial" related work here, while weave is not even mention later (see further marginpar).}
 
@@
 communicate through IPC, using Python built-in object serialization. This
-approach is only viable for computation intensive application. Furthermore, the
+approach is only viable for computation intensive application, since the
 communication and synchronization overheads are much greater than a light-weight
@@
 several light-weight threads within the same interpreter, this approach is not
-relevant for HPC, because of a specificity of CPython, the \emph{Global
+applicable to HPC, because of a specificity of CPython, the \emph{Global
 Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
-IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
+IronPython or Jython, do not have a GIL.} 
+\marginpar{The fact that others Python implementations do not have a GIL make all the argumentation dubious, especially the last paragraph.}
+This lock ensures that only one thread
 is active at a time in the interpreter. While it makes it possible to have
@@
 version. So the GIL contention actually \emph{increases} the execution time.
+\marginpar{MA: shortcut! What proves that the issue is the GIL? You may just have written bad code...}
 
@@
 \subsubsection{Backward-Incompatible Approaches}
+\marginpar{Why weave does not fit this section and is mention in the introduction.}
 
@@
 The main issue of these approaches is that they imply to modify, in a more or
-less complex way, the original code. They require to learn a new dialect,
+less complex way, the original code.\marginpar{just as pythran} They require to learn a new dialect,
 and the long-term preservation of this investment is not ensured. Moreover,
@@
 identify hot-spots that are JIT-compiled to native code.
+\marginpar{JIT is never defined.}
 
@@
 Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
-turn regular Python codes into optimized ones, but they do not offer support for
+turn regular Python codes into optimized ones \marginpar{"optimized"? I would rather say "native"/"bytecode"}, but they do not offer support for
 fine-grained parallelism beyond what the standard library proposes.
@@
 in an environment that is not aware of the existence of the parallelizing
-solution. This principle is not respected by parallel libraries, but code
-annotations satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
+solution\marginpar{First sentence is unclear to me.}. This principle is not respected by parallel libraries, but code
+annotations satisfy it \marginpar{It is obviously not true for OpenMP. I don't see why it is not for parallel libraries!}. This is where OpenMP~\cite{openmp3.1} parallel
 annotations shine: the original code remains compatible with a compiler that is
-not aware of OpenMP.
+not aware of OpenMP.\marginpar{This is wrong.}
 
@@
 to ensure no variable gets forgotten when building such lists.
+\marginpar{I'm not sure what's the point with this paragraph. Just drop it?}
 
@@
 the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
+\marginpar{Unclear to me what you're trying to explain. Python doesn't have \&\&, how is it relevant? C/C++ does not have  By the way OpenMP 4.0 has "user-defined" reduction operator.}
 
@@
 threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order. Integers used as loop
+described by a random access iterator with a total order.\marginpar{Is total order a real need?} Integers used as loop
 indices in Fortran and C satisfy this conditions, as well as C++ iterators with
@@
 associated with its index. These generators are random iterators only if their
-input iterator is a random iterator itself.
+input iterator is a random iterator itself.\marginpar{Necessary condition, but is a sufficient one?}
 
@@
 and forward iterators for generators.
+\marginpar{This sentence is like a "cheveu sur la soupe", the whole section talks about parallelism issue, while this sentence is about language restriction in general?}
 
@@
 \texttt{omp} module is not found.
+\marginpar{omp module? First time you're talking about that. How is involved? Is it standard? ...}
 
@@
 that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
-(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
+(e.g.\ \texttt{eval}).\marginpar{class?} A few standard modules are supported in addition to the
 core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
@@
 remain explicit.
+\marginpar{I've never seen "explicit" dependencies in my C++ code, what's that?}
 
@@
 to solely rely on reference counting for memory management. It can be
-implemented through the thread-safe shared pointer mechanism provided in C++11
+implemented through the thread-safe shared pointer mechanism provided by the C++11
 standard library.
@@
 reference or const reference to prevent this overhead.
+\marginpar{what is the point about reference or const refence? I mean: who cares? Nobody reads the generated code...
+Moreover, what about copy elision made by the compiler? You're tied to a single compilation unit so the C++ compiler should be able to make it anyway.}
 
@@
 Pythran compatible kernel and parallelized the Python version using the approach
-described in this paper. We also measure the SLOC of the two versions.
+described in this paper. We also measure the SLOC of the two versions. \marginpar{unfair, and arguably relevant.}
 Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
@@
 All results are normalized against Cython sequential execution time. They show
-that while being of a higher level than Cython, Pythran achieves comparable
+that while handling code at a higher level than Cython, Pythran achieves comparable
 results.
@@
     \includegraphics[width=\textwidth]{cython}
-    \caption{Comparison of Cython en Pythran generated code performance.}
+    \caption{Comparison of Cython and Pythran generated code performance.}
     \label{fig:cython-pythran}

### 2013-02-11 doc/papers/europar2013/paper.tex
Take into account most of RK's and MA's comment.
Just for posterity:

* I do not put weave in the biblio because I only use it as an
  introduction to the concept of C interleaved with Python
* Yes, IronPython having no GIL is disturbing. I did not dig too much
  into that part
* I am pretty confident about my threading implementation. No room to say
  everything in the paper, it's just a short intro. It's in the git if
  you're interested :-)
* Simple use of OpenMP is backward compatible, medium advanced us is
  not. I updated the paper accordingly.
* It is not unfair to compare SLOC for C an

--- 2013-02-11 doc/papers/europar2013/paper.tex
@@
 
-\marginpar{The title is to be changed}
 %
@@
 several light-weight threads within the same interpreter, this approach is not
-relevant for HPC, because of a specificity of CPython, the \emph{Global
+applicable to HPC, because of a specificity of CPython, the \emph{Global
 Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
@@
 implementation on 4 cores is $\times 1.46$ the execution time of the sequential
-version. So the GIL contention actually \emph{increases} the execution time.
+version. GIL contention actually \emph{increases} the execution time.
 
@@
 approaches restrain the Python language to add a static typing overlay.  Also,
-only two types of integers (64-bit integers and multi-precision integers) and
+only two types of integers (64-bits integers and multi-precision integers) and
 one type of floating point type (double-precision floats) are available in
@@
 
-The main issue of these approaches is that they imply to modify, in a more or
-less complex way, the original code. They require to learn a new dialect,
-and the long-term preservation of this investment is not ensured. Moreover,
-there is no falling back if the code were to be deployed in an environment
-where the parallelizing tool/module is not available. For instance, once a code
-has been ported to PyCUDA and transformed into CUDA code embedded in Python
-strings, there is no way back and the developer has to maintain two versions of
-the algorithm.
+The main issue of these approaches is that they imply to modify, in a
+backward-incompatible way, the original code. They require to learn a new
+dialect, and the long-term preservation of this investment is not ensured.
+Moreover, there is no falling back if the code were to be deployed in an
+environment where the parallelizing tool/module is not available. For instance,
+once a code has been ported to PyCUDA and transformed into CUDA code embedded in
+Python strings, there is no way back and the developer has to maintain two
+versions of the algorithm.
 
@@
 Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
-Shed Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
-turn regular Python codes into optimized ones, but they do not offer support for
+Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
+turn regular Python codes into native ones, but they do not offer support for
 fine-grained parallelism beyond what the standard library proposes.
@@
 solution. This principle is not respected by parallel libraries, but code
-annotations satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
-annotations shine: the original code remains compatible with a compiler that is
-not aware of OpenMP.
+annotations partially satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
+annotations shine: the original code remains mostly compatible with a compiler
+that is not aware of OpenMP.
 
@@
 with a Python subset called Pythran to bring fine-grain parallelism to Python,
-while being backward compatible with both the Python language and the
+While being backward compatible with both the Python language, and the
 sequential algorithm. It means that:
@@
 are considered when building such variable lists: there is no concept of
-variable local to a block. Using the \texttt{default(none)} clause is possible
-to ensure no variable gets forgotten when building such lists.
+variable local to a block. %Using the \texttt{default(none)} clause is possible
+%to ensure no variable gets forgotten when building such lists.
 
@@
 
-Pythran runtime is aware of these three kind of iterators: random access
-iterators for standard containers, adaptor iterators for generator expressions
-and forward iterators for generators.
+Pythran runtime is aware of these three kind of iterators and accepts supports
+parallel iteration over random access iterators and iterators adapting random
+ones.
 
@@
 
-Providing a binding to these library in Python does not raise particular
-problems, as the signature of these functions only involves integers passed as
-parameters or return values, except for the mutex manipulation. In that case an
-opaque type is used to represent the native type.
+Providing a binding to these library in Python as an \texttt{omp} module does
+not raise particular problems, as the signature of these functions only involves
+integers passed as parameters or return values, except for the mutex
+manipulation. In that case an opaque type is used to represent the native type.
 
@@
 core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
-is currently only partially supported. The associated compiler turns Pythran
-code, possibly annotated with OpenMP and a few function type annotations, into
-C++ code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
-relies on type inference to remove most dynamic behavior, combined with a
-high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
-between Python and C++ constructs. The purpose of this section is not to dig
-into the internals of the compiler, but rather to focus on the impact of the
-parallelism layer, especially on the runtime library.
+is currently only partially supported. User classes are not supported. The
+associated compiler turns Pythran code, possibly annotated with OpenMP and a few
+function type annotations, into C++ code. Its processing is summarized in
+Figure~\ref{fig:pythran-compiler} and relies on type inference to remove most
+dynamic behavior, combined with a high-level runtime library,
+\texttt{pythonic++}, to allow a one-to-one mapping between Python and C++
+constructs. The purpose of this section is not to dig into the internals of the
+compiler, but rather to focus on the impact of the parallelism layer, especially
+on the runtime library.
 
@@
 Pythran runtime is based on the STL. The standard library of C++11, unlike its
-C++03 version, is thread-safe, so the data dependencies in the original code
-remain explicit.
+C++03 version, is thread-safe, so no data dependencies are added to the original
+code.
 
@@
 backward-compatible with Python. They translate explicit fine-grained
-parallelism through code annotations or specific python constructs,
+parallelism through code annotations or specific Python constructs,
 respectively.
@@
 Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
-mathematical, image-processing or linear algebra kernels. All these kernels
-have been written in Cython and Python ---compatible with Pythran--- and
-annotated through the mechanism of each language to exhibit parallelism, then
-compiled into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.
-Their execution time when called from the Python interpreter is measured
-through the \texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.
-All results are normalized against Cython sequential execution time. They show
-that while being of a higher level than Cython, Pythran achieves comparable
+mathematical, image-processing or linear algebra kernels. All these kernels have
+been written in Cython and Python ---compatible with Pythran--- and annotated
+through the mechanism of each language to exhibit parallelism, then compiled
+into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.  Their
+execution time when called from the Python interpreter is measured through the
+\texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.  All results
+are normalized against Cython sequential execution time. They show that while
+handling code at a higher level than Cython, Pythran achieves comparable
 results.
@@
 \end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% ispell-local-dictionary: "american"
-%%% TeX-auto-untabify: t
-%%% TeX-PDF-mode: t
-%%% End:
-
 % vim:spell spelllang=en

### 2013-04-05 doc/papers/cpc2013/paper.tex
First draft for CPC abstract submission
To be reviewed by Joël & Pierrick, and updated accordingly.

\documentclass{llncs}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}

\begin{document}
\title{Hybrid compiler-library approach to efficiently compile Numpy expressions}

\author{Serge Guelton\inst{1} \and Joël Falcou\inst{2} \and Pierrick Brunet\inst{3}}
\institute{ENS ULM, Paris, France \and LRI, University Paris Sud XI, France \and Télécom Bretagne, Plouzané, France}

\maketitle

\begin{abstract}

    The Numpy module for the Python language is growing in popularity among the
    scientific community as a scientific computing tool. It most noticeably
    provides a N-dimensional array object and a set of mathematical arrays
    using this structure that provide efficient computations thanks to the use
    of natively compiled implementation.

    Experiments shows that although more efficient than interpreted Python code,
    this module does not optimize complex array expressions and does not fully
    take advantage of multi-core machines with powerful vector instruction
    units. This paper examines the aggregation of complex Numpy expressions and
    their static compilation to efficient, hardware-aware native code to
    speedup mathematical computations.

    It shows that the combination of expression templates with a generic
    vectorisation library and OpenMP constructs can be used to turn the
    multiple sequential loops involved in complex Numpy expressions into a
    single parallel, vectorized loop that efficiently take advantage of the
    hardware capability.

    The proposed techniques are used to turn polymorphic Python functions to
    polymorphic C++ functions through template meta-programming. Specialization
    for given types is done during C++ template instantiation by the C++
    compiler.

    As the efficiency of the operator aggregation highly depends on the memory
    accesses over number of operations ratio, the paper also evaluates the
    relevancy of using forward substitution and eventually introducing
    redundant computations to increase computation intensity and hiding the
    memory latency implied by the multiple loads and stores to and from vector
    registers.

    The ideas presented in the paper are implemented in the Pythran compiler
    infrastructure using the Numeric Template Toolbox(NT$^2$). The static
    approach is compared with the Just-In-Time approach taken by the Numba
    compiler and the numexpr module, two tools used to dynamically optimize
    Numpy expressions.

\end{abstract}

%\bibliographystyle{splncs}
%\bibliography{biblio}

\end{document}

% vim:spell spelllang=en

### 2013-04-05 doc/papers/cpc2013/paper.tex

@@
 
-\author{Serge Guelton\inst{1} \and Joël Falcou\inst{2} \and Pierrick Brunet\inst{3}}
-\institute{ENS ULM, Paris, France \and LRI, University Paris Sud XI, France \and Télécom Bretagne, Plouzané, France}
+\author{Serge Guelton\inst{1} \and Pierrick Brunet\inst{2} \and Pierre Esterie\inst{3}\\
+ \and Mathias Gaunard\inst{4} \and Joël Falcou\inst{3} }
+\institute{ENS ULM, Paris, France \and Télécom Bretagne, Plouzané, France \and LRI, University Paris Sud XI, France  \and MetaScale SAS, Orsay, France}
 

### 2013-04-05 pythran/pythonic++/core/numpy_expr.h
Finish numpy ufunc.

@@ -9,11 +9,13 @@
 #include <nt2/include/functions/atan.hpp>
 #include <nt2/include/functions/atan2.hpp>
 #include <nt2/include/functions/atanh.hpp>
+#include <nt2/include/functions/bitofsign.hpp>
 #include <nt2/include/functions/ceil.hpp>
 #include <nt2/include/functions/copysign.hpp>
 #include <nt2/include/functions/cos.hpp>
 #include <nt2/include/functions/cosh.hpp>
 #include <nt2/include/functions/divfloor.hpp>
+#include <nt2/include/functions/eps.hpp>
 #include <nt2/include/functions/erf.hpp>
 #include <nt2/include/functions/erfc.hpp>
 #include <nt2/include/functions/exp.hpp>
@@ -38,7 +40,9 @@
 #include <nt2/include/functions/mod.hpp>
 #include <nt2/include/functions/nextafter.hpp>
 #include <nt2/include/functions/pow.hpp>
+#include <nt2/include/functions/rec.hpp>
 #include <nt2/include/functions/remainder.hpp>
+#include <nt2/include/functions/sign.hpp>
 #include <nt2/include/functions/sin.hpp>
 #include <nt2/include/functions/sinh.hpp>
 #include <nt2/include/functions/sqrt.hpp>
@@ -158,18 +162,17 @@ namespace  pythonic {
                     return t0 != t1;
                 }
 
-            template<class T >
-                auto reciprocal(T const& t) -> decltype(T(1)/t) {
-                    return T(1)/t;
-                }
-
             template<class T0, class T1>
                 auto right_shift(T0 const& t0, T1 const& t1) -> decltype(t0 >> t1) {
                     return t0 >> t1;
                 }
+            template<class T>
+                auto square(T const& t) -> decltype(t*t) {
+                    return t*t;
+                }
 
             template<class T0, class T1>
-                auto minus(T0 const& t0, T1 const& t1) -> decltype(t0 - t1) {
[…]

### 2013-04-05 pythran/pythonic++/modules/numpy.h
Finish numpy ufunc.

@@ -652,7 +652,7 @@ namespace pythonic {
 
         NP_PROXY_ALIAS(radians, nt2::inrad);
 
-        NP_PROXY_ALIAS(reciprocal, pythonic::numpy_expr::ops::reciprocal);
+        NP_PROXY_ALIAS(reciprocal, nt2::rec);
 
         NP_PROXY(remainder);
 
@@ -660,9 +660,29 @@ namespace pythonic {
 
         NP_PROXY_ALIAS(rint, nt2::iround)
 
+        NP_PROXY(sign);
+
+        NP_PROXY_ALIAS(signbit, nt2::bitofsign)
+
         NP_PROXY(sin);
 
+        NP_PROXY(sinh);
+
+        NP_PROXY_ALIAS(spacing, nt2::eps)
+
+        NP_PROXY(sqrt);
+
+        NP_PROXY_ALIAS(square, pythonic::numpy_expr::ops::square);
+
+        NP_PROXY_ALIAS(subtract, pythonic::numpy_expr::ops::subtract);
+
         NP_PROXY(tan);
+
+        NP_PROXY(tanh);
+
+        NP_PROXY_ALIAS(true_divide, pythonic::numpy_expr::ops::divide); 
+
+        NP_PROXY(trunc);
 
 #undef NP_PROXY
 #undef NAMED_OPERATOR

### 2013-04-05 pythran/tables.py

@@ -338,10 +338,20 @@ modules = {
             "right_shift": ConstFunctionIntr(),
             "rint": ConstFunctionIntr(),
             "reshape": ConstMethodIntr(),
+            "sign": ConstFunctionIntr(),
+            "signbit": ConstFunctionIntr(),
             "sin": ConstFunctionIntr(),
+            "sinh": ConstFunctionIntr(),
+            "spacing": ConstFunctionIntr(),
+            "sqrt": ConstFunctionIntr(),
+            "square": ConstFunctionIntr(),
+            "subtract": ConstFunctionIntr(),
             "sum": ConstMethodIntr(),
             "tan": ConstFunctionIntr(),
+            "tanh": ConstFunctionIntr(),
             "transpose": ConstMethodIntr(),
+            "true_divide": ConstFunctionIntr(),
+            "trunc": ConstFunctionIntr(),
             "uint8" : ConstFunctionIntr(),
             "uint16" : ConstFunctionIntr(),
             "uint32" : ConstFunctionIntr(),

### 2013-04-05 pythran/tests/test_numpy.py

@@ -301,6 +301,8 @@ binary_ufunc = (
         #'nextafter','not_equal', 
         #'power',
         #'remainder','right_shift', 
+        #'subtract',
+        'true_divide',
         )
 
 unary_ufunc = (
@@ -313,7 +315,10 @@ unary_ufunc = (
         #'isinf', 'isnan', 'invert', 'isfinite',
         #'log10', 'log1p', 'log2', 'logical_not',
         #'negative',
-        'rint', #'rad2deg', 'radians','reciprocal', 
+        #'rad2deg', 'radians','reciprocal', 'rint', 
+        #'sign', 'signbit', 
+        # 'sin', 'sinh', 'spacing', 'sqrt', 'square', 
+        #'tan', 'tanh','trunc', 
         )
 
 for f in unary_ufunc:
@@ -325,7 +330,7 @@ for f in unary_ufunc:
         setattr(TestNumpy, 'test_' + f + '_scalar', eval("lambda self: self.run_test('def np_{0}_scalar(): from numpy import {0} ; a = 0.5 ; return {0}(a+0.5)', np_{0}_scalar=[])".format(f)))
 
 for f in binary_ufunc:
-    if 'bitwise_' in f or 'ldexp' in f or '_shift' in f:
+    if 'bitwise_' in f or 'ldexp' in f or '_shift' in f :
         setattr(TestNumpy, 'test_' + f, eval("lambda self: self.run_test('def np_{0}(): from numpy import ones, int32, {0} ; a = ones(10, int32) ; return {0}(a,a)', np_{0}=[])".format(f)))
         setattr(TestNumpy, 'test_' + f + '_scalar', eval("lambda self: self.run_test('def np_{0}_scalar(): from numpy import {0} ; a = 1 ; return {0}(a, a-1)', np_{0}_scalar=[])".format(f)))
     else:

### 2013-04-06 doc/papers/cpc2013/paper.tex
Take into account Pierrick's advice.

@@
     scientific community as a scientific computing tool. It most noticeably
-    provides a N-dimensional array object and a set of mathematical arrays
+    provides a N-dimensional array object and a set of mathematical operators
     using this structure that provide efficient computations thanks to the use

### 2013-06-09 doc/papers/dls2013/paper.tex
Add dls paper

--- 2013-02-11 doc/papers/europar2013/paper.tex
@@
 % max 12 pages
-\documentclass{llncs}
+\documentclass{sigplanconf}
 %
@@
 %
-\title{Pythran: Bringing OpenMP to a Python Subset}
+\title{Experimenting OpenMP/Python Coupling}
 
-\author{Serge Guelton\inst{1,2} \and Pierrick Brunet\inst{2}}
-
-\institute{
-    {\'E}cole Normale Sup{\'e}rieure, Paris, France
-    \and
-    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France
-}
+\authorinfo{Serge Guelton}{{\'E}cole Normale Sup{\'e}rieure, Paris, France}{serge.guelton@telecom-bretagne.eu}
+\authorinfo{Pierrick Brunet}{T{\'e}l{\'e}com Bretagne, Plouzan{\'e},France}{pierrick.brunet@telecom-bretagne.eu}
 
@@
     fast as its naive C equivalent, while keeping the input code totally
-    compatible with the python interpreter. The approach is validated through
+    compatible with the Python interpreter. The approach is validated through
     the Pythran compiler.
 
+\end{abstract}
 \keywords{Python, OpenMP, C++, static compilation, scripting language.}
-\end{abstract}
 
@@
 within the Python language in order to add fine-grain parallelism support to
-Python in the context of a Python subset called Pythran. Static compilation of
-this language into C++ and the underlying runtime is described in
-Section~\ref{sec:python-static}. The Pythran compiler is benchmarked on several
-scientific kernels and compared to Cython in Section~\ref{sec:validation}.
+Python in the context of a Python subset called Pythran~\cite{pythran2013}.
+Static compilation of this language into C++ and the underlying runtime is
+described in Section~\ref{sec:python-static}. The Pythran compiler is
+benchmarked on several scientific kernels and compared to Cython in
+Section~\ref{sec:validation}.
 
@@
 In the many cores area, turning to parallelism to balance the performance
-limitations of scripting languages, as described 
-% in~\cite{choy05} in the context of the MATLAB language, or
-in~\cite{mals07} in the context of the R language, seems legitimate. In the
-context of Python, most approaches have focused on fork-based parallelism.
+limitations of scripting languages, as described in~\cite{choy05} in the context
+of the MATLAB language, or in~\cite{mals07} in the context of the R language,
+seems legitimate. In the context of Python, most approaches have focused on
+fork-based parallelism.
 
@@
 
+There have also been several approaches to replace the GIL by Transactional
+Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
+mainstream interpreters.
+
 As a consequence, Python developers need to write multi-threaded native
@@
 computations referred as hybrid computations.
-
-%There have been several approaches to replace the GIL by Transactional
-%Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
-%mainstream interpreters.
 
@@
 similar approach that mixes Python with C, using raw strings to hold the C code
-and the parallel directives, that are limited to parallel for loops.
-%The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
-%compiler uses additional type information to generate sequential LLVM bytecode.
-PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
-Python with kernels embedded as raw strings containing accelerator code.
+and the parallel directives, that are limited to parallel for loops.  The
+numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
+uses additional type information to generate sequential LLVM bytecode.  PyCUDA
+and PyOpenCL~\cite{klockner2012} also target accelerators by mixing Python with
+kernels embedded as raw strings containing accelerator code.
 
@@
 fine-grained parallelism beyond what the standard library proposes.
-%~\footnote{Shedskin is compatible with several Python modules that provide
-%coarse-grained, fork-based parallelism.}
 
@@
 OpenMP is a standard API for parallel programming for Fortran, C and C++. It
-consists in a set of parallelizing directives and a few runtime library calls.
+consists of a set of parallelizing directives and a few runtime library calls.
 If OpenMP is not activated, the directives are ignored, thus enabling
@@
 
-While Pythran has a decorator mechanism\footnote{\see PEP 318,
+While Python has a decorator mechanism\footnote{\see PEP 318,
 \url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
 Python decorators.}, it only applies to functions, methods and classes and does
-not make it possible to attach decorations to other statements. Pythran uses
-the path taken by PLW~\cite{dongara2007}, that uses string instructions to hold
-such decorations. As Python does not have anonymous block,
-%~\footnote{\see PEP340, \url{http://www.python.org/dev/peps/pep-0340/} for a
-%discussion concerning anonymous block support in Python.}
-one has to create a dummy \texttt{if 1:} instruction to apply an annotation to a
-whole block. Alternatively, the syntax \texttt{if 'my annotation':} is also
-supported.
+not make it possible to attach decorations to other statements. Pythran uses the
+path taken by PLW~\cite{dongara2007}, that uses string instructions to hold such
+decorations. As Python does not have anonymous block,~\footnote{\see PEP340,
+    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
+anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
+instruction to apply an annotation to a whole block. Alternatively, the syntax
+\texttt{if 'my annotation':} is also supported.
 
@@
 \texttt{enumerate} builtin, that yields each element of the enumerated iterator
-associated with its index. These generators are random iterators only if their
+associated with its index. These generators are random access iterators only if their
 input iterator is a random iterator itself.
@@
 'omp parallel for'
-for i, v in enumerate([2, 3, 5, 7, 11]): print i, ':', v
+for i, v in enumerate([2, 3, 5, 7, 11]):
+    print i, ':', v
   \end{lstlisting}
@@
 
-Pythran runtime is aware of these three kind of iterators and accepts supports
+Pythran's runtime is aware of these three kind of iterators and supports
 parallel iteration over random access iterators and iterators adapting random
@@
     sum += i
-  known_sum = (loop_count * (loop_count + 1)) / 2
+  known_sum = (loop_count *
+              (loop_count + 1)) / 2
   return known_sum == sum and num_threads == 1
@@
     \centering
-    \includegraphics[width=\textwidth]{hyantes}
+    \includegraphics[width=.5\textwidth]{hyantes}
 \end{figure}
@@
     with nogil, parallel():
-        for i in prange(10000000): r += sqrt(i)
+        for i in prange(10000000):
+            r += sqrt(i)
     return r
@@
 def sum_sqrt(r):
-    "omp parallel for private(i) reduction(+:r)"
-    for i in xrange(10000000): r += math.sqrt(i)
+    "omp parallel for reduction(+:r)"
+    for i in xrange(10000000):
+        r += math.sqrt(i)
     return r
@@
 The source codes used for these benchmarks are available on the Pythran
-repository.\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran} on \texttt{europar2013}
+repository.\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran}
+    on \texttt{dls2013}
 branch.}
@@
 \begin{figure}[ht]
-    \includegraphics[width=\textwidth]{cython}
+    \includegraphics[width=.5\textwidth]{cython}
     \caption{Comparison of Cython and Pythran generated code performance.}
@@
 %=========================================================
-\section*{Acknowledgments}
+\acks
 %=========================================================
@@
 
-\bibliographystyle{splncs03}
+\bibliographystyle{abbrvnat}
 \bibliography{biblio}

### 2013-06-09 doc/papers/dls2013/bench/cython/mandel_cython.pyx

@@ -1,0 +1,31 @@
+cimport cython
+cimport numpy as np
+import numpy as np
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+
+@cython.boundscheck(False)
+def mandel(int size, int x_center, int y_center, int max_iteration):
+    cdef int i, j, iteration, color_value
+    cdef double x, y, a, b
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] out = np.zeros((size,size))
+    for i in xrange(size):
+        for j in xrange(size):
+            x = x_center + 4.0*float(i-size/2)/size
+            y = y_center + 4.0*float(j-size/2)/size
+
+            a = 0.0
+            b = 0.0
+            iteration = 0
+
+            while (a**2 + b**2 <= 4.0 and iteration < max_iteration):
+                a= a**2 - b**2 + x
+                b= 2*a*b + y
+                iteration += 1
+            if iteration == max_iteration:
+                color_value = 255
+            else:
+                color_value = iteration*10 % 255
+            out[i,j]=color_value
+    return out

### 2013-06-09 doc/papers/dls2013/bench/cython/pi_buffon_cython.pyx

@@ -1,0 +1,20 @@
+#runas pi_estimate(40000000)
+#pythran export pi_estimate(int)
+from libc.math import sqrt, pow
+from random import random
+cimport cython
+
+def pi_estimate(int DARTS):
+    cdef double hits = 0.0
+    cdef double x, y, dist, pi
+    "omp parallel for private(i,x,y,dist), reduction(+:hits)"
+    for i in xrange (0, DARTS):
+        x = random()
+        y = random()
+        dist = sqrt(pow(x, 2) + pow(y, 2))
+        if dist <= 1.0:
+            hits += 1.0
+    # hits / throws = 1/4 Pi
+    pi = 4 * (hits / DARTS)
+    return pi
+

### 2013-06-09 doc/papers/dls2013/bench/cython/pi_buffon_para.pyx

@@ -1,0 +1,21 @@
+#runas pi_estimate(40000000)
+#pythran export pi_estimate(int)
+from libc.math cimport sqrt, pow
+from random import random
+from cython.parallel import parallel, prange
+cimport cython
+
+def pi_estimate(int DARTS):
+    cdef double hits = 0.0
+    cdef double x, y, dist, pi, i
+    with nogil, parallel():
+        for i in prange (DARTS):
+            x = random()
+            y = random()
+            dist = sqrt(pow(x, 2.) + pow(y, 2.))
+            if dist <= 1.0:
+                hits += 1.0
+    # hits / throws = 1/4 Pi
+    pi = 4 * (hits / DARTS)
+    return pi
+

### 2013-06-09 doc/papers/dls2013/bench/cython/wave_.pyx

@@ -1,0 +1,32 @@
+import numpy as np
+cimport numpy as np
+cimport cython
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+@cython.boundscheck(False)
+def wave(double t, double t_stop, double dt, double dx, double dy, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] u, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] um, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] k):
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] new_u
+    while t <= t_stop:
+        t += dt
+        new_u = calculate_u(dt, dx, dy, u, um, k)
+        um = u
+        u = new_u
+    return u
+
+@cython.boundscheck(False)
+def calculate_u(double dt, double dx, double dy, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] u, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] um, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] k):
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] up = np.zeros((u.shape[0], u.shape[1]))
+    cdef int i,j
+    "omp parallel for private(i,j)"
+    for i in xrange(1, u.shape[0]-1):
+        for j in xrange(1, u.shape[1]-1):
+            up[i,j] = 2*u[i,j] - um[i,j] + \
+                (dt/dx)**2*(
+                        (0.5*(k[i+1,j] + k[i,j])*(u[i+1,j] - u[i,j]) -
+                            0.5*(k[i,j] + k[i-1,j])*(u[i,j] - u[i-1,j]))) + \
+                        (dt/dy)**2*(
+                                (0.5*(k[i,j+1] + k[i,j])*(u[i,j+1] - u[i,j]) -
+                                    0.5*(k[i,j] + k[i,j-1])*(u[i,j] - u[i,j-1])))
+    return up
+

### 2013-06-09 doc/papers/dls2013/bench/cython/wave_para.pyx

@@ -1,0 +1,34 @@
+import numpy as np
+cimport numpy as np
+cimport cython
+from cython.parallel import parallel, prange
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+@cython.boundscheck(False)
+def wave(double t, double t_stop, double dt, double dx, double dy, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] u, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] um, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] k):
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] new_u
+    while t <= t_stop:
+        t += dt
+        new_u = calculate_u(dt, dx, dy, u, um, k)
+        um = u
+        u = new_u
+    return u
+
+@cython.boundscheck(False)
+@cython.cdivision(True)
+def calculate_u(double dt, double dx, double dy, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] u, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] um, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] k):
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] up = np.zeros((u.shape[0], u.shape[1]))
+    cdef int i,j
+    with nogil, parallel():
+        for i in prange(1, u.shape[0]-1):
+            for j in xrange(1, u.shape[1]-1):
+                up[i,j] = 2*u[i,j] - um[i,j] + \
+                    (dt/dx)**2*(
+                            (0.5*(k[i+1,j] + k[i,j])*(u[i+1,j] - u[i,j]) -
+                                0.5*(k[i,j] + k[i-1,j])*(u[i,j] - u[i-1,j]))) + \
+                            (dt/dy)**2*(
+                                    (0.5*(k[i,j+1] + k[i,j])*(u[i,j+1] - u[i,j]) -
+                                        0.5*(k[i,j] + k[i,j-1])*(u[i,j] - u[i,j-1])))
+    return up
+

### 2013-06-09 doc/papers/dls2013/bench/pythran/pairwise.py

@@ -1,0 +1,19 @@
+#from http://jakevdp.github.com/blog/2012/08/24/numba-vs-cython/
+#runas import random,numpy ; X = [ [random.random() for i in xrange(10000) ] for j in xrange(100) ] ; X = numpy.array(X) ; pairwise(X)
+#pythran export pairwise(float [] [])
+
+import math, numpy
+def pairwise(X):
+    M = X.shape[0]
+    N = X.shape[1]
+    D = numpy.zeros((M,N))
+    "omp parallel for private(i,j,d,k,tmp)"
+    for i in xrange(M):
+        for j in xrange(M):
+            d = 0.0
+            for k in xrange(N):
+                tmp = X[i,k] - X[j,k]
+                d += tmp * tmp
+            D[i,j] = math.sqrt(d)
+    return D
+

### 2013-06-09 doc/papers/dls2013/bench/pythran/pi_buffon.py

@@ -1,0 +1,18 @@
+#skip.runas pi_estimate(40000000)
+#pythran export pi_estimate(int)
+from math import sqrt, pow
+from random import random
+
+def pi_estimate(DARTS):
+    hits = 0
+    "omp parallel for private(i,x,y,dist), reduction(+:hits)"
+    for i in xrange (0, DARTS):
+    	x = random()
+    	y = random()
+    	dist = sqrt(pow(x, 2) + pow(y, 2))
+    	if dist <= 1.0:
+            hits += 1.0
+    # hits / throws = 1/4 Pi
+    pi = 4 * (hits / DARTS)
+    return pi
+

### 2013-06-09 doc/papers/dls2013/bench/pythran/setup.py

@@ -1,0 +1,61 @@
+from distutils.core import setup
+from distutils.core import Command
+from distutils.extension import Extension
+from os import path
+import pythran
+
+class build_ext(Command):
+
+    description = 'build pythran extension',
+    user_options = []
+
+    def initialize_options(self):
+        pass
+
+    def finalize_options(self):
+        self.extensions = self.distribution.ext_modules
+
+    def run(self):
+        import os
+        for extension in self.extensions:
+            source = extension.sources[0]
+            out = extension.name + '.so'
+            if not path.isfile(out) or path.getmtime(source) > path.getmtime(out):
+                print 'building pythran extension', extension.name
+                content = file(source).read()
+
+                module = pythran.cxx_generator(
+                        extension.name,
+                        content,
+                        pythran.spec_parser(content),
+                        []
+                        )
+
+                pythran.compile(
+                        os.environ.get('CXX','c++'),
+                        module,
+                        out,
+                        cxxflags = ['-O2', '-g'] + extension.extra_compile_args + extension.extra_link_args
+                        )
+
+
+setup(
+        cmdclass = {'build_ext': build_ext},
+        ext_modules = [
+            Extension("convolution", ["convolution.py"]),
+            Extension("convolution_omp", ["convolution.py"], extra_compile_args=['-fopenmp']),
+            Extension("fbcorr",   ["fbcorr.py"]),
+            Extension("fbcorr_omp",   ["fbcorr.py"], extra_compile_args=['-fopenmp']),
+            Extension("mandel", ["mandel.py"]),
+            Extension("mandel_omp", ["mandel.py"], extra_compile_args=['-fopenmp']),
+            Extension("matmul", ["matmul.py"]),
+            Extension("matmul_omp", ["matmul.py"], extra_compile_args=['-fopenmp']),
+            Extension("wave_", ["wave_.py"]),
+            Extension("wave__omp", ["wave_.py"], extra_compile_args=['-fopenmp']),
+            Extension("pairwise", ["pairwise.py"]),
+            Extension("pairwise_omp", ["pairwise.py"], extra_compile_args=['-fopenmp']),
+            Extension("hyantes", ["hyantes.py"]),
+            Extension("hyantes_omp", ["hyantes.py"], extra_compile_args=['-fopenmp']),
+            Extension("pi_buffon_omp", ["pi_buffon.py"], extra_compile_args=['-fopenmp']),
+            ]
+     )

### 2013-06-09 doc/papers/dls2013/pi_buffon.py

@@ -1,0 +1,50 @@
+from math import sqrt, pow
+from random import random
+import timeit
+
+def pi_estimate(darts):
+    hits = 0
+    for i in xrange (0, darts):
+    	x = random()
+    	y = random()
+    	dist = sqrt(pow(x, 2) + pow(y, 2))
+    	if dist <= 1.0:
+            hits += 1.0
+    # hits / throws = 1/4 Pi
+    pi = 4 * (hits / darts)
+    return pi
+
+from threading import Thread
+def threaded_pi_estimate(darts):
+    t1 = Thread(target=pi_estimate, args=(darts//4,))
+    t2 = Thread(target=pi_estimate, args=(darts//4,))
+    t3 = Thread(target=pi_estimate, args=(darts//4,))
+    t4 = Thread(target=pi_estimate, args=(darts//4,))
+    t1.start() ; t2.start()
+    t3.start() ; t4.start()
+    t1.join() ; t2.join()
+    t3.join() ; t4.join()
+
+from multiprocessing import Pool
+def multi_pi_estimate(darts):
+    p = Pool(4)
+    p.map(pi_estimate,[darts//4]*4)
+
+if __name__ == '__main__':
+    median = lambda x: sorted(x)[len(x) / 2]
+    nb_iter = 21
+
+    ti = timeit.Timer("pi_estimate(400000)", "from __main__ import pi_estimate")
+    stiming = median(ti.repeat(nb_iter, number=1))
+    print "sequential", stiming
+
+    ti = timeit.Timer("threaded_pi_estimate(400000)", "from __main__ import threaded_pi_estimate")
+    ttiming = median(ti.repeat(nb_iter, number=1))
+    print "threaded", ttiming
+    print "ratio", ttiming/stiming
+
+    ti = timeit.Timer("multi_pi_estimate(400000)", "from __main__ import multi_pi_estimate")
+    mtiming = median(ti.repeat(nb_iter, number=1))
+    print "multi", mtiming
+    print "ratio", mtiming/stiming
+

### 2013-09-01 doc/papers/sc2013/paper.tex
Add first draft for the pyhpc workshop paper
http://www.dlr.de/sc/en/desktopdefault.aspx/tabid-8686/14946_read-37159/

Deadline: September 6!

--- 2013-06-09 doc/papers/dls2013/paper.tex
@@
-% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
-% max 12 pages
-\documentclass{sigplanconf}
+% to submit via
+% https://www.easychair.org/account/signin.cgi?timeout=1;conf=pyhpc2013
+% max 10 pages
+\documentclass[conference]{IEEEtran}
 %
@@
 \lstset{language=python,morekeywords={cimport, cdef, with},
-basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray}}
+basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray},
+captionpos=b}
 
@@
 %
-\title{Experimenting OpenMP/Python Coupling}
+\title{Compiling Python Modules to Native Parallel Modules Using Pythran and
+OpenMP Annotations}
 
-\authorinfo{Serge Guelton}{{\'E}cole Normale Sup{\'e}rieure, Paris, France}{serge.guelton@telecom-bretagne.eu}
-\authorinfo{Pierrick Brunet}{T{\'e}l{\'e}com Bretagne, Plouzan{\'e},France}{pierrick.brunet@telecom-bretagne.eu}
+\author{
+    \IEEEauthorblockN{Serge Guelton}
+    \IEEEauthorblockA{
+        {\'E}cole Normale Sup{\'e}rieure, Paris, France\\
+        Email: \texttt{serge.guelton@telecom-bretagne.eu}\\
+    }
+    \and
+    \IEEEauthorblockN{Pierrick Brunet}
+    \IEEEauthorblockA{
+        T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France\\
+        Email: \texttt{pierrick.brunet@telecom-bretagne.eu}\\
+    }
+    \and
+    \IEEEauthorblockN{Mehdi Amini}
+    \IEEEauthorblockA{
+        Silkan, USA\\
+        Email: \texttt{mehdi.amini@silkan.com}
+    }
+}
 
@@
 
-    The Python language offers an interesting solution as a scientific
-    computing language thanks to the SciPy package. However, the lack of
-    fine-grained parallelism support remains an important shortcoming. This
-    paper proposes to add OpenMP directives to the Python language and shows
-    that it is possible to turn a module written in a large Python subset
-    enhanced with these directives into a native parallel module that runs as
-    fast as its naive C equivalent, while keeping the input code totally
-    compatible with the Python interpreter. The approach is validated through
-    the Pythran compiler.
+    HPC users traditionally rely on low-level, compiled language such as C or
+    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
+    common situation to have HPC application written in a high-level language
+    such as Python, calling native routines for computation-intensive tasks. To
+    improve development speed and reduce maintenance costs, several solutions
+    such as Cython, Numba, Parakeet or Pythran, have appeared to make it
+    possible to automatically or semi-automatically turn Python functions into
+    native ones.
+
+    One of the key point to match the performance of native applications is the
+    ability to write parallel applications. This paper studies the addition of
+    OpenMP directives, a popular model to describe the parallelism of
+    C/C++/FORTRAN applications, to Pythran, an automatic compiler from a subset
+    of Python to C++. It shows that scientific Python applications annotated
+    with OpenMP directives can be turned by an automatic compiler into native
+    applications that run within the same order of magnitude than
+    manually-written ones.
 
 \end{abstract}
-\keywords{Python, OpenMP, C++, static compilation, scripting language.}
 
@@
 
-The prohibitive performance overhead of the language implied by its interpreted
-and highly dynamic nature does not prevent its usage as a language for high
-performance computing. SciPy overcomes this issue through low-level routines
-written in C or Fortran and encapsulated in Python \emph{native} modules.
-Moreover, its core data structure, the multi-dimensional
-array~\cite{numpyarray2011}, has been designed so that the underlying data are
-available to both native modules and the Python interpreter without conversion
-cost.
+It is now heading toward High Performance Computing(HPC), either as a glue
+language used to bind several native libraries together, or to write whole
+applications. However, the prohibitive performance overhead of the language
+implied by its interpreted and highly dynamic nature should prevent its usage
+for critical code sections, where native code is generally used. The SciPy
+modules overcomes this issue through low-level routines written in C or Fortran
+and encapsulated in Python \emph{native} modules.  Moreover, its core data
+structure, the multi-dimensional array~\cite{numpyarray2011}, has been designed
+so that the underlying data are available to both native modules and the Python
+interpreter without conversion cost.
 
@@
 This hybrid approach, based on a mix of interpreted and native code, is getting
-widespread in the Python landscape. Section~\ref{sec:python-parallel} studies
-the existing approaches and shows a critical lack of parallelism support, and
-emphasises on the need for a backward-compatible approach.
-Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
-within the Python language in order to add fine-grain parallelism support to
-Python in the context of a Python subset called Pythran~\cite{pythran2013}.
-Static compilation of this language into C++ and the underlying runtime is
-described in Section~\ref{sec:python-static}. The Pythran compiler is
-benchmarked on several scientific kernels and compared to Cython in
-Section~\ref{sec:validation}.
+widespread in the Python landscape. Section~\ref{sec:python-parallel} gives an
+overview of existing approaches to compile Python functions or modules and shows
+a critical lack of parallelism support, and emphasises on the need for a
+backward-compatible approach.  Section~\ref{sec:python-static} presents the
+Pythran~\cite{pythran2013} compiler, a tool that turns Python modules into C++
+meta-programs, and the underlying runtime. Section~\ref{sec:python-openmp}
+studies the validity of using OpenMP directives within the Python language in
+order to add fine-grain parallelism support to Python in the context of Pythran.
+The Pythran compiler is benchmarked on several scientific kernels and compared
+to Cython in Section~\ref{sec:validation}.
 
@@
 
-Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
-typing informations, calls to native functions from third party libraries, and a
-limited set of parallelism constructs, such as the possibility to define
-parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
-similar approach that mixes Python with C, using raw strings to hold the C code
-and the parallel directives, that are limited to parallel for loops.  The
-numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}} compiler
-uses additional type information to generate sequential LLVM bytecode.  PyCUDA
-and PyOpenCL~\cite{klockner2012} also target accelerators by mixing Python with
-kernels embedded as raw strings containing accelerator code.
+Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
+syntax with typing informations, calls to native functions from third party
+libraries, and a limited set of parallelism constructs, such as the possibility
+to define parallel loops, but no task parallelism. When possible, it unboxes
+Python variable to improve performance. Without type annotation, the performance
+improvement is not terrific, but given enough type annotation, Cython can
+generates code that runs as fast as its C equivalent, while maintaining a syntax
+close to Python.
+
+PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose an other approach
+that directly mixes Python with C, using raw strings to hold the C code. PLW
+also supports parallel directives that are limited to parallel for loops.
+PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
+Python with kernels embedded as raw strings containing accelerator code.
 
@@
 
+The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
+compiler uses additional type information to generate sequential LLVM bytecode.
+Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
+Python to LLVM bytecode translation, but limits its scope to the numpy package,
+only supporting a small subset of the Python language. Additionally, it supports
+implicit parallelism using an implicit mapping between numpy functions and a set
+of parallel primitives including \texttt{map}s \texttt{scan}s and
+\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
+not suffer from the backward-incompatibility issues. When meeting an unsupported
+construct, numba falls back to Python C-API calls, while parakeet just raises an
+exception.
+
 Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
@@
 %=========================================================
-\section{OpenMP Semantical Adaptations}\label{sec:python-openmp}
-%=========================================================
-
-OpenMP is a standard API for parallel programming for Fortran, C and C++. It
-consists of a set of parallelizing directives and a few runtime library calls.
-If OpenMP is not activated, the directives are ignored, thus enabling
-incremental parallelization of the original source code while keeping the
-original code structure. The languages targeted by OpenMP are statically
-compiled. This section studies the semantical adaptation required to use the
-same API for a scripting language such as Python.
-
-\subsection{Directives}
-
-OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them
-apply to structured blocks, represented by an instruction in C/C++ and delimited
-by comments in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
-attached to a specific instruction.
-
-While Python has a decorator mechanism\footnote{\see PEP 318,
-\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
-Python decorators.}, it only applies to functions, methods and classes and does
-not make it possible to attach decorations to other statements. Pythran uses the
-path taken by PLW~\cite{dongara2007}, that uses string instructions to hold such
-decorations. As Python does not have anonymous block,~\footnote{\see PEP340,
-    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
-anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
-instruction to apply an annotation to a whole block. Alternatively, the syntax
-\texttt{if 'my annotation':} is also supported.
-
-%Listing~\ref{lst:instruction-annotation} shows how to annotate a
-%parallel loop with reduction this way.
-%
-%\begin{figure}
-%    \begin{lstlisting}[label={lst:instruction-annotation}, caption={Parallel
-%    loop with reduction annotated with OpenMP directive.}]
-%'omp parallel for reduction(+:r)'
-%for auto in my_container: r += x
-%  \end{lstlisting}
-%\end{figure}
-
-Many OpenMP annotations are parametrized by clauses that list variables,
-specifying their behavior with respect to parallel regions, e.g.\
-\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
-variables that have already been declared. However, there is no variable
-declaration in Python, and all variables assigned in a function have the
-function scope. Consequently, all variables that are referenced in a function
-are considered when building such variable lists: there is no concept of
-variable local to a block. %Using the \texttt{default(none)} clause is possible
-%to ensure no variable gets forgotten when building such lists.
-
-The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
-characterize some data dependencies when performing a parallel reduction. The
-list of supported operators depend on the input and backend languages: Python
-has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
-\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
-the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
-
-\subsubsection{Parallel For and Iterators}
-
-The core directive to handle data parallelism is the \texttt{for} directive that
-distributes the iteration space of the associated loop among the existing
-threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order. Integers used as loop
-indices in Fortran and C satisfy this conditions, as well as C++ iterators with
-the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
-step of one until it is exhausted, in which case it raises an exception: it
-behaves as a forward iterator. To be compatible with OpenMP, these iterators
-must be turned into random access iterators.
-
-The extension of Python iterator to random access iterators is direct for the
-standard containers: \texttt{list}\footnote{Python \texttt{list}s behave as C++
-\texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
-care.
-
-Generators, Python objects that behave like iterators, are commonly used in
-Python. The simplest one, \texttt{xrange(\emph{start}, \emph{stop},
-\emph{step})}, successively yields value starting from \texttt{\emph{start}} to
-\texttt{\emph{stop}} by a step of \texttt{\emph{step}}. It is easily extended to
-support random access, but it generally does not make sense to use a generator
-as a loop iterator, as the relation between two random states of the iterator
-may be of an arbitrary complexity.
-
-Generator expressions are generators whose content is built from an other
-iterator. For instance \texttt{(x*x for x in l)} successively yields the square
-of each element in \texttt{l}. They behave like adaptors: they apply a
-particular expression on each value of the iterator. It is also the case of the
-\texttt{enumerate} builtin, that yields each element of the enumerated iterator
-associated with its index. These generators are random access iterators only if their
-input iterator is a random iterator itself.
-
-Finally, if the iterator yields a tuple, it is possible to unpack it inside the
-\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
-all the unpacked variables are considered as iterators, especially with respect
-to default privatization rules: in the given example, \texttt{i} and \texttt{v}
-are private, and the parallel iteration is valid because the input of
-\texttt{enumerate} is a list, which allows random access iteration.
-
-\begin{figure}
-    \begin{lstlisting}[label={lst:complex-omp-for}, caption={Parallel
-    loop in Pythran with tuple unpacking.}]
-'omp parallel for'
-for i, v in enumerate([2, 3, 5, 7, 11]):
-    print i, ':', v
-  \end{lstlisting}
-\end{figure}
-
-Pythran's runtime is aware of these three kind of iterators and supports
-parallel iteration over random access iterators and iterators adapting random
-ones.
-
-\subsection{Runtime Library}
-
-OpenMP provides a small runtime library that, for instance, makes it possible to
-retrieve the active thread id. All the functions are declared in the
-\texttt{<omp.h>} header, and have a default behavior when OpenMP is not
-activated.
-
-Providing a binding to these library in Python as an \texttt{omp} module does
-not raise particular problems, as the signature of these functions only involves
-integers passed as parameters or return values, except for the mutex
-manipulation. In that case an opaque type is used to represent the native type.
-
-The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
-activated, and can be used to detect when OpenMP is not available. Python does
-not have a preprocessor, but it is possible to catch the import exception if the
-\texttt{omp} module is not found.
-
-\subsection{Validation}
-
-A validation suite for OpenMP is proposed in~\cite{wang2012} for C and Fortran.
-We ported it to Python, and also extended it to validate the corner cases
-specific to Python described in this section. A typical test case is given in
-Listing~\ref{lst:openmp-validation}.
-
-\begin{figure}
-    \begin{lstlisting}[label={lst:openmp-validation},caption={Example of Python
-    OpenMP validation test case.}]
-def omp_parallel_for_if(loop_count):
-  import omp
-  using = sum = 0
-  'omp parallel for if(using == 1)'
-  for i in range(loop_count + 1):
-    num_threads = omp.get_num_threads()
-    sum += i
-  known_sum = (loop_count *
-              (loop_count + 1)) / 2
-  return known_sum == sum and num_threads == 1
-    \end{lstlisting}
-\end{figure}
-
-We have used the Pythran tool described in the following section to turn each
-Python test function into a C++ function with the same directives and runtime
-calls. Apart from the \texttt{threadprivate} directive and the
-\texttt{collapse(\emph{n})} clause, all tests were successful.
-\texttt{threadprivate} directives were held by global variables not supported in
-Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
-loop nesting required by the \texttt{collapse} clause.
-
-%=========================================================
 \section{Static Translation of Python Programs}\label{sec:python-static}
@@
 file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
-node distance=0.6cm},
+node distance=0.7cm},
     tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
-distance=0.6cm}]
+distance=0.7cm}]
 \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
-\node[file] (annotation)	   [right=of python] {\textbf{Type Info}};
-\node[file] (omp)	   [left=of python] {\textbf{OpenMP}};
+\node[file] (omp)	   [xshift=-2em, left=of python] {\textbf{OpenMP}};
 \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
-\node[file] (cxx) [below=of pythran] {\textbf{C++}};
-\node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
-\node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
-\node[tool] (gxx) [below=of cxx] {\textbf{g++}};
-\node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
+\node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
+\node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
+\node (empty) [xshift=-1em, left=of gxx] {};
+\node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic++}}};
+\node[file] (annotation)	   [above=of pythonic] {\textbf{Type Info}};
+\node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
+\node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
 
 \draw[very thick, ->] (python) -- (pythran);
-\draw[very thick,dashed, ->] (annotation) -- (python);
+\draw[very thick, ->] (annotation) -| (empty);
 \draw[very thick,dashed, ->] (omp) -- (python);
-\draw[very thick, ->] (pythran) -- (cxx);
-\draw[very thick, ->] (cxx) -- (gxx);
-\draw[very thick, ->] (boost) -- (gxx);
+\draw[very thick, ->] (pythran) -- (meta-cxx);
+\draw[very thick, ->] (meta-cxx) -- (gxx);
+\draw[very thick, ->] (boost) -| (empty);
 \draw[very thick, ->] (pythonic) -- (gxx);
@@
 \end{tikzpicture}
+
 \caption{Pythran compiler workflow.}
@@
 %=========================================================
+\section{OpenMP Semantical Adaptations}\label{sec:python-openmp}
+%=========================================================
+
+OpenMP is a standard API for parallel programming for Fortran, C and C++. It
+consists of a set of parallelizing directives and a few runtime library calls.
+If OpenMP is not activated, the directives are ignored, thus enabling
+incremental parallelization of the original source code while keeping the
+original code structure. The languages targeted by OpenMP are statically
+compiled. This section studies the semantical adaptation required to use the
+same API for a scripting language such as Python.
+Listing~\ref{lst:motivating-example} introduces a motivating example through the
+computation of $\pi$, featuring a parallel reduction. The C equivalent with
+OpenMP directives is given in Listing~\ref{lst:motivating-example-c} for
+reference.
+
+\begin{lstlisting}[float, label={lst:motivating-example}, caption={Motivating
+example: computing $\pi$ in Python.}]
+def pi(nsteps):
+  sum, step = 0., 1. / nsteps
+  for i in range(nsteps):
+      x = (i - 0.5) * step
+      sum += 4. / (1. + x**2)
+  return step * sum
+\end{lstlisting}
+
+\begin{lstlisting}[float, label={lst:motivating-example-c}, caption={Motivating
+example: computing $\pi$ in C with OpenMP.}]
+double pi(size_t nsteps) {
+  double sum = 0., step = 1. / nsteps;
+  #pragma omp parallel for reduction(+:sum)
+  for(size_t i = 0; i < nsteps ; ++i)
+  {
+      double x = (i - 0.5) * step;
+      sum += 4. / (1. + x * x);
+  }
+  return step * sum;
+}
+\end{lstlisting}
+
+\subsection{Directives}
+
+OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them
+apply to structured blocks, represented by an instruction in C/C++ and delimited
+by comments in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
+attached to a specific instruction.
+
+While Python has a decorator mechanism\footnote{\see PEP 318,
+\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
+Python decorators.}, it only applies to functions, methods and classes and does
+not make it possible to attach decorations to other statements.
+PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
+Python does not have anonymous block,~\footnote{\see PEP340,
+    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
+anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
+instruction to apply an annotation to a whole block. Pythran uses comment to
+hold OpenMP directives, and internally stores them as string instructions.
+Alternatively, the syntax \texttt{if 'my annotation':} is also supported.
+
+Many OpenMP annotations are parametrized by clauses that list variables,
+specifying their behavior with respect to parallel regions, e.g.\
+\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
+variables that have already been declared. However, there is no variable
+declaration in Python, and all variables assigned in a function have the
+function scope. Consequently, all variables that are referenced in a function
+are considered when building such variable lists: there is no concept of
+variable local to a block. Using the \texttt{default(none)} clause is possible
+to ensure no variable gets forgotten when building such lists.
+
+The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
+characterize some data dependencies when performing a parallel reduction. The
+list of supported operators depend on the input and backend languages: Python
+has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
+\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
+the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
+
+The annotated motivating example is given in
+Listing~\ref{lst:motivating-example-annotated-v0}. Note that unlike in C,
+\texttt{x} has to be listed in a \texttt{private} clause.
+
+\begin{lstlisting}[float, label={lst:motivating-example-annotated-v0}, caption={Motivating
+example: computing $\pi$ in Python with OpenMP.}]
+def pi(nsteps):
+  sum, step = 0., 1. / nsteps
+#omp parallel for reduction(+:sum) private(x)
+  for i in range(nsteps):
+      x = (i - 0.5) * step
+      sum += 4. / (1. + x**2)
+  return step * sum
+\end{lstlisting}
+
+\subsection{Automatic Scoping Computation}
+
+Using function scope for all variables may lead to very long \texttt{private}
+clauses. We propose an algorithm to automatically limit the scope of a variable
+declaration so that it is can be automatically marked as private. The whole idea
+is to compute the minimal nesting depth of a variable usage. To do so, a
+\emph{nesting depth} is first associated to each expression. It lexically
+corresponds to the indentation level of its instruction, to the notable
+exception of the iterator declaration in a for loop, that holds a
+nesting depth equals to 1 plus the loop declaration indentation level.
+
+Tracking the nesting depth of all variable use, for instance using def-use
+chains, makes it possible to build a list of nesting depths, one per variable
+reference. Computing the min of this list yields the scope of the variable.
+
+For instance, in the motivating example, the \texttt{nsteps} variable is first
+defined as argument, then read twice, which gives a def-use chain of
+\{DEF, USE, USE\}. The associated nesting depths is
+\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator then
+used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
+depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
+of \{D, U\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
+automatically declared local to the loop, thus not needing to be listed as
+\texttt{private}.
+
+To be able to list a variable marked as local, for instance in a
+\texttt{lastprivate} clause, the directive accesses are taken into account in
+the def-use computations. For instance in
+Listing~\ref{lst:motivating-example-annotated-v0}, the chain for \texttt{x} is
+\{DEF, DEF, USE\} and the associated nested depths are \{1, 2, 2\} so its scope
+is 1 and the variable is not declared local to the loop.
+
+Thanks to this automatic scoping computation, it is possible to write shorted
+OpenMP directive in a similar manner to C, as illustrated in
+Listing~\ref{lst:motivating-example-annotated-v1}
+
+\begin{lstlisting}[float, label={lst:motivating-example-annotated-v1}, caption={Motivating
+example: computing $\pi$ in Python with OpenMP and automatic scoping.}]
+def pi(nsteps):
+  sum, step = 0., 1. / nsteps
+  #omp parallel for reduction(+:sum)
+  for i in range(nsteps):
+    x = (i - 0.5) * step
+    sum += 4. / (1. + x**2)
+  return step * sum
+\end{lstlisting}
+
+\subsubsection{Parallel For and Iterators}
+
+The core directive to handle data parallelism is the \texttt{for} directive that
+distributes the iteration space of the associated loop among the existing
+threads. To be compatible with OpenMP, the loop iteration space must be
+described by a random access iterator with a total order. Integers used as loop
+indices in Fortran and C satisfy this conditions, as well as C++ iterators with
+the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
+step of one until it is exhausted, in which case it raises an exception: it
+behaves as a forward iterator. To be compatible with OpenMP, these iterators
+must be turned into random access iterators.
+
+The extension of Python iterator to random access iterators is direct for the
+standard containers: \texttt{list}\footnote{Python \texttt{list}s behave as C++
+\texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
+care.
+
+Generators, Python objects that behave like iterators, are commonly used in
+Python. The simplest one, \texttt{xrange(\emph{start}, \emph{stop},
+\emph{step})}, successively yields value starting from \texttt{\emph{start}} to
+\texttt{\emph{stop}} by a step of \texttt{\emph{step}}. It is easily extended to
+support random access, but it generally does not make sense to use a generator
+as a loop iterator, as the relation between two random states of the iterator
+may be of an arbitrary complexity.
+
+Generator expressions are generators whose content is built from an other
+iterator. For instance \texttt{(x*x for x in l)} successively yields the square
+of each element in \texttt{l}. They behave like adaptors: they apply a
+particular expression on each value of the iterator. It is also the case of the
+\texttt{enumerate} builtin, that yields each element of the enumerated iterator
+associated with its index. These generators are random access iterators only if their
+input iterator is a random iterator itself.
+
+Finally, if the iterator yields a tuple, it is possible to unpack it inside the
+\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
+all the unpacked variables are considered as iterators, especially with respect
+to default privatization rules: in the given example, \texttt{i} and \texttt{v}
+are private, and the parallel iteration is valid because the input of
+\texttt{enumerate} is a list, which allows random access iteration.
+
+\begin{lstlisting}[float, label={lst:complex-omp-for}, caption={Parallel
+    loop in Pythran with tuple unpacking.}]
+'omp parallel for'
+for i, v in enumerate([2, 3, 5, 7, 11]):
+    print i, ':', v
+\end{lstlisting}
+
+Pythran's runtime is aware of these three kind of iterators and supports
+parallel iteration over random access iterators and iterators adapting random
+ones.
+
+\subsection{OpenMP Runtime Library}
+
+OpenMP provides a small runtime library that, for instance, makes it possible to
+retrieve the active thread id. All the functions are declared in the
+\texttt{<omp.h>} header, and have a default behavior when OpenMP is not
+activated.
+
+Providing a binding to these library in Python as an \texttt{omp} module does
+not raise particular problems, as the signature of these functions only involves
+integers passed as parameters or return values, except for the mutex
+manipulation. In that case an opaque type is used to represent the native type.
+
+The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
+activated, and can be used to detect when OpenMP is not available. Python does
+not have a preprocessor, but it is possible to catch the import exception if the
+\texttt{omp} module is not found. Listing~\ref{lst:sample-api-usage} showcase
+such a call using an example converted from the OpenMP validation suite
+presented in next section. The code starts a parallel section using the
+\texttt{parallel} directive, spawning several threads. Then each started thread
+increases the \texttt{nthreads} variable that has unspecified visibility, thus
+\texttt{shared}. The incrementation is protected by a guard through the
+\texttt{critical} directive. Then one of the thread retrieves the number of
+active thread in the parallel region through the \texttt{get\_num\_threads}
+function from the \texttt{omp} module. The function should always return
+\texttt{True}.
+
+\begin{lstlisting}[float,label={lst:sample-api-usage},caption={Example of
+        OpenMP API usage from Python.}]
+import omp
+def omp_get_num_threads():
+  nthreads, nthreads_lib = 0, -1
+  #omp parallel
+  if 1
+    #omp critical
+    nthreads += 1
+    #omp single
+    nthreads_lib = omp.get_num_threads()
+  return nthreads == nthreads_lib
+\end{lstlisting}
+
+\subsubsection{Interaction With Pythran Runtime}
+
+Pythran already uses OpenMP to parallelize some Python functions. For instance
+the \texttt{sum} function from the \texttt{\_\_builtin\_\_} is implemented using
+the OpenMP \texttt{reduction} clause. When it can prove that its first argument
+is a \emph{pure} function, it can also call a parallel version of the
+\texttt{map} function that internally spawns a parallel region. This leads to a
+well known problem with OpenMP: nested parallel regions. OpenMP provides several
+ways to handle this situation:
+
+\begin{enumerate}
+
+    \item A
+
+    \item B
+
+    \item C
+
+\end{enumerate}
+
+\subsection{Validation}
+
+A validation suite for OpenMP is proposed in~\cite{wang2012} for C and Fortran.
+We ported it to Python, and also extended it to validate the corner cases
+specific to Python described in this section. A typical test case is given in
+Listing~\ref{lst:openmp-validation}.
+
+\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of Python
+    OpenMP validation test case.}]
+def omp_parallel_for_if(loop_count):
+  import omp
+  using = sum = 0
+  'omp parallel for if(using == 1)'
+  for i in range(loop_count + 1):
+    num_threads = omp.get_num_threads()
+    sum += i
+  known_sum = (loop_count *
+              (loop_count + 1)) / 2
+  return known_sum == sum and num_threads == 1
+\end{lstlisting}
+
+We have used the Pythran tool described in the following section to turn each
+Python test function into a C++ function with the same directives and runtime
+calls. Apart from the \texttt{threadprivate} directive and the
+\texttt{collapse(\emph{n})} clause, all tests were successful.
+\texttt{threadprivate} directives were held by global variables not supported in
+Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
+loop nesting required by the \texttt{collapse} clause.
+
+%=========================================================
 \section{Validation}\label{sec:validation}
@@
 
-To validate the approach proposed in this paper, we first performed a simple
-experiment. Starting from the C code of a small geomatics application, Hyantes,
-we successively parallelized it with OpenMP, turned it into Python with a
-Pythran compatible kernel and parallelized the Python version using the approach
-described in this paper. We also measure the SLOC of the two versions.
-Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
-laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It not
-only emphasises the use of a high level language prototyping, but also shows
-that it is possible to turn this prototype into reasonably efficient code
-through Pythran. It is also possible to prototype the parallel version while
-remaining at the Python level.
+In addition to the formal validation done using the OpenMP validation suite
+presented in Section~\ref{sec:python-openmp}, the performance of parallel module
+generated by Pythran from Python module with explicit parallelization using
+OpenMP directives has been studied on several situations: first on the
+motivating example, then on a synthetic geomatic application. Python and Cython
+versions are then compared and finally the effect of OpenMP directives on Pythran
+generated code are illustrated in the context of the Python Benchmarks
+suite~\footnote{\url{https://github.com/numfocus/python-benchmarks}}.
+
+\subsection{Motivating Example}
+
+To validate the approach proposed in this paper, let first consider the
+motivating example. Figure~\ref{fig:motivating-example-scale} shows how the C
+and Python version of the program scale when changing the number of core used.
+As a reference, the original Python code runs 41 times slower than the
+equivalent C code. As the program is not memory bound and exhibit trivial
+parallelism, the C + OpenMP version scales linearly. The Pythran + OpenMP
+version follows the same pattern.
+
+\begin{figure}
+
+    \caption{Comparison of the scaling of $\pi$ computation function for C and
+    Python with OpenMP, depending on the number of active threads.}
+    \label{fig:motivating-example-scale}
+
+    \includegraphics[width=.5\textwidth]{pi_omp_bench.pdf}
+
+\end{figure}
+
+\subsection{Hyantes: a Geomatic Application}
+
+To get closer to a realistic example, we have considered the code of a small
+geomatics application, Hyantes. Starting from the C code, we successively
+parallelized it with OpenMP, turned it into Python with a Pythran compatible
+kernel and parallelized the Python version using the approach described in this
+paper. We also measure the SLOC of the two versions.  Table~\ref{tbl:hyantes}
+summarizes the results of the experimentation made on a laptop with 4 non
+hyperthreaded cores using \texttt{-O2} compiler flag. It not only emphasises the
+use of a high level language prototyping, but also shows that it is possible to
+turn this prototype into reasonably efficient code through Pythran. It is also
+possible to prototype the parallel version while remaining at the Python level.
 
@@
     \centering
-    \begin{tabular}{|l|l|l|l|}
+    \begin{tabular}{|l|r|}
         \hline
-        Language & SLOC & wall time (s) [gcc] & wall time (s) [clang]\\
+        Language & SLOC  \\
         \hline
-        C       & 102   & 25 & 22 \\
-        Python  & 30    & 676 & --\\
-        Pythran & 30    & 38 &  22 \\
-        Pythran+OMP    & 31    & 11 & no OpenMP support\\
+        C       & 102   \\
+        Python  & 30    \\
+        Pythran & 30     \\
+        Pythran+OMP    & 31  \\
         \hline
@@
 
-\begin{figure}[ht]
-    \caption{Relative speedups of Pythran+OpenMP version of the Hyantes prototype.}
+\begin{figure}
+
+    \caption{Comparison of the scaling of the hyantes application for C and
+    Python with OpenMP, depending on the number of active threads.}
     \label{fig:hyantes-speedup}
-    \centering
-    \includegraphics[width=.5\textwidth]{hyantes}
+
+    \includegraphics[width=.5\textwidth]{hyantes_omp_bench.pdf}
+
 \end{figure}
+
+\subsection{Comparison with Cython}
 
@@
 execution time when called from the Python interpreter is measured through the
-\texttt{timeit} module on a machine using 4 AMD Opteron 6176 SE.  All results
+\texttt{timeit} module on a machine using 8 AMD Opteron 6176 SE.  All results
 are normalized against Cython sequential execution time. They show that while
@@
 \end{figure}
+
+\subsection{Python Benchmarks Test Bed}
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{growcut}
+    \caption{Comparison of Various python Compilers on the Growcut benchmark}
+    \label{fig:pb-growcut}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{julia}
+    \caption{Comparison of Various python Compilers on the Julia benchmark}
+    \label{fig:pb-julia}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{rosen_der}
+    \caption{Comparison of Various python Compilers on the rosen\_der benchmark}
+    \label{fig:pb-rosen}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{pairwise}
+    \caption{Comparison of Various python Compilers on the pairwise benchmark}
+    \label{fig:pb-pairwise}
+\end{figure}
+growcut: $\times 2.65$
+julia: $\times 3.95$
+rosen\_der: $\times 1.78$
+pairwise: $\times 4.27$
 
@@
 %=========================================================
-\acks
+\section*{Acknowledgment}
 %=========================================================
 
-This work has received fundings from \textsc{silkan} and the French ANR through the
-CARP project. The authors thank Adrien \textsc{Merlini} and Alan
+This work has received fundings from \textsc{silkan} and the French ANR through
+the CARP project. The authors thank Adrien \textsc{Merlini} and Alan
 \textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
-B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Mehdi \textsc{Amini},
-Fran{\,c}ois \textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable
-reviews.
+B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Fran{\,c}ois
+\textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable reviews.
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{IEEEtranS}
 \bibliography{biblio}

### 2013-09-01 doc/papers/sc2013/bench/cython/convolution_para.pyx

@@ -1,0 +1,31 @@
+import numpy as np
+cimport numpy as np
+cimport cython
+from cython.parallel import parallel, prange
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+
+@cython.boundscheck(False)
+def convolution(np.ndarray[DTYPE_t, ndim=2, negative_indices=False] A):
+    cdef int m = A.shape[0]
+    cdef int n = A.shape[1]
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] B = np.zeros((m,n))
+
+    cdef double c11 = 2.0
+    cdef double c21 = 5.0
+    cdef double c31 = -8.0
+    cdef double c12 = -3.0
+    cdef double c22 = 6.0
+    cdef double c32 = -9.0
+    cdef double c13 = 4.0
+    cdef double c23 = 7.0
+    cdef double c33 = 10.0
+    cdef int i,j
+    with nogil, parallel():
+        for i in prange(1, m - 1):
+            for j in xrange(1, n - 1):
+              B[i,j] = c11 * A[i - 1,j - 1] + c12 * A[i + 0,j - 1] + c13 * A[i + 1,j - 1]\
+              + c21 * A[i - 1,j + 0] + c22 * A[i + 0,j + 0] + c23 * A[i + 1,j + 0]\
+              + c31 * A[i - 1,j + 1] + c32 * A[i + 0,j + 1] + c33 * A[i + 1,j + 1]
+    return B 

### 2013-09-01 doc/papers/sc2013/bench/cython/matmul_cython.pyx

@@ -1,0 +1,20 @@
+import numpy as np
+cimport numpy as np
+cimport cython
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+
+@cython.boundscheck(False)
+def matrix_multiply(np.ndarray[DTYPE_t, ndim=2, negative_indices=False] m0, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] m1):
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] new_matrix = np.zeros((m0.shape[0],m1.shape[1]))
+    cdef int i, j, k
+    cdef double r
+    "omp parallel for private(i,j,k,r)"
+    for i in xrange(m0.shape[0]):
+        for j in xrange(m1.shape[1]):
+            r=0
+            for k in xrange(m1.shape[0]):
+                r += m0[i,k]*m1[k,j]
+            new_matrix[i,j]=r
+    return new_matrix

### 2013-09-01 doc/papers/sc2013/bench/cython/setup.py

@@ -1,0 +1,23 @@
+from distutils.core import setup
+from distutils.extension import Extension
+from Cython.Distutils import build_ext
+
+setup(
+        cmdclass = {'build_ext': build_ext},
+        ext_modules = [
+            Extension("cconvolution", ["convolution.pyx"]),
+            Extension("cconvolution_omp",   ["convolution_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            Extension("cfbcorr", ["fbcorr_cython.pyx"]),
+            Extension("cfbcorr_omp",   ["fbcorr_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            Extension("cmandel", ["mandel_cython.pyx"]),
+            Extension("cmandel_omp",   ["mandel_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            Extension("cmatmul", ["matmul_cython.pyx"]),
+            Extension("cmatmul_omp",   ["matmul_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            Extension("cpairwise", ["pairwise.pyx"]),
+            Extension("cpairwise_omp", ["pairwise_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            Extension("cwave_", ["wave_.pyx"]),
+            Extension("cwave__omp", ["wave_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            Extension("chyantes", ["hyantes.pyx"]),
+            Extension("chyantes_omp", ["hyantes_para.pyx"], extra_compile_args=['-fopenmp'], extra_link_args=['-fopenmp']),
+            ]
+     )

### 2013-09-01 doc/papers/sc2013/bench/pythran/convolution.py

@@ -1,0 +1,24 @@
+import numpy
+#runas import numpy ; a = numpy.ones((5000,5000)) ; convolution(a)
+#pythran export convolution(float[][])
+def convolution(A):
+    m = A.shape[0]
+    n = A.shape[1]
+    B = numpy.zeros((m,n))
+
+    c11 = 2.0
+    c21 = 5.0
+    c31 = -8.0
+    c12 = -3.0
+    c22 = 6.0
+    c32 = -9.0
+    c13 = 4.0
+    c23 = 7.0
+    c33 = 10.0
+    "omp parallel for private(j,i)"
+    for i in xrange(1, m - 1):
+        for j in xrange(1, n - 1):
+          B[i,j] = c11 * A[i - 1,j - 1] + c12 * A[i + 0,j - 1] + c13 * A[i + 1,j - 1]\
+          + c21 * A[i - 1,j + 0] + c22 * A[i + 0,j + 0] + c23 * A[i + 1,j + 0]\
+          + c31 * A[i - 1,j + 1] + c32 * A[i + 0,j + 1] + c33 * A[i + 1,j + 1]
+    return B 

### 2013-09-01 doc/papers/sc2013/bench/pythran/fbcorr.py

@@ -1,0 +1,21 @@
+# from https://github.com/numba/numba/blob/master/examples/fbcorr.py
+#pythran export fbcorr(float [] [] [] [], float [] [] [] [])
+#runas from random import random ; imgs = [ [ [ [ random() for i in xrange(3) ] for j in xrange(32) ] for j in xrange(32) ] for k in xrange(32) ] ; filt = [ [ [ [ random() for i in xrange(3) ] for j in xrange(5) ] for j in xrange(5) ] for k in xrange(6) ] ; import numpy as np ; imgs = np.array(imgs) ; filt = np.array(filt) ; fbcorr(imgs, filt)
+
+import numpy
+def fbcorr(imgs, filters):
+    n_imgs, n_rows, n_cols, n_channels = imgs.shape
+    n_filters, height, width, n_ch2 = filters.shape
+    output = numpy.zeros((n_imgs, n_filters, (n_rows - height + 1),(n_cols - width + 1)))
+    "omp parallel for private(rr,cc,hh,ww,jj,ff,imgval, filterval)"
+    for ii in xrange(n_imgs):
+        for rr in xrange(n_rows - height + 1):
+            for cc in xrange(n_cols - width + 1):
+                for hh in xrange(height):
+                    for ww in xrange(width):
+                        for jj in xrange(n_channels):
+                            for ff in xrange(n_filters):
+                                imgval = imgs[ii,rr + hh,cc + ww,jj]
+                                filterval = filters[ff,hh,ww,jj]
+                                output[ii,ff,rr,cc] += imgval * filterval
+    return output

### 2013-09-01 doc/papers/sc2013/bench/pythran/parsetab.py

@@ -1,0 +1,49 @@
+
+# parsetab.py
+# This file is automatically generated. Do not edit.
+_tabversion = '3.2'
+
+_lr_method = 'LALR'
+
+_lr_signature = 'Li5-\x8b+j.\xdf\x84\x1c\x1e&_\x05c'
+    
+_lr_action_items = {'FLOAT':([9,10,11,25,27,],[20,20,20,20,20,]),'RPAREN':([9,10,12,13,14,15,16,17,18,19,20,21,22,23,24,26,31,33,34,35,],[-5,-5,-17,-19,-7,-15,-16,-6,-9,29,-20,-18,30,31,-11,-10,-14,-8,-12,-13,]),'STR':([9,10,11,25,27,],[15,15,15,15,15,]),'INT':([9,10,11,25,27,],[21,21,21,21,21,]),'PYTHRAN':([3,],[5,]),'LIST':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,26,-15,-16,-9,-20,-18,-11,-10,-14,26,-12,-13,]),'RARRAY':([28,],[34,]),'LONG':([9,10,11,25,27,],[13,13,13,13,13,]),'COLUMN':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,25,-15,-16,-9,-20,-18,-11,-10,-14,25,-12,-13,]),'LARRAY':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,28,-15,-16,-9,-20,-18,-11,-10,-14,28,-12,-13,]),'COMPLEX':([9,10,11,25,27,],[12,12,12,12,12,]),'EXPORT':([5,6,],[6,7,]),'BOOL':([9,10,11,25,27,],[16,16,16,16,16,]),'LPAREN':([7,8,9,10,11,25,27,],[9,10,11,11,11,11,11,]),'SET':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,24,-15,-16,-9,-20,-18,-11,-10,-14,24,-12,-13,]),'SHARP':([0,2,29,30,],[3,3,-4,-3,]),'IDENTIFIER':([6,],[8,]),'DICT':([12,13,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,-15,-16,-9,-20,-18,-11,-10,-14,35,-12,-13,]),'COMMA':([12,13,14,15,16,
[…]

### 2013-09-01 doc/papers/sc2013/benchmark.sh

@@ -1,0 +1,120 @@
+#!/bin/bash
+NBITER=1000000
+NBREP=5
+NBPROCS=8
+CXX=g++ #clang++
+CC=gcc
+LIBOMP=gomp #iomp5
+
+# select custom clang version with openMP support
+export PATH=$HOME/tmp/llvm/bin/:$PATH
+export LD_LIBRARY_PATH=$HOME/sources/libomp_oss/exports/lin_32e/lib
+export CPLUS_INCLUDE_PATH=$HOME/sources/libomp_oss/exports/common/include
+
+##############################
+# PI
+##############################
+
+# cleanup
+rm -f *.so pi
+
+# python part
+python -m timeit -n 1 -r $NBREP -s "import pi" "pi.pi($NBITER)" | sed -e 's/^.*: //' -e 's/ per loop//' -e 's/ msec/\/1000/' | bc -l > pi.python.seq
+python ../../../scripts/pythran pi.py
+python -m timeit -n 1 -r $NBREP -s "import pi" "pi.pi($NBITER)" | sed -e 's/^.*: //' -e 's/ per loop//' -e 's/ msec/\/1000/' -e 's/ usec/\/1000000/' -e 's/ sec//' | bc -l > pi.pythran.seq
+
+python ../../../scripts/pythran pi.py -fopenmp 
+for i in `seq 1 $NBPROCS` ; do
+    echo -n "$i "
+    OMP_NUM_THREADS=$i python -m timeit -n 1 -r $NBREP -s "import pi" "pi.pi($NBITER)" | sed -e 's/^.*: //' -e 's/ per loop//' -e 's/ msec/\/1000/' -e 's/ usec/\/1000000/' -e 's/ sec//' | bc -l
+done > pi.pythran.omp
+
+# c part
+$CXX -l$LIBOMP pi.cc -o pi -O2
+for i in `seq 1 $NBREP` ; do
+    ./pi $NBITER
+done | numbound -l > pi.c.seq
+
+$CXX -fopenmp pi.cc -o pi -O2
+for i in `seq 1 $NBPROCS` ; do
+    echo -n "$i "
+    for j in `seq 1 $NBREP` ; do
+        OMP_NUM_THREADS=$i ./pi $NBITER
+    done | numbound -l
+done > pi.c.omp
+
+#plotting
+gnuplot << EOF
+set terminal pdf
+set output "pi_omp_bench.pdf"
+set xlabel "Number of Active Threads"
+set ylabel "Elapsed Time(s)"
[…]

### 2013-09-01 doc/papers/sc2013/hyantes.c

@@ -1,0 +1,123 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <omp.h>
+
+#ifdef USE_FLOAT
+typedef float data_t;
+#define M_PI 3.141592654f
+#define INPUT_FORMAT "%f%*[ \t]%f%*[ \t]%f"
+#define OUTPUT_FORMAT "%f %f %f\n"
+#else
+typedef double data_t;
+#define M_PI 3.141592654
+#define INPUT_FORMAT "%lf%*[ \t]%lf%*[ \t]%lf"
+#define OUTPUT_FORMAT "%lf %lf %lf\n"
+#endif
+
+typedef data_t town[3];
+typedef struct {
+    size_t nb;
+    town *data;
+} towns;
+
+towns read_towns(const char fname[])
+{
+    FILE * fd = fopen(fname,"r");
+    size_t curr=0;
+    char c;
+    towns the_towns = { 1 , malloc(sizeof(town)) };
+
+    while(!feof(fd))
+    {
+        if(the_towns.nb==curr)
+        {
+            the_towns.nb*=2;
+            the_towns.data=realloc(the_towns.data,the_towns.nb*sizeof(town));
+        }
+        if(fscanf(fd,INPUT_FORMAT,&the_towns.data[curr][0],&the_towns.data[curr][1],&the_towns.data[curr][2]) !=3 )
+        {
+            while(!feof(fd))
+            {
+                c=(char)fgetc(fd);
+                if(c=='\n' || c=='\r') break;
+            }
+        }
+        else
+        {
+            the_towns.data[curr][0]*=M_PI/180;
+            the_towns.data[curr][1]*=M_PI/180;
+            ++curr;
+        }
+    }
+    fclose(fd);
+    the_towns.data=realloc(the_towns.data,curr*sizeof(town));
+    the_towns.nb=curr;
+    /*
+    for(curr=0;curr<the_towns.nb;curr++)
+        fprintf(stderr,OUTPUT_FORMAT,the_towns.data[curr][0],the_towns.data[curr][1],the_towns.data[curr][2]);
+    */
+    return the_towns;
+}
+
+
+towns run(towns t,data_t xmin, data_t ymin, data_t xmax, data_t ymax, data_t step,data_t range)
+{
+    size_t i,j;
+    size_t k;
+    size_t rangex=( (xmax - xmin )/step ),
+        rangey=( (ymax - ymin )/step );
[…]

### 2013-09-01 doc/papers/sc2013/hyantes_core.py

@@ -1,0 +1,14 @@
+#pythran export run(float, float, float, float, float, float, int, int, float [][])
+import math
+from numpy import zeros
+def run(xmin, ymin, xmax, ymax, step, range_, range_x, range_y, t):
+    pt = zeros((range_x, range_y, 3))
+    "omp parallel for private(i,j,k,tmp)"
+    for i in xrange(range_x):
+        for j in xrange(range_y):
+            pt[i,j,0], pt[i,j,1] = (xmin+step*i)*180/math.pi, (ymin+step*j)*180/math.pi
+            for k in xrange(t.shape[0]):
+                tmp = 6368.* math.acos( math.cos(xmin+step*i)*math.cos( t[k,0] ) * math.cos((ymin+step*j)-t[k,1])+  math.sin(xmin+step*i)*math.sin(t[k,0]))
+                if tmp < range_:
+                    pt[i,j,2]+= t[k,2] / (1+tmp)
+    return pt

### 2013-09-01 doc/papers/sc2013/parsetab.py

@@ -1,0 +1,49 @@
+
+# parsetab.py
+# This file is automatically generated. Do not edit.
+_tabversion = '3.2'
+
+_lr_method = 'LALR'
+
+_lr_signature = 'Li5-\x8b+j.\xdf\x84\x1c\x1e&_\x05c'
+    
+_lr_action_items = {'FLOAT':([9,10,11,25,27,],[20,20,20,20,20,]),'RPAREN':([9,10,12,13,14,15,16,17,18,19,20,21,22,23,24,26,31,33,34,35,],[-5,-5,-17,-19,-7,-15,-16,-6,-9,29,-20,-18,30,31,-11,-10,-14,-8,-12,-13,]),'STR':([9,10,11,25,27,],[15,15,15,15,15,]),'INT':([9,10,11,25,27,],[21,21,21,21,21,]),'PYTHRAN':([3,],[5,]),'LIST':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,26,-15,-16,-9,-20,-18,-11,-10,-14,26,-12,-13,]),'RARRAY':([28,],[34,]),'LONG':([9,10,11,25,27,],[13,13,13,13,13,]),'COLUMN':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,25,-15,-16,-9,-20,-18,-11,-10,-14,25,-12,-13,]),'LARRAY':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,28,-15,-16,-9,-20,-18,-11,-10,-14,28,-12,-13,]),'COMPLEX':([9,10,11,25,27,],[12,12,12,12,12,]),'EXPORT':([5,6,],[6,7,]),'BOOL':([9,10,11,25,27,],[16,16,16,16,16,]),'LPAREN':([7,8,9,10,11,25,27,],[9,10,11,11,11,11,11,]),'SET':([12,13,14,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,24,-15,-16,-9,-20,-18,-11,-10,-14,24,-12,-13,]),'SHARP':([0,2,29,30,],[3,3,-4,-3,]),'IDENTIFIER':([6,],[8,]),'DICT':([12,13,15,16,18,20,21,24,26,31,32,34,35,],[-17,-19,-15,-16,-9,-20,-18,-11,-10,-14,35,-12,-13,]),'COMMA':([12,13,14,15,16,
[…]

### 2013-09-01 doc/papers/sc2013/pi.py

@@ -1,0 +1,8 @@
+#pythran export pi(int)
+def pi(nsteps):
+  sum, step = 0., 1. / nsteps
+  "omp parallel for reduction(+:sum) private(x)"
+  for i in range(nsteps):
+      x = (i - 0.5) * step
+      sum += 4. / (1. + x**2)
+  return step * sum

### 2013-09-01 doc/papers/sc2013/bench/cython/hyantes.pyx
Add first draft for the pyhpc workshop paper
http://www.dlr.de/sc/en/desktopdefault.aspx/tabid-8686/14946_read-37159/

Deadline: September 6!

@@ -1,0 +1,25 @@
+#pythran export run(double, double, double, double, double, double, int, int, double [] [])
+#runas import numpy as np ; x = [ [i/10., i/10., i/20.] for i in xrange(400) ] ; x = np.array(x) ; run(0,0,90,90, 1, 0.0001, 400, 400, x)
+cimport libc.math as math
+import numpy as np
+cimport numpy as np
+cimport cython
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+
+@cython.boundscheck(False)
+def run(double xmin, double ymin, double xmax, double ymax, double step, double range_, int range_x, int range_y, np.ndarray[DTYPE_t, ndim=2, negative_indices=False] t):
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] pt = np.zeros((range_x, range_y))
+    cdef int i,j,k
+    cdef double tmp, xi, yj
+    "omp parallel for private(j,k,tmp)"
+    for i in xrange(range_x):
+        for j in xrange(range_y):
+            xi = xmin+step*i
+            yj = ymin+step*j
+            for k in xrange(t.shape[0]):
+                tmp = 6368.* math.acos( math.cos(xi)*math.cos( t[k,0] ) * math.cos(yj-t[k,1])+  math.sin(xi)*math.sin(t[k,0]))
+                if tmp < range_:
+                    pt[i,j]+=t[k,2] / (1+tmp)
+    return pt

### 2013-09-01 doc/papers/sc2013/bench/cython/mandel_para.pyx

@@ -1,0 +1,34 @@
+cimport cython
+cimport numpy as np
+import numpy as np
+from cython.parallel import parallel, prange
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+
+@cython.boundscheck(False)
+@cython.cdivision(True)
+def mandel(int size, int x_center, int y_center, int max_iteration):
+    cdef int i, j, iteration, color_value
+    cdef double x, y, a, b
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] out = np.zeros((size,size))
+    with nogil, parallel():
+        for i in prange(size):#, schedule="static", chunksize=5):
+            for j in xrange(size):
+                x = x_center + 4.0*float(i-size/2)/size
+                y = y_center + 4.0*float(j-size/2)/size
+
+                a = 0.0
+                b = 0.0
+                iteration = 0
+
+                while (a**2 + b**2 <= 4.0 and iteration < max_iteration):
+                    a= a**2 - b**2 + x
+                    b= 2*a*b + y
+                    iteration = iteration + 1
+                if iteration == max_iteration:
+                    color_value = 255
+                else:
+                    color_value = iteration*10 % 255
+                out[i,j]=color_value
+    return out

### 2013-09-01 doc/papers/sc2013/bench/cython/pairwise.c

@@ -1,0 +1,5124 @@
+/* Generated by Cython 0.15.1 on Mon Feb 11 15:34:07 2013 */
+
+#define PY_SSIZE_T_CLEAN
+#include "Python.h"
+#ifndef Py_PYTHON_H
+    #error Python headers needed to compile C extensions, please install development version of Python.
+#else
+
+#include <stddef.h> /* For offsetof */
+#ifndef offsetof
+#define offsetof(type, member) ( (size_t) & ((type*)0) -> member )
+#endif
+
+#if !defined(WIN32) && !defined(MS_WINDOWS)
+  #ifndef __stdcall
+    #define __stdcall
+  #endif
+  #ifndef __cdecl
+    #define __cdecl
+  #endif
+  #ifndef __fastcall
+    #define __fastcall
+  #endif
+#endif
+
+#ifndef DL_IMPORT
+  #define DL_IMPORT(t) t
+#endif
+#ifndef DL_EXPORT
+  #define DL_EXPORT(t) t
+#endif
+
+#ifndef PY_LONG_LONG
+  #define PY_LONG_LONG LONG_LONG
+#endif
+
+#if PY_VERSION_HEX < 0x02040000
+  #define METH_COEXIST 0
+  #define PyDict_CheckExact(op) (Py_TYPE(op) == &PyDict_Type)
+  #define PyDict_Contains(d,o)   PySequence_Contains(d,o)
+#endif
+
+#if PY_VERSION_HEX < 0x02050000
+  typedef int Py_ssize_t;
+  #define PY_SSIZE_T_MAX INT_MAX
+  #define PY_SSIZE_T_MIN INT_MIN
+  #define PY_FORMAT_SIZE_T ""
+  #define PyInt_FromSsize_t(z) PyInt_FromLong(z)
+  #define PyInt_AsSsize_t(o)   __Pyx_PyInt_AsInt(o)
+  #define PyNumber_Index(o)    PyNumber_Int(o)
+  #define PyIndex_Check(o)     PyNumber_Check(o)
+  #define PyErr_WarnEx(category, message, stacklevel) PyErr_Warn(category, message)
+#endif
+
+#if PY_VERSION_HEX < 0x02060000
+  #define Py_REFCNT(ob) (((PyObject*)(ob))->ob_refcnt)
+  #define Py_TYPE(ob)   (((PyObject*)(ob))->ob_type)
[…]

### 2013-09-01 doc/papers/sc2013/bench/cython/pairwise_para.pyx

@@ -1,0 +1,30 @@
+#from http://jakevdp.github.com/blog/2012/08/24/numba-vs-cython/
+#runas import random,numpy ; X = [ [random.random() for i in xrange(10000) ] for j in xrange(30) ] ; X = numpy.array(X) ; pairwise(X)
+#pythran export pairwise(double [] [])
+
+from libc.math cimport sqrt
+import numpy as np
+cimport numpy as np
+cimport cython
+from cython.parallel import parallel, prange
+
+DTYPE = np.double
+ctypedef np.double_t DTYPE_t
+
+@cython.boundscheck(False)
+def pairwise(np.ndarray[DTYPE_t, ndim=2, negative_indices=False] X):
+    cdef long i,j,k, M, N
+    cdef double d, tmp
+    M = X.shape[0]
+    N = X.shape[1]
+    cdef np.ndarray[DTYPE_t, ndim=2, negative_indices=False] D = np.zeros((M,N))
+    with nogil, parallel():
+        for i in prange(M):
+            for j in xrange(M):
+                d = 0.0
+                for k in xrange(N):
+                    tmp = X[i,k] - X[j,k]
+                    d = d + tmp * tmp
+                D[i,j] = sqrt(d)
+    return D
+

### 2013-09-01 doc/papers/sc2013/bench/pythran/wave_.py

@@ -1,0 +1,26 @@
+# from the paper `using cython to speedup numerical python programs'
+#pythran export wave(float, float, float, float, float, float [] [], float [] [], float [] [])
+#runas A=[map(float,range(2000)) for i in xrange(100)] ; B=[map(float,range(2000)) for i in xrange(100)] ; C=[map(float,range(2000)) for i in xrange(100)] ; import numpy ; A = numpy.array(A) ; B = numpy.array(B) ; C = numpy.array(C) ; wave(1,2,.01,.1,.18, A,B,C )
+def wave(t, t_stop, dt, dx, dy, u, um, k):
+    while t <= t_stop:
+        t += dt
+        new_u = calculate_u(dt, dx, dy, u, um, k)
+        um = u
+        u = new_u
+    return u
+import numpy
+
+def calculate_u(dt, dx, dy, u, um, k):
+    up = numpy.zeros((u.shape[0], u.shape[1]))
+    "omp parallel for private(i,j)"
+    for i in xrange(1, u.shape[0]-1):
+        for j in xrange(1, u.shape[1]-1):
+            up[i,j] = 2*u[i,j] - um[i,j] + \
+                (dt/dx)**2*(
+                        (0.5*(k[i+1,j] + k[i,j])*(u[i+1,j] - u[i,j]) -
+                            0.5*(k[i,j] + k[i-1,j])*(u[i,j] - u[i-1,j]))) + \
+                        (dt/dy)**2*(
+                                (0.5*(k[i,j+1] + k[i,j])*(u[i,j+1] - u[i,j]) -
+                                    0.5*(k[i,j] + k[i,j-1])*(u[i,j] - u[i,j-1])))
+    return up
+

### 2013-09-03 doc/papers/sc2013/paper.tex
Almost full revision of the paper
last benchmark part still needs to be written...

@@
 \author{
-    \IEEEauthorblockN{Serge Guelton}
-    \IEEEauthorblockA{
-        {\'E}cole Normale Sup{\'e}rieure, Paris, France\\
-        Email: \texttt{serge.guelton@telecom-bretagne.eu}\\
-    }
-    \and
-    \IEEEauthorblockN{Pierrick Brunet}
-    \IEEEauthorblockA{
-        T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France\\
-        Email: \texttt{pierrick.brunet@telecom-bretagne.eu}\\
-    }
-    \and
-    \IEEEauthorblockN{Mehdi Amini}
-    \IEEEauthorblockA{
-        Silkan, USA\\
-        Email: \texttt{mehdi.amini@silkan.com}
-    }
+    \IEEEauthorblockN{Serge Guelton\IEEEauthorrefmark{1}, \IEEEauthorrefmark{2}}
+    \IEEEauthorblockN{Pierrick Brunet\IEEEauthorrefmark{2}}
+    \IEEEauthorblockN{Mehdi Amini\IEEEauthorrefmark{3}}
+    \newline
+    \IEEEauthorblockA{\newline\IEEEauthorrefmark{1}
+        {\'E}cole Normale Sup{\'e}rieure, D\'{e}partement d'Informatique, Paris, France}
+    \IEEEauthorblockA{\newline\IEEEauthorrefmark{2}
+    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France}
+    \IEEEauthorblockA{\newline\IEEEauthorrefmark{3}
+    Silkan, Meudon la For\^et, France}
 }
@@
 computing~\cite{Oliphant2007}, mainly thanks to the SciPy~\cite{scipy} project
-that adds MATLAB-like functionalities to the language.
+that provides a set of scientific and numeric tools, ranging from linear algebra
+to ordinary differential equation solvers or generic algorithms.
 
@@
 language used to bind several native libraries together, or to write whole
-applications. However, the prohibitive performance overhead of the language
-implied by its interpreted and highly dynamic nature should prevent its usage
-for critical code sections, where native code is generally used. The SciPy
-modules overcomes this issue through low-level routines written in C or Fortran
-and encapsulated in Python \emph{native} modules.  Moreover, its core data
-structure, the multi-dimensional array~\cite{numpyarray2011}, has been designed
-so that the underlying data are available to both native modules and the Python
-interpreter without conversion cost.
+applications in Pure Python. However, the prohibitive performance overhead of
+the language implied by its interpreted and highly dynamic nature prevents its
+usage for the most performance - critical code sections, where native code is
+generally used. The SciPy modules partly overcome this issue through the usage
+of low-level routines written in C or Fortran and encapsulated in Python
+\emph{native} modules. Moreover, its core data structure, the multi-dimensional
+array~\cite{numpyarray2011}, has been designed so that the underlying data are
+available to both native modules and the Python interpreter without conversion
+cost. Yet when a new function not already part of an existing module is
+required, one has to write it in C or FORTRAN.
 
-To enable the user to write native functions, without having to write the glue
+To enable the user to write native functions without having to write the glue
 code that turns Python object into native structures back and forth, the SciPy
 package provides the \texttt{weave} module that makes it possible to bundle C
-code snippet into Python code and compile then load them at runtime.
+code snippet into Python code and compile then load them at runtime. This hybrid
+approach, based on a mix of interpreted and native code, is getting widespread
+in the Python landscape. It is also the recommended approach to write
+application that use fine grain parallelism, has Python parallelism support
+targets coarse-grain parallelism.
 
-This hybrid approach, based on a mix of interpreted and native code, is getting
-widespread in the Python landscape. Section~\ref{sec:python-parallel} gives an
-overview of existing approaches to compile Python functions or modules and shows
-a critical lack of parallelism support, and emphasises on the need for a
+This paper presents the addition of OpenMP, a popular standard to turn
+sequential C, C++ and FORTRAN application into parallel applications, to the
+Pythran~\cite{pythran2013} compiler, a compiler that turn Python modules into
+C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
+existing approaches to compile Python functions or modules and shows a critical
+lack of parallelism support, and emphasises on the need for a
 backward-compatible approach.  Section~\ref{sec:python-static} presents the
-Pythran~\cite{pythran2013} compiler, a tool that turns Python modules into C++
-meta-programs, and the underlying runtime. Section~\ref{sec:python-openmp}
+Pythran compiler and the underlying runtime. Section~\ref{sec:python-openmp}
 studies the validity of using OpenMP directives within the Python language in
 order to add fine-grain parallelism support to Python in the context of Pythran.
-The Pythran compiler is benchmarked on several scientific kernels and compared
-to Cython in Section~\ref{sec:validation}.
+The Pythran compiler is benchmarked on several scientific kernels, compared
+to Cython, Numba and Parakeet in Section~\ref{sec:validation}.
 
@@
 \texttt{multiprocessing} module. It spawns several interpreters that can
-communicate through IPC, using Python built-in object serialization. This
-approach is only viable for computation intensive application, since the
-communication and synchronization overheads are much greater than a light-weight
-processes approach.
+communicate through Inter Process Communication(IPC), using Python built-in
+object serialization. This approach is only viable for computation intensive
+application, since the communication and synchronization overheads are much
+greater than a light-weight processes approach.
 
-Although the standard \texttt{threading} module makes it possible to start
-several light-weight threads within the same interpreter, this approach is not
+The standard \texttt{threading} module also makes it possible to start
+several light-weight threads within the same interpreter, but this lower-level approach is not
 applicable to HPC, because of a specificity of CPython, the \emph{Global
@@
 
-To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
-estimate the value of $\pi$. Its execution time using a multi-threaded
-implementation on 4 cores is $\times 1.46$ the execution time of the sequential
-version. GIL contention actually \emph{increases} the execution time.
+To illustrate the limitations of these two approaches, we used the CPU-bound
+Buffon's needle algorithm to estimate the value of $\pi$. A sequential version
+parallelized using the two approaches is illustrated in
+Listing~\ref{lst:buffon-python}. When comparing their respective execution time
+using 4 cores, it appears that the version that uses the \texttt{threading}
+module runs $\times 1.46$ the execution time of the sequential version. GIL
+contention actually \emph{increases} the execution time. The version that uses
+\texttt{multiprocessing} provides a speedup of 3 over the sequential version
+while a speedup of 4 would be expected for this kind of application.
+
+\begin{lstlisting}[float, label={lst:buffon-python}, caption={Implementation of
+sequential and parallel version of the Buffon algorithm in Python.}]
+def buffon(darts):
+  hits = 0
+  for i in xrange (0, darts):
+    x, y = random(), random()
+    dist = sqrt(pow(x, 2) + pow(y, 2))
+    if dist <= 1.0:
+      hits += 1.0
+  # hits / throws = 1/4 Pi
+  pi = 4 * (hits / darts)
+  return pi
+
+from threading import Thread
+from multiprocessing import Queue
+def threaded_buffon(d):
+  def work(darts, queue):
+    queue.put(pi_estimate(darts))
+  n = 4
+  q = Queue()
+  threads = [
+    Thread(target=work, args=(d//n, q)),
+    Thread(target=work, args=(d//n, q)),
+    Thread(target=work, args=(d//n, q)),
+    Thread(target=work, args=(d//n, q)),
+  ]
+  map(Thread.start, threads)
+  map(Thread.join, threads)
+  return sum(q.get() for _ in threads)/n
+
+from multiprocessing import Pool
+def multi_buffon(darts):
+  p = Pool(4)
+  return sum(p.map(buffon, [darts//4]*4))/4
+\end{lstlisting}
 
@@
 Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
-mainstream interpreters.
-
-As a consequence, Python developers need to write multi-threaded native
-modules in order to fully benefit from multiple cores. This leads to a kind of
-computations referred as hybrid computations.
+mainstream interpreters.  As a consequence, Python developers need to write
+multi-threaded native modules in order to fully benefit from multiple cores.
+This leads to a kind of computations referred as \emph{hybrid computations}.
 
@@
 language is its type system. It implies that each method call is resolved
-dynamically, even a simple add operation! It comes at no surprise that many
+dynamically, even a simple add operation. It comes at no surprise that many
 approaches restrain the Python language to add a static typing overlay.  Also,
@@
 not suffer from the backward-incompatibility issues. When meeting an unsupported
-construct, numba falls back to Python C-API calls, while parakeet just raises an
+construct, numba falls back to Python C-API calls, while parakeet raises an
 exception.
@@
 Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
-turn regular Python codes into native ones, but they do not offer support for
-fine-grained parallelism beyond what the standard library proposes.
+enhance Python performance. However Shed~Skin does not provide support for
+fine-grained parallelism beyond what the standard library proposes. PyPy is
+heading toward STM for parallelism support.
 
@@
 solution. This principle is not respected by parallel libraries, but code
-annotations partially satisfy it. This is where OpenMP~\cite{openmp3.1} parallel
-annotations shine: the original code remains mostly compatible with a compiler
-that is not aware of OpenMP.
+annotations partially satisfy it. This is where parallel annotations shine: the
+original code remains mostly compatible with a compiler that is not aware of the
+annotations, while taking the annotation into account turns the sequential code
+into parallel code.
 
 This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
-with a Python subset called Pythran to bring fine-grain parallelism to Python,
-While being backward compatible with both the Python language, and the
-sequential algorithm. It means that:
-%
+with the Python subset supported by the Pythran compiler to make it possible to
+write fine-grain parallel applications in Python, while being fully backward
+compatible with both the Python language and the sequential algorithm. It means
+that:
+
 \begin{enumerate}
+
     \item any Pythran code can be run (sequentially), with no module dependency
         or code change, by any Python interpreter;
+
     \item parallelism is explicit and incrementally added to the original code
         through directives.
+
 \end{enumerate}
@@
 
-Pythran is a subset of Python designed for scientific computing. It is
-implicitly statically typed and supports most Python constructs except those
-that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
-(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
-core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
-is currently only partially supported. User classes are not supported. The
-associated compiler turns Pythran code, possibly annotated with OpenMP and a few
-function type annotations, into C++ code. Its processing is summarized in
-Figure~\ref{fig:pythran-compiler} and relies on type inference to remove most
-dynamic behavior, combined with a high-level runtime library,
+Pythran~\cite{pythran2013} is a subset of Python designed for scientific
+computing. It is implicitly statically typed and supports most Python constructs
+except those that involve introspection (e.g.\ \texttt{getattr}) or runtime
+compilation (e.g.\ \texttt{eval}). A few standard modules are supported in
+addition to the core language (e.g.\ \texttt{math}, \texttt{random}). The
+\texttt{numpy} module is currently only partially supported. User classes are
+not supported. The associated compiler turns Pythran code, possibly annotated
+with OpenMP and a few function type annotations, into C++ code. Its processing
+is summarized in Figure~\ref{fig:pythran-compiler} and relies on type inference
+to remove most dynamic behavior, combined with a high-level runtime library,
 \texttt{pythonic++}, to allow a one-to-one mapping between Python and C++
@@
 \draw[very thick, ->] (python) -- (pythran);
-\draw[very thick, ->] (annotation) -| (empty);
+\draw[very thick] (annotation) -| (empty.center);
 \draw[very thick,dashed, ->] (omp) -- (python);
@@
 \draw[very thick, ->] (meta-cxx) -- (gxx);
-\draw[very thick, ->] (boost) -| (empty);
+\draw[very thick] (boost) -| (empty.center);
 \draw[very thick, ->] (pythonic) -- (gxx);
@@
 passing still implies copies. To avoid a reference increment, one can pass
-parameters by reference. Using an out-of-scope interprocedural memory effect
-analysis, Pythran determines for each argument whether it must be passed using
-reference or const reference to prevent this overhead.
+parameters by reference. Using an inter-procedural memory effect analysis not
+presented in the paper, Pythran determines for each argument whether it must be
+passed using reference or const reference to prevent this overhead.
 
@@
 context-insensitive part ---the clauses---, and attaches them to the proper
-instruction in the AST.
+instruction in the Abstract Syntax tree(AST).
 
@@
 at the Python level. As a tool for scientific computing, Pythran supports such a
-structure.
+structure, implemented as a class wrapping a raw pointer.
 
@@
 
-OpenMP directives are held by C/C++ pragma, or by Fortran comments. Most of them
+OpenMP directives are held by C/C++ \texttt{\#pragma}, or by Fortran comments. Most of them
 apply to structured blocks, represented by an instruction in C/C++ and delimited
@@
 anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
-instruction to apply an annotation to a whole block. Pythran uses comment to
+instruction to apply an annotation to a whole block. Pythran uses comments to
 hold OpenMP directives, and internally stores them as string instructions.
@@
 declaration in Python, and all variables assigned in a function have the
-function scope. Consequently, all variables that are referenced in a function
-are considered when building such variable lists: there is no concept of
-variable local to a block. Using the \texttt{default(none)} clause is possible
-to ensure no variable gets forgotten when building such lists.
+function scope, called local scope and available through the \texttt{local()}
+built-in. As a consequence, all variables that are referenced in a function are
+considered when building such variable lists: there is no concept of variable
+local to a block. Using the \texttt{default(none)} clause is possible to ensure
+no variable gets forgotten when building such lists.
 
@@
 \texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
-the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
+the \texttt{and} and \texttt{or} keywords have a special meaning in Python.} The
+latest OpenMP specifications~\cite{openmp4} describe a way to declare
+user-defined reduction but is not implemented in any compiler yet.
 
@@
 Using function scope for all variables may lead to very long \texttt{private}
-clauses. We propose an algorithm to automatically limit the scope of a variable
-declaration so that it is can be automatically marked as private. The whole idea
-is to compute the minimal nesting depth of a variable usage. To do so, a
-\emph{nesting depth} is first associated to each expression. It lexically
-corresponds to the indentation level of its instruction, to the notable
-exception of the iterator declaration in a for loop, that holds a
+clauses. This section describes an algorithm to automatically limit the scope of
+a variable declaration so that it is can be automatically marked as private, in
+a similar manner to C/C++
+
+The whole idea is to compute the minimal nesting depth of a variable usage. To
+do so, a \emph{nesting depth} is first associated to each expression. It
+lexically corresponds to the indentation level of its instruction, to the
+notable exception of the iterator declaration in a for loop, that holds a
 nesting depth equals to 1 plus the loop declaration indentation level.
@@
 
-Thanks to this automatic scoping computation, it is possible to write shorted
-OpenMP directive in a similar manner to C, as illustrated in
-Listing~\ref{lst:motivating-example-annotated-v1}
+Thanks to this automatic scoping computation, it is possible to write shorter
+OpenMP directive, as illustrated in
+Listing~\ref{lst:motivating-example-annotated-v1}.
 
@@
 \texttt{map} function that internally spawns a parallel region. This leads to a
-well known problem with OpenMP: nested parallel regions. OpenMP provides several
-ways to handle this situation:
+well known situation with OpenMP, where too many threads may be spawned: nested
+parallel regions. OpenMP 4 provides several ways to handle this situation:
 
@@
 
-    \item A
+    \item Disabling nested parallelism, through the \texttt{OMP\_NESTED}
+        environment variable or the \texttt{omp\_set\_nested} routine.
 
-    \item B
+    \item Setting the maximum number of active levels, through the
+        \texttt{OMP\_MAX\_ACTIVE\_LEVELS} environment variable or the
+        \texttt{omp\_set\_max\_active\_levels} routine.
 
-    \item C
+    \item Conditionally activating regions or conditionally setting the number
+        of threads in a parallel region using the \texttt{omp\_in\_parallel}
 
 \end{enumerate}
+
+The decision is left to the user, depending on its application and OpenMP
+implementation.
 
@@
 
+The target machine has  8 AMD Opteron 6176 SE cores and 8 GO of RAM. It is
+running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
+\texttt{g++} version 4.8.1 and all applicationsa re compiled using the
+\texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
+version 1.54. The Pythran version is extracted from the \texttt{sc2013} branch
+of the git repository \url{https://github.com/serge-sans-paille/pythran}. Python
+programs are run using the CPython implementation, version 2.7.5. Cython
+programs are generated using version 0.19.1. Numba and Parakeet are installed
+from the \texttt{master} branch of their respective repositories.
+
 \subsection{Motivating Example}
@@
 paper. We also measure the SLOC of the two versions.  Table~\ref{tbl:hyantes}
-summarizes the results of the experimentation made on a laptop with 4 non
-hyperthreaded cores using \texttt{-O2} compiler flag. It not only emphasises the
+summarizes the results of the experimentation. It not only emphasises the
 use of a high level language prototyping, but also shows that it is possible to
@@
 the relative speedup of the application when increasing the number of OpenMP
-threads using up to 8 AMD Opteron 6176 SE.
+threads using up to 8 cores.
 
@@
 def sum_sqrt(r):
-    "omp parallel for reduction(+:r)"
+    #omp parallel for reduction(+:r)
     for i in xrange(10000000):
@@
 through the mechanism of each language to exhibit parallelism, then compiled
-into native code using GCC 4.7.2 and the \texttt{-O2 -fopenmp} flag.  Their
-execution time when called from the Python interpreter is measured through the
-\texttt{timeit} module on a machine using 8 AMD Opteron 6176 SE.  All results
-are normalized against Cython sequential execution time. They show that while
-handling code at a higher level than Cython, Pythran achieves comparable
-results.
+into native code.  Their execution time when called from the Python interpreter
+is measured through the \texttt{timeit} module. All results are normalized
+against Cython sequential execution time. They show that while handling code at
+a higher level than Cython, Pythran achieves comparable results.
 
 The source codes used for these benchmarks are available on the Pythran
-repository.\footnote{\see \texttt{https://github.com/serge-sans-paille/pythran}
-    on \texttt{dls2013}
-branch.}
+repository.\footnote{\see the \texttt{sc2013} branch of the git repository.}
 
@@
 \subsection{Python Benchmarks Test Bed}
+
+To Be FINISHED
 
@@
 
-This paper studies the addition of OpenMP annotations to Python. It shows that
+This paper studies the addition of OpenMP annotations to Pythran. It shows that
 providing minor semantic adaptations, these annotations make it possible for
@@
 OpenMP directives and a few type annotations into native parallel module. The
-input module remains compatible with the standard interpreter.
+input module remains compatible with the standard interpreter and the underlying
+runtime is compatible with parallel constructs.
 
@@
 compatibility does not prevent the achievement of comparable performances.
+Comparisons to Numba and Parakeet shows the benefit of using parallelism on the
+benchmarks from the Python Benchmark project.
 
-Future work includes the extension of the approach to OpenMP 4 and the
+Future work will focus on the extension of the approach to OpenMP 4 and the
 \texttt{target} clause that should make it possible to target MIC from Python.
-There is also many vectorization possibility in Python, especially in the list
-comprehension construction, that needs to be explored.
+The \texttt{simd} clause also brings interesting vectorization capabilities that
+can be exposed at the python level.  There are also several implicit vectorization
+opportunities in Python, especially in the list comprehension construction, that
+needs to be explored.
 

### 2013-09-03 doc/papers/sc2013/pi_buffon.py
Almost full revision of the paper
last benchmark part still needs to be written...

@@ -15,20 +15,26 @@ def pi_estimate(darts):
     return pi
 
 from threading import Thread
-def threaded_pi_estimate(darts):
-    t1 = Thread(target=pi_estimate, args=(darts//4,))
-    t2 = Thread(target=pi_estimate, args=(darts//4,))
-    t3 = Thread(target=pi_estimate, args=(darts//4,))
-    t4 = Thread(target=pi_estimate, args=(darts//4,))
-    t1.start() ; t2.start()
-    t3.start() ; t4.start()
-    t1.join() ; t2.join()
-    t3.join() ; t4.join()
+from multiprocessing import Queue
+def threaded_pi_estimate(d):
+    def work(darts, queue):
+        queue.put(pi_estimate(darts))
+    n = 4
+    q = Queue()
+    threads = [
+      Thread(target=work, args=(d//n, q)),
+      Thread(target=work, args=(d//n, q)),
+      Thread(target=work, args=(d//n, q)),
+      Thread(target=work, args=(d//n, q)),
+    ]
+    map(Thread.start, threads)
+    map(Thread.join, threads)
+    return sum(q.get() for _ in threads)/n
 
 from multiprocessing import Pool
 def multi_pi_estimate(darts):
     p = Pool(4)
-    p.map(pi_estimate,[darts//4]*4)
+    return sum(p.map(pi_estimate,[darts//4]*4)) / 4
 
 if __name__ == '__main__':
     median = lambda x: sorted(x)[len(x) / 2]

### 2013-09-04 doc/papers/sc2013/paper.tex
SC2013: fix affilitation

@@
     \IEEEauthorblockA{\newline\IEEEauthorrefmark{3}
-    Silkan, Meudon la For\^et, France}
+    SILKAN Inc., Los-Altos, USA}
 }
@@
     common situation to have HPC application written in a high-level language
-    such as Python, calling native routines for computation-intensive tasks. To
-    improve development speed and reduce maintenance costs, several solutions
-    such as Cython, Numba, Parakeet or Pythran, have appeared to make it
-    possible to automatically or semi-automatically turn Python functions into
-    native ones.
+    such as Python, calling native routines for computate-intensive tasks. To
+    improve development speed and reduce maintenance costs, using a higher-level
+    language like Python seems attractive. While it is usually associated with
+    low performance, several solutions such as Cython, Numba, Parakeet or 
+    Pythran offer to automatically or semi-automatically turn Python functions 
+    into native ones.
 
-    One of the key point to match the performance of native applications is the
-    ability to write parallel applications. This paper studies the addition of
-    OpenMP directives, a popular model to describe the parallelism of
-    C/C++/FORTRAN applications, to Pythran, an automatic compiler from a subset
-    of Python to C++. It shows that scientific Python applications annotated
-    with OpenMP directives can be turned by an automatic compiler into native
-    applications that run within the same order of magnitude than
+    One of the key points required to match the performance of native 
+    applications is the ability to write parallel applications. This paper
+    studies the addition of OpenMP directives, a popular model to describe
+    parallelism in C/C++/FORTRAN applications, to Pythran, an automatic compiler     
+    from a subset of Python to C++. It shows that scientific Python applications
+    annotated with OpenMP directives can be turned by an automatic compiler into 
+    native applications that run within the same order of magnitude than
     manually-written ones.
-
 \end{abstract}
@@
 services, thanks to its dynamicity, expressiveness, rich ecosystem and
-battery-included standard library. It is also getting widely used in scientific
-computing~\cite{Oliphant2007}, mainly thanks to the SciPy~\cite{scipy} project
-that provides a set of scientific and numeric tools, ranging from linear algebra
-to ordinary differential equation solvers or generic algorithms.
+% Mehdi : battery-included ??????
+battery-included standard library. It is also widely used in scientific
+computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
+which provides a set of scientific and numeric tools, ranging from linear algebra to ordinary differential equation solvers or generic algorithms.
 
-It is now heading toward High Performance Computing(HPC), either as a glue
-language used to bind several native libraries together, or to write whole
-applications in Pure Python. However, the prohibitive performance overhead of
-the language implied by its interpreted and highly dynamic nature prevents its
-usage for the most performance - critical code sections, where native code is
-generally used. The SciPy modules partly overcome this issue through the usage
-of low-level routines written in C or Fortran and encapsulated in Python
-\emph{native} modules. Moreover, its core data structure, the multi-dimensional
+It is now heading toward High Performance Computing (HPC), either as a glue
+language used to bind several native libraries together, or to support whole
+applications in Pure Python. However, the prohibitive overhead of the language
+implied by its interpreted and highly dynamic nature prevents its usage for the
+most performance-critical code sections, where native code is generally used.
+The SciPy modules partly overcome this issue through the usage of low-level 
+routines written in C or Fortran and encapsulated in Python \emph{native} 
+modules. Moreover, its core data structure, the multi-dimensional
 array~\cite{numpyarray2011}, has been designed so that the underlying data are
@@
 
-To enable the user to write native functions without having to write the glue
-code that turns Python object into native structures back and forth, the SciPy
-package provides the \texttt{weave} module that makes it possible to bundle C
-code snippet into Python code and compile then load them at runtime. This hybrid
-approach, based on a mix of interpreted and native code, is getting widespread
-in the Python landscape. It is also the recommended approach to write
-application that use fine grain parallelism, has Python parallelism support
-targets coarse-grain parallelism.
+To allow the user to write native functions without having to write the boilerplate glue code that turns Python object into native structures back and
+forth, the SciPy package provides the \texttt{weave} module. It makes
+possible to bundle C code snippets into Python code. They are then compiled and 
+loaded at runtime. This hybrid approach, based on a mix of interpreted and
+native code, is getting widespread in the Python landscape. It is also the
+recommended approach to write applications that make use of fine grain
+parallelism, as Python parallelism support is only suitable for coarse-grain
+parallelism.
 
@@
 existing approaches to compile Python functions or modules and shows a critical
-lack of parallelism support, and emphasises on the need for a
+lack of parallelism support. It also emphasises on the need for a
 backward-compatible approach. Section~\ref{sec:python-static} presents the
@@
 studies the validity of using OpenMP directives within the Python language in
+% I would dispute the fact that OpenMP is "fine-grain", we're not talking about % vectorization here.
 order to add fine-grain parallelism support to Python in the context of Pythran.
-The Pythran compiler is benchmarked on several scientific kernels, compared
-to Cython, Numba and Parakeet in Section~\ref{sec:validation}.
+Finally, the Pythran compiler is benchmarked on several scientific kernels and
+compared to Cython, Numba and Parakeet in Section~\ref{sec:validation}.
 
@@
 
-In the many cores area, turning to parallelism to balance the performance
-limitations of scripting languages, as described in~\cite{choy05} in the context
-of the MATLAB language, or in~\cite{mals07} in the context of the R language,
-seems legitimate. In the context of Python, most approaches have focused on
+% Mehdi area => era, Serge: validate?
+% I also wonder the rational of using parallelism when performance are bad 
+% because language is bad…
+In the many-cores era, it is mandatory to exhibit parallelism to balance the 
+performance limitations of scripting languages, as described in~\cite{choy05}
+in the context of the MATLAB language, or in~\cite{mals07} in the context of 
+the R language. In the context of Python, most approaches have focused on
 fork-based parallelism.
@@
 Parallel computations are supported by the Python standard library through the
-\texttt{multiprocessing} module. It spawns several interpreters that can
+\texttt{multiprocessing} module. It spawns several interpreters that
 communicate through Inter Process Communication(IPC), using Python built-in
 object serialization. This approach is only viable for computation intensive
-application, since the communication and synchronization overheads are much
-greater than a light-weight processes approach.
+independent tasks, since the communication and synchronization overheads are 
+much greater than a light-weight threaded approach.
 
-The standard \texttt{threading} module also makes it possible to start
-several light-weight threads within the same interpreter, but this lower-level approach is not
-applicable to HPC, because of a specificity of CPython, the \emph{Global
+The standard \texttt{threading} module also makes possible to start several
+light-weight threads within the same interpreter, but this lower-level approach
+is not applicable to HPC, because of a specificity of CPython, the \emph{Global
 Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
 IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
-is active at a time in the interpreter. While it makes it possible to have
+is active at a time in the interpreter. While it enables the possibility of
 cooperative threads, say for a GUI, it does not take advantage of multiple
@@
 def multi_buffon(darts):
-  p = Pool(4)
-  return sum(p.map(buffon, [darts//4]*4))/4
+  n = 4
+  p = Pool(n)
+  return sum(p.map(buffon, [darts//n]*n))/n
 \end{lstlisting}
+% Mehdi: I don't get the [darts//4]*4 at the last line of the listing
 
@@
 
-In the context of interpreted languages, \cite{dongara2007} defines a computation
-as \emph{hybrid} when part of the code is interpreted, and part of it is
-executed natively.
+In the context of interpreted languages, \cite{dongara2007} defines a 
+computation as \emph{hybrid} when part of the code is interpreted, and part of
+it is executed natively.
 
@@
 parallel C/C++ functions and the associated boilerplate based on the Python C
-API~\cite{pythoncapi}. Tools have been developed to relieve the user from
-writing the encapsulating code, notably SWIG~\cite{swig2003} that relies on an
-enhanced interface specification, or
-\texttt{boost::python}~\cite{boostpython2007} that relies on C++ facilities to
-guide translation.
+API~\cite{pythoncapi}. Tools have been developed to relieve the user from this
+cumbersome task, notably SWIG~\cite{swig2003} that relies on an interface specification supplied by the programmer to generate the glue, or 
+\texttt{boost::python}~\cite{boostpython2007} that relies on C++ templates and 
+type overloading facilities to guide translation.
 
 An opposite approach consists in using the host language ---herein Python--- to
-describe both parts of the system, and to let an automated tool perform the
-translation to native code of a part of the application, generally the
-computation-intensive one where parallelism has been expressed in some ways.
-Thus, developers not familiar with lower level languages or not eager to invest
-the additional development time can still benefit from a fair performance boost.
+describe both parts of the system, i.e.\ the hybrid and the native. An automated 
+tool performs the translation to native code of a specific part of the 
+application, generally the computation-intensive one where parallelism has been
+expressed in some ways.
+Therefore, developers not familiar with lower level languages or not eager to 
+invest the additional development time can still benefit from a fair performance
+boost.
 This approach is the subject of many studies that can be classified according to
@@
 to define parallel loops, but no task parallelism. When possible, it unboxes
-Python variable to improve performance. Without type annotation, the performance
-improvement is not terrific, but given enough type annotation, Cython can
-generates code that runs as fast as its C equivalent, while maintaining a syntax
-close to Python.
+Python variables to improve performance. Without type annotations, the 
+performance improvement is not terrific, but given enough type annotation, Cython can generates code that runs as fast as its C equivalent, while 
+maintaining a syntax close to Python.
 
@@
 
-The main issue of these approaches is that they imply to modify, in a
-backward-incompatible way, the original code. They require to learn a new
+The main drawback of these approaches is that they require to modify in a
+backward-incompatible way the original code. They require to learn a new
 dialect, and the long-term preservation of this investment is not ensured.
-Moreover, there is no falling back if the code were to be deployed in an
+Moreover, there is no fall back if the code were to be deployed in an
 environment where the parallelizing tool/module is not available. For instance,
@@
 \texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
-not suffer from the backward-incompatibility issues. When meeting an unsupported
+not suffer from backward-incompatibility issues. When meeting an unsupported
 construct, numba falls back to Python C-API calls, while parakeet raises an
@@
 
+%FIXME: why not cite OpenMP 4.0 (aka the latest)?
 This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
@@
 write fine-grain parallel applications in Python, while being fully backward
-compatible with both the Python language and the sequential algorithm. It means
-that:
+compatible with both the Python language and the sequential algorithm. It 
+implies that:
 

### 2013-09-04 doc/papers/sc2013/paper.tex
Finish benchmark section
Convert the whole paper back to linux EOL, comment the python benchmarks
example.

@@
 % Mehdi : battery-included ??????
-battery-included standard library. It is also widely used in scientific
+``batteries included'' standard library. It is also widely used in scientific
 computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
@@
   p = Pool(n)
-  return sum(p.map(buffon, [darts//n]*n))/n
+  return sum(p.map(buffon, darts//n))/n
 \end{lstlisting}
@@
 generated code are illustrated in the context of the Python Benchmarks
-suite~\footnote{\url{https://github.com/numfocus/python-benchmarks}}.
+suite.
 
@@
 
-To Be FINISHED
+The Python Benchmark test bed
+(\url{https://github.com/numfocus/python-benchmarks}) is an innitiative to
+gather relevant numerical python benchmarks and compare several implementations
+and compilers. The original benchmarks are slightly adapted to match each
+compiler restriction (e.g. lack of support for a particular construct), or
+completely rewritten in the target's paradigm (e.g. using PyOpenCL or Theano).
+We picked up the targets that did not imply a full rewrite of their input:
+Cython, Numba, Pythran and Parakeet, when available and supported.
+Additionnaly, the original Python code is run.
+
+The best execution time out of five run is shown for each code version. If the
+code does not compile with a particular compiler, the column is left blank. Two
+code versions are shown for Pytrhan: one compiler without the OpenMP flag set
+and one with the OpenMP flag set. The code itself is not modified, only the
+compilation flag. A logarithmic scale is used due to the generally poor
+performance of the pure Python version.
+
+Figure~\ref{fig:pb-julia} shows the execution time of the computation of the
+julia fractal. It's a highly parallel, computate-intensive code. Pythran and
+Cython's version have almost the same execution time, but activating OpenMP
+yields an extra $\times 3.95$ speedup when compared to Pythran alone.
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{julia}
+    \caption{Comparison of Various python Compilers on the Julia benchmark}
+    \label{fig:pb-julia}
+\end{figure}
+
+Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a distance matrix between two random vectors, using a parallel nested loop.
+On that example, Pytrhan slightly outperforms Cython, parakeet and Numba.
+Taking into account the OpenMP annotation that flags the outermost loop as parallel yields an $\times 4.27$ speedup to Pythran.
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{pairwise}
+    \caption{Comparison of Various python Compilers on the pairwise benchmark}
+    \label{fig:pb-pairwise}
+\end{figure}
+
+Figure~\ref{fig:pb-rosen} shows the execution time of the computation of the third derivative of the Rosenbrock function for a random array input, using Numpy's vector operations.
+There is no explicit parallelism in this function but the vector operation is implictly parallel.
+On this example, Cython is on par with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the Pythran version run $\times 1.78$ faster than sequential Pythran.
+
+\begin{figure}[ht]
+    \includegraphics[width=.5\textwidth]{rosen_der}
+    \caption{Comparison of Various python Compilers on the rosen\_der benchmark}
+    \label{fig:pb-rosen}
+\end{figure}
+
+Figure~\ref{fig:pb-rosen} shows the execution time of the computation of one step of the GrowCut algorithm used in image processing.
+On this example, Parakeet performs better than Cython and Pytrhan and way better than Numba, but paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to Pythran that helps it outperform Parakeet.
 
@@
 
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{julia}
-    \caption{Comparison of Various python Compilers on the Julia benchmark}
-    \label{fig:pb-julia}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{rosen_der}
-    \caption{Comparison of Various python Compilers on the rosen\_der benchmark}
-    \label{fig:pb-rosen}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{pairwise}
-    \caption{Comparison of Various python Compilers on the pairwise benchmark}
-    \label{fig:pb-pairwise}
-\end{figure}
-growcut: $\times 2.65$
-julia: $\times 3.95$
-rosen\_der: $\times 1.78$
-pairwise: $\times 4.27$
+It appears that all Python compilers manage to yield significant speedups over
+the Python interepreter, each compiler managing some cases better than others.
+All the benchmarks exhibt some level of parallelism that are easy to capture
+using OpenMP without changing the original algorithm and while maintaining
+backward compatibility with the Python interpreter. Some even have a degree of
+implicit parallism that makes parallelization fully transparent to the user.
+Taking advantage of this parallelism appears to be the key to reach an additional performance level.
 

### 2013-09-05 doc/papers/sc2013/paper.tex
Pierrick review

@@
 
-    HPC users traditionally rely on low-level, compiled language such as C or
+    High Performance Computing users traditionally rely on low-level,
+    compiled language such as C or
     FORTRAN to perform computation-intensive tasks. As a consequence, it is a
-    common situation to have HPC application written in a high-level language
+    common situation to have High Performance Computing application written in a high-level language
     such as Python, calling native routines for computate-intensive tasks. To
@@
 \texttt{multiprocessing} module. It spawns several interpreters that
-communicate through Inter Process Communication(IPC), using Python built-in
+communicate through Inter Process Communication (IPC), using Python built-in
 object serialization. This approach is only viable for computation intensive
@@
 PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
-Python does not have anonymous block,~\footnote{\see PEP340,
+Python does not have anonymous block,\footnote{\see PEP340,
     \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
@@
 
+%what about flush clause which is not linked to a statement?
 Many OpenMP annotations are parametrized by clauses that list variables,
@@
 depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
-of \{D, U\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
+of \{DEF, USE\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
 automatically declared local to the loop, thus not needing to be listed as
@@
 The extension of Python iterator to random access iterators is direct for the
-standard containers: \texttt{list}\footnote{Python \texttt{list}s behave as C++
-\texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
+standard containers: \texttt{list},\footnote{Python \texttt{list}s behave as C++
+\texttt{vector}s.} \texttt{set} or \texttt{dict}. Other iterators require more
 care.
@@
 running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
-\texttt{g++} version 4.8.1 and all applicationsa re compiled using the
+\texttt{g++} version 4.8.1 and all applications are compiled using the
 \texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
@@
 
+%Ce n'est pas ce code qui ne donne pas les mêmes résultats avec Cython et Python et Pythran?
+
 The performance of the two approaches is shown in
@@
 code does not compile with a particular compiler, the column is left blank. Two
-code versions are shown for Pytrhan: one compiler without the OpenMP flag set
+code versions are shown for Pythran: one compiler without the OpenMP flag set
 and one with the OpenMP flag set. The code itself is not modified, only the
@@
 Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a distance matrix between two random vectors, using a parallel nested loop.
-On that example, Pytrhan slightly outperforms Cython, parakeet and Numba.
+On that example, Pythran slightly outperforms Cython, parakeet and Numba.
 Taking into account the OpenMP annotation that flags the outermost loop as parallel yields an $\times 4.27$ speedup to Pythran.

### 2013-09-06 doc/papers/sc2013/paper.tex
Update SC2013 paper

@@
 It is now common for scripting language to have C bindings. To take advantage of
-compiled code, and to overcome the GIL limitations, Python developers must write
+compiled code, and to overcome the GIL limitations, Python developers has to write
 parallel C/C++ functions and the associated boilerplate based on the Python C
@@
 
-To be completely backward compatible, it must be possible to run the input code
+To be completely backward compatible, it has to be possible to run the input code
 in an environment that is not aware of the existence of the parallelizing
@@
 This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
-with the Python subset supported by the Pythran compiler to make it possible to
+with the Python subset supported by the Pythran compiler to make possible to
 write fine-grain parallel applications in Python, while being fully backward
@@
 not supported. The associated compiler turns Pythran code, possibly annotated
-with OpenMP and a few function type annotations, into C++ code. Its processing
-is summarized in Figure~\ref{fig:pythran-compiler} and relies on type inference
-to remove most dynamic behavior, combined with a high-level runtime library,
-\texttt{pythonic++}, to allow a one-to-one mapping between Python and C++
-constructs. The purpose of this section is not to dig into the internals of the
-compiler, but rather to focus on the impact of the parallelism layer, especially
-on the runtime library.
+with OpenMP and a few function type annotations, into C++ code.
+Figure~\ref{fig:pythran-compiler} summarizes its processing: type inference
+helps removing most dynamic behavior, and is combined with a high-level runtime
+library ---\texttt{pythonic++}--- to allow a one-to-one mapping between Python and
+C++ constructs. The purpose of this section is not to dig into the internals of
+the compiler, but rather to focus on the impact of the parallelism layer,
+especially on the runtime library.
 
@@
 
-Pythran runtime is based on the STL. The standard library of C++11, unlike its
-C++03 version, is thread-safe, so no data dependencies are added to the original
-code.
+% STL != standard library
+Pythran runtime is based on the C++11 standard library which, unlike its C++03
+version, is thread-safe. It implies that no data dependencies are added to the
+original code.
 
@@
 
-Pythran handles the problem by refusing recursive types, which makes it possible
+Pythran handles the problem by refusing recursive types, which makes possible
 to solely rely on reference counting for memory management. It can be
@@
 Using shared references simplifies memory management, but the counterpart is an
-extra atomic operation for each copy. As it limits parallelism, reducing the
-number of copies becomes an important goal. The move semantic introduced in
+extra atomic operation for each copy.
+% UNCLEAR
+As it limits parallelism, reducing the number of copies becomes an important
+goal.
+% end unclear
+The move semantic introduced in
 C++11 avoids a few copies when working on temporary objects, but argument
@@
 parameters by reference. Using an inter-procedural memory effect analysis not
-presented in the paper, Pythran determines for each argument whether it must be
-passed using reference or const reference to prevent this overhead.
+presented in the paper, Pythran determines for each argument whether it has to
+be passed using reference or const reference to prevent this overhead.
 
@@
 context-insensitive part ---the clauses---, and attaches them to the proper
-instruction in the Abstract Syntax tree(AST).
-
-The Pythran compiler ensures a bijective translation between Python instruction
-and generated C++ instruction so that the OpenMP directive is regenerated on the
-proper instruction. The same approach is used at the expression level, to be
+construct in the Abstract Syntax tree (AST).
+
+The Pythran compiler ensures a bijective translation between Python constructs
+and generated C++ constructs so that the OpenMP directive is regenerated on the
+proper construct.
+%next sentence unclear
+The same approach is used at the expression level, to be
 compatible with the \texttt{if} clause.
@@
 OpenMP directives are held by C/C++ \texttt{\#pragma}, or by Fortran comments. Most of them
-apply to structured blocks, represented by an instruction in C/C++ and delimited
-by comments in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
+apply to structured blocks in C/C++ and delimited by comments in Fortran.
+A few directives (e.g.\ \texttt{threadprivate}) are not
 attached to a specific instruction.
@@
 Python decorators.}, it only applies to functions, methods and classes and does
-not make it possible to attach decorations to other statements.
+not allow to attach decorations to other statements.
 PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
@@
 list of supported operators depend on the input and backend languages: Python
-has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
+has \texttt{min}/\texttt{max} operators but C/C++ does not. % C++11 has...
+C/C++ have
 \texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
@@
 clauses. This section describes an algorithm to automatically limit the scope of
-a variable declaration so that it is can be automatically marked as private, in
+a variable declaration so that it can be automatically marked as private, in
 a similar manner to C/C++
@@
 
-Tracking the nesting depth of all variable use, for instance using def-use
-chains, makes it possible to build a list of nesting depths, one per variable
+Tracking the nesting depth of all variable use, for instance using def-use % I don't see the need for def-use chains for this! Unless you include some "unaliasing algo"
+chains, makes possible to build a list of nesting depths, one per variable
 reference. Computing the min of this list yields the scope of the variable.
@@
 distributes the iteration space of the associated loop among the existing
-threads. To be compatible with OpenMP, the loop iteration space must be
+threads. To be compatible with OpenMP, the loop iteration space has to be
 described by a random access iterator with a total order. Integers used as loop
@@
 behaves as a forward iterator. To be compatible with OpenMP, these iterators
-must be turned into random access iterators.
+has to be turned into random access iterators.
 
@@
 
-Pythran's runtime is aware of these three kind of iterators and supports
+Pythran's runtime is aware of these three kinds of iterators and supports
 parallel iteration over random access iterators and iterators adapting random
@@
 
-OpenMP provides a small runtime library that, for instance, makes it possible to
+OpenMP provides a small runtime library that, for instance, makes possible to
 retrieve the active thread id. All the functions are declared in the
@@
 
-Providing a binding to these library in Python as an \texttt{omp} module does
+Providing a binding to these libraries in Python as an \texttt{omp} module does
 not raise particular problems, as the signature of these functions only involves
-integers passed as parameters or return values, except for the mutex
+integers, except for the mutex
 manipulation. In that case an opaque type is used to represent the native type.
@@
 not have a preprocessor, but it is possible to catch the import exception if the
-\texttt{omp} module is not found. Listing~\ref{lst:sample-api-usage} showcase
+\texttt{omp} module is not found. Listing~\ref{lst:sample-api-usage} showcases
 such a call using an example converted from the OpenMP validation suite
@@
 increases the \texttt{nthreads} variable that has unspecified visibility, thus
-\texttt{shared}. The incrementation is protected by a guard through the
+is \texttt{shared}. A guard protects the incrementation using the
 \texttt{critical} directive. Then one of the thread retrieves the number of
-active thread in the parallel region through the \texttt{get\_num\_threads}
+active threads in the parallel region through the \texttt{get\_num\_threads}
 function from the \texttt{omp} module. The function should always return
@@
 the \texttt{sum} function from the \texttt{\_\_builtin\_\_} is implemented using
-the OpenMP \texttt{reduction} clause. When it can prove that its first argument
-is a \emph{pure} function, it can also call a parallel version of the
+the OpenMP \texttt{reduction} clause. When its first argument proves to be a
+\emph{pure} function, it can also call a parallel version of the
 \texttt{map} function that internally spawns a parallel region. This leads to a
@@
 
-We have used the Pythran tool described in the following section to turn each
+We have used the Pythran tool described in the following % following??
+ section to turn each
 Python test function into a C++ function with the same directives and runtime
@@
 
-In addition to the formal validation done using the OpenMP validation suite
+In addition to the experimental validation done using the OpenMP validation suite
 presented in Section~\ref{sec:python-openmp}, the performance of parallel module
@@
 from the \texttt{master} branch of their respective repositories.
+% dommage pour numba et parakeet qui sont les seuls "mal spécifiés"...
 
@@
 equivalent C code. As the program is not memory bound and exhibit trivial
-parallelism, the C + OpenMP version scales linearly. The Pythran + OpenMP
+parallelism, the C + OpenMP version scales almost linearly. The Pythran + OpenMP
 version follows the same pattern.
@@
 
+% geomatic or geomatics?? Unify everywhere.
 To get closer to a realistic example, we have considered the code of a small
 geomatics application, Hyantes. Starting from the C code, we successively
-parallelized it with OpenMP, turned it into Python with a Pythran compatible
+parallelized it using OpenMP, turned it into Python with a Pythran compatible
 kernel and parallelized the Python version using the approach described in this
-paper. We also measure the SLOC of the two versions.  Table~\ref{tbl:hyantes}
-summarizes the results of the experimentation. It not only emphasises the
-use of a high level language prototyping, but also shows that it is possible to
+paper. We also measure the number of Source Lines Of Code (SLOC) of the two
+versions.  Table~\ref{tbl:hyantes} summarizes the result of this
+experimentation. It not only exhibit the use of a high level language
+for prototyping, but also shows that it is possible to
 turn this prototype into reasonably efficient code through Pythran. It is also
@@
 
-We then compare Pythran with Cython. The approaches share some similarities:
+We then compare Pythran with Cython. These approaches share some similarities:
 they both generate code written in a lower level language, with OpenMP
@@
 a higher level than Cython, Pythran achieves comparable results.
+% pythran est bien plus lent quand même :(
 
@@
 
+%% For the two paragraph below. it is not specified if openmp annotations were
+% added by hand. It is misleading since the previous paragraph said
+% "The code itself is not modified, only the compilation flag."
 Figure~\ref{fig:pb-julia} shows the execution time of the computation of the
@@
 This paper studies the addition of OpenMP annotations to Pythran. It shows that
-providing minor semantic adaptations, these annotations make it possible for
-Python code to benefit from multi-cores while retaining backward compatibility
-and without worrying about the Global Interpreter Lock.
+providing minor semantic adaptations, these annotations enable Python code
+to benefit from multi-cores while retaining backward compatibility and without
+worrying about the Global Interpreter Lock.
 
@@
 Future work will focus on the extension of the approach to OpenMP 4 and the
-\texttt{target} clause that should make it possible to target MIC from Python.
+\texttt{target} clause that should allow to target accelerator like Intel MIC
+from Python.
 The \texttt{simd} clause also brings interesting vectorization capabilities that
-can be exposed at the python level.  There are also several implicit vectorization
+could be exposed at the python level. There are also several implicit vectorization
 opportunities in Python, especially in the list comprehension construction, that
-needs to be explored.
+need to be explored.
 

### 2013-09-08 doc/papers/sc2013/paper.tex
Minor grammar fix to SC2013

@@
     compiled language such as C or
-    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
+    FORTRAN to perform compute-intensive tasks. As a consequence, it is a
     common situation to have High Performance Computing application written in a high-level language
-    such as Python, calling native routines for computate-intensive tasks. To
+    such as Python, calling native routines for compute-intensive tasks. To
     improve development speed and reduce maintenance costs, using a higher-level
@@
 communicate through Inter Process Communication (IPC), using Python built-in
-object serialization. This approach is only viable for computation intensive
+object serialization. This approach is only viable for compute-intensive
 independent tasks, since the communication and synchronization overheads are
@@
 tool performs the translation to native code of a specific part of the
-application, generally the computation-intensive one where parallelism has been
+application, generally the compute-intensive one where parallelism has been
 expressed in some ways.
@@
 Figure~\ref{fig:pb-julia} shows the execution time of the computation of the
-julia fractal. It's a highly parallel, computate-intensive code. Pythran and
+julia fractal. It's a highly parallel, compute-intensive code. Pythran and
 Cython's version have almost the same execution time, but activating OpenMP

### 2013-09-27 doc/papers/cc2014/paper.tex
Setup format for cc2014 submission

--- 2013-09-05 doc/papers/sc2013/paper.tex
@@
-% to submit via
-% https://www.easychair.org/account/signin.cgi?timeout=1;conf=pyhpc2013
-% max 10 pages
-\documentclass[conference]{IEEEtran}
-%
+\documentclass{llncs}
+
 \usepackage{listings}
@@
 \usepackage{todo}
-\usepackage{xcolor}
-\newcommand\see{\emph{cf.\ }}
 
@@
 
+\bibliographystyle{splncs}
 \lstset{language=python,morekeywords={cimport, cdef, with},
@@
 
+\title{Pythran: Turning Python Modules into Optimized C++ Meta-Programs}
+
+\author{Serge Guelton\inst{1,2} \and Pierrick Brunet\inst{2} \and Mehdi Amini\inst{3}}
+
+\institute{École Normale Supérieure, Département d'Informatique, Paris, France
+\and
+Télécom Bretagne, Plouzané, France
+\and
+SILKAN Inc., Los-Altos, USA
+}
 \begin{document}
-% update listing numbering
-\renewcommand{\thelstlisting}{\arabic{lstlisting}}
-%
-\title{Compiling Python Modules to Native Parallel Modules Using Pythran and
-OpenMP Annotations}
-
-\author{
-    \IEEEauthorblockN{Serge Guelton\IEEEauthorrefmark{1}, \IEEEauthorrefmark{2}}
-    \IEEEauthorblockN{Pierrick Brunet\IEEEauthorrefmark{2}}
-    \IEEEauthorblockN{Mehdi Amini\IEEEauthorrefmark{3}}
-    \newline
-    \IEEEauthorblockA{\newline\IEEEauthorrefmark{1}
-        {\'E}cole Normale Sup{\'e}rieure, D\'{e}partement d'Informatique, Paris, France}
-    \IEEEauthorblockA{\newline\IEEEauthorrefmark{2}
-    T{\'e}l{\'e}com Bretagne, Plouzan{\'e}, France}
-    \IEEEauthorblockA{\newline\IEEEauthorrefmark{3}
-    SILKAN Inc., Los-Altos, USA}
-}
 
@@
 
-%
 \begin{abstract}
-
-    High Performance Computing users traditionally rely on low-level,
-    compiled language such as C or
-    FORTRAN to perform computation-intensive tasks. As a consequence, it is a
-    common situation to have High Performance Computing application written in a high-level language
-    such as Python, calling native routines for computate-intensive tasks. To
-    improve development speed and reduce maintenance costs, using a higher-level
-    language like Python seems attractive. While it is usually associated with
-    low performance, several solutions such as Cython, Numba, Parakeet or 
-    Pythran offer to automatically or semi-automatically turn Python functions 
-    into native ones.
-
-    One of the key points required to match the performance of native 
-    applications is the ability to write parallel applications. This paper
-    studies the addition of OpenMP directives, a popular model to describe
-    parallelism in C/C++/FORTRAN applications, to Pythran, an automatic compiler     
-    from a subset of Python to C++. It shows that scientific Python applications
-    annotated with OpenMP directives can be turned by an automatic compiler into 
-    native applications that run within the same order of magnitude than
-    manually-written ones.
+...
 \end{abstract}
 
-%
+\section{Introduction}
+...
 \section{Introduction}
 
-Since its birth in December 1989, the Python language~\cite{rossum97} has proved
-to be useful in various domains, ranging from system administration to web
-services, thanks to its dynamicity, expressiveness, rich ecosystem and
-% Mehdi : battery-included ??????
-``batteries included'' standard library. It is also widely used in scientific
-computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
-which provides a set of scientific and numeric tools, ranging from linear algebra to ordinary differential equation solvers or generic algorithms.
-
-It is now heading toward High Performance Computing (HPC), either as a glue
-language used to bind several native libraries together, or to support whole
-applications in Pure Python. However, the prohibitive overhead of the language
-implied by its interpreted and highly dynamic nature prevents its usage for the
-most performance-critical code sections, where native code is generally used.
-The SciPy modules partly overcome this issue through the usage of low-level 
-routines written in C or Fortran and encapsulated in Python \emph{native} 
-modules. Moreover, its core data structure, the multi-dimensional
-array~\cite{numpyarray2011}, has been designed so that the underlying data are
-available to both native modules and the Python interpreter without conversion
-cost. Yet when a new function not already part of an existing module is
-required, one has to write it in C or FORTRAN.
-
-To allow the user to write native functions without having to write the boilerplate glue code that turns Python object into native structures back and
-forth, the SciPy package provides the \texttt{weave} module. It makes
-possible to bundle C code snippets into Python code. They are then compiled and 
-loaded at runtime. This hybrid approach, based on a mix of interpreted and
-native code, is getting widespread in the Python landscape. It is also the
-recommended approach to write applications that make use of fine grain
-parallelism, as Python parallelism support is only suitable for coarse-grain
-parallelism.
-
-This paper presents the addition of OpenMP, a popular standard to turn
-sequential C, C++ and FORTRAN application into parallel applications, to the
-Pythran~\cite{pythran2013} compiler, a compiler that turn Python modules into
-C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
-existing approaches to compile Python functions or modules and shows a critical
-lack of parallelism support. It also emphasises on the need for a
-backward-compatible approach. Section~\ref{sec:python-static} presents the
-Pythran compiler and the underlying runtime. Section~\ref{sec:python-openmp}
-studies the validity of using OpenMP directives within the Python language in
-% I would dispute the fact that OpenMP is "fine-grain", we're not talking about % vectorization here.
-order to add fine-grain parallelism support to Python in the context of Pythran.
-Finally, the Pythran compiler is benchmarked on several scientific kernels and
-compared to Cython, Numba and Parakeet in Section~\ref{sec:validation}.
-
-%=========================================================
-\section{Parallel Computations in Python}\label{sec:python-parallel}
-%=========================================================
-
-% Mehdi area => era, Serge: validate?
-% I also wonder the rational of using parallelism when performance are bad 
-% because language is bad…
-In the many-cores era, it is mandatory to exhibit parallelism to balance the 
-performance limitations of scripting languages, as described in~\cite{choy05}
-in the context of the MATLAB language, or in~\cite{mals07} in the context of 
-the R language. In the context of Python, most approaches have focused on
-fork-based parallelism.
-
-\subsection{Python and Parallelism}
-
-Parallel computations are supported by the Python standard library through the
-\texttt{multiprocessing} module. It spawns several interpreters that
-communicate through Inter Process Communication (IPC), using Python built-in
-object serialization. This approach is only viable for computation intensive
-independent tasks, since the communication and synchronization overheads are 
-much greater than a light-weight threaded approach.
-
-The standard \texttt{threading} module also makes possible to start several
-light-weight threads within the same interpreter, but this lower-level approach
-is not applicable to HPC, because of a specificity of CPython, the \emph{Global
-Interpreter Lock}~\cite{gil2012}.\footnote{Other Python interpreters, such as
-IronPython or Jython, do not have a GIL.} This lock ensures that only one thread
-is active at a time in the interpreter. While it enables the possibility of
-cooperative threads, say for a GUI, it does not take advantage of multiple
-cores. However, there are two notable exceptions: the GIL is released on I/O,
-and the GIL does not prevent the use of threads inside native modules, where the
-user has full control.
-
-To illustrate the limitations of these two approaches, we used the CPU-bound
-Buffon's needle algorithm to estimate the value of $\pi$. A sequential version
-parallelized using the two approaches is illustrated in
-Listing~\ref{lst:buffon-python}. When comparing their respective execution time
-using 4 cores, it appears that the version that uses the \texttt{threading}
-module runs $\times 1.46$ the execution time of the sequential version. GIL
-contention actually \emph{increases} the execution time. The version that uses
-\texttt{multiprocessing} provides a speedup of 3 over the sequential version
-while a speedup of 4 would be expected for this kind of application.
-
-\begin{lstlisting}[float, label={lst:buffon-python}, caption={Implementation of
-sequential and parallel version of the Buffon algorithm in Python.}]
-def buffon(darts):
-  hits = 0
-  for i in xrange (0, darts):
-    x, y = random(), random()
-    dist = sqrt(pow(x, 2) + pow(y, 2))
-    if dist <= 1.0:
-      hits += 1.0
-  # hits / throws = 1/4 Pi
-  pi = 4 * (hits / darts)
-  return pi
-
-from threading import Thread
-from multiprocessing import Queue
-def threaded_buffon(d):
-  def work(darts, queue):
-    queue.put(pi_estimate(darts))
-  n = 4
-  q = Queue()
-  threads = [
-    Thread(target=work, args=(d//n, q)),
-    Thread(target=work, args=(d//n, q)),
-    Thread(target=work, args=(d//n, q)),
-    Thread(target=work, args=(d//n, q)),
-  ]
-  map(Thread.start, threads)
-  map(Thread.join, threads)
-  return sum(q.get() for _ in threads)/n
-
-from multiprocessing import Pool
-def multi_buffon(darts):
-  n = 4
-  p = Pool(n)
-  return sum(p.map(buffon, darts//n))/n
-\end{lstlisting}
-% Mehdi: I don't get the [darts//4]*4 at the last line of the listing
-
-There have also been several approaches to replace the GIL by Transactional
-Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
-mainstream interpreters. As a consequence, Python developers need to write
-multi-threaded native modules in order to fully benefit from multiple cores.
-This leads to a kind of computations referred as \emph{hybrid computations}.
-
-\subsection{Hybrid Computations}
-
-In the context of interpreted languages, \cite{dongara2007} defines a 
-computation as \emph{hybrid} when part of the code is interpreted, and part of
-it is executed natively.
-
-It is now common for scripting language to have C bindings. To take advantage of
-compiled code, and to overcome the GIL limitations, Python developers must write
-parallel C/C++ functions and the associated boilerplate based on the Python C
-API~\cite{pythoncapi}. Tools have been developed to relieve the user from this
-cumbersome task, notably SWIG~\cite{swig2003} that relies on an interface specification supplied by the programmer to generate the glue, or 
-\texttt{boost::python}~\cite{boostpython2007} that relies on C++ templates and 
-type overloading facilities to guide translation.
-
-An opposite approach consists in using the host language ---herein Python--- to
-describe both parts of the system, i.e.\ the hybrid and the native. An automated 
-tool performs the translation to native code of a specific part of the 
-application, generally the computation-intensive one where parallelism has been
-expressed in some ways.
-Therefore, developers not familiar with lower level languages or not eager to 
-invest the additional development time can still benefit from a fair performance
-boost.
-This approach is the subject of many studies that can be classified according to
-their compatibility with the host language.
-
-\subsubsection{Backward-Incompatible Approaches}
-
-A constraining (from the performance point of view) aspect of the Python
-language is its type system. It implies that each method call is resolved
-dynamically, even a simple add operation. It comes at no surprise that many
-approaches restrain the Python language to add a static typing overlay.  Also,
-only two types of integers (64-bits integers and multi-precision integers) and
-one type of floating point type (double-precision floats) are available in
-Python, so using a type with the appropriate size may lead to significant
-performance boost.
-
-Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
-syntax with typing informations, calls to native functions from third party
-libraries, and a limited set of parallelism constructs, such as the possibility
-to define parallel loops, but no task parallelism. When possible, it unboxes
-Python variables to improve performance. Without type annotations, the 
-performance improvement is not terrific, but given enough type annotation, Cython can generates code that runs as fast as its C equivalent, while 
-maintaining a syntax close to Python.
-
-PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose an other approach
-that directly mixes Python with C, using raw strings to hold the C code. PLW
-also supports parallel directives that are limited to parallel for loops.
-PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
-Python with kernels embedded as raw strings containing accelerator code.
-
-The main drawback of these approaches is that they require to modify in a
-backward-incompatible way the original code. They require to learn a new
-dialect, and the long-term preservation of this investment is not ensured.
-Moreover, there is no fall back if the code were to be deployed in an
-environment where the parallelizing tool/module is not available. For instance,
-once a code has been ported to PyCUDA and transformed into CUDA code embedded in
-Python strings, there is no way back and the developer has to maintain two
-versions of the algorithm.
-
-\subsubsection{Backward-Compatible Approaches}
-
-Most backward-compatible approaches also require to modify the input program.
-They do not extend the Python language, but restrict it. As a consequence, they
-remain compatible with the original language and do not suffer from the
-drawbacks of the previous approaches. They also benefit from existing tools
-associated to the language.
-
-Copperhead~\cite{copperhead2011} is a functional, data parallel language
-embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
-structure and prohibits usage of many control-flow operators such as loops,
-enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
-intrinsics to exhibit parallelism. But it can be efficiently compiled to either
-CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
-\texttt{http://thrust.github.com/}} library. Python decorators are used to
-identify hot-spots that are JIT-compiled to native code.
-
-The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
-compiler uses additional type information to generate sequential LLVM bytecode.
-Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
-Python to LLVM bytecode translation, but limits its scope to the numpy package,
-only supporting a small subset of the Python language. Additionally, it supports
-implicit parallelism using an implicit mapping between numpy functions and a set
-of parallel primitives including \texttt{map}s \texttt{scan}s and
-\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
-not suffer from backward-incompatibility issues. When meeting an unsupported
-construct, numba falls back to Python C-API calls, while parakeet raises an
-exception.
-
-Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
-Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
-enhance Python performance. However Shed~Skin does not provide support for
-fine-grained parallelism beyond what the standard library proposes. PyPy is
-heading toward STM for parallelism support.
-
-To be completely backward compatible, it must be possible to run the input code
-in an environment that is not aware of the existence of the parallelizing
-solution. This principle is not respected by parallel libraries, but code
-annotations partially satisfy it. This is where parallel annotations shine: the
-original code remains mostly compatible with a compiler that is not aware of the
-annotations, while taking the annotation into account turns the sequential code
-into parallel code.
-
-%FIXME: why not cite OpenMP 4.0 (aka the latest)?
-This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
-with the Python subset supported by the Pythran compiler to make it possible to
-write fine-grain parallel applications in Python, while being fully backward
-compatible with both the Python language and the sequential algorithm. It 
-implies that:
-
-\begin{enumerate}
-
-    \item any Pythran code can be run (sequentially), with no module dependency
-        or code change, by any Python interpreter;
-
-    \item parallelism is explicit and incrementally added to the original code
-        through directives.
-
-\end{enumerate}
-
-%=========================================================
-\section{Static Translation of Python Programs}\label{sec:python-static}
-%=========================================================
-
-Pythran~\cite{pythran2013} is a subset of Python designed for scientific
-computing. It is implicitly statically typed and supports most Python constructs
-except those that involve introspection (e.g.\ \texttt{getattr}) or runtime
-compilation (e.g.\ \texttt{eval}). A few standard modules are supported in
-addition to the core language (e.g.\ \texttt{math}, \texttt{random}). The
-\texttt{numpy} module is currently only partially supported. User classes are
-not supported. The associated compiler turns Pythran code, possibly annotated
-with OpenMP and a few function type annotations, into C++ code. Its processing
-is summarized in Figure~\ref{fig:pythran-compiler} and relies on type inference
-to remove most dynamic behavior, combined with a high-level runtime library,
-\texttt{pythonic++}, to allow a one-to-one mapping between Python and C++
-constructs. The purpose of this section is not to dig into the internals of the
-compiler, but rather to focus on the impact of the parallelism layer, especially
-on the runtime library.
-
-\begin{figure}
-
-\centering
-\begin{tikzpicture}[
-file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
-node distance=0.7cm},
-    tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
-distance=0.7cm}]
-\node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
-\node[file] (omp)	   [xshift=-2em, left=of python] {\textbf{OpenMP}};
-\node[tool] (pythran) [below=of python] {\textbf{Pythran}};
-\node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
-\node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
-\node (empty) [xshift=-1em, left=of gxx] {};
-\node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic++}}};
-\node[file] (annotation)	   [above=of pythonic] {\textbf{Type Info}};
-\node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
-\node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
-
-\draw[very thick, ->] (python) -- (pythran);
-\draw[very thick] (annotation) -| (empty.center);
-\draw[very thick,dashed, ->] (omp) -- (python);
-\draw[very thick, ->] (pythran) -- (meta-cxx);
-\draw[very thick, ->] (meta-cxx) -- (gxx);
-\draw[very thick] (boost) -| (empty.center);
-\draw[very thick, ->] (pythonic) -- (gxx);
-\draw[very thick, ->] (gxx) -- (so);
-\end{tikzpicture}
-
-\caption{Pythran compiler workflow.}
-\label{fig:pythran-compiler}
-\end{figure}
-
-\subsection{Runtime Support}
-
-Pythran runtime is based on the STL. The standard library of C++11, unlike its
-C++03 version, is thread-safe, so no data dependencies are added to the original
-code.
-
-A critical point of the design of the \texttt{pythonic++} runtime is memory
-management, the very same aspect that led to the use of a GIL in CPython.
-Memory management is implemented in Shed Skin through the general-purpose Boehm
-garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
-objects inside parallel regions, thus making memory management explicit for
-these parts.
-
-Pythran handles the problem by refusing recursive types, which makes it possible
-to solely rely on reference counting for memory management. It can be
-implemented through the thread-safe shared pointer mechanism provided by the C++11
-standard library.
-
-Using shared references simplifies memory management, but the counterpart is an
-extra atomic operation for each copy. As it limits parallelism, reducing the
-number of copies becomes an important goal. The move semantic introduced in
-C++11 avoids a few copies when working on temporary objects, but argument
-passing still implies copies. To avoid a reference increment, one can pass
-parameters by reference. Using an inter-procedural memory effect analysis not
-presented in the paper, Pythran determines for each argument whether it must be
-passed using reference or const reference to prevent this overhead.
-
-\subsection{Directive Oblivious Translation}
-
-During Python to C++ translation, Pythran adopts a blind strategy: it does not
-understand the semantics of the annotations. Instead, it just splits each
-annotation into a context-sensitive part ---the variable names--- and a
-context-insensitive part ---the clauses---, and attaches them to the proper
-instruction in the Abstract Syntax tree(AST).
-
-The Pythran compiler ensures a bijective translation between Python instruction
-and generated C++ instruction so that the OpenMP directive is regenerated on the
-proper instruction. The same approach is used at the expression level, to be
-compatible with the \texttt{if} clause.
-
-At the AST level, it means that the Python AST is first reduced to a tree where
-all nodes not available in C++ are transformed into a compatible representation.
-For instance, list comprehension expressions are turned into function calls or
-tuple unpacking is turned in multiple assignments. During this transformation
-process, an expression is always transformed into a single expression and a
-statement is always transformed into a single statement. Then this AST is
-converted into a C++ AST meant to be pretty-printed.
-
-\subsection{Transfer Costs}
-
-When passing containers from Python to C back and forth, a copy of the whole
-container is made to turn the type agnostic, non-contiguous Python data into
-dense typed ones. This extra copy implies an extra cost that is not negligible:
-Passing two lists of float from Python to C++ requires as much as half the time
-to compute the dot product of the same lists directly in Python. Following
-Amdahl's law, this copy overhead greatly hinders the benefits of
-parallelization, and is a well known issue.
-
-The traditional solution is to use a native type that exposes a Python interface
-and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
-implementation of this concept and is the keystone of the Scipy and NumPy
-packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
-at the Python level. As a tool for scientific computing, Pythran supports such a
-structure, implemented as a class wrapping a raw pointer.
-
-%=========================================================
-\section{OpenMP Semantical Adaptations}\label{sec:python-openmp}
-%=========================================================
-
-OpenMP is a standard API for parallel programming for Fortran, C and C++. It
-consists of a set of parallelizing directives and a few runtime library calls.
-If OpenMP is not activated, the directives are ignored, thus enabling
-incremental parallelization of the original source code while keeping the
-original code structure. The languages targeted by OpenMP are statically
-compiled. This section studies the semantical adaptation required to use the
-same API for a scripting language such as Python.
-Listing~\ref{lst:motivating-example} introduces a motivating example through the
-computation of $\pi$, featuring a parallel reduction. The C equivalent with
-OpenMP directives is given in Listing~\ref{lst:motivating-example-c} for
-reference.
-
-\begin{lstlisting}[float, label={lst:motivating-example}, caption={Motivating
-example: computing $\pi$ in Python.}]
-def pi(nsteps):
-  sum, step = 0., 1. / nsteps
-  for i in range(nsteps):
-      x = (i - 0.5) * step
-      sum += 4. / (1. + x**2)
-  return step * sum
-\end{lstlisting}
-
-\begin{lstlisting}[float, label={lst:motivating-example-c}, caption={Motivating
-example: computing $\pi$ in C with OpenMP.}]
-double pi(size_t nsteps) {
-  double sum = 0., step = 1. / nsteps;
-  #pragma omp parallel for reduction(+:sum)
-  for(size_t i = 0; i < nsteps ; ++i)
-  {
-      double x = (i - 0.5) * step;
-      sum += 4. / (1. + x * x);
-  }
-  return step * sum;
-}
-\end{lstlisting}
-
-\subsection{Directives}
-
-OpenMP directives are held by C/C++ \texttt{\#pragma}, or by Fortran comments. Most of them
-apply to structured blocks, represented by an instruction in C/C++ and delimited
-by comments in Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not
-attached to a specific instruction.
-
-While Python has a decorator mechanism\footnote{\see PEP 318,
-\url{http://www.python.org/dev/peps/pep-0318/} for a more detail explanation of
-Python decorators.}, it only applies to functions, methods and classes and does
-not make it possible to attach decorations to other statements.
-PLW~\cite{dongara2007}, uses string instructions to hold such decorations. As
-Python does not have anonymous block,\footnote{\see PEP340,
-    \url{http://www.python.org/dev/peps/pep-0340/} for a discussion concerning
-anonymous block support in Python.} one has to create a dummy \texttt{if 1:}
-instruction to apply an annotation to a whole block. Pythran uses comments to
-hold OpenMP directives, and internally stores them as string instructions.
-Alternatively, the syntax \texttt{if 'my annotation':} is also supported.
-
-%what about flush clause which is not linked to a statement?
-Many OpenMP annotations are parametrized by clauses that list variables,
-specifying their behavior with respect to parallel regions, e.g.\
-\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
-variables that have already been declared. However, there is no variable
-declaration in Python, and all variables assigned in a function have the
-function scope, called local scope and available through the \texttt{local()}
-built-in. As a consequence, all variables that are referenced in a function are
-considered when building such variable lists: there is no concept of variable
-local to a block. Using the \texttt{default(none)} clause is possible to ensure
-no variable gets forgotten when building such lists.
-
-The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
-characterize some data dependencies when performing a parallel reduction. The
-list of supported operators depend on the input and backend languages: Python
-has \texttt{min}/\texttt{max} operators but C/C++ does not. C/C++ have
-\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
-the \texttt{and} and \texttt{or} keywords have a special meaning in Python.} The
-latest OpenMP specifications~\cite{openmp4} describe a way to declare
-user-defined reduction but is not implemented in any compiler yet.
-
-The annotated motivating example is given in
-Listing~\ref{lst:motivating-example-annotated-v0}. Note that unlike in C,
-\texttt{x} has to be listed in a \texttt{private} clause.
-
-\begin{lstlisting}[float, label={lst:motivating-example-annotated-v0}, caption={Motivating
-example: computing $\pi$ in Python with OpenMP.}]
-def pi(nsteps):
-  sum, step = 0., 1. / nsteps
-#omp parallel for reduction(+:sum) private(x)
-  for i in range(nsteps):
-      x = (i - 0.5) * step
-      sum += 4. / (1. + x**2)
-  return step * sum
-\end{lstlisting}
-
-\subsection{Automatic Scoping Computation}
-
-Using function scope for all variables may lead to very long \texttt{private}
-clauses. This section describes an algorithm to automatically limit the scope of
-a variable declaration so that it is can be automatically marked as private, in
-a similar manner to C/C++
-
-The whole idea is to compute the minimal nesting depth of a variable usage. To
-do so, a \emph{nesting depth} is first associated to each expression. It
-lexically corresponds to the indentation level of its instruction, to the
-notable exception of the iterator declaration in a for loop, that holds a
-nesting depth equals to 1 plus the loop declaration indentation level.
-
-Tracking the nesting depth of all variable use, for instance using def-use
-chains, makes it possible to build a list of nesting depths, one per variable
-reference. Computing the min of this list yields the scope of the variable.
-
-For instance, in the motivating example, the \texttt{nsteps} variable is first
-defined as argument, then read twice, which gives a def-use chain of
-\{DEF, USE, USE\}. The associated nesting depths is
-\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator then
-used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
-depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
-of \{DEF, USE\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
-automatically declared local to the loop, thus not needing to be listed as
-\texttt{private}.
-
-To be able to list a variable marked as local, for instance in a
-\texttt{lastprivate} clause, the directive accesses are taken into account in
-the def-use computations. For instance in
-Listing~\ref{lst:motivating-example-annotated-v0}, the chain for \texttt{x} is
-\{DEF, DEF, USE\} and the associated nested depths are \{1, 2, 2\} so its scope
-is 1 and the variable is not declared local to the loop.
-
-Thanks to this automatic scoping computation, it is possible to write shorter
-OpenMP directive, as illustrated in
-Listing~\ref{lst:motivating-example-annotated-v1}.
-
-\begin{lstlisting}[float, label={lst:motivating-example-annotated-v1}, caption={Motivating
-example: computing $\pi$ in Python with OpenMP and automatic scoping.}]
-def pi(nsteps):
-  sum, step = 0., 1. / nsteps
-  #omp parallel for reduction(+:sum)
-  for i in range(nsteps):
-    x = (i - 0.5) * step
-    sum += 4. / (1. + x**2)
-  return step * sum
-\end{lstlisting}
-
-\subsubsection{Parallel For and Iterators}
-
-The core directive to handle data parallelism is the \texttt{for} directive that
-distributes the iteration space of the associated loop among the existing
-threads. To be compatible with OpenMP, the loop iteration space must be
-described by a random access iterator with a total order. Integers used as loop
-indices in Fortran and C satisfy this conditions, as well as C++ iterators with
-the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
-step of one until it is exhausted, in which case it raises an exception: it
-behaves as a forward iterator. To be compatible with OpenMP, these iterators
-must be turned into random access iterators.
-
-The extension of Python iterator to random access iterators is direct for the
-standard containers: \texttt{list},\footnote{Python \texttt{list}s behave as C++
-\texttt{vector}s.} \texttt{set} or \texttt{dict}. Other iterators require more
-care.
-
-Generators, Python objects that behave like iterators, are commonly used in
-Python. The simplest one, \texttt{xrange(\emph{start}, \emph{stop},
-\emph{step})}, successively yields value starting from \texttt{\emph{start}} to
-\texttt{\emph{stop}} by a step of \texttt{\emph{step}}. It is easily extended to
-support random access, but it generally does not make sense to use a generator
-as a loop iterator, as the relation between two random states of the iterator
-may be of an arbitrary complexity.
-
-Generator expressions are generators whose content is built from an other
-iterator. For instance \texttt{(x*x for x in l)} successively yields the square
-of each element in \texttt{l}. They behave like adaptors: they apply a
-particular expression on each value of the iterator. It is also the case of the
-\texttt{enumerate} builtin, that yields each element of the enumerated iterator
-associated with its index. These generators are random access iterators only if their
-input iterator is a random iterator itself.
-
-Finally, if the iterator yields a tuple, it is possible to unpack it inside the
-\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
-all the unpacked variables are considered as iterators, especially with respect
-to default privatization rules: in the given example, \texttt{i} and \texttt{v}
-are private, and the parallel iteration is valid because the input of
-\texttt{enumerate} is a list, which allows random access iteration.
-
-\begin{lstlisting}[float, label={lst:complex-omp-for}, caption={Parallel
-    loop in Pythran with tuple unpacking.}]
-'omp parallel for'
-for i, v in enumerate([2, 3, 5, 7, 11]):
-    print i, ':', v
-\end{lstlisting}
-
-Pythran's runtime is aware of these three kind of iterators and supports
-parallel iteration over random access iterators and iterators adapting random
-ones.
-
-\subsection{OpenMP Runtime Library}
-
-OpenMP provides a small runtime library that, for instance, makes it possible to
-retrieve the active thread id. All the functions are declared in the
-\texttt{<omp.h>} header, and have a default behavior when OpenMP is not
-activated.
-
-Providing a binding to these library in Python as an \texttt{omp} module does
-not raise particular problems, as the signature of these functions only involves
-integers passed as parameters or return values, except for the mutex
-manipulation. In that case an opaque type is used to represent the native type.
-
-The \texttt{\_OPENMP} macro definition is always provided when OpenMP is
-activated, and can be used to detect when OpenMP is not available. Python does
-not have a preprocessor, but it is possible to catch the import exception if the
-\texttt{omp} module is not found. Listing~\ref{lst:sample-api-usage} showcase
-such a call using an example converted from the OpenMP validation suite
-presented in next section. The code starts a parallel section using the
-\texttt{parallel} directive, spawning several threads. Then each started thread
-increases the \texttt{nthreads} variable that has unspecified visibility, thus
-\texttt{shared}. The incrementation is protected by a guard through the
-\texttt{critical} directive. Then one of the thread retrieves the number of
-active thread in the parallel region through the \texttt{get\_num\_threads}
-function from the \texttt{omp} module. The function should always return
-\texttt{True}.
-
-\begin{lstlisting}[float,label={lst:sample-api-usage},caption={Example of
-        OpenMP API usage from Python.}]
-import omp
-def omp_get_num_threads():
-  nthreads, nthreads_lib = 0, -1
-  #omp parallel
-  if 1
-    #omp critical
-    nthreads += 1
-    #omp single
-    nthreads_lib = omp.get_num_threads()
-  return nthreads == nthreads_lib
-\end{lstlisting}
-
-\subsubsection{Interaction With Pythran Runtime}
-
-Pythran already uses OpenMP to parallelize some Python functions. For instance
-the \texttt{sum} function from the \texttt{\_\_builtin\_\_} is implemented using
-the OpenMP \texttt{reduction} clause. When it can prove that its first argument
-is a \emph{pure} function, it can also call a parallel version of the
-\texttt{map} function that internally spawns a parallel region. This leads to a
-well known situation with OpenMP, where too many threads may be spawned: nested
-parallel regions. OpenMP 4 provides several ways to handle this situation:
-
-\begin{enumerate}
-
-    \item Disabling nested parallelism, through the \texttt{OMP\_NESTED}
-        environment variable or the \texttt{omp\_set\_nested} routine.
-
-    \item Setting the maximum number of active levels, through the
-        \texttt{OMP\_MAX\_ACTIVE\_LEVELS} environment variable or the
-        \texttt{omp\_set\_max\_active\_levels} routine.
-
-    \item Conditionally activating regions or conditionally setting the number
-        of threads in a parallel region using the \texttt{omp\_in\_parallel}
-
-\end{enumerate}
-
-The decision is left to the user, depending on its application and OpenMP
-implementation.
-
-\subsection{Validation}
-
-A validation suite for OpenMP is proposed in~\cite{wang2012} for C and Fortran.
-We ported it to Python, and also extended it to validate the corner cases
-specific to Python described in this section. A typical test case is given in
-Listing~\ref{lst:openmp-validation}.
-
-\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of Python
-    OpenMP validation test case.}]
-def omp_parallel_for_if(loop_count):
-  import omp
-  using = sum = 0
-  'omp parallel for if(using == 1)'
-  for i in range(loop_count + 1):
-    num_threads = omp.get_num_threads()
-    sum += i
-  known_sum = (loop_count *
-              (loop_count + 1)) / 2
-  return known_sum == sum and num_threads == 1
-\end{lstlisting}
-
-We have used the Pythran tool described in the following section to turn each
-Python test function into a C++ function with the same directives and runtime
-calls. Apart from the \texttt{threadprivate} directive and the
-\texttt{collapse(\emph{n})} clause, all tests were successful.
-\texttt{threadprivate} directives were held by global variables not supported in
-Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
-loop nesting required by the \texttt{collapse} clause.
-
-%=========================================================
-\section{Validation}\label{sec:validation}
-%=========================================================
-
-In addition to the formal validation done using the OpenMP validation suite
-presented in Section~\ref{sec:python-openmp}, the performance of parallel module
-generated by Pythran from Python module with explicit parallelization using
-OpenMP directives has been studied on several situations: first on the
-motivating example, then on a synthetic geomatic application. Python and Cython
-versions are then compared and finally the effect of OpenMP directives on Pythran
-generated code are illustrated in the context of the Python Benchmarks
-suite.
-
-The target machine has  8 AMD Opteron 6176 SE cores and 8 GO of RAM. It is
-running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
-\texttt{g++} version 4.8.1 and all applications are compiled using the
-\texttt{-Ofast} flag. Pythran generated programs are linked with Boost.Python
-version 1.54. The Pythran version is extracted from the \texttt{sc2013} branch
-of the git repository \url{https://github.com/serge-sans-paille/pythran}. Python
-programs are run using the CPython implementation, version 2.7.5. Cython
-programs are generated using version 0.19.1. Numba and Parakeet are installed
-from the \texttt{master} branch of their respective repositories.
-
-\subsection{Motivating Example}
-
-To validate the approach proposed in this paper, let first consider the
-motivating example. Figure~\ref{fig:motivating-example-scale} shows how the C
-and Python version of the program scale when changing the number of core used.
-As a reference, the original Python code runs 41 times slower than the
-equivalent C code. As the program is not memory bound and exhibit trivial
-parallelism, the C + OpenMP version scales linearly. The Pythran + OpenMP
-version follows the same pattern.
-
-\begin{figure}
-
-    \caption{Comparison of the scaling of $\pi$ computation function for C and
-    Python with OpenMP, depending on the number of active threads.}
-    \label{fig:motivating-example-scale}
-
-    \includegraphics[width=.5\textwidth]{pi_omp_bench.pdf}
-
-\end{figure}
-
-\subsection{Hyantes: a Geomatic Application}
-
-To get closer to a realistic example, we have considered the code of a small
-geomatics application, Hyantes. Starting from the C code, we successively
-parallelized it with OpenMP, turned it into Python with a Pythran compatible
-kernel and parallelized the Python version using the approach described in this
-paper. We also measure the SLOC of the two versions.  Table~\ref{tbl:hyantes}
-summarizes the results of the experimentation. It not only emphasises the
-use of a high level language prototyping, but also shows that it is possible to
-turn this prototype into reasonably efficient code through Pythran. It is also
-possible to prototype the parallel version while remaining at the Python level.
-
-\begin{table}
-
-    \caption{Comparison of several versions of the Hyantes prototype.}
-    \label{tbl:hyantes}
-
-    \centering
-    \begin{tabular}{|l|r|}
-        \hline
-        Language & SLOC  \\
-        \hline
-        C       & 102   \\
-        Python  & 30    \\
-        Pythran & 30     \\
-        Pythran+OMP    & 31  \\
-        \hline
-    \end{tabular}
-
-\end{table}
-
-The Hyantes programs scales relatively well. Figure~\ref{fig:hyantes-speedup} shows
-the relative speedup of the application when increasing the number of OpenMP
-threads using up to 8 cores.
-
-\begin{figure}
-
-    \caption{Comparison of the scaling of the hyantes application for C and
-    Python with OpenMP, depending on the number of active threads.}
-    \label{fig:hyantes-speedup}
-
-    \includegraphics[width=.5\textwidth]{hyantes_omp_bench.pdf}
-
-\end{figure}
-
-\subsection{Comparison with Cython}
-
-We then compare Pythran with Cython. The approaches share some similarities:
-they both generate code written in a lower level language, with OpenMP
-directives. However, the inputs differ as Cython requires more typing
-information than Pythran to generate efficient code, and Cython is not
-backward-compatible with Python. They translate explicit fine-grained
-parallelism through code annotations or specific Python constructs,
-respectively.
-
-Pythran exposes the full OpenMP interface to the user, thus enabling both data
-and task parallelism, as described in Section~\ref{sec:python-openmp}. It is
-not the case in Cython:
-%
-\begin{itemize}
-
-    \item Only loops can be made parallel, using a new \texttt{prange} generator.
-
-    \item Reductions and variable privacy are inferred, but it chokes on
-        reduction on private variables.
-
-    \item The user is responsible from releasing the GIL inside parallel
-        regions.
-
-    \item It is impossible to use a function imported from a Python module in a
-        parallel region, but it is still possible to use native C functions.
-
-\end{itemize}
-%
-Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrate the
-difference between the two and illustrates the intrusive behavior of Cython.
-
-\begin{figure}
-
-    \begin{lstlisting}[label={lst:cython-sample}, caption={Cython implementation
-    of a parallel reduction.}]
-from libc.math cimport sqrt
-from cython.parallel import parallel, prange
-def sum_sqrt(double r):
-    cdef int i
-    with nogil, parallel():
-        for i in prange(10000000):
-            r += sqrt(i)
-    return r
-    \end{lstlisting}
-%
-    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran implementation
-    of a parallel reduction.}]
-#pythran export sum_sqrt(float)
-import math
-def sum_sqrt(r):
-    #omp parallel for reduction(+:r)
-    for i in xrange(10000000):
-        r += math.sqrt(i)
-    return r
-    \end{lstlisting}
-\end{figure}
-
-%Ce n'est pas ce code qui ne donne pas les mêmes résultats avec Cython et Python et Pythran?
-
-The performance of the two approaches is shown in
-Figure~\ref{fig:cython-pythran}.  The benchmarked codes are typical
-mathematical, image-processing or linear algebra kernels. All these kernels have
-been written in Cython and Python ---compatible with Pythran--- and annotated
-through the mechanism of each language to exhibit parallelism, then compiled
-into native code.  Their execution time when called from the Python interpreter
-is measured through the \texttt{timeit} module. All results are normalized
-against Cython sequential execution time. They show that while handling code at
-a higher level than Cython, Pythran achieves comparable results.
-
-The source codes used for these benchmarks are available on the Pythran
-repository.\footnote{\see the \texttt{sc2013} branch of the git repository.}
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{cython}
-    \caption{Comparison of Cython and Pythran generated code performance.}
-    \label{fig:cython-pythran}
-\end{figure}
-
-\subsection{Python Benchmarks Test Bed}
-
-The Python Benchmark test bed
-(\url{https://github.com/numfocus/python-benchmarks}) is an innitiative to
-gather relevant numerical python benchmarks and compare several implementations
-and compilers. The original benchmarks are slightly adapted to match each
-compiler restriction (e.g. lack of support for a particular construct), or
-completely rewritten in the target's paradigm (e.g. using PyOpenCL or Theano).
-We picked up the targets that did not imply a full rewrite of their input:
-Cython, Numba, Pythran and Parakeet, when available and supported.
-Additionnaly, the original Python code is run.
-
-The best execution time out of five run is shown for each code version. If the
-code does not compile with a particular compiler, the column is left blank. Two
-code versions are shown for Pythran: one compiler without the OpenMP flag set
-and one with the OpenMP flag set. The code itself is not modified, only the
-compilation flag. A logarithmic scale is used due to the generally poor
-performance of the pure Python version.
-
-Figure~\ref{fig:pb-julia} shows the execution time of the computation of the
-julia fractal. It's a highly parallel, computate-intensive code. Pythran and
-Cython's version have almost the same execution time, but activating OpenMP
-yields an extra $\times 3.95$ speedup when compared to Pythran alone.
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{julia}
-    \caption{Comparison of Various python Compilers on the Julia benchmark}
-    \label{fig:pb-julia}
-\end{figure}
-
-Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a distance matrix between two random vectors, using a parallel nested loop.
-On that example, Pythran slightly outperforms Cython, parakeet and Numba.
-Taking into account the OpenMP annotation that flags the outermost loop as parallel yields an $\times 4.27$ speedup to Pythran.
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{pairwise}
-    \caption{Comparison of Various python Compilers on the pairwise benchmark}
-    \label{fig:pb-pairwise}
-\end{figure}
-
-Figure~\ref{fig:pb-rosen} shows the execution time of the computation of the third derivative of the Rosenbrock function for a random array input, using Numpy's vector operations.
-There is no explicit parallelism in this function but the vector operation is implictly parallel.
-On this example, Cython is on par with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the Pythran version run $\times 1.78$ faster than sequential Pythran.
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{rosen_der}
-    \caption{Comparison of Various python Compilers on the rosen\_der benchmark}
-    \label{fig:pb-rosen}
-\end{figure}
-
-Figure~\ref{fig:pb-rosen} shows the execution time of the computation of one step of the GrowCut algorithm used in image processing.
-On this example, Parakeet performs better than Cython and Pytrhan and way better than Numba, but paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to Pythran that helps it outperform Parakeet.
-
-\begin{figure}[ht]
-    \includegraphics[width=.5\textwidth]{growcut}
-    \caption{Comparison of Various python Compilers on the Growcut benchmark}
-    \label{fig:pb-growcut}
-\end{figure}
-
-It appears that all Python compilers manage to yield significant speedups over
-the Python interepreter, each compiler managing some cases better than others.
-All the benchmarks exhibt some level of parallelism that are easy to capture
-using OpenMP without changing the original algorithm and while maintaining
-backward compatibility with the Python interpreter. Some even have a degree of
-implicit parallism that makes parallelization fully transparent to the user.
-Taking advantage of this parallelism appears to be the key to reach an additional performance level.
-
-%=========================================================
-\section{Conclusion and Future Work}
-%=========================================================
-
-This paper studies the addition of OpenMP annotations to Pythran. It shows that
-providing minor semantic adaptations, these annotations make it possible for
-Python code to benefit from multi-cores while retaining backward compatibility
-and without worrying about the Global Interpreter Lock.
-
-To achieve this goal, a translator from a subset of the Python language to C++
-has been designed. This translator turns regular Python modules annotated with
-OpenMP directives and a few type annotations into native parallel module. The
-input module remains compatible with the standard interpreter and the underlying
-runtime is compatible with parallel constructs.
-
-The approach is compared with Cython, an extension of the Python language used
-to generate native module with an hybrid Python-C syntax that also provides
-means to exhibit fine grained parallelism. It shows that retaining Python
-compatibility does not prevent the achievement of comparable performances.
-Comparisons to Numba and Parakeet shows the benefit of using parallelism on the
-benchmarks from the Python Benchmark project.
-
-Future work will focus on the extension of the approach to OpenMP 4 and the
-\texttt{target} clause that should make it possible to target MIC from Python.
-The \texttt{simd} clause also brings interesting vectorization capabilities that
-can be exposed at the python level.  There are also several implicit vectorization
-opportunities in Python, especially in the list comprehension construction, that
-needs to be explored.
-
-%=========================================================
-\section*{Acknowledgment}
-%=========================================================
-
-This work has received fundings from \textsc{silkan} and the French ANR through
-the CARP project. The authors thank Adrien \textsc{Merlini} and Alan
-\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
-B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Fran{\,c}ois
-\textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable reviews.
-
-\bibliographystyle{IEEEtranS}
 \bibliography{biblio}
@@
 \end{document}
-% vim:spell spelllang=en
-% vim:tw=80

### 2013-09-29 doc/papers/cc2014/paper.tex
First draft of the CC tool paper

@@
+%http://www.etaps.org/index.php/2014/cc
+%to be submtitted to https://www.easychair.org/account/signin.cgi?timeout=1;conf=cc2014
+% abstract: 4th october 2013
+% paper: 11th october 2013
+%
+%Submissions should consist of two parts:
+%
+%    The first part, at most 4 pages, should describe the tool presented.
+Please include the URL of the tool (if available) and provide information that illustrates the maturity and robustness of the tool (this part will be included in the proceedings).
+%        The second part, at most 6 pages, should explain how the demonstration will be carried out and what it will show, including screen dumps and examples.
+(This part will be not be included in the proceedings, but will be evaluated.)
+%
 \documentclass{llncs}
@@
 \usepackage{url}
+\usepackage{acronym}
 \usepackage{todo}
+\usepackage{paralist}
+\usepackage[english]{babel}
 
 \usepackage{tikz}
+\usetikzlibrary{backgrounds}
+\usetikzlibrary{calc}
 \usetikzlibrary{positioning}
@@
 \usetikzlibrary{shapes}
+\usetikzlibrary{fit}
 
@@
 
+\acrodef{AST}{Abstract Syntax Tree}
+\acrodef{API}{Application Programming Interface}
+\acrodef{CLI}{Command Line Interface}
+\acrodef{IR}{Internal Representation}
+\acrodef{JIT}{Just-In-Time}
+\acrodef{AOT}{Ahead-Of-Time}
+\acrodef{OO}{Object Oriented}
+\acrodef{OSS}{Open Source Software}
+
 \title{Pythran: Turning Python Modules into Optimized C++ Meta-Programs}
 
-\author{Serge Guelton\inst{1,2} \and Pierrick Brunet\inst{2} \and Mehdi Amini\inst{3}}
+\author{Serge Guelton\inst{1,2} \and Pierrick Brunet\inst{2}}
 
@@
 Télécom Bretagne, Plouzané, France
-\and
-SILKAN Inc., Los-Altos, USA
 }
@@
 \begin{abstract}
-...
+
+    Pythran is a compiler tool that turns source files written in a subset of
+    the Python language into C++11 meta-programs. It also extract from the
+    original source files type annotations used to instantiate the meta-programs
+    into a Python native module that runs typically faster than its interpreted
+    counterpart.
+
+    This paper presents the optimization techniques the Pythran compiler applies
+    to Python code: unboxing, deforestation, inter procedural constant folding,
+    explicit and implicit parallelization and implicit vectorization.
+
 \end{abstract}
@@
 \section{Introduction}
-...
-\section{Introduction}
+
+Python is a renowned programming language that supports multiple paradigms. It
+is highly dynamic (e.g.\ typing and name resolution are dynamic) at the expense
+of slower execution than equivalent programs written in static languages.
+
+Several tools exist to counterbalance the execution time limitations. They can
+be classified depending on
+\begin{inparaenum}[1)]
+\item the restriction they impose on their input and
+\item the compilation approach.
+\end{inparaenum}
+PyPy\cite{pypy2009} targets the whole Python
+language using a tracing \ac{JIT} approach, Nuitka\cite{nuitka2012} sets the same
+goal but using translation to C++ and \ac{AOT} compilation. Shed
+Skin\cite{shedskin2006} translates to C++ but only targets an implicitly statically
+typed subset of Python and several modules from the standard library. In the
+field of numerical computations, Parakeet\cite{parakeet2012} and
+Numba\cite{numba2013} are \ac{JIT} compilers based on LLVM that
+only supports the Python subset required to run numpy-centric programs.
+
+Pythran\cite{pythran2013} is an \ac{AOT} compiler that focuses on the
+optimization of the functional and procedural paradigms offered by Python,
+leaving aside dynamicity and most of \ac{OO} features.  Indeed, the Python
+language and its standard library offers comprehensive services to the
+programmers, but the language proves to be useful even with a limited subset of
+features:
+\begin{inparaenum}[1)]
+\item classical data types (\texttt{bool}eans, \texttt{int}egers,
+\texttt{long} integers, \texttt{float}ing point numbers, \texttt{str}ings) ;
+\item standard containers (\texttt{list}s, \texttt{set}s, \texttt{tuple}s,
+\texttt{dict}ionaries) ;
+\item a few features or function inherited from
+functional programming such as list/set/dict comprehension and \texttt{reduce},
+\texttt{map}, \texttt{filter} operators and
+\item the \texttt{numpy} module that
+provides a multi-dimensional array structure targeted at scientific computing. 
+\end{inparaenum}
+
+The paper is organized as follows: Section~\ref{sec:architecture} presents the
+architecture of the Pythran compiler, Section~\ref{sec:functionnal-optim}
+details some high-level optimizations related to the functional aspects of
+Python, while Section~\ref{sec:numeric-optim} focuses on optimizations for
+numerical computations.
+
+\section{Architecture of the Pythran Compiler}
+\label{sec:architecture}
+
+Pythran is a mature \ac{OSS} available through its source code
+repository (see \url{https://githib.com/serge-sans-paille/pythran}), the
+Python Package Index (see \url{https://pypi.python.org/pypi/pythran}) and
+a Debian repository (see \url{http://ridee.enstb.org/debian}). It can be
+used as a source-to-source compiler infrastructure for a subset of the Python
+language, a C++11 meta-programs generator or a native module generators. Each
+of this target corresponds to a layer in the tool architecture described in
+Figure~\ref{fig:archi}.
+
+The \ac{IR} used by Pythran is a strict subset of Python's \ac{AST}, simplified
+to ease the optimization process. Using a high-level \ac{IR} is well suited for
+high-level transformations, and provides source-to-source facilities for free.
+As the primary backend of Pythran is C++11, a high-level language, there is no
+need of low-level optimizations that are performed by the backend compiler.
+
+\begin{figure}[ht]
+    \centering
+    \begin{tikzpicture}[
+            file/.style={align=center, node distance=0.4cm},
+            state/.style={draw=black!50,fill=black!5,rectangle, align=center, node distance=0.4cm},
+            box/.style={draw=black!50,fill=black!5,rectangle, rounded corners, drop shadow, align=center,
+            node distance=0.4cm, inner sep=.2cm},
+            mbox/.style={draw=black!50,fill=black!5,rectangle, rounded corners, drop shadow, align=center,
+            node distance=0.4cm, inner sep=.2cm},
+        ]
+        \node[file] (python)                      {\textbf{Python Module [\texttt{.py}]}};
+        \node[state] (ast)       [below=of python]       {\ac{AST}};
+        \node[state] (ir)        [below=of ast]          {\ac{IR}};
+        \node[state] (ir2)       [below=of ir]           {Optimized \ac{IR}};
+        \node[state] (be0)       [xshift=1.5cm, below=of ir2]          {Python};
+        \node[state] (be1)       [xshift=-1.5cm,below=of ir2]          {C++};
+        \begin{pgfonlayer}{background} 
+            \node[box]  (pythran)   [fit=(ast) (ir) (be0) (be1)]  {};
+        \end{pgfonlayer}
+        \node[mbox] (cxx)        [yshift=-.1cm, below=of be1]          {\texttt{cxx}};
+        \node[file] (so)        [below=of cxx]          {\textbf{Native[\texttt{.so}]}};
+        \node[file] (py)        [yshift=-.1cm, below=of be0]          {\textbf{Python[\texttt{.py}]}};
+
+        \draw[very thick, ->] (python) -- (pythran);
+        \draw[very thick, ->] (pythran.south -| cxx.north) -- (cxx.north);
+        \draw[very thick, ->] (cxx) -- (so);
+        \draw[very thick, ->] (pythran.south -| py.north) -- (py.north);
+
+        \draw[thick, dotted, ->] (ast) -- (ir);
+        \draw[thick, dotted, ->] (ir) -- (ir2);
+        \draw[thick, dotted, ->] (ir2.south) -- ($(ir2.south) - (0, .25)$) -| (be0.north);
+        \draw[thick, dotted, ->] (ir2.south) -- ($(ir2.south) - (0, .25)$) -| (be1.north);
+
+    \end{tikzpicture}
+
+    \caption{Architecture of the Pythran compiler.}
+    \label{fig:archi}
+\end{figure}
+
+The compilation process used by Pythran follows several steps:
+\begin{inparaenum}[1)]
+\item\label{en:ast} Turn Python code into an \ac{AST} using the \texttt{ast}
+    module.
+\item Refine Python \ac{AST} into Pythran \ac{IR}. For instance this makes
+    \emph{destructuring} or \emph{list comprehension} explicit and removes the
+    corresponding constructs from the \ac{AST}. \emph{Partial functions} or
+    \emph{lambda functions} are also converted to top-level functions after
+    closure computations at this step.
+\item Perform several optimizations on the \ac{IR}, using the optimization
+    sequence provided by the user.
+\item Type the \ac{IR} using \emph{parametric types}.
+\item Dump the \ac{IR} into Python code or templated C++ code.
+\item Optionally use type annotations gathered during Step~\ref{en:ast} to
+    instantiate the C++ meta-functions and compile them to native code.
+\end{inparaenum}
+
+The compiler can be used from the \ac{CLI} as a regular Python-to-native-code
+compiler that takes a \texttt{.py} module and turns it into a \texttt{.so}
+native module. It can also be used as a compiler infrastructure, in which case
+the developer interacts with the various passes and analyse through the
+\texttt{PassManager} class.
+
+\section{Procedural and Functional Python Optimizations}
+\label{sec:functionnal-optim}
+
+Python compilers generally use \emph{unboxing}---the process of removing the
+object layer Python adds around any value--- as the primary way to obtain
+C-like performances. This is achieved either through the discovery of variable
+types at runtime or through static typing. Pythran, in a similar manner to
+Sched Skin, relies on static type inference, with the slight difference that it
+performs function-level parametric typing and relies on C++ template
+instantiantion to verify the code is correctly type. In the end, valid Pythran
+programs are implicitly statically typed and all variables are unboxed.
+
+Additionally, Pythran performs several optimizations inherited from functional
+programing to its \ac{IR}. It uses a variant of
+\emph{deforestation}~\cite{Wadler1988} to avoid usage of intermediate storage
+when combining some intrinsics. Practically, this relies on aliasing analysis
+to detect calls to the \texttt{map} (resp.\ \texttt{range}, \texttt{zip} or
+\texttt{filter}) functions and lazyness analysis to turn them into calls to the
+\texttt{imap} function from the \texttt{itertools} module (resp.\
+\texttt{xrange}, \texttt{izip} or \texttt{ifilter}). The transformation engine
+is aware of the semantics of several Python builtins, so that a call like
+\texttt{sum(map(sqrt, map(abs, l))} is turned into the equivalent---but more
+memory-efficient---\texttt{sum(imap(sqrt, imap(abs, l))}. During its \ac{AST}
+to \ac{IR} conversion, Pythran turns some list comprehension constructs into
+\texttt{map} calls, so list comprehension also benefit from the optimization:
+\texttt{sum([x*y for x,y in zip(l0, l1)]} is turned into \texttt{sum(imap(foo,
+izip(l0, l1)))}, where \texttt{foo} is an helper function generated on the fly.
+
+Using the introspection capability of Python, Pythran also performs
+inter-procedural constant folding. The optimization relies on an analysis that
+flags some functions as ``pure'' when they neither update their arguments not
+have global side effects (which include I/O and random generator state). When a
+pure function is called without arguments or with literal arguments, it can be
+replaced by its compile-time evaluation. Technically, as Pythran's\ac{IR} is a
+subset of Python's \ac{AST}, a simple call to the \texttt{eval} builtin yields
+the result to substitute.
+
+\section{Numeric Python Optimizations}
+\label{sec:numeric-optim}
+
+Thanks to the \texttt{scipy} and \texttt{numpy} packages, Python is also used
+for fast prototyping of numerical applications. Pythran
+implements part of the \acs{API} of the \texttt{numpy} package, with a focus on
+the multi-dimensional array point-to-point expressions. Given two arrays of
+similar shape \texttt{a} and \texttt{b}, it turns the expression
+\texttt{cos(a)*sin(2*b)} into a single parallelized and vectorized loop, taking
+advantage of multiple codes and vector instruction units if available. This is
+done at the C++ level using algorithmic patterns parallelized with OpenMP,
+vectorized through calls to \texttt{Boost.simd}~\cite{boostsimd2012} and
+combined with \emph{expression template}s.
+
+To parallelize explicit loops, Pythran makes it possible to use the OpenMP
+syntax at the Python level, as described in~\cite{pythranOMP2013}. Following
+the original OpenMP philosophy, OpenMP annotations are added to the sequential
+program to give it an additional parallel semantic, either loop-based or
+task-based. Annotations can be discarded by the compiler to keep the original
+sequential semantics.
+
+\section{Conclusion}
+
+The Pythran compiler offers a robust (the validation suite holds more than 1500
+test cases) tool to translate procedural Python modules into natives ones. It
+can also be used as a compiler infrastructure that offers high-level analysis
+like points-to analysis, global effects analysis etc.
+
+It is of interest for Python user that needs to prototype algorithms and make
+them run at C-speed without the burden of writing low-level code. It can also
+be used as a clean compiler infrastructure, using Python \ac{AST} as a
+common \ac{IR}.
 

### 2013-10-01 doc/papers/cc2014/paper.tex
First draft of the CC tool paper
Pierrick review

@@
 
-\institute{École Normale Supérieure, Département d'Informatique, Paris, France
+\institute{\'Ecole Normale Sup\'erieure, D\'epartement d'Informatique, Paris, France
 \and
-Télécom Bretagne, Plouzané, France
+T\'el\'ecom Bretagne, Plouzan\'e, France
 }
@@
 be classified depending on
+% PB : Inline enumeration looks strange for me but may be it is the english way to do
 \begin{inparaenum}[1)]
@@
 language using a tracing \ac{JIT} approach, Nuitka\cite{nuitka2012} sets the same
-goal but using translation to C++ and \ac{AOT} compilation. Shed
-Skin\cite{shedskin2006} translates to C++ but only targets an implicitly statically
+goal but using translation to C++ and \ac{AOT} compilation.
+Shed~Skin\cite{shedskin2006} translates to C++ but only targets an implicitly statically
 typed subset of Python and several modules from the standard library. In the
@@
 
+% we don't say we focus on scientifical python?
+% Ex : ...and most of \ac{OO} features to focus on scientific python modules
 Pythran\cite{pythran2013} is an \ac{AOT} compiler that focuses on the
@@
 \item classical data types (\texttt{bool}eans, \texttt{int}egers,
-\texttt{long} integers, \texttt{float}ing point numbers, \texttt{str}ings) ;
+\texttt{long} integers, \texttt{float}ing point numbers, \texttt{str}ings)~;
 \item standard containers (\texttt{list}s, \texttt{set}s, \texttt{tuple}s,
@@
 \item a few features or function inherited from
-functional programming such as list/set/dict comprehension and \texttt{reduce},
-\texttt{map}, \texttt{filter} operators and
+functional programming such as \texttt{list}/\texttt{set}/\texttt{dict}
+comprehension and \texttt{reduce}, \texttt{map}, \texttt{filter} operators and
 \item the \texttt{numpy} module that
 provides a multi-dimensional array structure targeted at scientific computing. 
+% "at scientific computing" looks strange for me but it may be my bad english...
 \end{inparaenum}
@@
 
+% mature?
 Pythran is a mature \ac{OSS} available through its source code
@@
 
+% It looks like pythran generate a *.py file as an output. What did you mean with
+% this arrow?
 \begin{figure}[ht]
@@
 The compilation process used by Pythran follows several steps:
+% In previous enumeration, all items are in a same sentence, here we have
+% a sentence for each item.
 \begin{inparaenum}[1)]
@@
     corresponding constructs from the \ac{AST}. \emph{Partial functions} or
+    %what is partial function in python? Is it sub-function
+    % (function declared in another function)?
     \emph{lambda functions} are also converted to top-level functions after
@@
     sequence provided by the user.
+    %after user optimisation, it applies pythran default optimisation, isn't it?
 \item Type the \ac{IR} using \emph{parametric types}.
@@
 
+% PB : --- are ugly for me :-) but it is only cosmetic
 Python compilers generally use \emph{unboxing}---the process of removing the
-object layer Python adds around any value--- as the primary way to obtain
+object layer Python adds around any value---as the primary way to obtain
 C-like performances. This is achieved either through the discovery of variable
@@
 programs are implicitly statically typed and all variables are unboxed.
+% PB : I don't understand where pythran unbox values.
 
@@
 \texttt{filter}) functions and lazyness analysis to turn them into calls to the
+% and pure function detection?
 \texttt{imap} function from the \texttt{itertools} module (resp.\
@@
 \texttt{sum(map(sqrt, map(abs, l))} is turned into the equivalent---but more
+% PB : explains why it is more efficient?
 memory-efficient---\texttt{sum(imap(sqrt, imap(abs, l))}. During its \ac{AST}
@@
 
-Using the introspection capability of Python, Pythran also performs
+Using the introspection ability of Python, Pythran also performs
 inter-procedural constant folding. The optimization relies on an analysis that
@@
 \texttt{cos(a)*sin(2*b)} into a single parallelized and vectorized loop, taking
-advantage of multiple codes and vector instruction units if available. This is
+advantage of multiple cores and vector instruction units if available. This is
 done at the C++ level using algorithmic patterns parallelized with OpenMP,

### 2013-10-02 doc/papers/cc2014/tool.tex
Update CC paper: tool part

%http://www.etaps.org/index.php/2014/cc
%to be submtitted to https://www.easychair.org/account/signin.cgi?timeout=1;conf=cc2014
% abstract: 4th october 2013
% paper: 11th october 2013
%
%Submissions should consist of two parts:
%
%    The first part, at most 4 pages, should describe the tool presented. Please include the URL of the tool (if available) and provide information that illustrates the maturity and robustness of the tool (this part will be included in the proceedings).
%        The second part, at most 6 pages, should explain how the demonstration will be carried out and what it will show, including screen dumps and examples. (This part will be not be included in the proceedings, but will be evaluated.)
%
\documentclass{llncs}

\usepackage{listings}
\usepackage{url}
\usepackage{acronym}
\usepackage{todo}
\usepackage{paralist}
\usepackage[english]{babel}

\usepackage{tikz}
\usetikzlibrary{backgrounds}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}
\usetikzlibrary{fit}

\bibliographystyle{splncs}
\lstset{language=python,morekeywords={cimport, cdef, with},
basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray},
captionpos=b}

\acrodef{AST}{Abstract Syntax Tree}
\acrodef{API}{Application Programming Interface}
\acrodef{CLI}{Command Line Interface}
\acrodef{IR}{Internal Representation}
\acrodef{JIT}{Just-In-Time}
\acrodef{AOT}{Ahead-Of-Time}
\acrodef{OO}{Object Oriented}
\acrodef{OSS}{Open Source Software}

\begin{document}
\setcounter{page}{6}

%\maketitle

%\begin{abstract}
%
%    Pythran is a compiler tool that turns source files written in a subset of
%    the Python language into C++11 meta-programs. It also extract from the
%    original source files type annotations used to instantiate the meta-programs
%    into a Python native module that runs typically faster than its interpreted
%    counterpart.
%
%    This paper illustrates the demonstration of Pythran.
%
%\end{abstract}

\section{Tool Paper Motivation}

The purpose of the tool demonstration is to showcase the core functionalities
of the Pythran compiler infrastructure, for three kind of audience:
\begin{inparaenum} \item Python users who want to benefit from the execution
        speed of a compiled language while keeping the advantages of a dynamic,
        high-level language, or who want to prototype C++ code in Python and
        include it in a C++ codebase.  \item Python compiler
        researcher/developers who want to share analyse and passes in a common
    compiler infrastructure.  \item Dynamic language researcher who want to
        explore new or existing optimizations in the context of Python, or want
        to transfer concepts from one language to another.  \end{inparaenum}

To achieve these goal, the presentation relies on \begin{inparaenum} \item User
    stories that illustrates the basic \ac{CLI} of the Pythran compiler,
(\S\ref{sec:cli}).  \item Code snippets related to analyse and passes
    implemented in Pythran (\S\ref{sec:code-analyse},
\S\ref{sec:code-transformation}). \item A sample implementation of parallel map
detection in the Pythran compiler infrastructure (\S\ref{sec:pmap}).  \item
Benchmarks from the \emph{Python Benchmark} codebase that compare several
Python compiler in the context of numerical applications
(\S\ref{sec:python-benchmarks}). \end{inparaenum}

In order to demonstrate the robustness of the tool, the whole presentation
relies on IPython Notebooks, a modern tool that makes it possible to
run code snippets from within the presentation, a feature that helps building
up the confidence in the tool itself. Figure~\ref{fig:notebook} illustrates the
output of a notebook. Input boxes contain code snippet that can be executed
during the presentation, and their output is directly integrated in the
document. This offers a good trade off between interactivity (the audience sees
the code running but the speaker does not to type anything) and reproducibility
(the whole presentation can be executed and check as a test, avoiding the
``demonstration effect'').

\begin{figure}
    \includegraphics[width=\textwidth]{pythran-notebook}
    \caption{Example of IPython Notebook.}
    \label{fig:notebook}
\end{figure}

%\section*{Conventions}
%
%Python code run from the interpreter is preceded by a triple chevron, but console output is not, as in:
%
%\begin{lstlisting}
%>>> print 7 * 3 * 2
%42
%\end{lstlisting}
%
%Shell code run from the interpreter is preceded by a single chevron, as in:
%\begin{lstlisting}
%> echo graphie
%graphie
%\end{lstlisting}

\section*{Installing Pythran}
\label{sec:install}

Pythran is relatively easy to install. One can rely on the prepackaged version in the Python Package Index, and run:

\begin{lstlisting}
> pip install pythran
\end{lstlisting}

A Debian package is also available, so after a slight modification of \texttt{sources.list} (see \url{http://ridee.enstb.org/debian/info.html}):

\begin{lstlisting}
> sudo apt-get install pythran
\end{lstlisting}

Compilation from sources starts is the best way to get last modifications:

\begin{lstlisting}
> git clone https://github.com/serge-sans-paille/pythran
> cd pythran && python setup.py install
\end{lstlisting}

Pythran-generated binaries depend on \texttt{libboost-python} for interfacing
Python with C++, \texttt{libgmpxx} for multi precision arithmetic, and a modern
C++11 compiler (clang version 3.4 or g++ version 4.8).

\section{Pythran \acf{CLI}}
\label{sec:cli}

Pythran is a source-to-source compiler that turns Python code into C++ meta
programs then instantiate it for given types and compile it to a native module
that can be imported using Python native module import.

\subsection{Python to C++ Conversion}

Let us consider a very simple module, \texttt{hello.py}:

\begin{lstlisting}
def hello(name):
    print 'hello', name
\end{lstlisting}

Pythran can turn it into a pure C++ source file:

\begin{lstlisting}
> pythran -e hello.py -o hello.hpp
> cat hello.hpp
#include <pythran/pythran.h>
namespace __pythran_hello {
  struct hello {
    typedef void callable;
    template <typename argument_type0 >
    struct type {
        typedef typename assignable<
          typename std::remove_cv<
            typename std::remove_reference<
              decltype(__builtin__::None)
            >::type
          >::type
        >::type result_type;
    }; 
    template <typename argument_type0 >
    typename type<argument_type0>::result_type
    operator()(argument_type0 const &) const;
  };
  template <typename argument_type0 >
  typename hello::type<argument_type0>::result_type
  hello::operator()(argument_type0 const & name) const {
    print(core::string("hello"), name);
    return __builtin__::None;
  }
}
\end{lstlisting}

Generated header code contains two parts: \begin{inparaenum} \item A meta class
\texttt{hello} that reflects the Python function, \item a templated nested
class \texttt{type} that contains the type inference logic. \end{inparaenum}

The header can be instantiated from regular C++ code, and template argument
detection emulates Python duck typing:

\begin{lstlisting}
> cat hello.cpp
#include "hello.hpp"
using namespace __pythran_hello;
int main() {
  hello h;
  h("world"); h(42);
  return 0;
}
> cxx -std=c++11 -I... hello.cpp -o hello
> ./hello
hello world
hello 42
\end{lstlisting}

\subsection{Python to Native Module Conversion}

To turn a Python module into a native module, Pythran needs type information
concerning the module interface. These type informations are provided by the
user in the form of comments, as in:

\begin{lstlisting}
#pytrhan export hello(str)
#pytrhan export hello(int)
def hello(name):
    print 'hello', name
\end{lstlisting}

These type annotations are used by the Pythran compiler to automatically
instanciate the generated C++ meta program and generate the Python foreign
interface glue. The associated code can be inspected using:

\begin{lstlisting}
> pythran -E hello.py
\end{lstlisting}

\noindent or directly compiled using:

\begin{lstlisting}
> pythran -O2 -g hello.py
\end{lstlisting}

Note that any compiler flags not relevant to Pythran is forwarded to the C++
compiler. the defaults are parametrized through a user configuration file
\texttt{.pythranrc} and its system-wide counterpart.

\section{Python Code Analysis in Pythran}
\label{sec:code-analyse}

Pythran does not only perform Python-to-C++ translation. It also applies
several high-level optimizations backed up by several code analyse.

One such analysis is the argument effect analysis. Informally, the analyse
traverses the \ac{IR} in a per function basis, building intra-function
argument effects as well as an inter-function dependency graph, then performs a
flooding pass over the dependency graph to compute a inter-procedural argument
effects. One can observe the result of this analyse using the \texttt{pythran} module:

\begin{lstlisting}
>>> from pythran import passmanager, analysis
>>> pm = passmanager.PassManager("tutorial_module")
>>> code = '''
def foo(l,a):
  l+=[a]
def bar(g):
  foo(g, 1)'''
>>> tree = ast.parse(code)
>>> ae = pm.gather(analysis.ArgumentEffects, tree)
>>> foo, bar = tree.body
>>> ae[foo]
[True, False]
>>> ae[bar]
[True]
\end{lstlisting}

Pythran also implements an intra-procedural aliasing analysis that provides
critical piece of informations:

\begin{lstlisting}
>>> code = '''
def foo(c, d):
    b= c or d
    return b'''
>>> tree = ast.parse(code)
>>> al = pm.gather(analysis.Aliases, tree)
>>> returned = tree.body[-1].body[-1].value
>>> print ast.dump(returned)
Name(id='b', ctx=Load())
>>> sorted(a.id for a in al[returned].aliases)
['c', 'd']
\end{lstlisting}

\section{Python Code Transformation in Pythran}
\label{sec:code-transformation}

Pythran provides two kind of passes: passes that transform Python's \ac{AST}
into Pythran \ac{IR}, and optimizations passes. An example of \ac{AST} to
\ac{IR} transformation is \emph{destructuring}

\begin{lstlisting}
>>> from pythran import passes
>>> tree = ast.parse("a, b = 1, 3.5")
>>> pm.apply(passes.NormalizeTuples, tree)
>>> print pm.dump(backend.Python, tree)
if 1:
    __tuple10 = (1, 3.5)
    a = __tuple10[0]
    b = __tuple10[1]
\end{lstlisting}

Among available optimizations, Pythran can use a variant of \emph{deforestation}, as showcased in the following snippet:
\begin{lstlisting}
>>> code = '''
def norm(l):
  return sum(n*n for n in l)'''
>>> tree = ast.parse(code)
>>> pm.apply(optim.GenExpToImap, tree)
>>> print pm.dump(backend.Python, tree)
import itertools
def norm(l):
    return sum(itertools.imap((lambda n: (n * n)), l))
\end{lstlisting}

\section{Live Exercise: Implement Parallel Map Detection in Pythran}
\label{sec:pmap}

Pythran makes it fairly easy to add new analyse that depend from existing
analyse. Let us take the example of parallel map detection. First step consist
in declaring a new analysis, that works at the function level:

\begin{lstlisting}
class ParallelMaps(ModuleAnalysis):
\end{lstlisting}

Then one need to list the analyse this analysis depends on:
\texttt{PureFunctions} to detect pure functions and \texttt{Aliases} to use
alaising informations:

\begin{lstlisting}
  def __init__(self):
    self.result = set()
    super(ParallelMaps, self).__init__(PureFunctions, Aliases)
\end{lstlisting}

The \texttt{result} member variable is used to store the result of the
analysis. Next one need to inspect each function call, select those calling the
\texttt{map} builtin and verify the first argument always aliases to a pure
function:

\begin{lstlisting}
  def visit_Call(self, node):
    if all(alias == modules['__builtin__']['map']
           for alias in self.aliases[node.func].aliases):
      if all(self.pure_functions.__contains__(f)
             for f in self.aliases[node.args[0]].aliases):
        self.result.add(node)
\end{lstlisting}

Matching calls are stored as the result of the analyse.

\section{Python Benchmarks Project}
\label{sec:python-benchmarks}

Python Benchmarks is a project that aims at comparing the existing approaches to
speedup numerical computations in Python. The project and its result can be
found on \url{https://github.com/numfocus/python-benchmarks}.
Figure~\ref{fig:python-benchmarks} illustrates the result of one of the
benchmark, showing that Pythran performs well compared to existing approaches.

\begin{figure}

    \includegraphics[width=\textwidth]{rosen_der_logscale}
    \caption{Timings for the \texttt{rosen\_der} function from Python Benchmarks.}
    \label{fig:python-benchmarks}

\end{figure}

\section{Conclusion}

The Pythran compiler is an open source project that offers a compilation
infrastructure to analyze and optimize a large class of Python programs.
Benchmarks shows it can produce C++ code whose performance match those of
annotated or JIT compiled code, and it is mature enough to host third part
contributions.

\end{document}

### 2013-10-02 doc/papers/cc2014/paper.tex
Take Into account Pierrick's remark

@@
 
-% we don't say we focus on scientifical python?
-% Ex : ...and most of \ac{OO} features to focus on scientific python modules
 Pythran\cite{pythran2013} is an \ac{AOT} compiler that focuses on the
 optimization of the functional and procedural paradigms offered by Python,
-leaving aside dynamicity and most of \ac{OO} features.  Indeed, the Python
+leaving aside dynamicity and most \ac{OO} features to focus on scientific python modules.  Indeed, the Python
 language and its standard library offers comprehensive services to the
@@
 % mature?
-Pythran is a mature \ac{OSS} available through its source code
+% yes, mature! 1500+ validation test looks mature enough to me ;-)
+Pythran is an \ac{OSS} available through its source code
 repository (see \url{https://githib.com/serge-sans-paille/pythran}), the
@@
 % this arrow?
+% SG: Pythran can generate a.py file if using the Python backend
 \begin{figure}[ht]

### 2013-10-02 doc/papers/cc2014/paper.tex
Slightly update tool paper

@@
 
-    This paper presents the optimization techniques the Pythran compiler applies
-    to Python code: unboxing, deforestation, inter procedural constant folding,
-    explicit and implicit parallelization and implicit vectorization.
+    This tool paper presents the optimization techniques the Pythran compiler
+    applies to Python code: unboxing, deforestation, inter procedural constant
+    folding, explicit and implicit parallelization and implicit vectorization.
 

### 2013-10-02 doc/papers/cc2014/tool.tex
Slightly update tool paper

@@
 
-\bibliographystyle{splncs}
 \lstset{language=python,morekeywords={cimport, cdef, with},
@@
 
-\section{Tool Paper Motivation}
+\section*{Tool Paper Motivation}
 
@@
 \begin{figure}
+    \vspace{-1em}
     \includegraphics[width=\textwidth]{pythran-notebook}
@@
     \label{fig:notebook}
+    \vspace{-1em}
 \end{figure}
@@
 [True, False]
->>> ae[bar]
-[True]
 \end{lstlisting}
@@
 def foo(c, d):
-    b= c or d
-    return b'''
+    b= c or d ; return b'''
 >>> tree = ast.parse(code)
@@
 
-Python Benchmarks is a project that aims at comparing the existing approaches to
-speedup numerical computations in Python. The project and its result can be
-found on \url{https://github.com/numfocus/python-benchmarks}.
+Python Benchmarks is a project that aims at comparing the existing approaches
+to speedup numerical computations in Python. Any project can contribute to new
+motivated benchmarks, resulting in more than five different tools
+beingrepresented. The project and its result are hosted at
+\url{https://github.com/numfocus/python-benchmarks}.
+
 Figure~\ref{fig:python-benchmarks} illustrates the result of one of the
 benchmark, showing that Pythran performs well compared to existing approaches.
+Several other benchmarks are available and are to be shown during the
+presentation.
 
-\begin{figure}
-
-    \includegraphics[width=\textwidth]{rosen_der_logscale}
+\begin{figure}[ht]
+    \vspace{-2em}
+    \hspace{-.05\textwidth}
+    \includegraphics[width=1.1\textwidth]{rosen_der_logscale}
     \caption{Timings for the \texttt{rosen\_der} function from Python Benchmarks.}
     \label{fig:python-benchmarks}
+    \vspace{-2em}
 

### 2013-10-02 doc/papers/cc2014/tool.tex

@@
 
-To achieve these goal, the presentation relies on \begin{inparaenum} \item User
+To achieve these goals, the presentation relies on \begin{inparaenum} \item User
     stories that illustrates the basic \ac{CLI} of the Pythran compiler,

### 2013-10-26 doc/papers/sc2013/paper.tex
SC2013: typo (make possible => make it possible)

@@
 computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
-which provides a set of scientific and numeric tools, ranging from linear algebra to ordinary differential equation solvers or generic algorithms.
+which provides a set of scientific and numeric tools, ranging from linear
+algebra to ordinary differential equation solvers or generic algorithms.
 
@@
 
-To allow the user to write native functions without having to write the boilerplate glue code that turns Python object into native structures back and
-forth, the SciPy package provides the \texttt{weave} module. It makes
+To allow the user to write native functions without having to write the
+boilerplate glue code that turns Python object into native structures back and
+forth, the SciPy package provides the \texttt{weave} module. It makes it
 possible to bundle C code snippets into Python code. They are then compiled and
@@
 
-The standard \texttt{threading} module also makes possible to start several
+The standard \texttt{threading} module also makes it possible to start several
 light-weight threads within the same interpreter, but this lower-level approach
@@
 This article proposes to combine OpenMP~\cite{openmp3.1} parallel annotations
-with the Python subset supported by the Pythran compiler to make possible to
+with the Python subset supported by the Pythran compiler to make it possible to
 write fine-grain parallel applications in Python, while being fully backward
@@
 
-Pythran handles the problem by refusing recursive types, which makes possible
+Pythran handles the problem by refusing recursive types, which makes it possible
 to solely rely on reference counting for memory management. It can be
@@
 Tracking the nesting depth of all variable use, for instance using def-use % I don't see the need for def-use chains for this! Unless you include some "unaliasing algo"
-chains, makes possible to build a list of nesting depths, one per variable
+chains, makes it possible to build a list of nesting depths, one per variable
 reference. Computing the min of this list yields the scope of the variable.
@@
 
-OpenMP provides a small runtime library that, for instance, makes possible to
+OpenMP provides a small runtime library that, for instance, makes it possible to
 retrieve the active thread id. All the functions are declared in the

### 2013-10-26 doc/papers/sc2013/paper.tex
SC2013: limit line length

@@
     \IEEEauthorblockA{\newline\IEEEauthorrefmark{1}
-        {\'E}cole Normale Sup{\'e}rieure, D\'{e}partement d'Informatique, Paris, France}
+        {\'E}cole Normale Sup{\'e}rieure, D\'{e}partement d'Informatique, Paris,
+        France}
     \IEEEauthorblockA{\newline\IEEEauthorrefmark{2}
@@
 
-    High Performance Computing users traditionally rely on low-level,
-    compiled language such as C or
-    FORTRAN to perform compute-intensive tasks. As a consequence, it is a
-    common situation to have High Performance Computing application written in a high-level language
-    such as Python, calling native routines for compute-intensive tasks. To
-    improve development speed and reduce maintenance costs, using a higher-level
-    language like Python seems attractive. While it is usually associated with
-    low performance, several solutions such as Cython, Numba, Parakeet or
-    Pythran offer to automatically or semi-automatically turn Python functions
-    into native ones.
+    High Performance Computing users traditionally rely on low-level, compiled
+    language such as C or FORTRAN to perform compute-intensive tasks. As a
+    consequence, it is a common situation to have High Performance Computing
+    application written in a high-level language such as Python, calling native
+    routines for compute-intensive tasks. To improve development speed and
+    reduce maintenance costs, using a higher-level language like Python seems
+    attractive. While it is usually associated with low performance, several
+    solutions such as Cython, Numba, Parakeet or Pythran offer to automatically
+    or semi-automatically turn Python functions into native ones.
 
@@
 services, thanks to its dynamicity, expressiveness, rich ecosystem and
-% Mehdi : battery-included ??????
 ``batteries included'' standard library. It is also widely used in scientific
@@
 It is now common for scripting language to have C bindings. To take advantage of
-compiled code, and to overcome the GIL limitations, Python developers has to write
-parallel C/C++ functions and the associated boilerplate based on the Python C
-API~\cite{pythoncapi}. Tools have been developed to relieve the user from this
-cumbersome task, notably SWIG~\cite{swig2003} that relies on an interface specification supplied by the programmer to generate the glue, or
+compiled code, and to overcome the GIL limitations, Python developers have to
+write parallel C/C++ functions and the associated boilerplate based on the
+Python C API~\cite{pythoncapi}. Tools have been developed to relieve the user
+from this cumbersome task, notably SWIG~\cite{swig2003} that relies on an
+interface specification supplied by the programmer to generate the glue, or
 \texttt{boost::python}~\cite{boostpython2007} that relies on C++ templates and
@@
 Python variables to improve performance. Without type annotations, the
-performance improvement is not terrific, but given enough type annotation, Cython can generates code that runs as fast as its C equivalent, while
+performance improvement is not terrific, but given enough type annotation,
+Cython can generates code that runs as fast as its C equivalent, while
 maintaining a syntax close to Python.
@@
 intrinsics to exhibit parallelism. But it can be efficiently compiled to either
-CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
+CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
 \texttt{http://thrust.github.com/}} library. Python decorators are used to
@@
 
-To be completely backward compatible, it has to be possible to run the input code
-in an environment that is not aware of the existence of the parallelizing
+To be completely backward compatible, it has to be possible to run the input
+code in an environment that is not aware of the existence of the parallelizing
 solution. This principle is not respected by parallel libraries, but code
@@
 helps removing most dynamic behavior, and is combined with a high-level runtime
-library ---\texttt{pythonic++}--- to allow a one-to-one mapping between Python and
-C++ constructs. The purpose of this section is not to dig into the internals of
-the compiler, but rather to focus on the impact of the parallelism layer,
+library ---\texttt{pythonic++}--- to allow a one-to-one mapping between Python
+and C++ constructs. The purpose of this section is not to dig into the internals
+of the compiler, but rather to focus on the impact of the parallelism layer,
 especially on the runtime library.
@@
 \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
-\node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
+\node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module
+[\texttt{.so}]}};
 
@@
 to solely rely on reference counting for memory management. It can be
-implemented through the thread-safe shared pointer mechanism provided by the C++11
-standard library.
+implemented through the thread-safe shared pointer mechanism provided by the
+C++11 standard library.
 
@@
 
-OpenMP directives are held by C/C++ \texttt{\#pragma}, or by Fortran comments. Most of them
-apply to structured blocks in C/C++ and delimited by comments in Fortran.
-A few directives (e.g.\ \texttt{threadprivate}) are not
-attached to a specific instruction.
+OpenMP directives are held by C/C++ \texttt{\#pragma}, or by Fortran comments.
+Most of them apply to structured blocks in C/C++ and delimited by comments in
+Fortran. A few directives (e.g.\ \texttt{threadprivate}) are not attached to a
+specific instruction.
 
@@
 Many OpenMP annotations are parametrized by clauses that list variables,
-specifying their behavior with respect to parallel regions, e.g.\
-\texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
+specifying their behavior with respect to parallel regions,
+e.g.\ \texttt{private}, \texttt{shared}, \texttt{copyin}. They can only refer to
 variables that have already been declared. However, there is no variable
@@
 
-\begin{lstlisting}[float, label={lst:motivating-example-annotated-v0}, caption={Motivating
-example: computing $\pi$ in Python with OpenMP.}]
+\begin{lstlisting}[float, label={lst:motivating-example-annotated-v0},
+  caption={Motivating example: computing $\pi$ in Python with OpenMP.}]
 def pi(nsteps):
@@
 \{DEF, USE, USE\}. The associated nesting depths is
-\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator then
-used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
-depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain
-of \{DEF, USE\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
+\{0, 1, 1\} so the variable scope is 0. \texttt{i} is defined as a loop iterator
+then used in the loop body, which gives a chain of \{DEF, USE\}, with nesting
+depths of \{2, 2\} and a scope of 2. Likewise \texttt{x} has a chain of
+\{DEF, USE\}, nesting depths of \{2, 2\} and a scope of 2, which means it can be
 automatically declared local to the loop, thus not needing to be listed as
@@
 
-\begin{lstlisting}[float, label={lst:motivating-example-annotated-v1}, caption={Motivating
-example: computing $\pi$ in Python with OpenMP and automatic scoping.}]
+\begin{lstlisting}[float, label={lst:motivating-example-annotated-v1},
+  caption={Motivating example: computing $\pi$ in Python with OpenMP and
+  automatic scoping.}]
 def pi(nsteps):
@@
 \texttt{enumerate} builtin, that yields each element of the enumerated iterator
-associated with its index. These generators are random access iterators only if their
-input iterator is a random iterator itself.
+associated with its index. These generators are random access iterators only if
+their input iterator is a random iterator itself.
 
 Finally, if the iterator yields a tuple, it is possible to unpack it inside the
-\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that case
-all the unpacked variables are considered as iterators, especially with respect
-to default privatization rules: in the given example, \texttt{i} and \texttt{v}
-are private, and the parallel iteration is valid because the input of
+\texttt{for} construct, as shown in Listing~\ref{lst:complex-omp-for}. In that
+case all the unpacked variables are considered as iterators, especially with
+respect to default privatization rules: in the given example, \texttt{i} and
+\texttt{v} are private, and the parallel iteration is valid because the input of
 \texttt{enumerate} is a list, which allows random access iteration.
@@
 
-\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of Python
-    OpenMP validation test case.}]
+\begin{lstlisting}[float,label={lst:openmp-validation},caption={Example of
+  Python OpenMP validation test case.}]
 def omp_parallel_for_if(loop_count):
@@
 \texttt{threadprivate} directives were held by global variables not supported in
-Pythran yet~; and the C++ code generated by Pythran does not preserve the perfect
-loop nesting required by the \texttt{collapse} clause.
+Pythran yet~; and the C++ code generated by Pythran does not preserve the
+perfect loop nesting required by the \texttt{collapse} clause.
 
@@
 
-In addition to the experimental validation done using the OpenMP validation suite
-presented in Section~\ref{sec:python-openmp}, the performance of parallel module
-generated by Pythran from Python module with explicit parallelization using
-OpenMP directives has been studied on several situations: first on the
+In addition to the experimental validation done using the OpenMP validation
+suite presented in Section~\ref{sec:python-openmp}, the performance of parallel
+module generated by Pythran from Python module with explicit parallelization
+using OpenMP directives has been studied on several situations: first on the
 motivating example, then on a synthetic geomatic application. Python and Cython
-versions are then compared and finally the effect of OpenMP directives on Pythran
-generated code are illustrated in the context of the Python Benchmarks
+versions are then compared and finally the effect of OpenMP directives on
+Pythran generated code are illustrated in the context of the Python Benchmarks
 suite.
@@
 
-The Hyantes programs scales relatively well. Figure~\ref{fig:hyantes-speedup} shows
-the relative speedup of the application when increasing the number of OpenMP
-threads using up to 8 cores.
+The Hyantes programs scales relatively well. Figure~\ref{fig:hyantes-speedup}
+shows the relative speedup of the application when increasing the number of
+OpenMP threads using up to 8 cores.
 
@@
 %
-    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran implementation
-    of a parallel reduction.}]
+    \begin{lstlisting}[label={lst:pythran-sample}, caption={Pythran
+      implementation of a parallel reduction.}]
 #pythran export sum_sqrt(float)
@@
 
-Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a distance matrix between two random vectors, using a parallel nested loop.
-On that example, Pythran slightly outperforms Cython, parakeet and Numba.
-Taking into account the OpenMP annotation that flags the outermost loop as parallel yields an $\times 4.27$ speedup to Pythran.
+Figure~\ref{fig:pb-pairwise} shows the execution time of the computation of a
+distance matrix between two random vectors, using a parallel nested loop. On
+that example, Pythran slightly outperforms Cython, parakeet and Numba. Taking
+into account the OpenMP annotation that flags the outermost loop as parallel
+yields an $\times 4.27$ speedup to Pythran.
 
@@
 
-Figure~\ref{fig:pb-rosen} shows the execution time of the computation of the third derivative of the Rosenbrock function for a random array input, using Numpy's vector operations.
-There is no explicit parallelism in this function but the vector operation is implictly parallel.
-On this example, Cython is on par with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the Pythran version run $\times 1.78$ faster than sequential Pythran.
+Figure~\ref{fig:pb-rosen} shows the execution time of the computation of the
+third derivative of the Rosenbrock function for a random array input, using
+Numpy's vector operations. There is no explicit parallelism in this function but
+the vector operation is implictly parallel. On this example, Cython is on par
+with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the
+Pythran version run $\times 1.78$ faster than sequential Pythran.
 
@@
 
-Figure~\ref{fig:pb-rosen} shows the execution time of the computation of one step of the GrowCut algorithm used in image processing.
-On this example, Parakeet performs better than Cython and Pytrhan and way better than Numba, but paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to Pythran that helps it outperform Parakeet.
+Figure~\ref{fig:pb-rosen} shows the execution time of the computation of one
+step of the GrowCut algorithm used in image processing. On this example,
+Parakeet performs better than Cython and Pytrhan and way better than Numba, but
+paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
+Pythran that helps it outperform Parakeet.
 
@@
 implicit parallism that makes parallelization fully transparent to the user.
-Taking advantage of this parallelism appears to be the key to reach an additional performance level.
+Taking advantage of this parallelism appears to be the key to reach an
+additional performance level.
 
@@
 The \texttt{simd} clause also brings interesting vectorization capabilities that
-could be exposed at the python level. There are also several implicit vectorization
-opportunities in Python, especially in the list comprehension construction, that
-need to be explored.
+could be exposed at the python level. There are also several implicit
+vectorization opportunities in Python, especially in the list comprehension
+construction, that need to be explored.
 

### 2013-10-26 doc/papers/sc2013/paper.tex
SC2013: takes into account reviewer 1 comments

@@
 with OpenMP and a few function type annotations, into C++ code.
+
 Figure~\ref{fig:pythran-compiler} summarizes its processing: type inference
 helps removing most dynamic behavior, and is combined with a high-level runtime
-library ---\texttt{pythonic++}--- to allow a one-to-one mapping between Python
-and C++ constructs. The purpose of this section is not to dig into the internals
+library ---\texttt{Pythonic++}--- to allow a one-to-one mapping between Python
+and C++ constructs. Pythonic++ is a template library, i.e.\ it consists
+exclusively of a set of headers. Therefore the generated C++ file is easily
+distribuable and does not depend on Pythran. One could even preprocess this
+C++ file to avoid the need of distributing along the Pythonic++ headers.
+
+The purpose of this section is not to dig into the internals
 of the compiler, but rather to focus on the impact of the parallelism layer,
@@
 
-To achieve this goal, a translator from a subset of the Python language to C++
-has been designed. This translator turns regular Python modules annotated with
+To achieve this goal, we designed Pythran, a translator from a subset of the
+Python language to C++. Pythran turns regular Python modules annotated with
 OpenMP directives and a few type annotations into native parallel module. The
 input module remains compatible with the standard interpreter and the underlying
-runtime is compatible with parallel constructs.
+runtime is compatible with parallel constructs. While Pythran accepts a limited
+input, we extend it progressively and with a great care on its stability. As
+such, even if its development is community-based, it is more than a research
+prototype and should be usable on real world code.
 

### 2013-10-26 pythran/pythonic++/core/shared_ref.h
Never use dynamic exception specification, it is costly!

@@ -27,23 +27,23 @@ namespace pythonic {
                             memory(Types&&... args) : ptr(args...),count(1) {}
                     } *mem;
                 public:
-                    shared_ref(no_memory const&) throw()
+                    shared_ref(no_memory const&) noexcept
                         : mem(nullptr)
                     {}
-                    shared_ref(shared_ref<T>&& p) throw()
+                    shared_ref(shared_ref<T>&& p) noexcept
                         : mem(p.mem)
                     {p.mem=nullptr;}
-                    shared_ref(shared_ref<T> const& p) throw()
+                    shared_ref(shared_ref<T> const& p) noexcept
                         : mem(p.mem)
                     {if(mem) acquire();}
 
                     template<class... Types>
                         shared_ref(Types... args) : mem(new memory(args...)) {}
 
-                    ~shared_ref() throw()
+                    ~shared_ref() noexcept
                     {dispose();}
 
-                    shared_ref<T>& operator=(shared_ref<T> const & p) throw()
+                    shared_ref<T>& operator=(shared_ref<T> const & p) noexcept
                     {
                         if(mem != p.mem){
                             dispose();
@@ -52,7 +52,7 @@ namespace pythonic {
                         }
                         return *this;
                     }
-                    shared_ref<T>& operator=(shared_ref<T> && p) throw()
+                    shared_ref<T>& operator=(shared_ref<T> && p) noexcept
                     {
                         if(mem != p.mem){
                             dispose();
@@ -62,10 +62,10 @@ namespace pythonic {
                         return *this;
                     }
 
-                    T& operator*() const throw()
+                    T& operator*() const noexcept
                     {return mem->ptr;}
 
-                    T* operator->() const throw()
+                    T* operator->() const noexcept
                     {return &mem->ptr;}
 
                     bool operator!=(shared_ref<T> const & other) const {

### 2013-10-26 doc/papers/sc2013/paper.tex
SC2013: take into account reviewer3 comments

@@
 sequential C, C++ and FORTRAN application into parallel applications, to the
-Pythran~\cite{pythran2013} compiler, a compiler that turn Python modules into
+Pythran~\cite{pythran2013} compiler, a compiler that turns Python modules into
 C++ meta-programs.  Section~\ref{sec:python-parallel} gives an overview of
@@
 backward-compatible approach. Section~\ref{sec:python-static} presents the
-Pythran compiler and the underlying runtime. Section~\ref{sec:python-openmp}
-studies the validity of using OpenMP directives within the Python language in
+Pythran compiler and the underlying runtime library.
+Section~\ref{sec:python-openmp} studies the validity of using OpenMP directives
+within the Python language in
 % I would dispute the fact that OpenMP is "fine-grain", we're not talking about % vectorization here.
@@
 There have also been several approaches to replace the GIL by Transactional
-Memories~\cite{Riley2006,Tabba2010} but none of them made there way to the
+Memories~\cite{Riley2006,Tabba2010} but none of them made their way to the
 mainstream interpreters. As a consequence, Python developers need to write
@@
 
-\subsection{Runtime Support}
+\subsection{Runtime Library Support}
 
 % STL != standard library
-Pythran runtime is based on the C++11 standard library which, unlike its C++03
-version, is thread-safe. It implies that no data dependencies are added to the
-original code.
-
-A critical point of the design of the \texttt{pythonic++} runtime is memory
-management, the very same aspect that led to the use of a GIL in CPython.
+Pythran runtime library is based on the C++11 standard library which, unlike its
+C++03 counterpart, is thread-safe. It implies that no data dependencies are
+added to the original code.
+
+A critical point of the design of the \texttt{pythonic++} runtime library is
+memory management, the very same aspect that led to the use of a GIL in CPython.
 Memory management is implemented in Shed Skin through the general-purpose Boehm
@@
 
-Pythran's runtime is aware of these three kinds of iterators and supports
-parallel iteration over random access iterators and iterators adapting random
-ones.
+Pythran's runtime library is aware of these three kinds of iterators and
+supports parallel iteration over random access iterators and iterators adapting
+random ones.
 
@@
 
-\subsubsection{Interaction With Pythran Runtime}
+\subsubsection{Interaction With Pythran Runtime Library}
 
@@
 Python test function into a C++ function with the same directives and runtime
-calls. Apart from the \texttt{threadprivate} directive and the
+library calls. Apart from the \texttt{threadprivate} directive and the
 \texttt{collapse(\emph{n})} clause, all tests were successful.
@@
 
-The target machine has  8 AMD Opteron 6176 SE cores and 8 GO of RAM. It is
+The target machine has  8 AMD Opteron 6176 SE cores and 8 GB of RAM. It is
 running a Linux kernel 3.10-2-amd64, x86\_64 GNU/Linux. The C++ compiler used is
@@
     \centering
-    \begin{tabular}{|l|r|}
+    \begin{tabular}{|l|c|}
         \hline
-        Language & SLOC     \\
+        Language & Source Lines Of Code (SLOC) \\
         \hline
@@
 
-Figure~\ref{fig:pb-rosen} shows the execution time of the computation of one
+Figure~\ref{fig:pb-growcut} shows the execution time of the computation of one
 step of the GrowCut algorithm used in image processing. On this example,
@@
 input module remains compatible with the standard interpreter and the underlying
-runtime is compatible with parallel constructs. While Pythran accepts a limited
-input, we extend it progressively and with a great care on its stability. As
-such, even if its development is community-based, it is more than a research
-prototype and should be usable on real world code.
+runtime library is compatible with parallel constructs. While Pythran accepts a
+limited input, we extend it progressively and with a great care on its
+stability. As such, even if its development is community-based, it is more than
+a research prototype and should be usable on real world code.
 
@@
 the CARP project. The authors thank Adrien \textsc{Merlini} and Alan
-\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
+\textsc{Raynaud} for their work on the runtime library, Albert \textsc{Cohen},
 B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Fran{\,c}ois

### 2013-10-27 doc/papers/sc2013/paper.tex
SC2013: update taking reviews into account
Still incomplete, missing points are added as comments
(look for FIXME)

@@
 ``batteries included'' standard library. It is also widely used in scientific
-computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
-which provides a set of scientific and numeric tools, ranging from linear
-algebra to ordinary differential equation solvers or generic algorithms.
+computing~\cite{Oliphant2007}, thanks to the \texttt{numpy} module and the
+SciPy~\cite{scipy} project which provide a set of scientific and numeric tools,
+ranging from linear algebra to ordinary differential equation solvers or generic
+algorithms.
 
@@
 
+The hybrid approach, where an application contains a mix of interpreted and
+native code, is getting widespread in the Python landscape.
+It is also the recommended approach to write applications that make use of fine
+grain parallelism, as Python parallelism support is only suitable for
+coarse-grain parallelism.
 To allow the user to write native functions without having to write the
@@
 possible to bundle C code snippets into Python code. They are then compiled and
-loaded at runtime. This hybrid approach, based on a mix of interpreted and
-native code, is getting widespread in the Python landscape. It is also the
-recommended approach to write applications that make use of fine grain
-parallelism, as Python parallelism support is only suitable for coarse-grain
-parallelism.
+loaded at runtime. Other project like PyCuda or PyOpenCL offers the same
+convenience to target hardware accelerators.
 
@@
 sequential and parallel version of the Buffon algorithm in Python.}]
-def buffon(darts):
+def buffon(darts, _ = 0):
   hits = 0
@@
   def work(darts, queue):
-    queue.put(pi_estimate(darts))
+    queue.put(buffon(darts))
   n = 4
@@
 
+import functools
 from multiprocessing import Pool
@@
   p = Pool(n)
-  return sum(p.map(buffon, darts//n))/n
+  return sum(p.map(functools.partial(buffon,darts/n), range(n)))/n
 \end{lstlisting}
-% Mehdi: I don't get the [darts//4]*4 at the last line of the listing
+
+Another widely used solution is the module IPython.parallel, which supports many
+different styles of parallelism including: single program, multiple data (SPMD)
+parallelism, multiple program, multiple data (MPMD) parallelism. message passing
+using MPI, task farming, and data parallel.
+FIXME: to be continued...
 
 There have also been several approaches to replace the GIL by Transactional
-Memories~\cite{Riley2006,Tabba2010} but none of them made their way to the
+Memories~\cite{Riley2006,Tabba2010} but none of them made its way to the
 mainstream interpreters. As a consequence, Python developers need to write
@@
 Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
-syntax with typing informations, calls to native functions from third party
+syntax with typing information, calls to native functions from third party
 libraries, and a limited set of parallelism constructs, such as the possibility
@@
 
-PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose an other approach
+PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose another approach
 that directly mixes Python with C, using raw strings to hold the C code. PLW
@@
 
-The main drawback of these approaches is that they require to modify in a
+The main drawback of these approaches % Reviewer 4 not happy with this
+is that they require to modify in a
 backward-incompatible way the original code. They require to learn a new
 dialect, and the long-term preservation of this investment is not ensured.
-Moreover, there is no fall back if the code were to be deployed in an
-environment where the parallelizing tool/module is not available. For instance,
-once a code has been ported to PyCUDA and transformed into CUDA code embedded in
-Python strings, there is no way back and the developer has to maintain two
-versions of the algorithm.
+Moreover, some of these approaches have no fallback if the code were to be
+deployed in an environment where the parallelizing tool/module is not available.
+For instance, once a code has been ported to PyCUDA and transformed into CUDA
+code embedded in Python strings, there is no way back and the developer has to
+maintain two versions of the algorithm.
 
@@
 
-Most backward-compatible approaches also require to modify the input program.
+Most backward-compatible approaches also % Reviewer 4 not happy with this
+require to modify the input program.
 They do not extend the Python language, but restrict it. As a consequence, they
@@
 process, an expression is always transformed into a single expression and a
-statement is always transformed into a single statement. Then this AST is
-converted into a C++ AST meant to be pretty-printed.
+statement is always transformed into a single statement.
+% FIXME as noted by reviewer 4, the previous sentence is not exact. Just a few
+% lines above we said that "tuple unpacking is turned in multiple assignments"
+Then this AST is converted into a C++ AST meant to be pretty-printed.
 
@@
 packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
-at the Python level. As a tool for scientific computing, Pythran supports such a
+at the Python level. The PEP 3118 defined the API that allows to share
+efficiently the embeded data between Python and the native world without
+involving any copy. As a tool for scientific computing, Pythran supports such a
 structure, implemented as a class wrapping a raw pointer.
@@
 \texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
-the \texttt{and} and \texttt{or} keywords have a special meaning in Python.} The
-latest OpenMP specifications~\cite{openmp4} describe a way to declare
+the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
+% FIXME reviewer 4 doesn't seem to agree with and/or C++ vs Python
+The latest OpenMP specifications~\cite{openmp4} describe a way to declare
 user-defined reduction but is not implemented in any compiler yet.
@@
 the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
-step of one until it is exhausted, in which case it raises an exception: it
+step of one until it is exhausted,
+% reviewer 4 disagree with the previous statement
+in which case it raises an exception: it
 behaves as a forward iterator. To be compatible with OpenMP, these iterators
@@
 
-Generator expressions are generators whose content is built from an other
+Generator expressions are generators whose content is built from another
 iterator. For instance \texttt{(x*x for x in l)} successively yields the square
@@
   #omp parallel
-  if 1
+  if 1:
     #omp critical
@@
 
+% FIXME reviewer 4
+% Fig 2, 3, etc: Please use a different color scheme, or add plot markers to
+% the plots for the colorblind among us who can't distinguish thin green and
+% red lines from each other.
+
+% FIXME reviewer 4
+% Fig 2, 3: I believe a log-log plot or a plot of the overall speedup versus
+% number of active threads would be a better way to display the data.
+
+% Remaining reviewer's comments:
+% - Is it the intent of the Pythran authors to expose the entirety of the
+%   OpenMP API, or just a subset?  Task parallelism?
+% - V.C: "Cython is not backwards compatible with Python".  This is false:
+%   Cython's Pure Python mode is fully backwards compatible.
+% - Figs 5,6,7,8: to do apples-to-apples, what about Cython + prange?
+% - Perhaps the only complain is that I would have liked to see the timing for
+%   compilation to be included in the article.  Pythran and Cython are very
+%   different beasts than Numba or Parakeet in this sense, and the later maybe
+%   more useful to compile code that still not exists at compilation time (like
+%   user queries, for example).
+
 \end{document}

### 2013-11-22 doc/papers/wpmvp14/paper.tex
Add paper skeleton for wpmvp'14

% submit to https://sites.google.com/site/wpmvp2014/home
\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

%\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
%\copyrightyear{20yy} 
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

\title{Vectorizing Python Constructs Using Pythran and Boost SIMD}

\authorinfo{Serge Guelton}
           {QuarksLab, T{\'e}l{\'e}com Bretagne}
           {sguelton@quarkslab.com}
\authorinfo{Jo{\"e}l Falcou}
           {MetaScale}
           {joel.falcou@metascale.org}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%term1, term2

\keywords
Vectorization, Meta-Programming, Python, C++

%%
%%
\section{Python, Unboxing and Pythran}

\subsection{The Unboxing Problem}

\subsection{The Pythran Compiler}

%%
%%
\section{Changing Python Semantic for Vectorization}

\subsection{Intrinsics}

\subsection{List Comprehension}

\subsection{The Numpy Module}

%%
%%
\section{Vectorization of a Meta-Program}

\subsection{Boost SIMD}

\subsection{Load/Store Protocol}

%%
%%
\section{Validation of the Approach using the Pythran Compiler}

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.
\cite{*}

\bibliographystyle{abbrvnat}
\bibliography{biblio}

\end{document}

### 2013-11-22 doc/papers/wpmvp14/paper.tex
Abstract!

@@
 \usepackage{amsmath}
+\usepackage{listings}
 
@@
 \begin{abstract}
-This is the text of the abstract.
+
+    The Python language is highly dynamic, most notably due to late binding. As
+    a consequence, program run using Python typically run an order of magnitude
+    slower than their C counterpart. It is also a high level languages whose
+    semantic can be made more static without much change from a user point of
+    view in the case of mathematical applications. In that case, the language
+    provides several vectorization opportunities that are studied in this
+    paper, and evaluated in the context of Pythran, an ahead-of-time compiler
+    that turns Python module into C++ meta-programs.
+
 \end{abstract}
@@
 \section{Python, Unboxing and Pythran}
+
+The Python language has grown in audience for the past ten years, even reaching
+the world of scientific computations thanks to the \texttt{numpy} module, a
+module that provides a MATLAB-like API. As a consequence, more and more code is
+being written either in pure Python, generally to prototype an application, or
+as a Python and native code mix when performance matters. However, Python
+trades performance for dynamicity and does not particularly shines in terms of
+performance.
+
+For instance, the natural solution to Project Euler sixth problem ``Find the
+difference between the sum of the squares of the first one hundred natural
+numbers and the square of the sum`` can be coded as in
+Listing~\ref{lst:euler06-py} in Python and as in Listing~\ref{lst:euler06-c} in
+C.
+
+\begin{lstlisting}[language=python, label={lst:euler06-py}, caption={Solution to the Project Euler Sixth Problem in Python}]
+def solve();
+  r = xrange(1, 101)
+  a = sum(r)
+  return a * a - sum(i*i for i in r)
+\end{lstlisting}
+
+\begin{lstlisting}[language=c, label={lst:euler06-c}, caption={Solution to the Project Euler Sixth Problem in C}]
+size_t solve() {
+  size_t a = 0, s = 0;
+  for(size_t i = 0; i< 101; ++i) {
+    a += i;
+    s += i*i;
+  }
+  return a * a - s
+}
+\end{lstlisting}
 

### 2013-11-22 doc/papers/wpmvp14/paper.tex
Intro: first XP

@@
 \usepackage{listings}
+\usepackage{hyperref}
 
@@
 Listing~\ref{lst:euler06-py} in Python and as in Listing~\ref{lst:euler06-c} in
-C.
+C. A comparison of the performance of the Python function and the C function
+called through the \texttt{ctypes} module shows that the C version runs more
+than $\times250$ faster than the Python version. Enabling compiler
+auto-vectorization makes the C version $\times400$ faster than the Python
+version.~\footnote{To perform more reliable time measurements, \texttt{n} has
+    been increased to 100001 instead of 101. The script used to perform the
+    measures as well as all other experimental data are available at
+\url{https://github.com/serge-sans-paille/pythran/tree/wpmvp14/doc/papers/wpmvp14/experiments}.}
 
-\begin{lstlisting}[language=python, label={lst:euler06-py}, caption={Solution to the Project Euler Sixth Problem in Python}]
-def solve();
-  r = xrange(1, 101)
-  a = sum(r)
-  return a * a - sum(i*i for i in r)
-\end{lstlisting}
+\lstinputlisting[language=python, label={lst:euler06-py}, caption={Solution to the Project Euler Sixth Problem in Python}]{experiments/euler06.py}
 
-\begin{lstlisting}[language=c, label={lst:euler06-c}, caption={Solution to the Project Euler Sixth Problem in C}]
-size_t solve() {
-  size_t a = 0, s = 0;
-  for(size_t i = 0; i< 101; ++i) {
-    a += i;
-    s += i*i;
-  }
-  return a * a - s
-}
-\end{lstlisting}
+\lstinputlisting[language=c, label={lst:euler06-c}, caption={Solution to the Project Euler Sixth Problem in C}]{experiments/euler06.c}
 

### 2013-11-22 doc/papers/wpmvp14/experiments/euler06.c

@@ -1,0 +1,8 @@
+unsigned int solve(unsigned int n) {
+  unsigned int a = 0, s = 0;
+  for(unsigned int i = 0; i< n; ++i) {
+    a += i;
+    s += i*i;
+  }
+  return a * a - s;
+}

### 2013-11-22 doc/papers/wpmvp14/experiments/euler06.py

@@ -1,0 +1,4 @@
+def solve(n):
+  r = range(1, n)
+  a = sum(r)
+  return a * a - sum(i*i for i in r)

### 2013-11-22 doc/papers/wpmvp14/experiments/run_xp_euler06.sh

@@ -1,0 +1,5 @@
+python -m timeit -s 'import euler06 as e; s = e.solve' 's(100001)'
+cc -shared -O3 -std=c99 euler06.c -o libeuler06.so
+python -m timeit -s "from ctypes import cdll ; d = cdll.LoadLibrary('./libeuler06.so') ; s = d.solve" "s(100001)"
+cc -shared -O3 -std=c99 -march=native euler06.c -o libeuler06.so
+python -m timeit -s "from ctypes import cdll ; d = cdll.LoadLibrary('./libeuler06.so') ; s = d.solve" "s(100001)"

### 2013-11-24 doc/papers/wpmvp14/paper.tex
++introduction

@@
 
+There are two main reason for the poor performance of Python:
+
+\begin{description}
+
+    \item[dynamic binding] each variable access is made through a dictionnary
+        look-up to bind variable names to variable instances~;
+
+    \item[boxing] each variable value is encapsulated into a heap-allocated
+        generic object ---the \texttt{PyObject}--- which implies extra
+        indirections for each operation.
+
+\end{description}
+
+In that context, it does not make sense to speak about vectorization: the
+nature of an operation is unknown to the interpreter until its execution, and
+data are not contiguous in memory.
+
 \lstinputlisting[language=python, label={lst:euler06-py}, caption={Solution to the Project Euler Sixth Problem in Python}]{experiments/euler06.py}
@@
 
-\subsection{The Unboxing Problem}
+Several approaches have been proposed to solve these issues: using native
+modules, relying on Ahead-Of-Time (AOT) or Just-In-Time (JIT) compilation.
 
-\subsection{The Pythran Compiler}
+A native module is a shared library whose interface make it callable from the
+Python interpreter using the regular \texttt{import} mechanism. It makes it
+possible to call FORTRAN/C/C++ code from Python in a transparent way. Typically
+Python object are unboxed when transferred from Python to C, where efficient
+computations can happen. The the result is boxed when getting back to the
+Python world. In between, vectorization can happen.  This is the approach taken
+by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
+contiguous unboxed elemnts, and provides a great deal of high-level functions
+to manipulate them. All the computation-intensive part is done in pure C.
+Vectorization can eventually happen at this level, for instance when performing
+a point-to-point operation like the sum of two arrays.  However the granuarity
+of the operations limits the ratio of load / stores versus operation. For
+instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum
+\texttt{a + b +c} is computed using two loops and an intermediate array.
+
+AOT compilation uses type inference or type annotation to statically compute
+the type of all (like in Shedskin, Parakeet) or part of (like in Cython or
+Numba) variables, effectively unboxing them to speed-up computations. Using JIT
+compilation, type information is available at runtime and the unboxing decision
+is based on profitablity, e.g.\ according to a trace analysis. This is the
+approach of the PyPy project. In both cases, generation of vector instruction
+is possible, but none of the cited project use it.  Indeed automatic
+vectorization is a complex topic, and as stated in~\cite{maleki2011}, many cases
+are not yet correctly supported even for mainstream compilers.
 

### 2013-11-25 doc/papers/wpmvp14/paper.tex
++Introduction

@@
 
+Pythran is also an AOT compiler, which holds the uinique property of turning
+Python modules into C++ \emph{meta-programs}. It does not need type annotations
+and keeps the polymorphism of original functions, respecting Python's duck
+typing. Like other AOT compilers, it unboxes all variables for better
+performance, and performs a wide variety of optimizations such as lazy
+evaluation, constant folding, expression fusion etc.  It supports core
+\texttt{numpy} constructs and most Python constructs to the notable exception
+of user classes. Constructs such as \texttt{eval}, are not supported.
+
+This paper introduces vectorization of Python programs based on the idea that
+many Python constructs already exhibit good vectorization opportunities. The
+goal of the paper switches from automatic vectorization to vectorization of
+high-level Python intrisics, as described in Section~\ref{sec:python-semantic}.
+One of the difficulties of vectorizing Python functions is linked to function
+polymorphism. Vectorization in the context of C++ meta-programs is examined in
+Section~\ref{sec:meta-vectorization}. The approach is validated using the
+Pythran compiler on several micro-applications from signal processing, physics
+and mathematics presented in Section~\ref{sec:benchs}.
+
 %%
 %%
-\section{Changing Python Semantic for Vectorization}
+\section{Using Python Semantic for Vectorization}
+\label{sec:python-semantic}
 
@@
 \section{Vectorization of a Meta-Program}
+\label{sec:meta-vectorization}
 
@@
 \section{Validation of the Approach using the Pythran Compiler}
+\label{sec:benchs}
 

### 2013-11-25 doc/papers/wpmvp14/paper.tex
++Python

@@
 \usepackage{hyperref}
+
+\lstset{captionpos=b, float}
 
@@
 
+Automatic vectorization generally relies on explicit loop-level vectorization, as
+extensively described in~\cite{bik04}, or on block-level vectorization, also
+called Superword Level Parallelism (SLP) as introduced in~\cite{larsen00}.
+Modern compilers for C/C++/FORTRAN generally implement a mixture of both.
+
+A high-level language such as Python exhibit another level of vectorization
+opportunity in implict loop. Implicit loops can be foud in many places, but
+this paper focuses on \emph{list comprehension} and \emph{generator
+expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
+through the \texttt{numpy} module.
+
 \subsection{Intrinsics}
+
+\subsubsection{\texttt{sum}}
+
+The \texttt{sum} intrinsic is a typical case of potential vectorization of an
+implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
+value} and its behavior is equivalent to the Python code in
+Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
+wether one needs to strictly conforms to the IEEE Standard for Floating-Point
+Arithmetic (IEEE 754).
+
+\begin{lstlisting}[language=python, label={lst:sum-py}, caption={Pseudo code of the \texttt{sum} intrinsic.}]
+def sum(sequence, value=0):
+  for e in sequence:
+    value += e
+  return value
+\end{lstlisting}
+
+Appart from standard compliancy, to be able to efficiently vectorize this code in C++, one must check that:
+
+\begin{enumerate}
+    \item\label{enu:same-types} All element of \texttt{sequence} have the same types;
+    \item\label{enu:scalar-types} \texttt{sequence} contains scalar types;
+    \item\label{enu:contiguous} \texttt{sequence} contains contiguous elements;
+    \item\label{enu:random-access} \texttt{sequence} provides a random access iterator. in particular one can computes its size in $\mathcal{O}(1)$.
+\end{enumerate}
+
+Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
+unboxing. For instance, any code that does not match it will fail to compile
+using Pythran or Shed-Skin. It is also satisified for the \texttt{ndarray}
+container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
+checked at compile time. Python supports two native types candidates for
+vectorization: \texttt{int} (64 bits integer) and \texttt{float}
+(double-precision floating point numbers). The \texttt{numpy} model supports a
+wide range of signed and unsigned integers as well as single, double or
+quadruple precision float. Depending on the target architecture, only some of
+them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
+is a static property of a container. It is verified for C++ alternative to
+Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
+for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not matched by \emph{generator}s, the Python incarnation of co-routines.
+
+Depending on the container properties, it is possible to select a vectorized
+implementation of fallback to a generic one, e.g.\ using Substitution Failure
+Is Not An Error (SFINAE).
+
+A possible C++ meta-implementation of \texttt{sum} in that context is presented
+in listing~\ref{lst:sum-cxx}. It makes use of the parametric vector register
+introduced by Boost.SIMD and presented in Section~\ref{sec:meta-vectorization}.
+There is nothing new in the vectorization scheme itself. However, the
+possibility to write a meta-function that provides a vectorized implementation
+of the \texttt{sum} intrinsic makes it possible to efficiently turn a Python
+call into a C++ call.
+
+\begin{lstlisting}[language=c++, label={lst:sum-cxx}, caption={Meta-implementation of the \texttt{sum} intrinsic in C++11.}]
+template<class T, class V=long>
+typename std::enable_if<is_vectorizable<T>::value,
+                        typename T::value_type
+                       >::type
+sum(T const& sequence, V value=0L)
+{
+  // E is the type of an element of T
+  typedef typename T::value_type E;
+  // vE is a vector type of vsize elements of type E
+  typedef typename simd::native<E,
+    BOOST_SIMD_DEFAULT_EXTENSION> vE;
+  static const size_t vsize =
+    simd::meta::cardinal_of<vE>::value;
+
+  // vector_value is for the parallel reduction
+  vE vector_value = simd::splat<vE>(E());
+  const size_t n = sequence.size();
+  const size_t bound = n / vsize * vsize;
+  for(size_t i = 0; i < bound; i+= vsize)
+    vp += sequence.load(i);
+  // inner sum of the vector
+  value += boost::simd::sum(vp);
+
+  // handle the tail
+  for(size_t i = bound; i < n; ++i)
+    value += sequence.at(i);
+  return value;
+}
+\end{lstlisting}
+
+\subsubsection{\texttt{map}}
+
+The \texttt{map} intrinsic simply creates a new \texttt{list} and fills it by
+repeatedly applying a function to the items of its argument sequence(s). Its
+prototype is: \texttt{map(function, sequence[, sequence, \dots]) $\rightarrow$
+list}. Again, this is a typical case of vectorizable implicit loops, providing some conditions are met:
+
+\begin{enumerate}
+
+    \item[\label{enu:pure}] \texttt{function} must be vectorizable;
+    \item[\label{enu:sequence}] all \texttt{sequence}s verify the same constraints as stated for the \texttt{sum} intrinsic.
+
+\end{enumerate}
+
+Deciding if a given function is vectorizable can be approximated through a
+composition rules. First, some functions are known to be vectorizable. This is
+generally the case for the addition, substraction etc. Depending on the target
+vector instruction set, some operations may or may not be available as vector
+instructions, e.g.\ trigonometric functions, but they can be provided as
+composite instructions. Some are just not available, as quadruple precision
+float addition. Starting from this set of vectorizable function, one can use
+the \texttt{;} as a closure binary relation to compute a subset of all
+vectorizable functions. Said otherwise, any sequence of operation that only
+involve vector operations is a vector operation and a function whose body is a
+sequence of vector operation or vectorizable functions is a vectorizable
+function itself.
+
+The preceding statemnt is only valid if the function's body only references
+vector functions and formal parameters, as the function must be valid for both
+scalar types and vector types. Note that this perfectly matches the translation
+of polymorphic Python functions into C++ meta-functions, as detailed in
+Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}.
+
+\begin{lstlisting}[language=python, label={lst:square-python}, caption={Python implementation of the square function.}]
+def square(x):
+  return x * x
+\end{lstlisting}
+
+\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={C++ meta-implementation of the square function.}]
+template<class T>
+auto square(T const& x) -> decltype(x * x) {
+  return x * x;
+}
+\end{lstlisting}
 

### 2013-11-26 doc/papers/wpmvp14/paper.tex
++Python

@@
 %%
+\section{Vectorization of a Meta-Program}
+\label{sec:meta-vectorization}
+
+\subsection{Boost SIMD}
+
+\subsection{Load/Store Protocol}
+
+%%
+%%
 \section{Using Python Semantic for Vectorization}
@@
 Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
-for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not matched by \emph{generator}s, the Python incarnation of co-routines.
+for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
+matched by \emph{generator}s, the Python incarnation of co-routines.
 
@@
 prototype is: \texttt{map(function, sequence[, sequence, \dots]) $\rightarrow$
-list}. Again, this is a typical case of vectorizable implicit loops, providing some conditions are met:
+list}. Again, this is a typical case of vectorizable implicit loops, providing
+some conditions are met:
 
@@
 composition rules. First, some functions are known to be vectorizable. This is
-generally the case for the addition, substraction etc. Depending on the target
+generally the case for the addition, subtraction etc. Depending on the target
 vector instruction set, some operations may or may not be available as vector
@@
 
-The preceding statemnt is only valid if the function's body only references
+The preceding statement is only valid if the function's body only references
 vector functions and formal parameters, as the function must be valid for both
@@
 of polymorphic Python functions into C++ meta-functions, as detailed in
-Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}.
+Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}. The core idea is that
+the \texttt{square} function can be called indifferently on floating point
+values or on vector of floating point values. Through (automatic) template
+instantiation, the C++ compiler generates the actual scalar and vector version
+from the same meta-function. As Pythran always translates Python's polymorphic
+function in that form, writing a vector version of the \texttt{map} intrinsic is
+slightly more difficult than for the \texttt{sum} intrinsic due to its variadic
+number of arguments, but it also relies on the \texttt{load}/\texttt{store}
+protocol and SFINAE to select it.
 
@@
 \begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={C++ meta-implementation of the square function.}]
+struct square {
     template<class T>
-auto square(T const& x) -> decltype(x * x) {
+    auto operator()(T const& x)
+    -> decltype(x * x)
+    {
       return x * x;
     }
+};
 \end{lstlisting}
@@
 
+Python provides \emph{list comprehension} as a way to create \texttt{list} from
+any iterable. For instance \texttt{[abs(x) for x in l]} creates a new
+\texttt{list} that holds the absolute value of each element of \texttt{l}. A
+filter can be set up as in  \texttt{[abs(x) for x in l if x \% 3]}, where only
+the elements non multiple of 3 are taken into account. Let us call
+\emph{comprehesion rule} the expression on the left of the
+\texttt{for}.
+
+List comprehension can be translated into a combination of \texttt{map} and
+(optionnaly) \texttt{filter}, by transformation of the comprehension rule into
+a lambda function. For instance the previous example can be re-written as
+\texttt{map(abs, filter(lambda x: x\% 3, l))}. Once in this form, the same
+    vectorization opportunities as in the previous section arise.
+
+Generator expression is somehow similar to list comprehension, to the notable
+exception that the value returned is an \emph{iterator adaptator} on the
+container argument. In the context of vectorization, a generator expression
+supports the vector protocol if the associated container also supports it. In
+that case the  \emph{iterator adaptator} implements the vector protocol by
+applying the comprehension rule to each vctor element from this container, in a
+streaming fashion.
+
+Set comprehensions and dict comprehensions can also be rewritten in the form of
+generator expression using the eponym function: \texttt{\{ x + 3 for x in l\}}
+is equivalent to \texttt{set(x + 3 for x in l)}.
+
 \subsection{The Numpy Module}
-
-%%
-%%
-\section{Vectorization of a Meta-Program}
-\label{sec:meta-vectorization}
-
-\subsection{Boost SIMD}
-
-\subsection{Load/Store Protocol}
 

### 2013-11-26 doc/papers/wpmvp14/paper.tex
++Python

@@
 \lstset{captionpos=b, float}
+
+\usepackage{tikz}
+\usetikzlibrary{positioning}
+\usetikzlibrary{shadows}
+\usetikzlibrary{arrows}
+\usetikzlibrary{shapes}
 
@@
            {joel.falcou@metascale.org}
+\authorinfo{Pierrick Brunet}
+           {INRIA/MOAIS}
+           {pierrick.brunet@inria.fr}
 
@@
 
-Pythran is also an AOT compiler, which holds the uinique property of turning
+Pythran is also an AOT compiler, which holds the unique property of turning
 Python modules into C++ \emph{meta-programs}. It does not need type annotations
 and keeps the polymorphism of original functions, respecting Python's duck
-typing. Like other AOT compilers, it unboxes all variables for better
-performance, and performs a wide variety of optimizations such as lazy
-evaluation, constant folding, expression fusion etc.  It supports core
-\texttt{numpy} constructs and most Python constructs to the notable exception
-of user classes. Constructs such as \texttt{eval}, are not supported.
+typing. It compilation frow is described in Figure~\ref{fig:pythran-compiler}.
+Like other AOT compilers, it unboxes all variables for better performance, and
+performs a wide variety of optimizations such as lazy evaluation, constant
+folding, expression fusion etc.  It supports core \texttt{numpy} constructs and
+most Python constructs to the notable exception of user classes. Constructs
+such as \texttt{eval}, are not supported.
+
+\begin{figure}
+
+\centering
+\begin{tikzpicture}[
+file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
+node distance=0.7cm},
+    tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
+    distance=0.7cm}]
+    \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
+    \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
+    \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
+    \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
+    \node (empty) [xshift=-1em, left=of gxx] {};
+    \node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic++}}};
+    \node[file] (annotation)     [above=of pythonic] {\textbf{Type Info}};
+    \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
+    \node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module
+    [\texttt{.so}]}};
+
+\draw[very thick, ->] (python) -- (pythran);
+\draw[very thick] (annotation) -| (empty.center);
+\draw[very thick, ->] (pythran) -- (meta-cxx);
+\draw[very thick, ->] (meta-cxx) -- (gxx);
+\draw[very thick] (boost) -| (empty.center);
+\draw[very thick, ->] (pythonic) -- (gxx);
+\draw[very thick, ->] (gxx) -- (so);
+\end{tikzpicture}
+
+\caption{Pythran compiler workflow.}
+\label{fig:pythran-compiler}
+\end{figure}
 
@@
 
-Set comprehensions and dict comprehensions can also be rewritten in the form of
+Set comprehension can also be rewritten in the form of
 generator expression using the eponym function: \texttt{\{ x + 3 for x in l\}}
@@
 
+Dictionnary comprehension and multi-level comprehension (i.e.\ containing
+multiple \texttt{for} clauses) are not considered in this preliminary work.
+
 \subsection{The Numpy Module}
+
+The \texttt{numpy} module played a major role in the adoption of Python by the
+scientific community. In a nutshell, this is a thin Python wrapper above a raw
+array that provides an array language embeded into Python. An example of
+\texttt{numpy} code is show on Listing~\ref{lst:numpy-code}.
+
+\begin{lstlisting}[language=python, label={lst:numpy-code}, caption={Sample \texttt{numpy} code.}]
+import numpy as np
+def rosen(x):
+  return np.sum(100*(x[1:] - x[:-1] ** 2) ** 2
+                + (1 - x[:-1]) ** 2)
+\end{lstlisting}
+
+Common features of array language include indexing, slicing, point-to-point
+operation etc. The latter is traditionnaly a good source of vectorization.
+However, because of the module layer, \texttt{numpy} always create a new array
+to hold intermediate result. For instance in the expression \texttt{(1 -
+x[:-1]) ** 2)} two array are created ---slicing does not create an array but a
+view. The implied extra cost would hide the benefits of vectorization.
+
+The typical way to solve this issue in C++ is to use \emph{expression
+templates}, a technique described in~\cite{} that uses the C++ type system to
+perform lazy evaluation of an expression. However, this approach is limited to
+a single expression. Listing~\ref{lst:numpy-code-fs} illustrates the situation:
+expression templates would not prevent the creation of temporary arrays
+\texttt{t0} and \texttt{t1}. Fortunately, it is possible to apply \emph{forward
+substitution} to fallback to the expression from Listing~\ref{lst:numpy-code}
+and avoid the creattion of any temporary.
+
+\begin{lstlisting}[language=python, label={lst:numpy-code-fs}, caption={Sample \texttt{numpy} code requiring forward subsitution.}]
+import numpy as np
+def rosen(x):
+  t0 = 100*(x[1:] - x[:-1] ** 2) ** 2
+  t1 = (1 - x[:-1]) ** 2
+  return np.sum(t0 + t1)
+\end{lstlisting}
+
+It is possible to hook in the expression template to add vectorization support.
+Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
+loop, that can be explicitly vectorized using Bosst.SIMD in a similar manner as
+the \texttt{map} and \texttt{sum} intrinsics.
+
+One potential drawback of the approach would be the generation of multiple load
+from the same location, as in \texttt{1. / a + a}, where \texttt{a} is loaded
+twice on each loop iteration. Fortunately, the backend C++ compiler is capable
+of detecting the redundant load and perform \emph{redundant load store
+elimination}, effectively issuing a single load.
 

### 2013-11-27 doc/papers/wpmvp14/paper.tex
Piece of Boost.SIMD

@@
 
+\providecommand{\boostsimd}{\textsc{Boost.SIMD}}
+\providecommand{\cpp}[1][~]{\textsc{C++}#1}
+\providecommand{\ie}[1][~]{\textit{i.e.}#1}
+\providecommand{\eg}[1][~]{\textit{e.g.#1}}
+
 \begin{document}
@@
 \authorinfo{Jo{\"e}l Falcou}
-           {MetaScale}
-           {joel.falcou@metascale.org}
+           {LRI, Universit\'e Paris-Sud}
+           {joel.falcou@lri.fr}
 \authorinfo{Pierrick Brunet}
@@
 \label{sec:meta-vectorization}
+SIMD extensions have been a feature of choice for processor manufacturers for a
+couple of decades. However, programming applications that take advantage of the 
+SIMD extension available on the current target remains a complex task. Programmers 
+that use low-level intrinsics have to deal with a verbose programming style due 
+to the fact that SIMD instructions sets cover a few common functionnalities, requiring
+to bury the initial algorithms in architecture specific implementation details.
+Furthermore, these efforts have to be repeated for every different extension
+that one may want to support, making design and maintenance of such applications 
+very time consuming.
 
-\subsection{Boost SIMD}
+\boostsimd{} is a high-level C++ library to program SIMD architectures. 
+Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
+provides both expresiveness and performance by using generic programming 
+to handle vectorization. \boostsimd{} does not only provide a portable 
+way but also enable the use of common programming idioms when designing 
+SIMD-aware algorithms. Two components are provided:
+
+\begin{itemize}
+\item \textbf{An abstraction of SIMD registers} for portable algorithm design,
+coupled with a large set of functions covering the classical set of operators
+along with a sensible amount (200+) of mathematical functions and utility functions,
+\item \textbf{A support for \cpp standard components} like \texttt{Iterator}
+over SIMD \texttt{Range}, SIMD-aware allocators and SIMD-aware STL algorithms.
+\end{itemize}
+
+\subsection{SIMD register abstraction}
+The first level of abstraction introduced by \textsc{Boost.} \textsc{SIMD} is
+the \texttt{pack} class. For a given type \texttt{T} and a given static integral 
+value \texttt{N} (\texttt{N} being a power of 2), a \texttt{pack} encapsulates
+the best type able to store a sequence of \texttt{N} elements of type \texttt{T}.
+For arbitrary \texttt{T} and \texttt{N}, this type is simply \texttt{std::array<T,N>}
+but when \texttt{T} and \texttt{N} matches the type and width of a SIMD register,
+the architecture-specific type used to represent this register is used instead.
+This semantic provides a way to use arbitrarily large SIMD registers on any system
+and let the library select the best vectorizable type to handle them. By default,
+if \texttt{N} is not provided, \texttt{pack} will automatically select a value
+that will trigger the selection of the native SIMD register type. Moreover, by
+carrying informations about its underlying scalar type, \texttt{pack} enables
+proper instruction selection even when used on extensions (like SSE2 and above)
+that map all integral type to a single SIMD type (\texttt{\_\_m128i} for SSE2).\\
+
+\texttt{pack} handles these low-level SIMD register type as regular objects with
+value semantics, which includes the ability to be constructed or copied from a
+single scalar value, list of scalar values, iterator or range. In each case, the
+proper register loading strategy (splat, set, load or gather) will be issued.
+\texttt{pack} also takes care of issues like boolean predicates support and provides
+both a range and a tuple-like interface.
+
+\subsection{Interaction with Standard Algorithm}
+Modern \cpp programming style based on Generic Programming usually leads to an
+intensive use of various STL components like Iterators. \boostsimd{} provides
+iterator adaptors that turn regular random access iterators into iterators
+suitable for SIMD processing. These adaptors act as free functions taking
+regular iterators as parameters and return iterators that output \texttt{pack}
+whenever dereferenced. These iterators are then usable directly in usual STL
+algorithms such as \texttt{transform} or \texttt{fold} (Listing \ref{stl_example}).
+
+\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={SIMD Iterator with STL algorithm},label=stl_example]
+std::vector<int,allocator<int> > v(N),r(N);
+
+std::transform ( simd::begin(v.begin())
+               , simd::end(v.end())
+               , simd::begin(r.begin())
+               , [](pack<int> const& p)
+                 { 
+                   return p*p; 
+                 } 
+              );
+\end{lstlisting}
+
+Previous examples of integration within standard algorithm are still limited.
+Applying \texttt{transform} or \texttt{fold} algorithm on SIMD aware data
+requires the size of the data to be an exact multiple of the SIMD register width
+and, in the case of fold, to potentially perform additional operations at the
+end of its call. To alleviate this limitation, \boostsimd{}
+provides its own overload for both \texttt{transform} and \texttt{fold} that take
+care of potential trailing data and performs proper completion of \texttt{fold}.
+Moreover, as \boostsimd{} provides default implementation of all its function for scalar values,
+we can rely on polymorphic callable object to write a mixed scalar/SIMD call to algortihm
+like \texttt{transform}. By using the \texttt{boost::simd::transform} algorithm, code
+handlign alignment, scalar prologue and epilogue is generated.
+
+\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={Polymorphic SIMD algorithm call},label=stl_poly]
+std::vector<int,allocator<int> > v(128), r(128);
+
+struct f
+{
+  template<class T>
+  auto operator()(T const& p)
+  -> decltype(p * p)
+  {
+    return p * p;
+  }
+};
+
+simd::transform( v.begin(), v.end()
+               , r.begin()
+               , f() 
+               );
+\end{lstlisting}
+
+Code written this way keeps a conventional structure and facilitate an usage of template
+functors for both scalar and SIMD cases also helps maximizing code reuse.
 

### 2013-11-27 doc/papers/wpmvp14/paper.tex
First Pierrick review

--- 2013-11-26 doc/papers/wpmvp14/paper.tex
@@
 the world of scientific computations thanks to the \texttt{numpy} module, a
+%% what about the scipy module? It is not really used?
 module that provides a MATLAB-like API. As a consequence, more and more code is
@@
 Python interpreter using the regular \texttt{import} mechanism. It makes it
-possible to call FORTRAN/C/C++ code from Python in a transparent way. Typically
+possible to call native code from Python in a transparent way. Typically
+%% It may be others languages, isn't it?
 Python object are unboxed when transferred from Python to C, where efficient
-computations can happen. The the result is boxed when getting back to the
+computations can happen. Then the result is boxed when getting back to the
 Python world. In between, vectorization can happen. This is the approach taken
 by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
-contiguous unboxed elemnts, and provides a great deal of high-level functions
+contiguous unboxed elements, and provides a great deal of high-level functions
 to manipulate them. All the computation-intensive part is done in pure C.
 Vectorization can eventually happen at this level, for instance when performing
-a point-to-point operation like the sum of two arrays.  However the granuarity
-of the operations limits the ratio of load / stores versus operation. For
+a point-to-point operation like the sum of two arrays. However the granularity
+of the operations limits the ratio of \texttt{load}/\texttt{store} versus operation. For
 instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum
-\texttt{a + b +c} is computed using two loops and an intermediate array.
+\texttt{a + b + c} is computed using two loops and an intermediate array.
+% It means there are 4 load / stores and only two loops of addition.
 
@@
 compilation, type information is available at runtime and the unboxing decision
-is based on profitablity, e.g.\ according to a trace analysis. This is the
+is based on profitability, e.g.\ according to a trace analysis. This is the
 approach of the PyPy project. In both cases, generation of vector instruction
@@
 and keeps the polymorphism of original functions, respecting Python's duck
-typing. It compilation frow is described in Figure~\ref{fig:pythran-compiler}.
-Like other AOT compilers, it unboxes all variables for better performance, and
+typing. Its compilation workflow is described in Figure~\ref{fig:pythran-compiler}.
+Like others AOT compilers, it unboxes all variables for better performance, and
 performs a wide variety of optimizations such as lazy evaluation, constant
@@
 
-\subsection{Load/Store Protocol}
+\subsection{\texttt{Load}/\texttt{Store} Protocol}
 
@@
 A high-level language such as Python exhibit another level of vectorization
-opportunity in implict loop. Implicit loops can be foud in many places, but
+opportunity in implict loop. Implicit loops can be found in many places, but
 this paper focuses on \emph{list comprehension} and \emph{generator
@@
 \subsubsection{\texttt{sum}}
+% only sum as texttt in a title is not really beautiful
 
@@
 Arithmetic (IEEE 754).
+% I don't understand what this last sentence is doing here :-s
 
@@
     \item\label{enu:contiguous} \texttt{sequence} contains contiguous elements;
-    \item\label{enu:random-access} \texttt{sequence} provides a random access iterator. in particular one can computes its size in $\mathcal{O}(1)$.
+    \item\label{enu:random-access} \texttt{sequence} provides a random access iterator. In particular one can computes its size in $\mathcal{O}(1)$.
 \end{enumerate}
@@
 container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
-checked at compile time. Python supports two native types candidates for
+checked at compile time. Python supports two native scalar types candidates for
 vectorization: \texttt{int} (64 bits integer) and \texttt{float}
-(double-precision floating point numbers). The \texttt{numpy} model supports a
+(double-precision floating point numbers). The \texttt{numpy} model add supports for a
 wide range of signed and unsigned integers as well as single, double or
@@
 
-Depending on the container properties, it is possible to select a vectorized
-implementation of fallback to a generic one, e.g.\ using Substitution Failure
+%we know a work have to be done about random acces but you speak about something else now ...
+%It is disturbing and we don't really know why you speak about this new topic.
+
+As function have to be generic but some containers do not match requierement
+properties, it is possible to select a vectorized
+implementation of fallback to a generic one using Substitution Failure
 Is Not An Error (SFINAE).
@@
 that case the  \emph{iterator adaptator} implements the vector protocol by
-applying the comprehension rule to each vctor element from this container, in a
+applying the comprehension rule to each vector element from this container, in a
 streaming fashion.
+
+Python also provide same features for set and dictionnary.
 

### 2013-11-27 doc/papers/wpmvp14/experiments/euler06.c
Update pythran to match the paper
this is quick'n dirty and should *never* make its way to master

@@ -1,6 +1,6 @@
-unsigned int solve(unsigned int n) {
-  unsigned int a = 0, s = 0;
-  for(unsigned int i = 0; i< n; ++i) {
+unsigned int solve(long n) {
+  long a = 0, s = 0;
+  for(long i = 0; i< n; ++i) {
     a += i;
     s += i*i;
   }

### 2013-11-27 doc/papers/wpmvp14/experiments/euler06.py

@@ -1,3 +1,4 @@
+#pythran export solve(int)
 def solve(n):
   r = range(1, n)
   a = sum(r)

### 2013-11-27 doc/papers/wpmvp14/experiments/map0.py

@@ -1,0 +1,5 @@
+#pythran export map0(int)
+def map0(n):
+    r = range(n)
+    l = vmap(lambda x: 5*x + 3, r)
+    return l[len(l)/2]

### 2013-11-27 doc/papers/wpmvp14/experiments/map1.py

@@ -1,0 +1,5 @@
+#pythran export map1(int)
+def map1(n):
+    r = range(n)
+    l = vmap(abs, r)
+    return l[len(l)/2]

### 2013-11-27 doc/papers/wpmvp14/experiments/np1.py

@@ -1,0 +1,5 @@
+#pythran export np1(float [])
+import numpy as np
+def np1(x):
+  return np.sum(100.*(x[1:] - x[:-1] ** 2) ** 2
+                + (1. - x[:-1]) ** 2)

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_euler06.sh

@@ -1,5 +1,11 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
 python -m timeit -s 'import euler06 as e; s = e.solve' 's(100001)'
 cc -shared -O3 -std=c99 euler06.c -o libeuler06.so
 python -m timeit -s "from ctypes import cdll ; d = cdll.LoadLibrary('./libeuler06.so') ; s = d.solve" "s(100001)"
 cc -shared -O3 -std=c99 -march=native euler06.c -o libeuler06.so
 python -m timeit -s "from ctypes import cdll ; d = cdll.LoadLibrary('./libeuler06.so') ; s = d.solve" "s(100001)"
+pythran -O3 -march=native euler06.py
+python -m timeit -s 'import euler06 as e; s = e.solve' 's(100001)'
+

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_map0.sh

@@ -1,0 +1,10 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+sed 's/vmap/map/' map0.py > smap0.py
+python -m timeit -s 'from smap0 import map0 as s' 's(1000000)'
+rm -f smap0.py
+pythran -Ofast map0.py
+python -m timeit -s 'from map0 import map0 as s' 's(1000000)'
+pythran -Ofast -DUSE_BOOST_SIMD -march=native map0.py
+python -m timeit -s 'from map0 import map0 as s ' 's(1000000)'

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_map1.sh

@@ -1,0 +1,10 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+sed 's/vmap/map/' map1.py > smap1.py
+python -m timeit -s 'from smap1 import map1 as s' 's(1000000)'
+rm -f smap1.py
+pythran -Ofast map1.py
+python -m timeit -s 'from map1 import map1 as s' 's(1000000)'
+pythran -Ofast -DUSE_BOOST_SIMD -march=native map1.py
+python -m timeit -s 'from map1 import map1 as s ' 's(1000000)'

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_np0.sh

@@ -1,0 +1,8 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+python -m timeit -s 'from np0 import np0 as s ; import numpy as np ; a = np.random.rand(1000000)' 's(a)'
+pythran -Ofast np0.py
+python -m timeit -s 'from np0 import np0 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'
+pythran -Ofast -DUSE_BOOST_SIMD -march=native np0.py
+python -m timeit -s 'from np0 import np0 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_np1.sh

@@ -1,0 +1,8 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+python -m timeit -s 'from np1 import np1 as s ; import numpy as np ; a = np.random.rand(1000000)' 's(a)'
+pythran -Ofast np1.py
+python -m timeit -s 'from np1 import np1 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'
+pythran -Ofast -DUSE_BOOST_SIMD -march=native np1.py
+python -m timeit -s 'from np1 import np1 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_np2.sh

@@ -1,0 +1,8 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+python -m timeit -s 'from np2 import np2 as s ; import numpy as np ; a = np.random.rand(1000000); b = np.random.rand(len(a)); c = np.random.rand(1000000); d = np.random.rand(len(a))' 's(a,b,c,d)'
+pythran -Ofast np2.py
+python -m timeit -s 'from np2 import np2 as s ; import numpy as np ; a = np.random.rand(1000000); b = np.random.rand(len(a)); c = np.random.rand(1000000); d = np.random.rand(len(a))' 's(a,b,c,d)'
+pythran -Ofast -DUSE_BOOST_SIMD -march=native np2.py
+python -m timeit -s 'from np2 import np2 as s ; import numpy as np ; a = np.random.rand(1000000); b = np.random.rand(len(a)); c = np.random.rand(1000000); d = np.random.rand(len(a))' 's(a,b,c,d)'

### 2013-11-27 doc/papers/wpmvp14/experiments/run_xp_sum0.sh

@@ -1,0 +1,10 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+sed 's/vsum/sum/' sum0.py > ssum0.py
+python -m timeit -s 'from ssum0 import sum0 as s' 's(1000000)'
+rm -f ssum0.py
+pythran -Ofast sum0.py
+python -m timeit -s 'from sum0 import sum0 as s' 's(1000000)'
+pythran -Ofast -DUSE_BOOST_SIMD -march=native sum0.py
+python -m timeit -s 'from sum0 import sum0 as s ' 's(1000000)'

### 2013-11-27 doc/papers/wpmvp14/experiments/sum0.py

@@ -1,0 +1,4 @@
+#pythran export sum0(int)
+def sum0(n):
+    r = range(n)
+    return vsum(r)

### 2013-11-27 pythran/pythonic++/core/intrinsics.h

@@ -6,6 +6,10 @@
 #include <complex>
 #include <nt2/include/functions/abs.hpp>
 
+#ifdef USE_BOOST_SIMD
+#include <boost/simd/include/functions/sum.hpp>
+#endif
+
 namespace pythonic {
 
     double const nan = std::numeric_limits<double>::quiet_NaN();
@@ -483,6 +487,33 @@ namespace pythonic {
 
         PROXY(pythonic::__builtin__,map);
 
+        template<class F, class E>
+        core::list<decltype(std::declval<F>()(std::declval<typename E::value_type>()))>
+        vmap(F function, E const & expr) {
+               typedef typename E::value_type T;
+               long n= expr.size();
+               core::list<decltype(std::declval<F>()(std::declval<typename E::value_type>()))> out(n);
+#ifdef USE_BOOST_SIMD
+               long i;
+               typedef typename boost::simd::native<T, BOOST_SIMD_DEFAULT_EXTENSION> vT;
+               static const std::size_t vN = boost::simd::meta::cardinal_of< vT >::value;
+               const long bound = n/vN*vN;
+#endif
+#ifdef USE_BOOST_SIMD
+               for(i=0;i< bound; i+= vN) {
+                   out.store(i, expr.load(i));
+               }
+               for(;i< n; ++i)
+                   out.at(i) = expr.at(i);
+               return out;
+#else
+               for(long i=0 ; i<n; ++i) {
+                   out.at(i) = expr.at(i);
+               }
+               return out;
+#endif
+           }
+        PROXY(pythonic::__builtin__,vmap);
         /* max */
         template <int n, class... Types>
             struct Max;
@@ -829,6 +860,37 @@ namespace pythonic {
             auto sum(std::tuple<Types...> const & t) -> decltype(tuple_sum(t, ::pythonic::int_<sizeof...(Types)-1>())) {
                 return tuple_sum(t, ::pythonic::int_<sizeof...(Types)-1>());
             }
+        template<class E>
+        typename E::value_type
+        vsum(E const & expr) {
+               typedef typename E::value_type T;
+               long n= expr.size();
+               core::list<typename E::value_type> out(n);
+#ifdef USE_BOOST_SIMD
+               long i;
+               typedef typename boost::simd::native<T, BOOST_SIMD_DEFAULT_EXTENSION> vT;
+               static const std::size_t vN = boost::simd::meta::cardinal_of< vT >::value;
+               const long bound = n/vN*vN;
[…]

### 2013-11-27 pythran/pythonic++/core/list.h

@@ -11,6 +11,14 @@
 #include "shared_ref.h"
 #include "slice.h"
 
+#ifdef USE_BOOST_SIMD
+#include <boost/simd/sdk/simd/native.hpp>
+#include <boost/simd/include/functions/unaligned_load.hpp>
+#include <boost/simd/include/functions/unaligned_store.hpp>
+#include <boost/simd/include/functions/load.hpp>
+#include <boost/simd/include/functions/store.hpp>
+#endif
+
 #define DEFAULT_LIST_CAPACITY 16
 
 namespace  pythonic {
@@ -245,6 +253,19 @@ namespace  pythonic {
                 const_reference at( long n ) const {
                     return (*data)[n];
                 }
+                reference at( long n ) {
+                    return (*data)[n];
+                }
+#ifdef USE_BOOST_SIMD
+                template<class L>
+                auto load( L n ) const -> decltype(boost::simd::unaligned_load<boost::simd::native<T, BOOST_SIMD_DEFAULT_EXTENSION>>(data->data(), n)){
+                    return (boost::simd::unaligned_load<boost::simd::native<T, BOOST_SIMD_DEFAULT_EXTENSION>>(data->data(), n));
+                }
+                template<class vT>
+                void store(long n, vT&& v) {
+                    //boost::simd::unaligned_store<typename std::remove_cv<typename std::remove_reference<vT>::type>::type>(v, &(*data)[n], n);
+                }
+#endif
 
                 list<T> operator[]( slice const &s ) const {
                     core::normalized_slice norm = s.normalize(size());

### 2013-11-27 pythran/pythonic++/core/ndarray.h

@@ -627,17 +627,23 @@ namespace  pythonic {
 
                 sliced_ndarray(T const& data, slice const& s) : normalized_slice(s.normalize(data.shape[0])), shape(data.shape), data(data) {
                     shape[0] = normalized_slice::size();
-                    jump = 1;
+                    jump = lower;
                     for(size_t i=1;i<value; ++i)
                         jump*=shape[i];
                 }
 
                 auto at(long i) const -> decltype(data.at(0)) {
-                    return data.at(jump*lower+i*step);
+                    return data.at(jump+i*step);
                 }
+#ifdef USE_BOOST_SIMD
+                /* warning : step must be equal to 1! */
+                auto load(long i) const -> decltype(data.load(jump+i)) {
+                    return data.load(jump+i);
+                }
+#endif
                 long size() const { return (data.size() / data.shape[0]) * shape[0] ; }
-                reference operator[](long i) { return data[jump*lower+i*step]; }
-                const_reference operator[](long i) const { return data[jump*lower+i*step]; }
+                reference operator[](long i) { return data[jump+i*step]; }
+                const_reference operator[](long i) const { return data[jump+i*step]; }
 
                 sliced_ndarray<T> operator[](slice const& s) const {
                     core::normalized_slice norm = s.normalize(normalized_slice::size());

### 2013-11-27 pythran/tables.py

@@ -175,6 +175,7 @@ modules = {
             "list": ReadOnceFunctionIntr(),
             "long_": ConstFunctionIntr(),
             "map": ReadOnceFunctionIntr(),
+            "vmap": ReadOnceFunctionIntr(),
             "max": ReadOnceFunctionIntr(),
             "min": ReadOnceFunctionIntr(),
             "next": FunctionIntr(),  # not const
@@ -191,6 +192,7 @@ modules = {
             "sorted": ConstFunctionIntr(),
             "str": ConstFunctionIntr(),
             "sum": ReadOnceFunctionIntr(),
+            "vsum": ReadOnceFunctionIntr(),
             "tuple": ReadOnceFunctionIntr(),
             "xrange": ConstFunctionIntr(),
             "zip": ReadOnceFunctionIntr(),

### 2013-11-29 doc/papers/wpmvp14/paper.tex
Update experiments

@@
 
+To validate the approach, a number of code snippets presenting vectorization
+opportunities have been selected and benchmarked. The original Python version
+is used as a reference, then code is compiled with Pythran without
+vectorization support and finally with Pythran and vectorization support. Time
+measurements are made using the \texttt{timeit} module.
+
+Experiments are run on a computer running a Linux kernel version 3.10. The
+processors are Intel Core i7, so the AVX instruction set is available. Python
+version is 2.7.5, Pythran version is the \texttt{HEAD} of the \texttt{wpmvp14}
+branch of the official Pythran repository
+\url{https://github.com/serge-sans-paille/pythran}. The backend C++ compiler is
+Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
+vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
+vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
+performances.
+
+\subsection{Vectorization of the \texttt{map} intrisinc}
+
+To evaluate the vectorization of the \texttt{map} intrisinc, two expressions
+are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
+\texttt{r} is a \texttt{ndarray} of random double-precision floating point
+numbers. Table~\ref{tbl:map-timings} summarizes the result. The execution time
+for the Cython version is extremly long due to the ndarray indexing and
+function call overheads in CPython. Thanks to static typing and aggressive
+inlining, Pythran does not suffer from these issues, as shown by its sequential
+execution time. Vectorization is beneficial and yields a $\times2.93$ speedup.
+
+\begin{table}
+
+    \centering
+    \begin{tabular}{|l|r|r|r|}
+        \hline
+                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
+        \hline
+        \texttt{map(abs, r)}                    &   1530ms  & 1.72ms    & 0.587ms   \\
+        \hline
+        \texttt{map(f, r)}    &   141ms   & 1.72ms    & 0.587ms  \\
+        \hline
+
+    \end{tabular}
+    \caption{Timings of various \texttt{map} calls using CPython, Pythran, and Pythran with vectorization.}
+    \label{tbl:map-timings}
+
+\end{table}
+
+\subsection{Vectorization of the \texttt{sum} intrisinc}
+
+To evaluate the vectorization of the \texttt{sum} intrisinc, two expressions
+are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
+\texttt{ndarray} of random double-precision floating point numbers.
+Table~\ref{tbl:sum-timings} summarizes the result. Again, CPython is extremely
+slow. It is interesting to note that the extra array multiplication does not
+impacts the excution time much, simply because it is an operation pefromed at
+the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
+interpreted. The vectorized reductions yields a decent $\times2.41$ speedup
+over the sequential version.
+
+\begin{table}
+
+    \centering
+    \begin{tabular}{|l|r|r|r|}
+        \hline
+                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
+        \hline
+        \texttt{sum(r)}                    &   277ms  & 1.07ms    & 0.416ms   \\
+        \hline
+        \texttt{sum(r * r)}    &   292ms   & 1.09ms    & 0.452ms  \\
+        \hline
+
+    \end{tabular}
+    \caption{Timings of various \texttt{sum} calls using CPython, Pythran, and Pythran with vectorization.}
+    \label{tbl:sum-timings}
+
+\end{table}
+
+\subsection{Vectorization of \texttt{ndarray} point-to-point operations}
+
+To evaluate the vectorization of \texttt{ndarray} point-to-point operations,
+three expressions are used: \texttt{1. / a + a}, the example from
+Listing~\ref{lst:numpy-code} and a computation of the point-to-point arc
+distance  between random values. Again, all inputs are large randomized
+\texttt{ndarray} of double precision floating point numbers.
+Table~\ref{tbl:numpy-timings} summarizes the results. Because all computations
+are done in native space, there is not much speedup when running them through
+Pythran. The only benefits comes from the elimination of temporary arrays and
+loop fusion. As a consequence, the more elements in the \texttt{numpy}
+expression, the more Pythran accelerates the code. The vectorization speedups
+follow the same process, as larger expression tend to increase the operation
+over memory movement ratio.
+
+\begin{table}
+
+    \centering
+    \begin{tabular}{|l|r|r|r|}
+        \hline
+                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
+        \hline
+        \texttt{1. / a + a}                    &   7.48ms  & 6.19ms    & 4.05ms   \\
+        \hline
+        \emph{rosen}                           &   13.9ms   & 2.98ms   & 1.15ms  \\
+        \hline
+        \emph{arc distance}                    &   178ms   & 146ms    & 53.8ms  \\
+        \hline
+
+    \end{tabular}
+    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
+    \label{tbl:numpy-timings}
+
+\end{table}
+
+\subsection{Vectorization of an Euler Problem}
+
+This experiment considers the code from Listing~\ref{lst:euler06-py}.
+
+\begin{table}
+
+    \centering
+    \begin{tabular}{|l|r|r|r|}
+        \hline
+                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
+        \hline
+        \emph{euler06}                    &   12ms   & 0.34ms    & 0.30ms  \\
+        \hline
+
+    \end{tabular}
+    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
+    \label{tbl:numpy-timings}
+
+\end{table}
+
 \acks

### 2013-11-29 doc/papers/wpmvp14/experiments/map0.py

@@ -1,5 +1,4 @@
-#pythran export map0(int)
-def map0(n):
-    r = range(n)
+#pythran export map0(float [])
+def map0(r):
     l = vmap(lambda x: 5*x + 3, r)
     return l[len(l)/2]

### 2013-11-29 doc/papers/wpmvp14/experiments/map1.py

@@ -1,5 +1,4 @@
-#pythran export map1(int)
-def map1(n):
-    r = range(n)
+#pythran export map1(float [])
+def map1(r):
     l = vmap(abs, r)
     return l[len(l)/2]

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_map0.sh

@@ -2,9 +2,9 @@ rm *.so
 export PATH=../../../../scripts:$PATH
 export PYTHONPATH=../../../..
 sed 's/vmap/map/' map0.py > smap0.py
-python -m timeit -s 'from smap0 import map0 as s' 's(1000000)'
+python -m timeit -s 'from smap0 import map0 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
 rm -f smap0.py
-pythran -Ofast map0.py
-python -m timeit -s 'from map0 import map0 as s' 's(1000000)'
-pythran -Ofast -DUSE_BOOST_SIMD -march=native map0.py
-python -m timeit -s 'from map0 import map0 as s ' 's(1000000)'
+pythran -O2 map0.py
+python -m timeit -s 'from map0 import map0 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
+pythran -O2 -DUSE_BOOST_SIMD -march=native map0.py
+python -m timeit -s 'from map0 import map0 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_map1.sh

@@ -2,9 +2,9 @@ rm *.so
 export PATH=../../../../scripts:$PATH
 export PYTHONPATH=../../../..
 sed 's/vmap/map/' map1.py > smap1.py
-python -m timeit -s 'from smap1 import map1 as s' 's(1000000)'
+python -m timeit -s 'from smap1 import map1 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
 rm -f smap1.py
-pythran -Ofast map1.py
-python -m timeit -s 'from map1 import map1 as s' 's(1000000)'
-pythran -Ofast -DUSE_BOOST_SIMD -march=native map1.py
-python -m timeit -s 'from map1 import map1 as s ' 's(1000000)'
+pythran -O2 map1.py
+python -m timeit -s 'from map1 import map1 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
+pythran -O2 -DUSE_BOOST_SIMD -march=native map1.py
+python -m timeit -s 'from map1 import map1 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_np0.sh

@@ -2,7 +2,7 @@ rm *.so
 export PATH=../../../../scripts:$PATH
 export PYTHONPATH=../../../..
 python -m timeit -s 'from np0 import np0 as s ; import numpy as np ; a = np.random.rand(1000000)' 's(a)'
-pythran -Ofast np0.py
+pythran -O2 np0.py
 python -m timeit -s 'from np0 import np0 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'
-pythran -Ofast -DUSE_BOOST_SIMD -march=native np0.py
+pythran -O2 -DUSE_BOOST_SIMD -march=native np0.py
 python -m timeit -s 'from np0 import np0 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_np1.sh

@@ -2,7 +2,7 @@ rm *.so
 export PATH=../../../../scripts:$PATH
 export PYTHONPATH=../../../..
 python -m timeit -s 'from np1 import np1 as s ; import numpy as np ; a = np.random.rand(1000000)' 's(a)'
-pythran -Ofast np1.py
+pythran -O2 np1.py
 python -m timeit -s 'from np1 import np1 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'
-pythran -Ofast -DUSE_BOOST_SIMD -march=native np1.py
+pythran -O2 -DUSE_BOOST_SIMD -march=native np1.py
 python -m timeit -s 'from np1 import np1 as s ;  import numpy as np ; a = np.random.rand(1000000)' 's(a)'

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_np2.sh

@@ -2,7 +2,7 @@ rm *.so
 export PATH=../../../../scripts:$PATH
 export PYTHONPATH=../../../..
 python -m timeit -s 'from np2 import np2 as s ; import numpy as np ; a = np.random.rand(1000000); b = np.random.rand(len(a)); c = np.random.rand(1000000); d = np.random.rand(len(a))' 's(a,b,c,d)'
-pythran -Ofast np2.py
+pythran -O2 np2.py
 python -m timeit -s 'from np2 import np2 as s ; import numpy as np ; a = np.random.rand(1000000); b = np.random.rand(len(a)); c = np.random.rand(1000000); d = np.random.rand(len(a))' 's(a,b,c,d)'
-pythran -Ofast -DUSE_BOOST_SIMD -march=native np2.py
+pythran -O2 -DUSE_BOOST_SIMD -march=native np2.py
 python -m timeit -s 'from np2 import np2 as s ; import numpy as np ; a = np.random.rand(1000000); b = np.random.rand(len(a)); c = np.random.rand(1000000); d = np.random.rand(len(a))' 's(a,b,c,d)'

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_sum0.sh

@@ -2,9 +2,9 @@ rm *.so
 export PATH=../../../../scripts:$PATH
 export PYTHONPATH=../../../..
 sed 's/vsum/sum/' sum0.py > ssum0.py
-python -m timeit -s 'from ssum0 import sum0 as s' 's(1000000)'
+python -m timeit -s 'from ssum0 import sum0 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
 rm -f ssum0.py
-pythran -Ofast sum0.py
-python -m timeit -s 'from sum0 import sum0 as s' 's(1000000)'
-pythran -Ofast -DUSE_BOOST_SIMD -march=native sum0.py
-python -m timeit -s 'from sum0 import sum0 as s ' 's(1000000)'
+pythran -O2 sum0.py
+python -m timeit -s 'from sum0 import sum0 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
+pythran -O2 -DUSE_BOOST_SIMD -march=native sum0.py
+python -m timeit -s 'from sum0 import sum0 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'

### 2013-11-29 doc/papers/wpmvp14/experiments/run_xp_sum1.sh

@@ -1,0 +1,10 @@
+rm *.so
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+sed 's/vsum/sum/' sum1.py > ssum1.py
+python -m timeit -s 'from ssum1 import sum1 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
+rm -f ssum1.py
+pythran -O2 sum1.py
+python -m timeit -s 'from sum1 import sum1 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'
+pythran -O2 -DUSE_BOOST_SIMD -march=native sum1.py
+python -m timeit -s 'from sum1 import sum1 as s; import numpy as np ; r = np.random.rand(1000000)' 's(r)'

### 2013-11-29 doc/papers/wpmvp14/experiments/sum0.py

@@ -1,4 +1,3 @@
-#pythran export sum0(int)
-def sum0(n):
-    r = range(n)
+#pythran export sum0(float [])
+def sum0(r):
     return vsum(r)

### 2013-11-29 doc/papers/wpmvp14/experiments/sum1.py

@@ -1,0 +1,3 @@
+#pythran export sum1(float [])
+def sum1(r):
+    return vsum(r*r)

### 2013-11-29 pythran/pythonic++/core/intrinsics.h

@@ -865,7 +865,6 @@ namespace pythonic {
         vsum(E const & expr) {
                typedef typename E::value_type T;
                long n= expr.size();
-               core::list<typename E::value_type> out(n);
 #ifdef USE_BOOST_SIMD
                long i;
                typedef typename boost::simd::native<T, BOOST_SIMD_DEFAULT_EXTENSION> vT;
@@ -881,7 +880,7 @@ namespace pythonic {
                }
                T p = boost::simd::sum(vp);
                for(;i< n; ++i)
-                   p += expr.at(i);
+                   p = p + expr.at(i);
                return p;
 #else
                for(long i=0 ; i<n; ++i) {

### 2013-11-29 doc/papers/wpmvp14/paper.tex
Update whole version

@@
 
-\title{Vectorizing Python Constructs Using Pythran and Boost SIMD}
+\title{Exploring the Vectorization of Python Constructs Using Pythran and Boost SIMD}
 
@@
 
-The Python language has grown in audience for the past ten years, even reaching
-the world of scientific computations thanks to the \texttt{numpy} module, a
-%% what about the scipy module? It is not really used?
-module that provides a MATLAB-like API. As a consequence, more and more code is
-being written either in pure Python, generally to prototype an application, or
-as a Python and native code mix when performance matters. However, Python
-trades performance for dynamicity and does not particularly shines in terms of
-performance.
+The Python language~\cite{rossum97} has grown in audience for the past ten
+years, even reaching the world of scientific computations~\cite{oliphant2007}
+thanks to the \texttt{numpy} module~\cite{numpyarray2011}, a module that
+provides an efficient multi-dimensionnal array type, and the \texttt{scipy}
+package~\cite{scipy} that provides a MATLAB-like API. As a consequence, more
+and more code is being written either in pure Python, generally to prototype an
+application, or as a Python and native code mix when performance matters.
+However, Python trades performance for dynamicity and does not particularly
+shines in terms of performance.
+
+\subsection{Python Performance Shortcomings}
 
@@
 auto-vectorization makes the C version $\times400$ faster than the Python
-version.~\footnote{To perform more reliable time measurements, \texttt{n} has
-    been increased to 100001 instead of 101. The script used to perform the
-    measures as well as all other experimental data are available at
-\url{https://github.com/serge-sans-paille/pythran/tree/wpmvp14/doc/papers/wpmvp14/experiments}.}
+version. To perform more reliable time measurements, \texttt{n} has been
+increased to 100001 instead of 101.
 
@@
 
-Several approaches have been proposed to solve these issues: using native
-modules, relying on Ahead-Of-Time (AOT) or Just-In-Time (JIT) compilation.
+\subsection{Compiling Python Code}
+
+Several approaches have been proposed to solve the performance issues of
+Python: using native modules, relying on Ahead-Of-Time (AOT) or Just-In-Time
+(JIT) compilation.
 
@@
 Python interpreter using the regular \texttt{import} mechanism. It makes it
-possible to call native code from Python in a transparent way. Typically
-%% It may be others languages, isn't it?
-Python object are unboxed when transferred from Python to C, where efficient
+possible to call native code from Python in a transparent way. Typically Python
+object are unboxed when transferred from Python to C, where efficient
 computations can happen. Then the result is boxed when getting back to the
@@
 a point-to-point operation like the sum of two arrays. However the granularity
-of the operations limits the ratio of \texttt{load}/\texttt{store} versus operation. For
-instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum
-\texttt{a + b + c} is computed using two loops and an intermediate array.
-% It means there are 4 load / stores and only two loops of addition.
+of the operations limits the ratio of \texttt{load}/\texttt{store} versus
+operation. For instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D
+arrays, the sum \texttt{a + b + c} is computed using two loops and an
+intermediate array.  It means there are 4 load / stores and only two additions
+per loop iteration.
 
 AOT compilation uses type inference or type annotation to statically compute
-the type of all (like in Shedskin, Parakeet) or part of (like in Cython or
-Numba) variables, effectively unboxing them to speed-up computations. Using JIT
-compilation, type information is available at runtime and the unboxing decision
-is based on profitability, e.g.\ according to a trace analysis. This is the
-approach of the PyPy project. In both cases, generation of vector instruction
-is possible, but none of the cited project use it. Indeed automatic
-vectorization is a complex topic, and as stated in~\cite{maleki2011}, many cases
-are not yet correctly supported even for mainstream compilers.
+the type of all (like in Shedskin~\cite{shedskin2006},
+Parakeet~\cite{parakeet2012}) or part of (like in Cython~\cite{cython2010} or
+Numba~\cite{numba}) variables, effectively unboxing them to speed-up
+computations. Using JIT compilation, type information is available at runtime
+and the unboxing decision is based on profitability, e.g.\ according to a trace
+analysis. This is the approach of the PyPy project~\cite{pypy2009}. In both
+cases, generation of vector instruction is possible, but none of the cited
+project use it. Indeed automatic vectorization is a complex topic, and as
+stated in~\cite{maleki2011}, many cases are not yet correctly supported even
+for mainstream compilers.
 
-Pythran is also an AOT compiler, which holds the unique property of turning
-Python modules into C++ \emph{meta-programs}. It does not need type annotations
-and keeps the polymorphism of original functions, respecting Python's duck
-typing. Its compilation workflow is described in Figure~\ref{fig:pythran-compiler}.
-Like others AOT compilers, it unboxes all variables for better performance, and
-performs a wide variety of optimizations such as lazy evaluation, constant
-folding, expression fusion etc. It supports core \texttt{numpy} constructs and
-most Python constructs to the notable exception of user classes. Constructs
-such as \texttt{eval}, are not supported.
+\subsection{The Pythran Compiler}
+
+Pythran~\cite{pythran2013, pyhpc2013} is also an AOT compiler, bit it holds the
+unique property of turning Python modules into C++ \emph{meta-programs}. It
+does not need type annotations and keeps the polymorphism of original
+functions, respecting Python's duck typing. Its compilation workflow is
+described in Figure~\ref{fig:pythran-compiler}.  Like others AOT compilers, it
+unboxes all variables for better performance, and performs a wide variety of
+optimizations such as lazy evaluation, constant folding, expression fusion etc.
+It supports core \texttt{numpy} constructs and most Python constructs to the
+notable exception of user classes. Constructs such as \texttt{eval}, are not
+supported either.
 
@@
 
-This paper introduces vectorization of Python programs based on the idea that
+This paper explores the vectorization of Python programs based on the idea that
 many Python constructs already exhibit good vectorization opportunities. The
 goal of the paper switches from automatic vectorization to vectorization of
-high-level Python intrisics, as described in Section~\ref{sec:python-semantic}.
+high-level Python intrinsics, as described in Section~\ref{sec:python-semantic}.
 One of the difficulties of vectorizing Python functions is linked to function
@@
 Section~\ref{sec:meta-vectorization}. The approach is validated using the
-Pythran compiler on several micro-applications from signal processing, physics
-and mathematics presented in Section~\ref{sec:benchs}.
+Pythran compiler on several kernels presented in Section~\ref{sec:benchs}.
 
@@
 \label{sec:meta-vectorization}
-SIMD extensions have been a feature of choice for processor manufacturers for a
-couple of decades. However, programming applications that take advantage of the 
-SIMD extension available on the current target remains a complex task. Programmers 
-that use low-level intrinsics have to deal with a verbose programming style due 
-to the fact that SIMD instructions sets cover a few common functionnalities, requiring
-to bury the initial algorithms in architecture specific implementation details.
+
+Single Instruction Multiple Data (SIMD) extensions have been a feature of
+choice for processor manufacturers for a couple of decades. However,
+programming applications that take advantage of the SIMD extension available on
+the current target remains a complex task. Programmers that use low-level
+intrinsics have to deal with a verbose programming style due to the fact that
+SIMD instructions sets cover a few common functionnalities, requiring to bury
+the initial algorithms in architecture specific implementation details.
 Furthermore, these efforts have to be repeated for every different extension
-that one may want to support, making design and maintenance of such applications 
-very time consuming.
+that one may want to support, making design and maintenance of such
+applications very time consuming.
 
@@
 
-\subsection{\texttt{Load}/\texttt{Store} Protocol}
+\subsection{Interaction with Pythran}
+
+The first attempt to have Pytrhan generate vectorized code directly relied on
+SSE intrinsics. This approach quickly proved to be inappropriate for several
+reasons:
+
+\begin{description}
+
+    \item[portability] Only SSE was supported, moving to AVX or another vector
+        instruction set would require a great deal of engineering;
+
+    \item[completness] Many basic mathematic functions such as cosinus or sinus
+        are not directly available in the target instruction sets;
+
+    \item[typing] Pytrhan manipulates meta-functions with parametric types,
+        while a given instruction set only provide concretes type.
+
+\end{description}
+
+The \boostsimd{} abstraction layer solves these three issues to the expense of
+an increase in compilation time, which still remains acceptable for an AOT
+compiler.
+
+Pythran uses function objects to represent any Python functions, either
+user-functions or functions from the standard library. The former are
+automatically generated from the user code analysis, and the latter are
+hand-coded in a library called \texttt{pythonic++} provided with the compiler.
+For instance the Python function \texttt{def foo(x): return 3 * x} is compiled
+into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
+actual would be more compilcated to handle type inference across several
+statements, but the structure remains valid.
+
+Thanks to this representation, both the Python and C++ function can be called
+on scalar types, \texttt{list} type or any other type that supports
+multiplication by an integer constant. The only difference is that the Python
+version uses \emph{dynamic polymorphism} while the C++ version uses
+\emph{static polymorphism}.
+
+\begin{lstlisting}[language=c++, caption={Simplified Pythran Translation of a Polymorphic Function}, label={lst:simple-pythran-output}]
+struct foo {
+    template<class T>
+    auto operator()(T&& x)
+    -> decltype(3 * x)
+    {
+        return x * 3;
+    }
+};
+\end{lstlisting}
+
+This representation perfectly matches the requirements of \boostsimd, as
+presented in Listing~\ref{lst:square-cxx}. The \texttt{foo} function can be
+used indifferently on scalar types and vector types.
 
@@
 
-A high-level language such as Python exhibit another level of vectorization
+A high-level language such as Python exhibits another level of vectorization
 opportunity in implict loop. Implicit loops can be found in many places, but
@@
 
-\subsubsection{\texttt{sum}}
-% only sum as texttt in a title is not really beautiful
+\subsubsection{The \texttt{sum} Intrinsic}
 
@@
 wether one needs to strictly conforms to the IEEE Standard for Floating-Point
-Arithmetic (IEEE 754).
-% I don't understand what this last sentence is doing here :-s
+Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.
 
@@
 implementation of fallback to a generic one using Substitution Failure
-Is Not An Error (SFINAE).
+Is Not An Error (SFINAE)~\cite{metaprogramming2002}.
 
@@
 vectorization support and finally with Pythran and vectorization support. Time
-measurements are made using the \texttt{timeit} module.
+measurements are made using the \texttt{timeit} module, which runs the piece of
+code a number of time and measure its average execution time. It does this
+three times and picks the best average execution time among those.
 
@@
 
+The scripts used to perform the measures as well as all other experimental data
+used in this paper are available online at
+\url{https://github.com/serge-sans-paille/pythran/tree/wpmvp14/doc/papers/wpmvp14/experiments}.
+
+\subsection{Vectorization of an Euler Problem}
+
+This experiment considers the code from Listing~\ref{lst:euler06-py}. This code
+uses \emph{generator expression} that have not been explored in this paper, so
+it has been modified to use \emph{list comprehension} instead. As a result, a
+lot of temporay array are used, which leads to the performance show on
+Figure~\ref{fig:euler06-timings}. Although Pythran does a decent job at
+optimizing the code, vectorization does not provide significant speedup. This
+is due to the poor computation over memory transfer ratio, and the lack of
+optimization done by Pythran on that aspect for \texttt{list}s. A similar setup
+is evaluated in the context of \texttt{numpy} arrays in
+\S~\ref{sec:bench-numpy} that use temporary array elimination.
+
+\begin{figure}[ht]
+
+    \includegraphics[width=.5\textwidth]{euler06}
+    \caption{Timings of euler06 problem using CPython, Pythran, and Pythran with vectorization.}
+    \label{fig:euler06-timings}
+
+\end{figure}
+
 \subsection{Vectorization of the \texttt{map} intrisinc}
@@
 \texttt{r} is a \texttt{ndarray} of random double-precision floating point
-numbers. Table~\ref{tbl:map-timings} summarizes the result. The execution time
+numbers. Figure~\ref{fig:map-timings} summarizes the result. The execution time
 for the Cython version is extremly long due to the ndarray indexing and
@@
 
-\begin{table}
+\begin{figure}[ht]
 
-    \centering
-    \begin{tabular}{|l|r|r|r|}
-        \hline
-                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
-        \hline
-        \texttt{map(abs, r)}                    &   1530ms  & 1.72ms    & 0.587ms   \\
-        \hline
-        \texttt{map(f, r)}    &   141ms   & 1.72ms    & 0.587ms  \\
-        \hline
+    \includegraphics[width=.5\textwidth]{map_abs}
+    \includegraphics[width=.5\textwidth]{map_fun}
+    \caption{Timings of various \texttt{map} calls using CPython, Pythran, and Pythran with vectorization.}
+    \label{fig:map-timings}
 
-    \end{tabular}
-    \caption{Timings of various \texttt{map} calls using CPython, Pythran, and Pythran with vectorization.}
-    \label{tbl:map-timings}
-
-\end{table}
+\end{figure}
 
@@
 \texttt{ndarray} of random double-precision floating point numbers.
-Table~\ref{tbl:sum-timings} summarizes the result. Again, CPython is extremely
+Figure~\ref{fig:sum-timings} summarizes the result. Again, CPython is extremely
 slow. It is interesting to note that the extra array multiplication does not
@@
 
-\begin{table}
+\begin{figure}[ht]
 
-    \centering
-    \begin{tabular}{|l|r|r|r|}
-        \hline
-                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
-        \hline
-        \texttt{sum(r)}                    &   277ms  & 1.07ms    & 0.416ms   \\
-        \hline
-        \texttt{sum(r * r)}    &   292ms   & 1.09ms    & 0.452ms  \\
-        \hline
+    \includegraphics[width=.5\textwidth]{sum}
+    \includegraphics[width=.5\textwidth]{sum_square}
+    \caption{Timings of various \texttt{sum} calls using CPython, Pythran, and Pythran with vectorization.}
+    \label{fig:sum-timings}
 
-    \end{tabular}
-    \caption{Timings of various \texttt{sum} calls using CPython, Pythran, and Pythran with vectorization.}
-    \label{tbl:sum-timings}
-
-\end{table}
+\end{figure}
 
 \subsection{Vectorization of \texttt{ndarray} point-to-point operations}
+\label{sec:bench-numpy}
 
@@
 \texttt{ndarray} of double precision floating point numbers.
-Table~\ref{tbl:numpy-timings} summarizes the results. Because all computations
+Figure~\ref{fig:numpy-timings} summarizes the results. Because all computations
 are done in native space, there is not much speedup when running them through
@@
 
-\begin{table}
+\begin{figure}[ht]
 
-    \centering
-    \begin{tabular}{|l|r|r|r|}
-        \hline
-                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
-        \hline
-        \texttt{1. / a + a}                    &   7.48ms  & 6.19ms    & 4.05ms   \\
-        \hline
-        \emph{rosen}                           &   13.9ms   & 2.98ms   & 1.15ms  \\
-        \hline
-        \emph{arc distance}                    &   178ms   & 146ms    & 53.8ms  \\
-        \hline
+    \includegraphics[width=.5\textwidth]{numexpr}
+    \includegraphics[width=.5\textwidth]{rosen}
+    \includegraphics[width=.5\textwidth]{arc}
+    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
+    \label{fig:numpy-timings}
 
-    \end{tabular}
-    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
-    \label{tbl:numpy-timings}
+\end{figure}
 
-\end{table}
+\section*{Conclusion}
 
-\subsection{Vectorization of an Euler Problem}
+This paper explores the design and implementation of a vectorizer for the
+Python language. It targets high-level constructs from the language and from
+the \texttt{numpy} module that already have a vector semantic, efficiently
+vectorizing them thanks to the combination of Pythran, a Python to C++ compiler
+that turns Python modules into C++ meta-programs and \boostsimd, a C++ library
+to program SIMD architectures.
 
-This experiment considers the code from Listing~\ref{lst:euler06-py}.
+Common Python constructions such as \emph{list comprehension}, the \texttt{map}
+and \texttt{sum} intrinsics and array operations from \texttt{numpy} are
+considered and the approach is validated on several kernels, showing
+promising results.
 
-\begin{table}
-
-    \centering
-    \begin{tabular}{|l|r|r|r|}
-        \hline
-                                               & \textbf{CPython}   & \textbf{Pythran}   & \textbf{Pythran+AVX} \\
-        \hline
-        \emph{euler06}                    &   12ms   & 0.34ms    & 0.30ms  \\
-        \hline
-
-    \end{tabular}
-    \caption{Timings of various \texttt{numpy} expressions using CPython, Pythran, and Pythran with vectorization.}
-    \label{tbl:numpy-timings}
-
-\end{table}
+Future works include the case of Python generators, that have the benfit of not
+creating intermediate container to hold the data, but require more analyse in
+the context of vectorization. Handling \texttt{list} operations in an efficient
+manner as done for \texttt{numpy} arrays is also a required developpment axis.
 
@@
 
-Acknowledgments, if needed.
+Several students from Télécom Bretagne have been working on this project, the
+authors would like to thank Alan Raynaud, Adrien Merlini and Yuanchang Peng for
+their contribution to the Pytrhan project. They are also in debt to Mehdi
+Aminin for his careful code reviews.
 
 % We recommend abbrvnat bibliography style.
-\cite{*}
 

### 2013-11-29 doc/papers/wpmvp14/bench.sh

@@ -1,0 +1,89 @@
+#!/bin/bash
+cat > map_abs.dat << EOF
+CPython 1530
+Pythran 1.72
+Pythran+AVX 0.587
+EOF
+
+cat > map_fun.dat << EOF
+CPython 141
+Pythran 1.72
+Pythran+AVX 0.587
+EOF
+
+cat > sum.dat << EOF
+CPython 277
+Pythran 1.07
+Pythran+AVX 0.416
+EOF
+
+cat > sum_square.dat << EOF
+CPython 292
+Pythran 1.09
+Pythran+AVX 0.452
+EOF
+
+cat > numexpr.dat << EOF
+CPython 7.48
+Pythran 6.19
+Pythran+AVX 4.05
+EOF
+
+cat > rosen.dat << EOF
+CPython 13.9
+Pythran 2.98
+Pythran+AVX 1.15
+EOF
+
+cat > arc.dat << EOF
+CPython 178
+Pythran 146
+Pythran+AVX 53.8
+EOF
+
+cat > euler06.dat << EOF
+CPython 12
+Pythran 0.34
+Pythran+AVX 0.30
+EOF
+
+function logplot {
+gnuplot << EOF
+set terminal pdf
+set logscale y
+set grid
+set style histogram
+set style data histogram
+set output "$1.pdf"
+set ylabel "Elapsed Time(s)"
+set xtics rotate by -45
+set style fill solid 1.00 border 0
+set title "$2"
+plot '$1.dat' using 2:xtic(1) ti ""
+EOF
+}
+function plot {
+gnuplot << EOF
+set terminal pdf
+#set logscale y
+set grid
+set style histogram
+set style data histogram
+set output "$1.pdf"
+set ylabel "Elapsed Time(s)"
+set xtics rotate by -45
+set yrange [0:]
+set style fill solid 1.00 border 0
+set title "$2"
+plot '$1.dat' using 2:xtic(1) ti ""
+EOF
+}
+
+logplot map_abs 'mapping abs on a ndarray'
+logplot map_fun 'mapping a linear function on a ndarray'
+logplot sum 'taking the sum of the elements of an ndarray'
+logplot sum_square 'taking the sum of the square of the elements of an ndarray'
+plot numexpr 'ternary numpy expression'
+plot rosen 'rosen function'
+plot arc 'arc distances'
+logplot euler06 'euler06 problem'

### 2013-11-29 doc/papers/wpmvp14/paper.tex
+Spellchecking

@@
 thanks to the \texttt{numpy} module~\cite{numpyarray2011}, a module that
-provides an efficient multi-dimensionnal array type, and the \texttt{scipy}
+provides an efficient multi-dimensional array type, and the \texttt{scipy}
 package~\cite{scipy} that provides a MATLAB-like API. As a consequence, more
@@
 
-    \item[dynamic binding] each variable access is made through a dictionnary
+    \item[dynamic binding] each variable access is made through a dictionary
         look-up to bind variable names to variable instances~;
@@
 intrinsics have to deal with a verbose programming style due to the fact that
-SIMD instructions sets cover a few common functionnalities, requiring to bury
+SIMD instructions sets cover a few common functionalities, requiring to bury
 the initial algorithms in architecture specific implementation details.
@@
 Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
-provides both expresiveness and performance by using generic programming 
+provides both expressiveness and performance by using generic programming 
 to handle vectorization. \boostsimd{} does not only provide a portable 
@@
 Moreover, as \boostsimd{} provides default implementation of all its function for scalar values,
-we can rely on polymorphic callable object to write a mixed scalar/SIMD call to algortihm
+we can rely on polymorphic callable object to write a mixed scalar/SIMD call to algorithm
 like \texttt{transform}. By using the \texttt{boost::simd::transform} algorithm, code
-handlign alignment, scalar prologue and epilogue is generated.
+handling alignment, scalar prologue and epilogue is generated.
 
@@
 
-The first attempt to have Pytrhan generate vectorized code directly relied on
+The first attempt to have Pythran generate vectorized code directly relied on
 SSE intrinsics. This approach quickly proved to be inappropriate for several
@@
 
-    \item[completness] Many basic mathematic functions such as cosinus or sinus
+    \item[completness] Many basic mathematic functions such as cos or sin
         are not directly available in the target instruction sets;
@@
 into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
-actual would be more compilcated to handle type inference across several
+actual would be more complicated to handle type inference across several
 statements, but the structure remains valid.
@@
 A high-level language such as Python exhibits another level of vectorization
-opportunity in implict loop. Implicit loops can be found in many places, but
+opportunity in implicit loop. Implicit loops can be found in many places, but
 this paper focuses on \emph{list comprehension} and \emph{generator
@@
 The \texttt{sum} intrinsic is a typical case of potential vectorization of an
-implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
+implicit loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
 value} and its behavior is equivalent to the Python code in
 Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
-wether one needs to strictly conforms to the IEEE Standard for Floating-Point
+whether one needs to strictly conforms to the IEEE Standard for Floating-Point
 Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.
@@
 
-Appart from standard compliancy, to be able to efficiently vectorize this code in C++, one must check that:
+Apart from standard compliance, to be able to efficiently vectorize this code in C++, one must check that:
 
@@
 unboxing. For instance, any code that does not match it will fail to compile
-using Pythran or Shed-Skin. It is also satisified for the \texttt{ndarray}
+using Pythran or Shed-Skin. It is also satisfied for the \texttt{ndarray}
 container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
@@
 
-As function have to be generic but some containers do not match requierement
+As function have to be generic but some containers do not match required
 properties, it is possible to select a vectorized
-implementation of fallback to a generic one using Substitution Failure
+implementation of fall-back to a generic one using Substitution Failure
 Is Not An Error (SFINAE)~\cite{metaprogramming2002}.
@@
 the elements non multiple of 3 are taken into account. Let us call
-\emph{comprehesion rule} the expression on the left of the
+\emph{comprehension rule} the expression on the left of the
 \texttt{for}.
@@
 List comprehension can be translated into a combination of \texttt{map} and
-(optionnaly) \texttt{filter}, by transformation of the comprehension rule into
+(optionally) \texttt{filter}, by transformation of the comprehension rule into
 a lambda function. For instance the previous example can be re-written as
@@
 
-Python also provide same features for set and dictionnary.
-
-Set comprehension can also be rewritten in the form of
-generator expression using the eponym function: \texttt{\{ x + 3 for x in l\}}
-is equivalent to \texttt{set(x + 3 for x in l)}.
-
-Dictionnary comprehension and multi-level comprehension (i.e.\ containing
+Python also provide same features for set and dictionary.  Set comprehension
+can be rewritten in the form of generator expression using the eponym function:
+\texttt{\{ x + 3 for x in l\}} is equivalent to \texttt{set(x + 3 for x in l)}.
+Dictionary comprehension and multi-level comprehension (i.e.\ containing
 multiple \texttt{for} clauses) are not considered in this preliminary work.
@@
 scientific community. In a nutshell, this is a thin Python wrapper above a raw
-array that provides an array language embeded into Python. An example of
+array that provides an array language embedded into Python. An example of
 \texttt{numpy} code is show on Listing~\ref{lst:numpy-code}.
@@
 Common features of array language include indexing, slicing, point-to-point
-operation etc. The latter is traditionnaly a good source of vectorization.
+operation etc. The latter is traditionally a good source of vectorization.
 However, because of the module layer, \texttt{numpy} always create a new array
@@
 \texttt{t0} and \texttt{t1}. Fortunately, it is possible to apply \emph{forward
-substitution} to fallback to the expression from Listing~\ref{lst:numpy-code}
-and avoid the creattion of any temporary.
+substitution} to fall-back to the expression from Listing~\ref{lst:numpy-code}
+and avoid the creation of any temporary.
 
@@
 Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
-loop, that can be explicitly vectorized using Bosst.SIMD in a similar manner as
+loop, that can be explicitly vectorized using \boostsimd in a similar manner as
 the \texttt{map} and \texttt{sum} intrinsics.
@@
 from the same location, as in \texttt{1. / a + a}, where \texttt{a} is loaded
-twice on each loop iteration. Fortunately, the backend C++ compiler is capable
+twice on each loop iteration. Fortunately, the back-end C++ compiler is capable
 of detecting the redundant load and perform \emph{redundant load store
@@
 branch of the official Pythran repository
-\url{https://github.com/serge-sans-paille/pythran}. The backend C++ compiler is
+\url{https://github.com/serge-sans-paille/pythran}. The back-end C++ compiler is
 Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
@@
 it has been modified to use \emph{list comprehension} instead. As a result, a
-lot of temporay array are used, which leads to the performance show on
+lot of temporary array are used, which leads to the performance show on
 Figure~\ref{fig:euler06-timings}. Although Pythran does a decent job at
@@
 
-\subsection{Vectorization of the \texttt{map} intrisinc}
+\subsection{Vectorization of the \texttt{map} intrinsic}
 
-To evaluate the vectorization of the \texttt{map} intrisinc, two expressions
+To evaluate the vectorization of the \texttt{map} intrinsic, two expressions
 are used: \texttt{map(abs, r)} and \texttt{map(lambda x: 5 * x + 3, r)}, where
@@
 numbers. Figure~\ref{fig:map-timings} summarizes the result. The execution time
-for the Cython version is extremly long due to the ndarray indexing and
+for the CPython version is extremely long due to the \texttt{ndarray} indexing and
 function call overheads in CPython. Thanks to static typing and aggressive
@@
 
-\subsection{Vectorization of the \texttt{sum} intrisinc}
+\subsection{Vectorization of the \texttt{sum} intrinsic}
 
-To evaluate the vectorization of the \texttt{sum} intrisinc, two expressions
+To evaluate the vectorization of the \texttt{sum} intrinsic, two expressions
 are used: \texttt{sum(r)} and \texttt{sum(r * r)}, where \texttt{r} is a
@@
 slow. It is interesting to note that the extra array multiplication does not
-impacts the excution time much, simply because it is an operation pefromed at
+impacts the execution time much, simply because it is an operation performed at
 the native level by the \texttt{numpy} module, while the \texttt{sum} is fully
@@
 
-Future works include the case of Python generators, that have the benfit of not
+Future works include the case of Python generators, that have the benefit of not
 creating intermediate container to hold the data, but require more analyse in
 the context of vectorization. Handling \texttt{list} operations in an efficient
-manner as done for \texttt{numpy} arrays is also a required developpment axis.
+manner as done for \texttt{numpy} arrays is also a required development axis.
 
@@
 
-Several students from Télécom Bretagne have been working on this project, the
+Several students from T{\'e}l{\'e}com Bretagne have been working on this project, the
 authors would like to thank Alan Raynaud, Adrien Merlini and Yuanchang Peng for
-their contribution to the Pytrhan project. They are also in debt to Mehdi
-Aminin for his careful code reviews.
+their contribution to the Pythran project. They are also in debt to Mehdi
+Amini for his careful code reviews.
 

### 2013-11-29 doc/papers/wpmvp14/paper.tex

@@
 substitution} to fall-back to the expression from Listing~\ref{lst:numpy-code}
-and avoid the creation of any temporary.
+and avoid the creation of any temporary in some case detected with static analysis
+of the source code.
 
@@
 
+\subsection{Vectorization of \texttt{ndarray} point-to-point operations with
+intermediate array}
+
+To have good performance even with intermediate array, we use forward substitution
+detect through a static analysis. It can be done only if it is not use multipe time
+--- as it would lead to compute the same thing many times --- and if values used
+to compute it are not changed between definition and use.
+This detection transform 3 SIMD Loop computation in one intensive computation in
+the case of Listing~\ref{numpy-code-fs}
+
+\begin{figure}[ht]
+
+    \includegraphics[width=.5\textwidth]{rosen_fs}
+    \caption{Timings of rosen computation with mutlpiple assignment using CPython, Pythran, Pythran with vectorization, Pythran with vectorization and forward substitution.}
+    \label{fig:numpy-timings}
+
+\end{figure}
+
 \section*{Conclusion}

### 2013-11-29 doc/papers/wpmvp14/paper.tex
++Proof reading

--- 2013-11-29 doc/papers/wpmvp14/paper.tex
@@
     The Python language is highly dynamic, most notably due to late binding. As
-    a consequence, program run using Python typically run an order of magnitude
-    slower than their C counterpart. It is also a high level languages whose
+    a consequence, programs run using Python typically run an order of magnitude
+    slower than their C counterpart. It is also a high level language whose
     semantic can be made more static without much change from a user point of
@@
 
-For instance, the natural solution to Project Euler sixth problem ``Find the
-difference between the sum of the squares of the first one hundred natural
-numbers and the square of the sum`` can be coded as in
-Listing~\ref{lst:euler06-py} in Python and as in Listing~\ref{lst:euler06-c} in
-C. A comparison of the performance of the Python function and the C function
-called through the \texttt{ctypes} module shows that the C version runs more
-than $\times250$ faster than the Python version. Enabling compiler
-auto-vectorization makes the C version $\times400$ faster than the Python
-version. To perform more reliable time measurements, \texttt{n} has been
-increased to 100001 instead of 101.
+To illustrate the relatively bad performance of Python for numerical
+computations, let us consider the sixth problem from the Project Euler: ``Find
+the difference between the sum of the squares of the first one hundred natural
+numbers and the square of the sum``. A straight-forward solution can be coded
+as in Listing~\ref{lst:euler06-py} in Python or as in
+Listing~\ref{lst:euler06-c} in C. A comparison of the performance of the Python
+function and the C function called through the \texttt{ctypes} module shows
+that the C version runs more than $\times250$ faster than the Python version.
+Enabling compiler auto-vectorization even makes the C version $\times400$
+faster than the Python version (Note that to perform more reliable time
+measurements, the size of the problem has been increased to 100001 instead of
+101).
 
-There are two main reason for the poor performance of Python:
+There are two main reasons for the poor performance of Python:
 
@@
     \item[dynamic binding] each variable access is made through a dictionary
-        look-up to bind variable names to variable instances~;
+        look-up that binds variable names to variable instances;
 
@@
 nature of an operation is unknown to the interpreter until its execution, and
-data are not contiguous in memory.
+data are not even contiguous in memory.
 
@@
 
-Several approaches have been proposed to solve the performance issues of
-Python: using native modules, relying on Ahead-Of-Time (AOT) or Just-In-Time
-(JIT) compilation.
+Several approaches have been proposed to solve the performance issue of Python:
+using native modules, relying on Ahead-Of-Time (AOT) compilation or
+Just-In-Time (JIT) compilation.
 
@@
 possible to call native code from Python in a transparent way. Typically Python
-object are unboxed when transferred from Python to C, where efficient
+object are unboxed when transfered from Python to C, where efficient
 computations can happen. Then the result is boxed when getting back to the
@@
 by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
-contiguous unboxed elements, and provides a great deal of high-level functions
-to manipulate them. All the computation-intensive part is done in pure C.
-Vectorization can eventually happen at this level, for instance when performing
-a point-to-point operation like the sum of two arrays. However the granularity
-of the operations limits the ratio of \texttt{load}/\texttt{store} versus
-operation. For instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D
-arrays, the sum \texttt{a + b + c} is computed using two loops and an
-intermediate array.  It means there are 4 load / stores and only two additions
-per loop iteration.
+contiguous unboxed elements, say single precision floating point numbers, and
+provides a great deal of high-level functions to manipulate them. All the
+computation-intensive part is done in pure C.  Vectorization can eventually
+happen at this level, for instance when performing a point-to-point operation
+like the sum of two arrays. However the granularity of the operations limits
+the ratio of \texttt{load}/\texttt{store} versus operation. For instance, if
+\texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum \texttt{a + b +
+c} is computed using two loops and an intermediate array. It means there are 4
+load / stores and only two additions per loop iteration and vectorization is
+unlikely to be profitable.
 
@@
 cases, generation of vector instruction is possible, but none of the cited
-project use it. Indeed automatic vectorization is a complex topic, and as
+project uses it. Indeed automatic vectorization is a complex topic, and as
 stated in~\cite{maleki2011}, many cases are not yet correctly supported even
-for mainstream compilers.
+for mainstream C/C++/FORTRAN compilers.
 
@@
 
-Pythran~\cite{pythran2013, pyhpc2013} is also an AOT compiler, bit it holds the
-unique property of turning Python modules into C++ \emph{meta-programs}. It
-does not need type annotations and keeps the polymorphism of original
-functions, respecting Python's duck typing. Its compilation workflow is
-described in Figure~\ref{fig:pythran-compiler}.  Like others AOT compilers, it
-unboxes all variables for better performance, and performs a wide variety of
-optimizations such as lazy evaluation, constant folding, expression fusion etc.
-It supports core \texttt{numpy} constructs and most Python constructs to the
-notable exception of user classes. Constructs such as \texttt{eval}, are not
-supported either.
+Pythran~\cite{pythran2013, pyhpc2013} is also an AOT compiler, but it holds the
+unique property of turning Python modules into C++11 \emph{meta-programs}. It
+does not need any type annotations during translation and keeps the
+polymorphism of original functions, respecting Python's duck typing. Its
+compilation work-flow is described in Figure~\ref{fig:pythran-compiler}: Python
+source is turned into generic C++11 code that is then instantiated for the
+given types provided as extra annotations.  Like other AOT compilers, it
+unboxes all variables for better performance, but also performs a wide variety
+of optimizations such as lazy evaluation, constant folding, expression fusion
+etc.  It supports core \texttt{numpy} constructs and most Python constructs, to
+the notable exception of user classes. Instructions such as \texttt{eval},
+\texttt{exec} or \texttt{getattr} are not supported either.
 
@@
 
-\caption{Pythran compiler workflow.}
+\caption{Pythran compiler work-flow.}
 \label{fig:pythran-compiler}
@@
 This paper explores the vectorization of Python programs based on the idea that
-many Python constructs already exhibit good vectorization opportunities. The
-goal of the paper switches from automatic vectorization to vectorization of
-high-level Python intrinsics, as described in Section~\ref{sec:python-semantic}.
-One of the difficulties of vectorizing Python functions is linked to function
-polymorphism. Vectorization in the context of C++ meta-programs is examined in
-Section~\ref{sec:meta-vectorization}. The approach is validated using the
-Pythran compiler on several kernels presented in Section~\ref{sec:benchs}.
+many Python constructs already exhibit good vectorization opportunities. Thus
+the goal of the compiler switches from automatic vectorization to vectorization
+of high-level Python intrinsics. One of the difficulties of vectorizing Python
+functions is linked to function polymorphism: vectorization in the context of
+C++ meta-programs is first examined in Section~\ref{sec:meta-vectorization},
+and Python constructs that can benefit from vectorization are then described in
+Section~\ref{sec:python-semantic}. The approach is validated using the Pythran
+compiler on several kernels presented in Section~\ref{sec:benchs}.
 
@@
 to handle vectorization. \boostsimd{} does not only provide a portable 
-way but also enable the use of common programming idioms when designing 
+way but also enables the use of common programming idioms when designing 
 SIMD-aware algorithms. Two components are provided:
@@
     \item[portability] Only SSE was supported, moving to AVX or another vector
-        instruction set would require a great deal of engineering;
+        instruction set would have required a great deal of engineering;
 
     \item[completness] Many basic mathematic functions such as cos or sin
-        are not directly available in the target instruction sets;
+        were not directly available in the target instruction sets;
 
-    \item[typing] Pytrhan manipulates meta-functions with parametric types,
-        while a given instruction set only provide concretes type.
+    \item[typing] Pythran manipulates meta-functions with parametric types,
+        while a given instruction set only provides concretes type.
 
@@
 hand-coded in a library called \texttt{pythonic++} provided with the compiler.
-For instance the Python function \texttt{def foo(x): return 3 * x} is compiled
+As an example, the Python function \texttt{def foo(x): return 3 * x} is compiled
 into a code similar to the one in Listing~\ref{lst:simple-pythran-output}. The
-actual would be more complicated to handle type inference across several
+actual is slightly more complicated to handle type inference across several
 statements, but the structure remains valid.
@@
 
-This representation perfectly matches the requirements of \boostsimd, as
-presented in Listing~\ref{lst:square-cxx}. The \texttt{foo} function can be
-used indifferently on scalar types and vector types.
+A critical point is that this representation perfectly matches the requirements
+of \boostsimd, as presented in Listing~\ref{lst:square-cxx}. The \texttt{foo}
+function can be used indifferently on scalar types and vector types. This paves
+the way for automatic vectorization of high level idioms.
 
@@
 
-A high-level language such as Python exhibits another level of vectorization
-opportunity in implicit loop. Implicit loops can be found in many places, but
-this paper focuses on \emph{list comprehension} and \emph{generator
-expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
-through the \texttt{numpy} module.
+However, a high-level language such as Python exhibits another level of
+vectorization opportunity: implicit loops. Implicit loops can be found in many
+places, but this paper focuses on \emph{list comprehension} and \emph{set
+comprehension}, the \emph{map} and \emph{sum} intrinsics, and \emph{array
+operations} through the \texttt{numpy} module.
 
@@
 whether one needs to strictly conforms to the IEEE Standard for Floating-Point
-Arithmetic (IEEE 754), vectorization of this reduction is a comm case of \texttt{fold}.
+Arithmetic (IEEE 754), vectorization of this reduction is a valid application
+of \boostsimd's \texttt{fold}.
 
@@
 
-Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
+Condition~\ref{enu:same-types} is a prerequisite for static typing and efficient
 unboxing. For instance, any code that does not match it will fail to compile
-using Pythran or Shed-Skin. It is also satisfied for the \texttt{ndarray}
-container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
+using Pythran or Shed-Skin. It is always satisfied for the \texttt{ndarray}
+container as it wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
 checked at compile time. Python supports two native scalar types candidates for
 vectorization: \texttt{int} (64 bits integer) and \texttt{float}
-(double-precision floating point numbers). The \texttt{numpy} model add supports for a
+(double-precision floating point numbers). The \texttt{numpy} module adds support for a
 wide range of signed and unsigned integers as well as single, double or
@@
 As function have to be generic but some containers do not match required
-properties, it is possible to select a vectorized
-implementation of fall-back to a generic one using Substitution Failure
-Is Not An Error (SFINAE)~\cite{metaprogramming2002}.
+properties, it is necessary to select a vectorized implementation of fall back
+to a generic one using Substitution Failure Is Not An Error
+(SFINAE)~\cite{metaprogramming2002}. Vectorized versions are implement in this manner.
 
@@
 template<class T, class V=long>
-typename std::enable_if<is_vectorizable<T>::value,
+typename std::enable_if<
+  is_vectorizable<T>::value,
   typename T::value_type
@@
   typedef typename T::value_type E;
-  // vE is a vector type of vsize elements of type E
+  // vE is a vector type of vsize elements
   typedef typename simd::native<E,
@@
 
-  // vector_value is for the parallel reduction
+  // parallel reduction
   vE vector_value = simd::splat<vE>(E());
@@
 instructions, e.g.\ trigonometric functions, but they can be provided as
-composite instructions. Some are just not available, as quadruple precision
-float addition. Starting from this set of vectorizable function, one can use
-the \texttt{;} as a closure binary relation to compute a subset of all
-vectorizable functions. Said otherwise, any sequence of operation that only
-involve vector operations is a vector operation and a function whose body is a
-sequence of vector operation or vectorizable functions is a vectorizable
-function itself.
+composite instructions, as done by \boostsimd. Some are just not available, as
+quadruple precision float addition. Starting from this set of vectorizable
+function, one can use the \texttt{;} as a closure binary relation to compute a
+subset of all vectorizable functions. Said otherwise, any sequence of operation
+that only involve vector operations is a vector operation and a function whose
+body is a sequence of vector operation or vectorizable functions is a
+vectorizable function itself.
 
-The preceding statement is only valid if the function's body only references
-vector functions and formal parameters, as the function must be valid for both
-scalar types and vector types. Note that this perfectly matches the translation
-of polymorphic Python functions into C++ meta-functions, as detailed in
-Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}. The core idea is that
-the \texttt{square} function can be called indifferently on floating point
-values or on vector of floating point values. Through (automatic) template
-instantiation, the C++ compiler generates the actual scalar and vector version
-from the same meta-function. As Pythran always translates Python's polymorphic
-function in that form, writing a vector version of the \texttt{map} intrinsic is
-slightly more difficult than for the \texttt{sum} intrinsic due to its variadic
-number of arguments, but it also relies on the \texttt{load}/\texttt{store}
-protocol and SFINAE to select it.
+The preceding statement is only valid if the body of the function only
+references vector functions and formal parameters, as the function must be
+valid for both scalar types and vector types. Note that this perfectly matches
+the translation of polymorphic Python functions into C++ meta-functions, as
+detailed in Listings~\ref{lst:square-python} and~\ref{lst:square-cxx}. The core
+idea is that the \texttt{square} function can be called indifferently on
+floating point values or on vector of floating point values. Through
+(automatic) template instantiation, the C++ compiler generates the actual
+scalar and vector version from the same meta-function. As Pythran always
+translates Python's polymorphic function in that form, writing a vector version
+of the \texttt{map} intrinsic becomes possible. It is slightly more technical
+than for the \texttt{sum} intrinsic due to its variadic number of arguments,
+but it relies on the same concepts.
 
@@
 
-Generator expression is somehow similar to list comprehension, to the notable
-exception that the value returned is an \emph{iterator adaptator} on the
-container argument. In the context of vectorization, a generator expression
-supports the vector protocol if the associated container also supports it. In
-that case the  \emph{iterator adaptator} implements the vector protocol by
-applying the comprehension rule to each vector element from this container, in a
-streaming fashion.
+% Generator expression is somehow similar to list comprehension, to the notable
+% exception that the value returned is an \emph{iterator adaptator} on the
+% container argument. In the context of vectorization, a generator expression
+% supports the vector protocol if the associated container also supports it. In
+% that case the  \emph{iterator adaptator} implements the vector protocol by
+% applying the comprehension rule to each vector element from this container, in a
+% streaming fashion.
 
@@
 \texttt{\{ x + 3 for x in l\}} is equivalent to \texttt{set(x + 3 for x in l)}.
-Dictionary comprehension and multi-level comprehension (i.e.\ containing
-multiple \texttt{for} clauses) are not considered in this preliminary work.
+
+Dictionary comprehension, generator expressions and multi-level comprehension
+(i.e.\ containing multiple \texttt{for} clauses) are not considered in this
+paper.
 
@@
 The \texttt{numpy} module played a major role in the adoption of Python by the
-scientific community. In a nutshell, this is a thin Python wrapper above a raw
+scientific community. In a nutshell, it is a thin Python wrapper above a raw
 array that provides an array language embedded into Python. An example of
@@
 Common features of array language include indexing, slicing, point-to-point
-operation etc. The latter is traditionally a good source of vectorization.
-However, because of the module layer, \texttt{numpy} always create a new array
-to hold intermediate result. For instance in the expression \texttt{(1 -
-x[:-1]) ** 2)} two array are created ---slicing does not create an array but a
-view. The implied extra cost would hide the benefits of vectorization.
+operation etc. Point-to-point operations are traditionally a good source of
+vectorization. However, because of the mandatory module layer, \texttt{numpy}
+always create a new array to hold intermediate result. For instance in the
+expression \texttt{(1 - x[:-1]) ** 2)} two array are created ---slicing does
+not create an array but a view. The implied extra cost would hide the benefits
+of vectorization.
 
 The typical way to solve this issue in C++ is to use \emph{expression
-templates}, a technique described in~\cite{} that uses the C++ type system to
-perform lazy evaluation of an expression. However, this approach is limited to
-a single expression. Listing~\ref{lst:numpy-code-fs} illustrates the situation:
-expression templates would not prevent the creation of temporary arrays
-\texttt{t0} and \texttt{t1}. Fortunately, it is possible to apply \emph{forward
-substitution} to fall-back to the expression from Listing~\ref{lst:numpy-code}
-and avoid the creation of any temporary.
+templates}, a technique described in~\cite{expression_templates} that uses the
+C++ type system to perform lazy evaluation of an expression. Yet, this
+approach is limited to a single expression, as illustrated by Listing~\ref{lst:numpy-code-fs}: in that case expression templates would not prevent the creation
+of temporary arrays \texttt{t0} and \texttt{t1}. Fortunately, it is possible to
+apply \emph{forward substitution} to fall-back to the expression from
+Listing~\ref{lst:numpy-code} and avoid the creation of any temporary in some
+case detected with static analysis of the source code.
 
@@
 
-It is possible to hook in the expression template to add vectorization support.
+It is possible to hook in the expression template code to add vectorization support.
 Indeed the expression from Listing~\ref{lst:numpy-code} hides a single vector
@@
 
+\subsection{Vectorization of \texttt{ndarray} point-to-point operations with
+intermediate array}
+
+To have good performance even with intermediate array, Pythran uses forward
+substitution. This is not always beneficial if the substituted variable is used multiple time ---it
+would lead to the computation of the expression several times--- and it is only legal if no dependency are violated. This detection
+transforms three SIMD loop computations in one intensive computation in the case of Listing~\ref{lst:numpy-code}. Figure~\ref{fig:numpy-code-fs} shows the benefits of the approach.
+
+\begin{figure}[ht]
+
+    \includegraphics[width=.5\textwidth]{rosen_fs}
+    \caption{Timings of rosen computation with multiple assignment using CPython, Pythran, Pythran with vectorization, Pythran with vectorization and forward substitution.}
+    \label{fig:numpy-code-fs}
+
+\end{figure}
+
 \section*{Conclusion}

### 2013-11-29 doc/papers/wpmvp14/bench.sh

@@ -51,7 +51,7 @@ cat > rosen_fs.dat << EOF
 CPython 1.64
 Pythran 0.49
 Pythran+AVX 0.41
-Pythran+AVX+forward_substitution 0.29
+Pythran+AVX+fs 0.29
 EOF
 
 

### 2013-11-29 doc/papers/wpmvp14/bench.sh
Update bench results

@@ -48,10 +48,10 @@ Pythran+AVX 0.30
 EOF
 
 cat > rosen_fs.dat << EOF
-CPython 1.64
-Pythran 0.49
-Pythran+AVX 0.41
-Pythran+AVX+fs 0.29
+CPython 14.6
+Pythran 4.47
+Pythran+AVX 3.07
+Pythran+AVX+fs 2.88
 EOF
 
 

### 2013-11-29 doc/papers/wpmvp14/paper.tex
Reading

@@
 c} is computed using two loops and an intermediate array. It means there are 4
-load / stores and only two additions per loop iteration and vectorization is
+\texttt{load}/\texttt{store} and only two additions per loop iteration and vectorization is
 unlikely to be profitable.
@@
 Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
-for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
+for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is matched by
+the C++ alternative of contiguous container but it is not
 matched by \emph{generator}s, the Python incarnation of co-routines.
-
-%we know a work have to be done about random acces but you speak about something else now ...
-%It is disturbing and we don't really know why you speak about this new topic.
 
@@
 apply \emph{forward substitution} to fall-back to the expression from
-Listing~\ref{lst:numpy-code} and avoid the creation of any temporary in some
-case detected with static analysis of the source code.
+Listing~\ref{lst:numpy-code} and avoid the creation of any temporary based on
+static analysis of the source code.
 

### 2014-01-19 doc/papers/wpmvp14/paper.tex
Take into account reviewers comment
Most notably, emphasise the fact we are using doubles and not floats
during the experiments...

@@
 \providecommand{\eg}[1][~]{\textit{e.g.#1}}
+
+%\doi{}
 
@@
     The Python language is highly dynamic, most notably due to late binding. As
-    a consequence, programs run using Python typically run an order of magnitude
+    a consequence, programs using Python typically run an order of magnitude
     slower than their C counterpart. It is also a high level language whose
@@
 
-A native module is a shared library whose interface make it callable from the
+A native module is a shared library whose interface makes it callable from the
 Python interpreter using the regular \texttt{import} mechanism. It makes it
 possible to call native code from Python in a transparent way. Typically Python
-object are unboxed when transfered from Python to C, where efficient
+objects are unboxed when transfered from Python to C, where efficient
 computations can happen. Then the result is boxed when getting back to the
@@
 
+All the benchmarks use Numpy arrays of \emph{double precision} random floats as
+input, which means a maximum theoritical speedup of $\times4$ using AVX.
+
 The scripts used to perform the measures as well as all other experimental data

### 2014-01-21 doc/papers/wpmvp14/paper.tex
Last update of the WPMVP paper
Real final version, as submitted to the ACM, version
Some code had to be reformatted to fit in the new page layout, some were
pruned due to an overgrowth of the paper.

@@
 % submit to https://sites.google.com/site/wpmvp2014/home
-\documentclass[preprint]{sigplanconf}
+\documentclass[10pt]{sigplanconf}
 
@@
 \usepackage{hyperref}
+
+\special{papersize=8.5in,11in}
+\setlength{\pdfpageheight}{\paperheight}
+\setlength{\pdfpagewidth}{\paperwidth}
+
+\conferenceinfo{WPMVP '14}{February 16, 2014, Orlando, FL, USA} 
+\copyrightyear{2014} 
+\copyrightdata{978-1-4503-2653-7/14/02}
+\doi{2568058.2568060}
 
@@
 
-%\doi{}
-
 \begin{document}
-
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-
-%\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-%\copyrightyear{20yy} 
-%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-%\doi{nnnnnnn.nnnnnnn}
 
@@
 whenever dereferenced. These iterators are then usable directly in usual STL
-algorithms such as \texttt{transform} or \texttt{fold} (Listing \ref{stl_example}).
+algorithms such as \texttt{transform} or \texttt{fold} (Listing \ref{lst:tl_example}).
 
-\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={SIMD Iterator with STL algorithm},label=stl_example]
-std::vector<int,allocator<int> > v(N),r(N);
-
-std::transform ( simd::begin(v.begin())
-               , simd::end(v.end())
-               , simd::begin(r.begin())
-               , [](pack<int> const& p)
-                 { 
-                   return p*p; 
-                 } 
+\begin{lstlisting}[language=c++, label={lst:stl_example}, caption={SIMD Iterator with STL algorithm},label=stl_example]
+std::vector<int,allocator<int>> v(N),r(N);
+std::transform (
+ simd::begin(v.begin()),simd::end(v.end()),
+ simd::begin(r.begin()),
+ [](pack<int> const& p) { return p*p; } 
 );
@@
 \begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={Polymorphic SIMD algorithm call},label=stl_poly]
-std::vector<int,allocator<int> > v(128), r(128);
-
-struct f
-{
+std::vector<int,allocator<int>> v(9), r(9);
+struct f {
  template<class T>
  auto operator()(T const& p)
-  -> decltype(p * p)
-  {
-    return p * p;
-  }
+ -> decltype(p*p)
+ { return p * p; }
 };
 
-simd::transform( v.begin(), v.end()
-               , r.begin()
-               , f() 
-               );
+simd::transform( v.begin(), v.end(),
+                 r.begin(), f());
 \end{lstlisting}
@@
  template<class T>
-    auto operator()(T&& x)
-    -> decltype(3 * x)
-    {
-        return x * 3;
-    }
+ auto operator()(T&& x) -> decltype(3*x)
+ { return x * 3; }
 };
@@
 \begin{lstlisting}[language=python, label={lst:square-python}, caption={Python implementation of the square function.}]
-def square(x):
-  return x * x
-\end{lstlisting}
-
-\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={C++ meta-implementation of the square function.}]
-struct square {
-    template<class T>
-    auto operator()(T const& x)
-    -> decltype(x * x)
-    {
-      return x * x;
-    }
-};
+square = lambda x: x * x
 \end{lstlisting}
@@
 def rosen(x):
-  return np.sum(100*(x[1:] - x[:-1] ** 2) ** 2
-                + (1 - x[:-1]) ** 2)
+  return np.sum(100*(x[1:]-x[:-1]**2)**2
+                + (1 - x[:-1])**2)
 \end{lstlisting}
@@
 branch of the official Pythran repository
-\url{https://github.com/serge-sans-paille/pythran}. The back-end C++ compiler is
-Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
+\url{https://github.com/serge-sans-paille/pythran}. The back-end C++ compiler
+is Clang version 3.4-1. The compilation flags used are \texttt{-O2} without
 vectorisation support and \texttt{-O2 -march=native -DUSE\_BOOST\_SIMD} with
 vectorisation support. Using \texttt{-O3} or \texttt{-Ofast} actually degraded
-performances.
+performances. Using the compiler automatic vectorizer did not end up with any
+speedup, most probably due to expression templates confusing the vectorizer.
 
@@
 \texttt{ndarray} of double precision floating point numbers.
-Figure~\ref{fig:numpy-timings} summarizes the results. Because all computations
-are done in native space, there is not much speedup when running them through
-Pythran. The only benefits comes from the elimination of temporary arrays and
-loop fusion. As a consequence, the more elements in the \texttt{numpy}
-expression, the more Pythran accelerates the code. The vectorization speedups
-follow the same process, as larger expression tend to increase the operation
-over memory movement ratio.
+Figures~\ref{fig:numpy-timings} and~\ref{fig:numpy-code-fs} summarize the
+results. Because all computations are done in native space, there is not much
+speedup when running them through Pythran. The only benefits comes from the
+elimination of temporary arrays and loop fusion. As a consequence, the more
+elements in the \texttt{numpy} expression, the more Pythran accelerates the
+code. The vectorization speedups follow the same process, as larger expression
+tend to increase the operation over memory movement ratio.
 
@@
     \includegraphics[width=.5\textwidth]{numexpr}
-    \includegraphics[width=.5\textwidth]{rosen}
+    %\includegraphics[width=.5\textwidth]{rosen}
     \includegraphics[width=.5\textwidth]{arc}

### 2014-01-27 doc/papers/CGO2014/paper.tex
Add CGO 2014 draft

\documentclass[10pt, onecolumn, preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{listings}
\usepackage{hyperref}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\lstset{captionpos=b, float, language=python}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}

\providecommand{\boostsimd}{\textsc{Boost.SIMD}}
\providecommand{\cpp}[1][~]{\textsc{C++}#1}
\providecommand{\ie}[1][~]{\textit{i.e.}#1}
\providecommand{\eg}[1][~]{\textit{e.g.#1}}

\begin{document}

\title{Pythran: Benchmarking parallel Scientific Python Programs}

\authorinfo{Serge Guelton}
           {T{\'e}l{\'e}com Bretagne}
           {serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}
           {INRIA/MOAIS}
           {pierrick.brunet@inria.fr}
\authorinfo{Mehdi Amini}
           {SILKAN}
           {mehdi.amini@silkan.com}

\maketitle

\begin{abstract}

    Turning Python module in native one is a common practice in the Python
    eco-system to improve performance of computation intensive kernel for
    scientific modules. Many Python compiler are born in the Python community
    such as the well known Cython trying bring in Python the holy C
    performances. To reach this goal, optimisations should be done
    using as much as possible computer capabilities instead of doing a
    straightforward conversion. Such optimisations should take into account
    multi-core architecture and vectorization available in common
    platforms as numpy - the reference scientific python module - do an
    intensive use of such functions. Thanks to these optimisation,
    Python module should be able to achieve higher performances.

    Pythran is an open source static compiler that turns modules written
    in a subset of Python into native ones~\cite{pythran}. It already apply
    some optimisations like constant-folding for functions or partial forward
    substitution. This paper introduce integration of vectorization and
    parallelism on numpy array based kernel. Pythran already provides
    possibility to add OpenMP annotation in Python program for explicit
    parallelism~\cite{pythran-compas} and now, it aims to add it through its
    C++ runtime library using the Numeric Template Toolbox (NT2) for vectorization
    and OpenMP for parallelism.

    This approach have been validated though benchmark between Pythran with
    vectorization and parallelism and Pythran without it. It also compares these
    measures with others Python compiler like Numba and Coperhead.

\end{abstract}

\keywords
static compilation, parallelization, Python, C++, SIMD

%%
%%
\section{Introduction}

L'optimisation de performances pour les programmes Python n'est pas récent
puisque les premiers compilateur de pseudo language proche de Python comme Pyrex
ont plus de 10 ans maintenant. Certains projets sont aujourd'hui arrété
alors que d'autre naissent pour apporter toujours plus de performances.

%Cython
%De nombreux compilateur s'intéresse à la vectorization. automatique. Dans la
%communauté Python, Cython prévois des améliorations à ce sujet bien que rien ne
%soit encore implémenté
%\footnote{\url{https://github.com/cython/cython/wiki/enhancements-simd}}.
%Pour le multi-core, Cython ajoute une librairy mais ne fait rien
%automatiquement (donc c'est des annotations cf papier renpar)
%
%Nuitka
%je ne sais pas ce qu'ils font. Rien sur le web, il faudrait regarder le code
%généré pour savoir... (Dans le code, aucune trace de parallelism SIMD/OpenMP)
%
%Numba
%ne peut faire du parallelism que si il ne repose pas du tout sur Python
%(qu'il compile tout le code) : \url{http://www.phi-node.com/2013/01/just-in-time-compilers-for-number.html}
%Ils ont un prange comme Cython pour avoir du parallélisme explicite et ils ont
%une library de threading a eux...
%Pour le SIMD, ils font de la fusion d'operateur SI l'utilisateur le demande et
%specifie correctement les dimensions/profondeur ou on applique la fusion:
%\url{http://numba.pydata.org/numba-doc/0.13.4/ufuncs.html}
%Ils font du déroulage partiel de boucle et du vectoriel ensuite mais on a pas
%beaucoup d'info je trouve... (J'ai trouvé ça dans le code avec du code :
%oldnumba/minivect/specializers.py:438: rhs = b.vector\_load(data\_pointer, self.sp.vector\_size)
%Ils prévoient aussi un support openmp apparement : "OpenMP is not yet implemented, only process the 'else' directives. llvm\_codegen.py l.168"
%Ils ont un bon support cuda par contre : \url{https://developer.nvidia.com/how-to-cuda-python}
%
%Numexpr
%fait de la composition d'opérateur mais pas d'utilisation de vectorization
%Ils ont un support de parallelisation multicore automatique a base de pthread.
%
%Copperhead
%font du parallelism dans tous les sens :
%\url{http://www.eecs.berkeley.edu/Pubs/TechRpts/2010/EECS-2010-124.pdf}
%mais c'est une syntax particuliére. => ils disent ne aps trop se préoccupé des perfs
%mais plus de la scalabilité
%The runtime operates on CuArrays, which extend Numpy.ndarray
%
%Theano
%C'est un truc fait pour le parallélisme MAIS
%197 * SIMD parallelism on the CPU comes from the compiler.                          
%198 * Multi-core parallelism is only supported by Conv2d(not by default).           
%199   If the external BLAS implementation supports it,                              
%200   there are also, gemm, gemv and ger that are parallelized
%tiré de leurs docs -> D'après le code ils fusionne les opérateur pour que le
%backend puisse essayer de faire du SIMD avec ça.
%Pour le parallel : 
%Multi-core parallelism is only supported by Conv2d(not by default)
%
%Shedskin
%Apparement, pas de SIMD, mais du parallelism explicite a base de Parallel
%Python ou pprocess (fork)
%
%parakeet
%fusion d'opérateur, OpenMP mais pas de SIMD
%\url{http://cs.nyu.edu/web/Research/Theses/eric_h.pdf}
%Attention, ils ont des contraites fortes : si on utilise map, reduce, ..., on
%concidére que c'est paralléle sans vérification. (2.4 de la thèse)
%(D'ailleurs dans Pythran, on vérifie la pureté de la fonction mais on ne
%vérifie pas la sortie :
%ex a = map(foo, xrange(12)); if a == [foo(0), foo(1), ...]: ....
%ne fonctionnerait pas avec l'analyse pmap)
%
%
%Numpypy
%Ils ne font pas ce genre d'optim
%
%
%ATTENTION : le papier risque de faire quand même très répétitif par rapport a WPMVP'14

%TODO ICI UN PARAGRAPH SUR COMMENT LES AUTRES FONT DE la vectorization ET DU PARALLELISME

%TODO un paragraph sur le parallélisme dans numpy

In this papers, we will speak about the Pythran compiler that turn pure Python
code to C++ to create native module and gather high level information to
profiter pleinement de la parallélisation des opérations.

This paper highlight two major contributions. We first introduce how vectoriation
have been added numpy computation intensive kernel thank to
\boostsimd library provided by NT2. We present performances results for Pythran
with and without this auto-vectorization to show performances gains.
Secondly, we introduce parallelization on multi-core architecture for these
computation intensive kernel and evaluates it.

The remainder of this paper is organized as follows. Section \ref{sec:numpy}
describes the differents numpy constructs that are highly parallel and how to
"profiter" from this parallelism. Section \ref{sec:optim} describes how Pythran
optimisation enhance this parallelism increasing computation kernel. Section 
\ref{sec:SIMD} introduces how \boostsimd is used to add auto-vectorization in
Pythran. Section \ref{sec:omp} show use of OpenMP in these kernel for
mutli-core architecture. We will discuss performance result for those
optimization using two different hardware configurations in Section
\ref{sec:perfs}. We finally present some related works in section
\ref{sec:related-work} before concluding.

\section{Data parallelism in Numpy}
\label{sec:numpy}

\section{Optimization feed each other}
\label{sec:optim}

\section{Vectorization using \boostsimd}
\label{sec:SIMD}

\section{Multi core parallelism using OpenMP}
\label{sec:omp}

\section{Benchmarks}
\label{sec:perfs}

\subsection{Pythran with/without Vectorization}

\subsection{Pythran with/without OpenMP}

\subsection{Combining OpenMP and Vectorization}

% Une autre section pour la comparaison avec les autres compilo?

\section{Related works}
\label{sec:related-work}

% On aura pas le temps pour le papier je pense mais faire une comparaison avec du
% C pure pour voir ce que les compilo de C arrive a faire.

\section*{Conclusion}

\acks

This project has been partially funded by the CARP Project and the SILKAN
Company.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{biblio}

\end{document}

### 2014-01-27 doc/papers/iop2014/paper.tex
Add IOP 2014 draft
From the scipy proceedings, updated until the benchmarks.

% submit to https://sites.google.com/site/wpmvp2014/home
\documentclass[10pt, onecolumn, preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{listings}
\usepackage{hyperref}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

%\conferenceinfo{WPMVP '14}{February 16, 2014, Orlando, FL, USA} 
%\copyrightyear{2014} 
%\copyrightdata{978-1-4503-2653-7/14/02}
%\doi{2568058.2568060}

\lstset{captionpos=b, float, language=python}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}

\providecommand{\boostsimd}{\textsc{Boost.SIMD}}
\providecommand{\cpp}[1][~]{\textsc{C++}#1}
\providecommand{\ie}[1][~]{\textit{i.e.}#1}
\providecommand{\eg}[1][~]{\textit{e.g.#1}}

\begin{document}

\title{Pythran: Enabling Static Optimization of Scientific Python Programs}

\authorinfo{Serge Guelton}
           {T{\'e}l{\'e}com Bretagne}
           {serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}
           {INRIA/MOAIS}
           {pierrick.brunet@inria.fr}
\authorinfo{Alan Raynaud}
           {T{\'e}l{\'e}com Bretagne}
           {alan.raynaud@telecom-bretagne.eu}
\authorinfo{Mehdi Amini}
           {SILKAN}
           {mehdi.amini@silkan.com}
\authorinfo{Adrien Merlini}
           {T{\'e}l{\'e}com Bretagne}
           {adrien.merlini@telecom-bretagne.eu}
\authorinfo{Eliott Coyac}
           {T{\'e}l{\'e}com Bretagne}
           {eliott.coyac@telecom-bretagne.eu}

\maketitle

\begin{abstract}

    Pythran is an open source static compiler that turns modules written
    in a subset of Python into native ones. Based on the fact that scientific
    modules do not rely much on the dynamic features of the language, it trades
    them in favor of powerful, eventually inter procedural, optimizations.
    These include detection of pure functions, temporary allocation removal,
    constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
    thread-level parallelism through OpenMP annotations, false variable
    polymorphism pruning, and automatic vector instruction generation such as
    AVX or SSE.

    In addition to these compilation steps, Pythran provides a C++ runtime library that
    leverages the C++ STL to provide generic containers, and the Numeric
    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
    features such as variadic templates, type inference, move semantics and
    perfect forwarding, as well as classical ones such as expression templates.

    The input code remains compatible with the Python interpreter, and output
    code is generally as efficient as the annotated Cython equivalent, if not
    more, without the backward compatibility loss of Cython. Numpy expressions
    run faster than when compiled with numexpr, or numba, without any change to
    the original code.

\end{abstract}

\keywords
static compilation, parallelization, Python, C++

%%
%%
\section{Introduction}

The Python language is growing in popularity as a language for scientific
computing, mainly thanks to a concise syntax, a high level standard library and
several scientific packages.

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high, due to numerous dynamic lookup and interpretation cost inherent
in high level languages. Additionally, the Python compiler performs no
optimization on the byte code, while scientific applications are first-class
candidates for many of them.

Following the saying that scientific applications spend 90\% of their time in
10\% of the code, it is natural to focus on computation-intensive piece of code.
The aim may not be to optimize the full Python application, but rather a
small subset of the application.

Several tools have been proposed by an active community to fill the performance
gap met when running these computation-intensive piece of code, either through
static compilation or Just In Time (JIT) compilation.

An approach used by Cython\cite{cython2010} is to suppress the interpretation
overhead by translating Python Programs to C programs calling the Python C
API\cite{pythoncapi}. More recently, Nuitka\cite{nuitka}  has taken the same
approach using C++ has a back-end. Going a step further Cython also provides an
hybrid C/Python language that can efficiently be translated to C code, relying
on the Python C API for some parts and on plain C for others.
ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python program
into C++, without any call to the Python C API, but fails to compile dynamic
codes or codes for which static type inference is impossible.

The alternate approach consists in writing a JIT compiler, embedded into the
interpreter, to dynamically turn the computation intensive parts into native
code. The \texttt{numexpr} module\cite{numexpr} does so for Numpy expressions
by JIT-compiling them from a string representation to native code.
Numba\cite{numba} and Parakeet\cite{parakeet2012} extend this approach to
Numpy-centric applications while PyPy\cite{pypy2009} applies it to the whole
language, including the Numpy module through the in-progress \texttt{numpypy}
branch.

To the notable exception of PyPy, these compilers do not apply any of the
static optimization techniques that have been known for decades and
successfully applied to statically compiled language such as C or C++.
Translators to statically compiled languages do take advantage of them
indirectly, but the quality of generated code may prevent advanced
optimizations, such as vectorization, while they are available at higher level,
i.e. at the Python level. Taking into account the specificities of the Python
language can unlock many new transformations. For instance, PyPy automates the
conversion of the \texttt{range} builtin into \texttt{xrange} through the use
of a dedicated structure called \texttt{range-list}.

This article presents Pythran, an optimizing compiler for a subset of the
Python language that turns implicitly statically typed modules into parametric
C++ code. It supports many high-level constructs of the 2.7 version of the
Python language such as list comprehension, set comprehension, dict
comprehension, generator expression, lambda functions, nested functions,
polymorphic functions and global variables. It does \textbf{not} support user
classes or any dynamic feature such as introspection or polymorphic variables.

Unlike existing alternatives, Pythran does not solely perform static typing of
Python programs. It also performs various compiler optimizations such as
detection of pure functions, temporary allocation removal or constant folding.
These transformations are backed up by code analysis such as aliasing,
inter-procedural memory effect computations or use-def chains.

The article is structured as follows: Section~\ref{sec:infrastructure}
introduces the Pythran compiler compilation flow and internal representation.
Section~\ref{sec:analysis} presents several code analysis while
Section~\ref{sec:optimizations} focuses on code optimizations.
Section~\ref{sec:backend} presents back-end optimizations for Numpy
expressions. Section~\ref{sec:openmp}  briefly introduces OpenMP-like
annotations for explicit parallelization of Python programs and
section~\ref{sec:benchmarks} presents the performance obtained on a few
synthetic benchmarks and concludes.

\section{Pythran Compiler Infrastructure}
\label{sec:infrastructure}

Pythran is a compiler for a subset of the Python language. In this paper, the
name \textbf{Pythran} will be used indifferently to refer to the language or
the associated compiler. The input of the Pythran compiler is a Python module
---not a Python program--- meant to be turned into a native module. Typically,
computation-intensive parts of the program are moved to a module fed to
Pythran. The module is translated to polymorphic C++ code through the
generation of templated classes in place of Python functions, then these
classes are either used from regular C++ code or instantiated for given types
to generate a native Python module.

It is a strong requirement for Pythran to maintain backward compatibility with
Python, that is unlike Cython, all code compilable by Pythran are still
executable by a standard-conforming Python interpreter. For that reason, in
addition to language restrictions detailed in the following, Pythran
understands special comments such as:

\begin{lstlisting}
#pythran export foo(int list, float)
\end{lstlisting}

as optional module signature. One does not need to list all the module
functions in an \texttt{export} directive, only the functions meant to be used
outside of the module. Polymorphic functions can be listed several times with
different types. These type annotations are only used for native code generation.

The Pythran compiler is built as a traditional static compiler: a front-end
turns Python code into an Internal Representation (IR), a middle-end performs
various code optimizations on this IR, and a back-end turns the IR into C++
code. The front-end performs two steps:

\begin{enumerate}

    \item turn Python code into Python Abstract Syntax Tree (AST) thanks to the \texttt{ast}
   module from the standard library;

    \item turn the Python AST into a type-agnostic Pythran IR, which remains a subset
   of the Python AST.

\end{enumerate}

Pythran IR is similar to Python AST, as defined in the \texttt{ast} module, except
that several nodes are forbidden (most notably Pythran does not support
user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
to others to form a simpler AST easier to deal with for further analyses and
optimizations. The transformations applied by Pythran on Python AST are the
following:

\begin{itemize}
    \item list/set/dict comprehension are expanded into loops wrapped into a function call;

    \item destructuring is expanded into several variable assignments;

    \item lambda functions are turned into named nested functions;

    \item the closure of nested functions is statically computed to turn nested
        functions into global functions taking the closure elements as
        parameters, and the initial definition is replaced by a \emph{partial
        function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;

    \item implicit \texttt{return None} are made explicit;

    \item all imports are fully expanded to make function and global variables access paths explicit,

    \item method calls are turned into function calls;

    \item implicit \texttt{\_\_builtin\_\_} function calls are made explicit;

    \item \texttt{try \dots finally} constructs are turned into nested \texttt{try \dots except} blocks;

    \item identifiers whose name may clash with C++ keywords are renamed.

\end{itemize}

The back-end works in three steps:

\begin{enumerate}

    \item turning Pythran IR into parametric C++ code;

    \item instantiating the C++ code for the desired types;

    \item compiling the generated C++ code into native code.

\end{enumerate}

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its life
time, but it cannot be assigned to a string if it has already been assigned to
an integer. As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected, which is very
similar to Python's duck typing, except that it is checked at compile time, as
illustrated by the following implementation of a generic dot product in Python:

\begin{lstlisting}
def dot(l0, l1):
    return sum(x*y for x,y in zip(l0,l1))
\end{lstlisting}

\noindent and in C++:

\begin{lstlisting}[language=c++]
template<class T0, class T1>
    auto dot(T0&& l0, T1&& l1) -> decltype(/* skipped */)
    {
        return pythonic::sum(
            pythonic::map(
                operator_::multiply(),
                    pythonic::zip(
                        std::forward<T0>(l0),
                        std::forward<T1>(l1))
            )
        );
    }
\end{lstlisting}

Although far more verbose than the Python version, the C++ version also uses a
form of structural typing: the only assumption these two versions make are that
\texttt{l0} and \texttt{l1} are iterable, their content can be multiplied and the result of
the multiplication is accumulatable.

Finally, Pythran computes the set of functions and types used for the processed
module, and generate the minimal set of headers that need to be included to
provide the definitions needed to compile the generated C++ code. This step is
critical to prevent typical slowdown from C++ compilation.

The second step only consists in the instantiation of the top-level functions of the
module, using user-provided signatures. Template instantiation then triggers the
different correctly typed instantiations for all functions written in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites.

The last step involves a template library, called \texttt{pythonic} that contains a
polymorphic implementation of many functions from the Python standard library
in the form of C++ template functions. Several optimizations, most notably
expression template, are delegated to this library. Pythran relies on the
C++11\cite{isocxx11} language, as it makes heavy use of recent features such as
move semantics, type inference through \texttt{decltype(\dots)} and variadic templates.
As a consequence it requires a compatible C++ compiler for the native code
generation. Boost.Python\cite{boostpython2007} is involved for the automatic
Python-to-C++ boilerplate.  Generated C++ code is compatible with g++ 4.7.2 and
clang++ 3.2.

It is important to note that all Pythran analyses and optimizations are
type-agnostic, i.e. they do not assume any type for the variables manipulated
by the program. Type specialization is only done in the back-end, right before
native code generation. Said otherwise, the Pythran compiler analyzes
polymorphic functions and turns Python modules into C++ meta-programs.

Figure\ref{fig:pythran-compiler} summarizes the compilation flow and the involved
tools.

\begin{figure}

    \centering
    \begin{tikzpicture}[
            file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
            node distance=0.7cm},
            tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
        distance=0.7cm}]
        \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
        \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
        \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
        \node (empty) [xshift=-1em, left=of gxx] {};
        \node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic}}};
        \node[file] (annotation)     [above=of pythonic] {\textbf{Type Info}};
        \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
        \node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

        \draw[very thick, ->] (python) -- (pythran);
        \draw[very thick] (annotation) -| (empty.center);
        \draw[very thick, ->] (pythran) -- (meta-cxx);
        \draw[very thick, ->] (meta-cxx) -- (gxx);
        \draw[very thick] (boost) -| (empty.center);
        \draw[very thick, ->] (pythonic) -- (gxx);
        \draw[very thick, ->] (gxx) -- (so);
    \end{tikzpicture}

    \caption{Pythran compilation flow.}
    \label{fig:pythran-compiler}

\end{figure}

\section{Code Analyses}
\label{sec:analysis}

A code analysis is a function that takes a part of the IR (or the whole
module's IR) as input and returns aggregated high-level information. For
instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
of all identifiers used throughout the program. This information is later used
when the creation of new identifiers is required so that no conflict occurs
with existing ones.

One of the most important analysis in Pythran is the \textbf{alias analysis}, sometimes
referred as \textbf{points-to} analysis. For each identifier, it computes an
approximation of the set of locations this identifier may points to. For
instance, let us consider the polymorphic function \texttt{foo} defined as follows:

\begin{lstlisting}
def foo(a,b):
    c = a or b
    return c * 2
\end{lstlisting}

The identifier \texttt{c} involved in the multiplication may refer to

\begin{itemize}
    \item a fresh location if \texttt{a} and \texttt{b} are scalars

    \item the same location as \texttt{a} if \texttt{a} evaluates to \texttt{True}

    \item the same location as \texttt{b} otherwise.

\end{itemize}

As we do not specialise the analysis for different types and the truth value of
\texttt{a} is unknown at compilation time, the alias analysis yields the approximated
result that \texttt{c} may point to a fresh location, \texttt{a} or \texttt{b}.

Without this kind of information, even a simple instruction like
\texttt{sum(a)} would yield very few informations as there is no guarantee that
the \texttt{sum} identifiers points to the \texttt{sum} built-in, as someone
could have previously stated that \texttt{sum = len}.

When turning Python AST to Pythran IR, nested functions are turned into global
functions taking their expanded closure as parameter. This closure is computed using the
information provided by the \texttt{Globals} analysis that statically computes the
state of the dictionary of globals, and \texttt{ImportedIds} that computes the set of
identifiers used by an instruction but not declared in this instruction. For
instance in the following snippet:

\begin{lstlisting}
def outer(outer_argument):
    def inner(inner_argument):
        return cos(outer_argument) + inner_argument
        return inner(3)
\end{lstlisting}

The \texttt{Globals} analysis called on the \texttt{inner} function definition
marks \texttt{cos} as a global variable, and \texttt{ImportedIds} marks
\texttt{outer\_argument} and \texttt{cos} as imported identifiers. The result
of the transformation of nested function into global functions on that
particular example is:

\begin{lstlisting}
import functools

def _inner(outer_argument, inner_argument):
    return cos(outer_argument) + inner_argument

def outer(outer_argument):
    inner = functools.partial(_inner, outer_argument)
    return inner(3)
\end{lstlisting}

A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
set of functions declared in the module that are pure, i.e. whose return value
only depends from the value of their argument. This analysis depends on two
other analyses, namely \texttt{GlobalEffects} that computes for each function whether
this function modifies the global state (including I/O, random generators, etc.)
and \texttt{ArgumentEffects} that computes for each argument of each function whether
this argument may be updated in the function body. These three analyses work
inter-procedurally, as illustrated by the following example:

\begin{lstlisting}
def fibo(n):
    return n if n < 2 else fibo(n - 1) + fibo(n - 2)

def bar(l):
    return map(fibo, l)

def foo(l):
    return map(fibo, random.sample(l, 3))
\end{lstlisting}

The \texttt{fibo} function is pure as it has no global effects or argument effects and
only calls itself. As a consequence the \texttt{bar} function is also pure as the
\texttt{map} intrinsic is pure when its first argument is pure. However the \texttt{foo}
function is not pure as it calls the \texttt{sample} function from the \texttt{random}
module, which has a global effect (on the underlying random number generator
internal state).

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because the \texttt{print b} involves an \textbf{I/O}.

\begin{lstlisting}
def pure(a):
    return a ** 2

def guilty(a):
    b = pure(a)
    print b
    return b

l = list(...)
map(pure, l)
map(guilty, l)
\end{lstlisting}

\texttt{ConstantExpressions} uses function purity to decide whether a given expression
is constant, i.e. its value only depends on literals. For instance the
expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
argument is a literal.

\texttt{UseDefChains} is a classical analysis from the static compilation world. For
each variable defined in a function, it computes the chain of \textbf{use} and \textbf{def}.
The result can be used to drive various code transformations, for instance to
remove dead code, as a \textbf{def} followed by a \textbf{def} or nothing is useless. It is
used in Pythran to avoid false polymorphism. An intuitive way to represent
use-def chains is illustrated on next code snippet:

\begin{lstlisting}
a = 1
if cond:
    a = a + 2
else:
    a = 3
print a
a = 4
\end{lstlisting}

In this example, there are two possible chains starting from the first
assignment. Using \texttt{U} to denote \textbf{use} and \texttt{D} to denote \textbf{def}, one gets:

\begin{center}
\begin{tikzpicture}[D/.style={black!50,fill=black!10, circle, drop shadow, align=center},
    U/.style={black!50,fill=black!10, rectangle, drop shadow, align=center}]
    \node[D] (d0) {D};
    \node[U] (u0) [right=of d0] {U};
    \node[D] (d1) [right=of u0] {D};
    \node[U] (u1) [right=of d1] {U};
    \node[D] (d2) [right=of u1] {D};

    \node[D] (dp0)[below=of u0, yshift=2em] {D};

    \draw[very thick, ->] (d0) -- (u0);
    \draw[very thick, ->] (d0) |- (dp0);
    \draw[very thick, ->] (u0) -- (d1);
    \draw[very thick, ->] (d1) -- (u1);
    \draw[very thick, ->] (dp0) -| (u1);
    \draw[very thick, ->] (u1) -- (d2);
\end{tikzpicture}
\end{center}

The fact that all paths finish by a \textbf{def} indicates that the last
assignment can be removed (but not necessarily its right hand part that could
have a side-effect). The fact that the last \textbf{use} is always preceded by
a \textbf{def}taht only point to it indicates that these \textbf{def}s can
safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
renaming).

All the above analyses are used by the Pythran developer to build code
transformations that improve the execution time of the generated code.

\section{Code Optimizations}
\label{sec:optimizations}

One of the benefits of translating Python code to C++ code is that it removes
most of the dynamic lookups. It also unveils all the optimizations available at
C++ level. For instance, a function call is quite costly in Python, which
advocates in favor of using inlining. This transformation comes at no cost when
using C++ as the back-end language, as the C++ compiler does it.

However, there are some informations available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functor with an
internal state and a goto dispatch table to represent generators. Although
effective, this approach is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]+i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i]+i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

A similar optimization consists in turning \texttt{map}, \texttt{zip} or
\texttt{filter} into their equivalent version from the \texttt{itertool}
module. The benefit is twofold: first it removes a temporary allocation, second
it gives an opportunity to the compiler to replace list accesses by scalar
accesses. This transformation is not always valid, nor profitable. It is not
valid if the content of the output list is written later on, and not profitable
if the content of the output list is read several times, as each read implies
the (re) computation, as illustrated in the following code:

\begin{lstlisting}
def valid_conversion(n):
    # this map can be converted to imap
    l = map(math.cos, range(n))
    return sum(l) # sum iterates once on its input

def invalid_conversion(n):
    # this map cannot be converted to imap
    l = map(math.cos, range(n))
    l[0] = 1  # invalid assignment
    return sum(l) + max(l) # sum iterates once
\end{lstlisting}

The information concerning constant expressions is used to perform a classical
transformation called \texttt{ConstantFolding}, which consists in the compile-time
evaluation of constant expressions. The validity is guaranteed by the
\texttt{ConstantExpressions} analysis, and the evaluation relies on Python ability to
compile an AST into byte code and run it, benefiting from the fact that Pythran
IR is a subset of Python AST. A typical illustration is the initialization of a
cache at compile-time:

\begin{lstlisting}
def esieve(n):
    candidates = range(2, n+1)
    return sorted(
        set(candidates) - set(p*i
                              for p in candidates
                              for i in range(p, n+1))
        )

cache = esieve(100)
\end{lstlisting}

Pythran automatically detects that \texttt{eseive} is a pure function and evaluates
the \texttt{cache} variable value at compile time.

\texttt{ConstantFolding} may create a lot of candidates for \texttt{LoopUnrolling}. In Python, loops are used to iterate over sequences. If such a sequence is an explicit list, then Pythran unrolls the loop accordingly. Thanks to \texttt{ConstantFolding}, a call to \texttt{range(4)} will statically turn into \texttt{[0, 1, 2, 3]} (after a check that \texttt{range} actually points to \texttt{\_\_builtin\_\_.range}) making it possible to unroll the loop statically.

Sometimes, coders use the same variable in a function to represent value with
different types, which leads to false polymorphism, as in:

\begin{lstlisting}
a = cos(1)
a = str(a)
\end{lstlisting}

These instructions cannot be translated to C++ directly because \texttt{a}
would have both \texttt{double} and \texttt{str} type. However, using
\texttt{UseDefChains} it is possible to assert the validity of the renaming of
the instructions into:

\begin{lstlisting}
a = cos(1)
a_ = str(a)
\end{lstlisting}

that does not have the same typing issue.

In addition to these python-level optimizations, the Pythran back end library,
\texttt{pythonic}, uses several well known optimizations, especially for Numpy
expressions.

\section{Library Level Optimizations}
\label{sec:backend}

Using the proper library, the C++ language provides an abstraction level close
to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
that leverage on the C++ Standard Template Library (STL), the GNU Multiple
Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
standard library. The STL is used to provide a typed version of the standard
containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
well as reference-based memory management through \texttt{shared\_ptr}. Generic
algorithms such as \texttt{accumulate} are used when possible. GMP is the
natural pick to represent Python's \texttt{long} in C++. NT2 provides a generic
vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
the vector instruction units of modern processors in a generic way. It is used
to efficiently compile Numpy expressions.

Numpy expressions are the perfect candidates for library level optimizations.
Pythran implements three optimizations on such expressions:

\begin{enumerate}

    \item Expression templates\cite{expression_templates} are used to avoid
        multiple iterations and the creation of intermediate arrays. Because
        they aggregates all \texttt{ufunc} into a single expression at compile
        time, they also increase the computation intensity of the loop body,
        which increases the impact of the two following optimizations.

    \item Loop vectorization. All modern processors have vector instruction
        units capable of applying the same operation on a vector of data
        instead of a single data. For instance Intel Sandy Bridge can run 8
        single-precision additions per instruction. One can directly use the
        vector instruction set assembly to use these vector units, or use C/C++
        intrinsics. Pythran relies on \texttt{boost.simd} from NT2 that offers
        a generic vector implementation of all standard math functions to
        generate a vectorized version of Numpy expressions. Again, the
        aggregation of operators performed by the expression templates proves
        to be beneficial, as it reduces the number of (costly) loads from the
        main memory to the vector unit.

    \item Loop parallelization through OpenMP\cite{openmp3.1}. Numpy expression
        computation do not carry any loop-dependency. They are perfect
        candidates for loop parallelization, especially after the expression
        templates aggregation, as OpenMP generally performs better on loops
        with higher computation intensity that masks the scheduling overhead.

\end{enumerate}

To illustrate the benefits of these three optimizations combined, let us
consider the simple Numpy expression:

\begin{lstlisting}
d = numpy.sqrt(b*b+c*c)
\end{lstlisting}

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core i7, the
pure Python execution yields:

\begin{lstlisting}
>>> %timeit np.sqrt(b*b+c*c)
1000 loops, best of 3: 1.23 ms per loop
\end{lstlisting}

\noindent then after Pythran processing and using expression templates:

\begin{lstlisting}
>>> %timeit my.pythranized(b,c)
1000 loops, best of 3: 621 us per loop
\end{lstlisting}

The speed-up is due to the fact that expression templates replace 4 temporary
array creations and 4 loops by a single allocation and a single loop.

Going a step further and vectorizing the generated loop yields an extra
performance boost:

\begin{lstlisting}
>>> %timeit my.pythranized(b,c)
1000 loops, best of 3: 418 us per loop
\end{lstlisting}

Although the AVX instruction set makes it possible to store 4 double precision
floats, one does not get a 4x speed up because of the (unaligned) memory transfers
to and from vector registers.

Finally, using both expression templates, vectorization and OpenMP:

\begin{lstlisting}
>>> %timeit my.pythranized(b,c)
1000 loops, best of 3: 105 us per loop
\end{lstlisting}

The 4 hyper-threaded cores give an extra performance boost. Unfortunately, the
load is not sufficient to get more than an average 4x speed up compared to the
vectorized version. In the end, Pythran generates a native module that performs
roughly 11 times faster than the original version.

As a reference, the \texttt{numexpr} module that performs JIT optimization of the
expression yields the following timing:

\begin{lstlisting}
>>> %timeit numexpr.evaluate("sqrt(b*b+c*c)")
1000 loops, best of 3: 395 us per loop
\end{lstlisting}

\noindent Pythran still performs almost four times faster.

\section{Explicit Parallelization}
\label{sec:openmp}

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developer to
be able to take advantage of them.

As explained in the previous section, Pythran takes advantage of multiple cores
when compiling Numpy expressions. However, when possible, it is often more
profitable to parallelize the outermost loops rather than the inner loops
---the Numpy expressions--- because it avoids the synchronization barrier at
the end of each parallel section, and generally offers more computation
intensive computations.

The OpenMP standard\cite{openmp3.1} is a widely used solution for Fortran, C
and C++ to describe loop-based and task-based parallelism. It consists of a few
directives attached to the code, that describe parallel loops and parallel code
sections in a shared memory model.

Pythran makes this directives available at the Python level through string
instructions. The semantic is roughly similar to the original semantics,
assuming that all variables have function level scope.

The following listing gives a simple example of explicit loop-based
parallelism.  OpenMP 3.0 task-based parallelism form is also supported, the
interested reader may refer to~\cite{pyhpc2013} for more details.

\begin{lstlisting}
def pi_estimate(darts):
    hits = 0
    #omp parallel for reduction(+:hits)"
    for i in xrange(darts):
        x,y = random(), random()
        dist = sqrt(pow(x, 2) + pow(y, 2))
        if dist <= 1.0:
            hits += 1.0
    pi = 4 * (hits / DARTS)
    return pi
\end{lstlisting}

The loop is flagged as parallel, performing a reduction using the \texttt{+}
operator on the \texttt{hits} variable. Variables \texttt{x}, \texttt{y} and
\texttt{dist} are automatically made local by Pythran after a
\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, i.e. local to
a thread and not shared with other threads, by OpenMP.

Next section performs an in-depth comparison of Pythran with for Python
optimizers: Numba, PyPy, ShedSkin and numexpr.

\section{Benchmarks}
\label{sec:benchmarks}

All benchmarks presented in this section are ran on an hyper-threaded quad-core
i7, using examples shipped along Pythran sources, available at
https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
depending on the target.

Cython is not considered in most benchmarks, because to get an efficient
binary, one needs to rewrite the original code, while all the considered tools
are running the very same Python code that remains compatible with CPython. The
experiment was only done to have a comparison with Numexpr.

Pystone is a Python translation of whetstone, a famous floating point number
benchmarks that dates back to Algol60 and the 70's. Although non representative
of real applications, it illustrates the general performance of floating point
number manipulations. Table \ref{tbl:pystone}  illustrates the benchmark
result for CPython, PyPy, ShedSkin and Pythran, using an input value of
\texttt{10**3}. Note that the original version has been updated to replace the user
class by a function call.

\begin{table}
    \centering

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin  \\
    \hline
     Timing  &  861ms      &   11.8ms      &     29.1ms &  24.7ms    \\
    \hline
     Speedup &  $\times$1         &   $\times$72.9       &    $\times$29.6   &  $\times$34.8     \\
    \hline
\end{tabular}

    \caption{Benchmarking result on the Pystone program.}
    \label{tbl:pystone}

\end{table}

It comes at no surprise that all tools get more than decent results on this
benchmark. PyPy generates a code almost as efficient as ShedSkin. Altough both
generate C++, Pythran outperforms ShedSkin thanks to a higher level generated
code. For instance all arrays are represented in ShedSkin by pointers to arrays
that likely disturbs the g++ optimizer, while Pythran uses a vector class wrapping
shared pointers.

Nqueen is a benchmark extracted from the former Unladen Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
is particularly interesting as it makes an intensive use of non-trivial
generator expressions and integer sets. Table\ref{tbl:nqueen} illustrates
the benchmark results for CPython, PyPy, ShedSkin and Pythran. The code had to
be slightly updated to run with ShedSkin because type inference in ShedSkin does
not support mixed scalar and \textbf{None} variables. The input value is \texttt{9}.

\begin{table}
    \centering

    \begin{tabular}{|l|c|c|c|c|}
        \hline
     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin \\
    \hline
     Timing  &  1904.6ms   &   358.3ms     &    546.1ms &  701.5ms  \\
    \hline
     Speedup &  $\times$1         &    $\times$5.31      &    $\times$3.49   &  $\times$2.71    \\
    \hline
\end{tabular}
\caption{Benchmarking result on the NQueen program.}
\label{tbl:nqueen}

\end{table}

It seems that compilers have difficulties to take advantage of high level
constructs such as generator expressions, as the overall speedup is not
breathtaking. Pythran benefits from the conversion to \texttt{itertools.imap} here,
while ShedSkin and PyPy rely on more costly constructs. A deeper look at the
Pythran profiling trace shows that more than half of the execution time is
spent allocating and deallocating a \texttt{set} used in the internal loop. There is a
memory allocation invariant that could be taken advantage of there, but none of
the compiler does.

Hyantes\footnote{\url{http://hyantes.gforge.inria.fr}} is a geomatic application that exhibits typical usage of arrays
using loops instead of generalized expressions. It is helpful to measure the
performance of direct array indexing.

Table\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
ShedSkin and Pythran, when using lists as the data container. The output window
used is \texttt{100x100}.

\begin{table}
    \centering

    \begin{tabular}{|l|c|c|c|c|}

        \hline
     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin  \\
    \hline
     Timing  &  1295.4ms   &   270.5ms     &    277.5ms &  281.5ms   \\
    \hline
     Speedup &  x1         &    x4.79      &    x4.67   &  x4.60     \\
    \hline
\end{tabular}
\caption{Benchmarking result on the hyantes kernel, list version.}
\label{tbl:hyantes}

\end{table}

The speed ups are not amazing for a numerical application. there are two
reasons for this poor speedups. First, the \texttt{hyantes} benchmark makes heavy
usage of trigonometric functions, and there is not much gain there. Second, and
most important, the benchmark produces a big 2D array stored as a list of list,
so the application suffers from the heavy overhead of converting them from C++
to Python. Running the same benchmark using Numpy arrays as core containers
confirms this assumption, as illustrated by Table \ref{tbl:np-hyantes}. This
table also demonstrates the benefits of manual parallelization using OpenMP.

\begin{table}
    \centering

    \begin{tabular}{|l|c|c|c|c|}

        \hline
     Tool    &  CPython    &   Pythran     & Pythran+OpenMP   \\
    \hline
     Timing  &  450.0ms    &   4.8ms       &      2.3ms       \\
    \hline
     Speedup &  x1         &    x93.8      &    x195.7        \\
    \hline
\end{tabular}
\caption{Benchmarking result on the hyantes kernel, numpy version.}
\label{tbl:np-hyantes}
\end{table}

Finally, \texttt{arc\_distance} \footnote{The \texttt{arc\_distance} testbed is taken from to \url{https://bitbucket.org/FedericoV/numpy-tip-complex-modeling}.} presents a classical usage of Numpy expression. It
is typically more efficient than its loop alternative as all the iterations are
done directly in C. Its code is reproduced below:

\begin{lstlisting}
def arc_distance(theta_1, phi_1, theta_2, phi_2):
    """
    Calculates the pairwise arc distance
    between all points in vector a and b.
    """
    temp = (np.sin((theta_2-theta_1)/2)**2
        + np.cos(theta_1)*np.cos(theta_2)
          * np.sin((phi_2-phi_1)/2)**2)
    distance_matrix = 2 * np.arctan2(
            sqrt(temp),sqrt(1-temp))
    return distance_matrix
\end{lstlisting}

Table \ref{tbl:arc-distance} illustrates the benchmark result for CPython,
Cython, Numexpr and Pythran, using random input arrays of \texttt{10**6} elements.
Table \ref{tbl:arc-distance-2} details the Pythran performance. Cython code
is written using the \texttt{parallel.prange} feature and compiled with
\texttt{-fopenmp -O2 -march=native}.

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
     Tool    &  CPython    &  Cython  &  Numexpr    & Pythran   \\
    \hline
     Timing  &  192.2ms    &  36.0ms  &    41.2ms   &  17.1ms   \\
    \hline
     Speedup &  x1         &  x5.33   &  x4.67      &  x11.23   \\
    \hline
\end{tabular}
\caption{Benchmarking result on the arc distance kernel.}
\label{tbl:arc-distance}
\end{table}

\begin{table}
    \centering
    \begin{tabular}{|l|c|c|c|c|}
    \hline
    Tool & Pythran (raw) & Pythran (+AVX) & Pythran (+OMP) & Pythran (full)  \\
    \hline
    Timing & 186.3ms     &    75.4ms      &    41.1ms      &  17.1ms         \\
    \hline
    Speedup&    x1.03      &    x2.54       &    x4.67       &  x11.23         \\
    \hline
\end{tabular}
\caption{Benchmarking result on the arc distance kernel, Pythran details.}
\label{tbl:arc-distance-2}
\end{table}

It shows a small benefit from using expression templates on their own, most
certainly because the loop control overhead is negligible in front of the
trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
using it. The benefit of OpenMP, although related to the number of cores, makes
a whole speedup greater than x11 over the original Numpy version, without
changing the input code. Quite the opposite, Numexpr requires rewriting the input
and does not achieve the same level of performance as Pythran when OpenMP and
AVX are combined.

Writing efficient Cython code requires more work than just typing the variable
declarations using Cython's specific syntax: it only takes advantage of
parallelism because we made it explicit. Without explicit parallelization,
the generated code runs around 176ms instead of 36ms. Cython does not generate
vectorized code, and \texttt{gcc} does not vectorize the inner loop, which explains
the better result obtained with Pythran.

\section{Future Work}

Although Pythran focuses on a subset of Python and its standard library, many
optimizations opportunities are still possible. Using as Domain Specific
Language(DSL) approach, one could use rewriting rules to optimize several
Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
\texttt{count\_uniq} that would iterate only once on the input sequence.

There is naturally more work to be done at the Numpy level, for instance to
support more functions from the original module. The extraction of Numpy
expressions from \texttt{for loops} is also a natural optimization candidate, which
shares similarities with code refactoring.

Numpy expressions also fit perfectly well in the polyhedral model. Exploring
the coupling of polyhedral tools with the code generated from Pythran offers
enthusiastic perspectives.

\section{Conclusion}

This paper presents the Pythran compiler, a translator, and an optimizer, that
converts Python to C++. Unlike existing static compilers for Python, Pythran
leverages several function-level or module-level analyses to provide several
generic or Python-centric code optimizations. Additionally, it uses a C++
library that makes heavy usage of template programming to provide an efficient
API similar to a subset of Python standard library. This library takes
advantage of modern hardware capabilities ---vector instruction units and
multi-cores--- in its implementation of parts of the \texttt{numpy} package.

This paper gives an overview of the compilation flow, the analyses involved and
the optimizations used. It also compares the performance of compiled Pythran
modules against CPython and other optimizers: ShedSkin, PyPy and numexpr.

To conclude, limiting Python to a statically typed subset does not hinders the
expressivity when it comes to scientific or mathematic computations, but makes
it possible to use a wide variety of classical optimizations to help Python
match the performance of statically compiled language. Moreover, one can use
high level information to generate efficient code that would be difficult to write for the average programmer.

\acks

This project has been partially funded by the CARP Project and the SILKAN
Company.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{biblio}

\end{document}

### 2014-01-28 doc/papers/iop2014/paper.tex
Review IOP 2014 paper

@@
     perfect forwarding, as well as classical ones such as expression templates.
+    % FIXME Expression template is not a "feature" of the language IMO
 
@@
     more, without the backward compatibility loss of Cython. Numpy expressions
-    run faster than when compiled with numexpr, or numba, without any change to
+    run faster than when compiled with numexpr or numba, without any change to
     the original code.
@@
 Several tools have been proposed by an active community to fill the performance
-gap met when running these computation-intensive piece of code, either through
-static compilation or Just In Time (JIT) compilation.
+gap encountered when running these computation-intensive piece of code, either
+through static compilation or Just In Time (JIT) compilation.
 
@@
 static optimization techniques that have been known for decades and
+% FIXME And Numba? It generates LLVM, which uses static optimization techniques
 successfully applied to statically compiled language such as C or C++.
@@
 detection of pure functions, temporary allocation removal or constant folding.
-These transformations are backed up by code analysis such as aliasing,
+These transformations are backed up by code analyses such as aliasing,
 inter-procedural memory effect computations or use-def chains.
@@
 introduces the Pythran compiler compilation flow and internal representation.
-Section~\ref{sec:analysis} presents several code analysis while
+Section~\ref{sec:analysis} presents several code analyses while
 Section~\ref{sec:optimizations} focuses on code optimizations.
@@
 annotations for explicit parallelization of Python programs and
-section~\ref{sec:benchmarks} presents the performance obtained on a few
+Section~\ref{sec:benchmarks} presents the performance obtained on a few
 synthetic benchmarks and concludes.
@@
 \begin{lstlisting}
-def dot(l0, l1):
-    return sum(x*y for x,y in zip(l0,l1))
+def dot(L0, L1):
+    return sum(x*y for x,y in zip(L0,L1))
 \end{lstlisting}
@@
 template<class T0, class T1>
-    auto dot(T0&& l0, T1&& l1) -> decltype(/* skipped */)
+    auto dot(T0&& L0, T1&& L1) -> decltype(/* skipped */)
     {
@@
                     pythonic::zip(
-                        std::forward<T0>(l0),
-                        std::forward<T1>(l1))
+                        std::forward<T0>(L0),
+                        std::forward<T1>(L1))
             )
@@
 form of structural typing: the only assumption these two versions make are that
-\texttt{l0} and \texttt{l1} are iterable, their content can be multiplied and the result of
+\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and the result of
 the multiplication is accumulatable.
@@
 
-The second step only consists in the instantiation of the top-level functions of the
-module, using user-provided signatures. Template instantiation then triggers the
-different correctly typed instantiations for all functions written in the
+The second step only consists in the instantiation of the top-level functions of
+the module, using user-provided signatures. Template instantiation then triggers
+the different correctly typed instantiations for all functions written in the
 module. Note that the user only needs to provide the type of the functions
 exported outside the module. The possible types of all internal functions are
-then inferred from the call sites.
-
-The last step involves a template library, called \texttt{pythonic} that contains a
-polymorphic implementation of many functions from the Python standard library
-in the form of C++ template functions. Several optimizations, most notably
-expression template, are delegated to this library. Pythran relies on the
-C++11\cite{isocxx11} language, as it makes heavy use of recent features such as
-move semantics, type inference through \texttt{decltype(\dots)} and variadic templates.
-As a consequence it requires a compatible C++ compiler for the native code
+then inferred from the call sites. Note that this complex step leverage the 
+C++11 compiler capabilities: Pythran generates only one line using user-provided
+signature to trigger the template instanciation by the C++ compiler.
+
+The last step involves a template library, called \texttt{pythonic} that 
+contains a polymorphic implementation of many functions from the Python standard
+library in the form of C++ template functions. Several optimizations, most
+notably using expression template, are delegated to this library. Pythran relies
+on the C++11\cite{isocxx11} language, as it makes heavy use of recent features 
+such as move semantics, type inference through \texttt{decltype(\dots)} and 
+variadic templates.
+As a consequence it requires a C++11 compatible compiler for the native code
 generation. Boost.Python\cite{boostpython2007} is involved for the automatic
-Python-to-C++ boilerplate.  Generated C++ code is compatible with g++ 4.7.2 and
-clang++ 3.2.
+Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
+4.7.2 and clang++ 3.2.
 
@@
 type-agnostic, i.e. they do not assume any type for the variables manipulated
-by the program. Type specialization is only done in the back-end, right before
-native code generation. Said otherwise, the Pythran compiler analyzes
-polymorphic functions and turns Python modules into C++ meta-programs.
-
-Figure\ref{fig:pythran-compiler} summarizes the compilation flow and the involved
-tools.
+by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
+Said otherwise, the Pythran compiler analyzes polymorphic functions and turns
+Python modules into C++ meta-programs.
+
+Figure~\ref{fig:pythran-compiler} summarizes the compilation flow and the 
+involved tools.
 
@@
 A code analysis is a function that takes a part of the IR (or the whole
-module's IR) as input and returns aggregated high-level information. For
+module's IR) as an input and returns aggregated high-level information. For
 instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
@@
 have a side-effect). The fact that the last \textbf{use} is always preceded by
-a \textbf{def}taht only point to it indicates that these \textbf{def}s can
+a \textbf{def} that only point to it indicates that these \textbf{def}s can
 safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
 renaming).
+% FIXME The scalar renaming part is unclear
 
@@
 advocates in favor of using inlining. This transformation comes at no cost when
-using C++ as the back-end language, as the C++ compiler does it.
+using C++ as the back-end language, as the C++ compiler performs this 
+transformation triggered by fine tuned heuristics.
 
@@
     # this map can be converted to imap
-    l = map(math.cos, range(n))
+    L = map(math.cos, range(n))
     return sum(l) # sum iterates once on its input
@@
     # this map cannot be converted to imap
-    l = map(math.cos, range(n))
-    l[0] = 1  # invalid assignment
-    return sum(l) + max(l) # sum iterates once
+    L = map(math.cos, range(n))
+    L[0] = 1  # invalid assignment if converted to imap
+    return sum(L) + max(L) # two iterations over L prevents imap transformation
 \end{lstlisting}
@@
 Pythran makes this directives available at the Python level through string
-instructions. The semantic is roughly similar to the original semantics,
-assuming that all variables have function level scope.
+instructions. The semantic is roughly similar to the semantics defined by the
+standard, assuming that all variables have function-level scope.
 
@@
     hits = 0
-    #omp parallel for reduction(+:hits)"
+    "#omp parallel for reduction(+:hits)"
     for i in xrange(darts):
@@
 
-All benchmarks presented in this section are ran on an hyper-threaded quad-core
-i7, using examples shipped along Pythran sources, available at
+All benchmarks presented in this section are run on an hyper-threaded quad-core
+i7, using examples shipped along with Pythran sources, available at
 https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
@@
 
-Pystone is a Python translation of whetstone, a famous floating point number
+Pystone is a Python translation of Whetstone, a famous floating point number
 benchmarks that dates back to Algol60 and the 70's. Although non representative
 of real applications, it illustrates the general performance of floating point
-number manipulations. Table \ref{tbl:pystone}  illustrates the benchmark
+number manipulations. Table~\ref{tbl:pystone}  illustrates the benchmark
 result for CPython, PyPy, ShedSkin and Pythran, using an input value of
@@
 is particularly interesting as it makes an intensive use of non-trivial
-generator expressions and integer sets. Table\ref{tbl:nqueen} illustrates
+generator expressions and integer sets. Table~\ref{tbl:nqueen} illustrates
 the benchmark results for CPython, PyPy, ShedSkin and Pythran. The code had to
@@
 
-Table\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
+Table~\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
 ShedSkin and Pythran, when using lists as the data container. The output window
@@
 to Python. Running the same benchmark using Numpy arrays as core containers
-confirms this assumption, as illustrated by Table \ref{tbl:np-hyantes}. This
+confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
 table also demonstrates the benefits of manual parallelization using OpenMP.
@@
 
-Table \ref{tbl:arc-distance} illustrates the benchmark result for CPython,
+Table~\ref{tbl:arc-distance} illustrates the benchmark result for CPython,
 Cython, Numexpr and Pythran, using random input arrays of \texttt{10**6} elements.
-Table \ref{tbl:arc-distance-2} details the Pythran performance. Cython code
+Table~\ref{tbl:arc-distance-2} details the Pythran performance. Cython code
 is written using the \texttt{parallel.prange} feature and compiled with
@@
 \end{tabular}
-\caption{Benchmarking result on the arc distance kernel, Pythran details.}
+\caption{Detailed benchmarking result on the arc distance kernel for Pythran.}
 \label{tbl:arc-distance-2}
@@
 a whole speedup greater than x11 over the original Numpy version, without
-changing the input code. Quite the opposite, Numexpr requires rewriting the input
-and does not achieve the same level of performance as Pythran when OpenMP and
-AVX are combined.
+changing the input code. Almost on the opposite, Numexpr requires rewriting the 
+input and does not achieve the same level of performance as Pythran when OpenMP
+and AVX are combined.
 
@@
 the generated code runs around 176ms instead of 36ms. Cython does not generate
-vectorized code, and \texttt{gcc} does not vectorize the inner loop, which explains
-the better result obtained with Pythran.
+vectorized code, and \texttt{gcc} does not vectorize the inner loop, which 
+explains the better result obtained with Pythran.
 
@@
 Although Pythran focuses on a subset of Python and its standard library, many
-optimizations opportunities are still possible. Using as Domain Specific
+optimizations opportunities are still possible. Using a Domain Specific
 Language(DSL) approach, one could use rewriting rules to optimize several

### 2014-01-28 doc/papers/iop2014/paper.tex
X. Corbillon proofread
Update IOP2014 paper with merge from Adrien's diff

@@
     Pythran is an open source static compiler that turns modules written
-    in a subset of Python into native ones. Based on the fact that scientific
+    in a subset of Python language into native ones. Assuming that scientific
     modules do not rely much on the dynamic features of the language, it trades
-    them in favor of powerful, eventually inter procedural, optimizations.
-    These include detection of pure functions, temporary allocation removal,
+    them for powerful, eventually inter procedural, optimizations.
+    These optimizations include detection of pure functions, temporary
+    allocation removal,
     constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
@@
 
-    In addition to these compilation steps, Pythran provides a C++ runtime library that
+    In addition to these compilation steps, Pythran provides a C++ runtime 
+    library that
     leverages the C++ STL to provide generic containers, and the Numeric 
+    % FIXME (phrase trop longuuuuuuuuue) 
     Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
@@
 
+    % FIXME  (explain that annotations can be added to the code?)
     The input code remains compatible with the Python interpreter, and output
     code is generally as efficient as the annotated Cython equivalent, if not
-    more, without the backward compatibility loss of Cython. Numpy expressions
+    more, % FIXME (cut sentence) 
+    without the backward compatibility loss of Cython. Numpy expressions
     run faster than when compiled with numexpr or numba, without any change to
@@
 The Python language is growing in popularity as a language for scientific
-computing, mainly thanks to a concise syntax, a high level standard library and
+computing, thanks to % FIXME (sounds strange for a publication)
+ its concise syntax, a high level standard library and
 several scientific packages.
@@
 compared to the same algorithm written in a statically compiled language such
-as C is high, due to numerous dynamic lookup and interpretation cost inherent
-in high level languages. Additionally, the Python compiler performs no
+as C is high, % FIXME cut sentence
+due to numerous dynamic lookup and interpretation cost inherent
+in high level languages. Additionally, the Python compiler does not perform
 optimization on the byte code, while scientific applications are first-class
@@
 
-Following the saying that scientific applications spend 90\% of their time in
-10\% of the code, it is natural to focus on computation-intensive piece of code.
-The aim may not be to optimize the full Python application, but rather a
-small subset of the application.
-
-Several tools have been proposed by an active community to fill the performance
+Since scientific applications are often told to spend 90\% of their time in
+10\% of the code, it is natural to focus on those computation-intensive pieces 
+of code. The aim % FIXME (of what) 
+may not be to optimize the full Python application, but rather a subset of the application.
+
+Several tools have been developed by an active community to fill the performance
 gap encountered when running these computation-intensive piece of code, either
+% FIXME after? (perfs issues are not on the compilation (except for...))
 through static compilation or Just In Time (JIT) compilation.
 
-An approach used by Cython\cite{cython2010} is to suppress the interpretation
+One approach, used by Cython \cite{cython2010} is to suppress the interpretation
 overhead by translating Python Programs to C programs calling the Python C
 API\cite{pythoncapi}. More recently, Nuitka\cite{nuitka}  has taken the same
-approach using C++ has a back-end. Going a step further Cython also provides an
-hybrid C/Python language that can efficiently be translated to C code, relying
-on the Python C API for some parts and on plain C for others.
-ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python program
+approach using C++ as a back-end. Going a step further Cython also provides an
+hybrid C/Python language that can efficiently be translated into C code. It 
+relies both on the Python C API and on plain C.
+ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python programs
 into C++, without any call to the Python C API, but fails to compile dynamic
@@
 interpreter, to dynamically turn the computation intensive parts into native
-code. The \texttt{numexpr} module\cite{numexpr} does so for Numpy expressions
-by JIT-compiling them from a string representation to native code.
+code. This is what the \texttt{numexpr} module\cite{numexpr} does for Numpy 
+expressions, JIT-compiling them from a string representation to native code.
 Numba\cite{numba} and Parakeet\cite{parakeet2012} extend this approach to
 Numpy-centric applications while PyPy\cite{pypy2009} applies it to the whole
-language, including the Numpy module through the in-progress \texttt{numpypy}
+language - including the Numpy module through the in-progress \texttt{numpypy}
 branch.
 
-To the notable exception of PyPy, these compilers do not apply any of the
-static optimization techniques that have been known for decades and
+Except for PyPy, none of these compilers apply any of the
+static optimization techniques that have been known and successfully applied 
 % FIXME And Numba? It generates LLVM, which uses static optimization techniques
-successfully applied to statically compiled language such as C or C++.
+for decades to statically compiled language such as C or C++.
 Translators to statically compiled languages do take advantage of them
-indirectly, but the quality of generated code may prevent advanced
-optimizations, such as vectorization, while they are available at higher level,
+indirectly, but the quality of the generated code may prevent advanced
+optimizations, such as vectorization, %FIXME too long
+while they are available at higher level,
 i.e. at the Python level. Taking into account the specificities of the Python
-language can unlock many new transformations. For instance, PyPy automates the
-conversion of the \texttt{range} builtin into \texttt{xrange} through the use
-of a dedicated structure called \texttt{range-list}.
+language can unlock many new transformations. %FIXME sentence without subject
+For instance, PyPy automatically
+converts the \texttt{range} builtin into \texttt{xrange} by using a dedicated
+structure called \texttt{range-list}.
 
@@
 computation-intensive parts of the program are moved to a module fed to
-Pythran. The module is translated to polymorphic C++ code through the
-generation of templated classes in place of Python functions, then these
-classes are either used from regular C++ code or instantiated for given types
-to generate a native Python module.
+Pythran. The module is then translated to polymorphic C++ code through the
+generation of templated classes in place of Python functions. These
+classes are either used from regular C++ % FIXME (not clear what do you mean?)
+code or instantiated for given types to generate a native Python module.
 
 It is a strong requirement for Pythran to maintain backward compatibility with
-Python, that is unlike Cython, all code compilable by Pythran are still
-executable by a standard-conforming Python interpreter. For that reason, in
+Python, that is, unlike Cython, all code compilable by Pythran is still
+executable by a standard-conforming Python interpreter. For this reason, in
 addition to language restrictions detailed in the following, Pythran
@@
 
-as optional module signature. One does not need to list all the module
+as an optional module signature. One does not need to list all the module
 functions in an \texttt{export} directive, only the functions meant to be used
-outside of the module. Polymorphic functions can be listed several times with
+outside of the module, i.e imported by another module.
+Polymorphic functions can be listed several times with
 different types. These type annotations are only used for native code generation.
@@
 
-    \item turn Python code into Python Abstract Syntax Tree (AST) thanks to the \texttt{ast}
+    \item turn Python code into Python Abstract Syntax Tree (AST) by using the \texttt{ast}
    module from the standard library;
@@
 
-Pythran IR is similar to Python AST, as defined in the \texttt{ast} module, except
-that several nodes are forbidden (most notably Pythran does not support
+Pythran IR is similar to the Python AST, as defined in the \texttt{ast} module, except
+that several nodes are forbidden (for instance because Pythran does not support
 user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
-to others to form a simpler AST easier to deal with for further analyses and
+to others to form a simpler and easier to deal with AST for further analyses and
 optimizations. The transformations applied by Pythran on Python AST are the
@@
         function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;
+        % FIXME (this sentence makes my head hurt :s maybe cut into smaller sentences)
 
@@
 
-    \item all imports are fully expanded to make function and global variables access paths explicit,
+    \item all imports are fully expanded to make functions and global variables access paths explicit,
 
@@
     \item \texttt{try \dots finally} constructs are turned into nested \texttt{try \dots except} blocks;
-
-    \item identifiers whose name may clash with C++ keywords are renamed.
+        % FIXME (Is the difference between construct and block obvious for intended readers?)
+
+    \item identifiers whose name may be in conflict with C++ keywords are renamed.
 
@@
 
-The back-end works in three steps:
+The back-end operates in three steps:
 
@@
 time, but it cannot be assigned to a string if it has already been assigned to
-an integer. As shown later, it is possible to detect several false variable
+an integer. % FIXME (not sure if clear enough) 
+As shown later, it is possible to detect several false variable
 polymorphism cases using use-def chains. Function polymorphism is achieved
@@
 similar to Python's duck typing, except that it is checked at compile time, as
+% FIXME (tooooooooooooooooooooooo long) 
 illustrated by the following implementation of a generic dot product in Python:
@@
 
-Although far more verbose than the Python version, the C++ version also uses a
-form of structural typing: the only assumption these two versions make are that
-\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and the result of
-the multiplication is accumulatable.
+Although far more verbose % FIXME  (don't think you can use verbose this way)
+than the Python version, the C++ version also uses some
+form of structural typing: the only assumptions these two versions make are that
+\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and 
+that the result of the multiplication is accumulatable.
 
 Finally, Pythran computes the set of functions and types used for the processed
-module, and generate the minimal set of headers that need to be included to
+module, and generates the minimal set of headers that need to be included to
 provide the definitions needed to compile the generated C++ code. This step is
-critical to prevent typical slowdown from C++ compilation.
+critical to prevent typical slowdown during C++ compilation.
 
@@
 the different correctly typed instantiations for all functions written in the
+module. % FIXME (not only the functions called by top-level functs????)
 module. Note that the user only needs to provide the type of the functions
@@
 The last step involves a template library, called \texttt{pythonic} that 
-contains a polymorphic implementation of many functions from the Python standard
+contains a polymorphic implementation of many functions of the Python standard
 library in the form of C++ template functions. Several optimizations, most
 notably using expression template, are delegated to this library. Pythran relies
-on the C++11\cite{isocxx11} language, as it makes heavy use of recent features 
-such as move semantics, type inference through \texttt{decltype(\dots)} and 
-variadic templates.
+on the C++11\cite{isocxx11} language and leverage recent features such as
+move semantics, type inference through \texttt{decltype(\dots)} and variadic
+templates.
 As a consequence it requires a C++11 compatible compiler for the native code
-generation. Boost.Python\cite{boostpython2007} is involved for the automatic
+generation. Boost.Python\cite{boostpython2007} is used for the automatic
 Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
@@
 by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
-Said otherwise, the Pythran compiler analyzes polymorphic functions and turns
+In other words, the Pythran compiler analyzes polymorphic functions and turns
 Python modules into C++ meta-programs.
@@
 of all identifiers used throughout the program. This information is later used
-when the creation of new identifiers is required so that no conflict occurs
-with existing ones.
+when creating new identifiers so that no conflict occurs with already existing ones.
 
@@
 As we do not specialise the analysis for different types and the truth value of
-\texttt{a} is unknown at compilation time, the alias analysis yields the approximated
-result that \texttt{c} may point to a fresh location, \texttt{a} or \texttt{b}.
+\texttt{a} is unknown at compilation time, the alias analysis yields % FIXME (sounds strange) 
+the approximated result that \texttt{c} may point to a fresh location, 
+\texttt{a} or \texttt{b}.
 
 Without this kind of information, even a simple instruction like
-\texttt{sum(a)} would yield very few informations as there is no guarantee that
-the \texttt{sum} identifiers points to the \texttt{sum} built-in, as someone
-could have previously stated that \texttt{sum = len}.
-
-When turning Python AST to Pythran IR, nested functions are turned into global
-functions taking their expanded closure as parameter. This closure is computed using the
-information provided by the \texttt{Globals} analysis that statically computes the
-state of the dictionary of globals, and \texttt{ImportedIds} that computes the set of
+\texttt{sum(a)} would yield very few information as there is no guarantee that
+the \texttt{sum} identifiers point to the \texttt{sum} built-in, as it could have been 
+previously stated that \texttt{sum = len}.
+
+When turning Python AST into Pythran IR, nested functions are turned into global
+functions taking their expanded closure as parameter. This closure is computed 
+using the information provided by the \texttt{Globals} analysis which 
+statically computes the state of the dictionary of globals, and 
+\texttt{ImportedIds} that computes the set of % FIXME (too long)
 identifiers used by an instruction but not declared in this instruction. For
@@
 set of functions declared in the module that are pure, i.e. whose return value
-only depends from the value of their argument. This analysis depends on two
+only depends on the value of their arguments. This analysis depends on two
 other analyses, namely \texttt{GlobalEffects} that computes for each function whether
 this function modifies the global state (including I/O, random generators, etc.)
-and \texttt{ArgumentEffects} that computes for each argument of each function whether
-this argument may be updated in the function body. These three analyses work
-inter-procedurally, as illustrated by the following example:
+and \texttt{ArgumentEffects} that computes for each argument of each function 
+whether or not this argument may be updated in the function body. These three
+analyses work inter-procedurally, as illustrated in the following example:
 
@@
 
-The \texttt{fibo} function is pure as it has no global effects or argument effects and
-only calls itself. As a consequence the \texttt{bar} function is also pure as the
+The \texttt{fibo} function is pure as it has no global effect or argument effect and
+only calls itself. As consequence the \texttt{bar} function is also pure since the
 \texttt{map} intrinsic is pure when its first argument is pure. However the \texttt{foo}
-function is not pure as it calls the \texttt{sample} function from the \texttt{random}
+function is not pure since it calls the \texttt{sample} function from the \texttt{random}
 module, which has a global effect (on the underlying random number generator
@@
 This is the case for the first \texttt{map} in the following snippet, but not
-for the second because the \texttt{print b} involves an \textbf{I/O}.
+for the second because \texttt{print b} involves an \textbf{I/O}.
 
@@
 
-\texttt{ConstantExpressions} uses function purity to decide whether a given expression
+\texttt{ConstantExpressions} uses function purity to decide whether or not a given expression
 is constant, i.e. its value only depends on literals. For instance the
@@
 argument is a literal. 
+% FIXME (pourrait etre interressant de repondre a : et t'en fait quoi calcul a la compilation? calcul une fois si appele plusieurs fois ?... ou ref a la transformation qui fait ca)
 
@@
 The result can be used to drive various code transformations, for instance to
-remove dead code, as a \textbf{def} followed by a \textbf{def} or nothing is useless. It is
-used in Pythran to avoid false polymorphism. An intuitive way to represent
-use-def chains is illustrated on next code snippet:
+remove dead code, such as a \textbf{def} followed by aeither nothing or a 
+\textbf{def} is useless. 
+It is used in Pythran to avoid false polymorphism. An intuitive way to represent
+use-def chains is illustrated on the next code snippet:
 
@@
 assignment can be removed (but not necessarily its right hand part that could
-have a side-effect). The fact that the last \textbf{use} is always preceded by
-a \textbf{def} that only point to it indicates that these \textbf{def}s can
+have a side-effect) % FIXME example?
+The fact that the last \textbf{use} is always preceded by
+a \textbf{def} that only points to it indicates that these \textbf{def}s can
 safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
@@
 
-All the above analyses are used by the Pythran developer to build code
+All the above analyses are used by Pythran developers to build code
 transformations that improve the execution time of the generated code.
@@
 
-However, there are some informations available at the Python level that cannot
+However, there are some information available at the Python level that cannot
 be recovered at the C++ level. For instance, Pythran uses functor with an
-internal state and a goto dispatch table to represent generators. Although
-effective, this approach is not very efficient, especially for trivial cases.
+internal state and a goto dispatch table to represent generators. This approach
+is not very efficient, especially for trivial cases.
 Such trivial cases appear when a generator expression is converted, in the
@@
 
-The information concerning constant expressions is used to perform a classical
+The information regarding constant expressions is used to perform a classical
 transformation called \texttt{ConstantFolding}, which consists in the compile-time
@@
 
-\texttt{ConstantFolding} may create a lot of candidates for \texttt{LoopUnrolling}.
-In Python, loops are used to iterate over sequences.
-If such a sequence is an explicit list, then Pythran unrolls the loop accordingly.
-Thanks to \texttt{ConstantFolding}, a call to \texttt{range(4)} will statically turn into \texttt{[0, 1, 2, 3]} (after a check that \texttt{range} actually points to \texttt{\_\_builtin\_\_.range}) making it possible to unroll the loop statically.
-
-Sometimes, coders use the same variable in a function to represent value with
-different types, which leads to false polymorphism, as in:
+\texttt{ConstantFolding} may create a lot of candidates for 
+\texttt{LoopUnrolling}. In Python, loops are used to iterate over sequences. If
+such a sequence is an explicit list, then Pythran unrolls the loop accordingly.
+Thanks to \texttt{ConstantFolding}, a call to \texttt{range(4)} will statically
+turn into \texttt{[0, 1, 2, 3]} (after a check that \texttt{range} actually
+points to \texttt{\_\_builtin\_\_.range}) making it possible to unroll the loop
+statically.
+
+Sometimes, developers use the same variable in a function to represent value 
+with different types, which leads to false polymorphism, as in:
 
@@
 algorithms such as \texttt{accumulate} are used when possible. GMP is the
-natural pick to represent Python's \texttt{long} in C++. NT2 provides a generic
+natural choice to represent Python's \texttt{long} in C++. NT2 provides a generic
 vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
-the vector instruction units of modern processors in a generic way. It is used
+the use of vector instruction units of modern processors in a generic way. It is used
 to efficiently compile Numpy expressions.
@@
         multiple iterations and the creation of intermediate arrays. Because
-        they aggregates all \texttt{ufunc} into a single expression at compile
+        they aggregate all \texttt{ufunc} into a single expression at compile
         time, they also increase the computation intensity of the loop body,
-        which increases the impact of the two following optimizations.
+        and hence increase the impact of the two following optimizations.
 
@@
 
-When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core i7, the
-pure Python execution yields:
+When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
+i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
 
@@
 kernels. As modern computers generally feature several processors and several
-cores per processor, it is critical for the scientific application developer to
+cores per processor, it is critical for the scientific application developers to
 be able to take advantage of them.
@@
 and C++ to describe loop-based and task-based parallelism. It consists of a few
-directives attached to the code, that describe parallel loops and parallel code
+directives attached to the code, directives that describe parallel loops and parallel code
 sections in a shared memory model.
@@
 
+% FIXME (Ca semble tomber a plat, pas de resultats, pas de conclusion...)
+
 Next section performs an in-depth comparison of Pythran with for Python
@@
 All benchmarks presented in this section are run on an hyper-threaded quad-core
-i7, using examples shipped along with Pythran sources, available at
+i7, % FIXME sandy-brige, nehalem, what?
+using examples shipped along with Pythran sources, available at
 https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
@@
 are running the very same Python code that remains compatible with CPython. The
-experiment was only done to have a comparison with Numexpr.
+experiment was performed only to obtain a comparison with Numexpr.
 
@@
 It comes at no surprise that all tools get more than decent results on this
-benchmark. PyPy generates a code almost as efficient as ShedSkin. Altough both
+benchmark. PyPy generates a code almost as fast % FIXME (dunno if it's efficient) 
+as ShedSkin. Altough both
 generate C++, Pythran outperforms ShedSkin thanks to a higher level generated
@@
 memory allocation invariant that could be taken advantage of there, but none of
-the compiler does.
+the compiler does so.
 
@@
 
-The speed ups are not amazing for a numerical application. there are two
-reasons for this poor speedups. First, the \texttt{hyantes} benchmark makes heavy
-usage of trigonometric functions, and there is not much gain there. Second, and
+The speed ups are not amazing % FIXME (not formal enough, same for previous usage) 
+for a numerical application. There are two
+reasons for these poor speedups. First, the \texttt{hyantes} benchmark makes heavy
+usage of trigonometric functions, and there is not much to gain there. Second, and
 most important, the benchmark produces a big 2D array stored as a list of list,
-so the application suffers from the heavy overhead of converting them from C++
+hence the application suffers from the heavy overhead of converting them from C++
 to Python. Running the same benchmark using Numpy arrays as core containers
@@
 \end{tabular}
+% FIXME (You changed the order of the compilers, even though pythran is still first why?)
 \caption{Benchmarking result on the arc distance kernel.}
@@
 It shows a small benefit from using expression templates on their own, most
-certainly because the loop control overhead is negligible in front of the
+certainly because the loop control overhead is negligible compared to the
 trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
 using it. The benefit of OpenMP, although related to the number of cores, makes
-a whole speedup greater than x11 over the original Numpy version, without
+an overall speedup greater than x11 compared to the original Numpy version, 
+without
 changing the input code. Almost on the opposite, Numexpr requires rewriting the 
@@
 declarations using Cython's specific syntax: it only takes advantage of
-parallelism because we made it explicit. Without explicit parallelization,
-the generated code runs around 176ms instead of 36ms. Cython does not generate
+parallelism because we make it explicit. Without explicit parallelization,
+the generated code runs around 176ms instead of 36ms. %FIXME speedup in x?
+Cython does not generate
 vectorized code, and \texttt{gcc} does not vectorize the inner loop, which 
@@
 Although Pythran focuses on a subset of Python and its standard library, many
-optimizations opportunities are still possible. Using a Domain Specific
+optimizations opportunities are still present. Using a Domain Specific
 Language(DSL) approach, one could use rewriting rules to optimize several
@@
 
-This paper presents the Pythran compiler, a translator, and an optimizer, that
-converts Python to C++. Unlike existing static compilers for Python, Pythran
-leverages several function-level or module-level analyses to provide several
+This paper presents the Pythran compiler, a translator and an optimizer that
+converts Python into C++. Unlike existing static Python compilers, Pythran
+leverages several function-level or module-level analyses to provide
 generic or Python-centric code optimizations. Additionally, it uses a C++
@@
 advantage of modern hardware capabilities ---vector instruction units and
-multi-cores--- in its implementation of parts of the \texttt{numpy} package.
+multiple-cores--- in its implementation of parts of the \texttt{numpy} package.
 
@@
 it possible to use a wide variety of classical optimizations to help Python
-match the performance of statically compiled language. Moreover, one can use
+match the performances of statically compiled language. Moreover, one can use
 high level information to generate efficient code that would be difficult to write for the average programmer.
+
+% FIXME (just thinking about it, did you say what CPyhton?)
 

### 2014-01-28 doc/papers/iop2014/paper.tex
Pierrick's review

@@
     % FIXME Expression template is not a "feature" of the language IMO
+    % It is more a classical "use" of the language
 
@@
     the original code. 
+    % It may be a little too ... untactful?
 
@@
 generation of templated classes in place of Python functions. These
-classes are either used from regular C++ % FIXME (not clear what do you mean?)
+classes are either used to generate regular C++ % FIXME (not clear what do you mean?) PB : Better?
 code or instantiated for given types to generate a native Python module.
@@
 Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
-4.7.2 and clang++ 3.2.
+4.8.2 and clang++ 3.3.
 
 It is important to note that all Pythran analyses and optimizations are
-type-agnostic, i.e. they do not assume any type for the variables manipulated
-by the program. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
+type-agnostic, \ie{they do not assume any type for the variables manipulated
+by the program}. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
 In other words, the Pythran compiler analyzes polymorphic functions and turns
@@
 A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
-set of functions declared in the module that are pure, i.e. whose return value
-only depends on the value of their arguments. This analysis depends on two
+set of functions declared in the module that are pure, \ie{whose return value
+only depends from the value of their arguments and do not change any global state}.
+This analysis depends on two
 other analyses, namely \texttt{GlobalEffects} that computes for each function whether
@@
 for the second because \texttt{print b} involves an \textbf{I/O}.
+% Which is not really usefull as there is no way to say pythran we want to execut it in parallel
+% (but logging information is still displayed)
 
@@
 \texttt{ConstantExpressions} uses function purity to decide whether or not a given expression
-is constant, i.e. its value only depends on literals. For instance the
+is constant, \ie{its value only depends on literals}. For instance the
 expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
@@
 % FIXME (pourrait etre interressant de repondre a : et t'en fait quoi calcul a la compilation? calcul une fois si appele plusieurs fois ?... ou ref a la transformation qui fait ca)
+% PB : explain later I think.
 
@@
 The result can be used to drive various code transformations, for instance to
-remove dead code, such as a \textbf{def} followed by aeither nothing or a 
+remove dead code, such as a \textbf{def} followed by either nothing or a 
 \textbf{def} is useless. 
@@
 assignment can be removed (but not necessarily its right hand part that could
-have a side-effect) % FIXME example?
-The fact that the last \textbf{use} is always preceded by
-a \textbf{def} that only points to it indicates that these \textbf{def}s can
+have a side-effect). % FIXME example?
+In the use-def Chains, we can gather all nodes tied with a \textbf{use} node or
+preceded by a common \textbf{def} node.
+All these nodes refere to the same variable and can
 safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
-renaming).
+renaming). It avoid false polymorphism.
 % FIXME The scalar renaming part is unclear
+% PB : Is it clearer now? I don't speak about UD node but we don't care, do we?
 
@@
 dispatching table. This kind of transformation cannot be made by the C++
-compiler. For instance, the one-liner \texttt{len(set(vec[i]+i for i in cols))}
+compiler. For instance, the one-liner \texttt{len(set(vec[i]i + i for i in cols))}
 extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
-is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i]+i,cols)))}.
+is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i] + i,cols)))}.
 This new form is less efficient in pure Python (it implies one extra function
@@
     L = map(math.cos, range(n))
-    return sum(l) # sum iterates once on its input
+    return sum(L) # sum iterates once on its input
 
@@
 
-Using the proper library, the C++ language provides an abstraction level close
+Using the proper library, the \cpp language provides an abstraction level close
 to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
-that leverage on the C++ Standard Template Library (STL), the GNU Multiple
+that leverage on the \cpp Standard Template Library (STL), the GNU Multiple
 Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
@@
 algorithms such as \texttt{accumulate} are used when possible. GMP is the
-natural choice to represent Python's \texttt{long} in C++. NT2 provides a generic
+natural choice to represent Python's \texttt{long} in \cpp. NT2 provides a generic
 vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
@@
         and hence increase the impact of the two following optimizations.
+        % what about binary func?
 
@@
 i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
+% PB : Not sure it is needed as audiance don't really care about it.
 
@@
 to and from vector registers.
+% 1.33x speedup is not really good if the best expected is 4x. unaligned explain why it is not better 
+% than 2x but what about the others 0.66?
 
@@
 load is not sufficient to get more than an average 4x speed up compared to the
-vectorized version. In the end, Pythran generates a native module that performs
-roughly 11 times faster than the original version.
+vectorized version.
+% You complain about 4x because you only have 3.98x ???
+In the end, Pythran generates a native module that performs
+roughly 11\times faster than the original version.
 
@@
 be able to take advantage of them.
+%Does python programmer really use it? Or people using it may write their native code?
+% They may write it using python only to prototyping?
 
@@
 \end{lstlisting}
+% why #? I thought string was enough?
 
@@
 \texttt{dist} are automatically made local by Pythran after a
-\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, i.e. local to
-a thread and not shared with other threads, by OpenMP.
+\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, \ie{local to
+a thread and not shared with other threads, by OpenMP}.
 
@@
 
-Next section performs an in-depth comparison of Pythran with for Python
+Next section performs an in-depth comparison of Pythran with others Python
 optimizers: Numba, PyPy, ShedSkin and numexpr.
@@
 0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
-taking the best of all runs. All C++ codes are compiled with g++ 4.7.3, using
+taking the best of all runs. All C++ codes are compiled with g++ 4.8.2, using
 the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
@@
 table also demonstrates the benefits of manual parallelization using OpenMP.
+% Showing parallelization AND np.array at the same time confuse reader as we
+% don't know how much each optimisation contribute in the speed up
 
@@
     """
-    temp = (np.sin((theta_2-theta_1)/2)**2
-        + np.cos(theta_1)*np.cos(theta_2)
-          * np.sin((phi_2-phi_1)/2)**2)
+    temp = (np.sin((theta_2 - theta_1) / 2) ** 2
+        + np.cos(theta_1) * np.cos(theta_2)
+          * np.sin((phi_2 - phi_1) / 2) ** 2)
     distance_matrix = 2 * np.arctan2(
-            sqrt(temp),sqrt(1-temp))
+            sqrt(temp), sqrt(1 - temp))
     return distance_matrix
@@
 It shows a small benefit from using expression templates on their own, most
+% In fact there are small benefits for compiling this code using only pythran without AVX and openMP.
 certainly because the loop control overhead is negligible compared to the
@@
 % FIXME (just thinking about it, did you say what CPyhton?)
+% PB : I think it is ok as audience is Python programmers
 

### 2014-01-28 pythran/tests/cases/vibr_energy.py
Fix bad runas for benchmark

@@ -3,10 +3,10 @@
 #pythran export calculate_vibr_energy(float[], float[], int)
 #pythran export calculate_vibr_energy(float[], float[], float)
 #pythran export calculate_vibr_energy(float[], float[], float [])
-#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; ba = np.cos(np.ones(1000000)) ; n = np.arange(1000000); calculate_vibr_energy(a, b, n)
-#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; ba = np.cos(np.ones(1000000)) ; calculate_vibr_energy(a, b, 10)
-#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; ba = np.cos(np.ones(1000000)) ; calculate_vibr_energy(a, b, 10.)
-#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; ba = np.cos(np.ones(1000000)) ; n = np.arange(1000000, dtype=np.double); calculate_vibr_energy(a, b, n)
+#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; b = np.cos(np.ones(1000000)) ; n = np.arange(1000000); calculate_vibr_energy(a, b, n)
+#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; b = np.cos(np.ones(1000000)) ; calculate_vibr_energy(a, b, 10)
+#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; b = np.cos(np.ones(1000000)) ; calculate_vibr_energy(a, b, 10.)
+#runas import numpy as np ; a = np.sin(np.ones(1000000)) ; b = np.cos(np.ones(1000000)) ; n = np.arange(1000000, dtype=np.double); calculate_vibr_energy(a, b, n)
 import numpy
 def calculate_vibr_energy(harmonic, anharmonic, i):
     return numpy.exp(-harmonic * i - anharmonic * (i ** 2))

### 2014-01-29 doc/papers/iop2014/xp/cython/growcut.pyx
Add experimental setup
Not all tests are shown in the paper, but Pythran performs well on them.

@@ -1,0 +1,72 @@
+# Authors: Stefan van der Walt, Nathan Faggian, Aron Ahmadia
+# https://github.com/stefanv/growcut_py
+from __future__ import division
+ 
+import numpy as np
+cimport cython
+ 
+cdef extern from "math.h" nogil:
+    double sqrt(double)
+ 
+
+@cython.boundscheck(False)
+@cython.wraparound(False)
+cdef inline double distance(double[:, :, ::1] image,
+                            Py_ssize_t r0, Py_ssize_t c0,
+                            Py_ssize_t r1, Py_ssize_t c1) nogil:
+    cdef:
+        double s = 0, d
+        int i
+ 
+    for i in range(3):
+        d = image[r0, c0, i] - image[r1, c1, i]
+        s += d * d
+ 
+    return sqrt(s)
+ 
+
+cdef double s3 = sqrt(3)
+ 
+cdef inline double g(double d) nogil:
+    return 1 - (d / s3)
+ 
+@cython.boundscheck(False)
+@cython.wraparound(False)
+def growcut(double[:, :, ::1] image, double[:, :, ::1] state, double[:, :, ::1] state_next, Py_ssize_t window_radius):
+ 
+    cdef:
+        Py_ssize_t i, j, ii, jj, width, height
+        double gc, attack_strength, defense_strength, winning_colony
+        int changes
+ 
+    height, width = image.shape[0], image.shape[1]
+ 
+    changes = 0
+ 
+    for j in range(width):
+        for i in range(height):
+ 
+            winning_colony = state[i, j, 0]
+            defense_strength = state[i, j, 1]
+ 
+            for jj in xrange(max(0, j - window_radius), min(j + window_radius + 1, width)):
+                for ii in xrange(max(0, i - window_radius), min(i + window_radius + 1, height)):
+                    if ii == i and jj == j:
+                        continue
+ 
+                    # p -> current cell, (i, j)
+                    # q -> attacker, (ii, jj)
+ 
+                    gc = g(distance(image, i, j, ii, jj))
+ 
+                    attack_strength = gc * state[ii, jj, 1]
+ 
+                    if attack_strength > defense_strength:
+                        defense_strength = attack_strength
+                        winning_colony = state[ii, jj, 0]
+                        changes += 1
+ 
+                    state_next[i, j, 0] = winning_colony
+                    state_next[i, j, 1] = defense_strength
+ 
+    return changes

### 2014-01-29 doc/papers/iop2014/xp/cython/hyantes.pyx

@@ -1,0 +1,20 @@
+#pythran export run(float, float, float, float, float, float, int, int, float[][])
+#runas import numpy ; run(0,0,90,90, 1, 100, 80, 80, numpy.array([ [i/10., i/10., i/20.] for i in xrange(80)],dtype=numpy.double))
+import numpy as np
+cimport cython
+@cython.boundscheck(False)
+@cython.wraparound(False)
+def run(double xmin, double ymin, double xmax, double ymax, double step, double range_, int range_x, int range_y, double [:,:] t):
+    cdef:
+        int X, Y, i, j
+        double [:,:] pt
+        double tmp
+    X,Y = t.shape[0], t.shape[1]
+    pt = np.zeros((X,Y))
+    for i in range(X):
+        for j in range(Y):
+            for k in t:
+                tmp = 6368.* np.arccos( np.cos(xmin+step*i)*np.cos( k[0] ) * np.cos((ymin+step*j)-k[1])+  np.sin(xmin+step*i)*np.sin(k[0]))
+                if tmp < range_:
+                    pt[i][j]+=k[2] / (1+tmp)
+    return pt

### 2014-01-29 doc/papers/iop2014/xp/numba/hyantes.py

@@ -1,0 +1,14 @@
+#pythran export run(float, float, float, float, float, float, int, int, float[][])
+#runas import numpy ; run(0,0,90,90, 1, 100, 80, 80, numpy.array([ [i/10., i/10., i/20.] for i in xrange(80)],dtype=numpy.double))
+import numpy as np
+def run(xmin, ymin, xmax, ymax, step, range_, range_x, range_y, t):
+    X,Y = t.shape
+    pt = np.zeros((X,Y))
+    "omp parallel for"
+    for i in range(X):
+        for j in range(Y):
+            for k in t:
+                tmp = 6368.* np.arccos( np.cos(xmin+step*i)*np.cos( k[0] ) * np.cos((ymin+step*j)-k[1])+  np.sin(xmin+step*i)*np.sin(k[0]))
+                if tmp < range_:
+                    pt[i][j]+=k[2] / (1+tmp)
+    return pt

### 2014-01-29 doc/papers/iop2014/xp/numba/rosen_der.py

@@ -1,0 +1,18 @@
+# Authors: Travis E. Oliphant (numpy version), Serge Guelton (python version)
+# License: BSD
+# Source: https://github.com/scipy/scipy/blob/master/scipy/optimize/optimize.py
+#pythran export rosen_der(float[])
+import numpy
+
+
+def rosen_der(x):
+    xm = x[1:-1]
+    xm_m1 = x[:-2]
+    xm_p1 = x[2:]
+    der = numpy.zeros_like(x)
+    der[1:-1] = (+ 200 * (xm - xm_m1 ** 2)
+                 - 400 * (xm_p1 - xm ** 2) * xm
+                 - 2 * (1 - xm))
+    der[0] = -400 * x[0] * (x[1] - x[0] ** 2) - 2 * (1 - x[0])
+    der[-1] = 200 * (x[-1] - x[-2] ** 2)
+    return der

### 2014-01-29 doc/papers/iop2014/xp/run_benchmarks.sh

@@ -1,0 +1,76 @@
+#!/bin/sh
+set -e
+
+CXX=~/sources/llvm/_build/install/bin/clang++
+CXXFLAGS=-O3
+export PATH=../../../../scripts:$PATH
+export PYTHONPATH=../../../..
+
+# first clean all
+find -name '*.pyc' -o -name '*.so' -o -name '*.c' -o -name '*.cpp' | xargs rm -f || true
+
+# then build all cython stuff
+for target in cython/*.pyx
+do
+    cython --cplus $target -o ${target%.pyx}.cpp
+    $CXX $CXXFLAGS `python-config --cflags --libs` -shared ${target%.pyx}.cpp -fPIC -o ${target%.pyx}.so
+done
+
+# then build all pythran stuff
+for target in pythran/*.py
+do
+    pythran $CXXFLAGS $target -I../../../../build/lib.linux-x86_64-2.7/pythran -o ${target%.py}.so
+done
+
+# then run tests
+
+echo "==== arc distance ===="
+
+SETUP='import numpy as np; N=1000000 ; a,b,c,d = np.random.random(N), np.random.random(N), np.random.random(N), np.random.random(N) ; from arc_distance import arc_distance'
+RUN='arc_distance(a,b,c,d)'
+
+printf "python " ;  PYTHONPATH=./python python -m timeit -s "$SETUP" "$RUN"
+printf "cython " ;  PYTHONPATH=./cython python -m timeit -s "$SETUP" "$RUN"
+printf "pythran " ; PYTHONPATH=./pythran python -m timeit -s "$SETUP" "$RUN"
+printf "pypy " ;    PYTHONPATH=./pypy pypy -m timeit -s "from random import random ; import numpypy as np; N=1000000 ; a,b,c,d = np.array([random() for _ in range(N)]), np.array([random() for _ in range(N)]), np.array([random() for _ in range(N)]), np.array([random() for _ in range(N)]) ; from arc_distance import arc_distance" "$RUN"
+
+echo "==== blacksholes ===="
+
+SETUP='N=1000 ; from random import random ; a,b,c,d,e,f = [random() for _ in range(1,N)], [random() for _ in range(1,N)], [random() for _ in range(1,N)], 0.5, 0.76, 200 ; from blacksholes import BlackScholes'
[…]

### 2014-01-29 doc/papers/iop2014/paper.tex
Comment experiments & benchmarks.

@@
 
-The 4 hyper-threaded cores give an extra performance boost. Unfortunately, the
-load is not sufficient to get more than an average 4x speed up compared to the
-vectorized version.
-% You complain about 4x because you only have 3.98x ???
-In the end, Pythran generates a native module that performs
-roughly 11\times faster than the original version.
+The 4 hyper-threaded cores give an extra performance boost. In the end, Pythran
+generates a native module that performs roughly $\times11$ faster than the
+original version.
 
@@
     hits = 0
-    "#omp parallel for reduction(+:hits)"
+    #omp parallel for reduction(+:hits)
     for i in xrange(darts):
@@
 i7, % FIXME sandy-brige, nehalem, what?
-using examples shipped along with Pythran sources, available at
-https://github.com/serge-sans-paille/pythran in the \texttt{pythran/test/cases}
-directory. The Pythran version used is the \texttt{HEAD} of the \texttt{scipy2013} branch,
-ShedSkin 0.9.2, PyPy 2.0 compiled with the \texttt{-jit} flag, CPython 2.7.3, Cython
-0.19.1 and Numexpr 2.0.1. All timings are made using the \texttt{timeit} module,
-taking the best of all runs. All C++ codes are compiled with g++ 4.8.2, using
-the tool default compiler option, generally \texttt{-O2} plus a few optimizing flags
-depending on the target.
-
-Cython is not considered in most benchmarks, because to get an efficient
-binary, one needs to rewrite the original code, while all the considered tools
-are running the very same Python code that remains compatible with CPython. The
-experiment was performed only to obtain a comparison with Numexpr.
-
-Pystone is a Python translation of Whetstone, a famous floating point number
-benchmarks that dates back to Algol60 and the 70's. Although non representative
-of real applications, it illustrates the general performance of floating point
-number manipulations. Table~\ref{tbl:pystone}  illustrates the benchmark
-result for CPython, PyPy, ShedSkin and Pythran, using an input value of
-\texttt{10**3}. Note that the original version has been updated to replace the user
-class by a function call.
+using examples shipped along with
+Pythran sources, available at \url{https://github.com/serge-sans-paille/pythran} in
+the \texttt{pythran/test/cases} directory. The Pythran version used is the
+\texttt{HEAD} of the \texttt{iop2014} branch, PyPy 2.0.2 compiled with the
+\texttt{-jit} flag, CPython 2.7.3 and Cython 0.20. All timings are made using
+the \texttt{timeit} module, taking the best of all runs. All C++ codes are
+compiled with g++ 4.8.2, using the tool default compiler option, generally
+\texttt{-O3} plus a few optimizing flags depending on the target. To make a
+fair comparison, no parallelism annotations are added to any code.
+
+Provided each module has been properly compiled, each experiment is run using
+the following command from a shell:
+
+\begin{lstlisting}[language=sh]
+python -m timeit -s "${SETUP}" "${RUN}"
+\end{lstlisting}
+
+\noindent where \texttt{\$SETUP} and \texttt{\$RUN} depend on the benchmark and
+\texttt{pypy} is substituted to \texttt{python} when needed. The
+\texttt{\$SETUP} and \texttt{\$RUN} values are available to the reader on the
+Git repository in the \texttt{doc/papers/iop/xp/run\_benchamrks.sh} file with
+all the experimental setup.
+
+The same source is used for each benchmark, to the exception of Cython sources
+which needed a manual conversion. For these sources, bound checking and wrap
+around are always disable, all variables are declared using the proper C or C++
+type and all mathematical functions in use come from \texttt{libm}. Some calls
+to the \texttt{zeros\_like} function also have been replaced by calls to
+\texttt{zeros} for the PyPy version.
+
+\subsection{NQueens}
+
+``Nqueens'' is a benchmark extracted from the former Unladen
+Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
+is particularly interesting as it makes an intensive use of non-trivial
+generator expressions and integer sets.
+
+Table~\ref{tbl:nqueen} illustrates the benchmark results for CPython, Cython,
+PyPy and Pythran.  It showcases the fact that Pythran benefits a lot from
+high-level transformations, for instance \emph{lazy evaluation}, which prevents
+the many allocation / deallocation triggered by Cython and PyPy runs.  A deeper
+look at the Pythran profiling trace shows that more than half of the execution
+time is spent allocating and deallocating a \texttt{set} used in the internal
+loop. There is a memory allocation invariant that could be taken advantage of
+there, but none of the compiler does so.
 
@@
         \hline
-     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin  \\
-    \hline
-     Timing  &  861ms      &   11.8ms      &     29.1ms &  24.7ms    \\
-    \hline
-     Speedup &  $\times$1         &   $\times$72.9       &    $\times$29.6   &  $\times$34.8     \\
-    \hline
-\end{tabular}
-
-    \caption{Benchmarking result on the Pystone program.}
-    \label{tbl:pystone}
-
-\end{table}
-
-It comes at no surprise that all tools get more than decent results on this
-benchmark. PyPy generates a code almost as fast % FIXME (dunno if it's efficient) 
-as ShedSkin. Altough both
-generate C++, Pythran outperforms ShedSkin thanks to a higher level generated
-code. For instance all arrays are represented in ShedSkin by pointers to arrays
-that likely disturbs the g++ optimizer, while Pythran uses a vector class wrapping
-shared pointers.
-
-Nqueen is a benchmark extracted from the former Unladen Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
-is particularly interesting as it makes an intensive use of non-trivial
-generator expressions and integer sets. Table~\ref{tbl:nqueen} illustrates
-the benchmark results for CPython, PyPy, ShedSkin and Pythran. The code had to
-be slightly updated to run with ShedSkin because type inference in ShedSkin does
-not support mixed scalar and \textbf{None} variables. The input value is \texttt{9}.
-
-\begin{table}
-    \centering
-
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin \\
-    \hline
-     Timing  &  1904.6ms   &   358.3ms     &    546.1ms &  701.5ms  \\
-    \hline
-     Speedup &  $\times$1         &    $\times$5.31      &    $\times$3.49   &  $\times$2.71    \\
+     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
+    \hline
+    Timing  &  181ms   &   42.3ms     &    \textbf{22.5ms} &  51.3ms  \\
+    \hline
+    Speedup &  $\times$1         &    $\times$4.27      &    \textbf{$\times$8.04}   &  $\times$3.52    \\
     \hline
@@
 
-It seems that compilers have difficulties to take advantage of high level
-constructs such as generator expressions, as the overall speedup is not
-breathtaking. Pythran benefits from the conversion to \texttt{itertools.imap} here,
-while ShedSkin and PyPy rely on more costly constructs. A deeper look at the
-Pythran profiling trace shows that more than half of the execution time is
-spent allocating and deallocating a \texttt{set} used in the internal loop. There is a
-memory allocation invariant that could be taken advantage of there, but none of
-the compiler does so.
-
-Hyantes\footnote{\url{http://hyantes.gforge.inria.fr}} is a geomatic application that exhibits typical usage of arrays
-using loops instead of generalized expressions. It is helpful to measure the
-performance of direct array indexing.
-
-Table~\ref{tbl:hyantes} illustrates the benchmark result for CPython, PyPy,
-ShedSkin and Pythran, when using lists as the data container. The output window
-used is \texttt{100x100}.
+\subsection{Rosen Der}
+
+The ``rosen der'' kernel computes a complex numpy expression that accesses
+data through a one dimensional slice. It is interesting to evaluate how much of
+the expression can be compiled to regular, efficient loop while avoiding
+temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
+CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
+test case, but Cython code had to be manually transformed into an explicit
+loop. Moreover, activating automatic parallelization for Pythran yields an
+additional $\times4$ speedup without a change to the input, while Cython needs
+manual code annotation to achieve a similar result. It seems that PyPy support
+for numpy expression is still note mature for numpy expression, but there is a
+strong ungoing effort of the community to fix this.
 
@@
     \begin{tabular}{|l|c|c|c|c|}
-
-        \hline
-     Tool    &  CPython    &   Pythran     &     PyPy   &  ShedSkin  \\
-    \hline
-     Timing  &  1295.4ms   &   270.5ms     &    277.5ms &  281.5ms   \\
-    \hline
-     Speedup &  x1         &    x4.79      &    x4.67   &  x4.60     \\
+        \hline
+     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
+    \hline
+    Timing  &  22.3ms   &   \textbf{3.39ms}     &    3.81ms &  51.3ms  \\
+    \hline
+    Speedup &  $\times$1         &    \textbf{$\times$6.57}      &    $\times$5.85   &  $\times$0.43    \\
     \hline
 \end{tabular}
-\caption{Benchmarking result on the hyantes kernel, list version.}
-\label{tbl:hyantes}
+\caption{Benchmarking result on the Rosen der program.}
+\label{tbl:rosen}
 
@@
 
-The speed ups are not amazing % FIXME (not formal enough, same for previous usage) 
-for a numerical application. There are two
-reasons for these poor speedups. First, the \texttt{hyantes} benchmark makes heavy
-usage of trigonometric functions, and there is not much to gain there. Second, and
-most important, the benchmark produces a big 2D array stored as a list of list,
-hence the application suffers from the heavy overhead of converting them from C++
-to Python. Running the same benchmark using Numpy arrays as core containers
-confirms this assumption, as illustrated by Table~\ref{tbl:np-hyantes}. This
-table also demonstrates the benefits of manual parallelization using OpenMP.
-% Showing parallelization AND np.array at the same time confuse reader as we
-% don't know how much each optimisation contribute in the speed up
+\subsection{Growcut}
+
+The ``growcut'' application is an image processing code where all array indexing
+are explicit. In that sense, this benchmark is complementary of the previous
+one, as it evaluates how well explicit array access are compiled.
+Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
+speedups on that benchmark, but this is is merely due to the fact that
+accessing individual elements is extremely slow in numpy, and the direct C
+translation of the Python code is very close to what a C developer would write.
+Pythran appears to outperforms Cython once all access are explicit, most
+certainly due to cleaner code generation.
 
@@
     \begin{tabular}{|l|c|c|c|c|}
-
-        \hline
-     Tool    &  CPython    &   Pythran     & Pythran+OpenMP   \\
-    \hline
-     Timing  &  450.0ms    &   4.8ms       &      2.3ms       \\
-    \hline
-     Speedup &  x1         &    x93.8      &    x195.7        \\
+        \hline
+     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
+    \hline
+    Timing  &  398ms   &   991us     &    \textbf{526us} &  5.54ms  \\
+    \hline
+    Speedup &  $\times$1         &    $\times$401.6      &    $\times$756.7   &  $\times$71    \\
     \hline
 \end{tabular}
-\caption{Benchmarking result on the hyantes kernel, numpy version.}
-\label{tbl:np-hyantes}
+\caption{Benchmarking result on the growcut program.}
+\label{tbl:growcut}
+
 \end{table}
 
-Finally, \texttt{arc\_distance} \footnote{The \texttt{arc\_distance} testbed is taken from to \url{https://bitbucket.org/FedericoV/numpy-tip-complex-modeling}.} presents a classical usage of Numpy expression.
-It
-is typically more efficient than its loop alternative as all the iterations are
-done directly in C. Its code is reproduced below:
-
-\begin{lstlisting}
-def arc_distance(theta_1, phi_1, theta_2, phi_2):
-    """
-    Calculates the pairwise arc distance
-    between all points in vector a and b.
-    """
-    temp = (np.sin((theta_2 - theta_1) / 2) ** 2
-        + np.cos(theta_1) * np.cos(theta_2)
-          * np.sin((phi_2 - phi_1) / 2) ** 2)
-    distance_matrix = 2 * np.arctan2(
-            sqrt(temp), sqrt(1 - temp))
-    return distance_matrix
-\end{lstlisting}
-
-Table~\ref{tbl:arc-distance} illustrates the benchmark result for CPython,
-Cython, Numexpr and Pythran, using random input arrays of \texttt{10**6} elements.
-Table~\ref{tbl:arc-distance-2} details the Pythran performance. Cython code
-is written using the \texttt{parallel.prange} feature and compiled with
-\texttt{-fopenmp -O2 -march=native}.
+\subsection{Blacksholes}
+
+The ``blacksholes'' application is a benchmarks that showcases the
+translation cost from the Python space to native space. Indeed we forced the
+use of a \texttt{list} as input of the kernel instead of a numpy array. As a
+consequence there is a costly translation step from a structure that contains
+an array of pointers to generic Python objects to a dense structure involved.
+Results are shown in Table~\ref{tbl:black}. They show that conversion code is
+less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
+it does not suffer from a similar conversion penality.
 
@@
     \centering
+
     \begin{tabular}{|l|c|c|c|c|}
         \hline
-     Tool    &  CPython    &  Cython  &  Numexpr    & Pythran   \\
-    \hline
-     Timing  &  192.2ms    &  36.0ms  &    41.2ms   &  17.1ms   \\
-    \hline
-     Speedup &  x1         &  x5.33   &  x4.67      &  x11.23   \\
+     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
+    \hline
+    Timing  &  490us   &   74.7us     &    \textbf{34.8us} &  37.3us  \\
+    \hline
+    Speedup &  $\times$1         &    $\times$6.55      &    $\times$14.1   &  $\times$13.1    \\
     \hline
 \end{tabular}
-% FIXME (You changed the order of the compilers, even though pythran is still first why?)
-\caption{Benchmarking result on the arc distance kernel.}
-\label{tbl:arc-distance}
+\caption{Benchmarking result on the blacksholes program.}
+\label{tbl:black}
+
 \end{table}
 
-\begin{table}
-    \centering
-    \begin{tabular}{|l|c|c|c|c|}
-    \hline
-    Tool & Pythran (raw) & Pythran (+AVX) & Pythran (+OMP) & Pythran (full)  \\
-    \hline
-    Timing & 186.3ms     &    75.4ms      &    41.1ms      &  17.1ms         \\
-    \hline
-    Speedup&    x1.03      &    x2.54       &    x4.67       &  x11.23         \\
-    \hline
-\end{tabular}
-\caption{Detailed benchmarking result on the arc distance kernel for Pythran.}
-\label{tbl:arc-distance-2}
-\end{table}
-
-It shows a small benefit from using expression templates on their own, most
-% In fact there are small benefits for compiling this code using only pythran without AVX and openMP.
-certainly because the loop control overhead is negligible compared to the
-trigonometric functions. It gets a decent x2.5 speed-up when using AVX over not
-using it. The benefit of OpenMP, although related to the number of cores, makes
-an overall speedup greater than x11 compared to the original Numpy version, 
-without
-changing the input code. Almost on the opposite, Numexpr requires rewriting the 
-input and does not achieve the same level of performance as Pythran when OpenMP
-and AVX are combined.
-
-Writing efficient Cython code requires more work than just typing the variable
-declarations using Cython's specific syntax: it only takes advantage of
-parallelism because we make it explicit. Without explicit parallelization,
-the generated code runs around 176ms instead of 36ms. %FIXME speedup in x?
-Cython does not generate
-vectorized code, and \texttt{gcc} does not vectorize the inner loop, which 
-explains the better result obtained with Pythran.
-
 \section{Future Work}
@@
 Although Pythran focuses on a subset of Python and its standard library, many
-optimizations opportunities are still present. Using a Domain Specific
-Language(DSL) approach, one could use rewriting rules to optimize several
+optimizations opportunities are still available. Using a Domain Specific
+Language (DSL) approach, one could use rewriting rules to optimize several
 Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
@@
 match the performances of statically compiled language. Moreover, one can use
-high level information to generate efficient code that would be difficult to write for the average programmer.
+high level information to generate efficient code that would be difficult to
+write for the average programmer.
 

### 2014-01-29 pythran/backend.py
Improve slicing by making the step field a constant when possible
It appears that slices often have a step of one, which turns them into
contiguous slices. Many optimizations can be performed on contiguous
slices, so take advantage of it!

@@ -270,11 +270,11 @@ class Cxx(Backend):
                     ", ".join(formal_types)) if formal_types else "")
 
             operator_body.append(
-                Statement("{0}: return result_type();".format(
+                Statement("{0}: return;".format(
                     Cxx.final_statement)))
 
             next_declaration = [
-                FunctionDeclaration(Value("result_type", "next"), []),
+                FunctionDeclaration(Value("void", "next"), []),
                 EmptyStatement()]  # empty statement to force a comma ...
 
             # the constructors
@@ -353,10 +353,7 @@ class Cxx(Backend):
                     ]
             next_signature = templatize(
                     FunctionDeclaration(
-                        Value(
-                            "typename {0}::result_type".format(
-                                instanciated_next_name),
-                            "{0}::next".format(instanciated_next_name)),
+                        Value("void", instanciated_next_name + "::next"),
                         []),
                     formal_types)
 
@@ -541,9 +538,7 @@ class Cxx(Backend):
         num, label = self.yields[node]
         return "".join(n for n in Block([
             Assign(Cxx.generator_state_holder, num),
-            ReturnStatement("{0} = {1}".format(
-                Cxx.generator_state_value,
-                self.visit(node.value))),
+            Assign(Cxx.generator_state_value, self.visit(node.value)),
             Statement("{0}:".format(label))
             ]).generate())
 
@@ -947,7 +942,10 @@ class Cxx(Backend):
             arg = (self.visit(nfield) if nfield
                     else 'pythonic::__builtin__::None')
             args.append(arg)
-        return "pythonic::types::slice({},{},{})".format(*args)
+        if node.step is None or type(node.step) is ast.Num and node.step.n == 1:
+            return "pythonic::types::contiguous_slice({},{})".format(*args[0:2])
+        else:
+            return "pythonic::types::slice({},{},{})".format(*args)
 
     def visit_Index(self, node):
         return self.visit(node.value)

### 2014-01-29 pythran/pythonic/types/list.hpp

@@ -26,14 +26,14 @@ namespace pythonic {
         /* forward declaration */
         struct empty_list;
         template<class T> class list;
-        template<class T> class list_view;
+        template<class T, class S> class sliced_list;
 
         /* for type disambiguification */
         struct single_value {};
 
         /* list view */
-        template<class T>
-            class list_view {
+        template<class T, class S=slice>
+            class sliced_list {
 
                 // data holder
                 typedef  typename std::remove_cv< typename std::remove_reference<T>::type>::type  _type;
@@ -43,7 +43,7 @@ namespace pythonic {
                 template<class U>
                     friend class list;
 
-                normalized_slice slicing;
+                typename S::normalized_type slicing;
 
                 public:
                 //  types
@@ -61,15 +61,15 @@ namespace pythonic {
                 typedef typename container_type::const_reverse_iterator const_reverse_iterator;
 
                 // constructor
-                list_view(): data(utils::no_memory()) {}
-                list_view(list_view<T> const & s): data(s.data), slicing(s.slicing) {}
-                list_view(list<T> & other, slice const & s) : data(other.data), slicing(s.normalize(other.size())) {}
+                sliced_list(): data(utils::no_memory()) {}
+                sliced_list(sliced_list<T,S> const & s): data(s.data), slicing(s.slicing) {}
+                sliced_list(list<T> & other, S const & s) : data(other.data), slicing(s.normalize(other.size())) {}
 
                 // assignment
-                list_view& operator=(list<T> const & );
-                list_view& operator=(list_view<T> const & );
+                sliced_list& operator=(list<T> const & );
+                sliced_list& operator=(sliced_list<T,S> const & );
                 list<T> operator+(list<T> const & );
-                list<T> operator+(list_view<T> const & );
+                list<T> operator+(sliced_list<T,S> const & );
 
                 // iterators
                 iterator begin() { assert(slicing.step==1) ; return data->begin()+slicing.lower; }
@@ -104,8 +104,8 @@ namespace pythonic {
                 typedef std::vector< _type > container_type;
                 utils::shared_ref<container_type> data; 
 
-                template<class U>
-                    friend class list_view;
+                template<class U, class S>
+                    friend class sliced_list;
 
                 template<class U>
                     friend class list;
@@ -150,7 +150,8 @@ namespace pythonic {
[…]

### 2014-01-29 pythran/pythonic/types/ndarray.hpp

@@ -699,8 +699,8 @@ we assume that M>=N
 
         /* proxy type to hold the return of a slice
         */
-        template<class T>
-            struct sliced_ndarray : normalized_slice {
+        template<class T, class S=slice>
+            struct sliced_ndarray : S::normalized_type {
                 static constexpr size_t value = T::value;
                 typedef typename T::value_type value_type;
                 typedef typename T::reference reference;
@@ -712,31 +712,36 @@ we assume that M>=N
 
                 sliced_ndarray() {}
 
-                sliced_ndarray(T const& data, slice const& s) : normalized_slice(s.normalize(data.shape[0])), shape(data.shape), data(data) {
-                    shape[0] = normalized_slice::size();
-                    jump = 1;
+                sliced_ndarray(T const& data, S const& s) : S::normalized_type(s.normalize(data.shape[0])), shape(data.shape), data(data) {
+                    shape[0] = S::normalized_type::size();
+                    jump = S::normalized_type::lower;
                     for(size_t i=1;i<value; ++i)
                         jump*=shape[i];
                 }
 
+#ifdef USE_BOOST_SIMD
+                auto load(long i) const -> decltype(data.load(i)) {
+                    return data.load(jump + i); // FIXME
+                }
+#endif
                 auto at(long i) const -> decltype(data.at(0)) {
-                    return data.at(jump*lower+i*step);
+                    return data.at(jump + i * S::normalized_type::step);
                 }
                 long size() const { return (data.size() / data.shape[0]) * shape[0] ; }
-                reference operator[](long i) { return data[jump*lower+i*step]; }
-                const_reference operator[](long i) const { return data[jump*lower+i*step]; }
+                reference operator[](long i) { return data[jump+i*S::normalized_type::step]; }
+                const_reference operator[](long i) const { return data[jump+i*S::normalized_type::step]; }
 
-                sliced_ndarray<T> operator[](slice const& s) const {
-                    normalized_slice norm = s.normalize(normalized_slice::size());
-                    return sliced_ndarray(data, slice(lower + step*norm.lower, lower + step*norm.upper, step*norm.step));
+                sliced_ndarray<T,S> operator[](slice const& s) const {
+                    S norm = s.normalize(S::normalized_type::size());
[…]

### 2014-01-29 pythran/pythonic/types/slice.hpp

@@ -23,6 +23,7 @@ namespace pythonic {
         };
 
         struct slice {
+            typedef normalized_slice normalized_type;
             none<long> lower, upper, step;
             slice(none<long> lower, none<long> upper, none<long> step = 1)
                 : lower(lower), upper(upper), step(step) {}
@@ -57,6 +58,55 @@ namespace pythonic {
             {
                 assert( not upper.is_none and not lower.is_none and not step.is_none );
                 return std::max(0L, long(ceil(double(upper - lower)/double(step))));
+            }
+        };
+
+        struct contiguous_normalized_slice {
+            long lower, upper;
+            static const long step = 1;
+            contiguous_normalized_slice() {}
+            contiguous_normalized_slice(long lower, long upper)
+                : lower(lower), upper(upper) {}
+
+            long size() const
+            {
+                return std::max(0L, long(ceil(double(upper - lower))));
+            }
+        };
+
+        struct contiguous_slice {
+            typedef contiguous_normalized_slice normalized_type;
+            none<long> lower, upper;
+            static const long step = 1;
+            contiguous_slice(none<long> lower, none<long> upper)
+                : lower(lower), upper(upper) {}
+            contiguous_slice(){}
+
+            /*
+               Normalize change a[:-1] to a[:len(a)-1] to have positif index.
+               It also check for value bigger than len(a) to fit the size of the container
+               */
+            contiguous_normalized_slice normalize(long max_size) const
+            {
+                long normalized_upper;
+                if(upper.is_none) normalized_upper = max_size;
+                else if(upper<0L) normalized_upper = std::max(-1L, max_size + upper);
+                else if(upper> max_size) normalized_upper = max_size;
+                else normalized_upper = (long)upper;
+
+                long normalized_lower = (long)lower;
+                if(lower.is_none) normalized_lower = 0L;
+                else if(lower <0L) normalized_lower = std::max(0L, max_size + lower);
+                else if(lower > max_size) normalized_lower = max_size - 1L;
+                else normalized_lower = (long)lower;
+
+                return contiguous_normalized_slice(normalized_lower, normalized_upper);
+            }
+
+            long size() const
+            {
+                assert(not upper.is_none and not lower.is_none);
+                return std::max(0L, long(ceil(double(upper - lower))));
             }
         };
     }

### 2014-01-29 doc/papers/iop2014/paper.tex
Take into account other reviews.

@@
            {pierrick.brunet@inria.fr}
-\authorinfo{Alan Raynaud}
-           {T{\'e}l{\'e}com Bretagne}
-           {alan.raynaud@telecom-bretagne.eu}
 \authorinfo{Mehdi Amini}
@@
            {adrien.merlini@telecom-bretagne.eu}
-\authorinfo{Eliott Coyac}
+\authorinfo{Xavier Corbillon}
            {T{\'e}l{\'e}com Bretagne}
-           {eliott.coyac@telecom-bretagne.eu}
+           {xavier.corbillon@telecom-bretagne.eu}
+\authorinfo{Alan Raynaud}
+           {T{\'e}l{\'e}com Bretagne}
+           {alan.raynaud@telecom-bretagne.eu}
 
@@
     features such as variadic templates, type inference, move semantics and
-    perfect forwarding, as well as classical ones such as expression templates.
-    % FIXME Expression template is not a "feature" of the language IMO
-    % It is more a classical "use" of the language
+    perfect forwarding, as well as classical idioms such as expression templates.
 
     % FIXME  (explain that annotations can be added to the code?)
-    The input code remains compatible with the Python interpreter, and output
-    code is generally as efficient as the annotated Cython equivalent, if not
-    more, % FIXME (cut sentence) 
-    without the backward compatibility loss of Cython. Numpy expressions
-    run faster than when compiled with numexpr or numba, without any change to
-    the original code. 
-    % It may be a little too ... untactful?
+    Unlike the Cython approach , Pythran input code remains compatible with the
+    Python interpreter. Output code is generally as efficient as the annotated
+    Cython equivalent, if not more, but without the backward compatibility
+    loss.
 
@@
 compared to the same algorithm written in a statically compiled language such
-as C is high, % FIXME cut sentence
-due to numerous dynamic lookup and interpretation cost inherent
-in high level languages. Additionally, the Python compiler does not perform
-optimization on the byte code, while scientific applications are first-class
-candidates for many of them.
+as C is high. % FIXME cut sentence It is partly due to the interpration cost,
+but also to Python's dynamic binding. Additionally, the Python compiler does
+not perform any optimization on the byte code, while scientific applications
+are first-class candidates for many of them.
 
@@
 10\% of the code, it is natural to focus on those computation-intensive pieces 
-of code. The aim % FIXME (of what) 
+of code. The goal of an optimizing compiler % FIXME (of what) 
 may not be to optimize the full Python application, but rather a subset of the application.
@@
 
-One approach, used by Cython \cite{cython2010} is to suppress the interpretation
+One approach, used by Cython~\cite{cython2010} is to suppress the interpretation
 overhead by translating Python Programs to C programs calling the Python C
-API\cite{pythoncapi}. More recently, Nuitka\cite{nuitka}  has taken the same
+API~\cite{pythoncapi}. More recently, Nuitka~\cite{nuitka}  has taken the same
 approach using C++ as a back-end. Going a step further Cython also provides an
@@
 relies both on the Python C API and on plain C.
-ShedSkin\cite{shedskin2006} translates implicitly strongly typed Python programs
+ShedSkin~\cite{shedskin2006} translates implicitly strongly typed Python programs
 into C++, without any call to the Python C API, but fails to compile dynamic
@@
 interpreter, to dynamically turn the computation intensive parts into native
-code. This is what the \texttt{numexpr} module\cite{numexpr} does for Numpy 
+code. This is what the \texttt{numexpr} module~\cite{numexpr} does for Numpy 
 expressions, JIT-compiling them from a string representation to native code.
-Numba\cite{numba} and Parakeet\cite{parakeet2012} extend this approach to
-Numpy-centric applications while PyPy\cite{pypy2009} applies it to the whole
+Numba~\cite{numba} and Parakeet~\cite{parakeet2012} extend this approach to
+Numpy-centric applications while PyPy~\cite{pypy2009} applies it to the whole
 language - including the Numpy module through the in-progress \texttt{numpypy}
@@
 
-Except for PyPy, none of these compilers apply any of the
-static optimization techniques that have been known and successfully applied 
+Except for PyPy, none of these compilers apply any of the static optimization
+techniques that have been known and successfully applied 
 % FIXME And Numba? It generates LLVM, which uses static optimization techniques
-for decades to statically compiled language such as C or C++.
-Translators to statically compiled languages do take advantage of them
-indirectly, but the quality of the generated code may prevent advanced
-optimizations, such as vectorization, %FIXME too long
-while they are available at higher level,
-i.e. at the Python level. Taking into account the specificities of the Python
-language can unlock many new transformations. %FIXME sentence without subject
-For instance, PyPy automatically
-converts the \texttt{range} builtin into \texttt{xrange} by using a dedicated
-structure called \texttt{range-list}.
+for decades to statically compiled language such as C or C++.  Translators to
+statically compiled languages do take advantage of them indirectly, but the
+quality of the generated code may prevent advanced optimizations. For instance
+GCC hardly vectorizes the output of Cython, while the array semantic is
+avaialable at the Python level. Taking into account the specificities of the
+Python language can unlock many new transformations.  For instance, PyPy
+automatically converts the \texttt{range} builtin into \texttt{xrange} by using
+a dedicated structure called \texttt{range-list}.
 
@@
 Pythran. The module is then translated to polymorphic C++ code through the
-generation of templated classes in place of Python functions. These
-classes are either used to generate regular C++ % FIXME (not clear what do you mean?) PB : Better?
-code or instantiated for given types to generate a native Python module.
+generation of templated classes in place of Python functions. These classes are
+either used from regular C++ applications, or instantiated for given types to
+generate a native Python module.
 
@@
         functions into global functions taking the closure elements as
-        parameters, and the initial definition is replaced by a \emph{partial
+        parameters. The initial definition is replaced by a \emph{partial
         function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;
-        % FIXME (this sentence makes my head hurt :s maybe cut into smaller sentences)
 
@@
 
-    \item \texttt{try \dots finally} constructs are turned into nested \texttt{try \dots except} blocks;
-        % FIXME (Is the difference between construct and block obvious for intended readers?)
+    \item \texttt{try \dots finally} instriuction are turned into nested \texttt{try \dots except} instructions;
 
@@
 time, but it cannot be assigned to a string if it has already been assigned to
-an integer. % FIXME (not sure if clear enough) 
-As shown later, it is possible to detect several false variable
+an integer.  As shown later, it is possible to detect several false variable
 polymorphism cases using use-def chains. Function polymorphism is achieved
 through template parameters: a template function can be applied to several
-types as long as an implicit structural typing is respected, which is very
-similar to Python's duck typing, except that it is checked at compile time, as
-% FIXME (tooooooooooooooooooooooo long) 
-illustrated by the following implementation of a generic dot product in Python:
+types as long as an implicit structural typing is respected. This technique is
+very similar to Python's duck typing, except that it is checked at compile
+time, as illustrated by the following implementation of a generic dot product
+in Python:
 
@@
 
-Although far more verbose % FIXME  (don't think you can use verbose this way)
-than the Python version, the C++ version also uses some
-form of structural typing: the only assumptions these two versions make are that
-\texttt{L0} and \texttt{L1} are iterable, their content can be multiplied and 
-that the result of the multiplication is accumulatable.
+Although more verbose than the Python version, the C++ version also uses some
+form of structural typing: the only assumptions these two versions make are
+that \texttt{L0} and \texttt{L1} are iterable, their content can be multiplied
+and that the result of the multiplication is accumulatable.
 
@@
 the module, using user-provided signatures. Template instantiation then triggers
-the different correctly typed instantiations for all functions written in the
-module. % FIXME (not only the functions called by top-level functs????)
+the different correctly typed instantiations for all functions called in the
 module. Note that the user only needs to provide the type of the functions
@@
 notably using expression template, are delegated to this library. Pythran relies
-on the C++11\cite{isocxx11} language and leverage recent features such as
+on the C++11~\cite{isocxx11} language and leverage recent features such as
 move semantics, type inference through \texttt{decltype(\dots)} and variadic
@@
 As a consequence it requires a C++11 compatible compiler for the native code
-generation. Boost.Python\cite{boostpython2007} is used for the automatic
+generation. Boost.Python~\cite{boostpython2007} is used for the automatic
 Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
@@
 
-When turning Python AST into Pythran IR, nested functions are turned into global
-functions taking their expanded closure as parameter. This closure is computed 
-using the information provided by the \texttt{Globals} analysis which 
-statically computes the state of the dictionary of globals, and 
-\texttt{ImportedIds} that computes the set of % FIXME (too long)
-identifiers used by an instruction but not declared in this instruction. For
-instance in the following snippet:
+When turning Python AST into Pythran IR, nested functions are turned into
+global functions taking their expanded closure as parameter. This closure is
+computed using the information provided by the \texttt{Globals} and
+\texttt{ImportedIds} analyse. The former statically computes the state of the
+dictionary of globals, and the latter computes the set of identifiers used by
+an instruction but not declared in this instruction. For instance in the
+following snippet:
 
@@
 natural choice to represent Python's \texttt{long} in \cpp. NT2 provides a generic
-vector library called \texttt{boost.simd}\cite{esterie2012boost} that enables
+vector library called \texttt{boost.simd}~\cite{esterie2012boost} that enables
 the use of vector instruction units of modern processors in a generic way. It is used
@@
 
-    \item Expression templates\cite{expression_templates} are used to avoid
+    \item Expression templates~\cite{expression_templates} are used to avoid
         multiple iterations and the creation of intermediate arrays. Because
@@
         % what about binary func?
+        % ufunc -> micro function, in numpy ;-)
 
@@
 
-    \item Loop parallelization through OpenMP\cite{openmp3.1}. Numpy expression
+    \item Loop parallelization through OpenMP~\cite{openmp3.1}. Numpy expression
         computation do not carry any loop-dependency. They are perfect
@@
 
-The OpenMP standard\cite{openmp3.1} is a widely used solution for Fortran, C
+The OpenMP standard~\cite{openmp3.1} is a widely used solution for Fortran, C
 and C++ to describe loop-based and task-based parallelism. It consists of a few
@@
 
-The following listing gives a simple example of explicit loop-based 
-parallelism. OpenMP 3.0 task-based parallelism form is also supported, the
-interested reader may refer to~\cite{pyhpc2013} for more details.
-
-\begin{lstlisting}
-def pi_estimate(darts):
-    hits = 0
-    #omp parallel for reduction(+:hits)
-    for i in xrange(darts):
-        x,y = random(), random()
-        dist = sqrt(pow(x, 2) + pow(y, 2))
-        if dist <= 1.0:
-            hits += 1.0
-    pi = 4 * (hits / DARTS)
-    return pi
-\end{lstlisting}
-% why #? I thought string was enough?
-
-The loop is flagged as parallel, performing a reduction using the \texttt{+}
-operator on the \texttt{hits} variable. Variables \texttt{x}, \texttt{y} and
-\texttt{dist} are automatically made local by Pythran after a
-\texttt{ScopeAnalysis}, and are thus marked as \texttt{private}, \ie{local to
-a thread and not shared with other threads, by OpenMP}.
-
-% FIXME (Ca semble tomber a plat, pas de resultats, pas de conclusion...)
+The following listing gives a simple example of explicit loop-based parallelism
+(OpenMP 3.0 task-based parallelism form is also supported, the interested
+reader may refer to~\cite{pyhpc2013} for more details). the \texttt{kernel}
+function is pure and computation intensive, making the loop a perfect candidate
+for parallelization.
+
+\begin{lstlisting}
+def compute_julia(cr, ci, N, bound=1.5, lim=1000., cutoff=1e6):
+    julia = np.empty((N, N), np.uint32)
+    grid_x = np.linspace(-bound, bound, N)
+    #omp parallel for 
+    for i, x in enumerate(grid_x):
+        for j, y in enumerate(grid_x):
+            julia[i,j] = kernel(x, y, cr, ci, lim, cutoff)
+    return julia
+\end{lstlisting}
+
+As in in standard OpenMP, loop indices of the parallel loop are automatically
+made private, and Pythran extends the concept to any random access iterator,
+even after type destructuring. In the example, both \texttt{i} and \texttt{x}
+are private. A scope analysis is used to prove that \texttt{j} and \texttt{y}
+are also local to the loop. In the end, the whole computation achieves a
+perfect speedup of $\times4$
 
 Next section performs an in-depth comparison of Pythran with others Python
-optimizers: Numba, PyPy, ShedSkin and numexpr.
+optimizers: Cython and PyPy.
 
@@
 for numpy expression is still note mature for numpy expression, but there is a
-strong ungoing effort of the community to fix this.
+strong on-going effort of the community to fix this.
 
@@
 less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
-it does not suffer from a similar conversion penality.
+it does not suffer from a similar conversion penalty.
 
@@
 the optimizations used. It also compares the performance of compiled Pythran
-modules against CPython and other optimizers: ShedSkin, PyPy and numexpr.
+modules against CPython and other optimizers like Cython and PyPy.
 
 To conclude, limiting Python to a statically typed subset does not hinders the
-expressivity when it comes to scientific or mathematic computations, but makes
+expressiveness when it comes to scientific or mathematic computations, but makes
 it possible to use a wide variety of classical optimizations to help Python
@@
 This project has been partially funded by the CARP Project and the SILKAN
-Company.
+Company.  The authors would like to thank Yuancheng Peng and Eliott Coyac for
+their contributions to development of the Pythran compiler.
 

### 2014-01-30 doc/papers/iop2014/paper.tex
update paper

@@
     % FIXME  (explain that annotations can be added to the code?)
-    Unlike the Cython approach , Pythran input code remains compatible with the
-    Python interpreter. Output code is generally as efficient as the annotated
-    Cython equivalent, if not more, but without the backward compatibility
-    loss.
+    Unlike the Cython approach, code fed to Pythran remains compatible with the
+    Python interpreter. The result is generally as efficient as Cython with
+    the equivalent annotated Cython code, if not more, 
+    but without the backward compatibility loss.
 
@@
 compared to the same algorithm written in a statically compiled language such
-as C is high. % FIXME cut sentence It is partly due to the interpration cost,
-but also to Python's dynamic binding. Additionally, the Python compiler does
+as C is high. % FIXME cut sentence 
+It is partly due to the interpration cost,
+but also because of Python's dynamic binding. Additionally, the Python compiler does
 not perform any optimization on the byte code, while scientific applications
@@
 10\% of the code, it is natural to focus on those computation-intensive pieces 
-of code. The goal of an optimizing compiler % FIXME (of what) 
+of code. The goal of a specialized compiler % FIXME (of what) 
 may not be to optimize the full Python application, but rather a subset of the application.
@@
 % FIXME And Numba? It generates LLVM, which uses static optimization techniques
-for decades to statically compiled language such as C or C++.  Translators to
+for decades to statically compiled languages such as C or C++.  Translators to
 statically compiled languages do take advantage of them indirectly, but the
@@
 to others to form a simpler and easier to deal with AST for further analyses and
-optimizations. The transformations applied by Pythran on Python AST are the
+optimizations. The transformations applied by Pythran on the Python AST are the
 following:
@@
 
-    \item all imports are fully expanded to make functions and global variables access paths explicit,
+    \item all imports are fully expanded to make the access paths of functions and global variables explicit,
 
@@
 
-    \item compiling the generated C++ code into native code.
+    \item compiling the generated C++ code into native code.%FIXME maybe native module?
 
@@
 from the Python world to C++. Pythran only supports polymorphic variables for
-functions, i.e. a variable can hold several function objects during its life
-time, but it cannot be assigned to a string if it has already been assigned to
+functions, i.e. a variable can hold several function objects during its lifetime, 
+but it cannot be assigned to a string if it has already been assigned to
 an integer.  As shown later, it is possible to detect several false variable
@@
 exported outside the module. The possible types of all internal functions are
-then inferred from the call sites. Note that this complex step leverage the 
+then inferred from the call sites. Note that this complex step leverages the 
 C++11 compiler capabilities: Pythran generates only one line using user-provided
@@
 Without this kind of information, even a simple instruction like
-\texttt{sum(a)} would yield very few information as there is no guarantee that
+\texttt{sum(a)} would yield very little information as there is no guarantee that
 the \texttt{sum} identifiers point to the \texttt{sum} built-in, as it could have been 
@@
 computed using the information provided by the \texttt{Globals} and
-\texttt{ImportedIds} analyse. The former statically computes the state of the
+\texttt{ImportedIds} analyses. The former statically computes the state of the
 dictionary of globals, and the latter computes the set of identifiers used by
@@
 for the second because \texttt{print b} involves an \textbf{I/O}.
-% Which is not really usefull as there is no way to say pythran we want to execut it in parallel
+% Which is not really usefull as there is no way to say pythran we want to execute it in parallel
 % (but logging information is still displayed)
@@
 
-However, there are some information available at the Python level that cannot
-be recovered at the C++ level. For instance, Pythran uses functor with an
+However, there is some information available at the Python level that cannot
+be recovered at the C++ level. For instance, Pythran uses functors with an
 internal state and a goto dispatch table to represent generators. This approach
@@
 i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
-% PB : Not sure it is needed as audiance don't really care about it.
+% PB : Not sure it is needed as audience doesn't really care about it.
 
@@
 be able to take advantage of them.
-%Does python programmer really use it? Or people using it may write their native code?
+%Do python programmers really use it? Or people using it may write their native code?
 % They may write it using python only to prototyping?
@@
 loop. There is a memory allocation invariant that could be taken advantage of
-there, but none of the compiler does so.
+there, but none of the compilers do so.
 
@@
 manual code annotation to achieve a similar result. It seems that PyPy support
-for numpy expression is still note mature for numpy expression, but there is a
-strong on-going effort of the community to fix this.
+for numpy expressions is still not mature, but there is a
+strong ongoing effort of the community to fix this.
 
@@
 The ``growcut'' application is an image processing code where all array indexing
-are explicit. In that sense, this benchmark is complementary of the previous
+is explicit. In that sense, this benchmark complements the previous
 one, as it evaluates how well explicit array access are compiled.
-Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
+Table~\ref{tbl:growcut} summarizes the results. Every compiler yield impressive
 speedups on that benchmark, but this is is merely due to the fact that
@@
 
-The ``blacksholes'' application is a benchmarks that showcases the
+The ``blacksholes'' application is a benchmark that showcases the
 translation cost from the Python space to native space. Indeed we forced the
@@
 consequence there is a costly translation step from a structure that contains
-an array of pointers to generic Python objects to a dense structure involved.
-Results are shown in Table~\ref{tbl:black}. They show that conversion code is
+an array of generic Python objects pointers to a dense structure.
+Results are shown in Table~\ref{tbl:black}. They show that that conversion is
 less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
-it does not suffer from a similar conversion penalty.
+it does not suffer from that problem.
 
@@
 
-To conclude, limiting Python to a statically typed subset does not hinders the
+To conclude, limiting Python to a statically typed subset does not hinder the
 expressiveness when it comes to scientific or mathematic computations, but makes
@@
 Company.  The authors would like to thank Yuancheng Peng and Eliott Coyac for
-their contributions to development of the Pythran compiler.
+their contributions to the development of the Pythran compiler.
 

### 2014-05-04 doc/papers/pydata2014/talk.tex
PyData talk, pre-release

%\documentclass{beamer} % use draft mode for speed !
\documentclass[aspectratio=1610]{beamer}
\usepackage{pacman}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amsfonts, wasysym}
\usepackage{xcolor}
\usepackage{listings}
\definecolor{darkgreen}{rgb}{.1,.7,.1}
\definecolor{darkred}{rgb}{.7,.1,.1}
\definecolor{darkblue}{rgb}{.1,.1,.7}
\lstset{language=python, basicstyle=\ttfamily,moredelim=[is][\textcolor{darkblue}]{@}{@},keywordstyle=\textcolor{darkblue},commentstyle=\textcolor{darkgreen},stringstyle=\textcolor{darkred},
        backgroundcolor=\color[gray]{0.9},
                frame=single
}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\usetikzlibrary{trees}
\usetikzlibrary{shadows}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}

\author[Serge Guelton, Pierrick Brunet, Mehdi Amini]{Serge Guelton \& Pierrick Brunet \& Mehdi Amini}

\title[\texttt{pythran}]{\texttt{Pythran} -- C++ for Snakes\\\emph{\small Static Compiler for High Performance}}

\date{PyData - May 4th 2014}

\institute{}

\begin{document}

\begin{frame}
  \maketitle

  \begin{center}
    \includegraphics[height=3em]{logo}
  \end{center}
  \vfill
  \hfill
  \scriptsize{Get the slides: http://goo.gl/6dgra0}
\end{frame}

% \begin{frame}[fragile]
%   \frametitle{Warning}
% 
% \begin{lstlisting}[language=python]
% from audience import you
% l=(Python, CXX, Compilation, Parallelism)
% skills=(x in you.__class__.__bases__
%     for x in l)
% if not any( skills ):
%     print("That's gonna be tough")
% \end{lstlisting}
% 
% \end{frame}

\begin{frame}[fragile]
    \frametitle{Disclaimer}

All timings were performed using OSX 10.9, an i7-3740QM CPU @ 2.70GHz, Python 2.7.6, current Pythran (c7de8b9c), pypy 2.2.1, numba 0.13.1, gcc 4.8, Clang r207887.
\vfill
I am \textbf{not} Pythonista, but I'm interested in performance in general. 
Daily job: driver-level C code, assembly, multi-threaded C++, GPU, ...
\vfill
\uncover<2->{
\begin{minipage}{0.4\textwidth}
\textit{Note: I love to troll, so let's have a beer later and talk
about how awful Python is! ;-)}
\end{minipage}
\hspace{0.09\textwidth}
\begin{minipage}{0.4\textwidth}
\uncover<3->{
    \includegraphics[width=0.9\textwidth]{compiling.png}
}
\end{minipage}
}
\vfill
\uncover<4->{
  By the way this talk is written in Latex and takes more than 10 seconds to
  \textbf{compile}. 
}
\end{frame}

\begin{frame}{Prototyping Tools for Scientific Computing}
    \begin{tikzpicture}
        \node at (0,0) {\includegraphics[height=4em]{matlab}};
        \only<1>{
            \node at (3,4) {\includegraphics[height=4em]{python}};
        }
        \only<2>{
            \node[draw=darkblue, ultra thick,rounded corners] at (3,4) {\includegraphics[height=4em]{python}};
        }
        \node at (9,3) {\includegraphics[height=4em]{octave}};
        \node at (10,-1) {\includegraphics[height=4em]{scilab}};
        \node at (6,2) {\includegraphics[height=2em]{idl}};
    \end{tikzpicture}
\end{frame}

\begin{frame}{Tools for Scientific Computing in Python}
    \begin{tikzpicture}
        \node at (0,0) {\includegraphics[height=4em]{pypy}};
        \node at (5,2) {\includegraphics[height=4em]{numba}};
        \node at (10,5) {\includegraphics[height=4em]{cython}};
        \node at (-1,6) {\includegraphics[height=4em]{scipy_tool}};
        \node at (3,4.5) {\includegraphics[height=2em]{theano}};
        \node at (.5,3) {\texttt{FORTRAN + f2py}};
        \node at (9.5,-.5) {\texttt{C + SWIG}};
    \end{tikzpicture}
\end{frame}

\begin{frame}{I Do not Know Much About Python But it Does not Matter Because...}

  \begin{center}
    \includegraphics[width=0.6\textwidth]{hotspot}
  \end{center}

  I only care about the white spot.
\end{frame}

\begin{frame}[fragile]
    \frametitle{Regular IPython Session}
\begin{lstlisting}
>>> import numpy as np
>>> def rosen(x):
...    return sum(100.*(x[1:]-x[:-1]**2.)**2.
...         + (1-x[:-1])**2.)
>>> import numpy as np
>>> r = np.random.rand(100000)
>>> %timeit rosen(r)
10 loops, best of 3: @35.1 ms per loop@
\end{lstlisting}

\begin{quote}
In mathematical optimization, the Rosenbrock function is a non-convex function used as a performance test problem for optimization algorithms introduced by Howard H. Rosenbrock in 1960.[1] It is also known as Rosenbrock's valley or Rosenbrock's banana function. (Wikipedia)
\end{quote}

\end{frame}

\begin{frame}[fragile]
    \frametitle{IPython Session with Pythran}
\begin{lstlisting}
>>> %load_ext pythranmagic
>>> %%pythran
import numpy as np
#pythran export rosen(float[])
def rosen(x):
     return sum(100.*(x[1:]-x[:-1]**2.)**2.
        + (1-x[:-1])**2.)
>>> import numpy as np
>>> r = np.random.rand(100000)
>>> %timeit rosen(r)
10000 loops, best of 3: @121 us per loop@
\end{lstlisting}
\vfill
\uncover<2>{\centering{\Large That's a \alert{$\times$290} speedup!}}
\vfill

\end{frame}

\begin{frame}{Pythran's Meta Program Generator}
    \begin{center}
    \scalebox{1.1}{
    \begin{tikzpicture}[
            file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
            node distance=0.6cm},
                tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
            distance=0.6cm}]
            \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
            \node[file] (omp) [left=of python] {\textbf{OpenMP}};
            \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
            \node[file] (cxx) [below=of pythran] {\textbf{C++ meta-program}};
            \node[file] (boost) [left=of cxx] {\textbf{\texttt{boost::python}}};
            \node[file] (pythonic) [right=of cxx] {\textbf{\texttt{pythonic++}}};
            \node[tool] (gxx) [below=of cxx] {\textbf{g++}};
            \node[file] (ti) [left=of gxx, xshift=-1cm] {\textbf{\texttt{Type Info}}};
            \node[file] (nt2) [right=of gxx, xshift=1cm] {\textbf{\texttt{NT}$^2$}};
            \node[file] (so) [below=of gxx] {\textbf{Native Module [\texttt{.so}]}};

            \draw[very thick, ->] (python) -- (pythran);
            \draw[very thick,dashed, ->] (omp) -- (python);
            \draw[very thick, ->] (pythran) -- (cxx);
            \draw[very thick, ->] (cxx) -- (gxx);
            \draw[very thick, ->] (boost) -- (gxx);
            \draw[very thick, ->] (pythonic) -- (gxx);
            \draw[very thick, ->] (ti) -- (gxx);
            \draw[very thick, ->] (nt2) -- (gxx);
            \draw[very thick, ->] (gxx) -- (so);
        \end{tikzpicture}
    }
\end{center}
\end{frame}

\begin{frame}{Pythran Moto}
    \begin{block}{Goals}
        \begin{enumerate}
                \centering{
                \item \structure{Performance First} \\\emph{sacrifice language feature}
                \item \structure{Subset of Python} \\\emph{backward compatibility matters} \\
                pythran $\approx$ python - useless stuff
                \\ \scriptsize\textit{(i.e. class, eval, introspection, polymorphic variable)} \normalsize
                \item \structure{Modular Compilation} \\\emph{focus on numerical kernels}
                }
        \end{enumerate}
    \end{block}

    \begin{block}{Means}
        \begin{enumerate}
                \centering{
                \item Static Compilation
                \item Static Optimizations
                \item Vectorization \& Parallelization
                }
        \end{enumerate}
    \end{block}
\end{frame}

\begin{frame}[fragile]
    \frametitle{A Story of Python \& C++}
    \begin{lstlisting}
def dot(l0, l1):
        return sum(x*y for x,y in zip(l0,l1))
        \end{lstlisting}
        \vspace{-1em}
        \begin{center}
            \Large$\simeq$
        \end{center}
        \vspace{-1em}
    \begin{lstlisting}[language=c++]
template<class T0, class T1>
auto dot(T0&& l0, T1&& l1)
-> decltype(/* ... */)
{
  return pythonic::sum(
    pythonic::map(
      operator_::multiply(),
        pythonic::zip(std::forward<T0>(l0),
                      std::forward<T1>(l1))
    ) );
}
        \end{lstlisting}
\end{frame}

\begin{frame}{C++ as a Back-end}
    \begin{block}{A High Level Language}
        \begin{itemize}
            \item Rich \structure{S}tandard \structure{T}emplate \structure{L}ibrary
            \item Object Oriented Programming
            \item Meta-Programming
            \item Glimpses of type inference\hfill\alert{\small C++11}
            \item Variadic Templates\hfill\alert{\small C++11}
        \end{itemize}
    \end{block}
    \begin{block}{A Low Level Language}
        \begin{itemize}
            \item Few Hidden Costs: "you don't pay for what you don't use"
            \item Direct access to vector instruction units \hfill\alert{SSE, AVX, \dots}
            \item Good for Parallel Programming \hfill\alert{OpenMP, TBB, \dots} \\ \hfill\scriptsize{(and the parallel STL is proposed for C++17)}
        \end{itemize}
    \end{block}
\end{frame}

\begin{frame}[fragile]
  \frametitle{Let's Dive Into the Backend Runtime... \uncover<2->{I'm Kidding!}}
    \begin{tikzpicture}
        \node at (0,0) {
  \begin{lstlisting}[language=c++,basicstyle=\ttfamily\scriptsize]
template <class Op, class S0, class... Iters>
  auto map_aux (Op &op, S0 const &seq, Iters... iters)
    -> sequence <decltype(op(*seq.begin(), *iters...))>
  {
    decltype(_map(op, seq, iters...)) s;
    auto iter = std::back_inserter(s);
    for(auto& iseq : seq)
      *iter ++= op(iseq , *iters++...);
    return s;
  }
template <class Op, class S0, class... SN>
  auto map ( Op op, S0 const &seq, SN const &... seqs )
    -> decltype(_map( op, seq, seqs.begin ()...)) 
  {
    return _map(op, seq, seqs.begin()...);
  }
\end{lstlisting}
        };
        \uncover<2->{
            \node at (6.5,0) {
                \includegraphics[height=8em]{scream}
            };
        }
    \end{tikzpicture}
\end{frame}

\begin{frame}[fragile]
    \frametitle{Static Compilation}
    \begin{block}{Buys time for many time-consuming analyse}
        Points-to, used-def, variable scope, memory effects, function purity\dots
    \end{block}
    \begin{block}{Unleashes powerful C++ optimizations}
        Lazy Evaluation, Map Parallelizations, Constant Folding
    \end{block}
    \begin{block}{Requires static type inference}
        \begin{lstlisting}[lang=python]
        #pythran export foo(int list, float)
        \end{lstlisting}
        \centering Only annotate exported functions!
    \end{block}
\end{frame}

\begin{frame}{Pythran's Moto \textbf{1/3}}
\Large\centering Gather as many information as possible\\
\vfill
\normalsize(Typing is just one information among others)\\
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Analysis: Points-to}
    \begin{lstlisting}[lang=python]
def foo(a,b):
    c = a or b
    return c*2
    \end{lstlisting}
    \vfill
    \centering\Large Where does \lstinline|c| points to ?
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Analysis: Argument Effects}
    \begin{lstlisting}[lang=python]
def fib(n):
    return n if n<2 else fib(n-1) + fib(n-2)

def bar(l):
    return map(fib, l)

def foo(l):
    return map(fib, random.sample(l, 3))
    \end{lstlisting}
    \vfill
    \centering\Large Do \texttt{fibo}, \texttt{bar} and \texttt{foo} update their arguments?
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Analysis: Pure Functions}
    \begin{lstlisting}[lang=python]
def f0(a):
    return a**2

def f1(a):
    b = f0(a)
    print b
    return b

l = list(...)
map(f0, l)
map(f1, l)
    \end{lstlisting}
    \vfill\centering\Large Are \texttt{f0} and \texttt{f1} pure functions?
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Analysis: Use - Def Chains}
    \begin{lstlisting}[lang=python]
a = '1'
if cond:
    a = int(a)
else:
    a = 3
print a
a = 4
    \end{lstlisting}
    \vfill\centering\Large Which version of \texttt{a} is seen by the \texttt{print} statement?
\end{frame}

\begin{frame}{Pythran's Moto \textbf{2/3}}
    \begin{center}
        Gather as many information as possible\\
        \vfill
        \Large Turn them into Code Optimizations!
        \vfill
    \end{center}
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Code Optimization: False Polymorphism}
    \begin{lstlisting}
a = cos(1)
if cond:
    a = str(a)
else:
    a = None
foo(a)
\end{lstlisting}
\vfill\centering\Large Is this code snippet statically typable?
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Code Optimization: Lazy Iterators}
    \begin{lstlisting}
def valid_conversion(n):
    l = map(math.cos, range(n))
    return sum(l)

def invalid_conversion(n):
    l = map(math.cos, range(n))
    l[0] = 1
    return sum(l) + max(l)
    \end{lstlisting}
    \vfill\centering\Large Which \texttt{map} can be converted into an \texttt{imap}
\end{frame}

\begin{frame}[fragile]
    \frametitle{Example of Code Optimization: Constant Folding}
    \begin{lstlisting}[lang=python]
def esieve(n):
    candidates = range(2, n+1)
    return sorted(
        set(candidates) - set(p*i
                              for p in candidates
                              for i in range(p, n+1))
        )

cache = esieve(100)
    \end{lstlisting}
    \vfill\centering\Large Can we evalute \texttt{esieve} at compile time?

\end{frame}

\begin{frame}{Pythran's Moto \textbf{3/3}}
    \begin{center}
Gather as many information as possible\\
\vfill
Turn them into Code Optimizations\\
\vfill
\Large Vectorize! Parallelize!
\end{center}
\end{frame}

\begin{frame}[fragile]
    \frametitle{Explicit Parallelization}
            \begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
def hyantes(xmin, ymin, xmax, ymax, step, range_, range_x, range_y, t):
  pt = [[0]*range_y for _ in range(range_x)]
  "omp parallel for"
  for i in xrange(range_x):
    for j in xrange(range_y):
      s = 0
      for k in t:
        tmp = 6368.* math.acos(math.cos(xmin+step*i)*math.cos( k[0] ) *
                               math.cos((ymin+step*j)-k[1]) +
                               math.sin(xmin+step*i)*math.sin(k[0]))
        if tmp < range_:
          s+=k[2] / (1+tmp)
      pt[i][j] = s
  return pt
\end{lstlisting}
\vfill
    \begin{tabular}{|l|r|r|r|}
        \hline
     Tool    &  CPython    &   Pythran     & OpenMP   \\
    \hline
     Timing  &  639.0ms    &   44.8ms       &      11.2ms       \\
    \hline
     Speedup &  $\times$1         &    $\times$14.2      &    $\times$57.7   \\
    \hline
\end{tabular}
\end{frame}

\begin{frame}{Library Level Optimizations}
    \begin{block}{Numpy is the \structure{key}}
        \begin{itemize}
            \item Basic block of Python Scientific Computing
            \item High-level Array Manipulations
            \item Many common functions implemented
        \end{itemize}
    \end{block}
    \begin{enumerate}
        \item This smells like FORTRAN
        \item For the compiler guy, FORTRAN smells good
        \item Unlock vectorization \& parallelization of Numpy code!
    \end{enumerate}

    \vfill
    \begin{center}
    \scriptsize{Cython is known to be "as fast as C", the only way we found to beat it is to be "as fast as Fortran", hence: Py\textbf{thran}}
    \end{center}

\end{frame}

\begin{frame}{Efficient Numpy Expressions}
    \begin{block}{Expression Templates}
        \begin{enumerate}
            \item A classic C++ meta-programming optimization
            \item Brings Lazy Evaluation to C++
            \item Equivalent to loop fusion
        \end{enumerate}
    \end{block}

    \begin{block}{More Optimizations}
        \begin{itemize}
            \item vectorization through \texttt{boost::simd} and \texttt{nt2}
            \item parallelization through \texttt{\#pragma omp}
        \end{itemize}
    \end{block}
\end{frame}

% \begin{frame}[fragile]
%     \frametitle{Benchmark}
%     \begin{lstlisting}
% def arc_dist(theta_1, phi_1, theta_2, phi_2):
%     temp = (np.sin((theta_2-theta_1)/2)**2
%         + np.cos(theta_1)*np.cos(theta_2)
%         * np.sin((phi_2-phi_1)/2)**2)
%     distance_matrix = 2 * np.arctan2(
%             sqrt(temp),sqrt(1-temp))
%     return distance_matrix
%     \end{lstlisting}
% 
%     \vfill
% 
%     \begin{tabular}{|l|r|r|r|r|}
%         \hline
%         Tool    &  CPython    &  Cython  &  Numexpr    & Pythran   \\
%         \hline
%         Timing  &  192.2ms    &  36.0ms  &    41.2ms   &  17.1ms   \\
%         \hline
%         Speedup &  $\times$1         &  $\times$5.33   &  $\times$4.67      &  $\times$11.23   \\
%         \hline
%     \end{tabular}
% \end{frame}

\begin{frame}[fragile]
    \frametitle{Julia Set, a Cython Example}

\begin{lstlisting}
def run_julia(cr, ci, N, bound, lim, cutoff):
    julia = np.empty((N, N), np.uint32)
    grid_x = np.linspace(-bound, bound, N)
    t0 = time()
    "omp parallel for"
    for i, x in enumerate(grid_x):
        for j, y in enumerate(grid_x):
            julia[i,j] = kernel(x, y, cr, ci, lim, cutoff)
    return julia, time() - t0
\end{lstlisting}
From Scipy2013 Cython Tutorial.
\vfill
    \begin{tabular}{|l|r|r|r|r|}
        \hline
        Tool    &  CPython    &  Cython  &  Pythran & +OpenMP   \\
        \hline
        Timing  &  3630ms    &   4.3ms  &   3.71ms  & 1.52ms  \\
        \hline
        Speedup & $\times$1 &  $\times$837 &  $\times$970 & $\times$2368   \\
        \hline
    \end{tabular}

\end{frame}

\begin{frame}[fragile]
    \frametitle{Mandelbrot, a Numba example}
    \begin{minipage}{0.49\textwidth}
\begin{lstlisting}[language=Python,frame=none,backgroundcolor=\color{white},basicstyle=\ttfamily\tiny]
@autojit
def mandel(x, y, max_iters):
    """
    Given the real and imaginary parts of a complex number,
    determine if it is a candidate for membership in the Mandelbrot
    set given a fixed number of iterations.
    """
    i = 0
    c = complex(x, y)
    z = 0.0j
    for i in range(max_iters):
        z = z**2 + c
        if abs(z)**2 >= 4:
            return i

    return 255
    \end{lstlisting}
\end{minipage}
\begin{minipage}{0.49\textwidth} \begin{lstlisting}[language=Python,frame=none,backgroundcolor=\color{white},basicstyle=\ttfamily\tiny]
@autojit
def create_fractal(min_x, max_x, min_y, max_y, image, iters):
    height = image.shape[0]
    width = image.shape[1]

    pixel_size_x = (max_x - min_x) / width
    pixel_size_y = (max_y - min_y) / height

    for x in range(width):
        real = min_x + x * pixel_size_x
        for y in range(height):
            imag = min_y + y * pixel_size_y
            color = mandel(real, imag, iters)
            image[y, x] = color

    return image
\end{lstlisting}
\end{minipage}
    \vfill
    \begin{tabular}{|l|r|r|r|r|}
        \hline
     Tool    &  CPython    &   Numba    &   Pythran  \\
        \hline
     Timing  &  8170ms   &    56ms       &   83ms  \\
        \hline
     Speedup & $\times$1 & $\times$145 & $\times$98 \\
    \hline
\end{tabular}
\end{frame}

\begin{frame}[fragile]
    \frametitle{Nqueens, to Show Some Cool Python}
    \begin{minipage}{0.49\textwidth}
            \begin{lstlisting}[frame=none,backgroundcolor=\color{white},basicstyle=\ttfamily\tiny]
# Pure-Python implementation of
# itertools.permutations()
# Why? Because we can :-)
def permutations(iterable, r=None):
  pool = tuple(iterable)
  n = len(pool)
  if r is None:
    r = n
  indices = range(n)
  cycles = range(n-r+1, n+1)[::-1]
  yield tuple(pool[i] for i in indices[:r])
  while n:
    for i in reversed(xrange(r)):
      cycles[i] -= 1
      if cycles[i] == 0:
        indices[i:] = indices[i+1:] + indices[i:i+1]
        cycles[i] = n - i
      else:
        j = cycles[i]
        indices[i], indices[-j] = \
            indices[-j], indices[i]
        yield tuple(pool[i] for i in indices[:r])
        break
    else:
      return
    \end{lstlisting}
\end{minipage}
\begin{minipage}{0.49\textwidth}
            \begin{lstlisting}[frame=none,backgroundcolor=\color{white},basicstyle=\ttfamily\tiny]
#pythran export n_queens(int)
def n_queens(queen_count):
  """N-Queens solver.
  Args:
    queen_count: the number of queens to solve
    for. This is also the board size.

  Yields:
    Solutions to the problem. Each yielded 
    value is looks like (3, 8, 2, ..., 6) 
    where each number is the column position 
    for the queen, and the index into the 
    tuple indicates the row.
  """
  out =list()
  cols = range(queen_count)
  for vec in permutations(cols,None):
    if (queen_count == len(set(vec[i]+i 
                               for i in cols))):
      out.append(vec)
  return out
\end{lstlisting}
\end{minipage}

    \begin{minipage}{0.49\textwidth}
      \scriptsize
Solving the NQueen problem, using generator, generator expression, list comprehension, sets\dots
    \\ http://code.google.com/p/unladen-swallow/
    \normalsize
\end{minipage}
\begin{minipage}{0.49\textwidth}

    \begin{tabular}{|l|r|r|r|r|}
        \hline
     Tool    &  CPython    &     PyPy     &  Pythran  \\
        \hline
     Timing  &  2640.6ms   &    501.1ms   &  693.3ms  \\
        \hline
     Speedup & $\times$1   & $\times$5.27 & $\times$3.8 \\
    \hline
\end{tabular}
\end{minipage}

\end{frame}

\begin{frame}[fragile]
    \frametitle{Numpy Benchmarks}

   \resizebox {\textwidth} {!} {
    \includegraphics{timing}
   }
    \url{https://github.com/serge-sans-paille/numpy-benchmarks/}

\vfill
    \scriptsize{Made with Matplotlib last night!}
\vfill

    \tiny{Debian clang version 3.5-1~exp1, x86-64, i7-3520M CPU @ 2.90GHz. No OpenMP/Vectorization for Pythran.}
\end{frame}

\begin{frame}{Crushing the Head of the Snake}

  \begin{center}
    \includegraphics[width=0.8\textwidth]{crushing}

\vfill
  I would have titled it "What every Pythonista should know about Python!"
\vfill

  \tiny (hopefully we'll get the video soon).
\vfill
  \end{center}

\end{frame}

\begin{frame}[fragile]
  \frametitle{The Compiler Is not the Solution to Keyboard-Chair Interface, Is It?}
    \begin{minipage}{0.49\textwidth}
\begin{lstlisting}[language=Python,frame=none,backgroundcolor=\color{white},basicstyle=\ttfamily\tiny]
def total(arr):
    s = 0
    for j in range(len(arr)):
        s += arr[j]
    return s

def varsum(arr):
    vs = 0
    for j in range(len(arr)):
        mean = (total(arr) / len(arr))
        vs += (arr[j] - mean) ** 2
    return vs

\end{lstlisting}
\end{minipage}
\begin{minipage}{0.49\textwidth} \begin{lstlisting}[language=Python,frame=none,backgroundcolor=\color{white},basicstyle=\ttfamily\tiny]
#pythran export stddev(float64 list list)
def stddev(partitions):
    ddof=1
    final = 0.0
    for part in partitions:
        m = total(part) / len(part)
        # Find the mean of the entire grup.
        gtotal = total([total(p) for p in partitions])
        glength = total([len(p) for p in partitions])
        g = gtotal / glength
        adj = ((2 * total(part) * (m - g)) + ((g ** 2 - m ** 2) * len(part)))
        final += varsum(part) + adj
    return math.sqrt(final / (glength - ddof))
\end{lstlisting}
\end{minipage}
\vfill
\begin{tabular}{|l|r|r|r|r|r|r|}
  \hline
  Version  &  Awful     &  Less Awful  &  OK & \textit{Differently OK} & \textit{OK} & Numpy   \\
  \hline
  CPython  & 127s       &   150ms  &   54.3ms  & 53.8ms    & 47.6ms & 8.2ms  \\
  \hline
  Pythran  & 1.38s      &   4.7ms  &   4.8ms   &  5.8ms    &  4.7ms & 1.7ms \\
        \hline
    \end{tabular}

\end{frame}

\begin{frame}{Engineering Stuff}
    \begin{block}{Get It}
        \begin{itemize}
            \item Follow the source: \url{https://github.com/serge-sans-paille/pythran} \\ \textit{(we are waiting for your pull requests)}
            \item Debian repo: \texttt{deb http://ridee.enstb.org/debian unstable main}
            \item Join us: \texttt{\#pythran} on Freenode 
            \textit{(very active)}, \\ or \url{pythran@freelists.org} 
            \textit{(not very active)}
        \end{itemize}
    \end{block}

    Available on PyPI, using Python 2.7, +2000 test cases, PEP8 approved, clang++ \& g++ ($>=4.8$) friendly, Linux and OSX validated.
\end{frame}

\begin{frame}{Conclusions}
    \hfill\tiny{\url{http://pythonhosted.org/pythran/}}
    \normalsize
    \begin{center}
        \vfill
        Compiling Python means more than typing and translating\\
        \vfill
        \uncover<2-> {
        What next: 
        \begin{itemize}
          \item<+-> Release the PyData version (ooops we're late)
          \item<+-> User module import (pull request already issued)
          \item<+-> Set-up an daily performance regression test bot (ongoing work with codespeed)
          \item<+-> Re-enable vectorization through boost::simd and nt2 (soon)
          \item<+-> More Numpy support, start looking into Scipy
          \item<+-> Polyhedral transformations (for my next life...)
        \end{itemize}
        }
        \vfill
        % \uncover<3->{
        %     Need a compiler guy / HPC Engineer? I am looking for a job!\\
        % }
        \vfill
    \end{center}
\end{frame}

\end{document}
% vim:spell spelllang=fr

### 2014-05-04 doc/papers/pydata2014/talk.tex
Update, thanks Serge

@@
 
-\date{PyData - May 4th 2014}
+\date{\small{PyData - May 4th 2014}}
 
@@
   \maketitle
-
+\vspace{-0.5cm}
   \begin{center}
-    \includegraphics[height=3em]{logo}
+    \includegraphics[height=6em]{logo}
   \end{center}
@@
     \frametitle{Static Compilation}
-    \begin{block}{Buys time for many time-consuming analyse}
+    \begin{block}{Buys time for many time-consuming analyses}
         Points-to, used-def, variable scope, memory effects, function purity\dots
@@
   pt = [[0]*range_y for _ in range(range_x)]
-  "omp parallel for"
+  "#omp parallel for"
   for i in xrange(range_x):
@@
     t0 = time()
-    "omp parallel for"
+    "#omp parallel for"
     for i, x in enumerate(grid_x):
@@
   \hline
-  Pythran  & 1.38s      &   4.7ms  &   4.8ms   &  5.8ms    &  4.7ms & 1.7ms \\
+
+  Pythran  & \uncover<2-> {1.38s}      &   \uncover<3-> {4.7ms}  &   \uncover<4-> {4.8ms}   &  \uncover<5-> {5.8ms}    &  \uncover<6-> {4.7ms} & \uncover<7-> {1.7ms} \\
         \hline
@@
         \begin{itemize}
-          \item<+-> Release the PyData version (ooops we're late)
-          \item<+-> User module import (pull request already issued)
-          \item<+-> Set-up an daily performance regression test bot (ongoing work with codespeed)
-          \item<+-> Re-enable vectorization through boost::simd and nt2 (soon)
-          \item<+-> More Numpy support, start looking into Scipy
-          \item<+-> Polyhedral transformations (for my next life...)
+          \item<3-> Release the PyData version \scriptsize{(ooops we're late)}
+          \item<4-> User module import \scriptsize{(pull request already issued)}
+          \item<5-> Set-up an daily performance regression test bot \scriptsize{(ongoing work with codespeed)}
+          \item<6-> Re-enable vectorization through boost::simd and nt2 \scriptsize{(soon)}
+          \item<7-> More Numpy support, start looking into Scipy
+          \item<8-> Polyhedral transformations \scriptsize{(in another life...)}
         \end{itemize}

### 2014-05-04 doc/papers/pydata2014/talk.tex
Update timing: -Ofast helped a lot on Mandelbrot!
Finale version of the slides presented @PyData 2014

@@
 \usepackage{listings}
+\usepackage{ulem}
 \definecolor{darkgreen}{rgb}{.1,.7,.1}
@@
 
-All timings were performed using OSX 10.9, an i7-3740QM CPU @ 2.70GHz, Python 2.7.6, current Pythran (c7de8b9c), pypy 2.2.1, numba 0.13.1, gcc 4.8, Clang r207887.
+Timings were performed using OSX 10.9, an i7-3740QM CPU @ 2.70GHz, Python 2.7.6, current Pythran (branch Pydata2014), pypy 2.2.1, numba 0.13.1, gcc 4.8, Clang r207887.
 \vfill
 I am \textbf{not} Pythonista, but I'm interested in performance in general. 
-Daily job: driver-level C code, assembly, multi-threaded C++, GPU, ...
+\\ Daily job: driver-level C code, assembly, multi-threaded C++, GPU, ...
 \vfill
@@
 \vfill
-\uncover<4->{
+\uncover<3->{
+\scriptsize{
   By the way this talk is written in Latex and takes more than 10 seconds to
   \textbf{compile}.
+}
 }
@@
   pt = [[0]*range_y for _ in range(range_x)]
-  "#omp parallel for"
+  #omp parallel for
   for i in xrange(range_x):
@@
     t0 = time()
-    "#omp parallel for"
+    #omp parallel for
     for i, x in enumerate(grid_x):
@@
         \hline
-     Timing  &  8170ms   &    56ms       &   83ms  \\
+     Timing  &  8170ms   &    56ms       &   47.2ms  \\
         \hline
-     Speedup & $\times$1 & $\times$145 & $\times$98 \\
+     Speedup & $\times$1 & $\times$145 & $\times$173 \\
     \hline
 \end{tabular}
+
+\scriptsize{"g++ -Ofast" here}
 \end{frame}
@@
 
-\begin{frame}{Conclusions}
-    \hfill\tiny{\url{http://pythonhosted.org/pythran/}}
+\begin{frame}[fragile]{Conclusion}
     \normalsize
-    \begin{center}
         \vfill
@@
         \uncover<2-> {
+        \vfill
         What next: 
+        \vfill
         \begin{itemize}
-          \item<3-> Release the PyData version \scriptsize{(ooops we're late)}
+          \item<2-> Release the PyData version  
+          \only<3>{\scriptsize{(ooops we're late)}} \only<4-> {\scriptsize{(\sout{ooops we're late} I did not specify which year)}}
           \item<4-> User module import \scriptsize{(pull request already issued)}
@@
         \vfill
-    \end{center}
+    \hfill\tiny{\url{http://pythonhosted.org/pythran/}}
 \end{frame}

### 2014-06-27 doc/papers/iop2014/paper.tex
Update article

--- 2014-01-29 doc/papers/iop2014/paper.tex
@@
 
-    Pythran is an open source static compiler that turns modules written
-    in a subset of Python language into native ones. Assuming that scientific
+    Pythran is an open source static compiler that turns modules written in a
+    subset of Python language into native ones. Assuming that scientific
     modules do not rely much on the dynamic features of the language, it trades
-    them for powerful, eventually inter procedural, optimizations.
-    These optimizations include detection of pure functions, temporary
-    allocation removal,
-    constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
-    thread-level parallelism through OpenMP annotations, false variable
-    polymorphism pruning, and automatic vector instruction generation such as
-    AVX or SSE.
+    them for powerful, possibly inter procedural, optimizations.  These
+    optimizations include detection of pure functions, temporary allocation
+    removal, constant folding, Numpy \texttt{ufunc} fusion and parallelization,
+    explicit thread-level parallelism through OpenMP annotations, false
+    variable polymorphism pruning, and automatic vector instruction generation
+    such as AVX or SSE.
 
@@
     % FIXME  (explain that annotations can be added to the code?)
-    Unlike the Cython approach , Pythran input code remains compatible with the
+    Unlike the Cython approach, Pythran input code remains compatible with the
     Python interpreter. Output code is generally as efficient as the annotated
@@
 The Python language is growing in popularity as a language for scientific
-computing, thanks to % FIXME (sounds strange for a publication)
- its concise syntax, a high level standard library and
-several scientific packages.
+computing, thanks to % FIXME (sounds strange for a publication) its concise
+syntax, a high level standard library and several scientific packages. The
+combination of a relatively efficient array package --\texttt{numpy}--, a
+library for scientific computing --\texttt{scipy}--, and enhanced interactive
+console --\texttt{ipython}-- and a comprehensive 2D plotting package
+--\texttt{matplotlib}-- makes it a relevant candidate for many daily scientific
+tasks.
 
@@
 compared to the same algorithm written in a statically compiled language such
-as C is high. % FIXME cut sentence It is partly due to the interpration cost,
-but also to Python's dynamic binding. Additionally, the Python compiler does
-not perform any optimization on the byte code, while scientific applications
-are first-class candidates for many of them.
-
-Since scientific applications are often told to spend 90\% of their time in
-10\% of the code, it is natural to focus on those computation-intensive pieces 
-of code. The goal of an optimizing compiler % FIXME (of what) 
-may not be to optimize the full Python application, but rather a subset of the application.
+as C is high. It is partly due to the interpretation cost, but also to Python's
+dynamic binding. Additionally, the Python compiler does not perform any
+optimization on the byte code, while scientific applications are first-class
+candidates for many of them. The ratio between an efficient C application and
+its Python counterpart can range from a factor of two, when the Python code
+only calls C routine, to a factor of one hundred or more when the Python code
+makes explicit array accesses.
+
+Given the increasing amount of data that scientists may need to process for
+their experiments, even two can be a limiting factor for the adoption of Python
+for computation intensive experiments. But since scientific applications are
+often told to spend 90\% of their time in 10\% of the code, one can focus on
+those computation-intensive pieces of code. The goal of an optimizing compiler
+may not be to optimize the full Python application, but rather a subset of the
+application.
 
@@
 GCC hardly vectorizes the output of Cython, while the array semantic is
-avaialable at the Python level. Taking into account the specificities of the
+available at the Python level. Taking into account the specificities of the
 Python language can unlock many new transformations.  For instance, PyPy
@@
 Finally, Pythran computes the set of functions and types used for the processed
-module, and generates the minimal set of headers that need to be included to
-provide the definitions needed to compile the generated C++ code. This step is
-critical to prevent typical slowdown during C++ compilation.
-
-The second step only consists in the instantiation of the top-level functions of
-the module, using user-provided signatures. Template instantiation then triggers
-the different correctly typed instantiations for all functions called in the
-module. Note that the user only needs to provide the type of the functions
-exported outside the module. The possible types of all internal functions are
-then inferred from the call sites. Note that this complex step leverage the 
-C++11 compiler capabilities: Pythran generates only one line using user-provided
-signature to trigger the template instanciation by the C++ compiler.
+module, and generates the minimal set of headers that provide the definitions
+needed to compile the generated C++ code. This step is critical to prevent
+typical slowdown during C++ compilation.
+
+The second step only consists in the instantiation of the top-level functions
+of the module, using user-provided signatures. Template instantiation then
+triggers the different correctly typed instantiations for all functions called
+in the module. Note that the user only needs to provide the type of the
+functions exported outside the module. The possible types of all internal
+functions are then inferred from the call sites. This complex step leverage the
+C++11 compiler capabilities: Pythran generates only one line using
+user-provided signature to trigger the template instanciation by the C++
+compiler.
 
@@
 library in the form of C++ template functions. Several optimizations, most
-notably using expression template, are delegated to this library. Pythran relies
-on the C++11~\cite{isocxx11} language and leverage recent features such as
+notably using expression templates, are delegated to this library. Pythran relies
+on the C++11~\cite{isocxx11} language and leverages recent features such as
 move semantics, type inference through \texttt{decltype(\dots)} and variadic
@@
 type-agnostic, \ie{they do not assume any type for the variables manipulated
-by the program}. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
+by the program}. The specialization of the generated code occurs only by generating the template instanciation.
 In other words, the Pythran compiler analyzes polymorphic functions and turns
@@
 other analyses, namely \texttt{GlobalEffects} that computes for each function whether
-this function modifies the global state (including I/O, random generators, etc.)
+it modifies the global state (including I/O, random generators, etc.)
 and \texttt{ArgumentEffects} that computes for each argument of each function 
@@
 
-Next section performs an in-depth comparison of Pythran with others Python
-optimizers: Cython and PyPy.
+Next section performs an in-depth comparison of Pythran with three other Python
+optimizers: CPython, Cython and PyPy.
 
@@
 the \texttt{pythran/test/cases} directory. The Pythran version used is the
-\texttt{HEAD} of the \texttt{iop2014} branch, PyPy 2.0.2 compiled with the
+\texttt{HEAD} of the \texttt{iop2014} branch. The other tools are: PyPy 2.0.2 compiled with the
 \texttt{-jit} flag, CPython 2.7.3 and Cython 0.20. All timings are made using
@@
 manual code annotation to achieve a similar result. It seems that PyPy support
-for numpy expression is still note mature for numpy expression, but there is a
+for numpy expression is still not mature for numpy expression, but there is a
 strong on-going effort of the community to fix this.
@@
 The ``growcut'' application is an image processing code where all array indexing
-are explicit. In that sense, this benchmark is complementary of the previous
-one, as it evaluates how well explicit array access are compiled.
-Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
+is explicit. In that sense, this benchmark is complementary to the previous
+one, as it evaluates how well explicit array accesses are compiled.
+Table~\ref{tbl:growcut} summarizes the results. All compilers yield impressive
 speedups on that benchmark, but this is is merely due to the fact that
@@
 translation of the Python code is very close to what a C developer would write.
-Pythran appears to outperforms Cython once all access are explicit, most
+Pythran appears to outperforms Cython once all accesses are explicit, most
 certainly due to cleaner code generation.
@@
 
+\section{Integration into Existing Applications}
+
+The \texttt{distutils} is the standard way to package applications. In addition
+to its command line interface, Pythran provides a \texttt{distutils} extension
+to integrate Pythran extension into existing applications. Its use is similar
+to the \texttt{Extension} class:
+
+\begin{lstlisting}
+from distutils.core import setup
+from pythran.dist import PythranExtension
+
+module1 = PythranExtension('demo', sources = ['a.py'])
+
+setup(name = 'demo',
+      version = '1.0',
+      description = 'This is a demo package',
+      ext_modules = [module1])
+\end{lstlisting}
+
+As Pythran input language is a subset of Python, it is still possible to write
+the \texttt{setup.py} so that the Python code is used as a regular module if
+the Pythran import fails.
+
+Pythran is supported on Linux and MacOSX systems.  It is available in source
+code form on the git repository
+\url{https://github.com/serge-sans-paille/pythran}, and on the cheese-shop
+\url{https://pypi.python.org/pypi/pythran}.
+
 \section{Future Work}
@@
 Although Pythran focuses on a subset of Python and its standard library, many
-optimizations opportunities are still available. Using a Domain Specific
+optimization opportunities are still available. Using a Domain Specific
 Language (DSL) approach, one could use rewriting rules to optimize several
@@
 support more functions from the original module. The extraction of Numpy
-expressions from \texttt{for loops} is also a natural optimization candidate, which
+expressions from \texttt{for} loops is also a natural optimization candidate, which
 shares similarities with code refactoring.
@@
 To conclude, limiting Python to a statically typed subset does not hinders the
-expressiveness when it comes to scientific or mathematic computations, but makes
-it possible to use a wide variety of classical optimizations to help Python
-match the performances of statically compiled language. Moreover, one can use
-high level information to generate efficient code that would be difficult to
-write for the average programmer.
+expressiveness when it comes to scientific or mathematic computations, but
+makes it possible to use a wide variety of classical optimizations to help
+Python match the performances of statically compiled language. Moreover, one
+can use high level information to generate efficient code that would be
+difficult to write for the average programmer. In the end, the scientific
+community can benefit from a high-level language to quickly write their
+experiments, and still run them at decent speed.
 

### 2014-08-22 doc/papers/cgo2014/paper.tex
Paper Layout

--- 2014-01-29 doc/papers/iop2014/paper.tex
@@
-% submit to https://sites.google.com/site/wpmvp2014/home
-\documentclass[10pt, onecolumn, preprint]{sigplanconf}
+\documentclass[10pt, preprint]{sigplanconf}
 
@@
 
-%\conferenceinfo{WPMVP '14}{February 16, 2014, Orlando, FL, USA} 
-%\copyrightyear{2014} 
-%\copyrightdata{978-1-4503-2653-7/14/02}
-%\doi{2568058.2568060}
-
 \lstset{captionpos=b, float, language=python}
@@
 
-\title{Pythran: Enabling Static Optimization of Scientific Python Programs}
+\title{Efficient Compilation of High Level Python Numerical Programs with Pythran}
 
@@
            {mehdi.amini@silkan.com}
-\authorinfo{Adrien Merlini}
-           {T{\'e}l{\'e}com Bretagne}
-           {adrien.merlini@telecom-bretagne.eu}
-\authorinfo{Xavier Corbillon}
-           {T{\'e}l{\'e}com Bretagne}
-           {xavier.corbillon@telecom-bretagne.eu}
-\authorinfo{Alan Raynaud}
-           {T{\'e}l{\'e}com Bretagne}
-           {alan.raynaud@telecom-bretagne.eu}
 
@@
 
-    Pythran is an open source static compiler that turns modules written
-    in a subset of Python language into native ones. Assuming that scientific
-    modules do not rely much on the dynamic features of the language, it trades
-    them for powerful, eventually inter procedural, optimizations.
-    These optimizations include detection of pure functions, temporary
-    allocation removal,
-    constant folding, Numpy \texttt{ufunc} fusion and parallelization, explicit
-    thread-level parallelism through OpenMP annotations, false variable
-    polymorphism pruning, and automatic vector instruction generation such as
-    AVX or SSE.
+    For decades, FORTRAN and C(++) have dominated the landscape of High
+    Performance Computing languages, leaving interpreted language like Matlab,
+    R, Python or more recently Julia for experimentation and prototyping.
 
-    In addition to these compilation steps, Pythran provides a C++ runtime 
-    library that
-    leverages the C++ STL to provide generic containers, and the Numeric 
-    % FIXME (phrase trop longuuuuuuuuue) 
-    Template Toolbox (NT2) for Numpy support. It takes advantage of modern C++11
-    features such as variadic templates, type inference, move semantics and
-    perfect forwarding, as well as classical idioms such as expression templates.
+    As more scientists use these scripting language, there is an on-going
+    research effort to compile them into efficient native code. However the
+    computations kernels often differ from classical ones as these languages
+    favors an high-level coding style, where loops are often replaced by calls
+    to the relevant toolbox/module/library/package. As a consequence, classical
+    loop optimization techniques like loop fusion, tiling and related
+    vectorization or parallelization techniques are not directly applicable.
 
-    % FIXME  (explain that annotations can be added to the code?)
-    Unlike the Cython approach , Pythran input code remains compatible with the
-    Python interpreter. Output code is generally as efficient as the annotated
-    Cython equivalent, if not more, but without the backward compatibility
-    loss.
+    This papers focuses on the case of the Python language and especially the
+    Numpy package that provides core array data structure and basic linear
+    algebra routines. It first conducts a case study based on Numpy user
+    questions from the stackoverflow question and answer website, then studies
+    existing compilers for numerical Python. Optimization opportunities,
+    including parallelization and vectorization, and their implementation in
+    the Pythran compiler are presented and illustrated through several
+    benchmarks, showing very interesting speedup over the standard Python
+    interpreter. Speedup of factor greater than 10 are achieved despite the
+    fact the considered benchmarks mostly call Numpy routines that run C code.
 
@@
 \keywords
-static compilation, parallelization, Python, C++
+static compilation, parallelization, Python, C++, SIMD
 
@@
 \section{Introduction}
+% On aura pas le temps pour le papier je pense mais faire une comparaison avec du
+% C pure pour voir ce que les compilo de C arrive a faire.
 
-The Python language is growing in popularity as a language for scientific
-computing, thanks to % FIXME (sounds strange for a publication)
- its concise syntax, a high level standard library and
-several scientific packages.
+\cite{rossum97, scipy}
 
-However, the overhead of running a scientific application written in Python
-compared to the same algorithm written in a statically compiled language such
-as C is high. % FIXME cut sentence It is partly due to the interpration cost,
-but also to Python's dynamic binding. Additionally, the Python compiler does
-not perform any optimization on the byte code, while scientific applications
-are first-class candidates for many of them.
+%%
+%%
+\section{A Crowd-Sourced Numpy Benchmark}
 
-Since scientific applications are often told to spend 90\% of their time in
-10\% of the code, it is natural to focus on those computation-intensive pieces 
-of code. The goal of an optimizing compiler % FIXME (of what) 
-may not be to optimize the full Python application, but rather a subset of the application.
+%%
+%%
+\section{Optimizations Opportunities in a Typical Numpy Kernel}
 
-Several tools have been developed by an active community to fill the performance
-gap encountered when running these computation-intensive piece of code, either
-% FIXME after? (perfs issues are not on the compilation (except for...))
-through static compilation or Just In Time (JIT) compilation.
+\subsection{Temporaries Elemination}
 
-One approach, used by Cython~\cite{cython2010} is to suppress the interpretation
-overhead by translating Python Programs to C programs calling the Python C
-API~\cite{pythoncapi}. More recently, Nuitka~\cite{nuitka}  has taken the same
-approach using C++ as a back-end. Going a step further Cython also provides an
-hybrid C/Python language that can efficiently be translated into C code. It 
-relies both on the Python C API and on plain C.
-ShedSkin~\cite{shedskin2006} translates implicitly strongly typed Python programs
-into C++, without any call to the Python C API, but fails to compile dynamic
-codes or codes for which static type inference is impossible.
+\subsection{Operator Fusion}
 
-The alternate approach consists in writing a JIT compiler, embedded into the
-interpreter, to dynamically turn the computation intensive parts into native
-code. This is what the \texttt{numexpr} module~\cite{numexpr} does for Numpy 
-expressions, JIT-compiling them from a string representation to native code.
-Numba~\cite{numba} and Parakeet~\cite{parakeet2012} extend this approach to
-Numpy-centric applications while PyPy~\cite{pypy2009} applies it to the whole
-language - including the Numpy module through the in-progress \texttt{numpypy}
-branch.
+\subsection{Loop Vectorization}
 
-Except for PyPy, none of these compilers apply any of the static optimization
-techniques that have been known and successfully applied 
-% FIXME And Numba? It generates LLVM, which uses static optimization techniques
-for decades to statically compiled language such as C or C++.  Translators to
-statically compiled languages do take advantage of them indirectly, but the
-quality of the generated code may prevent advanced optimizations. For instance
-GCC hardly vectorizes the output of Cython, while the array semantic is
-avaialable at the Python level. Taking into account the specificities of the
-Python language can unlock many new transformations.  For instance, PyPy
-automatically converts the \texttt{range} builtin into \texttt{xrange} by using
-a dedicated structure called \texttt{range-list}.
+\subsection{Loop Parallelization}
 
-This article presents Pythran, an optimizing compiler for a subset of the
-Python language that turns implicitly statically typed modules into parametric
-C++ code. It supports many high-level constructs of the 2.7 version of the
-Python language such as list comprehension, set comprehension, dict
-comprehension, generator expression, lambda functions, nested functions,
-polymorphic functions and global variables. It does \textbf{not} support user
-classes or any dynamic feature such as introspection or polymorphic variables.
+%%
+%%
+\section{Existing Compilation Approach}
 
-Unlike existing alternatives, Pythran does not solely perform static typing of
-Python programs. It also performs various compiler optimizations such as
-detection of pure functions, temporary allocation removal or constant folding.
-These transformations are backed up by code analyses such as aliasing,
-inter-procedural memory effect computations or use-def chains.
+\subsection{Numpy}
+\cite{oliphant2007,numpyarray2011}
 
-The article is structured as follows: Section~\ref{sec:infrastructure}
-introduces the Pythran compiler compilation flow and internal representation.
-Section~\ref{sec:analysis} presents several code analyses while
-Section~\ref{sec:optimizations} focuses on code optimizations.
-Section~\ref{sec:backend} presents back-end optimizations for Numpy
-expressions. Section~\ref{sec:openmp}  briefly introduces OpenMP-like
-annotations for explicit parallelization of Python programs and
-Section~\ref{sec:benchmarks} presents the performance obtained on a few
-synthetic benchmarks and concludes.
+\subsection{Cython}
+\cite{cython2010}
 
-\section{Pythran Compiler Infrastructure}
-\label{sec:infrastructure}
+\subsection{Numba}
+\cite{numba}
 
-Pythran is a compiler for a subset of the Python language. In this paper, the
-name \textbf{Pythran} will be used indifferently to refer to the language or
-the associated compiler. The input of the Pythran compiler is a Python module
----not a Python program--- meant to be turned into a native module. Typically,
-computation-intensive parts of the program are moved to a module fed to
-Pythran. The module is then translated to polymorphic C++ code through the
-generation of templated classes in place of Python functions. These classes are
-either used from regular C++ applications, or instantiated for given types to
-generate a native Python module.
+\subsection{Parakeet}
+\cite{parakeet2012}
 
-It is a strong requirement for Pythran to maintain backward compatibility with
-Python, that is, unlike Cython, all code compilable by Pythran is still
-executable by a standard-conforming Python interpreter. For this reason, in
-addition to language restrictions detailed in the following, Pythran
-understands special comments such as:
+\subsection{Copperhead}
+\cite{copperhead2011}
 
-\begin{lstlisting}
-#pythran export foo(int list, float)
-\end{lstlisting}
+\subsection{PyPy}
+\cite{pypy2009}
 
-as an optional module signature. One does not need to list all the module
-functions in an \texttt{export} directive, only the functions meant to be used
-outside of the module, i.e imported by another module.
-Polymorphic functions can be listed several times with
-different types. These type annotations are only used for native code generation.
+%%
+%%
+\section{Compiler - Runtime Cooperation}
 
-The Pythran compiler is built as a traditional static compiler: a front-end
-turns Python code into an Internal Representation (IR), a middle-end performs
-various code optimizations on this IR, and a back-end turns the IR into C++
-code. The front-end performs two steps:
+\subsection{Expression Templates}
 
-\begin{enumerate}
+\cite{expression_templates}
 
-    \item turn Python code into Python Abstract Syntax Tree (AST) by using the \texttt{ast}
-   module from the standard library;
+\subsection{Forward Substitution}
 
-    \item turn the Python AST into a type-agnostic Pythran IR, which remains a subset
-   of the Python AST.
+\subsection{Range Analysis}
 
-\end{enumerate}
+\subsection{Constant Propagation and Loop Unrolling}
 
-Pythran IR is similar to the Python AST, as defined in the \texttt{ast} module, except
-that several nodes are forbidden (for instance because Pythran does not support
-user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
-to others to form a simpler and easier to deal with AST for further analyses and
-optimizations. The transformations applied by Pythran on Python AST are the
-following:
+\subsection{Vectorization and Parallelization}
 
-\begin{itemize}
-    \item list/set/dict comprehension are expanded into loops wrapped into a function call;
+\cite{esterie2012boost, pyhpc2013, wpmvp2014}
 
-    \item destructuring is expanded into several variable assignments;
+%%
+%%
+\section{Experiments}
 
-    \item lambda functions are turned into named nested functions;
-
-    \item the closure of nested functions is statically computed to turn nested
-        functions into global functions taking the closure elements as
-        parameters. The initial definition is replaced by a \emph{partial
-        function} creation, using the \texttt{partial} type from the standard \texttt{functools} module;
-
-    \item implicit \texttt{return None} are made explicit;
-
-    \item all imports are fully expanded to make functions and global variables access paths explicit,
-
-    \item method calls are turned into function calls;
-
-    \item implicit \texttt{\_\_builtin\_\_} function calls are made explicit;
-
-    \item \texttt{try \dots finally} instriuction are turned into nested \texttt{try \dots except} instructions;
-
-    \item identifiers whose name may be in conflict with C++ keywords are renamed.
-
-\end{itemize}
-
-The back-end operates in three steps:
-
-\begin{enumerate}
-
-    \item turning Pythran IR into parametric C++ code;
-
-    \item instantiating the C++ code for the desired types;
-
-    \item compiling the generated C++ code into native code.
-
-\end{enumerate}
-
-The first step requires to map polymorphic variables and polymorphic functions
-from the Python world to C++. Pythran only supports polymorphic variables for
-functions, i.e. a variable can hold several function objects during its life
-time, but it cannot be assigned to a string if it has already been assigned to
-an integer.  As shown later, it is possible to detect several false variable
-polymorphism cases using use-def chains. Function polymorphism is achieved
-through template parameters: a template function can be applied to several
-types as long as an implicit structural typing is respected. This technique is
-very similar to Python's duck typing, except that it is checked at compile
-time, as illustrated by the following implementation of a generic dot product
-in Python:
-
-\begin{lstlisting}
-def dot(L0, L1):
-    return sum(x*y for x,y in zip(L0,L1))
-\end{lstlisting}
-
-\noindent and in C++:
-
-\begin{lstlisting}[language=c++]
-template<class T0, class T1>
-    auto dot(T0&& L0, T1&& L1) -> decltype(/* skipped */)
-    {
-        return pythonic::sum(
-            pythonic::map(
-                operator_::multiply(),
-                    pythonic::zip(
-                        std::forward<T0>(L0),
-                        std::forward<T1>(L1))
-            )
-        );
-    }
-\end{lstlisting}
-
-Although more verbose than the Python version, the C++ version also uses some
-form of structural typing: the only assumptions these two versions make are
-that \texttt{L0} and \texttt{L1} are iterable, their content can be multiplied
-and that the result of the multiplication is accumulatable.
-
-Finally, Pythran computes the set of functions and types used for the processed
-module, and generates the minimal set of headers that need to be included to
-provide the definitions needed to compile the generated C++ code. This step is
-critical to prevent typical slowdown during C++ compilation.
-
-The second step only consists in the instantiation of the top-level functions of
-the module, using user-provided signatures. Template instantiation then triggers
-the different correctly typed instantiations for all functions called in the
-module. Note that the user only needs to provide the type of the functions
-exported outside the module. The possible types of all internal functions are
-then inferred from the call sites. Note that this complex step leverage the 
-C++11 compiler capabilities: Pythran generates only one line using user-provided
-signature to trigger the template instanciation by the C++ compiler.
-
-The last step involves a template library, called \texttt{pythonic} that 
-contains a polymorphic implementation of many functions of the Python standard
-library in the form of C++ template functions. Several optimizations, most
-notably using expression template, are delegated to this library. Pythran relies
-on the C++11~\cite{isocxx11} language and leverage recent features such as
-move semantics, type inference through \texttt{decltype(\dots)} and variadic
-templates.
-As a consequence it requires a C++11 compatible compiler for the native code
-generation. Boost.Python~\cite{boostpython2007} is used for the automatic
-Python-to-C++ boilerplate. Generated C++ code is compatible starting with g++ 
-4.8.2 and clang++ 3.3.
-
-It is important to note that all Pythran analyses and optimizations are
-type-agnostic, \ie{they do not assume any type for the variables manipulated
-by the program}. The generated C++ template is still type agnostic, the specialization occurs only by generating the template instanciation.
-In other words, the Pythran compiler analyzes polymorphic functions and turns
-Python modules into C++ meta-programs.
-
-Figure~\ref{fig:pythran-compiler} summarizes the compilation flow and the 
-involved tools.
+\subsection{The Pythran Compiler}
+\cite{pythran2013,isocxx11}
 
@@
 
-\section{Code Analyses}
-\label{sec:analysis}
+\subsection{Expermimental Setup}
 
-A code analysis is a function that takes a part of the IR (or the whole
-module's IR) as an input and returns aggregated high-level information. For
-instance, a simple Pythran analysis called \texttt{Identifiers} gathers the set
-of all identifiers used throughout the program. This information is later used
-when creating new identifiers so that no conflict occurs with already existing ones.
+\subsection{Results and Analysis}
 
-One of the most important analysis in Pythran is the \textbf{alias analysis}, sometimes
-referred as \textbf{points-to} analysis. For each identifier, it computes an
-approximation of the set of locations this identifier may points to. For
-instance, let us consider the polymorphic function \texttt{foo} defined as follows:
-
-\begin{lstlisting}
-def foo(a,b):
-    c = a or b
-    return c * 2
-\end{lstlisting}
-
-The identifier \texttt{c} involved in the multiplication may refer to
-
-\begin{itemize}
-    \item a fresh location if \texttt{a} and \texttt{b} are scalars
-
-    \item the same location as \texttt{a} if \texttt{a} evaluates to \texttt{True}
-
-    \item the same location as \texttt{b} otherwise.
-
-\end{itemize}
-
-As we do not specialise the analysis for different types and the truth value of
-\texttt{a} is unknown at compilation time, the alias analysis yields % FIXME (sounds strange) 
-the approximated result that \texttt{c} may point to a fresh location, 
-\texttt{a} or \texttt{b}.
-
-Without this kind of information, even a simple instruction like
-\texttt{sum(a)} would yield very few information as there is no guarantee that
-the \texttt{sum} identifiers point to the \texttt{sum} built-in, as it could have been 
-previously stated that \texttt{sum = len}.
-
-When turning Python AST into Pythran IR, nested functions are turned into
-global functions taking their expanded closure as parameter. This closure is
-computed using the information provided by the \texttt{Globals} and
-\texttt{ImportedIds} analyse. The former statically computes the state of the
-dictionary of globals, and the latter computes the set of identifiers used by
-an instruction but not declared in this instruction. For instance in the
-following snippet:
-
-\begin{lstlisting}
-def outer(outer_argument):
-    def inner(inner_argument):
-        return cos(outer_argument) + inner_argument
-    return inner(3)
-\end{lstlisting}
-
-The \texttt{Globals} analysis called on the \texttt{inner} function definition
-marks \texttt{cos} as a global variable, and \texttt{ImportedIds} marks
-\texttt{outer\_argument} and \texttt{cos} as imported identifiers. The result
-of the transformation of nested function into global functions on that
-particular example is:
-
-\begin{lstlisting}
-import functools
-
-def _inner(outer_argument, inner_argument):
-    return cos(outer_argument) + inner_argument
-
-def outer(outer_argument):
-    inner = functools.partial(_inner, outer_argument)
-    return inner(3)
-\end{lstlisting}
-
-A rather high-level analysis is the \texttt{PureFunctions} analysis, that computes the
-set of functions declared in the module that are pure, \ie{whose return value
-only depends from the value of their arguments and do not change any global state}.
-This analysis depends on two
-other analyses, namely \texttt{GlobalEffects} that computes for each function whether
-this function modifies the global state (including I/O, random generators, etc.)
-and \texttt{ArgumentEffects} that computes for each argument of each function 
-whether or not this argument may be updated in the function body. These three
-analyses work inter-procedurally, as illustrated in the following example:
-
-\begin{lstlisting}
-def fibo(n):
-    return n if n < 2 else fibo(n - 1) + fibo(n - 2)
-
-def bar(l):
-    return map(fibo, l)
-
-def foo(l):
-    return map(fibo, random.sample(l, 3))
-\end{lstlisting}
-
-The \texttt{fibo} function is pure as it has no global effect or argument effect and
-only calls itself. As consequence the \texttt{bar} function is also pure since the
-\texttt{map} intrinsic is pure when its first argument is pure. However the \texttt{foo}
-function is not pure since it calls the \texttt{sample} function from the \texttt{random}
-module, which has a global effect (on the underlying random number generator
-internal state).
-
-Several analyses depend on the \texttt{PureFunctions} analysis.
-\texttt{ParallelMaps} uses aliasing information to check if an identifier
-points to the \texttt{map} intrinsic, and checks if the first argument is a
-pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
-added to the set of parallel maps, because it can be executed in any order.
-This is the case for the first \texttt{map} in the following snippet, but not
-for the second because \texttt{print b} involves an \textbf{I/O}.
-% Which is not really usefull as there is no way to say pythran we want to execut it in parallel
-% (but logging information is still displayed)
-
-\begin{lstlisting}
-def pure(a):
-    return a ** 2
-
-def guilty(a):
-    b = pure(a)
-    print b
-    return b
-
-l = list(...)
-map(pure, l)
-map(guilty, l)
-\end{lstlisting}
-
-\texttt{ConstantExpressions} uses function purity to decide whether or not a given expression
-is constant, \ie{its value only depends on literals}. For instance the
-expression \texttt{fibo(12)} is a constant expression because \texttt{fibo} is pure and its
-argument is a literal. 
-% FIXME (pourrait etre interressant de repondre a : et t'en fait quoi calcul a la compilation? calcul une fois si appele plusieurs fois ?... ou ref a la transformation qui fait ca)
-% PB : explain later I think.
-
-\texttt{UseDefChains} is a classical analysis from the static compilation world. For
-each variable defined in a function, it computes the chain of \textbf{use} and \textbf{def}.
-The result can be used to drive various code transformations, for instance to
-remove dead code, such as a \textbf{def} followed by either nothing or a 
-\textbf{def} is useless. 
-It is used in Pythran to avoid false polymorphism. An intuitive way to represent
-use-def chains is illustrated on the next code snippet:
-
-\begin{lstlisting}
-a = 1
-if cond:
-    a = a + 2
-else:
-    a = 3
-print a
-a = 4
-\end{lstlisting}
-
-In this example, there are two possible chains starting from the first
-assignment. Using \texttt{U} to denote \textbf{use} and \texttt{D} to denote \textbf{def}, one gets:
-
-\begin{center}
-\begin{tikzpicture}[D/.style={black!50,fill=black!10, circle, drop shadow, align=center},
-    U/.style={black!50,fill=black!10, rectangle, drop shadow, align=center}]
-    \node[D] (d0) {D};
-    \node[U] (u0) [right=of d0] {U};
-    \node[D] (d1) [right=of u0] {D};
-    \node[U] (u1) [right=of d1] {U};
-    \node[D] (d2) [right=of u1] {D};
-
-    \node[D] (dp0)[below=of u0, yshift=2em] {D};
-
-    \draw[very thick, ->] (d0) -- (u0);
-    \draw[very thick, ->] (d0) |- (dp0);
-    \draw[very thick, ->] (u0) -- (d1);
-    \draw[very thick, ->] (d1) -- (u1);
-    \draw[very thick, ->] (dp0) -| (u1);
-    \draw[very thick, ->] (u1) -- (d2);
-\end{tikzpicture}
-\end{center}
-
-The fact that all paths finish by a \textbf{def} indicates that the last
-assignment can be removed (but not necessarily its right hand part that could
-have a side-effect). % FIXME example?
-In the use-def Chains, we can gather all nodes tied with a \textbf{use} node or
-preceded by a common \textbf{def} node.
-All these nodes refere to the same variable and can
-safely be renamed from \texttt{a} to \texttt{a}' (thus performing scalar
-renaming). It avoid false polymorphism.
-% FIXME The scalar renaming part is unclear
-% PB : Is it clearer now? I don't speak about UD node but we don't care, do we?
-
-All the above analyses are used by Pythran developers to build code
-transformations that improve the execution time of the generated code.
-
-\section{Code Optimizations}
-\label{sec:optimizations}
-
-One of the benefits of translating Python code to C++ code is that it removes
-most of the dynamic lookups. It also unveils all the optimizations available at
-C++ level. For instance, a function call is quite costly in Python, which
-advocates in favor of using inlining. This transformation comes at no cost when
-using C++ as the back-end language, as the C++ compiler performs this 
-transformation triggered by fine tuned heuristics.
-
-However, there are some information available at the Python level that cannot
-be recovered at the C++ level. For instance, Pythran uses functor with an
-internal state and a goto dispatch table to represent generators. This approach
-is not very efficient, especially for trivial cases.
-Such trivial cases appear when a generator expression is converted, in the
-front-end, to a looping generator. To avoid this extra cost, Pythran turns
-generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
-\texttt{itertools} module whenever possible, removing the unnecessary goto
-dispatching table. This kind of transformation cannot be made by the C++
-compiler. For instance, the one-liner \texttt{len(set(vec[i]i + i for i in cols))}
-extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
-is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i] + i,cols)))}.
-This new form is less efficient in pure Python (it implies one extra function
-call per iteration), but can be compiled into C++ more efficiently than a
-generic generator.
-
-A similar optimization consists in turning \texttt{map}, \texttt{zip} or
-\texttt{filter} into their equivalent version from the \texttt{itertool}
-module. The benefit is twofold: first it removes a temporary allocation, second
-it gives an opportunity to the compiler to replace list accesses by scalar
-accesses. This transformation is not always valid, nor profitable. It is not
-valid if the content of the output list is written later on, and not profitable
-if the content of the output list is read several times, as each read implies
-the (re) computation, as illustrated in the following code:
-
-\begin{lstlisting}
-def valid_conversion(n):
-    # this map can be converted to imap
-    L = map(math.cos, range(n))
-    return sum(L) # sum iterates once on its input
-
-def invalid_conversion(n):
-    # this map cannot be converted to imap
-    L = map(math.cos, range(n))
-    L[0] = 1  # invalid assignment if converted to imap
-    return sum(L) + max(L) # two iterations over L prevents imap transformation
-\end{lstlisting}
-
-The information regarding constant expressions is used to perform a classical
-transformation called \texttt{ConstantFolding}, which consists in the compile-time
-evaluation of constant expressions. The validity is guaranteed by the
-\texttt{ConstantExpressions} analysis, and the evaluation relies on Python ability to
-compile an AST into byte code and run it, benefiting from the fact that Pythran
-IR is a subset of Python AST. A typical illustration is the initialization of a
-cache at compile-time:
-
-\begin{lstlisting}
-def esieve(n):
-    candidates = range(2, n+1)
-    return sorted(
-        set(candidates) - set(p*i
-                              for p in candidates
-                              for i in range(p, n+1))
-        )
-
-cache = esieve(100)
-\end{lstlisting}
-
-Pythran automatically detects that \texttt{eseive} is a pure function and evaluates
-the \texttt{cache} variable value at compile time.
-
-\texttt{ConstantFolding} may create a lot of candidates for 
-\texttt{LoopUnrolling}. In Python, loops are used to iterate over sequences. If
-such a sequence is an explicit list, then Pythran unrolls the loop accordingly.
-Thanks to \texttt{ConstantFolding}, a call to \texttt{range(4)} will statically
-turn into \texttt{[0, 1, 2, 3]} (after a check that \texttt{range} actually
-points to \texttt{\_\_builtin\_\_.range}) making it possible to unroll the loop
-statically.
-
-Sometimes, developers use the same variable in a function to represent value 
-with different types, which leads to false polymorphism, as in:
-
-\begin{lstlisting}
-a = cos(1)
-a = str(a)
-\end{lstlisting}
-
-These instructions cannot be translated to C++ directly because \texttt{a}
-would have both \texttt{double} and \texttt{str} type. However, using
-\texttt{UseDefChains} it is possible to assert the validity of the renaming of
-the instructions into:
-
-\begin{lstlisting}
-a = cos(1)
-a_ = str(a)
-\end{lstlisting}
-
-that does not have the same typing issue.
-
-In addition to these python-level optimizations, the Pythran back end library,
-\texttt{pythonic}, uses several well known optimizations, especially for Numpy
-expressions.
-
-\section{Library Level Optimizations}
-\label{sec:backend}
-
-Using the proper library, the \cpp language provides an abstraction level close
-to what Python proposes. Pythran provides a wrapper library, \texttt{pythonic},
-that leverage on the \cpp Standard Template Library (STL), the GNU Multiple
-Precision Arithmetic Library (GMP) and the Numerical Template Toolbox
-(NT2)\footnote{\url{https://github.com/MetaScale/nt2}} to emulate Python
-standard library. The STL is used to provide a typed version of the standard
-containers (\texttt{list}, \texttt{set}, \texttt{dict} and \texttt{str}), as
-well as reference-based memory management through \texttt{shared\_ptr}. Generic
-algorithms such as \texttt{accumulate} are used when possible. GMP is the
-natural choice to represent Python's \texttt{long} in \cpp. NT2 provides a generic
-vector library called \texttt{boost.simd}~\cite{esterie2012boost} that enables
-the use of vector instruction units of modern processors in a generic way. It is used
-to efficiently compile Numpy expressions.
-
-Numpy expressions are the perfect candidates for library level optimizations.
-Pythran implements three optimizations on such expressions:
-
-\begin{enumerate}
-
-    \item Expression templates~\cite{expression_templates} are used to avoid
-        multiple iterations and the creation of intermediate arrays. Because
-        they aggregate all \texttt{ufunc} into a single expression at compile
-        time, they also increase the computation intensity of the loop body,
-        and hence increase the impact of the two following optimizations.
-        % what about binary func?
-        % ufunc -> micro function, in numpy ;-)
-
-    \item Loop vectorization. All modern processors have vector instruction
-        units capable of applying the same operation on a vector of data
-        instead of a single data. For instance Intel Sandy Bridge can run 8
-        single-precision additions per instruction. One can directly use the
-        vector instruction set assembly to use these vector units, or use C/C++
-        intrinsics. Pythran relies on \texttt{boost.simd} from NT2 that offers
-        a generic vector implementation of all standard math functions to
-        generate a vectorized version of Numpy expressions. Again, the
-        aggregation of operators performed by the expression templates proves
-        to be beneficial, as it reduces the number of (costly) loads from the
-        main memory to the vector unit.
-
-    \item Loop parallelization through OpenMP~\cite{openmp3.1}. Numpy expression
-        computation do not carry any loop-dependency. They are perfect
-        candidates for loop parallelization, especially after the expression
-        templates aggregation, as OpenMP generally performs better on loops
-        with higher computation intensity that masks the scheduling overhead.
-
-\end{enumerate}
-
-To illustrate the benefits of these three optimizations combined, let us
-consider the simple Numpy expression:
-
-\begin{lstlisting}
-d = numpy.sqrt(b*b+c*c)
-\end{lstlisting}
-
-When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
-i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
-% PB : Not sure it is needed as audiance don't really care about it.
-
-\begin{lstlisting}
->>> %timeit np.sqrt(b*b+c*c)
-1000 loops, best of 3: 1.23 ms per loop
-\end{lstlisting}
-
-\noindent then after Pythran processing and using expression templates:
-
-\begin{lstlisting}
->>> %timeit my.pythranized(b,c)
-1000 loops, best of 3: 621 us per loop
-\end{lstlisting}
-
-The speed-up is due to the fact that expression templates replace 4 temporary
-array creations and 4 loops by a single allocation and a single loop.
-
-Going a step further and vectorizing the generated loop yields an extra
-performance boost:
-
-\begin{lstlisting}
->>> %timeit my.pythranized(b,c)
-1000 loops, best of 3: 418 us per loop
-\end{lstlisting}
-
-Although the AVX instruction set makes it possible to store 4 double precision
-floats, one does not get a 4x speed up because of the (unaligned) memory transfers
-to and from vector registers.
-% 1.33x speedup is not really good if the best expected is 4x. unaligned explain why it is not better 
-% than 2x but what about the others 0.66?
-
-Finally, using both expression templates, vectorization and OpenMP:
-
-\begin{lstlisting}
->>> %timeit my.pythranized(b,c)
-1000 loops, best of 3: 105 us per loop
-\end{lstlisting}
-
-The 4 hyper-threaded cores give an extra performance boost. In the end, Pythran
-generates a native module that performs roughly $\times11$ faster than the
-original version.
-
-As a reference, the \texttt{numexpr} module that performs JIT optimization of the
-expression yields the following timing:
-
-\begin{lstlisting}
->>> %timeit numexpr.evaluate("sqrt(b*b+c*c)")
-1000 loops, best of 3: 395 us per loop
-\end{lstlisting}
-
-\noindent Pythran still performs almost four times faster.
-
-\section{Explicit Parallelization}
-\label{sec:openmp}
-
-Many scientific applications can benefit from the parallel execution of their
-kernels. As modern computers generally feature several processors and several
-cores per processor, it is critical for the scientific application developers to
-be able to take advantage of them.
-%Does python programmer really use it? Or people using it may write their native code?
-% They may write it using python only to prototyping?
-
-As explained in the previous section, Pythran takes advantage of multiple cores
-when compiling Numpy expressions. However, when possible, it is often more
-profitable to parallelize the outermost loops rather than the inner loops
----the Numpy expressions--- because it avoids the synchronization barrier at
-the end of each parallel section, and generally offers more computation
-intensive computations.
-
-The OpenMP standard~\cite{openmp3.1} is a widely used solution for Fortran, C
-and C++ to describe loop-based and task-based parallelism. It consists of a few
-directives attached to the code, directives that describe parallel loops and parallel code
-sections in a shared memory model.
-
-Pythran makes this directives available at the Python level through string
-instructions. The semantic is roughly similar to the semantics defined by the
-standard, assuming that all variables have function-level scope.
-
-The following listing gives a simple example of explicit loop-based parallelism
-(OpenMP 3.0 task-based parallelism form is also supported, the interested
-reader may refer to~\cite{pyhpc2013} for more details). the \texttt{kernel}
-function is pure and computation intensive, making the loop a perfect candidate
-for parallelization.
-
-\begin{lstlisting}
-def compute_julia(cr, ci, N, bound=1.5, lim=1000., cutoff=1e6):
-    julia = np.empty((N, N), np.uint32)
-    grid_x = np.linspace(-bound, bound, N)
-    #omp parallel for 
-    for i, x in enumerate(grid_x):
-        for j, y in enumerate(grid_x):
-            julia[i,j] = kernel(x, y, cr, ci, lim, cutoff)
-    return julia
-\end{lstlisting}
-
-As in in standard OpenMP, loop indices of the parallel loop are automatically
-made private, and Pythran extends the concept to any random access iterator,
-even after type destructuring. In the example, both \texttt{i} and \texttt{x}
-are private. A scope analysis is used to prove that \texttt{j} and \texttt{y}
-are also local to the loop. In the end, the whole computation achieves a
-perfect speedup of $\times4$
-
-Next section performs an in-depth comparison of Pythran with others Python
-optimizers: Cython and PyPy.
-
-\section{Benchmarks}
-\label{sec:benchmarks}
-
-All benchmarks presented in this section are run on an hyper-threaded quad-core
-i7, % FIXME sandy-brige, nehalem, what?
-using examples shipped along with
-Pythran sources, available at \url{https://github.com/serge-sans-paille/pythran} in
-the \texttt{pythran/test/cases} directory. The Pythran version used is the
-\texttt{HEAD} of the \texttt{iop2014} branch, PyPy 2.0.2 compiled with the
-\texttt{-jit} flag, CPython 2.7.3 and Cython 0.20. All timings are made using
-the \texttt{timeit} module, taking the best of all runs. All C++ codes are
-compiled with g++ 4.8.2, using the tool default compiler option, generally
-\texttt{-O3} plus a few optimizing flags depending on the target. To make a
-fair comparison, no parallelism annotations are added to any code.
-
-Provided each module has been properly compiled, each experiment is run using
-the following command from a shell:
-
-\begin{lstlisting}[language=sh]
-python -m timeit -s "${SETUP}" "${RUN}"
-\end{lstlisting}
-
-\noindent where \texttt{\$SETUP} and \texttt{\$RUN} depend on the benchmark and
-\texttt{pypy} is substituted to \texttt{python} when needed. The
-\texttt{\$SETUP} and \texttt{\$RUN} values are available to the reader on the
-Git repository in the \texttt{doc/papers/iop/xp/run\_benchamrks.sh} file with
-all the experimental setup.
-
-The same source is used for each benchmark, to the exception of Cython sources
-which needed a manual conversion. For these sources, bound checking and wrap
-around are always disable, all variables are declared using the proper C or C++
-type and all mathematical functions in use come from \texttt{libm}. Some calls
-to the \texttt{zeros\_like} function also have been replaced by calls to
-\texttt{zeros} for the PyPy version.
-
-\subsection{NQueens}
-
-``Nqueens'' is a benchmark extracted from the former Unladen
-Swallow\footnote{\url{http://code.google.com/p/unladen-swallow/}} project. It
-is particularly interesting as it makes an intensive use of non-trivial
-generator expressions and integer sets.
-
-Table~\ref{tbl:nqueen} illustrates the benchmark results for CPython, Cython,
-PyPy and Pythran.  It showcases the fact that Pythran benefits a lot from
-high-level transformations, for instance \emph{lazy evaluation}, which prevents
-the many allocation / deallocation triggered by Cython and PyPy runs.  A deeper
-look at the Pythran profiling trace shows that more than half of the execution
-time is spent allocating and deallocating a \texttt{set} used in the internal
-loop. There is a memory allocation invariant that could be taken advantage of
-there, but none of the compiler does so.
-
-\begin{table}
-    \centering
-
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
-    \hline
-    Timing  &  181ms   &   42.3ms     &    \textbf{22.5ms} &  51.3ms  \\
-    \hline
-    Speedup &  $\times$1         &    $\times$4.27      &    \textbf{$\times$8.04}   &  $\times$3.52    \\
-    \hline
-\end{tabular}
-\caption{Benchmarking result on the NQueen program.}
-\label{tbl:nqueen}
-
-\end{table}
-
-\subsection{Rosen Der}
-
-The ``rosen der'' kernel computes a complex numpy expression that accesses
-data through a one dimensional slice. It is interesting to evaluate how much of
-the expression can be compiled to regular, efficient loop while avoiding
-temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
-CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
-test case, but Cython code had to be manually transformed into an explicit
-loop. Moreover, activating automatic parallelization for Pythran yields an
-additional $\times4$ speedup without a change to the input, while Cython needs
-manual code annotation to achieve a similar result. It seems that PyPy support
-for numpy expression is still note mature for numpy expression, but there is a
-strong on-going effort of the community to fix this.
-
-\begin{table}
-    \centering
-
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
-    \hline
-    Timing  &  22.3ms   &   \textbf{3.39ms}     &    3.81ms &  51.3ms  \\
-    \hline
-    Speedup &  $\times$1         &    \textbf{$\times$6.57}      &    $\times$5.85   &  $\times$0.43    \\
-    \hline
-\end{tabular}
-\caption{Benchmarking result on the Rosen der program.}
-\label{tbl:rosen}
-
-\end{table}
-
-\subsection{Growcut}
-
-The ``growcut'' application is an image processing code where all array indexing
-are explicit. In that sense, this benchmark is complementary of the previous
-one, as it evaluates how well explicit array access are compiled.
-Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
-speedups on that benchmark, but this is is merely due to the fact that
-accessing individual elements is extremely slow in numpy, and the direct C
-translation of the Python code is very close to what a C developer would write.
-Pythran appears to outperforms Cython once all access are explicit, most
-certainly due to cleaner code generation.
-
-\begin{table}
-    \centering
-
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
-    \hline
-    Timing  &  398ms   &   991us     &    \textbf{526us} &  5.54ms  \\
-    \hline
-    Speedup &  $\times$1         &    $\times$401.6      &    $\times$756.7   &  $\times$71    \\
-    \hline
-\end{tabular}
-\caption{Benchmarking result on the growcut program.}
-\label{tbl:growcut}
-
-\end{table}
-
-\subsection{Blacksholes}
-
-The ``blacksholes'' application is a benchmarks that showcases the
-translation cost from the Python space to native space. Indeed we forced the
-use of a \texttt{list} as input of the kernel instead of a numpy array. As a
-consequence there is a costly translation step from a structure that contains
-an array of pointers to generic Python objects to a dense structure involved.
-Results are shown in Table~\ref{tbl:black}. They show that conversion code is
-less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
-it does not suffer from a similar conversion penalty.
-
-\begin{table}
-    \centering
-
-    \begin{tabular}{|l|c|c|c|c|}
-        \hline
-     Tool    &  CPython    &   Cython     &     Pythran   &  PyPy \\
-    \hline
-    Timing  &  490us   &   74.7us     &    \textbf{34.8us} &  37.3us  \\
-    \hline
-    Speedup &  $\times$1         &    $\times$6.55      &    $\times$14.1   &  $\times$13.1    \\
-    \hline
-\end{tabular}
-\caption{Benchmarking result on the blacksholes program.}
-\label{tbl:black}
-
-\end{table}
-
-\section{Future Work}
-
-Although Pythran focuses on a subset of Python and its standard library, many
-optimizations opportunities are still available. Using a Domain Specific
-Language (DSL) approach, one could use rewriting rules to optimize several
-Python idioms. For instance, \texttt{len(set(x))} could lead to an optimized
-\texttt{count\_uniq} that would iterate only once on the input sequence.
-
-There is naturally more work to be done at the Numpy level, for instance to
-support more functions from the original module. The extraction of Numpy
-expressions from \texttt{for loops} is also a natural optimization candidate, which
-shares similarities with code refactoring.
-
-Numpy expressions also fit perfectly well in the polyhedral model. Exploring
-the coupling of polyhedral tools with the code generated from Pythran offers
-enthusiastic perspectives.
-
-\section{Conclusion}
-
-This paper presents the Pythran compiler, a translator and an optimizer that
-converts Python into C++. Unlike existing static Python compilers, Pythran
-leverages several function-level or module-level analyses to provide
-generic or Python-centric code optimizations. Additionally, it uses a C++
-library that makes heavy usage of template programming to provide an efficient
-API similar to a subset of Python standard library. This library takes
-advantage of modern hardware capabilities ---vector instruction units and
-multiple-cores--- in its implementation of parts of the \texttt{numpy} package.
-
-This paper gives an overview of the compilation flow, the analyses involved and
-the optimizations used. It also compares the performance of compiled Pythran
-modules against CPython and other optimizers like Cython and PyPy.
-
-To conclude, limiting Python to a statically typed subset does not hinders the
-expressiveness when it comes to scientific or mathematic computations, but makes
-it possible to use a wide variety of classical optimizations to help Python
-match the performances of statically compiled language. Moreover, one can use
-high level information to generate efficient code that would be difficult to
-write for the average programmer.
-
-% FIXME (just thinking about it, did you say what CPyhton?)
-% PB : I think it is ok as audience is Python programmers
+%%
+%%
+\section*{Conclusion}
 
@@
 
-This project has been partially funded by the CARP Project and the SILKAN
-Company.  The authors would like to thank Yuancheng Peng and Eliott Coyac for
-their contributions to development of the Pythran compiler.
+The Pythran project is an independent Open Source initiative that has been
+partially funded by the CARP Project and the SILKAN Company.
+
+Several Students of Télécom Bretagne, namely Adrien Merlini, Alan Raynaud,
+Xavier Corbillon, Yuancheng Peng and Eliott Coyacc, have made significant
+contributions to the project.
 

### 2014-08-22 doc/papers/cgo2014/paper.tex
Improve Stackoverflow part

@@
 \section{A Crowd-Sourced Numpy Benchmark}
+
+One of the first steps required when designing an optimizing compiler is to
+gather enough test cases, or benchmarks, in order to drive the optimization
+process toward realistic examples. To achieve this goal, we have gathered a few
+exemple from existing compiler sources, but these may have been modified to
+overcome some compiler limitations. A less biased source of benchmarks should
+be independant from the compiler community and focused on matching user needs.
+To achieve that goal, we collected synthetic scientifc kernels from the
+stackoverflow website.
+
+StackOverflow~\footnote{\url{http://stackoverflow.com/}} is a question and
+answer site for programmers. Any registered user can submit questions and other
+users may answer. A voting system is then used to sort the answers. Said
+otherwise, the website provides a database of question and answers on
+programming topics, including Python.
+
+A query using the terms \emph{numpy} and \emph{slow} yields many results, many
+of which use the pattern ``Q: my code is slow'' followed by an upvoted answer
+``A: rewrite it that way'', where the original code performs explicit looping
+over an array, and the rewritten code uses the relevant combination of numpy
+function.
+
+For instance, question \url{http://stackoverflow.com/questions/7741878}
+proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
+row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
+the proposed answer is to use the more compact version presented in
+Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
+roughly achieves a $times 4$ speedup over the loop version.
+
+\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, float, label={lst:l2norm-loopy}]
+def slow(x):
+    r = np.empty(x.shape[0])
+    for i in xrange(x.shape[0]):
+        r[i] = np.sum(np.abs(x[i])**2)
+    return r
+\end{lstlisting}
+
+\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, float, label={lst:l2norm}]
+import numpy as np
+def l2norm(x):
+    return np.sum(np.abs(x)**2, axis=1)
+\end{lstlisting}
+
+Our guess is that new Numpy user are more likely to write loops and code
+similar to Listing~\ref{lst:l2norm-loopy}, but as their knowledge of the numpy
+API and of the good programming practice grows, they get to write code similar
+to Listing~\ref{lst:l2norm}. A direct consequence is that a compiler should
+focus on optimizing high-level constructs rather loops, both to encourage good
+practice with respect to Numpy code, and because that's the kind of code Numpy
+user ultimately write.
 

### 2014-08-25 doc/papers/cgo2014/paper.tex
Add section and ref
Typo corrections

@@
 
-    As more scientists use these scripting language, there is an on-going
+    As more scientists use these scripting languages, there is an on-going
     research effort to compile them into efficient native code. However the
-    computations kernels often differ from classical ones as these languages
-    favors an high-level coding style, where loops are often replaced by calls
-    to the relevant toolbox/module/library/package. As a consequence, classical
+    computation kernels often differ from the classical ones as these languages
+    favor a high-level coding style, where loops are often replaced by calls
+    to the relevant toolbox/module/library/package. As a consequence, traditional
     loop optimization techniques like loop fusion, tiling and related
@@
     Numpy package that provides core array data structure and basic linear
-    algebra routines. It first conducts a case study based on Numpy user
-    questions from the stackoverflow question and answer website, then studies
+    algebra routines. It first conducts a case study based on Numpy use
+    cases from the stackoverflow question and answer website, then studies
     existing compilers for numerical Python. Optimization opportunities,
@@
     the Pythran compiler are presented and illustrated through several
-    benchmarks, showing very interesting speedup over the standard Python
-    interpreter. Speedup of factor greater than 10 are achieved despite the
-    fact the considered benchmarks mostly call Numpy routines that run C code.
+    benchmarks, showing very interesting speedups over the standard Python
+    interpreter. Speedup greater than a factor of 10 are achieved despite the
+    fact that the considered benchmarks mostly call Numpy routines that run C code.
 
@@
 process toward realistic examples. To achieve this goal, we have gathered a few
-exemple from existing compiler sources, but these may have been modified to
+example from existing compilers' testsuite, but these may have been modified to
 overcome some compiler limitations. A less biased source of benchmarks should
@@
 
-For instance, question \url{http://stackoverflow.com/questions/7741878}
+For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
 proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
@@
 
-Our guess is that new Numpy user are more likely to write loops and code
+Our guess is that beginner Numpy users are more likely to make use of raw loops
 similar to Listing~\ref{lst:l2norm-loopy}, but as their knowledge of the numpy
 API and of the good programming practice grows, they get to write code similar
-to Listing~\ref{lst:l2norm}. A direct consequence is that a compiler should
-focus on optimizing high-level constructs rather loops, both to encourage good
-practice with respect to Numpy code, and because that's the kind of code Numpy
-user ultimately write.
+to Listing~\ref{lst:l2norm}. As a direct consequence, a compiler should
+focus on optimizing high-level constructs rather than raw loops. Both to encourage good
+practice with respect to Numpy code, and because this is the kind of code Numpy
+users ultimately write.
 
@@
 
+\subsection{Typing}
+
 \subsection{Expression Templates}
 
-\cite{expression_templates}
+\cite{expression_templates, et2012}
 

### 2014-08-25 doc/papers/cgo2014/paper.tex
Running example section

--- 2014-08-22 doc/papers/cgo2014/paper.tex
@@
 
-\subsection{Temporaries Elemination}
+Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapated
+from the Scipy source code for \texttt{scipy.optimize.rosen}. This kernel makes
+use of Numpy's \texttt{sum} function, Python's square notation and Numpy's
+array slicing. It is a good example of high-level Python kernel, altohough
+using the function \texttt{scipy.optimize.rosen} would naturally make sense.
+
+\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, float, label={lst:rosen}]
+def rosen(x):
+    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
+    t1 = (1 - x[:-1]) ** 2
+    return numpy.sum(t0 + t1)
+\end{lstlisting}
+
+This section goes through all the optimization opportunities in
+Listing~\ref{lst:rosen} as a showcase of what an optimizing compiler could do.
+
+\subsection{Temporaries Elimination}
+
+In Numpy, all point-to-point array operations allocate a new array that holds
+the computation result. This behavior is consistent with many Python standard
+module, but it is a very inefficient design choice, as it keeps on polluting
+the cache with potentially large fresh storage. In the rosen expression from
+Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
+create a temporary array but a view) to hold intermidiate steps. had the
+computation been lazy, then no temporary would have been needed.
 
@@
 
-\subsection{Loop Vectorization}
+As Numpy is a native library mostly written in C, each operator computation is
+performed by a function is implemented as a loop performing a single operation,
+while the operator chaining is done at the interpreter level. This a typical
+problem in library design: if only a small set of functions is provided, it
+prevents the optimization of multiple operators into a signle specialized
+operator, and provided many operator combinations as part of the library yields
+better performance to the price of API bloat. In that case, a loop is used for
+each temporary computation, plus an extra loop for the \texttt{numpy.sum}
+reduction, where a single loop would have been needed.
 
-\subsection{Loop Parallelization}
+\subsection{Loop Vectorization and Parallelization}
+
+Even if compiled with vectorization flags on, there would be very little
+benefit to generate SIMD instructions for the respective array operations used
+by each operator, as the memory load store would have dominated, especially as
+Numpy typically operates on double precision floats. In a similar manner,
+parallelization of each operator would yield terrible speedups because a
+synchronization fence would need to be issued between each temporary
+computation and the loop computation intensity would be very low compared to
+the memory pressure implied by two array reads and one array write for a single
+binary operator.
+
+On the opposite, if all computation were merged into a single loop, the whole
+loop would be triially vectorizable and parallelizable. No dependency checks
+are needed as parallelization and vectorization are deduced from the semantics
+of the array operations.
 

### 2014-08-25 doc/papers/cgo2014/paper.tex
Introduction

@@
 
-\cite{rossum97, scipy}
+The Python language~\cite{rossum97} now provides a full toolkit to carry out
+scientific experiments~\cite{scipy}, from core scientific routines with the
+Numpy package, to scientific packages with the Scipy package, plotting
+facilities with the Matplotlib package, enhanced terminal and notebooks with
+IPython. As a consequence, there has been a move from historical languages like
+Fortran to Python, as showcased by the success of the Scipy conference.
+
+As Python based scientific tools gets widely used, the question of High
+performance Computing naturally arises, and it is the focus of many recent
+research. Indeed, although there is a great gain in productivity when using
+these tools, there is also a performance gap that needs to be filled.
+
+This paper focuses on compilation techniques that are relevant for the
+optimization of high-level numerical kernels written in Python using the Numpy
+package. Section~\ref{sec:stackoverflow} presents a case study based on user
+input gathered on the stackoverflow question and answer site.
+Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
+showcases the optimization opportunities typically found in Numpy-based
+kernels. Existing compilation approach to optimize such kernels are discussed
+in Section~\ref{sec:compilers}, then Section~\ref{sec:pythran} emphasises on a
+compiler-runtime cooperative approach to solve the performance issues.
+Experimentations showcasing important performance gains are detailed in
+Section~\ref{sec:xp}.
 
@@
 \section{A Crowd-Sourced Numpy Benchmark}
+\label{sec:stackoverflow}
 
@@
 
+The benchmark is publicly available on
+\url{https://github.com/serge-sans-paille/numpy-benchmarks}.
+
 %%
@@
 \section{Optimizations Opportunities in a Typical Numpy Kernel}
+\label{sec:optimize}
 
@@
 \section{Existing Compilation Approach}
+\label{sec:compilers}
 
@@
 \section{Compiler - Runtime Cooperation}
+\label{sec:pythran}
 
@@
 \section{Experiments}
+\label{sec:xp}
 
@@
 
-\subsection{Expermimental Setup}
+\subsection{Experimental Setup}
+
+\subsection{Performance of the rosenbrock Function}
 

### 2014-08-26 doc/papers/cgo2014/paper.tex
Unifiy StackOverflow (camel) case

@@
     algebra routines. It first conducts a case study based on Numpy use
-    cases from the stackoverflow question and answer website, then studies
+    cases from the StackOverflow question and answer website, then studies
     existing compilers for numerical Python. Optimization opportunities,
@@
 package. Section~\ref{sec:stackoverflow} presents a case study based on user
-input gathered on the stackoverflow question and answer site.
+input gathered on the StackOverflow question and answer site.
 Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
@@
 To achieve that goal, we collected synthetic scientifc kernels from the
-stackoverflow website.
+StackOverflow website.
 

### 2014-08-26 doc/papers/cgo2014/paper.tex
Typo fixes and comments

@@
 
-Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapated
+Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapted
 from the Scipy source code for \texttt{scipy.optimize.rosen}. This kernel makes
 use of Numpy's \texttt{sum} function, Python's square notation and Numpy's
-array slicing. It is a good example of high-level Python kernel, altohough
+array slicing. It is a good example of high-level Python kernel, although
 using the function \texttt{scipy.optimize.rosen} would naturally make sense.
@@
 Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
-create a temporary array but a view) to hold intermidiate steps. had the
-computation been lazy, then no temporary would have been needed.
+create a temporary array but a view) to hold intermediate steps. Had the
+computation been lazy, no temporary would have been needed.
 
@@
 As Numpy is a native library mostly written in C, each operator computation is
-performed by a function is implemented as a loop performing a single operation,
-while the operator chaining is done at the interpreter level. This a typical
+performed by a function implemented as a loop performing a single operation,
+and the operator chaining is done at the interpreter level. This a typical
 problem in library design: if only a small set of functions is provided, it
@@
 Numpy typically operates on double precision floats. In a similar manner,
-parallelization of each operator would yield terrible speedups because a
+parallelization of each operator would yield a terrible speedup because a
 synchronization fence would need to be issued between each temporary
@@
 On the opposite, if all computation were merged into a single loop, the whole
-loop would be triially vectorizable and parallelizable. No dependency checks
-are needed as parallelization and vectorization are deduced from the semantics
-of the array operations.
+loop would be trivially vectorizable and parallelizable. 
+%% I don't get the logic in the previous sentence...
+No dependency check is needed as parallelization and vectorization are deduced 
+from the semantic of the array operations.
+%% you still need a dependency test in the absolute, don't you?
+%% I'm not sure about what you mean by "deduced"
+%% I guess the only reason that allows to get rid of any dependency test is that
+%% memory is never updated but new arrays are created?
+%% The question is to which point is it still true after "fusion"
 

### 2014-08-27 doc/papers/cgo2014/paper.tex
Import some description from Scipy article

@@
 
+The success of Python and the development of its large ecosystem triggered
+several attempts to improve its execution speed. The most significant basic 
+block for high performance numerical simulations is certainly Numpy, a 
+numerical library to manipulate dense arrays. Numerous routines are exposed in 
+Python and implemented in C, offering close-to-native performance in some cases.
+% FIXME: Numpy is part of this section but somehow covered earlier???
+% 
+However, this library-like approach has inherent limitations that were exposed 
+in the previous section. To address these limitations, several compiler-based 
+approaches were proposed.
+
 \subsection{Numpy}
 \cite{oliphant2007,numpyarray2011}
+
+% Not sure what to add here because of the previous section.
 
@@
 
-\subsection{Numba}
-\cite{numba}
+% 
+Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
+syntax with typing information, calls to native functions from third party
+libraries, and a limited set of parallelism constructs, such as the possibility
+to define parallel loops, but no task parallelism. When possible, it unboxes
+Python variables to improve performance. Without type annotations, the
+performance improvement is not terrific, but given enough type annotation,
+Cython can generates code that runs as fast as its C equivalent, while
+maintaining a syntax close to Python.
 
-\subsection{Parakeet}
-\cite{parakeet2012}
+\subsection{Numba and Parakeet}
+
+The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}~\cite{numba}
+compiler uses additional type information to generate sequential LLVM bytecode.
+Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
+Python to LLVM bytecode translation, but limits its scope to the numpy package,
+only supporting a small subset of the Python language. Additionally, it supports
+implicit parallelism using an implicit mapping between numpy functions and a set
+of parallel primitives including \texttt{map}s \texttt{scan}s and
+\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
+not suffer from backward-incompatibility issues. When meeting an unsupported
+construct, numba falls back to Python C-API calls, while parakeet raises an
+exception.
 
@@
 
+Copperhead~\cite{copperhead2011} is a functional, data parallel language
+embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
+structure and prohibits usage of many control-flow operators such as loops,
+enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
+intrinsics to exhibit parallelism. But it can be efficiently compiled to either
+CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
+\texttt{http://thrust.github.com/}} library. Python decorators are used to
+identify hot-spots that are JIT-compiled to native code.
+
 \subsection{PyPy}
 \cite{pypy2009}
+
+Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
+Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
+enhance Python performance. However Shed~Skin does not provide support for
+fine-grained parallelism beyond what the standard library proposes. PyPy is
+heading toward STM for parallelism support.
 

### 2014-08-27 doc/papers/cgo2014/paper.tex
Some info on expression templates

@@
 
-\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, float, label={lst:l2norm-loopy}]
+\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, label={lst:l2norm-loopy}]
 def slow(x):
@@
 
-\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, float, label={lst:l2norm}]
+\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, label={lst:l2norm}]
 import numpy as np
@@
 
-\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, float, label={lst:rosen}]
+\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, label={lst:rosen}]
 def rosen(x):
@@
 \subsection{Temporaries Elimination}
+\label{sec:temporaries-elimination}
 
@@
 \subsection{Operator Fusion}
+\label{sec:operator-fusion}
 
@@
 %%
-\section{Compiler - Runtime Cooperation}
+\section{Compilers Cooperation}
 \label{sec:pythran}
 
-\subsection{Typing}
+C/C++ have been at the hearth of high performance computing for many years.
+Compiler infrasturctures for these languages also have a long history and
+implement many optimizations on the control flow graph, basic blocks or
+loops~\cite{Aho2006} that limit usage of assembly code to a few hot spots.
+Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
+programming model and libraries like Boost.SIMD~\cite{boostsimd} provide good
+abstractions for explicit vectorization. Many highly tuned kernels are
+available in the form of thid-party libraies like GotoBlas~\cite{gotoblas2008}.
+
+There is no need to reinvent the wheel when compiling numerical Python kernels.
+A more profitable approach is to map Python concept to existing ones, or
+translate them into existing them. As a consequence, this section focuses on
+the translation of Python constructs to efficient C++ constructs. The choice of
+C++ instead of C is motivated by the opportunities offered by meta programming
+and the interesting similarities between C++11~\cite{isocxx11} features and
+Python constructs.
 
@@
 
-\cite{expression_templates, et2012}
+Many C++ linear algebra libraries~\cite{eigen,ublas} use \emph{expression
+templates}~\cite{expression_templates, et2012} as a way to efficently describe high
+level matrix expressions while avoid the creation of intermediate storage.
+Additionnaly, it has the same effect as loop fusion, solving the issues
+described in \S~\cite{sec:temporaries-elimination} and
+\S~\ref{sec:operator-fusion}.
+
+Thanks to variadic templates introduced in C++11, it is possible to write a
+generic C++ meta-class that implements the same API as numpy array and provides
+all unary and binary operators, as well as the various form of indexing and
+slicing. Listing~\ref{lst:rosen-cxx} illustrates the translation of the Python
+code from Listing~\ref{lst:rosen} into C++. Altough this version is more
+verbose than the Python one, the translation from one to the other does not
+raise any particular issue.
+
+\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
+constexpr size_t N = x::ndim;
+ndarray<N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
+ndarray<N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
+return numpy::sum{}((t0 + t1));
+\end{lstlisting}
 

### 2014-08-28 doc/papers/cgo2014/paper.tex
Add some perfs mesures
More info on forward substitution (not finished yet)

@@
 \subsection{Expression Templates}
+\label{sec:expression-templates}
 
@@
 
+This version does not supress all intermediate computations: typically, the
+assignments to \texttt{t0} and \texttt{t1} trigger the computation of the right
+hand side expression, thus creating two intermediates array. This usual
+limitation is motivated by the inability to prove that \texttt{x} is not
+modified before \texttt{t0} is read.
+
 \begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
+typedef typename x::dtype T;
 constexpr size_t N = x::ndim;
-ndarray<N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
-ndarray<N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
+ndarray<T, N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
+ndarray<T, N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
 return numpy::sum{}((t0 + t1));
@@
 \subsection{Forward Substitution}
+
+The limitation of expression templates presented in
+\S~\ref{sec:expression-templates} is easy to solve at the Python compiler
+level, using forward substitution. Typically, forward substitution is
+supported by the computation of a dependency graph that itself relies on
+read-write effects and alias analysis.
+
+Alias analysis is impossible for Python programs in general, because of late
+binding and possible monkey patching. In the context of a static compiler for
+numerical programs, it can be assumed that binding is done statically and no
+monkey patching is done.  In that context, it is relatively easy to design an
+intra-procedural alias analysis thanks to the absence of pointer arithmetic:
+the only language construction that can create aliasing is the assignment, thus
+a simple control-flow base alias analysis yields accurate results.
+
+The computation of read-write effects involve the annotation of numpy functions
+so that the Python compiler knows that most of them do not modify their
+argument, and the computation of argument effects for user functions. This can
+be done by considering each function individually and looking for the two kind
+of constructs that can modify a function argument: indexing and further
+function calls.
+
+To interprocedurally compute the read-write effects, intraprocedrual effects
+are first computed. Then we create a directed graph $G$ with as many nodes as
+functions. Each node contains a list of integer representing the indices of the
+formal argument that are modified intra-procedurally. Then we add an edge
+between two nodes, labeeld by an integer tuple $(i, j)$ if $i^\text{th}$ formal
+argument of the source is passed as $j^\text{th}$ effective argument of the
+sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
+writes $i$, an outcoming edge $(i,j)$ can be safely removed, as we know for
+sure that the $i^\text{th}$ parameter is written.  Similarly, an incoming edge
+$(j,i)$ can be removed and $j$ is added to the source's list. Iteratively
+applying this process removes all edges that can propagate a write, the only
+remaining edges propagate a read and can be ultmately removed.
 

### 2014-08-29 doc/papers/cgo2014/paper.tex
Fix citation

--- 2014-08-27 doc/papers/cgo2014/paper.tex
@@
 Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
-programming model and libraries like Boost.SIMD~\cite{boostsimd} provide good
-abstractions for explicit vectorization. Many highly tuned kernels are
+programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provide 
+good abstractions for explicit vectorization. Many highly tuned kernels are
 available in the form of thid-party libraies like GotoBlas~\cite{gotoblas2008}.
@@
 \label{sec:xp}
+
+% The first one holds 8 AMD Magny Cours processors for a
+% total of 48 cores. Each core has access to 64\,KB of L1 cache,
+% 512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
+% is shared between the 6 cores of a processor. This configuration
+% provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
+% up to SSE4. We
+% will refer to this configuration as \textbf{AMD48}.
+% On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
+% from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
+% from OpenMP improvement. This clang version is based on clang 3.4 release
+% (LLVM:233b1e3f034, clang-omp:6f358bcd87).
+
+% The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a ttal of
+% 8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
+% 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
+% is shared between the 8 processors. This configuration
+% provides a total of 63\,GB of main memory. It supports up to AVX. We will
+% refer to this configuration as \textbf{Intel8}.
+% On Intel8, backend compilers are gcc 4.9 (20140528) and the clang-omp version
+% from intel.
 

### 2014-08-29 doc/papers/cgo2014/paper.tex
Fix section reference
Improve forward substitution section

--- 2014-08-28 doc/papers/cgo2014/paper.tex
@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{listings}
 \usepackage{hyperref}
+\usepackage{subcaption}
 
@@
 \usetikzlibrary{shapes}
+\usetikzlibrary{calc}
 
@@
 \subsection{Forward Substitution}
+\label{sec:fs}
 
@@
 sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
-writes $i$, an outcoming edge $(i,j)$ can be safely removed, as we know for
+writes $i$, any outcoming edge $(i,j)$ can be safely removed, as we know for
 sure that the $i^\text{th}$ parameter is written.  Similarly, an incoming edge
@@
 
+Let us illustrate the algorithm on a simple case. Python code source and the
+initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
+only node wis write effects is \texttt{bar} that writes \texttt{n} through a
+subscript. Outcoming edges (1) and (2) can be removed because they are
+outcoming edges of \texttt{bar} starting by \texttt{n}. In coming edge (0) ends
+by \texttt{n} that is written by \texttt{bar} so it is possible to prune it
+while adding the start \texttt{n} to node \texttt{root}. No edge left, and we
+know that both \texttt{root} and \texttt{bar} update their argument, but not
+\texttt{foo}.
+
+\begin{figure}
+
+    \begin{subfigure}{.5\textwidth}
+        \begin{lstlisting}[language=python]
+def foo(s):
+  return s * 2
+
+def bar(n, m):
+  if m:
+    n[0] = bar(n, m / 2)
+  return foo(n)
+
+def root(r):
+  return bar(r, r[0])
+        \end{lstlisting}
+        \caption{Input Python program.}
+        \label{fig:argument-effects-code}
+    \end{subfigure}
+
+    \begin{subfigure}{.5\textwidth}
+        \centering
+        \begin{tikzpicture}[node/.style={rectangle, draw=black}]
+            \node[node]  (foo)  {$\texttt{foo} \rightarrow \varnothing$};
+            \node[node]  (bar) [above=of foo] {$\texttt{bar} \rightarrow \{\texttt{n}\}$};
+            \node[node]  (root) [above=of bar] {$\texttt{root} \rightarrow \{\texttt{r}\}$};
+
+            \draw[->] (bar) -- node[right] {$(\texttt{n,s})$} node[left] {(2)} (foo);
+            \draw[->] (bar.east) node[right, yshift=-1em] {(1)} -| ($(bar.east) + (.5,.75)$) node[right, yshift=-.8em] {$(\texttt{n,n})$} -| (bar.north) ;
+            \draw[->] (root) -- node[right, yshift=.8em] {$(\texttt{r,n})$} node[left, yshift=.8em] {(0)} (bar);
+        \end{tikzpicture}
+        \caption{Initial argument effect graph..}
+        \label{fig:argument-effects-graph}
+    \end{subfigure}
+
+    \caption{Illustration of the argument effect graph (\ref{fig:argument-effects-graph}) of a simple Python program~(\ref{fig:argument-effects-code}).}
+    \label{fig:argument-effects}
+
+\end{figure}
+
 \subsection{Range Analysis}
@@
 \subsection{Constant Propagation and Loop Unrolling}
+
+There are very few values that are considered as constant in Python, to the
+notable exception of strings and tuples. However thare are a lot of constant
+propagation opportunities, the most remarkable one being calls to the
+\texttt{range} builtin with literal parameters.
+
+A naïve translation of such call to a generic C++ function would generate a
+call to a C++ function that would allocate enough memory to hold three integer,
+fill the corresponding buffer and return it. A smarter version would generate a
+static array holding the relevant values. Figure~\ref{fig:unroll-range}
+compares two translation of a Python snippet that iterates over a fixed-size
+list. The Clang compiler successfully turns Listing~\ref{lst:stat} into a
+single constant. It also statically computes the return value for
+Listing~\ref{lst:dyna}, but still generates the intiialization code for the
+\texttt{std::vector}. The GCC compiler fails to statically compute the return
+value in both cases. This example shows that constant propagation should be
+done at a higher level than C++, and so be part of the Python optimizer.
+
+The argument effect analysis described in \S~\ref{sec:fs} can be extended to
+global variables to determine if a function as side effects on its argument or
+on global variables (including I/O). This information makes it possible to flag
+some function as pure, and to statically evaluate pure function calls with
+constant arguments. This evaluation is especially easy in Python thanks to the
+abitlity of the language to turn an AST expression into bytecode, evaluate it
+and retreive its constant value. For instance, this makes it possible to turn a
+call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
+value of \texttt{fibonnaci(4)} at compile-time.
+
+\begin{figure}
+
+    \begin{subfigure}{.5\textwidth}
+    \begin{lstlisting}[language=python]
+s = 0
+for i in [0,1,2]:
+    s += i
+return s
+\end{lstlisting}
+\caption{Python version.}
+\end{subfigure}
+
+    \begin{subfigure}{.5\textwidth}
+\begin{lstlisting}[language=c++]
+long s = 0;
+for(auto i: std::vector<long>{{0,1,2}}:
+    s += i;
+return s;
+\end{lstlisting}
+\caption{C++ version with dynamic allocation.}
+\label{lst:unroll-dyna}
+\end{subfigure}
+
+    \begin{subfigure}{.5\textwidth}
+\begin{lstlisting}[language=c++]
+long s = 0;
+for(auto i: std::array<long, 3>{{0,1,2}}:
+    s += i;
+return s;
+\end{lstlisting}
+\caption{C++ version with static allocation.}
+\label{lst:unroll-stat}
+\end{subfigure}
+
+    \caption{Different implementations of an iteration over a fixed size container.}
+    \label{fig:unroll-range}
+
+\end{figure}
+
+Once interprocedural constant evaluation is done (eventually followed by
+forward substitution), iteration over fixed size lists may appear, giving an
+opportunity to fully unroll these loops at the Python level, thus avoiding the
+suboptimal C++ translation observed for Listing~\ref{lst:unroll-stat}
+and~\ref{lst:unroll-dyna}.
 

### 2014-08-29 doc/papers/cgo2014/paper.tex

@@
 Additionnaly, it has the same effect as loop fusion, solving the issues
-described in \S~\cite{sec:temporaries-elimination} and
+described in \S~\ref{sec:temporaries-elimination} and
 \S~\ref{sec:operator-fusion}.

### 2014-08-31 doc/papers/cgo2014/paper.tex
CGO: Update section on existing work

@@
 \section{Introduction}
-% On aura pas le temps pour le papier je pense mais faire une comparaison avec du
-% C pure pour voir ce que les compilo de C arrive a faire.
+% On aura pas le temps pour le papier je pense mais faire une comparaison avec
+% du C pure pour voir ce que les compilo de C arrive a faire.
 
@@
 scientific experiments~\cite{scipy}, from core scientific routines with the
-Numpy package, to scientific packages with the Scipy package, plotting
-facilities with the Matplotlib package, enhanced terminal and notebooks with
-IPython. As a consequence, there has been a move from historical languages like
-Fortran to Python, as showcased by the success of the Scipy conference.
+Numpy package\cite{oliphant2007,numpyarray2011}, to scientific packages with
+the Scipy package, plotting facilities with the Matplotlib package, enhanced
+terminal and notebooks with IPython. As a consequence, there has been a move
+from historical languages like Fortran to Python, as showcased by the success
+of the Scipy conference.
 
@@
 block for high performance numerical simulations is certainly Numpy, a
-numerical library to manipulate dense arrays. Numerous routines are exposed in
-Python and implemented in C, offering close-to-native performance in some cases.
-% FIXME: Numpy is part of this section but somehow covered earlier???
-%
-However, this library-like approach has inherent limitations that were exposed
-in the previous section. To address these limitations, several compiler-based
-approaches were proposed.
+numerical library to manipulate dense arrays. However, this library-like
+approach has inherent limitations that were exposed in the previous section.
+To address these limitations, several compiler-based approaches were proposed.
 
-\subsection{Numpy}
-\cite{oliphant2007,numpyarray2011}
+\subsection{Numpy Limitations}
 
-% Not sure what to add here because of the previous section.
+Numpy programs typically makes use of high-level programming, mostly using
+array broadcasting techniques as shown listing~\ref{lst:pairwise}. This level
+of programming can be very efficiemt with Numpy when compared to regular Python
+code. However the main limitation is that the Numpy library approach comes with
+numerous temporary arrays allocated implicitely.
+
+\begin{lstlisting}[language=python, caption={\textit{pairwise} function that
+  exhibits array broadcasting.}, label={lst:pairwise}, breaklines=true]]
+def pairwise_numpy(X):
+    return np.sqrt(((X[:, None, :] - X) ** 2).sum(-1))
+\end{lstlisting}
+
+A solution to avoid temporary allocation can be to revert to a programming style
+using explicit raw loops as shown listing~/ref{lst:pairwise-raw}. Unfortunately
+this code runs almost 100 times slower due to the Python interpreter overhead.
+
+\begin{lstlisting}[language=python, caption={\textit{pairwise} function
+  using raw loops.}, label={lst:pairwise-raw}]
+def pairwise_python(X):
+    M = X.shape[0]
+    N = X.shape[1]
+    D = np.empty((M, M), dtype=np.float)
+    for i in range(M):
+        for j in range(M):
+            d = 0.0
+            for k in range(N):
+                tmp = X[i, k] - X[j, k]
+                d += tmp * tmp
+            D[i, j] = np.sqrt(d)
+    return D
+\end{lstlisting}
+
+\begin{lstlisting}[language=python]
+
+\end{lstlisting}
 
 \subsection{Cython}
-\cite{cython2010}
-
 %
@@
 performance improvement is not terrific, but given enough type annotation,
-Cython can generates code that runs as fast as its C equivalent, while
-maintaining a syntax close to Python.
+Cython can generates code that runs as fast as its C equivalent.
 
-\subsection{Numba and Parakeet}
+The drawback is that the programmer has to avoid \textit{Pythonic}-like
+high-level constructs and instead expresses his algorithm using raw loops.
+The \textit{pairwise} function rewritten for Cython is shown 
+listing~\ref{lst:pairwise-cython}.
+
+\begin{lstlisting}[language=python, caption={\textit{pairwise} function 
+  rewritten for Cython.}, label={lst:pairwise-cython}]
+%%cython
+import numpy as np
+cimport cython
+from libc.math cimport sqrt
+
+@cython.boundscheck(False)
+@cython.wraparound(False)
+def pairwise_cython(double[:, ::1] X):
+ cdef int M = X.shape[0]
+ cdef int N = X.shape[1]
+ cdef double tmp, d
+ cdef double[:, ::1] D
+ D = np.empty((M, M), dtype=np.float64)
+ for i in range(M):
+  for j in range(M):
+   d = 0.0
+    for k in range(N):
+     tmp = X[i, k] - X[j, k]
+     d += tmp * tmp
+    D[i, j] = sqrt(d)
+ return np.asarray(D)
+\end{lstlisting}
+
+\subsection{Numba}
 
@@
 compiler uses additional type information to generate sequential LLVM bytecode.
-Parakeet~\cite{parakeet2012} follows an approach similar to numba, that is
-Python to LLVM bytecode translation, but limits its scope to the numpy package,
+Numba uses Just-In-Time(JIT) compilation and do not suffer from
+backward-incompatibility issues. When meeting an unsupported construct, numba
+falls back to Python C-API calls.
+
+Unfortunately, at that time Numba is providing its best speedup on code with 
+raw loops, which is not encouraging programmers to express their code using 
+the higher-level construct available with Numpy.
+
+\subsection{Parakeet}
+
+Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is
+Python to LLVM bytecode translation, but limits its scope to the Numpy package,
 only supporting a small subset of the Python language. Additionally, it supports
-implicit parallelism using an implicit mapping between numpy functions and a set
+implicit parallelism using an implicit mapping between Numpy functions and a set
 of parallel primitives including \texttt{map}s \texttt{scan}s and
-\texttt{reduce}s. These two approaches use Just-In-Time(JIT) compilation and do
-not suffer from backward-incompatibility issues. When meeting an unsupported
-construct, numba falls back to Python C-API calls, while parakeet raises an
-exception.
+\texttt{reduce}s.
 
 \subsection{Copperhead}
-\cite{copperhead2011}
 
@@
 \subsection{PyPy}
-\cite{pypy2009}
 
-Tools such as PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
+PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
 Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to

### 2014-08-31 doc/papers/cgo2014/paper.tex
Fix undefined references

@@
 \caption{C++ version with dynamic allocation.}
-\label{lst:unroll-dyna}
+\label{lst:dyna}
 \end{subfigure}
@@
 \caption{C++ version with static allocation.}
-\label{lst:unroll-stat}
+\label{lst:stat}
 \end{subfigure}
@@
 opportunity to fully unroll these loops at the Python level, thus avoiding the
-suboptimal C++ translation observed for Listing~\ref{lst:unroll-stat}
-and~\ref{lst:unroll-dyna}.
+suboptimal C++ translation observed for Listing~\ref{lst:stat}
+and~\ref{lst:dyna}.
 

### 2014-08-31 doc/papers/cgo2014/paper.tex
Add OpenMP part.

@@
 \usepackage{subcaption}
+\usepackage{todo}
 
@@
 Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
-programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provide 
+programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provides
 good abstractions for explicit vectorization. Many highly tuned kernels are
@@
 raise any particular issue.
+
+Note that turning such Numpy expressions into expression templates is
+equivalent to converting from a strict evaluation to a non-strict, or lazy,
+evaluation. This is always legal in the contect of Numpy because none of the
+computations considered for expression template conversion have side effect.
 
@@
 
-\subsection{Vectorization and Parallelization}
+\subsection{Parallelization and Vectorization}
 
-\cite{esterie2012boost, pyhpc2013, wpmvp2014}
+A simplified version of the evaluation loop for an expression template is
+illustrated on Listing~\ref{lst:eval-et}. The loop over each dimension of the
+target array expression \texttt{to} is made explicit through a recursive
+template call. The computation of the origin expression is only triggered for
+the innermost loop.
+
+By construction of the expression template, the \texttt{eval} method used in
+\texttt{from.eval(i)} does not have side effects, and the Python to C++
+compiler can assert there is no shared memory between \texttt{to} and any
+component of \texttt{from}. In that case, all loops can be made parallel, and
+the innermost loop is eventually vectorizable.
+
+\begin{lstlisting}[language=c++, label={lst:eval-et}, caption={Evaluation body of an expression template}]
+template<size_t N> struct D {};
+
+// real evaluation loop
+template<class A, class E>
+void eval(A& to, E const& from, D<1>)
+{
+  size_t n = to.shape[A::ndims-1];
+  for(size_t i = 0; i < n; ++i)
+    to[i] = from.eval(i);
+}
+
+// recursive evaluation loops
+template<class A, class E, size_t Dim>
+void eval(A& to, E const& from, D<Dim>)
+{
+  size_t n = to.shape[A::ndims-Dim];
+  for(size_t i = 0; i < n; ++i)
+    eval(to[i], from[i], D<Dim-1>);
+}
+\end{lstlisting}
+
+\subsubsection{Parallelisation of Expression templates}
+
+A naive way to parallelize the parameteric loop nest from
+Listing~\ref{lst:eval-et} is to use a \texttt{parallel} directive from the
+OpenMP~\cite{openmp4} standard. Depending on the number of dimension of the
+input array, this may lead to nested parallel regions. When OpenMP nested
+parallel regions are activated, each parallel region spaws as many thread as
+possible, which leads to oversubscription and generally bad performances. When
+disabled, only the first parallel region spawns threads, which is likely to be
+inefficient depending on the input array shape, e.g. $(1,100000)$. Moreover,
+even if they do not start new threads, each nested parallel region implies a
+costly overhead. The \texttt{collapse} clause is not compatible with the
+parametric loop version.
+
+A first approach would be to use recursieve tasks and manual loop chunking. \todo{pierrick? c'est bête comme idée ?}
+
+A conservative approach is to only start a parallel region if the loop trip
+count is large enough. We avoid by using the \texttt{if} clause for parallel
+regions to do so in order to prevent the unnecessary overhead. Instead the loop
+statement is duplicated and guarded by the appropriate check, where only one
+loop holds the parallel annotations.
+
+Note that this parallelization scheme is compatible with the manual
+parallelization of Python programs proposed in~\cite{pyhpc2013}.
+
+\subsubsection{Vectorization of Expression Templates}
+
+A first exploration of vectorization of Python constructs is done in~\cite{wpmvp2014}.
 

### 2014-08-31 doc/papers/cgo2014/paper.tex
Pierrick review

@@
 \keywords
-static compilation, parallelization, Python, C++, SIMD
+static compilation, parallelization, Python, C++, SIMD, runtime
 
@@
 optimization of high-level numerical kernels written in Python using the Numpy
-package. Section~\ref{sec:stackoverflow} presents a case study based on user
+package. Its shows an implementation of these techniques through the Pythran
+compiler and provided a benchmarks testsuite for performance measures with
+others Python compilers.
+Section~\ref{sec:stackoverflow} presents a case study based on user
 input gathered on the StackOverflow question and answer site.
@@
 overcome some compiler limitations. A less biased source of benchmarks should
-be independant from the compiler community and focused on matching user needs.
-To achieve that goal, we collected synthetic scientifc kernels from the
+be independent from the compiler community and focused on matching user needs.
+To achieve that goal, we collected synthetic scientific kernels from the
 StackOverflow website.
@@
 module, but it is a very inefficient design choice, as it keeps on polluting
-the cache with potentially large fresh storage. In the rosen expression from
+the cache with potentially large fresh storage and add an extra
+allocation/disallocation costly operation. In the rosen expression from
 Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
@@
 problem in library design: if only a small set of functions is provided, it
-prevents the optimization of multiple operators into a signle specialized
+prevents the optimization of merging multiple operators into a single specialized
 operator, and provided many operator combinations as part of the library yields
-better performance to the price of API bloat. In that case, a loop is used for
+better performance to the price of API bloat. With a smamm set of functions, a loop is used for
 each temporary computation, plus an extra loop for the \texttt{numpy.sum}
-reduction, where a single loop would have been needed.
+reduction, where a single loop would have been needed with operator fusion.
 
@@
 
-On the opposite, if all computation were merged into a single loop, the whole
-loop would be trivially vectorizable and parallelizable.
-%% I don't get the logic in the previous sentence...
-No dependency check is needed as parallelization and vectorization are deduced
-from the semantic of the array operations.
-%% you still need a dependency test in the absolute, don't you?
-%% I'm not sure about what you mean by "deduced"
-%% I guess the only reason that allows to get rid of any dependency test is that
-%% memory is never updated but new arrays are created?
-%% The question is to which point is it still true after "fusion"
+At the opposite, if all computation were merged into a single loop using
+Temporaries Elimination and Operator fusion as described above, parallelization
+would be applicable as barrier between binary operation are removed and only
+one load and one store will be use for multiple operations.
 
@@
 array broadcasting techniques as shown listing~\ref{lst:pairwise}. This level
-of programming can be very efficiemt with Numpy when compared to regular Python
+of programming can be very efficient with Numpy when compared to regular Python
 code. However the main limitation is that the Numpy library approach comes with
-numerous temporary arrays allocated implicitely.
+numerous temporary arrays implicitly allocated.
 
@@
 libraries, and a limited set of parallelism constructs, such as the possibility
-to define parallel loops, but no task parallelism. When possible, it unboxes
+to define parallel loops, but no task parallelism. When possible, it unboxes % ref?
 Python variables to improve performance. Without type annotations, the
@@
 of parallel primitives including \texttt{map}s \texttt{scan}s and
-\texttt{reduce}s.
+\texttt{reduce}s. As it applies optimizations like operator fusion, it
+encourages programmer to express their code using the higher-level construct
+available with Numpy.
 
@@
 \texttt{http://thrust.github.com/}} library. Python decorators are used to
-identify hot-spots that are JIT-compiled to native code.
+identify hot-spots that are JIT-compiled to native code. As it requires to write
+functional code, it needs code modifications of \textit{Pythonic} code
 
-\subsection{PyPy}
+\subsection{PyPy and Shed~Skin}
 
@@
 fine-grained parallelism beyond what the standard library proposes. PyPy is
-heading toward STM for parallelism support.
+heading toward STM for parallelism support. As it supports most of the python
+construct, python code can be directly use with performance gain but not all
+performance optimizations can be applied.
+% STM?
 
@@
 C/C++ have been at the hearth of high performance computing for many years.
-Compiler infrasturctures for these languages also have a long history and
+Compiler infrastructures for these languages also have a long history and
 implement many optimizations on the control flow graph, basic blocks or
 loops~\cite{Aho2006} that limit usage of assembly code to a few hot spots.
+% You are speaking about user written assembly code? Or generated code?
 Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
@@
 good abstractions for explicit vectorization. Many highly tuned kernels are
-available in the form of thid-party libraies like GotoBlas~\cite{gotoblas2008}.
+available in the form of third-party libraries like GotoBlas~\cite{gotoblas2008}.
+% pas de nt2?
 
@@
 Many C++ linear algebra libraries~\cite{eigen,ublas} use \emph{expression
-templates}~\cite{expression_templates, et2012} as a way to efficently describe high
+templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
 level matrix expressions while avoid the creation of intermediate storage.
-Additionnaly, it has the same effect as loop fusion, solving the issues
+Additionally, it has the same effect as loop fusion, solving the issues
 described in \S~\ref{sec:temporaries-elimination} and
@@
 slicing. Listing~\ref{lst:rosen-cxx} illustrates the translation of the Python
-code from Listing~\ref{lst:rosen} into C++. Altough this version is more
+code from Listing~\ref{lst:rosen} into C++. Although this version is more
 verbose than the Python one, the translation from one to the other does not
@@
 equivalent to converting from a strict evaluation to a non-strict, or lazy,
-evaluation. This is always legal in the contect of Numpy because none of the
+evaluation. This is always legal in the context of Numpy because none of the
 computations considered for expression template conversion have side effect.
 
-This version does not supress all intermediate computations: typically, the
+This version does not suppress all intermediate computations: typically, the
 assignments to \texttt{t0} and \texttt{t1} trigger the computation of the right
@@
 binding and possible monkey patching. In the context of a static compiler for
+% What is monkey patching?
 numerical programs, it can be assumed that binding is done statically and no
@@
 so that the Python compiler knows that most of them do not modify their
-argument, and the computation of argument effects for user functions. This can
+arguments, and the computation of argument effects for user functions. This can
 be done by considering each function individually and looking for the two kind
@@
 formal argument that are modified intra-procedurally. Then we add an edge
-between two nodes, labeeld by an integer tuple $(i, j)$ if $i^\text{th}$ formal
+between two nodes, labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
 argument of the source is passed as $j^\text{th}$ effective argument of the
 sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
-writes $i$, any outcoming edge $(i,j)$ can be safely removed, as we know for
+writes $i$, any outcoming edge $(i, j)$ can be safely removed, as we know for
 sure that the $i^\text{th}$ parameter is written. Similarly, an incoming edge
-$(j,i)$ can be removed and $j$ is added to the source's list. Iteratively
+$(j, i)$ can be removed and $j$ is added to the source's list. Iteratively
 applying this process removes all edges that can propagate a write, the only
-remaining edges propagate a read and can be ultmately removed.
+remaining edges propagate a read and can be ultimately removed.
 
@@
 initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
-only node wis write effects is \texttt{bar} that writes \texttt{n} through a
+only node with write effects is \texttt{bar} that writes \texttt{n} through a
 subscript. Outcoming edges (1) and (2) can be removed because they are
@@
 while adding the start \texttt{n} to node \texttt{root}. No edge left, and we
-know that both \texttt{root} and \texttt{bar} update their argument, but not
-\texttt{foo}.
+know that both \texttt{root} update \texttt{r} and \texttt{bar} update \texttt{n},
+but \texttt{foo} doesn't update its argument.
 
@@
 
+Thanks to these analysis, we are able to say that \texttt{t0} and \texttt{t1}
+can be forwarded in the Listing~\ref{lst:rosen}.
+
 \subsection{Range Analysis}
@@
 
+% Cette partie n'a pas déjà été écrite dans un autre papier?
 There are very few values that are considered as constant in Python, to the
-notable exception of strings and tuples. However thare are a lot of constant
+notable exception of strings and tuples. However there are a lot of constant
 propagation opportunities, the most remarkable one being calls to the
@@
 single constant. It also statically computes the return value for
-Listing~\ref{lst:dyna}, but still generates the intiialization code for the
+Listing~\ref{lst:dyna}, but still generates the initialization code for the
 \texttt{std::vector}. The GCC compiler fails to statically compute the return
@@
 constant arguments. This evaluation is especially easy in Python thanks to the
-abitlity of the language to turn an AST expression into bytecode, evaluate it
-and retreive its constant value. For instance, this makes it possible to turn a
+ability of the language to turn an AST expression into bytecode, evaluate it
+and retrieve its constant value. For instance, this makes it possible to turn a
 call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
@@
 \end{subfigure}
+% Hum, on dirait du pattern transform : list => tuple pour l'iterable du for
 
@@
 
-A naive way to parallelize the parameteric loop nest from
+A naive way to parallelize the parametric loop nest from
 Listing~\ref{lst:eval-et} is to use a \texttt{parallel} directive from the
@@
 input array, this may lead to nested parallel regions. When OpenMP nested
-parallel regions are activated, each parallel region spaws as many thread as
-possible, which leads to oversubscription and generally bad performances. When
+parallel regions are activated, each parallel region spawns as many thread as
+possible up to the runtime upper bound, which leads to oversubscription with 
+more common implementation (libgomp and libiomp) and generally bad performances. When
 disabled, only the first parallel region spawns threads, which is likely to be
@@
 even if they do not start new threads, each nested parallel region implies a
-costly overhead. The \texttt{collapse} clause is not compatible with the
-parametric loop version.
+costly overhead for scheduling computation and datasharing. An easy way to solve
+the issue with small then large dimension is to use the \texttt{collapse}
+clause but it is not compatible with the parametric loop version.
 
-A first approach would be to use recursieve tasks and manual loop chunking. \todo{pierrick? c'est bête comme idée ?}
+A first approach would be to use recursive tasks and manual loop chunking. \todo{pierrick? c'est bête comme idée ?}
+% Les taches sont en général plus lente que les boucles for (malheureusement)
+% et il y a des fois ou on ne pourra jamais être plus rapide si on passe par
+% des taches.
+% Souvent, c'est a cause du scheduling static qui permet d'optimiser le cache
+% et aussi parce que les compilos, quand ils voient un omp for static, ils
+% font directement du déroulage de boucle. (Jusqu'au full unrolling des fois
+% avec autant de tache que de thread). 
+%
+% Dans notre cas, on ne connait pas forcement la taille de l'espace d'iteration
+% a la compilation. On ne pourra donc pas faire trop de unrolling mais bon, on
+% doit pouvoir faire comme gcc. Par contre, pour l'optimisation du cache, les
+% boucles for assurent que c'est le même noeud qui exécute le même "chunk" entre
+% les boucles for et ça, on le perd avec les taches. Je ne sais pas si ça joue
+% vraiment.
+% Autre problem. On va créé des taches en assez grande quantité. Ce que libgomp
+% supporte assez mal. Aussi, on va créé du datasharing pour des petites taches
+% (iteration de longueur 1 dans le cas du (1, 1000000)) ce qui est couteux et
+% assez mal supporté par libgomp. Il faudra voir comment ça se passe avec libiomp
+% qui est utilisé par clang.
+%
+% Au final : Ca doit être jouable. Il y a plusieurs raisons qui peuvent faire
+% qu'on a de mauvaise perfs mais sans mesures, on ne sait pas.
 
@@
 loop holds the parallel annotations.
+% usuellement appelé : manual if
 

### 2014-09-01 doc/papers/cgo2014/paper.tex
CGO: review pages 1-3

@@
     As more scientists use these scripting languages, there is an on-going
-    research effort to compile them into efficient native code. However the
-    computation kernels often differ from the classical ones as these languages
-    favor a high-level coding style, where loops are often replaced by calls
-    to the relevant toolbox/module/library/package. As a consequence, traditional
-    loop optimization techniques like loop fusion, tiling and related
+    research effort to compile them into efficient native code. However as 
+    these languages favor a high-level coding style, the computation kernels are
+    different from what we use to see in traditional C or Fortran programs:
+    explicit raw loops are replaced by high-level constructs and calls to the 
+    relevant toolbox/module/library/package. As a consequence, traditional
+    loop optimization techniques like fusion, tiling and related
     vectorization or parallelization techniques are not directly applicable.
@@
 
-The Python language~\cite{rossum97} now provides a full toolkit to carry out
-scientific experiments~\cite{scipy}, from core scientific routines with the
-Numpy package\cite{oliphant2007,numpyarray2011}, to scientific packages with
-the Scipy package, plotting facilities with the Matplotlib package, enhanced
-terminal and notebooks with IPython. As a consequence, there has been a move
-from historical languages like Fortran to Python, as showcased by the success
-of the Scipy conference.
+The Python language~\cite{rossum97} has a rich ecosystem that now provides a 
+full toolkit to carry out scientific experiments, from core scientific routines 
+with the Numpy package\cite{oliphant2007,numpyarray2011}, to scientific 
+packages with the Scipy package~\cite{scipy}, plotting facilities with the 
+Matplotlib package, enhanced terminal and notebooks with IPython. As a 
+consequence, there has been a move from historical languages like Fortran to 
+Python, as showcased by the success of the Scipy conference.
 
@@
 ``A: rewrite it that way'', where the original code performs explicit looping
-over an array, and the rewritten code uses the relevant combination of numpy
+over an array, and the rewritten code uses the relevant combination of Numpy
 function.
@@
 Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
-roughly achieves a $times 4$ speedup over the loop version.
+roughly achieves a $\times 4$ speedup over the loop version.
 
-\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in numpy.}, label={lst:l2norm-loopy}]
+\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in Numpy.}, label={lst:l2norm-loopy}]
 def slow(x):
@@
 
-\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in numpy.}, label={lst:l2norm}]
+\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in Numpy.}, label={lst:l2norm}]
 import numpy as np
@@
 
-Our guess is that beginner Numpy users are more likely to make use of raw loops
-similar to Listing~\ref{lst:l2norm-loopy}, but as their knowledge of the numpy
-API and of the good programming practice grows, they get to write code similar
-to Listing~\ref{lst:l2norm}. As a direct consequence, a compiler should
-focus on optimizing high-level constructs rather than raw loops. Both to encourage good
-practice with respect to Numpy code, and because this is the kind of code Numpy
-users ultimately write.
+Newcommers to Python with a prior C or Fortran experience are likely to make 
+use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their 
+knowledge of the Numpy API and of the good programming practice grows, they get to write code similar
+to Listing~\ref{lst:l2norm}. As a direct consequence, 
+\todo{Serge: don't you think we should also says that high-level constructs 
+enable better optimizations for the compiler?}
+a compiler should focus on optimizing high-level constructs rather than raw 
+loops. Both to encourage good practice with respect to Numpy code, and because 
+this is the kind of code Numpy users ultimately write.
 
@@
 
-\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, label={lst:rosen}]
+\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy}, label={lst:rosen}]
 def rosen(x):
@@
 the cache with potentially large fresh storage and add an extra
-allocation/disallocation costly operation. In the rosen expression from
+allocation/disallocation costly operation. In the \textit{rosen} function from
 Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
@@
 operator, and provided many operator combinations as part of the library yields
-better performance to the price of API bloat. With a smamm set of functions, a loop is used for
+better performance to the price of API bloat. The Listing~\ref{lst:rosen} 
+illustrates the use of a small set of functions: a loop is used for
 each temporary computation, plus an extra loop for the \texttt{numpy.sum}
@@
 
-Even if compiled with vectorization flags on, there would be very little
-benefit to generate SIMD instructions for the respective array operations used
-by each operator, as the memory load store would have dominated, especially as
-Numpy typically operates on double precision floats. In a similar manner,
-parallelization of each operator would yield a terrible speedup because a
-synchronization fence would need to be issued between each temporary
-computation and the loop computation intensity would be very low compared to
+Without operator fusion, there would be very little benefit to generate SIMD 
+instructions for the respective array operations used by each operator, as the
+memory load store would have dominated. This is even more important as
+Numpy typically operates on double precision floats. Parallelization also 
+suffers from the lack of operator fusion: a synchronization fence is needed
+between each temporary computation. 
+Hence the loop computation intensity would be very low compared to
 the memory pressure implied by two array reads and one array write for a single
@@
 
-At the opposite, if all computation were merged into a single loop using
-Temporaries Elimination and Operator fusion as described above, parallelization
-would be applicable as barrier between binary operation are removed and only
-one load and one store will be use for multiple operations.
+On the opposite, if all computations were merged into a single loop using
+temporaries elimination and operator fusion, parallelization would be more
+effective as barrier between binary operations are no longer needed and loads
+and stores will be used to perform multiple operations.
 
@@
 
-The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}~\cite{numba}
+The numba 
 compiler uses additional type information to generate sequential LLVM bytecode.
@@
 implicit parallelism using an implicit mapping between Numpy functions and a set
-of parallel primitives including \texttt{map}s \texttt{scan}s and
-\texttt{reduce}s. As it applies optimizations like operator fusion, it
+of parallel primitives including \texttt{map}, \texttt{scan}, and
+\texttt{reduce}. As it applies optimizations such as operator fusion, it
 encourages programmer to express their code using the higher-level construct
@@
 Copperhead~\cite{copperhead2011} is a functional, data parallel language
-embedded in Python. It uses $n$-uplets, NumPy arrays and lists as its core data
+embedded in Python. It uses $n$-uplets, Numpy arrays and lists as its core data
 structure and prohibits usage of many control-flow operators such as loops,
@@
 Thanks to variadic templates introduced in C++11, it is possible to write a
-generic C++ meta-class that implements the same API as numpy array and provides
+generic C++ meta-class that implements the same API as Numpy array and provides
 all unary and binary operators, as well as the various form of indexing and
@@
 
-\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
+\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the Rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
 typedef typename x::dtype T;
@@
 
-The computation of read-write effects involve the annotation of numpy functions
+The computation of read-write effects involve the annotation of Numpy functions
 so that the Python compiler knows that most of them do not modify their
@@
 
-A first approach would be to use recursive tasks and manual loop chunking. \todo{pierrick? c'est bête comme idée ?}
+A first approach would be to use recursive tasks and manual loop chunking.
+\todo{pierrick ? c'est bête comme idée ?}
 % Les taches sont en général plus lente que les boucles for (malheureusement)
@@
 
-\subsection{Performance of the rosenbrock Function}
+\subsection{Performance of the Rosenbrock Function}
 

### 2014-09-01 doc/papers/cgo2014/paper.tex
Proof read the paper once more. Lot of work left!

@@
 package. Its shows an implementation of these techniques through the Pythran
-compiler and provided a benchmarks testsuite for performance measures with
+compiler and provided a benchmarks test suite for performance measures with
 others Python compilers.
@@
 process toward realistic examples. To achieve this goal, we have gathered a few
-example from existing compilers' testsuite, but these may have been modified to
+examples from existing compilers' test suite, but these may have been modified to
 overcome some compiler limitations. A less biased source of benchmarks should
@@
 
-A query using the terms \emph{numpy} and \emph{slow} yields many results, many
+A query using the terms \emph{numpy} and \emph{slow} yields a lot of results, many
 of which use the pattern ``Q: my code is slow'' followed by an upvoted answer
@@
 
-Newcommers to Python with a prior C or Fortran experience are likely to make 
+Newcomers to Python with a prior C or Fortran experience are likely to make
 use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
-knowledge of the Numpy API and of the good programming practice grows, they get to write code similar
-to Listing~\ref{lst:l2norm}. As a direct consequence, 
-\todo{Serge: don't you think we should also says that high-level constructs 
-enable better optimizations for the compiler?}
-a compiler should focus on optimizing high-level constructs rather than raw 
+command of the Numpy API and of good programming practices grows, they get to write code similar
+to Listing~\ref{lst:l2norm}.
+The fact that in Numpy, high-level constructs run faster is also a strong motivation.
+As a direct consequence, a compiler should focus on optimizing high-level constructs rather than raw
 loops. Both to encourage good practice with respect to Numpy code, and because
@@
 
-The benchmark is publicly available on
-\url{https://github.com/serge-sans-paille/numpy-benchmarks}.
+The benchmarks collected during this process are publicly available on
+\url{https://github.com/serge-sans-paille/numpy-benchmarks}, with a set of
+benchmarking routine specialized for major Python/Numpy compilers.
 
@@
 Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapted
-from the Scipy source code for \texttt{scipy.optimize.rosen}. This kernel makes
+from the Scipy source code of \texttt{scipy.optimize.rosen}. This kernel makes
 use of Numpy's \texttt{sum} function, Python's square notation and Numpy's
 array slicing. It is a good example of high-level Python kernel, although
-using the function \texttt{scipy.optimize.rosen} would naturally make sense.
+using the function \texttt{scipy.optimize.rosen} directly would naturally make sense.
 
-\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy}, label={lst:rosen}]
+\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy.}, label={lst:rosen}]
 def rosen(x):
@@
 the cache with potentially large fresh storage and add an extra
-allocation/disallocation costly operation. In the \textit{rosen} function from
+allocation/deallocation costly operation. In the \textit{rosen} function from
 Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
@@
 instructions for the respective array operations used by each operator, as the
-memory load store would have dominated. This is even more important as
-Numpy typically operates on double precision floats. Parallelization also 
-suffers from the lack of operator fusion: a synchronization fence is needed
-between each temporary computation. 
-Hence the loop computation intensity would be very low compared to
-the memory pressure implied by two array reads and one array write for a single
-binary operator.
+memory load and store would have dominated the execution time. This is even more important as
+Numpy typically operates on double precision floats, which means only two (SSE) to four (AVX) scalar per vector registers.
+
+Parallelization also suffers from the lack of operator fusion: a
+synchronization fence is needed between each temporary computation.  Hence the
+loop computation intensity would be very low compared to the memory pressure
+implied by two array reads and one array write for a single binary operator.
 
@@
 numerical library to manipulate dense arrays. However, this library-like
-approach has inherent limitations that were exposed in the previous section.
-To address these limitations, several compiler-based approaches were proposed.
+approach has inherent limitations that have been exposed in \S~\ref{sec:optimize}.
+To address these limitations, several compiler-based approaches have been proposed.
 
-\subsection{Numpy Limitations}
+\subsection{Numpy}
 
-Numpy programs typically makes use of high-level programming, mostly using
-array broadcasting techniques as shown listing~\ref{lst:pairwise}. This level
-of programming can be very efficient with Numpy when compared to regular Python
-code. However the main limitation is that the Numpy library approach comes with
-numerous temporary arrays implicitly allocated.
+The reference implementation of Numpy is a native Module, written mostly in C.
+It uses the BLAS API whenever possible and provides a relatively efficient array abstraction in the form of the \texttt{ndarray} data structure.
+This data structures makes it possible to use various styles of slicing and broadcasting, as illustrated in Listing~\ref{lst:pairwise}.
+
+It also enforces a high-level programming style but it's very inefficient when
+explicit subscripts are used.  For instance, raw loops are used in
+Listing~\ref{lst:pairwise-raw}. It avoids the creation of temporary arrays but
+runs almost 100 times slower due to the Python interpreter overhead and the C to Python conversion involved in every subscript.
 
@@
 \end{lstlisting}
-
-A solution to avoid temporary allocation can be to revert to a programming style
-using explicit raw loops as shown listing~/ref{lst:pairwise-raw}. Unfortunately
-this code runs almost 100 times slower due to the Python interpreter overhead.
 
@@
 
-\begin{lstlisting}[language=python]
-
-\end{lstlisting}
-
 \subsection{Cython}
@@
 Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
-syntax with typing information, calls to native functions from third party
+syntax with typing information, calls to native functions from third party C
 libraries, and a limited set of parallelism constructs, such as the possibility
-to define parallel loops, but no task parallelism. When possible, it unboxes % ref?
-Python variables to improve performance. Without type annotations, the
-performance improvement is not terrific, but given enough type annotation,
+to define parallel loops, but no task parallelism. When possible, it unboxes %
+ref?  Python variables to improve performance. Without type annotations, the
+performance improvement is not significant, but given enough type annotation,
 Cython can generates code that runs as fast as its C equivalent.
 
-The drawback is that the programmer has to avoid \textit{Pythonic}-like
-high-level constructs and instead expresses his algorithm using raw loops.
-The \textit{pairwise} function rewritten for Cython is shown 
-listing~\ref{lst:pairwise-cython}.
+The drawback is that the programmer has to avoid high-level constructs and
+instead expresses his algorithm using raw loops.  The \textit{pairwise}
+function rewritten for Cython is shown listing~\ref{lst:pairwise-cython},
+showing usage of compiler annotations in the form of function decorators.
 
@@
 
-The numba 
-compiler uses additional type information to generate sequential LLVM bytecode.
-Numba uses Just-In-Time(JIT) compilation and do not suffer from
-backward-incompatibility issues. When meeting an unsupported construct, numba
-falls back to Python C-API calls.
+The Numba compiler uses additional type information collected at run-time to
+generate sequential LLVM bytecode.  Numba uses Just-In-Time(JIT) compilation
+and does not suffer from backward-incompatibility issues: a single
+\texttt{@jit} decorator is used to flag kernels to optimize. When meeting an
+unsupported construct, Numba silently falls back to Python C-API calls.
 
-Unfortunately, at that time Numba is providing its best speedup on code with 
-raw loops, which is not encouraging programmers to express their code using 
-the higher-level construct available with Numpy.
+Like Python, Numba is providing its best speedup on code with raw loops, which
+is not encouraging programmers to express their code using the higher-level
+construct available with Numpy. However, when input constraint are met, it
+exhibits performance close to Cython, or even better when conversion to GPU
+code is possible.
 
@@
 
-Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is
-Python to LLVM bytecode translation, but limits its scope to the Numpy package,
-only supporting a small subset of the Python language. Additionally, it supports
-implicit parallelism using an implicit mapping between Numpy functions and a set
-of parallel primitives including \texttt{map}, \texttt{scan}, and
-\texttt{reduce}. As it applies optimizations such as operator fusion, it
-encourages programmer to express their code using the higher-level construct
-available with Numpy.
+Parakeet~\cite{parakeet2012} follows an approach similar to Numba, that is JIT
+compilation of selected kernels. It uses C annotated with OpenMP directives as
+a target language, but limits its scope to the Numpy package, only supporting
+the subset of the Python language required to use Numpy functions.
+Additionally, it supports implicit parallelism using an implicit mapping
+between Numpy functions and a set of parallel primitives including
+\texttt{map}, \texttt{scan}, and \texttt{reduce}.  As it applies optimizations
+such as operator fusion, it encourages programmer to express their code using
+the higher-level construct available with Numpy.
 
@@
 CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.
-\texttt{http://thrust.github.com/}} library. Python decorators are used to
-identify hot-spots that are JIT-compiled to native code. As it requires to write
-functional code, it needs code modifications of \textit{Pythonic} code
+\texttt{http://thrust.github.com/}} library. Like Numba and Parakeet, it uses
+Python decorators to identify hot-spots that are JIT-compiled to native code.
+As it requires to write functional code, it is not compatible with the Numpy
+interface, although several Numpy functions could be rewritten in terms of
+Copperhead intrinsics.
 
@@
 Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
-enhance Python performance. However Shed~Skin does not provide support for
-fine-grained parallelism beyond what the standard library proposes. PyPy is
-heading toward STM for parallelism support. As it supports most of the python
-construct, python code can be directly use with performance gain but not all
-performance optimizations can be applied.
-% STM?
+enhance Python performance. However Shed~Skin does not provide support for the
+Numpy module. The is an important work in progress from the PyPy community to
+support Numpy, but it implies a partial rewrite of the Numpy API in RPython,
+the language used by PyPy, to enable PyPy's optimization.
 
@@
 
-C/C++ have been at the hearth of high performance computing for many years.
+Fortran/C/C++ have been at the hearth of high performance computing for many years.
 Compiler infrastructures for these languages also have a long history and
 implement many optimizations on the control flow graph, basic blocks or
-loops~\cite{Aho2006} that limit usage of assembly code to a few hot spots.
+loops~\cite{Aho2006} that limit usage of hand-written assembly to a few hot spots.
 % You are speaking about user written assembly code? Or generated code?
@@
 Alias analysis is impossible for Python programs in general, because of late
-binding and possible monkey patching. In the context of a static compiler for
-% What is monkey patching?
+binding and possible monkey patching---a way to modify the code of dynamic languages at run-time. In the context of a static compiler for
 numerical programs, it can be assumed that binding is done statically and no
@@
 
-To interprocedurally compute the read-write effects, intraprocedrual effects
+To inter procedurally compute the read-write effects, intra procedural effects
 are first computed. Then we create a directed graph $G$ with as many nodes as
@@
 % Cette partie n'a pas déjà été écrite dans un autre papier?
+%SG: non
 There are very few values that are considered as constant in Python, to the
@@
 propagation opportunities, the most remarkable one being calls to the
-\texttt{range} builtin with literal parameters.
+\texttt{range} built-in with literal parameters.
 
@@
 
-Once interprocedural constant evaluation is done (eventually followed by
+Once inter procedural constant evaluation is done (eventually followed by
 forward substitution), iteration over fixed size lists may appear, giving an
@@
 
-\subsubsection{Parallelisation of Expression templates}
+\subsubsection{Parallelization of Expression templates}
 
@@
 even if they do not start new threads, each nested parallel region implies a
-costly overhead for scheduling computation and datasharing. An easy way to solve
+costly overhead for scheduling computation and data sharing. An easy way to solve
 the issue with small then large dimension is to use the \texttt{collapse}
@@
 count is large enough. We avoid by using the \texttt{if} clause for parallel
-regions to do so in order to prevent the unnecessary overhead. Instead the loop
-statement is duplicated and guarded by the appropriate check, where only one
-loop holds the parallel annotations.
-% usuellement appelé : manual if
+regions to do so in order to prevent the unnecessary overhead. Instead a manual
+if is used: the loop statement is duplicated and guarded by the appropriate
+check, where only one loop holds the parallel annotations.
 

### 2014-09-02 doc/papers/cgo2014/paper.tex
Some slight improvements

@@
 
-PyPy~\cite{pypy2009}, a Python interpreter with a tracing JIT, or
-Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler are also viable ways to
-enhance Python performance. However Shed~Skin does not provide support for the
-Numpy module. The is an important work in progress from the PyPy community to
-support Numpy, but it implies a partial rewrite of the Numpy API in RPython,
-the language used by PyPy, to enable PyPy's optimization.
+PyPy~\cite{pypy2009}, a fully compatible Python interpreter with a tracing JIT,
+or Shed~Skin~\cite{shedskin2006}, a Python to C++ compiler, are also viable
+ways to enhance Python performance. However Shed~Skin does not provide support
+for the Numpy module. PyPy faces the issue of calling native libraries that
+cannot be optimized by a Python JIT compiler. There is an important work in
+progress from the community to support Numpy, but it implies a partial rewrite
+of the Numpy API in RPython, the language used by PyPy.
 
@@
 
-Many C++ linear algebra libraries~\cite{eigen,ublas} use \emph{expression
+Many C++ linear algebra libraries~\cite{eigen,ublas,esterie2014} use \emph{expression
 templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
@@
 
-\subsection{Range Analysis}
+%\subsection{Range Analysis}
 
@@
 
-A first exploration of vectorization of Python constructs is done in~\cite{wpmvp2014}.
+A first exploration of vectorization of Python constructs is done
+in~\cite{wpmvp2014}. It shows that language constructs like list comprehension
+offer many vectorization opportunities. This section focuses on the
+vectorization of complex Numpy expressions. In order to vectorize a numpy
+expression, the last dimension of the expression must be contiguous. This can
+be checked at compile time using type traits. The general form of a subscripted
+array can be expressed in C++ using a template type, say \texttt{numpy\_gexpr}.
+For instance the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix is
+represented by the C++ type in Listing~\ref{lst:gexpr}.
+
+\begin{lstlisting}[language=c++, caption={C++ type corresponding to the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix}, label={lst:gexpr}]
+numpy_gexpr<ndarray<double, 3>,
+            contiguous_slice,
+            long, slice>
+\end{lstlisting}
+
+This representation makes it easy to
+statically compute, in a vectorizable type trait, wether the last dimension of
+a subscript is contiguous or not. It can be generalized to the other array
+manipulations. BOOST.simd~\cite{esterie2012boost} provides a
+machine-independant abstraction for vector operation and supports all the C99
+math library. Thanks to this abstraction, all Numpy binary and unary functions
+and operators can be vectorized.
+
+The loop described in Listing~\ref{lst:eval-et} can be specialized using this
+vectorizable type trait, to automatically vectorize the innermost loop whenever
+it is legal. The choice is statically made at compile time.
 
@@
 %%
-\section{Experiments}
-\label{sec:xp}
+\section{Experiments} \label{sec:xp}
 
@@
 \subsection{The Pythran Compiler}
-\cite{pythran2013,isocxx11}
+
+The ideas described in this paper are implemnted in the Pythran
+compiler~\cite{pythran2013}, a translator from a subset of Python to
+C++11~\cite{isocxx11}. The architecture of the compiler is briefely described
+in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
+the Pythran subset. Pythran translates it in its internal representation, a
+simplified Python AST. It performs various optimization then outputs either
+Python code, in a source-to-source fashion, or C++ template code that makes
+call to the pythonic library that typically implements the \texttt{ndarray}
+interface. User annotations can be used to instanciate this code for the proper
+types and generate a native library. This library relies on Boost.Python
+library to match Python's C API.
+
+The Pythran compiler is an open source project publicly released under the BSD
+license~\footnote{\url{http://pythonhosted.org/pythran/}}.
 
@@
         \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
-        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{g++}};
+        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{c++}};
         \node (empty) [xshift=-1em, left=of gxx] {};

### 2014-09-02 doc/papers/cgo2014/paper.tex
Add bench part

@@
 command of the Numpy API and of good programming practices grows, they get to write code similar
-to Listing~\ref{lst:l2norm}.
-The fact that in Numpy, high-level constructs run faster is also a strong motivation.
-As a direct consequence, a compiler should focus on optimizing high-level constructs rather than raw
+to Listing~\ref{lst:l2norm}. The fact that in Numpy, high-level constructs run faster is also a
+strong motivation. As a direct consequence, a compiler should focus on optimizing
+high-level constructs rather than raw
 loops. Both to encourage good practice with respect to Numpy code, and because
@@
 performed by a function implemented as a loop performing a single operation,
-and the operator chaining is done at the interpreter level. This a typical
+and the operator chaining is done at the interpreter level. This is a typical
 problem in library design: if only a small set of functions is provided, it
@@
 memory load and store would have dominated the execution time. This is even more important as
-Numpy typically operates on double precision floats, which means only two (SSE) to four (AVX) scalar per vector registers.
+Numpy typically operates on double precision floats, which means only two (SSE)
+to four (AVX) scalar per vector registers.
 
@@
 The reference implementation of Numpy is a native Module, written mostly in C.
-It uses the BLAS API whenever possible and provides a relatively efficient array abstraction in the form of the \texttt{ndarray} data structure.
-This data structures makes it possible to use various styles of slicing and broadcasting, as illustrated in Listing~\ref{lst:pairwise}.
+It uses the BLAS API whenever possible and provides a relatively efficient array
+abstraction in the form of the \texttt{ndarray} data structure. This data
+structures makes it possible to use various styles of slicing and broadcasting,
+as illustrated in Listing~\ref{lst:pairwise}.
 
@@
 Listing~\ref{lst:pairwise-raw}. It avoids the creation of temporary arrays but
-runs almost 100 times slower due to the Python interpreter overhead and the C to Python conversion involved in every subscript.
+runs almost 100 times slower due to the Python interpreter overhead and the C
+to Python conversion involved in every subscript.
 
@@
 libraries, and a limited set of parallelism constructs, such as the possibility
-to define parallel loops, but no task parallelism. When possible, it unboxes %
-ref?  Python variables to improve performance. Without type annotations, the
+to define parallel loops, but no task parallelism. When possible, it unboxes % ref?
+Python variables to improve performance. Without type annotations, the
 performance improvement is not significant, but given enough type annotation,
@@
 loops~\cite{Aho2006} that limit usage of hand-written assembly to a few hot spots.
-% You are speaking about user written assembly code? Or generated code?
 Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
@@
 Alias analysis is impossible for Python programs in general, because of late
-binding and possible monkey patching---a way to modify the code of dynamic languages at run-time. In the context of a static compiler for
+binding and possible monkey patching---a way to modify the code of dynamic
+languages at run-time. In the context of a static compiler for
 numerical programs, it can be assumed that binding is done statically and no
@@
 The argument effect analysis described in \S~\ref{sec:fs} can be extended to
-global variables to determine if a function as side effects on its argument or
+global variables to determine if a function has side effects on its argument or
 on global variables (including I/O). This information makes it possible to flag
@@
 
-A first approach would be to use recursive tasks and manual loop chunking.
-\todo{pierrick ? c'est bête comme idée ?}
+%A first approach would be to use recursive tasks and manual loop chunking.
+%\todo{pierrick ? c'est bête comme idée ?}
 % Les taches sont en général plus lente que les boucles for (malheureusement)
@@
 
-% The first one holds 8 AMD Magny Cours processors for a
-% total of 48 cores. Each core has access to 64\,KB of L1 cache,
-% 512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
-% is shared between the 6 cores of a processor. This configuration
-% provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
-% up to SSE4. We
-% will refer to this configuration as \textbf{AMD48}.
-% On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
-% from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
-% from OpenMP improvement. This clang version is based on clang 3.4 release
-% (LLVM:233b1e3f034, clang-omp:6f358bcd87).
-
-% The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a ttal of
-% 8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
-% 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
-% is shared between the 8 processors. This configuration
-% provides a total of 63\,GB of main memory. It supports up to AVX. We will
-% refer to this configuration as \textbf{Intel8}.
-% On Intel8, backend compilers are gcc 4.9 (20140528) and the clang-omp version
-% from intel.
-
 \subsection{The Pythran Compiler}
 
-The ideas described in this paper are implemnted in the Pythran
+The ideas described in this paper are implemented in the Pythran
 compiler~\cite{pythran2013}, a translator from a subset of Python to
-C++11~\cite{isocxx11}. The architecture of the compiler is briefely described
+C++11~\cite{isocxx11}. The architecture of the compiler is briefly described
 in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
@@
 call to the pythonic library that typically implements the \texttt{ndarray}
-interface. User annotations can be used to instanciate this code for the proper
+interface. User annotations can be used to instantiate this code for the proper
 types and generate a native library. This library relies on Boost.Python
@@
 
+As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
+to evaluate these optimisation. It is use with Pythran (commit: 488b7ba)
+with optimization enabled/disabled. Finally, we compare result with Pypy (???),
+Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.
+
+Benchmarking is done one two architecture with different backend compilers.
+
+The first one holds 8 AMD Magny Cours processors for a
+total of 48 cores. Each core has access to 64\,KB of L1 cache,
+512\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
+is shared between the 6 cores of a processor. This configuration
+provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
+up to SSE4. We will refer to this configuration as \textbf{AMD48}.
+On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
+from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
+from OpenMP improvement. This clang version is based on clang 3.4 release
+(LLVM:233b1e3f034, clang-omp:6f358bcd87).
+
+The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
+8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
+256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
+is shared between the 8 processors. This configuration
+provides a total of 63\,GB of main memory. It supports up to AVX. We will
+refer to this configuration as \textbf{Intel8}.
+On Intel8, backend compilers is gcc 4.9 (20140528).
+
 \subsection{Performance of the Rosenbrock Function}
 
-\subsection{Results and Analysis}
+% For performance comparison between different Pythran version
+To compare optimisation result, we run the numpy-benchmarks test suite multiple
+time on the rosen.py benchmark. Results on Figure~\ref{rosenamd48.png} and \ref{rosenintel8.png} show difference
+one speed which could be explain by AVX support on the intel8 architecture while amd48
+support up to SSE4 only.
+% Haaaaa, I don't understand why OMP is not use...
+% tODO : I have to redo mesures without forward substitution to show difference.
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{rosen_amd48.png}
+\caption{Rosen on AMD48}
+\label{fig:rosenamd48}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{rosen_intel8.png}
+\caption{Rosen on Intel8}
+\label{fig:rosenintel8}
+\end{figure}
+
+\subsection{Comparison with others Python compiler}
+
+Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
+use of technique described in this paper. As we can see, both SIMD instruction
+and parallelization have an impact on execution time. We can notice that SIMD
+contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
+Also, we can notice that gcc as a backend compiler have more optimization opportunity.
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{arc_distance_amd48.png}
+\caption{Arc distance on AMD48}
+\label{fig:arc_distance}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{arc_distance_intel8.png}
+\caption{Arc distance on Intel8}
+\label{fig:arc_distance_intel8}
+\end{figure}
+
+At the opposite, Figure~\ref{fig:diffusion} show that clang is better when generated
+code make a more intensive use of templates (heavily use with gsliced\_ndarray).
+gsliced\_ndarray is a non aligned view on an array in this case so SIMD optimization
+can't be applied on the second dimension even if loop fusion and temporary
+elimination is performed.  We can notice that Parakeet perform really well on
+extensive gslice use.
+%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
+%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{diffusion_amd48.png}
+\caption{Diffusion on AMD48}
+\label{fig:diffusion}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{diffusion_intel8.png}
+\caption{Diffusion on Intel8}
+\label{fig:diffusion_intel8}
+\end{figure}
+
+Finally, Figure~\ref{fig:l2norme} show the l2norme algorithm introduce in
+\S~\ref{sec:stackoverflow}. Comparison with other compiler show that Pythran approach
+is correct for these kind of functions.
+%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
+%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{l2norm_amd48.png}
+\caption{L2norm on AMD48}
+\label{fig:l2norme}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[scale=0.4]{l2norm_intel8.png}
+\caption{L2norm on Intel8}
+\label{fig:l2norm_intel8}
+\end{figure}
 

### 2014-09-02 doc/papers/cgo2014/paper.tex
CGO: partial review

@@
 implement many optimizations on the control flow graph, basic blocks or
-loops~\cite{Aho2006} that limit usage of hand-written assembly to a few hot spots.
-Language extensions like OpenMP~\cite{openmp4} provide a convenient parallel
-programming model and libraries like Boost.SIMD~\cite{esterie2012boost} provides
-good abstractions for explicit vectorization. Many highly tuned kernels are
-available in the form of third-party libraries like GotoBlas~\cite{gotoblas2008}.
+loops~\cite{Aho2006} that limit the need of hand-written assembly to a few hot 
+spots.
+Language extensions like OpenMP~\cite{openmp4} provide a basic support for 
+parallel programming, while libraries like Boost.SIMD~\cite{esterie2012boost} 
+offer good abstractions for explicit vectorization. Many highly tuned kernels 
+are available in the form of third-party libraries like 
+GotoBlas~\cite{gotoblas2008}.
 % pas de nt2?
@@
 There is no need to reinvent the wheel when compiling numerical Python kernels.
-A more profitable approach is to map Python concept to existing ones, or
-translate them into existing them. As a consequence, this section focuses on
+A profitable approach is to map Python high-level to existing ones, or
+translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
+As a consequence, this section focuses on
 the translation of Python constructs to efficient C++ constructs. The choice of
@@
 
-\begin{lstlisting}[language=c++, caption={C++11 translated version of Python version for the Rosenbrock kernel.}, label={lst:rosen-cxx}, breaklines=true]
+\begin{lstlisting}[language=c++, 
+  caption={C++11 translated version of Python version for the Rosenbrock 
+           kernel.}, 
+  label={lst:rosen-cxx}, 
+  breaklines=true,
+  basicstyle=\small]
 typedef typename x::dtype T;
 constexpr size_t N = x::ndim;
-ndarray<T, N> t0 = 100L * numpy::square{}(x[slice(1L, None)] - numpy::square{}(x[slice(None,-1L)]));
-ndarray<T, N> t1 = numpy::square{}(1L - x[slice(None,-1L)]);
+ndarray<T, N> t0 = 100L * 
+  numpy::square{}(x[slice(1L, None)] - 
+  numpy::square{}(x[slice(None,-1L)]));
+ndarray<T, N> t1 = 
+  numpy::square{}(1L - x[slice(None,-1L)]);
 return numpy::sum{}((t0 + t1));
@@
 
-The computation of read-write effects involve the annotation of Numpy functions
+The computation of read-write effects involves the computation of argument 
+effects for user functions, and the annotation of Numpy functions
 so that the Python compiler knows that most of them do not modify their
-arguments, and the computation of argument effects for user functions. This can
+arguments. This can
 be done by considering each function individually and looking for the two kind
@@
 sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
-writes $i$, any outcoming edge $(i, j)$ can be safely removed, as we know for
+writes $i$, any outgoing edge $(i, j)$ can be safely removed, as we know for
 sure that the $i^\text{th}$ parameter is written. Similarly, an incoming edge
@@
 only node with write effects is \texttt{bar} that writes \texttt{n} through a
-subscript. Outcoming edges (1) and (2) can be removed because they are
-outcoming edges of \texttt{bar} starting by \texttt{n}. In coming edge (0) ends
+subscript. Outgoing edges (1) and (2) can be removed because they are
+outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
 by \texttt{n} that is written by \texttt{bar} so it is possible to prune it
@@
 The argument effect analysis described in \S~\ref{sec:fs} can be extended to
-global variables to determine if a function has side effects on its argument or
+global variables to determine if a function has side effects on its arguments or
 on global variables (including I/O). This information makes it possible to flag
@@
 As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
-to evaluate these optimisation. It is use with Pythran (commit: 488b7ba)
+to evaluate these optimizations. It is use with Pythran (commit: 488b7ba)
 with optimization enabled/disabled. Finally, we compare result with Pypy (???),
@@
 % For performance comparison between different Pythran version
-To compare optimisation result, we run the numpy-benchmarks test suite multiple
-time on the rosen.py benchmark. Results on Figure~\ref{rosenamd48.png} and \ref{rosenintel8.png} show difference
-one speed which could be explain by AVX support on the intel8 architecture while amd48
-support up to SSE4 only.
+To compare the different optimization schemes, we run the numpy-benchmarks test 
+suite multiple time on the \textit{rosen} benchmark.
+Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
+The difference between the two architectures can be explain by AVX support on 
+the Intel8 architecture while AMD48 support up to SSE4 only.
 % Haaaaa, I don't understand why OMP is not use...

### 2014-09-02 doc/papers/cgo2014/paper.tex
CGO: dedicated section for alias and effects analyses

@@
 The limitation of expression templates presented in
-\S~\ref{sec:expression-templates} is easy to solve at the Python compiler
-level, using forward substitution. Typically, forward substitution is
-supported by the computation of a dependency graph that itself relies on
-read-write effects and alias analysis.
+\S~\ref{sec:expression-templates} can be solved using forward substitution at 
+the Python compiler level. This optimization leverages the 
+\textit{def-use} chains~\cite{defuse} as well as interprocedural read-write 
+effects and alias analysis presented in the next sections. 
+Thanks to these analyses, the forward substitution implementation is quite 
+simple: \todo{it is not THAT simple indeed, there is "lazyness" analysis 
+involved}....
+
+This optimization is able to forward \texttt{t0} and \texttt{t1} in 
+Listing~\ref{lst:rosen}, avoiding a temporary allocation and providing more
+arithmetic intensity at the same time.
+
+\subsection{Alias Analysis}
+\label{sec:alias}
 
@@
 numerical programs, it can be assumed that binding is done statically and no
-monkey patching is done. In that context, it is relatively easy to design an
-intra-procedural alias analysis thanks to the absence of pointer arithmetic:
+monkey patching is done. Indeed the Pythran compiler operates on a subset of 
+Python that forbid this kind of dynamic patching. 
+In this context, it becomes possible to design an intra-procedural alias 
+analysis thanks to the absence of pointer arithmetic:
 the only language construction that can create aliasing is the assignment, thus
-a simple control-flow base alias analysis yields accurate results.
+a simple control-flow based alias analysis yields accurate results.
+
+\subsection{Interprocedural Effects Analysis}
+\label{sec:effects}
 
@@
 so that the Python compiler knows that most of them do not modify their
-arguments. This can
-be done by considering each function individually and looking for the two kind
-of constructs that can modify a function argument: indexing and further
-function calls.
+arguments. This can be done by considering each function individually and 
+looking for the two kind of constructs that can modify a function argument: 
+indexing and further function calls.
 
@@
 \end{figure}
-
-Thanks to these analysis, we are able to say that \texttt{t0} and \texttt{t1}
-can be forwarded in the Listing~\ref{lst:rosen}.
 

### 2014-09-02 doc/papers/cgo2014/paper.tex
CGO: Rephrase effects analysis

@@
 
-To inter procedurally compute the read-write effects, intra procedural effects
+To interprocedurally compute the read-write effects, intra procedural effects
 are first computed. Then we create a directed graph $G$ with as many nodes as
-functions. Each node contains a list of integer representing the indices of the
-formal argument that are modified intra-procedurally. Then we add an edge
-between two nodes, labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
+functions, and with as many edges as formal parameters between two nodes. Local
+variables used as arguments do not trigger an edge.
+Edges are labeled by an integer tuple $(i, j)$ if $i^\text{th}$ formal
 argument of the source is passed as $j^\text{th}$ effective argument of the
-sink. The goal is to prune all edges from $G$. To do so, given a node $n$ that
-writes $i$, any outgoing edge $(i, j)$ can be safely removed, as we know for
-sure that the $i^\text{th}$ parameter is written. Similarly, an incoming edge
-$(j, i)$ can be removed and $j$ is added to the source's list. Iteratively
-applying this process removes all edges that can propagate a write, the only
-remaining edges propagate a read and can be ultimately removed.
+sink. The analysis is iteratively pruning all edges from $G$ by considering any
+node for which the intraprocedural analysis shown that an argument $i$ is 
+written. Since the parameter is modified, any outgoing edge $(i, j)$ for this 
+node will not add any information to the analysis and then can be safely 
+removed. Incoming edges $(j, i)$ are used to propagate the effects along the 
+call graph: $j$ is added to the source's list of argument written and the source
+node is added to the list of nodes waiting to be processed. Iteratively
+applying this process removes all edges that can propagate a write, when no edge
+can be removed, the remaining ones propagate a read and can be ultimately 
+removed.
 
@@
 initial effect graphs are illustrated on Figure~\ref{fig:argument-effects}. The
-only node with write effects is \texttt{bar} that writes \texttt{n} through a
+only node with write effects is \texttt{bar} that modifies \texttt{n} through a
 subscript. Outgoing edges (1) and (2) can be removed because they are
 outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
-by \texttt{n} that is written by \texttt{bar} so it is possible to prune it
-while adding the start \texttt{n} to node \texttt{root}. No edge left, and we
-know that both \texttt{root} update \texttt{r} and \texttt{bar} update \texttt{n},
-but \texttt{foo} doesn't update its argument.
+by \texttt{n} that is written by \texttt{bar}, so it propagates the write 
+effects to the source node corresponding to the function \texttt{root}. After
+removing this edge, the only remaining edges are propagating read and the 
+analysis ands. The analysis informs that both \texttt{root} update \texttt{r} 
+and \texttt{bar} update \texttt{n}, but \texttt{foo} doesn't update its 
+argument.
 

### 2014-09-02 doc/papers/cgo2014/paper.tex
Comments + typos

@@
 
-As introduce in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
-to evaluate these optimizations. It is use with Pythran (commit: 488b7ba)
-with optimization enabled/disabled. Finally, we compare result with Pypy (???),
+As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
+to evaluate the optimizations described in\S~\ref{sec:pythran}. It uses Pythran (commit: 488b7ba)
+with optimization enabled/disabled. The results are compared with Pypy \todo{add}(???),
 Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.
 
-Benchmarking is done one two architecture with different backend compilers.
+Benchmarking is done on two architectures with different backend compilers. The
+\texttt{timeit} module is used to perform the measurments, taking the average
+execution time over several runs.
 
@@
 provides a total of 256\,GB (32\,GB per NUMA node) of main memory. It supports
-up to SSE4. We will refer to this configuration as \textbf{AMD48}.
+up to SSE4. We refer to this configuration as \textbf{AMD48}.
 On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
@@
 
+%FIXME: je me demande si on devrait pas utiliser juste cette archi. Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...
 The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
@@
 is shared between the 8 processors. This configuration
-provides a total of 63\,GB of main memory. It supports up to AVX. We will
+provides a total of 63\,GB of main memory. It supports up to AVX. We
 refer to this configuration as \textbf{Intel8}.
@@
 suite multiple time on the \textit{rosen} benchmark.
-Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
+Figures~\ref{fig:rosenamd48} and~\ref{fig:rosenintel8} illustrate the results.
 The difference between the two architectures can be explain by AVX support on
-the Intel8 architecture while AMD48 support up to SSE4 only.
+the Intel8 architecture while AMD48 supports up to SSE4 only.
 % Haaaaa, I don't understand why OMP is not use...
 % tODO : I have to redo mesures without forward substitution to show difference.
+% SG: may be the number of iteration is not enough? what array size did you use? (it should be specified in the text secion, BTW
 
@@
 \centering
-\includegraphics[scale=0.4]{rosen_amd48.png}
-\caption{Rosen on AMD48}
+\includegraphics[width=0.5\textwidth]{rosen_amd48.png}
+\caption{Execution time of the rosen kernel on AMD48.}
 \label{fig:rosenamd48}
@@
 \centering
-\includegraphics[scale=0.4]{rosen_intel8.png}
-\caption{Rosen on Intel8}
+\includegraphics[width=.5\textwidth]{rosen_intel8.png}
+\caption{Execution time of the rosen kernel on Intel8.}
 \label{fig:rosenintel8}
@@
 Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
-use of technique described in this paper. As we can see, both SIMD instruction
+use of the techniques described in this paper. As we can see, both SIMD instruction
 and parallelization have an impact on execution time. We can notice that SIMD
 contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
-Also, we can notice that gcc as a backend compiler have more optimization opportunity.
+Also, we can notice that gcc as a backend compiler finds more optimization opportunities.
+%FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
+% ça allègera aussi les graphes...
 
@@
 \centering
-\includegraphics[scale=0.4]{arc_distance_amd48.png}
-\caption{Arc distance on AMD48}
+\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
+\caption{Execution time of the arc distance kernel on AMD48.}
 \label{fig:arc_distance}
@@
 \centering
-\includegraphics[scale=0.4]{arc_distance_intel8.png}
-\caption{Arc distance on Intel8}
+\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
+\caption{Execution time of the arc distance kernel on Intel8.}
 \label{fig:arc_distance_intel8}
@@
 
-At the opposite, Figure~\ref{fig:diffusion} show that clang is better when generated
-code make a more intensive use of templates (heavily use with gsliced\_ndarray).
-gsliced\_ndarray is a non aligned view on an array in this case so SIMD optimization
+At the opposite, Figure~\ref{fig:diffusion} shows that clang is better when generated
+code make a more intensive use of templates (heavily used with \texttt{gsliced\_ndarray}).
+\texttt{gsliced\_ndarray} is a non aligned view on an array in this case so SIMD optimization
 can't be applied on the second dimension even if loop fusion and temporary
-elimination is performed.  We can notice that Parakeet perform really well on
+elimination is performed.  We can notice that Parakeet performs really well on
 extensive gslice use.
@@
 %pas du tout de parallelisme la. On note par contre l"overhead libgomp...
+%SG: +1
 
@@
 \centering
-\includegraphics[scale=0.4]{diffusion_amd48.png}
-\caption{Diffusion on AMD48}
+\includegraphics[width=0.5\textwidth]{diffusion_amd48.png}
+\caption{Execution time of the diffusion kernel on AMD48.}
 \label{fig:diffusion}
@@
 \centering
-\includegraphics[scale=0.4]{diffusion_intel8.png}
-\caption{Diffusion on Intel8}
+\includegraphics[width=0.5\textwidth]{diffusion_intel8.png}
+\caption{Execution time of the diffusion kernel on Intel8.}
 \label{fig:diffusion_intel8}
@@
 %pas du tout de parallelisme la. On note par contre l"overhead libgomp...
-
+%SG: +1
 \begin{figure}[t]
 \centering
-\includegraphics[scale=0.4]{l2norm_amd48.png}
-\caption{L2norm on AMD48}
+\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
+\caption{Execution time of the l2norm kernel on AMD48.}
 \label{fig:l2norme}
@@
 \centering
-\includegraphics[scale=0.4]{l2norm_intel8.png}
-\caption{L2norm on Intel8}
+\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
+\caption{Execution time of the l2norm kernel on Intel8.}
 \label{fig:l2norm_intel8}

### 2014-09-03 doc/papers/cgo2014/paper.tex
CGO: review

@@
     relevant toolbox/module/library/package. As a consequence, traditional
+    algorithms to perform
     loop optimization techniques like fusion, tiling and related
@@
 \keywords
-static compilation, parallelization, Python, C++, SIMD, runtime
+Static Compilation, Parallelization, Python, C++, SIMD, Runtime
 
@@
 translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
-As a consequence, this section focuses on
-the translation of Python constructs to efficient C++ constructs. The choice of
+As a consequence, this section focuses on the translation of Python constructs
+to efficient C++ constructs. The choice of
 C++ instead of C is motivated by the opportunities offered by meta programming
@@
 Python constructs.
+The meta-programming part of C++ makes it more than a traditional compiler
+runtime, closer to a code generator backend.
 
@@
 
-A naïve translation of such call to a generic C++ function would generate a
-call to a C++ function that would allocate enough memory to hold three integer,
+A naive translation of such call to a generic C++ function would generate a
+call to a C++ function that would allocate enough memory to hold three integers,
 fill the corresponding buffer and return it. A smarter version would generate a
 static array holding the relevant values. Figure~\ref{fig:unroll-range}
-compares two translation of a Python snippet that iterates over a fixed-size
+compares two translations of a Python snippet that iterates over a fixed-size
 list. The Clang compiler successfully turns Listing~\ref{lst:stat} into a
@@
 some function as pure, and to statically evaluate pure function calls with
-constant arguments. This evaluation is especially easy in Python thanks to the
-ability of the language to turn an AST expression into bytecode, evaluate it
+constant arguments. This evaluation is especially feasible in Python thanks to
+the ability of the language to turn an AST expression into bytecode, evaluate it
 and retrieve its constant value. For instance, this makes it possible to turn a
@@
 
-Once inter procedural constant evaluation is done (eventually followed by
-forward substitution), iteration over fixed size lists may appear, giving an
+Once interprocedural constant evaluation is done (eventually followed by
+forward substitution), iteration over fixed size lists may be exposed, giving an
 opportunity to fully unroll these loops at the Python level, thus avoiding the
@@
 target array expression \texttt{to} is made explicit through a recursive
-template call. The computation of the origin expression is only triggered for
-the innermost loop.
+template call. The computation of the \texttt{from} expression is only triggered
+for the innermost loop.
 
@@
 Listing~\ref{lst:eval-et} is to use a \texttt{parallel} directive from the
-OpenMP~\cite{openmp4} standard. Depending on the number of dimension of the
+OpenMP~\cite{openmp4} standard. Depending on the number of dimensions of the
 input array, this may lead to nested parallel regions. When OpenMP nested
-parallel regions are activated, each parallel region spawns as many thread as
+parallel regions are activated, each parallel region spawns as many threads as
 possible up to the runtime upper bound, which leads to oversubscription with
@@
 check, where only one loop holds the parallel annotations.
+\todo{this paragraph seems to be about a particular runtime issue, this shouldn't matter.
+Moreover, how is it relevant to this paper? It is an "implementation detail"}
 
@@
 parallelization of Python programs proposed in~\cite{pyhpc2013}.
+\todo{I think a specific section could be made on this topic}
 
@@
 offer many vectorization opportunities. This section focuses on the
-vectorization of complex Numpy expressions. In order to vectorize a numpy
+vectorization of complex Numpy expressions. In order to vectorize a Numpy
 expression, the last dimension of the expression must be contiguous. This can
@@
 
-\begin{lstlisting}[language=c++, caption={C++ type corresponding to the Numpy subscript \lstinline|a[:,0,::2]| on a 3D float matrix}, label={lst:gexpr}]
+\begin{lstlisting}[language=c++, caption={C++ type corresponding to the Numpy subscript \lstinline{a[:,0,::2]} on a 3D float matrix}, label={lst:gexpr}]
 numpy_gexpr<ndarray<double, 3>,
@@
 This representation makes it easy to
-statically compute, in a vectorizable type trait, wether the last dimension of
+statically compute, in a vectorizable type trait, whether the last dimension of
 a subscript is contiguous or not. It can be generalized to the other array
@@
 machine-independant abstraction for vector operation and supports all the C99
-math library. Thanks to this abstraction, all Numpy binary and unary functions
-and operators can be vectorized.
+math library. This abstraction helps to vectorize all Numpy binary and unary
+functions and operators in Pythran's runtime.
 
@@
 
-As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test suite
-to evaluate the optimizations described in\S~\ref{sec:pythran}. It uses Pythran (commit: 488b7ba)
-with optimization enabled/disabled. The results are compared with Pypy \todo{add}(???),
-Parakeet (commit: 4e55810), Numba (commit: b8a79c5) and Python 2.7.
+As introduced in \S~\ref{sec:stackoverflow}, we use the numpy-benchmark test 
+suite to evaluate the optimizations described in\S~\ref{sec:pythran}. 
+Pythran is used (commit: 488b7ba) with and without SIMD optimization. 
+The results are compared with Pypy \todo{add}(???), Parakeet (commit: 4e55810), 
+Numba (commit: b8a79c5) and Python 2.7.
 
@@
 up to SSE4. We refer to this configuration as \textbf{AMD48}.
-On AMD48, backend compilers are gcc (Debian 4.7.2-5) and the clang-omp version
-from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to benefits
-from OpenMP improvement. This clang version is based on clang 3.4 release
-(LLVM:233b1e3f034, clang-omp:6f358bcd87).
+On AMD48, the backend compilers are GCC (Debian 4.7.2-5) and the clang-omp
+version from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to
+benefits from OpenMP improvements. This Clang version is based on Clang 3.4
+release (LLVM:233b1e3f034, clang-omp:6f358bcd87).
 
-%FIXME: je me demande si on devrait pas utiliser juste cette archi. Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...
-The second one holds 8 Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
-8 cores (multithreading is not used). Each node can access to  64\,KB of L1 cache,
-256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
-is shared between the 8 processors. This configuration
-provides a total of 63\,GB of main memory. It supports up to AVX. We
-refer to this configuration as \textbf{Intel8}.
-On Intel8, backend compilers is gcc 4.9 (20140528).
+\todo{je me demande si on devrait pas utiliser juste cette archi. 
+Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...}
+\todo{Mehdi: agree, sauf si on peut avoir UN exemple avec 48 threads qui scale 
+"correctement"}
+
+The second one holds one Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
+8 cores (multithreading is not used). Each node can access to 64\,KB of L1
+cache, 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
+is shared between the 8 cores. This configuration provides a total of 64\,GB of 
+main memory. It supports up to AVX. We refer to this configuration as
+\textbf{Intel8}.
+On Intel8, backend compilers is GCC 4.9 (20140528).
 
@@
 To compare the different optimization schemes, we run the numpy-benchmarks test
-suite multiple time on the \textit{rosen} benchmark.
-Figures~\ref{fig:rosenamd48} and~\ref{fig:rosenintel8} illustrate the results.
+suite multiple times on the \textit{rosen} benchmark.
+Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
 The difference between the two architectures can be explain by AVX support on
@@
 % Haaaaa, I don't understand why OMP is not use...
-% tODO : I have to redo mesures without forward substitution to show difference.
+\todo{I have to redo mesures without forward substitution to show difference.}
 % SG: may be the number of iteration is not enough? what array size did you use? (it should be specified in the text secion, BTW
@@
 Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
-use of the techniques described in this paper. As we can see, both SIMD instruction
+use of the techniques described in this paper. As we can see, both vectorization
 and parallelization have an impact on execution time. We can notice that SIMD
-contribution is not as good as it could be on AMD48 as it doesn't have AVX support.
-Also, we can notice that gcc as a backend compiler finds more optimization opportunities.
+contribution is not as good on AMD48 as it is on Intel8, because it does not
+support AVX.
 %FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
@@
 
-At the opposite, Figure~\ref{fig:diffusion} shows that clang is better when generated
-code make a more intensive use of templates (heavily used with \texttt{gsliced\_ndarray}).
-\texttt{gsliced\_ndarray} is a non aligned view on an array in this case so SIMD optimization
-can't be applied on the second dimension even if loop fusion and temporary
-elimination is performed.  We can notice that Parakeet performs really well on
+Figure~\ref{fig:diffusion} show how Clang generated code is more performant
+when handling \lstinline{gsliced_ndarray}.
+This structure is a non aligned view on an array which prevent SIMD optimization
+on the second dimension even if loop fusion and temporary elimination is
+performed.  We can notice that Parakeet performs really well on
 extensive gslice use.
@@
 
-Finally, Figure~\ref{fig:l2norme} show the l2norme algorithm introduce in
-\S~\ref{sec:stackoverflow}. Comparison with other compiler show that Pythran approach
-is correct for these kind of functions.
+Finally, Figure~\ref{fig:l2norme} shows the L2norm algorithm introduced in
+\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
+approach taken by Pythran is relevant for these kind of functions.
 %TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a

### 2014-09-03 doc/papers/cgo2014/paper.tex
Add raw results

@@
 
+\begin{table*}
+
+    \centering{
+\begin{tabular}{|l||c|c|c|c|c|}
+\hline
+& Python & numba & parakeet & pypy & pythran \\
+\hline
+\hline
+       allpairs\_distances &   37727 &             0 &          1768 &         36855 & \textbf{1079} \\
+allpairs\_distances\_loops &   49799 &         55867 &          1781 &         50253 & \textbf{1758} \\
+             arc\_distance &    1365 &          1401 &          1596 &          1364 &  \textbf{494} \\
+                      conv & 1926454 &       2356675 &          2032 &       1899006 & \textbf{1944} \\
+              create\_grid &    3867 &          3950 &             0 & \textbf{3833} &          5512 \\
+                  cronbach &    1637 &          1681 &             0 &          1756 & \textbf{1136} \\
+                 diffusion &   22511 &         23564 &         21167 &         22450 & \textbf{9606} \\
+                    evolve &    5762 &             0 &             0 &          5522 & \textbf{3841} \\
+                      fdtd & 1558799 &       1824395 &             0 &       1538692 & \textbf{1198} \\
+                       fft &   25720 &             0 &             0 &         24771 & \textbf{1211} \\
+                  grouping &    2095 &             0 &             0 &          2104 &  \textbf{839} \\
+                   growcut & 1807849 &          5573 & \textbf{2200} &       1860926 &          3930 \\
+                    harris &    5516 &          5258 &          5084 &          5500 & \textbf{2994} \\
+                   hyantes &  259268 &             0 & \textbf{1865} &        261074 &          2359 \\
+                     julia & 2716770 &          2716 &          2930 &       2776183 & \textbf{2636} \\
+                    l2norm &    6057 &          6008 &         12574 &          6069 &  \textbf{540} \\
+             local\_maxima &   59822 &             0 &             0 &         60504 & \textbf{5084} \\
+                    lstsqr &    7886 &          8652 &             0 &          7891 & \textbf{1680} \\
+                    mandel &  437908 & \textbf{4361} &             0 &        439827 &          5075 \\
+             multiple\_sum &    2981 &          3194 &          8206 &          2975 & \textbf{2007} \\
+                  pairwise & 3888235 &          5777 &          3699 &       3873927 & \textbf{3435} \\
+            periodic\_dist &    1848 &          1983 &             0 &          1863 &  \textbf{951} \\
+                 repeating &    1216 &          1228 &             0 &          1226 &  \textbf{568} \\
+           reverse\_cumsum &    2620 &          2629 &             0 &          2614 & \textbf{2389} \\
+                     rosen &   14693 &         19419 &          1557 &         14611 &  \textbf{894} \\
+                 slowparts &    5687 &             0 & \textbf{1223} &          5900 &          1537 \\
+                 smoothing &  911126 & \textbf{4858} &          6258 &        935202 &          4924 \\
+           specialconvolve &    8211 &          6665 &         10340 &          8140 & \textbf{2449} \\
+              vibr\_energy &    2563 &          2450 &          2488 &          2548 & \textbf{1112} \\
+                      wave &   54234 &         54529 &             0 &         51471 & \textbf{1886} \\
+                     wdist &   82669 &         90274 & \textbf{1457} &         80406 &          1469 \\
+\hline
+\end{tabular}
+}
+\caption{Average execution times for all kernels of the numpy-benchamrks.}
+\end{table*}
+
 %%

### 2014-09-03 doc/papers/cgo2014/paper.tex
rework rosen measure and description

@@
 Pythran is used (commit: 488b7ba) with and without SIMD optimization. 
-The results are compared with Pypy \todo{add}(???), Parakeet (commit: 4e55810), 
+The results are compared with Pypy (commit: 632c832bd32f), Parakeet (commit: 4e55810), 
 Numba (commit: b8a79c5) and Python 2.7.
 
-Benchmarking is done on two architectures with different backend compilers. The
-\texttt{timeit} module is used to perform the measurments, taking the average
-execution time over several runs.
+numpy-benchmarks run 11 times benchmarks with an execution time around 1 seconde.
+THe best time, average time and standard deviation is return from these 11 runs.
+
+Benchmarking is done on two architectures with different backend compilers.
 
@@
 up to SSE4. We refer to this configuration as \textbf{AMD48}.
-On AMD48, the backend compilers are GCC (Debian 4.7.2-5) and the clang-omp
+On AMD48, the backend compilers is the clang-omp
 version from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to
@@
 release (LLVM:233b1e3f034, clang-omp:6f358bcd87).
-
-\todo{je me demande si on devrait pas utiliser juste cette archi. 
-Qu'est ce que l'autre apporte ? On sous utilise complètement les 48 coeurs...}
-\todo{Mehdi: agree, sauf si on peut avoir UN exemple avec 48 threads qui scale 
-"correctement"}
 
@@
 \textbf{Intel8}.
-On Intel8, backend compilers is GCC 4.9 (20140528).
+On Intel8, backend compilers is GCC 4.9 (20140528) with libgomp.
 
@@
 To compare the different optimization schemes, we run the numpy-benchmarks test
-suite multiple times on the \textit{rosen} benchmark.
+suite multiple times on the \textit{rosen} benchmark. The algorithm is the one
+on Listing~\ref{lst:rosen} and input is a raw array of 1,000,000 of double elements.
 Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
-The difference between the two architectures can be explain by AVX support on
+The speedup difference between the two architectures can be explain by AVX support on
 the Intel8 architecture while AMD48 supports up to SSE4 only.
-% Haaaaa, I don't understand why OMP is not use...
-\todo{I have to redo mesures without forward substitution to show difference.}
-% SG: may be the number of iteration is not enough? what array size did you use? (it should be specified in the text secion, BTW
+We can also notice importance of forward substitution on this benchmarks as
+we avoid extra loop, 2 temporary array assignment and avoid some load/store for
+SIMD instructions.
 
@@
 Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
-use of the techniques described in this paper. As we can see, both vectorization
+use of the techniques described in this paper. It measures performances on the
+arc\_distance benchmark. It is a kernel with intensive points to points computation
+and trigonometrics use. As we can see, both vectorization
 and parallelization have an impact on execution time. We can notice that SIMD
 contribution is not as good on AMD48 as it is on Intel8, because it does not
-support AVX.
+support AVX. This program also benefits from OpenMP. Input is 10000 elements 
+double raw array.
+
 %FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
 % ça allègera aussi les graphes...
+% On change juste le jeu de donnée plutot que de supprimer une archi
 

### 2014-09-03 doc/papers/cgo2014/paper.tex
update graphes

@@
 suite multiple times on the \textit{rosen} benchmark. The algorithm is the one
-on Listing~\ref{lst:rosen} and input is a raw array of 1,000,000 of double elements.
+on Listing~\ref{lst:rosen} and input is a raw array of 1,000,000 double elements.
 Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
@@
 \subsection{Comparison with others Python compiler}
+\todo{a better title have to be found}
 
-Figure~\ref{fig:arc_distance} is a typical example where we can make intensive
+Figure~\ref{fig:arc_distance} and \ref{fig:arc_distance_intel8} are typicals
+example where we can make intensive
 use of the techniques described in this paper. It measures performances on the
@@
 contribution is not as good on AMD48 as it is on Intel8, because it does not
-support AVX. This program also benefits from OpenMP. Input is 10000 elements 
-double raw array.
-
-%FIXME: mehdi propose de n'utiliser qu'un backend pour centrer le propos sur Pythran/Numpy.
-% ça allègera aussi les graphes...
-% On change juste le jeu de donnée plutot que de supprimer une archi
+support AVX. This program also benefits from OpenMP and is scalable up to 20 cores.
+The input is a raw array of 10,000,000 double elements.
 
@@
 \includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
-\caption{Execution time of the arc distance kernel on AMD48.}
+\caption{Speed up for the arc distance kernel on AMD48.}
 \label{fig:arc_distance}
@@
 \includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
-\caption{Execution time of the arc distance kernel on Intel8.}
+\caption{Speed up for the arc distance kernel on Intel48.}
 \label{fig:arc_distance_intel8}
@@
 
-Figure~\ref{fig:diffusion} show how Clang generated code is more performant
-when handling \lstinline{gsliced_ndarray}.
-This structure is a non aligned view on an array which prevent SIMD optimization
-on the second dimension even if loop fusion and temporary elimination is
-performed.  We can notice that Parakeet performs really well on
-extensive gslice use.
-%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
-%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
-%SG: +1
+Finally, Figure~\ref{fig:l2norme} and \ref{fig:l2norm_intel8} shows the L2norm algorithm introduced in
+\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
+approach taken by Pythran is relevant for these kind of functions. Just like for
+the arc\_distance benchmarks, we can notice use of AVX when it is available.
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{diffusion_amd48.png}
-\caption{Execution time of the diffusion kernel on AMD48.}
-\label{fig:diffusion}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{diffusion_intel8.png}
-\caption{Execution time of the diffusion kernel on Intel8.}
-\label{fig:diffusion_intel8}
-\end{figure}
-
-Finally, Figure~\ref{fig:l2norme} shows the L2norm algorithm introduced in
-\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
-approach taken by Pythran is relevant for these kind of functions.
-%TODO IL faut refaire les mesures en changeant la limite OpenMP. Parce qu'on a
-%pas du tout de parallelisme la. On note par contre l"overhead libgomp...
-%SG: +1
 \begin{figure}[t]

### 2014-09-04 doc/papers/cgo2014/paper.tex
Add a small snippet concerning the full benchmark

@@
 
-Figure~\ref{fig:arc_distance} and \ref{fig:arc_distance_intel8} are typicals
-example where we can make intensive
-use of the techniques described in this paper. It measures performances on the
-arc\_distance benchmark. It is a kernel with intensive points to points computation
-and trigonometrics use. As we can see, both vectorization
-and parallelization have an impact on execution time. We can notice that SIMD
-contribution is not as good on AMD48 as it is on Intel8, because it does not
-support AVX. This program also benefits from OpenMP and is scalable up to 20 cores.
-The input is a raw array of 10,000,000 double elements.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
-\caption{Speed up for the arc distance kernel on AMD48.}
-\label{fig:arc_distance}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
-\caption{Speed up for the arc distance kernel on Intel48.}
-\label{fig:arc_distance_intel8}
-\end{figure}
-
-Finally, Figure~\ref{fig:l2norme} and \ref{fig:l2norm_intel8} shows the L2norm algorithm introduced in
-\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
-approach taken by Pythran is relevant for these kind of functions. Just like for
-the arc\_distance benchmarks, we can notice use of AVX when it is available.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
-\caption{Execution time of the l2norm kernel on AMD48.}
-\label{fig:l2norme}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
-\caption{Execution time of the l2norm kernel on Intel8.}
-\label{fig:l2norm_intel8}
-\end{figure}
+Table~\ref{tbl:overview} presents the average execution time for each kernel
+and each compiler, with the lowest values highlighted in
+bold font. Zeros are used when the given compiler failed to
+compile the kernel. Concerning the Numba compiler, this
+benchmarks conﬁrms that it works best on explicit loops. It
+gives non performance gain when run on high-level kernels,
+either falling back to the generation of calls to Python C
+API, or stopping compilation. PyPy is the most robust of
+the compilers, but to a few exceptions, it does not generate
+code more efﬁcient than the original version. This reﬂects
+the fact that Numpy support is still a work-in-progress task.
+Parakeet performs extremely well on high-level benchmarks.
+Although it does not supports some Numpy constructs, it
+successfully parallelize many of them, and also generates
+efﬁcient code for explicit loops. Pythran still performs better
+than Parakeet on some kernels, whenever it successfully
+vectorizes operations.
 
@@
 \caption{Average execution times for all kernels of the numpy-benchamrks.}
+\label{tbl:overview}
 \end{table*}
+
+Figure~\ref{fig:arc_distance} and \ref{fig:arc_distance_intel8} are typicals
+example where we can make intensive
+use of the techniques described in this paper. It measures performances on the
+arc\_distance benchmark. It is a kernel with intensive points to points computation
+and trigonometrics use. As we can see, both vectorization
+and parallelization have an impact on execution time. We can notice that SIMD
+contribution is not as good on AMD48 as it is on Intel8, because it does not
+support AVX. This program also benefits from OpenMP and is scalable up to 20 cores.
+The input is a raw array of 10,000,000 double elements.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
+\caption{Speed up for the arc distance kernel on AMD48.}
+\label{fig:arc_distance}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
+\caption{Speed up for the arc distance kernel on Intel48.}
+\label{fig:arc_distance_intel8}
+\end{figure}
+
+Finally, Figure~\ref{fig:l2norme} and \ref{fig:l2norm_intel8} shows the L2norm algorithm introduced in
+\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
+approach taken by Pythran is relevant for these kind of functions. Just like for
+the arc\_distance benchmarks, we can notice use of AVX when it is available.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
+\caption{Execution time of the l2norm kernel on AMD48.}
+\label{fig:l2norme}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
+\caption{Execution time of the l2norm kernel on Intel8.}
+\label{fig:l2norm_intel8}
+\end{figure}
 

### 2014-09-04 doc/papers/cgo2014/paper.tex
Update table result
* Parakeet perform really well on julia...

@@
 \hline
-       allpairs\_distances &   37727 &             0 &          1768 &         36855 & \textbf{1079} \\
-allpairs\_distances\_loops &   49799 &         55867 &          1781 &         50253 & \textbf{1758} \\
-             arc\_distance &    1365 &          1401 &          1596 &          1364 &  \textbf{494} \\
-                      conv & 1926454 &       2356675 &          2032 &       1899006 & \textbf{1944} \\
-              create\_grid &    3867 &          3950 &             0 & \textbf{3833} &          5512 \\
-                  cronbach &    1637 &          1681 &             0 &          1756 & \textbf{1136} \\
-                 diffusion &   22511 &         23564 &         21167 &         22450 & \textbf{9606} \\
-                    evolve &    5762 &             0 &             0 &          5522 & \textbf{3841} \\
-                      fdtd & 1558799 &       1824395 &             0 &       1538692 & \textbf{1198} \\
-                       fft &   25720 &             0 &             0 &         24771 & \textbf{1211} \\
-                  grouping &    2095 &             0 &             0 &          2104 &  \textbf{839} \\
-                   growcut & 1807849 &          5573 & \textbf{2200} &       1860926 &          3930 \\
-                    harris &    5516 &          5258 &          5084 &          5500 & \textbf{2994} \\
-                   hyantes &  259268 &             0 & \textbf{1865} &        261074 &          2359 \\
-                     julia & 2716770 &          2716 &          2930 &       2776183 & \textbf{2636} \\
-                    l2norm &    6057 &          6008 &         12574 &          6069 &  \textbf{540} \\
-             local\_maxima &   59822 &             0 &             0 &         60504 & \textbf{5084} \\
-                    lstsqr &    7886 &          8652 &             0 &          7891 & \textbf{1680} \\
-                    mandel &  437908 & \textbf{4361} &             0 &        439827 &          5075 \\
-             multiple\_sum &    2981 &          3194 &          8206 &          2975 & \textbf{2007} \\
-                  pairwise & 3888235 &          5777 &          3699 &       3873927 & \textbf{3435} \\
-            periodic\_dist &    1848 &          1983 &             0 &          1863 &  \textbf{951} \\
-                 repeating &    1216 &          1228 &             0 &          1226 &  \textbf{568} \\
-           reverse\_cumsum &    2620 &          2629 &             0 &          2614 & \textbf{2389} \\
-                     rosen &   14693 &         19419 &          1557 &         14611 &  \textbf{894} \\
-                 slowparts &    5687 &             0 & \textbf{1223} &          5900 &          1537 \\
-                 smoothing &  911126 & \textbf{4858} &          6258 &        935202 &          4924 \\
-           specialconvolve &    8211 &          6665 &         10340 &          8140 & \textbf{2449} \\
-              vibr\_energy &    2563 &          2450 &          2488 &          2548 & \textbf{1112} \\
-                      wave &   54234 &         54529 &             0 &         51471 & \textbf{1886} \\
-                     wdist &   82669 &         90274 & \textbf{1457} &         80406 &          1469 \\
+         allpairs\_distances &   58293 &             0 &          3170 &   59232 & \textbf{1583} \\
+  allpairs\_distances\_loops &   64137 &         72748 &          3158 &   65287 & \textbf{3146} \\
+               arc\_distance &    2796 &          2792 &          3084 &    2791 &  \textbf{103} \\
+                        conv & 3260079 &             0 & \textbf{3284} & 3268298 &          3420 \\
+                create\_grid &    5231 & \textbf{5044} &             0 &    5240 &          5587 \\
+                    cronbach &    1905 &          2016 &             0 &    1909 & \textbf{1890} \\
+                   diffusion &   42998 &         45030 & \textbf{8843} &   43238 &         21704 \\
+                      evolve &    8236 &             0 &             0 &    8319 & \textbf{6242} \\
+                        fdtd & 1940884 &       1498849 &             0 & 2007839 & \textbf{2294} \\
+                         fft &   27120 &             0 &             0 &   27009 & \textbf{2151} \\
+                    grouping &    3739 &             0 &             0 &    3740 & \textbf{1747} \\
+                     growcut & 2112332 &         11617 & \textbf{5485} & 2141571 &         10338 \\
+                      harris &    7173 &          9116 & \textbf{4240} &    7322 &          4889 \\
+                     hyantes &  288852 &             0 & \textbf{2834} &  289074 &          3522 \\
+                       julia & 1744357 &          5168 &  \textbf{109} & 1771329 &          4255 \\
+                      l2norm &    5966 &          5948 &          9210 &    6559 &  \textbf{551} \\
+               local\_maxima &  132237 &             0 &             0 &  132047 & \textbf{8561} \\
+                      lstsqr &    9254 &          7528 &             0 &    8992 & \textbf{1414} \\
+                      mandel &  951612 & \textbf{8175} &             0 &  951687 &         10341 \\
+               multiple\_sum &    5477 &          5813 &          9805 &    5475 & \textbf{3503} \\
+                    pairwise & 3362150 &         10764 &          6199 & 3347574 & \textbf{6015} \\
+              periodic\_dist &    2559 &          2675 &             0 &    2562 & \textbf{2292} \\
+                   repeating &    2808 &          2836 &             0 &    2795 &  \textbf{810} \\
+             reverse\_cumsum &    4583 &          4828 &             0 &    4583 & \textbf{2379} \\
+                       rosen &   15382 &         14309 &          2763 &   16065 &  \textbf{939} \\
+                   slowparts &    8302 &             0 & \textbf{1610} &    8241 &          1927 \\
+                   smoothing &  834057 &          8478 &          9771 &  819827 & \textbf{8254} \\
+             specialconvolve &   12537 &         11978 &          6039 &   12570 & \textbf{5572} \\
+                vibr\_energy &    5277 &          5275 &          5334 &    5416 &  \textbf{146} \\
+                        wave &   72394 &         67590 &             0 &   72300 & \textbf{5186} \\
+                       wdist &  108572 &        119211 &          4197 &  108283 & \textbf{4010} \\
 \hline
@@
 }
-\caption{Average execution times for all kernels of the numpy-benchamrks.}
 \label{tbl:overview}
+\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
 \end{table*}
@@
 example where we can make intensive
-use of the techniques described in this paper. It measures performances on the
-arc\_distance benchmark. It is a kernel with intensive points to points computation
-and trigonometrics use. As we can see, both vectorization
-and parallelization have an impact on execution time. We can notice that SIMD
-contribution is not as good on AMD48 as it is on Intel8, because it does not
-support AVX. This program also benefits from OpenMP and is scalable up to 20 cores.
-The input is a raw array of 10,000,000 double elements.
+use of the techniques described in this paper. It shows speedup compare to one
+thread execution for the arc\_distance benchmark.This kernel performs intensive
+points to points computation and trigonometrics use. We can see that OpenMP
+improvement is scalable up to 20 threads with a raw array of 10,000,000 double
+elements as input.
 

### 2014-09-05 doc/papers/cgo2014/paper.tex
Fix dangling reference (in LaTeX :-)
CGO: partial Review

@@
     algebra routines. It first conducts a case study based on Numpy use
-    cases from the StackOverflow question and answer website, then studies
+    cases from the StackOverflow questions and answers website, then studies
     existing compilers for numerical Python. Optimization opportunities,
@@
 optimization of high-level numerical kernels written in Python using the Numpy
-package. Its shows an implementation of these techniques through the Pythran
-compiler and provided a benchmarks test suite for performance measures with
-others Python compilers.
+package. Its introduces the implementation of these techniques in the Pythran
+compiler and provides a benchmarks test suite for performance measures along
+several Python compilers.
 Section~\ref{sec:stackoverflow} presents a case study based on user
-input gathered on the StackOverflow question and answer site.
+input gathered on the StackOverflow questions and answers website.
 Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
 showcases the optimization opportunities typically found in Numpy-based
-kernels. Existing compilation approach to optimize such kernels are discussed
-in Section~\ref{sec:compilers}, then Section~\ref{sec:pythran} emphasises on a
+kernels. Existing compilation approaches to optimize such kernels are discussed
+in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on a
 compiler-runtime cooperative approach to solve the performance issues.
-Experimentations showcasing important performance gains are detailed in
+Finally, experimentations showcasing important performance gains are detailed in
 Section~\ref{sec:xp}.
@@
 users may answer. A voting system is then used to sort the answers. Said
-otherwise, the website provides a database of question and answers on
+otherwise, the website provides a database of questions and answers on
 programming topics, including Python.
@@
 A query using the terms \emph{numpy} and \emph{slow} yields a lot of results, many
-of which use the pattern ``Q: my code is slow'' followed by an upvoted answer
+of which expose the pattern ``Q: my code is slow'' followed by an upvoted answer
 ``A: rewrite it that way'', where the original code performs explicit looping
@@
 \begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in Numpy.}, label={lst:l2norm-loopy}]
-def slow(x):
+def l2norm_slow(x):
     r = np.empty(x.shape[0])
@@
 import numpy as np
-def l2norm(x):
+def l2norm_fast(x):
     return np.sum(np.abs(x)**2, axis=1)
@@
 use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
-command of the Numpy API and of good programming practices grows, they get to write code similar
-to Listing~\ref{lst:l2norm}. The fact that in Numpy, high-level constructs run faster is also a
-strong motivation. As a direct consequence, a compiler should focus on optimizing
+knowledge of the Numpy API and of good programming practices grows, they get to 
+write code similar to Listing~\ref{lst:l2norm}. 
+The fact that in Numpy, high-level constructs run faster is also a strong 
+motivation. As a direct consequence, a compiler should focus on optimizing
 high-level constructs rather than raw
@@
 temporaries elimination and operator fusion, parallelization would be more
-effective as barrier between binary operations are no longer needed and loads
+effective as barriers between binary operations are no longer needed and loads
 and stores will be used to perform multiple operations.
@@
 the Python compiler level. This optimization leverages the
-\textit{def-use} chains~\cite{defuse} as well as interprocedural read-write
+\textit{def-use} chains as well as interprocedural read-write
 effects and alias analysis presented in the next sections.
@@
 \subsection{Comparison with others Python compiler}
-\todo{a better title have to be found}
 
@@
 }
+\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
 \label{tbl:overview}
-\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
 \end{table*}

### 2014-09-05 doc/papers/cgo2014/paper.tex
Include BC's review

@@
     As more scientists use these scripting languages, there is an on-going
-    research effort to compile them into efficient native code. However as
-    these languages favor a high-level coding style, the computation kernels are
-    different from what we use to see in traditional C or Fortran programs:
+    research effort to compile them into efficient native code. However these
+    languages favor a high-level coding style, so the computation kernels are
+    different from those used in traditional C or Fortran programs:
     explicit raw loops are replaced by high-level constructs and calls to the
     relevant toolbox/module/library/package. As a consequence, traditional
-    algorithms to perform
-    loop optimization techniques like fusion, tiling and related
-    vectorization or parallelization techniques are not directly applicable.
+    algorithms to perform loop optimization techniques like fusion, tiling and
+    related vectorization or parallelization techniques are not directly
+    applicable.
 
@@
 with the Numpy package\cite{oliphant2007,numpyarray2011}, to scientific
-packages with the Scipy package~\cite{scipy}, plotting facilities with the
+packages with Scipy~\cite{scipy}, plotting facilities with the
 Matplotlib package, enhanced terminal and notebooks with IPython. As a
@@
 
-As Python based scientific tools gets widely used, the question of High
+As Python based scientific tools get widely used, the question of High
 performance Computing naturally arises, and it is the focus of many recent
@@
 package. Its introduces the implementation of these techniques in the Pythran
-compiler and provides a benchmarks test suite for performance measures along
+compiler and provides a benchmark test suite for performance measures along
 several Python compilers.
 Section~\ref{sec:stackoverflow} presents a case study based on user
-input gathered on the StackOverflow questions and answers website.
+input gathered from the StackOverflow questions and answers website.
 Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
@@
 kernels. Existing compilation approaches to optimize such kernels are discussed
-in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on a
-compiler-runtime cooperative approach to solve the performance issues.
+in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on
+compiler-runtime cooperation to solve the performance issues.
 Finally, experimentations showcasing important performance gains are detailed in
@@
 process toward realistic examples. To achieve this goal, we have gathered a few
-examples from existing compilers' test suite, but these may have been modified to
+examples from existing compilers' test suites, but these may have been modified to
 overcome some compiler limitations. A less biased source of benchmarks should
-be independent from the compiler community and focused on matching user needs.
+be independent from the compiler community and showed focus on matching user needs.
 To achieve that goal, we collected synthetic scientific kernels from the
@@
 
-StackOverflow~\footnote{\url{http://stackoverflow.com/}} is a question and
+StackOverflow\footnote{\url{http://stackoverflow.com/}} is a question and
 answer site for programmers. Any registered user can submit questions and other
@@
 the proposed answer is to use the more compact version presented in
-Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loops and
+Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loop and
 roughly achieves a $\times 4$ speedup over the loop version.
@@
 write code similar to Listing~\ref{lst:l2norm}. 
-The fact that in Numpy, high-level constructs run faster is also a strong 
+The fact that Numpy high-level constructs runs faster is also a strong 
 motivation. As a direct consequence, a compiler should focus on optimizing
@@
 
-The benchmarks collected during this process are publicly available on
+A selection of benchmarks collected during this process, selected for their
+conciseness and representativity of high-level calls, are publicly available on
 \url{https://github.com/serge-sans-paille/numpy-benchmarks}, with a set of
-benchmarking routine specialized for major Python/Numpy compilers.
+benchmarking routines specialized for major Python/Numpy compilers. This
+benchmark is used in \S~\ref{sec:xp} to evaluate the validity of the approach
+described in this paper.
 
@@
 
-Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapted
-from the Scipy source code of \texttt{scipy.optimize.rosen}. This kernel makes
-use of Numpy's \texttt{sum} function, Python's square notation and Numpy's
-array slicing. It is a good example of high-level Python kernel, although
-using the function \texttt{scipy.optimize.rosen} directly would naturally make sense.
-
-\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy.}, label={lst:rosen}]
-def rosen(x):
-    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
-    t1 = (1 - x[:-1]) ** 2
-    return numpy.sum(t0 + t1)
-\end{lstlisting}
-
-This section goes through all the optimization opportunities in
-Listing~\ref{lst:rosen} as a showcase of what an optimizing compiler could do.
-
-\subsection{Temporaries Elimination}
-\label{sec:temporaries-elimination}
-
-In Numpy, all point-to-point array operations allocate a new array that holds
-the computation result. This behavior is consistent with many Python standard
-module, but it is a very inefficient design choice, as it keeps on polluting
-the cache with potentially large fresh storage and add an extra
-allocation/deallocation costly operation. In the \textit{rosen} function from
-Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
-create a temporary array but a view) to hold intermediate steps. Had the
-computation been lazy, no temporary would have been needed.
-
-\subsection{Operator Fusion}
-\label{sec:operator-fusion}
-
-As Numpy is a native library mostly written in C, each operator computation is
-performed by a function implemented as a loop performing a single operation,
-and the operator chaining is done at the interpreter level. This is a typical
-problem in library design: if only a small set of functions is provided, it
-prevents the optimization of merging multiple operators into a single specialized
-operator, and provided many operator combinations as part of the library yields
-better performance to the price of API bloat. The Listing~\ref{lst:rosen}
-illustrates the use of a small set of functions: a loop is used for
-each temporary computation, plus an extra loop for the \texttt{numpy.sum}
-reduction, where a single loop would have been needed with operator fusion.
-
-\subsection{Loop Vectorization and Parallelization}
-
-Without operator fusion, there would be very little benefit to generate SIMD
-instructions for the respective array operations used by each operator, as the
-memory load and store would have dominated the execution time. This is even more important as
-Numpy typically operates on double precision floats, which means only two (SSE)
-to four (AVX) scalar per vector registers.
-
-Parallelization also suffers from the lack of operator fusion: a
-synchronization fence is needed between each temporary computation.  Hence the
-loop computation intensity would be very low compared to the memory pressure
-implied by two array reads and one array write for a single binary operator.
-
-On the opposite, if all computations were merged into a single loop using
-temporaries elimination and operator fusion, parallelization would be more
-effective as barriers between binary operations are no longer needed and loads
-and stores will be used to perform multiple operations.
-
-%%
-%%
-\section{Existing Compilation Approach}
-\label{sec:compilers}
-
-The success of Python and the development of its large ecosystem triggered
-several attempts to improve its execution speed. The most significant basic
-block for high performance numerical simulations is certainly Numpy, a
-numerical library to manipulate dense arrays. However, this library-like
-approach has inherent limitations that have been exposed in \S~\ref{sec:optimize}.
-To address these limitations, several compiler-based approaches have been proposed.
+This section briefly presents the core concepts of the Numpy package, then goes
+through all the optimization opportunities in a small kernel as a showcase of
+the optimization opportunities.
 
@@
 
+\subsection{The Rosenbrock Function}
+
+This section uses the kernel illustrated in Listing~\ref{lst:rosen} and adapted
+from the Scipy source code of \texttt{scipy.optimize.rosen}. It uses Numpy's
+\texttt{sum} function, Python's square notation and Numpy's array slicing. It
+is a good example of high-level Python kernel, although using the function
+\texttt{scipy.optimize.rosen} directly would naturally make sense.
+
+Note that due to dynamic typing, this function can take arrays of different
+shapes and types as input.
+
+\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy.}, label={lst:rosen}]
+def rosen(x):
+    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
+    t1 = (1 - x[:-1]) ** 2
+    return numpy.sum(t0 + t1)
+\end{lstlisting}
+
+\subsection{Temporaries Elimination}
+\label{sec:temporaries-elimination}
+
+In Numpy, any point-to-point array operation allocates a new array that holds
+the computation result. This behavior is consistent with many Python standard
+module, but it is a very inefficient design choice, as it keeps on polluting
+the cache with potentially large fresh storage and adds extra
+allocation/deallocation operations, that have a very bad caching effect. In the
+\textit{rosen} function from Listing~\ref{lst:rosen}, 7 temporary arrays are
+allocated (slicing does not create a temporary array but a view) to hold
+intermediate steps. Had the expression been evaluated lazily, no temporary
+would have been needed.
+
+\subsection{Operator Fusion}
+\label{sec:operator-fusion}
+
+As Numpy is a native library mostly written in C, each operator computation is
+performed by a function implemented as a loop performing a single operation,
+and the operator chaining is done at the interpreter level. This is a typical
+problem in library design: if only a small set of functions is provided, it
+prevents the optimization of merging multiple operators into a single specialized
+operator. On the contrary, providing many operator combinations as part of the library yields
+better performance to the price of API bloat. Listing~\ref{lst:rosen}
+illustrates the use of a small set of functions: a loop is generated for
+each temporary computation, plus an extra loop for the \texttt{numpy.sum}
+reduction, whereas a single loop would have been necessary with operator fusion.
+
+\subsection{Loop Vectorization and Parallelization}
+
+Without operator fusion, there would be very little benefit to generate SIMD
+instructions for the respective array operations used by each operator, as the
+memory loads and stores would have dominated the execution time. This is even more important as
+Numpy typically operates on double precision floats, which means only two (SSE)
+to four (AVX) scalars per vector registers.
+
+Parallelization would also suffer from the lack of operator fusion: a
+synchronization fence is needed between each temporary computation.  Hence the
+loop computation intensity would be very low compared to the memory pressure
+implied by two array reads and one array write for a single binary operator.
+
+On the opposite, if all computations were merged into a single loop using
+temporaries elimination and operator fusion, parallelization would be more
+effective as barriers between binary operations are no longer needed and loads
+and stores are counterbalanced by several operations.
+
+%%
+%%
+\section{Existing Compilation Approaches}
+\label{sec:compilers}
+
+The success of Python and the development of its large ecosystem triggered
+several attempts to improve its execution speed. The most significant basic
+block for high performance numerical simulations is certainly Numpy, a
+numerical library to manipulate dense arrays. However, this library-like
+approach has inherent limitations that have been exposed in \S~\ref{sec:optimize}.
+To address these limitations, several compiler-based approaches have been proposed.
+
 \subsection{Cython}
@@
 libraries, and a limited set of parallelism constructs, such as the possibility
-to define parallel loops, but no task parallelism. When possible, it unboxes % ref?
-Python variables to improve performance. Without type annotations, the
-performance improvement is not significant, but given enough type annotation,
-Cython can generates code that runs as fast as its C equivalent.
+to define parallel loops, but no task parallelism. When possible, it unboxes
+Python variables to improve performance, i.e. it removes the generic object
+layer around a variable to turn it into a native type. Without type
+annotations, the performance improvement is not significant, but given enough
+type annotations, Cython can generate code that runs as fast as its C
+equivalent.
 
 The drawback is that the programmer has to avoid high-level constructs and
-instead expresses his algorithm using raw loops.  The \textit{pairwise}
-function rewritten for Cython is shown listing~\ref{lst:pairwise-cython},
+must instead express his algorithm using raw loops.  The \textit{pairwise}
+function rewritten for Cython is shown in Listing~\ref{lst:pairwise-cython},
 showing usage of compiler annotations in the form of function decorators.
@@
 
-The Numba compiler uses additional type information collected at run-time to
+The Numba compiler collects additional type information at run-time to
 generate sequential LLVM bytecode.  Numba uses Just-In-Time(JIT) compilation
 and does not suffer from backward-incompatibility issues: a single
-\texttt{@jit} decorator is used to flag kernels to optimize. When meeting an
+\texttt{@jit} decorator is provided to flag kernels to optimize. When meeting an
 unsupported construct, Numba silently falls back to Python C-API calls.
 
-Like Python, Numba is providing its best speedup on code with raw loops, which
-is not encouraging programmers to express their code using the higher-level
-construct available with Numpy. However, when input constraint are met, it
+Like Cython, Numba achieves its best speedups on codes with raw loops, which
+does not encourage programmers to express their codes using the higher-level
+constructs available within Numpy. However, when input constraints are met, it
 exhibits performance close to Cython, or even better when conversion to GPU
@@
 such as operator fusion, it encourages programmer to express their code using
-the higher-level construct available with Numpy.
+the higher-level constructs available within Numpy.
 
@@
 Copperhead~\cite{copperhead2011} is a functional, data parallel language
-embedded in Python. It uses $n$-uplets, Numpy arrays and lists as its core data
-structure and prohibits usage of many control-flow operators such as loops,
+embedded in Python. It accepts $n$-uplets, Numpy arrays and lists as its core data
+structures and prohibits usage of many control-flow operators such as loops,
 enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
@@
 of the Numpy API in RPython, the language used by PyPy.
+
+\subsection{The Pythran Compiler}
+
+The ideas described in this paper are implemented in the Pythran
+compiler~\cite{pythran2013}, a translator from a subset of Python to
+C++11~\cite{isocxx11}. The architecture of the compiler is briefly described
+in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
+the subset accepted by the compiler. Pythran translates it in its internal representation, a
+simplified Python AST. It performs various optimizations then outputs either
+Python code, in a source-to-source fashion, or C++ templated code calling the pythonic library that typically implements the \texttt{ndarray}
+interface. User annotations can be used to instantiate this code for the proper
+types and generate a native library. This library relies on Boost.Python
+library to match Python's C API.
+
+The Pythran compiler is an open source project publicly released under the BSD
+license\footnote{\url{http://pythonhosted.org/pythran/}}.
+
+\begin{figure}
+
+    \centering
+    \begin{tikzpicture}[
+            file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
+            node distance=0.7cm},
+            tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
+        distance=0.7cm}]
+        \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
+        \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
+        \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
+        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{c++}};
+        \node (empty) [xshift=-1em, left=of gxx] {};
+        \node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic}}};
+        \node[file] (annotation)     [above=of pythonic] {\textbf{Type Info}};
+        \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
+        \node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
+
+        \draw[very thick, ->] (python) -- (pythran);
+        \draw[very thick] (annotation) -| (empty.center);
+        \draw[very thick, ->] (pythran) -- (meta-cxx);
+        \draw[very thick, ->] (meta-cxx) -- (gxx);
+        \draw[very thick] (boost) -| (empty.center);
+        \draw[very thick, ->] (pythonic) -- (gxx);
+        \draw[very thick, ->] (gxx) -- (so);
+    \end{tikzpicture}
+
+    \caption{Pythran compilation flow.}
+    \label{fig:pythran-compiler}
+
+\end{figure}
+
+Pythran keeps the static compilation approach used by Cython---JIT compilation
+is mainly used to statically type kernels before code generation by Numba and
+parakeet. Unlike Cython, it maintains full Python backward compatibility.
+Following the Parakeet approach, it focuses on high-level constructs, while
+still generating efficient code for explicit loop and subscripts.
 
@@
 There is no need to reinvent the wheel when compiling numerical Python kernels.
-A profitable approach is to map Python high-level to existing ones, or
-translate them into existing them.\todo{Mehdi: the previous sentence makes no sense to me}
+A profitable approach is to map Python high-level functions to existing ones.
 As a consequence, this section focuses on the translation of Python constructs
-to efficient C++ constructs. The choice of
-C++ instead of C is motivated by the opportunities offered by meta programming
-and the interesting similarities between C++11~\cite{isocxx11} features and
-Python constructs.
-The meta-programming part of C++ makes it more than a traditional compiler
-runtime, closer to a code generator backend.
+to efficient C++ constructs. The choice of C++ instead of C is motivated by the
+opportunities offered by meta programming and the interesting similarities
+between C++11~\cite{isocxx11} features and Python constructs.  The
+meta-programming part of C++ makes it more than a traditional compiler runtime,
+closer to a code generator back-end.
 
@@
 templates}~\cite{expression_templates, et2012} as a way to efficiently describe high
-level matrix expressions while avoid the creation of intermediate storage.
+level matrix expressions while avoiding the creation of intermediate storage.
 Additionally, it has the same effect as loop fusion, solving the issues
@@
 evaluation. This is always legal in the context of Numpy because none of the
-computations considered for expression template conversion have side effect.
+computations considered for expression template conversion have side effects.
 
@@
 subscript. Outgoing edges (1) and (2) can be removed because they are
-outgoing edges of \texttt{bar} starting by \texttt{n}. Incoming edge (0) ends
+outgoing edges of \texttt{bar} starting by \texttt{n}, which is already locally written by bar. Incoming edge (0) ends
 by \texttt{n} that is written by \texttt{bar}, so it propagates the write
 effects to the source node corresponding to the function \texttt{root}. After
-removing this edge, the only remaining edges are propagating read and the
-analysis ands. The analysis informs that both \texttt{root} update \texttt{r}
-and \texttt{bar} update \texttt{n}, but \texttt{foo} doesn't update its
+removing this edge, the only remaining edges are propagating reads and the
+analysis ends. The analysis informs that \texttt{root} updates \texttt{r}
+and \texttt{bar} updates \texttt{n}, but \texttt{foo} doesn't update its
 argument.
@@
             \draw[->] (bar) -- node[right] {$(\texttt{n,s})$} node[left] {(2)} (foo);
-            \draw[->] (bar.east) node[right, yshift=-1em] {(1)} -| ($(bar.east) + (.5,.75)$) node[right, yshift=-.8em] {$(\texttt{n,n})$} -| (bar.north) ;
+            \draw[->] (bar.east) node[right, yshift=-1em] {(1)} -| ($(bar.east) + (.5,.75)$) node[right, yshift=-.8em] {$(\texttt{n,n})$} -| ($(bar.north) + (.5,0)$) ;
             \draw[->] (root) -- node[right, yshift=.8em] {$(\texttt{r,n})$} node[left, yshift=.8em] {(0)} (bar);
@@
 
-A naive translation of such call to a generic C++ function would generate a
+A naive translation of such a call would generate a
 call to a C++ function that would allocate enough memory to hold three integers,
@@
 suboptimal C++ translation observed for Listing~\ref{lst:stat}
-and~\ref{lst:dyna}.
+and~\ref{lst:dyna} only.
 
@@
 component of \texttt{from}. In that case, all loops can be made parallel, and
-the innermost loop is eventually vectorizable.
+the innermost loop may be vectorized.
 
@@
 
-A conservative approach is to only start a parallel region if the loop trip
-count is large enough. We avoid by using the \texttt{if} clause for parallel
-regions to do so in order to prevent the unnecessary overhead. Instead a manual
-if is used: the loop statement is duplicated and guarded by the appropriate
-check, where only one loop holds the parallel annotations.
-\todo{this paragraph seems to be about a particular runtime issue, this shouldn't matter.
-Moreover, how is it relevant to this paper? It is an "implementation detail"}
+A conservative approach is to start a parallel region if the loop trip
+count is large enough. We avoid using the \texttt{if} clause for parallel
+regions in order to prevent the unnecessary overhead. Instead a manual
+\texttt{if} is used: the loop statement is duplicated and guarded by the appropriate
+check, and only one loop holds the parallel annotations.
+%\todo{this paragraph seems to be about a particular runtime issue, this shouldn't matter.
+%Moreover, how is it relevant to this paper? It is an "implementation detail"}
 
 Note that this parallelization scheme is compatible with the manual
-parallelization of Python programs proposed in~\cite{pyhpc2013}.
-\todo{I think a specific section could be made on this topic}
+parallelization of Python programs using OpenMP directives ported to the Python
+language as proposed in~\cite{pyhpc2013}.
+%\todo{I think a specific section could be made on this topic} SG: I lack time
 
@@
 
-\subsection{The Pythran Compiler}
-
-The ideas described in this paper are implemented in the Pythran
-compiler~\cite{pythran2013}, a translator from a subset of Python to
-C++11~\cite{isocxx11}. The architecture of the compiler is briefly described
-in Figure~\ref{fig:pythran-compiler}. The input is a Python module written in
-the Pythran subset. Pythran translates it in its internal representation, a
-simplified Python AST. It performs various optimization then outputs either
-Python code, in a source-to-source fashion, or C++ template code that makes
-call to the pythonic library that typically implements the \texttt{ndarray}
-interface. User annotations can be used to instantiate this code for the proper
-types and generate a native library. This library relies on Boost.Python
-library to match Python's C API.
-
-The Pythran compiler is an open source project publicly released under the BSD
-license~\footnote{\url{http://pythonhosted.org/pythran/}}.
-
-\begin{figure}
-
-    \centering
-    \begin{tikzpicture}[
-            file/.style={draw=black!50,fill=black!10,rectangle, drop shadow, align=center,
-            node distance=0.7cm},
-            tool/.style={draw=black!50,fill=black!10,ellipse, align=center, node
-        distance=0.7cm}]
-        \node[file] (python) {\textbf{Python Module [\texttt{.py}]}};
-        \node[tool] (pythran) [below=of python] {\textbf{Pythran}};
-        \node[file] (meta-cxx) [below=of pythran] {\textbf{C++} Meta Program};
-        \node[tool] (gxx) [yshift=-1.5em, below=of meta-cxx] {\textbf{c++}};
-        \node (empty) [xshift=-1em, left=of gxx] {};
-        \node[file] (pythonic) [left=of empty] {\textbf{\texttt{pythonic}}};
-        \node[file] (annotation)     [above=of pythonic] {\textbf{Type Info}};
-        \node[file] (boost) [below=of pythonic] {\textbf{\texttt{boost::python}}};
-        \node[file] (so) [yshift=-1.5em, below=of gxx] {\textbf{Native Module [\texttt{.so}]}};
-
-        \draw[very thick, ->] (python) -- (pythran);
-        \draw[very thick] (annotation) -| (empty.center);
-        \draw[very thick, ->] (pythran) -- (meta-cxx);
-        \draw[very thick, ->] (meta-cxx) -- (gxx);
-        \draw[very thick] (boost) -| (empty.center);
-        \draw[very thick, ->] (pythonic) -- (gxx);
-        \draw[very thick, ->] (gxx) -- (so);
-    \end{tikzpicture}
-
-    \caption{Pythran compilation flow.}
-    \label{fig:pythran-compiler}
-
-\end{figure}
-
 \subsection{Experimental Setup}
@@
 suite to evaluate the optimizations described in\S~\ref{sec:pythran}.
-Pythran is used (commit: 488b7ba) with and without SIMD optimization. 
+Pythran (commit: 488b7ba) is used  with and without SIMD optimization.
 The results are compared with Pypy (commit: 632c832bd32f), Parakeet (commit: 4e55810),
@@
 
-numpy-benchmarks run 11 times benchmarks with an execution time around 1 seconde.
-THe best time, average time and standard deviation is return from these 11 runs.
-
-Benchmarking is done on two architectures with different backend compilers.
+The evaluation process consists in  11 runs for each benchmark. The execution
+time of a single run is around 1 second.  The best time, average time and
+standard deviation are computed from these sequences.
+
+Benchmarking is done on two architectures with different back-end compilers.
 
@@
 up to SSE4. We refer to this configuration as \textbf{AMD48}.
-On AMD48, the backend compilers is the clang-omp
-version from intel~\footnote{\url{http://clang-omp.github.io/}} to be able to
-benefits from OpenMP improvements. This Clang version is based on Clang 3.4
+On AMD48, the back-end compiler is the clang-omp
+version from Intel\footnote{\url{http://clang-omp.github.io/}} to be able to
+benefit from OpenMP improvements. This Clang version is based on Clang 3.4
 release (LLVM:233b1e3f034, clang-omp:6f358bcd87).
 
-The second one holds one Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
-8 cores (multithreading is not used). Each node can access to 64\,KB of L1
+The second configuration holds one Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a total of
+8 cores (multithreading is not used). Each node has access to 64\,KB of L1
 cache, 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3 cache
@@
 % For performance comparison between different Pythran version
-To compare the different optimization schemes, we run the numpy-benchmarks test
-suite multiple times on the \textit{rosen} benchmark. The algorithm is the one
-on Listing~\ref{lst:rosen} and input is a raw array of 1,000,000 double elements.
+To compare the different optimization schemes proposed in \S~\ref{sec:pythran} and their effect with respect to the leading example from \S~\ref{sec:optimize}, we evaluated different optimization combination for Listing~\ref{lst:rosen}.
+The input is a raw array of 1,000,000 single precision floating point elements.
 Figures~\ref{fig:rosenamd48} and \ref{fig:rosenintel8} illustrate the results.
-The speedup difference between the two architectures can be explain by AVX support on
+The speedup difference between the two architectures are explained by AVX support on
 the Intel8 architecture while AMD48 supports up to SSE4 only.
-We can also notice importance of forward substitution on this benchmarks as
-we avoid extra loop, 2 temporary array assignment and avoid some load/store for
+We can also notice the importance of forward substitution on this benchmark as
+we avoid one extra loop, two temporary array assignments and some load/store for
 SIMD instructions.
@@
 
+\subsection{Benefits of Implicit Parallelism}
+
+The \texttt{arc\_distance} benchmark is a typical example for which we can make
+intensive use of the techniques described in this paper. This kernel performs
+intensive points to points computation and trigonometric use.
+Figures~\ref{fig:arc_distance} and~\ref{fig:arc_distance_intel8} show speedups
+compared to one thread execution. We can see that OpenMP improvement is
+scalable up to 20 threads with a raw array of 10,000,000 double precision
+floating point elements as input.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
+\caption{Speed up for the \texttt{arc\_distance} kernel on AMD48.}
+\label{fig:arc_distance}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
+\caption{Speed up for the \texttt{arc\_distance} kernel on Intel48.}
+\label{fig:arc_distance_intel8}
+\end{figure}
+
+%Finally, Figure~\ref{fig:l2norme} and \ref{fig:l2norm_intel8} illustrate the
+%execution time for the \texttt{l2norm} kernel introduced in
+%\S~\ref{sec:stackoverflow}. Comparison with other compilers shows that the
+%approach taken by Pythran is relevant for these kind of functions. Just like
+%for the \texttt{arc\_distance} benchmark, we notice the benefits of using the
+%AVX instruction set when available.
+%
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
+%\caption{Execution time of the l2norm kernel on AMD48.}
+%\label{fig:l2norme}
+%\end{figure}
+%
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
+%\caption{Execution time of the l2norm kernel on Intel8.}
+%\label{fig:l2norm_intel8}
+%\end{figure}
+
 \subsection{Comparison with others Python compiler}
 
-Table~\ref{tbl:overview} presents the average execution time for each kernel
-and each compiler, with the lowest values highlighted in
-bold font. Zeros are used when the given compiler failed to
-compile the kernel. Concerning the Numba compiler, this
-benchmarks conﬁrms that it works best on explicit loops. It
-gives non performance gain when run on high-level kernels,
-either falling back to the generation of calls to Python C
-API, or stopping compilation. PyPy is the most robust of
-the compilers, but to a few exceptions, it does not generate
-code more efﬁcient than the original version. This reﬂects
-the fact that Numpy support is still a work-in-progress task.
-Parakeet performs extremely well on high-level benchmarks.
-Although it does not supports some Numpy constructs, it
-successfully parallelize many of them, and also generates
-efﬁcient code for explicit loops. Pythran still performs better
-than Parakeet on some kernels, whenever it successfully
-vectorizes operations.
-
-\begin{table*}
+\begin{table*}[t]
 
@@
 
-Figure~\ref{fig:arc_distance} and \ref{fig:arc_distance_intel8} are typicals
-example where we can make intensive
-use of the techniques described in this paper. It shows speedup compare to one
-thread execution for the arc\_distance benchmark.This kernel performs intensive
-points to points computation and trigonometrics use. We can see that OpenMP
-improvement is scalable up to 20 threads with a raw array of 10,000,000 double
-elements as input.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{arc_distance_amd48.png}
-\caption{Speed up for the arc distance kernel on AMD48.}
-\label{fig:arc_distance}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{arc_distance_intel8.png}
-\caption{Speed up for the arc distance kernel on Intel48.}
-\label{fig:arc_distance_intel8}
-\end{figure}
-
-Finally, Figure~\ref{fig:l2norme} and \ref{fig:l2norm_intel8} shows the L2norm algorithm introduced in
-\S~\ref{sec:stackoverflow}. Comparison with other compilers illustrates how the
-approach taken by Pythran is relevant for these kind of functions. Just like for
-the arc\_distance benchmarks, we can notice use of AVX when it is available.
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{l2norm_amd48.png}
-\caption{Execution time of the l2norm kernel on AMD48.}
-\label{fig:l2norme}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.5\textwidth]{l2norm_intel8.png}
-\caption{Execution time of the l2norm kernel on Intel8.}
-\label{fig:l2norm_intel8}
-\end{figure}
+Table~\ref{tbl:overview} presents the average execution time for each kernel
+and each compiler, with the lowest values highlighted in
+bold font. Zeros are used when the given compiler failed to
+compile the kernel. Concerning the Numba compiler, this
+benchmarks confirms that it works best on explicit loops. It
+gives no performance gain when run on high-level kernels,
+either falling back to the generation of calls to Python C
+API, or stopping compilation. PyPy is the most robust of
+the compilers, but to a few exceptions, it does not generate
+code more efficient than the original version. This reflects
+the fact that Numpy support is still a work-in-progress.
+Parakeet performs extremely well on high-level benchmarks.
+Although it does not supports some Numpy constructs, it
+successfully parallelizes many of them, and also generates
+efficient code for explicit loops. Pythran still performs better
+than Parakeet on some kernels, whenever it successfully
+vectorizes operations.
 
@@
 
+The authors are very grateful to Béatrice Creusillet for her careful proof reading.
+
 % We recommend abbrvnat bibliography style.

### 2014-09-05 doc/papers/cgo2014/paper.tex
Additional rework of the paper

@@
 the Python compiler level. This optimization leverages the
-\textit{def-use} chains as well as interprocedural read-write
+\textit{def-use} chains as well as inter-procedural read-write
 effects and alias analysis presented in the next sections.
-Thanks to these analyses, the forward substitution implementation is quite
-simple: \todo{it is not THAT simple indeed, there is "lazyness" analysis
-involved}....
-
-This optimization is able to forward \texttt{t0} and \texttt{t1} in
+It is able to forward \texttt{t0} and \texttt{t1} in
 Listing~\ref{lst:rosen}, avoiding a temporary allocation and providing more
@@
 Python that forbid this kind of dynamic patching.
-In this context, it becomes possible to design an intra-procedural alias
+In this context, it becomes possible to design an intra procedural alias
 analysis thanks to the absence of pointer arithmetic:
@@
 
-\subsection{Interprocedural Effects Analysis}
+\subsection{Inter-procedural Effects Analysis}
 \label{sec:effects}
@@
 
-To interprocedurally compute the read-write effects, intra procedural effects
+To inter-procedurally compute the read-write effects, intra procedural effects
 are first computed. Then we create a directed graph $G$ with as many nodes as
@@
 sink. The analysis is iteratively pruning all edges from $G$ by considering any
-node for which the intraprocedural analysis shown that an argument $i$ is
+node for which the intra procedural analysis shown that an argument $i$ is
 written. Since the parameter is modified, any outgoing edge $(i, j)$ for this
@@
 call to \texttt{range(3)} into \texttt{[0, 1, 2]}, but also to compute the
-value of \texttt{fibonnaci(4)} at compile-time.
+value of \texttt{Fibonacci(4)} at compile-time.
 
@@
 
-Once interprocedural constant evaluation is done (eventually followed by
+Once inter-procedural constant evaluation is done (eventually followed by
 forward substitution), iteration over fixed size lists may be exposed, giving an
@@
 \hline
-         allpairs\_distances &   58293 &             0 &          3170 &   59232 & \textbf{1583} \\
+         allpairs\_distances &   58293 &             N/A &          3170 &   59232 & \textbf{1583} \\
   allpairs\_distances\_loops &   64137 &         72748 &          3158 &   65287 & \textbf{3146} \\
                arc\_distance &    2796 &          2792 &          3084 &    2791 &  \textbf{103} \\
-                        conv & 3260079 &             0 & \textbf{3284} & 3268298 &          3420 \\
-                create\_grid &    5231 & \textbf{5044} &             0 &    5240 &          5587 \\
-                    cronbach &    1905 &          2016 &             0 &    1909 & \textbf{1890} \\
+                        conv & 3260079 &             N/A & \textbf{3284} & 3268298 &          3420 \\
+                create\_grid &    5231 & \textbf{5044} &             N/A &    5240 &          5587 \\
+                    cronbach &    1905 &          2016 &             N/A &    1909 & \textbf{1890} \\
                    diffusion &   42998 &         45030 & \textbf{8843} &   43238 &         21704 \\
-                      evolve &    8236 &             0 &             0 &    8319 & \textbf{6242} \\
-                        fdtd & 1940884 &       1498849 &             0 & 2007839 & \textbf{2294} \\
-                         fft &   27120 &             0 &             0 &   27009 & \textbf{2151} \\
-                    grouping &    3739 &             0 &             0 &    3740 & \textbf{1747} \\
+                      evolve &    8236 &             N/A &             N/A &    8319 & \textbf{6242} \\
+                        fdtd & 1940884 &       1498849 &             N/A & 2007839 & \textbf{2294} \\
+                         fft &   27120 &             N/A &             N/A &   27009 & \textbf{2151} \\
+                    grouping &    3739 &             N/A &             N/A &    3740 & \textbf{1747} \\
                      growcut & 2112332 &         11617 & \textbf{5485} & 2141571 &         10338 \\
                       harris &    7173 &          9116 & \textbf{4240} &    7322 &          4889 \\
-                     hyantes &  288852 &             0 & \textbf{2834} &  289074 &          3522 \\
+                     hyantes &  288852 &             N/A & \textbf{2834} &  289074 &          3522 \\
                        julia & 1744357 &          5168 &  \textbf{109} & 1771329 &          4255 \\
                       l2norm &    5966 &          5948 &          9210 &    6559 &  \textbf{551} \\
-               local\_maxima &  132237 &             0 &             0 &  132047 & \textbf{8561} \\
-                      lstsqr &    9254 &          7528 &             0 &    8992 & \textbf{1414} \\
-                      mandel &  951612 & \textbf{8175} &             0 &  951687 &         10341 \\
+               local\_maxima &  132237 &             N/A &             N/A &  132047 & \textbf{8561} \\
+                      lstsqr &    9254 &          7528 &             N/A &    8992 & \textbf{1414} \\
+                      mandel &  951612 & \textbf{8175} &             N/A &  951687 &         10341 \\
                multiple\_sum &    5477 &          5813 &          9805 &    5475 & \textbf{3503} \\
                     pairwise & 3362150 &         10764 &          6199 & 3347574 & \textbf{6015} \\
-              periodic\_dist &    2559 &          2675 &             0 &    2562 & \textbf{2292} \\
-                   repeating &    2808 &          2836 &             0 &    2795 &  \textbf{810} \\
-             reverse\_cumsum &    4583 &          4828 &             0 &    4583 & \textbf{2379} \\
+              periodic\_dist &    2559 &          2675 &             N/A &    2562 & \textbf{2292} \\
+                   repeating &    2808 &          2836 &             N/A &    2795 &  \textbf{810} \\
+             reverse\_cumsum &    4583 &          4828 &             N/A &    4583 & \textbf{2379} \\
                        rosen &   15382 &         14309 &          2763 &   16065 &  \textbf{939} \\
-                   slowparts &    8302 &             0 & \textbf{1610} &    8241 &          1927 \\
+                   slowparts &    8302 &             N/A & \textbf{1610} &    8241 &          1927 \\
                    smoothing &  834057 &          8478 &          9771 &  819827 & \textbf{8254} \\
@@
                 vibr\_energy &    5277 &          5275 &          5334 &    5416 &  \textbf{146} \\
-                        wave &   72394 &         67590 &             0 &   72300 & \textbf{5186} \\
+                        wave &   72394 &         67590 &             N/A &   72300 & \textbf{5186} \\
                        wdist &  108572 &        119211 &          4197 &  108283 & \textbf{4010} \\
@@
 }
-\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
+\caption{Average execution times for all kernels of the numpy-benchmarks on Intel8 with OpenMP (8 threads) and SIMD extension.}
 \label{tbl:overview}
@@
 Table~\ref{tbl:overview} presents the average execution time for each kernel
-and each compiler, with the lowest values highlighted in
-bold font. Zeros are used when the given compiler failed to
-compile the kernel. Concerning the Numba compiler, this
-benchmarks confirms that it works best on explicit loops. It
-gives no performance gain when run on high-level kernels,
-either falling back to the generation of calls to Python C
-API, or stopping compilation. PyPy is the most robust of
-the compilers, but to a few exceptions, it does not generate
-code more efficient than the original version. This reflects
-the fact that Numpy support is still a work-in-progress.
-Parakeet performs extremely well on high-level benchmarks.
-Although it does not supports some Numpy constructs, it
-successfully parallelizes many of them, and also generates
-efficient code for explicit loops. Pythran still performs better
-than Parakeet on some kernels, whenever it successfully
-vectorizes operations.
+and each compiler, with the lowest values highlighted in bold font. N/A are
+used when the given compiler failed to compile the kernel. Standard deviations
+are not presented in the table, as the result computation were very stable, to
+the notable exception of \texttt{periodic\_dist} for parakeet. It comes out
+that parakeet leaks the memory of some of the \texttt{ndarray} it allocates,
+which results in the kernel swapping.
+
+Concerning the Numba compiler, this benchmarks confirms that it works best on
+explicit loops (benchmarks \texttt{growcut}, \texttt{mandel},
+\texttt{pariwise}). It does not show any significant improvements on high-level
+kernels (\texttt{l2norm}, \texttt{multiple\_sum}\dots), either falling back to
+the generation of calls to Python C API, or stopping compilation.
+
+PyPy is the most robust of the compilers, but to a few exceptions, it does not
+generate code more efficient than the original version. This reflects the fact
+that Numpy support is still a work-in-progress.
+
+Parakeet performs extremely well on both high-level benchmarks and explicit
+loops. Although it does not supports some Numpy constructs (e.g. it does no
+support calling the \texttt{numpy.cumsum} function in \texttt{reverse\_cumsum})
+, it successfully parallelizes many of them. It does not vectorize them though,
+which explains while Pythran still performs better by more than a factor of
+$\times2$ on some kernels.
+
+Pythran outperforms the other compilers, sometimes by a factor greater than
+$\times2$, thanks to the combination of vectorization and parallelization. Code
+generation for explicit loops can still be improved, as shown by the poor
+performance on the \texttt{growcut} kernel and \texttt{hyantes} kernels.
 
@@
 
+This paper presents a study of the optimization of Python/Numpy high level
+kernels in the context of high performance computing. It uses a set of
+realistic benchmarks adapted from the question and answers on the StackOverflow
+website to identify optimization opportunities. It focuses on the efficient
+implementation and optimization of array expressions within the ahead-of-time
+Pythran compiler, showing that a compilation step at Python level before
+generation of lower-level code makes it possible to generate vectorized,
+parallel C++ code. A comparison with existing JIT compiler for scientific
+Python validates the approach, showing significant performance improvements
+over the state of the art.
+
 \acks
 
-The Pythran project is an independent Open Source initiative that has been
-partially funded by the CARP Project and the SILKAN Company.
+The Pythran project is an independent Open Source initiative partially funded
+by the CARP Project and the SILKAN Company.
 

### 2014-09-05 doc/papers/cgo2014/paper.tex
Anonymize the paper

@@
 
-\authorinfo{Serge Guelton}
-           {T{\'e}l{\'e}com Bretagne}
-           {serge.guelton@telecom-bretagne.eu}
-\authorinfo{Pierrick Brunet}
-           {INRIA/MOAIS}
-           {pierrick.brunet@inria.fr}
-\authorinfo{Mehdi Amini}
-           {SILKAN}
-           {mehdi.amini@silkan.com}
+\authorinfo{\phantom{Serge Guelton}}
+{\phantom{T{\'e}l{\'e}com Bretagne}}
+{\phantom{serge.guelton@telecom-bretagne.eu}}
+\authorinfo{\phantom{Pierrick Brunet}}
+{\phantom{INRIA/MOAIS}}
+{\phantom{pierrick.brunet@inria.fr}}
+\authorinfo{\phantom{Mehdi Amini}}
+{\phantom{SILKAN}}
+{\phantom{mehdi.amini@silkan.com}}
 

### 2014-12-12 doc/papers/hpsl2015/paper.tex
Quick'n dirty extended abstract for http://polaris.cs.uiuc.edu/hpsl/

\documentclass[10pt, preprint, nocopyrightspace]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{todo}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\lstset{captionpos=b, float, language=python}

\usepackage{tikz}
\usetikzlibrary{positioning}
\usetikzlibrary{shadows}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes}
\usetikzlibrary{calc}

\providecommand{\boostsimd}{\textsc{Boost.SIMD}}
\providecommand{\cpp}[1][~]{\textsc{C++}#1}
\providecommand{\ie}[1][~]{\textit{i.e.}#1}
\providecommand{\eg}[1][~]{\textit{e.g.#1}}

\begin{document}

\title{Efficient Compilation of High Level Python Numerical Programs with Pythran}

\authorinfo{Serge Guelton}
{T{\'e}l{\'e}com Bretagne}
{serge.guelton@telecom-bretagne.eu}
\authorinfo{Pierrick Brunet}
{INRIA/MOAIS}
{pierrick.brunet@inria.fr}
\authorinfo{Mehdi Amini}
{Apple}
{mehdi.amini@apple.com}

\maketitle

%\begin{abstract}
%
%    For decades, FORTRAN and C(++) have dominated the landscape of High
%    Performance Computing languages, leaving interpreted language like Matlab,
%    R, Python or more recently Julia for experimentation and prototyping.
%
%    As more scientists use these scripting languages, there is an on-going
%    research effort to compile them into efficient native code. However these
%    languages favor a high-level coding style, so the computation kernels are
%    different from those used in traditional C or Fortran programs:
%    explicit raw loops are replaced by high-level constructs and calls to the
%    relevant toolbox/module/library/package. As a consequence, traditional
%    algorithms to perform loop optimization techniques like fusion, tiling and
%    related vectorization or parallelization techniques are not directly
%    applicable.
%
%    This papers focuses on the case of the Python language and especially the
%    Numpy package that provides core array data structure and basic linear
%    algebra routines. It first conducts a case study based on Numpy use
%    cases from the StackOverflow questions and answers website, then studies
%    existing compilers for numerical Python. Optimization opportunities,
%    including parallelization and vectorization, and their implementation in
%    the Pythran compiler are presented and illustrated through several
%    benchmarks, showing very interesting speedups over the standard Python
%    interpreter. Speedup greater than a factor of 10 are achieved despite the
%    fact that the considered benchmarks mostly call Numpy routines that run C code.
%
%\end{abstract}
%
%
%\keywords
%Static Compilation, Parallelization, Python, C++, SIMD, Runtime

%%
%%
\section*{Abstract}
% On aura pas le temps pour le papier je pense mais faire une comparaison avec
% du C pure pour voir ce que les compilo de C arrive a faire.

The Python language~\cite{rossum97} has a rich ecosystem that now provides a
full toolkit to carry out scientific experiments, from core scientific routines
with the Numpy package\cite{oliphant2007,numpyarray2011}, to scientific
packages with Scipy, plotting facilities with the
Matplotlib package, enhanced terminal and notebooks with IPython. As a
consequence, there has been a move from historical languages like Fortran to
Python, as showcased by the success of the Scipy conference.

As Python based scientific tools get widely used, the question of High
performance Computing naturally arises, and it is the focus of many recent
research. Indeed, although there is a great gain in productivity when using
these tools, there is also a performance gap that needs to be filled.

This extended abstract focuses on compilation techniques that are relevant for the
optimization of high-level numerical kernels written in Python using the Numpy
package, illustrated on a simple kernel.

%% Its introduces the implementation of these techniques in the Pythran
%% compiler and provides a benchmark test suite for performance measures along
%% several Python compilers.
%% Section~\ref{sec:stackoverflow} presents a case study based on user
%% input gathered from the StackOverflow questions and answers website.
%% Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
%% showcases the optimization opportunities typically found in Numpy-based
%% kernels. Existing compilation approaches to optimize such kernels are discussed
%% in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on
%% compiler-runtime cooperation to solve the performance issues.
%% Finally, experimentations showcasing important performance gains are detailed in
%% Section~\ref{sec:xp}.

%%% %%
%%% %%
%%% \section{A Crowd-Sourced Numpy Benchmark}
%%% \label{sec:stackoverflow}
%%% 
%%% One of the first steps required when designing an optimizing compiler is to
%%% gather enough test cases, or benchmarks, in order to drive the optimization
%%% process toward realistic examples. To achieve this goal, we have gathered a few
%%% examples from existing compilers' test suites, but these may have been modified to
%%% overcome some compiler limitations. A less biased source of benchmarks should
%%% be independent from the compiler community and showed focus on matching user needs.
%%% To achieve that goal, we collected synthetic scientific kernels from the
%%% StackOverflow website.
%%% 
%%% StackOverflow\footnote{\url{http://stackoverflow.com/}} is a question and
%%% answer site for programmers. Any registered user can submit questions and other
%%% users may answer. A voting system is then used to sort the answers. Said
%%% otherwise, the website provides a database of questions and answers on
%%% programming topics, including Python.
%%% 
%%% A query using the terms \emph{numpy} and \emph{slow} yields a lot of results, many
%%% of which expose the pattern ``Q: my code is slow'' followed by an upvoted answer
%%% ``A: rewrite it that way'', where the original code performs explicit looping
%%% over an array, and the rewritten code uses the relevant combination of Numpy
%%% function.
%%% 
%%% For instance, question \href{http://stackoverflow.com/questions/7741878}{7741878}
%%% proposes to use an explicit loop to iteratively compute the $L^2$ norm of each
%%% row of a matrix, as illustrated in Listing~\ref{lst:l2norm-loopy}, and one of
%%% the proposed answer is to use the more compact version presented in
%%% Listing~\ref{lst:l2norm}. This answer no longer uses any explicit loop and
%%% roughly achieves a $\times 4$ speedup over the loop version.
%%% 
%%% \begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in Numpy.}, label={lst:l2norm-loopy}]
%%% def l2norm_slow(x):
%%%     r = np.empty(x.shape[0])
%%%     for i in xrange(x.shape[0]):
%%%         r[i] = np.sum(np.abs(x[i])**2)
%%%     return r
%%% \end{lstlisting}
%%% 
%%% \begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in Numpy.}, label={lst:l2norm}]
%%% import numpy as np
%%% def l2norm_fast(x):
%%%     return np.sum(np.abs(x)**2, axis=1)
%%% \end{lstlisting}
%%% 
%%% Newcomers to Python with a prior C or Fortran experience are likely to make
%%% use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
%%% knowledge of the Numpy API and of good programming practices grows, they get to 
%%% write code similar to Listing~\ref{lst:l2norm}. 
%%% The fact that Numpy high-level constructs runs faster is also a strong 
%%% motivation. As a direct consequence, a compiler should focus on optimizing
%%% high-level constructs rather than raw
%%% loops. Both to encourage good practice with respect to Numpy code, and because
%%% this is the kind of code Numpy users ultimately write.
%%% 
%%% A selection of benchmarks collected during this process, selected for their
%%% conciseness and representativity of high-level calls, are publicly available on
%%% \url{https://github.com/serge-sans-paille/numpy-benchmarks}, with a set of
%%% benchmarking routines specialized for major Python/Numpy compilers. This
%%% benchmark is used in \S~\ref{sec:xp} to evaluate the validity of the approach
%%% described in this paper.

%%
%%
\section{Optimizations Opportunities in a Typical Numpy Kernel}
\label{sec:optimize}

This section briefly presents the core concepts of the Numpy package, then goes
through all the optimization opportunities in a small kernel as a showcase of
the optimization opportunities.

\subsection{Numpy}

The reference implementation of Numpy is a native Module, written mostly in C.
It uses the BLAS API whenever possible and provides a relatively efficient array
abstraction in the form of the \texttt{ndarray} data structure.

%% This data
%% structures makes it possible to use various styles of slicing and broadcasting,
%% as illustrated in Listing~\ref{lst:pairwise}.

It also enforces a high-level programming style but it's very inefficient when
explicit subscripts are used.

%% For instance, raw loops are used in
%% Listing~\ref{lst:pairwise-raw}. It avoids the creation of temporary arrays but
%% runs almost 100 times slower due to the Python interpreter overhead and the C
%% to Python conversion involved in every subscript.
%% 
%% \begin{lstlisting}[language=python, caption={\textit{pairwise} function that
%%   exhibits array broadcasting.}, label={lst:pairwise}, breaklines=true]]
%% def pairwise_numpy(X):
%%     return np.sqrt(((X[:, None, :] - X) ** 2).sum(-1))
%% \end{lstlisting}
%% 
%% 
%% \begin{lstlisting}[language=python, caption={\textit{pairwise} function
%%   using raw loops.}, label={lst:pairwise-raw}]
%% def pairwise_python(X):
%%     M = X.shape[0]
%%     N = X.shape[1]
%%     D = np.empty((M, M), dtype=np.float)
%%     for i in range(M):
%%         for j in range(M):
%%             d = 0.0
%%             for k in range(N):
%%                 tmp = X[i, k] - X[j, k]
%%                 d += tmp * tmp
%%             D[i, j] = np.sqrt(d)
%%     return D
%% \end{lstlisting}

\subsection{The Rosenbrock Function}

We use the kernel illustrated in Listing~\ref{lst:rosen} and adapted from the
Scipy source code of \texttt{scipy.optimize.rosen} as a leading example. It
uses Numpy's \texttt{sum} function, Python's square notation and Numpy's array
slicing. It is a good example of high-level Python kernel, although using the
original function directly would naturally make sense.

Note that due to dynamic typing, this function can take arrays of different
shapes and types as input.

\begin{lstlisting}[language=python, caption={High-level implementation of the Rosenbrock function in Numpy.}, label={lst:rosen}]
def rosen(x):
    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
    t1 = (1 - x[:-1]) ** 2
    return numpy.sum(t0 + t1)
\end{lstlisting}

\subsection{Temporaries Elimination}
\label{sec:temporaries-elimination}

In Numpy, any point-to-point array operation allocates a new array that holds
the computation result. This behavior is consistent with many Python standard
module, but it is a very inefficient design choice, as it keeps on polluting
the cache with potentially large fresh storage and adds extra
allocation/deallocation operations, that have a very bad caching effect. In the
\textit{rosen} function from Listing~\ref{lst:rosen}, 7 temporary arrays are
allocated (slicing does not create a temporary array but a view) to hold
intermediate steps. Had the expression been evaluated lazily, no temporary
would have been needed.

\subsection{Operator Fusion}
\label{sec:operator-fusion}

As Numpy is a native library mostly written in C, each operator computation is
performed by a function implemented as a loop performing a single operation,
and the operator chaining is done at the interpreter level. This is a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of merging multiple operators into a single specialized
operator. On the contrary, providing many operator combinations as part of the library yields
better performance to the price of API bloat. Listing~\ref{lst:rosen}
illustrates the use of a small set of functions: a loop is generated for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, whereas a single loop would have been necessary with operator fusion.

\subsection{Loop Vectorization and Parallelization}

Without operator fusion, there would be very little benefit to generate SIMD
instructions for the respective array operations used by each operator, as the
memory loads and stores would have dominated the execution time. This is even more important as
Numpy typically operates on double precision floats, which means only two (SSE)
to four (AVX) scalars per vector registers.

Parallelization would also suffer from the lack of operator fusion: a
synchronization fence is needed between each temporary computation.  Hence the
loop computation intensity would be very low compared to the memory pressure
implied by two array reads and one array write for a single binary operator.

On the opposite, if all computations were merged into a single loop using
temporaries elimination and operator fusion, parallelization would be more
effective as barriers between binary operations are no longer needed and loads
and stores are counterbalanced by several operations.

%%
%%
\section{Optimization with the Pythran Compiler}

\subsection{Pythran}

The optimizations presented in the previous section have been implemented in the Pythran
compiler~\cite{pythran2013}, a translator from a subset of Python to
C++11~\cite{isocxx11}. The input Optimization a Python module written in
the subset accepted by the compiler. Pythran translates it in its internal representation, a
simplified Python AST. It performs various optimizations then outputs either
Python code, in a source-to-source fashion, or C++ templated code calling the pythonic library that typically implements the \texttt{ndarray}
interface. User annotations can be used to instantiate this code for the proper
types and generate a native library. This library relies on Boost.Python
library to match Python's C API.

The Pythran compiler is an open source project publicly released under the BSD
license\footnote{\url{http://pythonhosted.org/pythran/}}.

Compared to existing alternatives, Pythran keeps the static compilation
approach used by Cython, when JIT compilation is mainly used to statically type
kernels before code generation in Numba and Parakeet. Unlike Cython, it
maintains full Python backward compatibility.  Following the Parakeet approach,
it focuses on high-level constructs, while still generating efficient code for
explicit loop and subscripts.

%%
%%
\subsection{Experiments} \label{sec:xp}

The experiments are run on an Intel(R) Xeon(R) CPU E5-2650 0 @ 2.00GHz for a
total of 8 cores (multithreading is not used). Each node has access to 64\,KB
of L1 cache, 256\,KB of L2 cache. Both L1 and L2 caches are private, while L3
cache is shared between the 8 cores. This configuration provides a total of
64\,GB of main memory. It supports up to AVX. The backend compiler is GCC 4.9
(20140528) with libgomp.

To compare the different optimization effects, we evaluated different
optimization combination for Listing~\ref{lst:rosen}. The input is a raw array
of 1,000,000 single precision floating point elements.
Figure~\ref{fig:rosenintel8} illustrates the results. We can notice the
importance of forward substitution on this benchmark as we avoid one extra
loop, two temporary array assignments and some load/store for SIMD
instructions. Vectorization is also extreemly profitable.

\begin{figure}[t]
\centering
\includegraphics[width=.5\textwidth]{rosen_intel8.png}
\caption{Execution time of the rosen kernel on Intel Xeon.}
\label{fig:rosenintel8}
\end{figure}

%%
%%
\section*{Conclusion}

This extended abstract presents a short study of the optimization of Python/Numpy high level
kernels in the context of high performance computing. It uses a real-worl synthetic kernel as leading example and focuses on the efficient
implementation and optimization of array expressions within the ahead-of-time
Pythran compiler, showing that a compilation step at Python level before
generation of lower-level code makes it possible to generate vectorized,
parallel C++ code. A comparison with existing JIT compilers for scientific
Python validates the approach, showing significant performance improvements
over the state of the art.

\acks

The Pythran project was partially funded
by the CARP Project and the SILKAN Company.

%Several Students of Télécom Bretagne, namely Adrien Merlini, Alan Raynaud,
%Xavier Corbillon, Yuancheng Peng and Eliott Coyacc, have made significant
%contributions to the project.

% The authors are very grateful to Béatrice Creusillet for her careful proof reading.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{biblio}

\end{document}

### 2017-10-26 docs/conf.py
Read The Docs integration

@@ -14,8 +14,101 @@
 
 import sys
 import os
+import re
 
 from pythran import __version__
+
+with open("../README.rst") as readme:
+    readme_body = readme.read()
+    toc = '''
+
+.. toctree::
+    :maxdepth: 1
+
+
+    MANUAL
+    CLI
+    SUPPORT
+    DEVGUIDE
+    TUTORIAL
+    INTERNAL
+    LICENSE
+    AUTHORS
+    Changelog
+'''
+
+    readme_body = readme_body.replace('http://pythran.readthedocs.io', toc)
+
+    with open("index.rst", "w") as index:
+        index.write(readme_body)
+    del readme_body
+
+with open("../LICENSE") as license:
+    with open('LICENSE.rst', 'w') as license_rst:
+        license_rst.write("=======\nLICENSE\n=======\n\n")
+        license_rst.write(license.read())
+
+with open("../Changelog") as changelog:
+    with open('Changelog.rst', 'w') as changelog_rst:
+        changelog_rst.write('=========\nChangelog\n=========\n\n')
+        changelog_rst.write(changelog.read())
+
+with open("../AUTHORS") as authors:
+    with open('AUTHORS.rst', 'w') as authors_rst:
+        authors_rst.write(authors.read())
+
+def make_support():
+    from pythran import tables
+
+    TITLE = "Supported Modules and Functions"
+
+    DEPTHS = '=*-+:~#.^"`'
+
+    body = []
+
+    body.append(DEPTHS[0]*len(TITLE))
+    body.append(TITLE)
+    body.append(DEPTHS[0]*len(TITLE))
+    body.append("")
+
+
+    def format_name(name):
+        if name.endswith('_') and not name.startswith('_'):
+            name = name[:-1]
+        return name
+
+
+    def isiterable(obj):
+        return hasattr(obj, '__iter__')
+
+
+    def dump_entry(entry_name, entry_value, depth):
+        if isiterable(entry_value):
+            body.append(entry_name)
+            body.append(DEPTHS[depth] * len(entry_name))
+            body.append("")
+            sym_entries, sub_entries = [], []
+            for sym in entry_value:
+                w = sub_entries if isiterable(entry_value[sym]) else sym_entries
+                w.append(sym)
+            for k in sorted(sym_entries):
+                dump_entry(format_name(k), entry_value[k], depth + 1)
+            body.append("")
+            for k in sorted(sub_entries):
+                dump_entry(format_name(k), entry_value[k], depth + 1)
+                body.append("")
+        else:
[…]

### 2017-10-26 docs/papers/sc2013/paper.tex
Read The Docs integration

--- 2013-10-27 doc/papers/sc2013/paper.tex
@@
 
-Another widely used solution is the module IPython.parallel, which supports many
+Another widely used solution is the module \texttt{IPython.parallel}, which supports many
 different styles of parallelism including: single program, multiple data (SPMD)
-parallelism, multiple program, multiple data (MPMD) parallelism. message passing
+parallelism, multiple program, multiple data (MPMD) parallelism, message passing
 using MPI, task farming, and data parallel.
-FIXME: to be continued...
+%FIXME: to be continued...
 
@@
 exclusively of a set of headers. Therefore the generated C++ file is easily
-distribuable and does not depend on Pythran. One could even preprocess this
+distributable and does not depend on Pythran. One could even preprocess this
 C++ file to avoid the need of distributing along the Pythonic++ headers.
@@
 For instance, list comprehension expressions are turned into function calls or
-tuple unpacking is turned in multiple assignments. During this transformation
+tuple unpacking is turned in multiple assignments enclosed in a dummy
+\texttt{if}. During this transformation
 process, an expression is always transformed into a single expression and a
 statement is always transformed into a single statement.
-% FIXME as noted by reviewer 4, the previous sentence is not exact. Just a few
-% lines above we said that "tuple unpacking is turned in multiple assignments"
 Then this AST is converted into a C++ AST meant to be pretty-printed.
@@
 at the Python level. The PEP 3118 defined the API that allows to share
-efficiently the embeded data between Python and the native world without
+efficiently the embedded data between Python and the native world without
 involving any copy. As a tool for scientific computing, Pythran supports such a
@@
 \texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
-the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
-% FIXME reviewer 4 doesn't seem to agree with and/or C++ vs Python
+the \texttt{and} and \texttt{or} keywords are not boolean operators in Python.}
 The latest OpenMP specifications~\cite{openmp4} describe a way to declare
@@
     loop in Pythran with tuple unpacking.}]
-'omp parallel for'
+#omp parallel for
 for i, v in enumerate([2, 3, 5, 7, 11]):
@@
   using = sum = 0
-  'omp parallel for if(using == 1)'
+  #omp parallel for if(using == 1)
   for i in range(loop_count + 1):
@@
 information than Pythran to generate efficient code, and Cython is not
-backward-compatible with Python. They translate explicit fine-grained
-parallelism through code annotations or specific Python constructs,
-respectively.
+backward-compatible with Python (outside of \emph{Pure mode}). They translate
+explicit fine-grained parallelism through code annotations or specific Python
+constructs, respectively.
 
@@
     cdef int i
-    with nogil, parallel():
-        for i in prange(10000000):
+    for i in prange(10000000, nogil=True):
         r += sqrt(i)
@@
 The Python Benchmark test bed
-(\url{https://github.com/numfocus/python-benchmarks}) is an innitiative to
+(\url{https://github.com/numfocus/python-benchmarks}) is an initiative to
 gather relevant numerical python benchmarks and compare several implementations
@@
 Cython, Numba, Pythran and Parakeet, when available and supported.
-Additionnaly, the original Python code is run.
+Additionally, the original Python code is run.
 
@@
 and one with the OpenMP flag set. The code itself is not modified, only the
-compilation flag. A logarithmic scale is used due to the generally poor
-performance of the pure Python version.
+compilation flag and the directives. A logarithmic scale is used due to the
+generally poor performance of the pure Python version.
 
@@
 Numpy's vector operations. There is no explicit parallelism in this function but
-the vector operation is implictly parallel. On this example, Cython is on par
+the vector operation is implicitly parallel. On this example, Cython is on par
 with Pythran and slightly faster than Parakeet, but turning OpenMP on makes the
@@
 step of the GrowCut algorithm used in image processing. On this example,
-Parakeet performs better than Cython and Pytrhan and way better than Numba, but
-paralleizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
+Parakeet performs better than Cython and Pythran and way better than Numba, but
+parallelizing the outermost loop with OpenMP gives a $\times 2.65$ speedup to
 Pythran that helps it outperform Parakeet.
@@
 It appears that all Python compilers manage to yield significant speedups over
-the Python interepreter, each compiler managing some cases better than others.
-All the benchmarks exhibt some level of parallelism that are easy to capture
+the Python interpreter, each compiler managing some cases better than others.
+All the benchmarks exhibit some level of parallelism that are easy to capture
 using OpenMP without changing the original algorithm and while maintaining
 backward compatibility with the Python interpreter. Some even have a degree of
-implicit parallism that makes parallelization fully transparent to the user.
+implicit parallelism that makes parallelization fully transparent to the user.
 Taking advantage of this parallelism appears to be the key to reach an
